---
title: "Time-course differential expression: THP-1 macrophages + LPS/IFN (GSE201354)"
author: "Crystal Shan"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
    fig_width: 8
    fig_height: 5
params:
  quant_dir: "/project/spott/cshan/fiber-seq/macrophage_project/annotations/GSE201354_RNA"
  # Base output directory; tables go to <out_dir>/results, figures to <out_dir>/figures.
  out_dir: "/project/spott/cshan/fiber-seq/macrophage_project/RNA-seq"
  ref_dir: "/project/spott/cshan/fiber-seq/macrophage_project/RNA-seq/reference"
  # Pre-built tx2gene (TXNAME/GENEID/SYMBOL/BIOTYPE). If it exists, GTFs are not needed.
  tx2gene_file: "/project/spott/cshan/fiber-seq/macrophage_project/RNA-seq/reference/tx2gene.tsv"
  gtf_quant_annotation: "/project/spott/cshan/annotations/Homo_sapiens.GRCh37.75.gtf.gz"       
  gtf_current: "/project/spott/cshan/annotations/gencode.v49.chr_patch_hapl_scaff.annotation.gtf.gz"  # GENCODE v49 = Ensembl 115, used only to refresh gene symbols
  alpha: 0.05
  lfc_threshold: 1
  lrt_alpha_for_clustering: 0.01
  n_clusters: 6
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dpi = 150, fig.align = "center")
```

# Overview

Differential expression across a **stimulation time course** of human THP-1-derived
macrophages treated with **LPS + IFN-$\gamma$** and sampled at 0, 0.5, 1, 1.5, 2, 4, 6
and 24 h (8 time points x 2 replicates = 16 samples). Input is the author-provided
Salmon `quant.sf` output from GEO series **GSE201354** (Reed et al. 2022, PMID 36323252).

The pipeline is:

1. Parse a sample sheet from the file names.
2. Build a transcript-to-gene map matched to the annotation the quantification used.
3. `tximport` to gene level, keeping average transcript lengths as DESeq2 offsets.
4. Two DESeq2 models: an **LRT** for any change across the series, and **Wald**
   contrasts of each time point against 0 h with `apeglm` shrinkage.
5. k-means clustering of temporal profiles into waves.
6. Figures and result tables.

Two points that matter for correctness:

- **DE runs on counts, not TPM.** The `TPM` column in `quant.sf` is already length- and
  composition-normalised; feeding it to DESeq2 violates the negative-binomial mean-variance
  relationship. `tximport` passes counts plus an average-transcript-length offset instead.
- **Replicate is a blocking factor.** The replicate labels (`_S_2` / `_S_3`) recur at every
  time point, so this is a blocked design. With n=2 per group, absorbing that term with
  `~ rep + time_f` rather than `~ time_f` materially affects power.

```{r libraries}
suppressPackageStartupMessages({
  library(tximport)
  library(DESeq2)
  library(data.table)
  library(ggplot2)
  library(ggrepel)
  library(patchwork)
  library(matrixStats)
  library(apeglm)
})

fig_dir <- file.path(params$out_dir, "figures")   # .png
tab_dir <- file.path(params$out_dir, "results")   # .tsv
for (d in c(params$out_dir, params$ref_dir, fig_dir, tab_dir))
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
set.seed(params$seed)

TP_LABEL <- c("30 min", "1 h", "1.5 h", "2 h", "4 h", "6 h", "24 h")
```



```{r theme}
theme_pub <- function(base_size = 9) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title      = element_text(size = base_size, hjust = 0, face = "plain"),
      plot.tag        = element_text(size = base_size + 3, face = "bold"),
      axis.text       = element_text(size = base_size - 2, colour = "black"),
      axis.title      = element_text(size = base_size),
      legend.text     = element_text(size = base_size - 2),
      legend.title    = element_text(size = base_size - 1),
      legend.key.size = unit(0.35, "cm"),
      strip.background = element_blank(),
      strip.text      = element_text(size = base_size)
    )
}
theme_set(theme_pub())

COL_UP   <- "#B2182B"
COL_DOWN <- "#2166AC"
COL_GREY <- "#7F7F7F"
```

# Sample sheet

Time point and replicate are parsed out of the file names
(`..._LPIF_<minutes>_S_<replicate>.1.1_quant.sf.gz`). The parse is asserted rather than
trusted: a silent mis-parse here would corrupt every downstream contrast.

```{r samples}
qfiles <- list.files(params$quant_dir, pattern = "quant\\.sf(\\.gz)?$", full.names = TRUE)
stopifnot(length(qfiles) > 0)

samples <- data.table(
  path     = qfiles,
  file     = basename(qfiles),
  gsm      = sub("^(GSM[0-9]+)_.*", "\\1", basename(qfiles)),
  time_min = as.integer(sub(".*_LPIF_([0-9]+)_S_.*", "\\1", basename(qfiles))),
  rep      = sub(".*_S_([0-9]+)\\..*", "R\\1", basename(qfiles))
)
samples[, time_hr := time_min / 60]
samples[, time_f  := factor(sprintf("t%04d", time_min),
                            levels = sprintf("t%04d", sort(unique(time_min))))]
samples[, rep     := factor(rep)]
setorder(samples, time_min, rep)

# fail loudly on a bad parse
stopifnot(!any(is.na(samples$time_min)), !any(is.na(samples$rep)),
          nlevels(samples$time_f) >= 3, all(table(samples$time_f) >= 2))

fwrite(samples[, .(gsm, file, time_min, time_hr, rep)],
       file.path(tab_dir, "sample_sheet.tsv"), sep = "\t")
knitr::kable(samples[, .(gsm, time_min, time_hr, rep)], caption = "Sample sheet")
```

Design table -- confirms the blocked layout (every replicate label appears at every time point):

```{r design-table}
knitr::kable(table(samples$time_f, samples$rep), caption = "Time point x replicate")
```

# Transcript-to-gene map

The transcript IDs in these `quant.sf` files come from an **older Ensembl release**, and the
version suffixes will not match a current GTF. Using the wrong release silently discards
reads at the gene-summarisation step, so the release is *chosen by measurement*, not assumed.

\

- **Ensembl 75 / GRCh37** (`params$gtf_quant_annotation`) must be *old*: it has to match the
  annotation the Salmon index was built on, which decides which reads are summarised into
  which gene. Measured match to this index: 99.9% of transcripts. (The GSE201354 coverage
  tracks are hg19, and the GRCh38 alternatives in `GTF_URLS` cover only ~90%.)
- **GENCODE v49 = Ensembl 115** (`params$gtf_current`) must be *current*: its only job is to
  replace the 2014-era gene symbols, which affects labels and not a single count. Without it
  `ENSG00000169429` stays `IL8` rather than `CXCL8`, and the marker panel in Figure 1 -- a
  silent `SYMBOL %in%` filter -- would quietly drop that gene.

Setting either param to NULL falls back to downloading the corresponding `GTF_URLS` entry.
If `params$tx2gene_file` already exists it is used directly and neither GTF is read at all --
useful on a cluster node without outbound internet.

```{r tx2gene-helpers}
# Parse transcript-level records out of an Ensembl GTF.
parse_gtf <- function(gtf_path) {
  con <- if (grepl("\\.gz$", gtf_path)) gzfile(gtf_path, "rt") else file(gtf_path, "rt")
  on.exit(close(con))
  dt <- fread(text = grep("^#", readLines(con), value = TRUE, invert = TRUE),
              sep = "\t", header = FALSE, quote = "",
              select = c(1, 3, 9), col.names = c("chr", "feature", "attr"))
  dt <- dt[feature == "transcript"]
  dt[, `:=`(
    TXNAME_nov = sub('.*transcript_id "([^"]+)".*', "\\1", attr),
    GENEID     = sub('.*gene_id "([^"]+)".*',       "\\1", attr),
    SYMBOL     = fifelse(grepl("gene_name", attr),
                         sub('.*gene_name "([^"]+)".*', "\\1", attr), NA_character_),
    # Ensembl spells it gene_biotype, GENCODE spells it gene_type.
    BIOTYPE    = fifelse(grepl("gene_biotype", attr),
                         sub('.*gene_biotype "([^"]+)".*', "\\1", attr),
                 fifelse(grepl("gene_type", attr),
                         sub('.*gene_type "([^"]+)".*', "\\1", attr), NA_character_))
  )]
  dt[, TXNAME_nov := sub("\\..*$", "", TXNAME_nov)]
  unique(dt[, .(TXNAME_nov, GENEID, SYMBOL, BIOTYPE, chr)], by = "TXNAME_nov")
}

cached_download <- function(url, dest) {
  if (!file.exists(dest)) download.file(url, dest, mode = "wb", quiet = TRUE)
  dest
}

# fread() needs R.utils to read .gz directly; go through a connection so the
# script works whether or not that optional dependency is installed.
read_quant <- function(path) {
  if (grepl("\\.gz$", path)) {
    con <- gzfile(path, "rt"); on.exit(close(con))
    as.data.table(read.delim(con, stringsAsFactors = FALSE))
  } else {
    fread(path)
  }
}
```

```{r tx2gene-build}
quant1 <- read_quant(samples$path[1])
quant1[, nov := sub("\\..*$", "", Name)]

if (file.exists(params$tx2gene_file)) {
  t2g <- fread(params$tx2gene_file)
  message("Using pre-built tx2gene: ", params$tx2gene_file)
} else {
  # Which release do these transcript IDs belong to? Measure the overlap.
  ref75 <- parse_gtf(if (!is.null(params$gtf_quant_annotation)) params$gtf_quant_annotation
                     else cached_download(GTF_URLS["r75"], file.path(params$ref_dir, "ens75.gtf.gz")))
  match_rate <- data.table(release = "75",
                           match = mean(quant1$nov %in% ref75$TXNAME_nov))
  print(match_rate)
  if (match_rate$match < 0.95)
    warning("Release 75 matches only ", round(100 * match_rate$match, 1),
            "% of transcripts -- check releases 84/96 (see GTF_URLS) before proceeding.")

  # Refresh gene symbols against a current release (2014-era symbols are stale).
  cur <- tryCatch(
    parse_gtf(if (!is.null(params$gtf_current)) params$gtf_current
              else cached_download(GTF_URLS["cur"], file.path(params$ref_dir, "ens_current.gtf.gz"))),
    error = function(e) { message("Current GTF unavailable; keeping original symbols."); NULL })

  t2g <- copy(ref75)
  if (!is.null(cur)) {
    cur_gene <- unique(cur[, .(g = sub("\\..*$", "", GENEID),
                               SYMBOL_cur = SYMBOL, BIOTYPE_cur = BIOTYPE)], by = "g")
    t2g <- merge(t2g, cur_gene, by.x = "GENEID", by.y = "g", all.x = TRUE)
    t2g[!is.na(SYMBOL_cur),  SYMBOL  := SYMBOL_cur]
    t2g[!is.na(BIOTYPE_cur), BIOTYPE := BIOTYPE_cur]
  }
  t2g[is.na(SYMBOL), SYMBOL := GENEID]

  # tximport matches on the exact string in quant.sf, so restore the version suffix.
  t2g[, TXNAME := quant1$Name[match(TXNAME_nov, quant1$nov)]]
  t2g <- t2g[!is.na(TXNAME), .(TXNAME, GENEID, SYMBOL, BIOTYPE, CHR = chr)]
  fwrite(t2g, params$tx2gene_file, sep = "\t")
}

cov_rate  <- mean(quant1$Name %in% t2g$TXNAME)
lost_frac <- quant1[!Name %in% t2g$TXNAME, sum(NumReads)] / quant1[, sum(NumReads)]
cat(sprintf("tx2gene: %s transcripts -> %s genes\ncoverage of quant index: %.1f%%\nreads on unmapped transcripts: %.2f%%\n",
            format(nrow(t2g), big.mark = ","), format(uniqueN(t2g$GENEID), big.mark = ","),
            100 * cov_rate, 100 * lost_frac))
stopifnot(cov_rate > 0.95)

gmap <- unique(t2g[, .(GENEID, SYMBOL, BIOTYPE)], by = "GENEID")
setkey(gmap, GENEID)
```

# Import and filtering

```{r tximport}
files <- samples$path
names(files) <- samples$gsm

txi <- tximport(files, type = "salmon", tx2gene = t2g[, .(TXNAME, GENEID)],
                countsFromAbundance = "no", ignoreTxVersion = FALSE)

cat("gene x sample:", paste(dim(txi$counts), collapse = " x "), "\n")
cat("library sizes (M):", paste(round(colSums(txi$counts) / 1e6, 1), collapse = " "), "\n")
```

Genes are kept if they reach 10 counts in at least as many samples as the smallest group
(here 2), which drops the untestable low-count tail without pre-selecting on the effect of
interest.

```{r dds-filter}
coldata <- as.data.frame(samples[, .(gsm, time_f, rep, time_min, time_hr)])
rownames(coldata) <- coldata$gsm

dds <- DESeqDataSetFromTximport(txi, colData = coldata, design = ~ rep + time_f)

min_group <- min(table(samples$time_f))
keep <- rowSums(counts(dds) >= 10) >= min_group
cat(sprintf("genes kept: %s of %s\n",
            format(sum(keep), big.mark = ","), format(length(keep), big.mark = ",")))
dds <- dds[keep, ]
```

# Quality control

```{r qc}
vsd <- vst(dds, blind = TRUE)
.
pca_raw <- plotPCA(vsd, intgroup = c("time_f", "rep"), returnData = TRUE)
pct     <- round(100 * attr(pca_raw, "percentVar"))
pca_dt  <- as.data.table(pca_raw[, !duplicated(colnames(pca_raw)), drop = FALSE])
pca_dt[, time_min := as.integer(sub("^t", "", as.character(time_f)))]

rv  <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 2000)
cm  <- cor(assay(vsd)[rv, ], method = "spearman")
rep_cor <- sapply(levels(samples$time_f), function(tt) {
  i <- which(samples$time_f == tt); cm[i[1], i[2]] })

cat(sprintf("PC1 = %d%%, PC2 = %d%%\nreplicate correlation (Spearman): %.3f - %.3f\n",
            pct[1], pct[2], min(rep_cor), max(rep_cor)))
knitr::kable(data.table(time_point = names(rep_cor), spearman = round(rep_cor, 4)),
             caption = "Replicate-pair correlation, top 2,000 variable genes")
```

# Differential expression

## Likelihood-ratio test: any change across the series

The LRT compares the full model against one that keeps the replicate term but drops time.
It answers "does this gene change *anywhere* in the series", pooling evidence over all
post-treatment points, and is the right test for identifying dynamic genes.

```{r lrt}
dds_lrt <- DESeq(dds, test = "LRT", reduced = ~ rep, quiet = TRUE)
res_lrt <- results(dds_lrt, alpha = params$alpha)

cat(sprintf("tested: %s | padj < %.2f: %s | padj < 0.01: %s\n",
            format(sum(!is.na(res_lrt$padj)), big.mark = ","), params$alpha,
            format(sum(res_lrt$padj < params$alpha, na.rm = TRUE), big.mark = ","),
            format(sum(res_lrt$padj < 0.01, na.rm = TRUE), big.mark = ",")))
```

## Wald contrasts: each time point against 0 h

`apeglm` shrinkage is applied so that fold changes for low-count genes are not inflated;
shrunken values are what the figures and the effect-size filter use.

```{r wald}
dds_w <- DESeq(dds, quiet = TRUE)
tps   <- levels(samples$time_f)[-1]

wald <- lapply(tps, function(tt) {
  lfcShrink(dds_w, coef = paste0("time_f_", tt, "_vs_t0000"), type = "apeglm", quiet = TRUE)
})
names(wald) <- tps

de_counts <- rbindlist(lapply(tps, function(tt) {
  r <- wald[[tt]]
  data.table(time_min = as.integer(sub("^t", "", tt)),
             up   = sum(r$padj < params$alpha & r$log2FoldChange >  params$lfc_threshold, na.rm = TRUE),
             down = sum(r$padj < params$alpha & r$log2FoldChange < -params$lfc_threshold, na.rm = TRUE))
}))
de_counts[, `:=`(time_hr = time_min / 60, total = up + down)]
fwrite(de_counts, file.path(tab_dir, "DE_counts_per_timepoint.tsv"), sep = "\t")
knitr::kable(de_counts[, .(time_hr, up, down, total)],
             caption = sprintf("padj < %.2f and |LFC| > %g vs 0 h", params$alpha, params$lfc_threshold))
```

## Temporal clustering

Genes that are both statistically dynamic (LRT padj < 0.01) and of appreciable amplitude
(max |shrunken LFC| > 1) are z-scored across time and partitioned by k-means. Clusters are
relabelled by the time of their peak so the numbering is interpretable.

```{r clustering}
lfc_mat <- sapply(wald, function(r) r$log2FoldChange)
rownames(lfc_mat) <- rownames(wald[[1]])

sig_lrt     <- rownames(res_lrt)[which(res_lrt$padj < params$lrt_alpha_for_clustering)]
max_abs     <- rowMaxs(abs(lfc_mat[sig_lrt, , drop = FALSE]), na.rm = TRUE)
dynamic     <- sig_lrt[max_abs > params$lfc_threshold]
cat(sprintf("dynamic genes (LRT padj < %.2f and max|LFC| > %g): %s\n",
            params$lrt_alpha_for_clustering, params$lfc_threshold,
            format(length(dynamic), big.mark = ",")))

Z <- t(scale(t(lfc_mat[dynamic, , drop = FALSE])))
set.seed(params$seed)
km <- kmeans(Z, centers = params$n_clusters, nstart = 50, iter.max = 100)

peak_idx <- apply(km$centers, 1, which.max)
relabel  <- setNames(order(order(peak_idx, -apply(km$centers, 1, max))),
                     rownames(km$centers))
wave     <- relabel[as.character(km$cluster)]
names(wave) <- names(km$cluster)

wave_dt <- data.table(gene_id = names(wave), wave = as.integer(wave))
wave_dt[, symbol := gmap[gene_id, SYMBOL]]
```

```{r wave-summary}
centroid <- rbindlist(lapply(sort(unique(wave_dt$wave)), function(w) {
  ids <- wave_dt[wave == w, gene_id]
  data.table(wave = w, time_f = tps, z = colMeans(Z[ids, , drop = FALSE], na.rm = TRUE))
}))

wave_summary <- rbindlist(lapply(sort(unique(wave_dt$wave)), function(w) {
  ids  <- wave_dt[wave == w, gene_id]
  zc   <- colMeans(Z[ids, , drop = FALSE], na.rm = TRUE)
  pk   <- which.max(zc)
  lf   <- lfc_mat[ids, pk]
  fin  <- lfc_mat[ids, length(tps)]
  nup  <- sum(lf >  0.5, na.rm = TRUE)
  ndn  <- sum(lf < -0.5, na.rm = TRUE)
  top  <- wave_dt[wave == w][order(-abs(lfc_mat[gene_id, pk]))][1:6, symbol]
  data.table(wave = w, n = length(ids), peak = TP_LABEL[pk],
             shape = if (nup > 3 * ndn) "induced" else if (ndn > 3 * nup) "repressed" else "mixed",
             median_LFC_at_peak = round(median(lf,  na.rm = TRUE), 2),
             median_LFC_at_24h  = round(median(fin, na.rm = TRUE), 2),
             examples = paste(top, collapse = ", "))
}))
fwrite(wave_summary, file.path(tab_dir, "cluster_summary.tsv"), sep = "\t")
knitr::kable(wave_summary, caption = "Temporal waves")
```

# Result tables

```{r tables}
lfc_dt <- as.data.table(lfc_mat, keep.rownames = "gene_id")
setnames(lfc_dt, c("gene_id", paste0("LFC_", sub("^t0*", "", tps), "min")))

lrt_out <- as.data.table(as.data.frame(res_lrt), keep.rownames = "gene_id")
lrt_out[, `:=`(symbol = gmap[gene_id, SYMBOL], biotype = gmap[gene_id, BIOTYPE])]
lrt_out <- lrt_out[, .(gene_id, symbol, biotype, baseMean = round(baseMean, 2),
                       stat = round(stat, 3), pvalue, padj)]
lrt_out <- merge(lrt_out, lfc_dt,  by = "gene_id", all.x = TRUE)
lrt_out <- merge(lrt_out, wave_dt[, .(gene_id, wave)], by = "gene_id", all.x = TRUE)
setorder(lrt_out, padj, na.last = TRUE)
fwrite(lrt_out, file.path(tab_dir, "DE_timecourse_LRT_all_genes.tsv"), sep = "\t")

wald_long <- rbindlist(lapply(tps, function(tt) {
  r <- as.data.table(as.data.frame(wald[[tt]]), keep.rownames = "gene_id")
  r[, `:=`(symbol = gmap[gene_id, SYMBOL],
           contrast = paste0(sub("^t0*", "", tt), "min_vs_0min"))]
  r[, .(gene_id, symbol, contrast, baseMean = round(baseMean, 2),
        log2FoldChange = round(log2FoldChange, 4), lfcSE = round(lfcSE, 4), pvalue, padj)]
}))
fwrite(wald_long, file.path(tab_dir, "DE_per_timepoint_vs_0min.tsv"), sep = "\t")
fwrite(wald_long[padj < params$alpha & abs(log2FoldChange) > params$lfc_threshold][
         order(contrast, -abs(log2FoldChange))],
       file.path(tab_dir, "DE_per_timepoint_significant.tsv"), sep = "\t")

fwrite(lrt_out[!is.na(wave)][order(wave, padj)],
       file.path(tab_dir, "dynamic_genes_by_wave.tsv"), sep = "\t")

vst_out <- as.data.table(assay(vsd), keep.rownames = "gene_id")
vst_out[, symbol := gmap[gene_id, SYMBOL]]
setcolorder(vst_out, c("gene_id", "symbol"))
fwrite(vst_out, file.path(tab_dir, "vst_normalized_counts.tsv"), sep = "\t")
fwrite(as.data.table(counts(dds_w, normalized = TRUE), keep.rownames = "gene_id"),
       file.path(tab_dir, "normalized_counts.tsv"), sep = "\t")

knitr::kable(head(lrt_out[, .(symbol, padj = signif(padj, 3),
                              LFC_30min, LFC_240min, LFC_1440min, wave)], 15),
             caption = "Top 15 genes by LRT")
```

# Figure 1 -- time-course overview

```{r fig1, fig.width = 8.2, fig.height = 7.2}
hr_lab <- function(m) ifelse(m %% 60 == 0, sprintf("%g", m / 60), sprintf("%g", m / 60))
tp_min <- as.integer(sub("^t", "", tps))

## (a) PCA -----------------------------------------------------------------
pca_mean <- pca_dt[, .(PC1 = mean(PC1), PC2 = mean(PC2)), by = time_min][order(time_min)]
pA <- ggplot(pca_dt, aes(PC1, PC2)) +
  geom_path(data = pca_mean, colour = COL_GREY, linewidth = 0.4) +
  geom_point(aes(fill = time_min), shape = 21, size = 2.6, colour = "white", stroke = 0.4) +
  geom_text(data = pca_mean, aes(label = paste0(time_min / 60, " h")),
            size = 2.4, vjust = -1.3) +
  scale_fill_viridis_c(trans = "sqrt", guide = "none") +
  scale_x_continuous(expand = expansion(mult = 0.14)) +
  scale_y_continuous(expand = expansion(mult = 0.14)) +
  labs(x = sprintf("PC1 (%d%% variance)", pct[1]),
       y = sprintf("PC2 (%d%% variance)", pct[2]),
       title = "Samples order by time along PC1")

## (b) DE counts -----------------------------------------------------------
cnt_long <- melt(de_counts[, .(time_min, up, down)], id.vars = "time_min",
                 variable.name = "direction", value.name = "n")
cnt_long[, x := match(time_min, tp_min)]
pB <- ggplot(cnt_long, aes(x, n, colour = direction, shape = direction)) +
  geom_line(linewidth = 0.5) + geom_point(size = 1.8) +
  geom_text(data = cnt_long[x == max(x)],
            aes(label = format(n, big.mark = ","),
                vjust = ifelse(direction == "up", -1.1, 1.9)),
            size = 2.3, hjust = 1, show.legend = FALSE) +
  scale_colour_manual(values = c(up = COL_UP, down = COL_DOWN), name = NULL) +
  scale_shape_manual(values = c(up = 16, down = 15), name = NULL) +
  scale_x_continuous(breaks = seq_along(tp_min), labels = tp_min / 60,
                     expand = expansion(mult = 0.09)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.16))) +
  theme(legend.position = c(0.12, 0.88), legend.background = element_blank()) +
  labs(x = expression("Time after LPS/IFN" * gamma * " (h)"),
       y = sprintf("Genes (padj<%.2f, |LFC|>%g)", params$alpha, params$lfc_threshold),
       title = "Response grows and turns bidirectional")

## (c) canonical markers ---------------------------------------------------
early <- c("TNF", "CCL4", "NFKBIA", "CXCL8")
late  <- c("IDO1", "IFIT1", "GBP1", "STAT1")
mk_ids <- gmap[SYMBOL %in% c(early, late) & GENEID %in% rownames(assay(vsd))]
vst_mean <- t(apply(assay(vsd)[mk_ids$GENEID, , drop = FALSE], 1,
                    function(v) tapply(v, samples$time_min, mean)))
rownames(vst_mean) <- mk_ids$SYMBOL

mk_dt <- melt(as.data.table(vst_mean, keep.rownames = "symbol"),
              id.vars = "symbol", variable.name = "time_min", value.name = "vst")
mk_dt[, time_min := as.integer(as.character(time_min))]
mk_dt[, delta := vst - vst[time_min == 0], by = symbol]
mk_dt[, x := match(time_min, sort(unique(time_min)))]
mk_dt[, class := ifelse(symbol %in% early, "NF-kB / immediate-early", "IFN-gamma / STAT1")]

pC <- ggplot(mk_dt, aes(x, delta, group = symbol, colour = class)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = COL_GREY, linewidth = 0.3) +
  geom_line(linewidth = 0.5) +
  geom_text_repel(data = mk_dt[x == max(x)], aes(label = symbol),
                  size = 2.2, fontface = "italic", direction = "y", hjust = 0,
                  nudge_x = 0.4, segment.size = 0.2, segment.colour = COL_GREY,
                  min.segment.length = 0, box.padding = 0.12, show.legend = FALSE) +
  scale_colour_manual(values = setNames(c(COL_UP, COL_DOWN),
                                        c("NF-kB / immediate-early", "IFN-gamma / STAT1")),
                      name = NULL) +
  scale_x_continuous(breaks = seq_along(unique(mk_dt$time_min)),
                     labels = sort(unique(mk_dt$time_min)) / 60,
                     expand = expansion(mult = c(0.03, 0.22))) +
  theme(legend.position = c(0.32, 0.93), legend.background = element_blank(),
        legend.spacing.y = unit(0.02, "cm")) +
  labs(x = expression("Time after LPS/IFN" * gamma * " (h)"),
       y = "Change vs 0 h (VST units)",
       title = expression("NF-" * kappa * "B genes fire first, IFN" * gamma * " targets later"))

## (d) wave centroids ------------------------------------------------------
centroid[, x := match(time_f, tps)]
wave_n <- wave_dt[, .N, by = wave]
cent_lab <- merge(centroid[x == max(x)], wave_n, by = "wave")
cent_lab[, lab := sprintf("C%d (n=%s)", wave, format(N, big.mark = ","))]

pD <- ggplot(centroid, aes(x, z, colour = factor(wave))) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = COL_GREY, linewidth = 0.3) +
  geom_line(linewidth = 0.5) + geom_point(size = 1.2) +
  geom_text_repel(data = cent_lab, aes(label = lab), size = 2.2, direction = "y",
                  hjust = 0, nudge_x = 0.35, segment.size = 0.2, min.segment.length = 0,
                  box.padding = 0.1, show.legend = FALSE) +
  scale_colour_viridis_d(option = "plasma", end = 0.85, guide = "none") +
  scale_x_continuous(breaks = seq_along(tp_min), labels = tp_min / 60,
                     expand = expansion(mult = c(0.03, 0.30))) +
  labs(x = expression("Time after LPS/IFN" * gamma * " (h)"), y = "Mean z-scored LFC",
       title = sprintf("%d temporal waves among %s dynamic genes",
                       params$n_clusters, format(length(dynamic), big.mark = ",")))

fig1 <- (pA | pB) / (pC | pD) + plot_annotation(tag_levels = "a")
ggsave(file.path(fig_dir, "fig1_timecourse_overview.png"), fig1,
       width = 8.2, height = 7.2, dpi = 300, bg = "white")
fig1
```

# Figure 2 -- volcano and wave heatmap

```{r fig2, fig.width = 8.4, fig.height = 4.0}
FOCUS <- "t0030"   # contrast to display
P_FLOOR <- 300

## (a) volcano -------------------------------------------------------------
volc <- as.data.table(as.data.frame(wald[[FOCUS]]), keep.rownames = "gene_id")
volc[, symbol := gmap[gene_id, SYMBOL]]
volc <- volc[!is.na(padj)]
volc[, mlp := pmin(-log10(pmax(padj, 1e-300)), P_FLOOR)]
volc[, cls := fifelse(padj < params$alpha & log2FoldChange >  params$lfc_threshold, "up",
              fifelse(padj < params$alpha & log2FoldChange < -params$lfc_threshold, "down", "n.s."))]

n_up <- volc[cls == "up", .N]; n_dn <- volc[cls == "down", .N]
lab_genes <- head(volc[cls == "up" & padj < 1e-20][order(-log2FoldChange)], 5)

pE <- ggplot(volc, aes(log2FoldChange, mlp, colour = cls)) +
  geom_vline(xintercept = c(-1, 1) * params$lfc_threshold, linetype = "dashed",
             colour = COL_GREY, linewidth = 0.3) +
  geom_hline(yintercept = -log10(params$alpha), linetype = "dashed",
             colour = COL_GREY, linewidth = 0.3) +
  geom_point(size = 0.35, alpha = 0.75) +
  geom_text_repel(data = lab_genes, aes(label = symbol), size = 2.2, fontface = "italic",
                  colour = "black", direction = "y", hjust = 0,
                  segment.size = 0.2, segment.colour = COL_GREY,
                  min.segment.length = 0, box.padding = 0.2, show.legend = FALSE) +
  scale_colour_manual(values = c(up = COL_UP, down = COL_DOWN, `n.s.` = "grey80"),
                      name = NULL,
                      breaks = c("up", "down", "n.s.")) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.22))) +
  guides(colour = guide_legend(override.aes = list(size = 1.6))) +
  theme(legend.position = c(0.09, 0.72), legend.background = element_blank()) +
  labs(x = sprintf("log2 fold change (%s h vs 0 h)",
                   as.integer(sub("^t", "", FOCUS)) / 60),
       y = "-log10 adjusted p",
       title = sprintf("%s up / %s down at %g h",
                       format(n_up, big.mark = ","), format(n_dn, big.mark = ","),
                       as.integer(sub("^t", "", FOCUS)) / 60))

## (b) heatmap -------------------------------------------------------------
ord <- wave_dt[order(wave)][, gene_id]
hm  <- melt(as.data.table(Z[ord, , drop = FALSE], keep.rownames = "gene_id"),
            id.vars = "gene_id", variable.name = "time_f", value.name = "z")
hm[, `:=`(row = match(gene_id, ord), x = match(time_f, tps))]
hm[, z := pmax(pmin(z, 2), -2)]

bounds <- cumsum(wave_dt[order(wave), .N, by = wave]$N)
mids   <- bounds - wave_dt[order(wave), .N, by = wave]$N / 2

pF <- ggplot(hm, aes(x, row, fill = z)) +
  geom_raster() +
  geom_hline(yintercept = head(bounds, -1), colour = "black", linewidth = 0.2) +
  annotate("label", x = 0.75, y = mids, label = paste0("C", seq_along(mids)),
           size = 2.1, label.size = 0, fill = "white", alpha = 0.75,
           label.padding = unit(0.08, "lines")) +
  scale_fill_distiller(palette = "RdBu", limits = c(-2, 2),
                       name = "z-scored\nlog2 FC vs 0 h") +
  scale_x_continuous(breaks = seq_along(tp_min), labels = tp_min / 60, expand = c(0, 0)) +
  scale_y_reverse(expand = c(0, 0)) +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  labs(x = expression("Time after LPS/IFN" * gamma * " (h)"),
       y = sprintf("%s dynamic genes, grouped by wave", format(length(ord), big.mark = ",")),
       title = "Distinct waves partition the response")

fig2 <- (pE | pF) + plot_layout(widths = c(1, 1.12)) + plot_annotation(tag_levels = "a")
ggsave(file.path(fig_dir, "fig2_volcano_heatmap.png"), fig2,
       width = 8.4, height = 4.0, dpi = 300, bg = "white")
fig2
```

# Session info

```{r session}
sessionInfo()
```
