---
title: "Regions with changes in FIRE frequency - Genome wide analysis"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = file.path(tempdir(), "fire-frequency-figures/"))
```

```{r load_packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggpubr)
library(ggrepel)
library(tidyverse)
library(DT)
```

# FIRE frequency

chrom start   end        region_id timepoint n_reads n_fire     freq
1  chr1 11261 11486 chr1:11261-11486     LPS_0      12      2 0.166667
2  chr1 21178 21359 chr1:21178-21359     LPS_0      72      9 0.125000
3  chr1 28997 29166 chr1:28997-29166     LPS_0      26     12 0.461538
4  chr1 29283 29478 chr1:29283-29478     LPS_0      25     23 0.920000
5  chr1 29964 30121 chr1:29964-30121     LPS_0      24      6 0.250000
6  chr1 38601 38804 chr1:38601-38804     LPS_0      19      5 0.263158

Each region is a union FIRE peak across all timepoints. `n_reads` counts reads
whose aligned span fully covers the region; `n_fire` counts those reads with a
called FIRE element covering at least 50% of the region.
```{r load_data}
input_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency"

fire_freq <- read.csv(file.path(input_dir, "fire_frequency_long.tsv.gz"), sep = "\t")
head(fire_freq)

# group by region_id and timepoint, and calculate the mean frequency
fire_freq_summary <- fire_freq %>%
  group_by(region_id, timepoint) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE)) %>%
  ungroup()


# rank region by mean frequency for each timepoint
fire_freq_summary <- fire_freq_summary %>%
  group_by(timepoint) %>%
  mutate(rank = rank(-mean_freq, ties.method = "first")) %>%
  ungroup()

head(fire_freq_summary)

```


# cumulative distribution of FIRE frequency for each timepoint
for a given frequency x, what fraction of FIRE-universe regions have frequency ≤ x?

Left-shifted curve = more regions accumulate at lower x values → overall lower FIRE frequencies
```{r CDF_fire_freq}

fire_freq_cdf <- fire_freq_summary %>%
  filter(!is.na(mean_freq)) %>%
  group_by(timepoint) %>%
  arrange(mean_freq, .by_group = TRUE) %>%
  mutate(
    cdf = row_number() / n()
  ) %>%
  ungroup()

head(fire_freq_cdf)



p_freq_cdf <- ggplot(fire_freq_cdf, aes(x = mean_freq, color = timepoint)) +
  stat_ecdf(na.rm = TRUE) +
  labs(
    x = "FIRE frequency",
    y = "Cumulative fraction of regions",
    color = "Timepoint"
  ) +
  theme_classic()
p_freq_cdf
```


# Pairwise differential FIRE frequency: top 100 regions

Top 100 regions per comparison, ranked by Fisher exact p-value among
coverage-passing regions (6 comparisons x 100 rows).

```{r load_pairwise}
pairwise <- read_tsv(file.path(input_dir, "pairwise_fisher.tsv.gz"),
                     show_col_types = FALSE)

comparison_levels <- c("LPS_0_vs_LPS_5", "LPS_0_vs_LPS_10", "LPS_0_vs_LPS_15",
                       "LPS_5_vs_LPS_10", "LPS_5_vs_LPS_15", "LPS_10_vs_LPS_15")
pairwise <- pairwise %>%
  mutate(comparison = factor(comparison, levels = comparison_levels))

# only coverage-passing regions were ranked in 04_pairwise_differential.R
passing <- pairwise %>% filter(pass_coverage)

# top 100 per comparison by raw p
top100 <- passing %>%
  group_by(comparison) %>%
  arrange(p_value, .by_group = TRUE) %>%
  slice_head(n = 100) %>%
  ungroup() %>%
  mutate(direction = ifelse(delta_freq > 0, "gain", "loss"))

# gain = higher FIRE frequency at the later timepoint
table(top100$comparison, top100$direction)
```


# Regions tested vs significant

Tested = coverage-passing regions in that comparison; significant = unadjusted
Fisher p < 0.05. "any comparison" is the union across the 6 comparisons, i.e.
the definition used in the cCRE section below.

```{r tested_vs_signif_bar, fig.width=8, fig.height=4.5}
p_sig <- 0.05

counts_by_comparison <- passing %>%
  group_by(comparison) %>%
  summarise(tested      = n(),
            significant = sum(p_value < p_sig),
            .groups = "drop") %>%
  mutate(group = as.character(comparison))

counts_union <- tibble(
  group       = "any comparison",
  tested      = n_distinct(passing$region_id),
  significant = n_distinct(passing$region_id[passing$p_value < p_sig]))

region_counts <- bind_rows(counts_by_comparison %>% select(group, tested, significant),
                           counts_union) %>%
  mutate(group = factor(group, levels = c(comparison_levels, "any comparison")))

region_counts_long <- region_counts %>%
  pivot_longer(c(tested, significant), names_to = "category", values_to = "n") %>%
  mutate(category = factor(category, levels = c("tested", "significant")))

p_region_counts <- ggplot(region_counts_long, aes(x = group, y = n, fill = category)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7,
           color = "grey30", linewidth = 0.2) +
  geom_text(aes(label = scales::comma(n)), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c(tested = "grey70", significant = "firebrick")) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "FIRE regions", fill = NULL,
       subtitle = paste0("FIRE universe: ", n_distinct(pairwise$region_id),
                         " regions; significant = p < ", p_sig)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_region_counts

region_counts %>% mutate(frac_signif = round(significant / tested, 3))
```


# P-value distribution per comparison

Histogram of the unadjusted Fisher p-values 

```{r pvalue_hist, fig.width=12, fig.height=6}
expected_bin <- passing %>%
  count(comparison) %>%
  mutate(expected = n * 0.025)

p_pval_hist <- ggplot(passing, aes(x = p_value)) +
  geom_histogram(binwidth = 0.025, boundary = 0,
                 fill = "grey70", color = "grey30", linewidth = 0.2) +
  geom_hline(data = expected_bin, aes(yintercept = expected),
             linetype = "dashed", color = "red") +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = "Fisher exact p (unadjusted)", y = "Regions") +
  theme_classic()
p_pval_hist
```


# Volcano: effect size vs significance

Effect size = delta_freq (later - earlier FIRE frequency); significance =
-log10 of the unadjusted Fisher p. Coverage-passing regions only. Points are
coloured by direction among the regions with p < 0.05.

```{r volcano_all, fig.width=12, fig.height=6}
volcano <- passing %>%
  mutate(neglog10_p = -log10(p_value),
         status = case_when(p_value >= p_sig  ~ "n.s.",
                            delta_freq > 0    ~ "gain (up in later)",
                            TRUE              ~ "loss (down in later)"),
         status = factor(status, levels = c("n.s.", "gain (up in later)",
                                            "loss (down in later)")))

status_cols <- c("n.s." = "grey75",
                 "gain (up in later)" = "firebrick",
                 "loss (down in later)" = "steelblue")

p_volcano <- ggplot(volcano, aes(x = delta_freq, y = neglog10_p, color = status)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_hline(yintercept = -log10(p_sig), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ comparison, nrow = 2) +
  scale_color_manual(values = status_cols) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = expression(Delta ~ "FIRE frequency (later - earlier)"),
       y = expression(-log[10] ~ "Fisher p (unadjusted)"), color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
p_volcano
```

## LPS_0 vs LPS_15: regions gained and lost at 15 min

Same volcano restricted to the 0 vs 15 min comparison. Labelled points are the
top 10 in each direction: p < 0.05 and the largest |delta_freq|, i.e. the
regions with the highest FIRE frequency at 15 min relative to 0 (gains) and the
lowest at 15 min relative to 0 (losses).

```{r volcano_0_vs_15, fig.width=8.5, fig.height=6}
n_label <- 10

v15 <- volcano %>% filter(comparison == "LPS_0_vs_LPS_15")

top_gain_15 <- v15 %>%
  filter(p_value < p_sig, delta_freq > 0) %>%
  arrange(desc(delta_freq), p_value) %>%
  mutate(rank = row_number(), direction = "gain")

top_loss_15 <- v15 %>%
  filter(p_value < p_sig, delta_freq < 0) %>%
  arrange(delta_freq, p_value) %>%
  mutate(rank = row_number(), direction = "loss")

labelled_15 <- bind_rows(head(top_gain_15, n_label), head(top_loss_15, n_label))

p_volcano_15 <- ggplot(v15, aes(x = delta_freq, y = neglog10_p, color = status)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_hline(yintercept = -log10(p_sig), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(data = labelled_15, color = "black", size = 1.4) +
  geom_text_repel(data = labelled_15, aes(label = region_id),
                  color = "black", size = 2.6, min.segment.length = 0,
                  max.overlaps = Inf, box.padding = 0.4) +
  scale_color_manual(values = status_cols) +

  scale_x_continuous(expand = expansion(mult = c(0.35, 0.35))) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = expression(Delta ~ "FIRE frequency (15 min - 0 min)"),
       y = expression(-log[10] ~ "Fisher p (unadjusted)"), color = NULL,
       subtitle = paste0(nrow(top_gain_15), " gains / ", nrow(top_loss_15),
                         " losses at p < ", p_sig,
                         "; top ", n_label, " each labelled")) +
  theme_classic() +
  theme(legend.position = "top")
p_volcano_15
```

```{r top_15_tables}
top_15 <- bind_rows(top_gain_15, top_loss_15)
write_tsv(top_15, file.path(input_dir, "top_changed_LPS_0_vs_LPS_15.tsv"))

# shared column layout for the two direction tables
format_top15 <- function(d) {
  d %>%
    transmute(rank, region_id,
              n_reads_0 = n_reads_A, freq_0 = round(freq_A, 3),
              n_reads_15 = n_reads_B, freq_15 = round(freq_B, 3),
              delta_freq = round(delta_freq, 3), log2_or = round(log2_or, 2),
              p_value = signif(p_value, 3))
}
```

Gains: higher FIRE frequency at 15 min than at 0, ranked by delta_freq.

```{r top_gain_15_table}
datatable(format_top15(top_gain_15), filter = "top", rownames = FALSE,
          caption = paste0("LPS_0 vs LPS_15 gains (p < ", p_sig, "): ",
                           nrow(top_gain_15), " regions"),
          options = list(pageLength = 10))
```

Losses: lower FIRE frequency at 15 min than at 0, ranked by the most negative
delta_freq.

```{r top_loss_15_table}
datatable(format_top15(top_loss_15), filter = "top", rownames = FALSE,
          caption = paste0("LPS_0 vs LPS_15 losses (p < ", p_sig, "): ",
                           nrow(top_loss_15), " regions"),
          options = list(pageLength = 10))
```


# QQ plot: observed vs null p-values

Do Fisher's p values from FIRE frequency comparisons contain more small p values than expected if there were no real differences between timepoints?

  1. For each pairwise comparison, all coverage passing regions are ranked from smallest to largest p values, the observed fisher p-vlaue for each region is converted to
  -log10(p), and is plotted against the -log10(p) expected at the same rank if all p values follow a unifomr (0,1) null distribution
  
  2. Points above the diagonal means there are more very small p values than expected
  
  
```{r pvalue_qq, fig.width=12, fig.height=6}
qq <- passing %>%
  group_by(comparison) %>%
  arrange(p_value, .by_group = TRUE) %>%
  # convert to -log(10) based p value
  mutate(observed = -log10(p_value),
         # expected p value at the same rank if all regions are null
         expected = -log10((row_number() - 0.5) / n())) %>%
  ungroup()

pi0_tab <- passing %>%
  group_by(comparison) %>%
  # count how many regions are tested in each pairwise comparison
  summarise(tested = n(),
            # fraction of tests behave like null hypothesis 
            pi0    = min(1, mean(p_value > 0.5) / 0.5),
            # number of regions with fisher p value < 0.05
            signif = sum(p_value < p_sig),
            .groups = "drop") %>%
  # estimated number of sig regions from the null hypothesis alone
  mutate(expected_null_signif = round(pi0 * tested * p_sig),
         excess               = signif - expected_null_signif)

# keeps the 5000 smallest p values
qq_thin <- qq %>%
  group_by(comparison) %>%
  filter(row_number() <= 5000 | row_number() %% 20 == 0) %>%
  ungroup()

# summary table
pi0_lab <- pi0_tab %>%
  mutate(label = paste0("pi0 = ", round(pi0, 2),
                        "\nexcess p<", p_sig, ": ", scales::comma(excess)))

# qq plot
  # x = expected -log10(p)
  # y = observed -log10(fisher p)
p_qq <- ggplot(qq_thin, aes(x = expected, y = observed)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(size = 0.4, alpha = 0.4, color = "grey30") +
  geom_text(data = pi0_lab, aes(x = 0, y = max(qq$observed), label = label),
            hjust = 0, vjust = 1, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = expression("Expected" ~ -log[10] ~ "p (uniform null)"),
       y = expression("Observed" ~ -log[10] ~ "Fisher p")) +
  theme_classic()
p_qq

pi0_tab %>% mutate(pi0 = round(pi0, 3))
```


# Significant regions vs ENCODE cCREs

Significant = unadjusted Fisher p < 0.05 in at least one of the 6 pairwise
comparisons

`ccre_class` / `ccre_id` report that best cCRE.

```{r ccre_overlap}
suppressPackageStartupMessages(library(GenomicRanges))

ccre_file <- "/project/spott/cshan/annotations/GRCh38-cCREs.bed"
p_sig <- 0.05

signif_rows <- passing %>% filter(p_value < p_sig)

# region level: keep the samllest p comparison for ranking
signif_regions <- signif_rows %>%
  group_by(region_id, chrom, start, end) %>%
  summarise(n_signif_comparisons = n(),
            min_p           = min(p_value),
            best_comparison = comparison[which.min(p_value)],
            delta_freq_best = delta_freq[which.min(p_value)],
            .groups = "drop")

ccre <- data.table::fread(ccre_file, header = FALSE,
                          col.names = c("chr", "start", "end",
                                        "ccre_dcc", "ccre_id", "ccre_class"))

reg.gr  <- GRanges(signif_regions$chrom,
                   IRanges(signif_regions$start + 1L, signif_regions$end))
ccre.gr <- GRanges(ccre$chr, IRanges(ccre$start + 1L, ccre$end))
hits <- findOverlaps(reg.gr, ccre.gr)
ov_frac <- width(pintersect(reg.gr[queryHits(hits)], ccre.gr[subjectHits(hits)])) /
  width(reg.gr)[queryHits(hits)]

# best cCRE per region (largest covered fraction)
best <- data.frame(ri = queryHits(hits), frac = ov_frac,
                   ccre_id = ccre$ccre_id[subjectHits(hits)],
                   ccre_class = ccre$ccre_class[subjectHits(hits)]) %>%
  group_by(ri) %>%
  slice_max(frac, n = 1, with_ties = FALSE) %>%
  ungroup()

signif_regions <- signif_regions %>%
  mutate(ccre_frac = 0, ccre_id = NA_character_, ccre_class = NA_character_)
signif_regions$ccre_frac[best$ri]  <- best$frac
signif_regions$ccre_id[best$ri]    <- best$ccre_id
signif_regions$ccre_class[best$ri] <- best$ccre_class

signif_ccre <- signif_regions %>% filter(ccre_frac >= 0.5)

tibble(n_signif_regions = nrow(signif_regions),
       any_ccre_overlap = sum(signif_regions$ccre_frac > 0),
       ccre_50          = nrow(signif_ccre),
       frac_ccre_50     = round(nrow(signif_ccre) / nrow(signif_regions), 3))
```

Nearest gencode v46 gene for the cCRE-supported regions (distance 0 =
overlapping the gene body)

```{r ccre_nearest_gene}
gtf_file <- "/project/spott/cshan/annotations/gencode.v46.annotation.gtf.gz"
gtf <- data.table::fread(cmd = paste("zcat", gtf_file, "| awk -F'\\t' '$3==\"gene\"'"),
                         header = FALSE, sep = "\t")
genes.gr <- GRanges(gtf$V1, IRanges(gtf$V4, gtf$V5),
                    gene_name = sub('.*gene_name "([^"]+)".*', "\\1", gtf$V9),
                    gene_type = sub('.*gene_type "([^"]+)".*', "\\1", gtf$V9))

sc.gr <- GRanges(signif_ccre$chrom, IRanges(signif_ccre$start + 1L, signif_ccre$end))
d <- distanceToNearest(sc.gr, genes.gr)
signif_ccre$nearest_gene  <- NA_character_
signif_ccre$nearest_gene_type <- NA_character_
signif_ccre$gene_distance <- NA_integer_
signif_ccre$nearest_gene[queryHits(d)]      <- genes.gr$gene_name[subjectHits(d)]
signif_ccre$nearest_gene_type[queryHits(d)] <- genes.gr$gene_type[subjectHits(d)]
signif_ccre$gene_distance[queryHits(d)] <- mcols(d)$distance

write_tsv(signif_ccre %>% arrange(min_p),
          file.path(input_dir, "signif_ccre_regions.tsv"))
```

```{r ccre_class_dist, fig.width=6, fig.height=4}
p_ccre_class <- signif_ccre %>%
  count(ccre_class) %>%
  ggplot(aes(x = reorder(ccre_class, -n), y = n)) +
  geom_col(fill = "grey70", color = "grey30") +
  labs(x = "ENCODE cCRE class",
       y = "Significant regions (>= 50% cCRE overlap)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_ccre_class
```

Top 100 cCRE-supported significant regions by best p (full table in
`signif_ccre_regions.tsv`).

```{r ccre_table}
signif_ccre %>%
  arrange(min_p) %>%
  head(100) %>%
  transmute(region_id, best_comparison, n_signif_comparisons,
            min_p = signif(min_p, 3),
            delta_freq_best = round(delta_freq_best, 3),
            ccre_id, ccre_class, ccre_frac = round(ccre_frac, 2),
            nearest_gene, gene_distance) %>%
  datatable(filter = "top", rownames = FALSE, options = list(pageLength = 20))
```


# Read-level plots: top 10 cCRE-supported differential regions

The 5 cCRE-supported significant regions with the smallest p-value and are associated with protein coding genes

```{r region_functions}
code_dir <- "/project/spott/cshan/fiber-seq/code/topic_model/read_region_extraction"
source(file.path(code_dir, "region_data_utils.R"))
source(file.path(code_dir, "region_plot_functions.R"))
```


```{r top10_select}
options(scipen = 999)

extract_script <- file.path(code_dir, "extract_region_result_macrophage.sh")
extract_root <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency/region_plots"
sample_names <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")
pad <- 1000

# Keep the existing protein-coding nearest-gene filter and rank by best p-value.
top10 <- signif_ccre %>%
  filter(nearest_gene_type == "protein_coding") %>%
  arrange(min_p) %>%
  head(10) %>%
  mutate(rank = row_number(),
         outname = paste(chrom, start, end, sep = "_"),
         window_start = as.integer(start - pad),
         window_end   = as.integer(end + pad))
```

```{r top10_table, echo=FALSE}
top10 %>%
  transmute(rank, region_id, best_comparison,
            min_p = formatC(min_p, format = "e", digits = 3),
            delta_freq_best = round(delta_freq_best, 3),
            ccre_id, ccre_class, ccre_frac = round(ccre_frac, 2),
            nearest_gene, gene_distance) %>%
  knitr::kable(caption = "Top 10 cCRE-supported differential regions selected for read-level plots")
```

```{r top10_extract, include=FALSE}
for (i in seq_len(nrow(top10))) {
  for (s in sample_names) {
    parsed <- file.path(extract_root, top10$outname[i], s, "parsed")
    if (file.exists(file.path(parsed, "fire.bed"))) next
    region_str <- paste0(top10$chrom[i], ":", top10$window_start[i], "-",
                         top10$window_end[i])
    cat("extracting", top10$outname[i], s, "\n")
    rc <- system2("bash", c(extract_script, s, region_str, top10$outname[i],
                            extract_root))
    if (rc != 0) stop("extraction failed for ", top10$outname[i], " ", s)
  }
}
```

```{r top10_region_plots, fig.width=10, fig.height=16}
for (i in seq_len(nrow(top10))) {
  row <- top10[i, ]
  cat("=== ", row$region_id, " (", row$ccre_class, ", nearest gene: ",
      row$nearest_gene, ") ===\n", sep = "")

  res <- load_region_results(
    region       = data.frame(chr = row$chrom, start = row$window_start,
                              end = row$window_end),
    sample_names = sample_names,
    result_dirs  = file.path(extract_root, row$outname, sample_names, "parsed"),
    fiberHMM_feature      = "tf",
    min_fiberHMM_fp_score = 50,
    fiberHMM_size_breaks  = c(10, 30, 60, 80))

  # facet the read-level panels and the pileup by timepoint
  res <- add_read_groups(res,
                         labels = setNames(as.character(res$rids_df$sample_name),
                                           as.character(res$rids_df$RID)),
                         group_col = "timepoint",
                         group_levels = sample_names)

  # cCREs in the window, one row per cCRE labelled with its class; the tested region
  # itself is the yellow band behind every panel
  peaks_df <- ccre %>%
    filter(chr == row$chrom, end > row$window_start, start < row$window_end) %>%
    arrange(ccre_class, start) %>%
    transmute(chr, start, end, group = ccre_class)

  p <- plot_region_panels(res,
                          plot_name = paste0(row$region_id, "  (", row$ccre_class,
                                             ", nearest gene: ", row$nearest_gene, ")"),
                          peaks_df = peaks_df,
                          highlight = c(row$start, row$end),
                          fiberHMM_cols = c("purple", "magenta", "#41c4e1"),
                          heights = c(pileup = 10, peaks = 2, fire_peaks = 1,
                                      reads = 8, fire_fiberHMM = 8))
  print(p$combined)
}
```


# Top 10 cCRE-supported regions: LPS time-course example figures

The same 10 regions as the read-level plots above (`top10`: cCRE-supported,
nearest gene protein coding, ranked by best pairwise Fisher p). Every panel uses
the same window, the tested FIRE region +/- `pad` bp (`window_start`-`window_end`),
and marks the tested region itself.

Molecules are the full-span reads of the Leiden + Manhattan workflow
(`assemble_region_m6a()` + `filter_met_mat()`: reads covering every m6A-called
site in the window), so panels A-D all describe the same reads. Each timepoint
is one merged sample (`LPS_0` ... `LPS_15`, biological replicates already
pooled), as in the rest of this notebook.

- **A** aggregate m6A per timepoint: the pileup of the read-level plots above
  (`pileup_reads()`: fraction of reads with an m6A call at each A, +/- 5 bp
  sliding mean), with ENCODE cCREs in the window.
- **B** every full-span molecule, by timepoint, not clustered (rows ordered by
  alignment start). Footprint encoding of the read-level plots above: FIRE /
  linker segmentation from `fire.bed`, FiberHMM 130-160 bp nucleosomes, and
  FIRE-polished FiberHMM TF footprints.
- **C** Leiden + Manhattan clustering of the molecules pooled over 0-15 min,
  with the settings of `marcophage_leiden_manhattan.Rmd` (site-level 0/1 m6A
  matrix, no windowing, k = 10, resolution 1). Timepoint is metadata only and no
  haplotype information is used. C1 = m6A heatmap, C2 = per-cluster m6A pileup
  and every molecule's footprints, by cluster.
- **D** cluster composition per timepoint, P(cluster | time).

Only PDFs are written, one folder per region under `top10_example_plots/`.

```{r top10_examples_setup, message=FALSE}
suppressPackageStartupMessages({
  library(Matrix)
  library(igraph)
  library(ComplexHeatmap)
  library(cowplot)
  library(patchwork)
})
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
ht_opt$message <- FALSE

example_dir   <- file.path(input_dir, "top10_example_plots")
ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)

# Leiden + Manhattan settings of marcophage_leiden_manhattan.Rmd
window_size       <- 0L
k_neighbors       <- 10
leiden_resolution <- 1
leiden_seed       <- 1
kernel_sigma      <- NULL
impute_missing    <- FALSE
impute_k          <- 5

time_levels <- c("0 min", "5 min", "10 min", "15 min")
time_labels <- setNames(time_levels, sample_names)
time_cols   <- setNames(unname(LEIDEN_TIMEPOINT_COLORS[sample_names]), time_levels)

# footprint colours of plot_region_panels() in the read-level plots above
fp_cols <- c("linker (non-FIRE MSP)"  = "darkgray",
             "FIRE element"           = "#FF8C00",
             "nucleosome 130-160 bp"  = "darkblue",
             "TF footprint 10-30 bp"  = "purple",
             "TF footprint 30-60 bp"  = "magenta",
             "TF footprint 60-80 bp"  = "#41c4e1")

# ENCODE SCREEN cCRE colours
ccre_cols <- c(PLS = "#FF0000", pELS = "#FFA700", dELS = "#FFCD00",
               "CA-H3K4me3" = "#FFAAAA", "CA-CTCF" = "#00B0F0", "CA-TF" = "#BE28E5",
               CA = "#06DA93", TF = "#D876EC", "tested FIRE region" = "black")
```

```{r top10_examples_regions}
top10_examples_regions <- top10 %>%
  mutate(example_id = sprintf("rank%02d_%s", rank, outname)) %>%
  select(rank, example_id, outname, region_id, chrom, start, end, window_start, window_end,
         best_comparison, n_signif_comparisons, min_p, delta_freq_best,
         ccre_id, ccre_class, ccre_frac, nearest_gene, gene_distance)

# FIRE frequency of the tested region at each timepoint (fire_frequency_long)
top10_fire_freq <- fire_freq %>%
  filter(region_id %in% top10_examples_regions$region_id) %>%
  mutate(time = factor(time_labels[timepoint], levels = time_levels)) %>%
  select(region_id, time, n_reads, n_fire, freq)

top10_examples_regions %>%
  mutate(min_p = signif(min_p, 3), delta_freq_best = round(delta_freq_best, 3),
         ccre_frac = round(ccre_frac, 2)) %>%
  select(-region_id)
```

## Leiden + Manhattan clustering, timepoints pooled

`leiden_manhattan_cluster_regions()` pools the reads of the four samples before
the distance, graph and Leiden steps; `sample_name` / `timepoint` are only
carried into `assignments`. `k_eff` < `k_neighbors` would mean the region had
too few reads and k was capped at n - 1.

```{r top10_examples_leiden}
top10_leiden <- leiden_manhattan_cluster_regions(
  sample_names   = sample_names,
  regions        = data.frame(chr = top10_examples_regions$chrom,
                              start = top10_examples_regions$window_start,
                              end = top10_examples_regions$window_end,
                              region_id = top10_examples_regions$example_id),
  ft_result_dir  = ft_result_dir,
  window_size    = window_size,
  k_neighbors    = k_neighbors,
  resolution     = leiden_resolution,
  impute_missing = impute_missing,
  impute_k       = impute_k,
  sigma          = kernel_sigma,
  seed           = leiden_seed,
  sample_table   = sample_table,
  verbose        = FALSE)

bind_rows(lapply(names(top10_leiden), function(id) {
  p <- top10_leiden[[id]]$params
  data.frame(example_id = id, n_reads = p$n_reads, n_features = p$n_features,
             k_neighbors = p$k_neighbors, k_eff = p$k_eff, resolution = p$resolution,
             sigma = round(p$sigma, 2), n_clusters = top10_leiden[[id]]$n_clusters)
}))
```

```{r top10_examples_functions}
theme_example <- function() {
  theme_classic(base_size = 9) +
    theme(strip.background = element_rect(fill = "grey95", colour = NA),
          strip.text.y = element_text(angle = 0, hjust = 0),
          plot.title = element_text(face = "bold", size = 10),
          plot.subtitle = element_text(size = 8, colour = "grey30"),
          legend.position = "bottom", legend.key.size = unit(3, "mm"),
          legend.text = element_text(size = 8))
}

window_breaks <- function(ex) {
  b <- pretty(c(ex$window_start, ex$window_end), n = 5)
  b[b >= ex$window_start & b <= ex$window_end]
}

example_title <- function(ex) {
  sprintf("#%d  %s:%s-%s  |  %s %s  |  nearest gene %s", ex$rank, ex$chrom,
          scales::comma(ex$start), scales::comma(ex$end), ex$ccre_class, ex$ccre_id,
          ex$nearest_gene)
}

# plotting row of each molecule within its facet, in the current row order
stack_rows <- function(df, group_col) {
  df$y <- as.integer(ave(seq_len(nrow(df)), df[[group_col]], FUN = seq_along))
  df
}

# FIRE / linker segmentation, FiberHMM nucleosomes and FIRE-polished TF footprints
# of a load_region_results() result, one row per interval
molecule_tracks <- function(res) {
  as_track <- function(df, class) {
    if (is.null(df) || nrow(df) == 0) return(NULL)
    data.frame(RID = as.character(df$RID), start = df$start, end = df$end, class = class)
  }
  fire <- res$fire[res$fire$class %in% c("FIRE", "linker"), , drop = FALSE]
  fps  <- res$fps_infire[!is.na(res$fps_infire$class), , drop = FALSE]
  tr <- bind_rows(
    as_track(fire, ifelse(fire$class == "FIRE", "FIRE element", "linker (non-FIRE MSP)")),
    as_track(res$nucs, "nucleosome 130-160 bp"),
    as_track(fps, paste("TF footprint", fps$class)))
  tr$class <- factor(tr$class, levels = names(fp_cols))
  tr[order(tr$class), , drop = FALSE]      # TF footprints drawn last, on top
}

# cCREs overlapping the window plus the tested region, one row each
plot_ccre_track <- function(ex, xlim) {
  cc <- ccre %>%
    filter(chr == ex$chrom, end > ex$window_start, start < ex$window_end) %>%
    arrange(start) %>%
    transmute(start, end, label = paste(ccre_class, ccre_id), class = ccre_class)
  tr <- bind_rows(tibble(start = ex$start, end = ex$end, label = "tested FIRE region",
                         class = "tested FIRE region"), cc)
  tr$y <- rev(seq_len(nrow(tr)))
  p <- ggplot(tr) +
    geom_rect(aes(xmin = pmax(start, ex$window_start), xmax = pmin(end, ex$window_end),
                  ymin = y - 0.35, ymax = y + 0.35, fill = class)) +
    scale_fill_manual(values = ccre_cols, guide = "none") +
    scale_y_continuous(breaks = tr$y, labels = tr$label, expand = expansion(add = 0.5)) +
    scale_x_continuous(breaks = window_breaks(ex), labels = scales::comma) +
    coord_cartesian(xlim = xlim, expand = FALSE) +
    labs(x = paste(ex$chrom, "position (bp)"), y = NULL) +
    theme_example() +
    theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())
  list(plot = p, n_tracks = nrow(tr))
}

# Panel A: per-timepoint m6A pileup on a shared y axis
plot_aggregate_m6a <- function(pileup, ex, n_reads, fire_freq_ex) {
  xlim <- c(ex$window_start, ex$window_end)
  strip <- setNames(sprintf("%s\nn = %d", time_levels, as.integer(n_reads[time_levels])),
                    time_levels)
  p_sig <- ggplot(pileup, aes(pos, smooth_frac)) +
    annotate("rect", xmin = ex$start, xmax = ex$end, ymin = -Inf, ymax = Inf,
             fill = "grey88") +
    geom_area(aes(fill = time), alpha = 0.9) +
    geom_line(linewidth = 0.25, colour = "grey15") +
    facet_grid(time ~ ., labeller = as_labeller(strip)) +
    scale_fill_manual(values = time_cols, guide = "none") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    coord_cartesian(xlim = xlim, expand = FALSE) +
    labs(x = NULL, y = "m6A fraction (+/- 5 bp mean)",
         title = example_title(ex),
         subtitle = sprintf(paste0("Aggregate m6A, full-span reads per timepoint; grey band = tested region\n",
                                   "FIRE frequency of the tested region 0 / 5 / 10 / 15 min: %s;  ",
                                   "best %s: p = %s, delta = %+.3f"),
                            paste(sprintf("%.2f", fire_freq_ex$freq[match(time_levels, fire_freq_ex$time)]),
                                  collapse = " / "),
                            ex$best_comparison, formatC(ex$min_p, format = "e", digits = 2),
                            ex$delta_freq_best)) +
    theme_example() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line.x = element_blank())
  cc <- plot_ccre_track(ex, xlim)
  list(plot = p_sig / cc$plot + plot_layout(heights = c(4 * 1.2, 0.3 * cc$n_tracks + 0.2)),
       height = 1.6 + 4 * 1.2 + 0.3 * cc$n_tracks + 0.6)
}

# x range of the molecule plots: the window plus room for n_bars colour bars
bar_width <- function(ex) 0.022 * (ex$window_end - ex$window_start)
bar_gap   <- function(ex) 0.006 * (ex$window_end - ex$window_start)
molecule_xlim <- function(ex, n_bars) {
  c(ex$window_start - n_bars * (bar_width(ex) + bar_gap(ex)) - bar_gap(ex), ex$window_end)
}

# Panels B and C2: one row per molecule, facetted by `facet_col`; `bar_cols`
# draws a colour bar per molecule left of the window for each named column
plot_molecules <- function(rows, tracks, ex, facet_col, facet_labels,
                           bar_cols = character(), bar_palette = NULL, legend_extra = NULL,
                           title = NULL, subtitle = NULL) {
  bw   <- bar_width(ex)
  gap  <- bar_gap(ex)
  xlim <- molecule_xlim(ex, length(bar_cols))

  rows$facet <- rows[[facet_col]]
  tr <- inner_join(tracks, rows[, c("RID", "facet", "y")], by = "RID")
  tr$start <- pmax(tr$start, ex$window_start)
  tr$end   <- pmin(tr$end, ex$window_end)
  tr <- tr[tr$end > tr$start, , drop = FALSE]

  bars <- bind_rows(lapply(seq_along(bar_cols), function(i) {
    xmax <- ex$window_start - gap - (i - 1) * (bw + gap)
    data.frame(facet = rows$facet, y = rows$y, xmin = xmax - bw, xmax = xmax,
               fill = as.character(rows[[bar_cols[i]]]))
  }))

  p <- ggplot() +
    geom_segment(data = rows, aes(x = ex$window_start, xend = ex$window_end, y = y, yend = y),
                 colour = "grey80", linewidth = 0.15) +
    geom_rect(data = tr, aes(xmin = start, xmax = end, ymin = y - 0.45, ymax = y + 0.45,
                             fill = class))
  if (nrow(bars) > 0)
    p <- p + geom_rect(data = bars, aes(xmin = xmin, xmax = xmax, ymin = y - 0.5,
                                        ymax = y + 0.5, fill = fill))
  p +
    geom_vline(xintercept = c(ex$start, ex$end), linetype = "dashed", linewidth = 0.3) +
    facet_grid(facet ~ ., scales = "free_y", space = "free_y",
               labeller = as_labeller(facet_labels)) +
    # clusters are named by the facets, so only classes and extra keys are listed
    scale_fill_manual(values = c(fp_cols, bar_palette),
                      breaks = c(names(fp_cols), legend_extra), name = NULL) +
    scale_x_continuous(breaks = window_breaks(ex), labels = scales::comma) +
    scale_y_reverse(expand = expansion(add = 0.6)) +
    coord_cartesian(xlim = xlim, expand = FALSE) +
    guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
    labs(x = paste(ex$chrom, "position (bp)"), y = NULL, title = title, subtitle = subtitle) +
    theme_example() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          axis.line.y = element_blank(), panel.spacing.y = unit(1.5, "mm"),
          plot.margin = margin(5.5, 5.5, 5.5, 20))
}

# Panel C2 top: per-cluster m6A pileup, same normalisation as panel A
plot_cluster_m6a <- function(pileup, ex, xlim, cl_cols) {
  ggplot(pileup, aes(pos, smooth_frac, colour = cluster)) +
    geom_vline(xintercept = c(ex$start, ex$end), linetype = "dashed", linewidth = 0.3) +
    geom_line(linewidth = 0.4) +
    scale_colour_manual(values = cl_cols, name = NULL) +
    scale_x_continuous(breaks = window_breaks(ex), labels = NULL) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    coord_cartesian(xlim = xlim, expand = FALSE) +
    labs(x = NULL, y = "m6A fraction\n(per cluster)") +
    theme_example() +
    theme(legend.position = "right", axis.ticks.x = element_blank())
}

# Panel C1: rows = molecules split by cluster (then time, then read start),
# columns = every bp of the window; 0/1 = m6A call from the clustering input
draw_leiden_heatmap <- function(lr, rows, ex, cl_cols) {
  pos  <- seq.int(ex$window_start, ex$window_end)
  site <- as.matrix(lr$site_met_mat[rows$RID, , drop = FALSE])
  m <- matrix(0L, nrow(rows), length(pos), dimnames = list(rows$RID, pos))
  m[, match(colnames(site), pos)] <- site
  ticks <- window_breaks(ex)
  n_cl  <- table(rows$cluster)
  in_region <- ifelse(pos > ex$start & pos <= ex$end, "tested FIRE region", "outside")

  ht <- Heatmap(
    m, name = "m6A", col = c("0" = "white", "1" = "black"),
    heatmap_legend_param = list(at = c(0, 1), labels = c("no m6A call", "m6A call"),
                                title = "m6A"),
    cluster_rows = FALSE, cluster_columns = FALSE,
    row_split = rows$cluster,
    row_title = sprintf("%s\nn = %d", names(n_cl), as.integer(n_cl)),
    row_title_rot = 0, row_title_gp = grid::gpar(fontsize = 8),
    row_gap = unit(1, "mm"), border = TRUE,
    show_row_names = FALSE, show_column_names = FALSE,
    use_raster = TRUE, raster_quality = 4,
    left_annotation = rowAnnotation(
      cluster = rows$cluster, time = rows$time,
      col = list(cluster = cl_cols, time = time_cols),
      annotation_name_gp = grid::gpar(fontsize = 8),
      annotation_legend_param = list(cluster = list(title = "Leiden cluster"),
                                     time = list(title = "Time"))),
    bottom_annotation = HeatmapAnnotation(
      region = anno_simple(in_region, col = c("tested FIRE region" = "black", outside = "white"),
                           height = unit(1.5, "mm")),
      position = anno_mark(at = match(ticks, pos), labels = scales::comma(ticks),
                           side = "bottom", labels_rot = 0,
                           labels_gp = grid::gpar(fontsize = 8)),
      show_annotation_name = c(region = TRUE, position = FALSE),
      annotation_name_gp = grid::gpar(fontsize = 8), annotation_label = c("tested region", "")),
    column_title = paste0(example_title(ex), "\nLeiden + Manhattan clusters, reads pooled 0-15 min (",
                          ex$chrom, ", ", nrow(rows), " molecules)"),
    column_title_gp = grid::gpar(fontsize = 9))

  draw(ht, merge_legend = TRUE, heatmap_legend_side = "bottom",
       annotation_legend_side = "bottom")
  # dashed lines at the tested-region boundaries, through every cluster slice
  x <- match(c(ex$start, ex$end), pos) / length(pos)
  for (s in seq_along(n_cl))
    decorate_heatmap_body("m6A", slice = s, {
      grid::grid.segments(x0 = x, x1 = x, y0 = 0, y1 = 1,
                          gp = grid::gpar(lty = 2, lwd = 0.8, col = "#D55E00"))
    })
  invisible(NULL)
}

# Panel D: P(cluster | time) as stacked bars
plot_cluster_proportions <- function(counts, ex, cl_cols) {
  totals <- counts %>% distinct(time, n_time)
  n_cl <- counts %>% group_by(cluster) %>% summarise(n = sum(n), .groups = "drop")
  ggplot(counts, aes(time, prop, fill = cluster)) +
    geom_col(width = 0.7, colour = "white", linewidth = 0.2,
             position = position_stack(reverse = TRUE)) +
    geom_text(data = totals, aes(time, 1.01, label = paste0("n = ", n_time)),
              inherit.aes = FALSE, vjust = 0, size = 2.8) +
    scale_fill_manual(values = cl_cols,
                      labels = setNames(sprintf("%s (n = %d)", n_cl$cluster, n_cl$n),
                                        n_cl$cluster),
                      name = "Leiden cluster") +
    scale_y_continuous(limits = c(0, 1.07), breaks = seq(0, 1, 0.25),
                       labels = scales::percent, expand = c(0, 0)) +
    labs(x = "Time after LPS", y = "Fraction of the timepoint's reads",
         title = example_title(ex),
         subtitle = "P(cluster | time); clusters called on reads pooled over 0-15 min") +
    theme_example() +
    theme(legend.position = "right", plot.title = element_text(size = 8))
}
```

```{r top10_examples_plots, message=FALSE}
if (!dir.exists(example_dir)) dir.create(example_dir, recursive = TRUE)
top10_examples <- list()

for (i in seq_len(nrow(top10_examples_regions))) {
  ex  <- top10_examples_regions[i, ]
  id  <- ex$example_id
  lr  <- top10_leiden[[id]]
  out <- file.path(example_dir, id)
  if (!dir.exists(out)) dir.create(out)
  f <- function(name) file.path(out, sprintf("rank%02d_%s.pdf", ex$rank, name))

  # footprints and m6A of every read in the window, as in the read-level plots
  res_ex <- load_region_results(
    region       = data.frame(chr = ex$chrom, start = ex$window_start, end = ex$window_end),
    sample_names = sample_names,
    result_dirs  = file.path(extract_root, ex$outname, sample_names, "parsed"),
    fiberHMM_feature      = "tf",
    min_fiberHMM_fp_score = 50,
    fiberHMM_size_breaks  = c(10, 30, 60, 80),
    verbose = FALSE)

  # the clustered (full-span) reads, with timepoint as metadata
  a <- lr$assignments
  a$time <- factor(time_labels[as.character(a$sample_name)], levels = time_levels)
  stopifnot(all(a$RID %in% levels(res_ex$rids_df$RID)))
  cl_cols <- cluster_palette(levels(a$cluster))
  tracks  <- molecule_tracks(res_ex)

  ## A: aggregate m6A per timepoint --------------------------------------------
  res_time <- add_read_groups(res_ex, labels = setNames(as.character(a$time), a$RID),
                              group_col = "time", group_levels = time_levels,
                              verbose = FALSE)
  pA <- plot_aggregate_m6a(res_time$pileup, ex, table(a$time),
                           top10_fire_freq[top10_fire_freq$region_id == ex$region_id, ])
  ggsave(f("panel_A_aggregate_m6a"), pA$plot, width = 8, height = pA$height)

  ## B: molecules by timepoint, not clustered -------------------------------------
  rows_B <- stack_rows(a[order(a$time, a$start, a$RID), ], "time")
  n_B <- table(rows_B$time)
  lab_B <- setNames(sprintf("%s\nn = %d", time_levels, as.integer(n_B)), time_levels)
  pB <- plot_molecules(rows_B, tracks, ex, "time", lab_B,
    title = example_title(ex),
    subtitle = sprintf(paste0("All %d full-span molecules by timepoint (not clustered; ordered by read start)\n",
                              "dashed lines = tested region"), nrow(rows_B)))
  ggsave(f("panel_B_timepoint_footprints"), pB, width = 8,
         height = 1.8 + 0.025 * nrow(rows_B) + 0.1 * length(time_levels), limitsize = FALSE)

  ## C1: Leiden m6A heatmap ---------------------------------------------------------
  rows_C <- a[order(a$cluster, a$time, a$start, a$RID), ]
  h_C1 <- min(14, 3 + 0.02 * nrow(rows_C))
  pdf(f("panel_C_leiden_heatmap"), width = 9, height = h_C1)
  draw_leiden_heatmap(lr, rows_C, ex, cl_cols)
  dev.off()

  ## C2: per-cluster aggregate + molecules --------------------------------------------
  res_cl <- add_read_groups(res_ex, labels = setNames(as.character(a$cluster), a$RID),
                            group_col = "cluster", group_levels = levels(a$cluster),
                            verbose = FALSE)
  rows_C2 <- stack_rows(rows_C, "cluster")
  n_C2 <- table(rows_C2$cluster)
  lab_C2 <- setNames(sprintf("%s\nn = %d", names(n_C2), as.integer(n_C2)), names(n_C2))
  pC2_mol <- plot_molecules(rows_C2, tracks, ex, "cluster", lab_C2,
                            bar_cols = c("cluster", "time"),
                            bar_palette = c(cl_cols, time_cols), legend_extra = time_levels)
  pC2 <- (plot_cluster_m6a(res_cl$pileup, ex, molecule_xlim(ex, 2), cl_cols) / pC2_mol) +
    plot_layout(heights = c(1, 0.025 * nrow(rows_C2) + 0.1 * nlevels(a$cluster))) +
    plot_annotation(
      title = example_title(ex),
      subtitle = sprintf(paste0("Leiden clusters: per-cluster m6A pileup (top) and all %d molecules (bottom; ",
                                "left bars = cluster, time; ordered by time then read start)\n",
                                "dashed lines = tested region"), nrow(rows_C2)),
      theme = theme(plot.title = element_text(face = "bold", size = 10),
                    plot.subtitle = element_text(size = 8, colour = "grey30")))
  h_C2 <- 3.2 + 0.025 * nrow(rows_C2) + 0.1 * nlevels(a$cluster)
  ggsave(f("panel_C_cluster_footprints"), pC2, width = 8.5, height = h_C2, limitsize = FALSE)

  ## C: heatmap and cluster footprints side by side --------------------------------------
  h_C <- max(h_C1, h_C2)
  g_C1 <- grid::grid.grabExpr(draw_leiden_heatmap(lr, rows_C, ex, cl_cols),
                              width = 9, height = h_C)
  pC <- cowplot::plot_grid(
    cowplot::plot_grid(NULL, NULL, nrow = 1, rel_widths = c(9, 8.5), labels = c("C1", "C2")),
    cowplot::plot_grid(g_C1, patchwork::patchworkGrob(pC2), nrow = 1, rel_widths = c(9, 8.5)),
    ncol = 1, rel_heights = c(0.3, h_C))
  ggsave(f("panel_C_combined"), pC, width = 17.5, height = h_C + 0.3, limitsize = FALSE)

  ## D: cluster proportions per timepoint ----------------------------------------------
  counts <- a %>%
    count(time, cluster, .drop = FALSE) %>%
    group_by(time) %>%
    mutate(n_time = sum(n), prop = n / n_time) %>%
    ungroup()
  ggsave(f("panel_D_cluster_proportions"), plot_cluster_proportions(counts, ex, cl_cols),
         width = 6, height = 4.5)

  top10_examples[[id]] <- list(
    region = ex,
    leiden = lr,
    molecules = data.frame(
      time = time_levels,
      n_reads_window = as.integer(table(factor(time_labels[as.character(res_ex$rids_df$sample_name)],
                                               levels = time_levels))),
      n_full_span = as.integer(table(a$time))),
    cluster_time_counts = counts,
    files = list.files(out, full.names = TRUE))
}
```

Molecules per timepoint: reads overlapping the window (`n_reads_window`) and the
full-span reads used, all of them, in every panel (`n_full_span`).

```{r top10_examples_molecules}
bind_rows(lapply(top10_examples, function(x)
  data.frame(example_id = x$region$example_id, x$molecules)))
```

Raw counts behind panel D (`n` reads of cluster c at time t; `prop` = n / n_time).

```{r top10_examples_counts}
bind_rows(lapply(top10_examples, function(x)
  data.frame(example_id = x$region$example_id, x$cluster_time_counts))) %>%
  mutate(prop = round(prop, 3))
```
