---
title: "footprints at FIRE regions"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---


# FIRE frequency vs absolute footprint counts, timepoints pooled

The four LPS timepoints are treated as one pooled sample. For each region, this notebook relates
FIRE frequency to the absolute number of TF footprints in that region. FIRE frequency is always
measured on the union region; the footprint-vs-FIRE correlations and plots count only the
footprints inside each timepoint's own FIRE peaks. Each region gets an absolute `fire_rank` and
`fp_rank`, and the regions are put in three rank orderings:

1. `open_fp_poor_gap` (largest first): high FIRE frequency, few footprints
2. `open_fp_rich_sum` (smallest first): high FIRE frequency, many footprints
3. `closed_fp_rich_gap` (largest first): low FIRE frequency, many footprints

All filtering, ranking, correlations, summaries and plots use the pooled values. There
are no cutoffs or categories; the pooled coverage window is the only filter

per read FIRE elements `/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/[SAMPLE]/additional-outputs-v0.1/fire-peaks/[SAMPLE]-v0.1-fire-elements.bed.gz`


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE}

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(patchwork)
})
source("/project/spott/cshan/fiber-seq/code/parsing_functions/plotting_functions.r", local = TRUE)

input_dir      <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency"
plot_dir       <- file.path(input_dir, "TF_FIRE", "footprints_in_FIRE")
annotation_dir <- "/project/spott/cshan/annotations"
timepoints     <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")
```

Parameters:

- `fp_rank_col`: the column used for `fp_rank`. Options are
  `fp_count_all`, `fp_count_fire` and `fp_count_nonfire` (union region), or `fp_peak_count_all`,
  `fp_peak_count_fire` and `fp_peak_count_nonfire` (per-timepoint FIRE peaks).
- `top_n_show`: how many rows the top tables show. It is for display only and filters nothing.
- `top_n_region_plots`: how many top regions per ordering get a read-level region plot (section 11).
- `region_pad`: how many bp of flank to show on each side of the region in those plots.

```{r parameters}
fp_rank_col <- "fp_count_all"
top_n_show  <- 20
top_n_region_plots <- 2
region_pad  <- 1000
```


# 1. Load the tables

- `feature_count_long.tsv.gz` (03b_feature_count.py): one row per region and timepoint
- `n_reads` = reads spanning the whole region, `n_fire` = those with a FIRE element covering >= 50% of it, `fp_*` = FiberHMM TF footprints
  (score >= 50, 10-80 bp) lying fully inside the region `_gated` columns count footprints on the FIRE-positive reads only.
- `peak_bp` = bp of the union region covered by that timepoint's own FIRE peaks, `fp_peak_*` = the same footprints,
  same reads, but only those lying fully inside one of those peaks
- the GENCODE v46 Ensembl-canonical TSSs (1 bp BED, start = 0-based TSS)

```{r load_tables}
load_tables <- function(input_dir, annotation_dir) {
  long <- read_tsv(file.path(input_dir, "feature_count_long.tsv.gz"), show_col_types = FALSE) %>%
    mutate(timepoint = factor(timepoint, levels = timepoints))
  tss <- read_tsv(file.path(annotation_dir, "gencodev46_Ensembl_canonical_TSS.bed"),
                  show_col_types = FALSE,
                  col_names = c("chrom", "start", "end", "name", "score", "strand"))

  # format checks: one long row per region x timepoint, n_fire is the FIRE read count
  stopifnot(!anyNA(long$timepoint),
            !anyDuplicated(long[c("region_id", "timepoint")]),
            nrow(long) == n_distinct(long$region_id) * length(timepoints),
            all(long$n_fire == long$fire_n_reads),
            all(tss$end - tss$start == 1))
  list(long = long, tss = tss)
}

tables <- load_tables(input_dir, annotation_dir)
```

| Column | Meaning |
|---|---|
| `region_id` | Unique region name, `chrom:start-end`. Use it to join to `fire_frequency_long`. |
| `chrom` | Chromosome of the region. |
| `start` | Region start (BED coordinates, 0-based). |
| `end` | Region end. |
| `width` | Region length in bp (`end - start`). |
| `timepoint` | LPS timepoint (LPS_0, LPS_5, LPS_10, LPS_15). |
| `n_reads` | Number of reads whose alignment covers 100% of the region. Denominator for the per-read rates. |
| `n_fire` | Number of spanning reads with a FIRE element covering at least 50% of the region (the FIRE-positive reads). Identical to `fire_n_reads`. |
| `fire_n_elements` | Total FIRE elements covering at least 50% of the region on spanning reads. |
| `fire_n_reads` | Spanning reads with at least one such FIRE element. |
| `fp_n_elements` | Total footprints lying fully inside the region on any spanning read, open or not. |
| `fp_n_reads` | Spanning reads with at least one footprint inside the region. |
| `fp_n_elements_gated` | Footprints inside the region on FIRE-positive reads only. Numerator for `fp_per_fire_read`. |
| `fp_n_reads_gated` | FIRE-positive reads with at least one footprint inside the region. Numerator for `fp_freq_given_fire`. |
| `fire_freq` | Fraction of spanning reads that are FIRE-positive (`fire_n_reads / n_reads`). Matches `freq` from script 03; measures how open the region is. |
| `fire_elements_per_read` | FIRE elements per spanning read (`fire_n_elements / n_reads`). |
| `fp_per_read` | Footprints per spanning read (`fp_n_elements / n_reads`), over all fibers. |
| `fp_per_fire_read` | Footprints per FIRE-positive read (`fp_n_elements_gated / n_fire`). Main footprint frequency; can exceed 1. NA when `n_fire = 0`. |
| `fp_freq_given_fire` | Fraction of FIRE-positive reads carrying at least one footprint (`fp_n_reads_gated / n_fire`). Bounded 0-1. NA when `n_fire = 0`. |
| `fp_per_fire_read_per_kb` | `fp_per_fire_read` divided by region width in kb, so wide and narrow regions can be compared. |
| `peak_bp` | bp of the region covered by this timepoint's FIRE peaks (peaks file columns 1-3, merged). 0 when this timepoint called no peak in the region. |
| `fp_peak_n_elements` | Footprints lying fully inside this timepoint's FIRE peaks, on any spanning read. 0 when `peak_bp = 0`. |
| `fp_peak_n_elements_gated` | The same, on FIRE-positive reads only. |


# 2. Pool the timepoints

For each region, the counts are summed over the four timepoints. Every pooled rate below is
recomputed from these summed counts; per-timepoint rates are never averaged.

A timepoint that called no FIRE peak in the region adds reads but no peak footprints, so the peak
rates are divided only by the reads from timepoints that did:

- `n_reads_peak`: spanning reads from timepoints with `peak_bp > 0`
- `peak_bp`: the mean `peak_bp` over those timepoints, weighted by `n_reads`
- `fire_read_peak_kb`: sum over timepoints of `n_fire * peak_bp / 1000`

```{r pool_timepoints}
pool_timepoints <- function(long) {
  long %>%
    group_by(region_id, chrom, start, end, width) %>%
    # the peak terms come first: they need the per-timepoint n_reads and n_fire
    summarise(n_reads_peak = sum(n_reads[peak_bp > 0]),
              fire_read_peak_kb = sum(n_fire * peak_bp) / 1000,
              peak_bp = if_else(n_reads_peak > 0, sum(n_reads * peak_bp) / n_reads_peak,
                                NA_real_),
              across(c(n_reads, n_fire, fp_n_elements, fp_n_elements_gated,
                       fp_peak_n_elements, fp_peak_n_elements_gated), sum),
              .groups = "drop")
}

pooled <- pool_timepoints(tables$long)
```


# 3. Per-timepoint information columns

timepoint information is attached in `<column>_<timepoint>`,  `min_timepoint_n_reads` and `n_timepoints_with_reads` for metadata only

```{r timepoint_info}
timepoint_info <- function(long) {
  per_timepoint <- long %>%
    select(region_id, timepoint, n_reads, n_fire, fp_n_elements, fp_n_elements_gated, fire_freq,
           peak_bp, fp_peak_n_elements) %>%
    pivot_wider(names_from = timepoint, names_sort = TRUE,
                values_from = c(n_reads, n_fire, fp_n_elements, fp_n_elements_gated, fire_freq,
                                peak_bp, fp_peak_n_elements))
  long %>%
    group_by(region_id) %>%
    summarise(min_timepoint_n_reads = min(n_reads),
              n_timepoints_with_reads = sum(n_reads > 0), .groups = "drop") %>%
    left_join(per_timepoint, by = "region_id")
}

info <- timepoint_info(tables$long)
info_cols <- setdiff(names(info), "region_id")
pooled <- left_join(pooled, info, by = "region_id")
head(pooled)
```


# 4. Coverage filter

It uses the pooled `n_reads`: a region is kept when max(M - 3 sqrt(M), 10) <= n_reads <= M + 3 sqrt(M), where M is the median pooled `n_reads` over all regions
  - coverage filter applied on all reads combining all timepoints


```{r coverage_filter}
cov_median <- median(pooled$n_reads)
cov_lower  <- max(cov_median - 3 * sqrt(cov_median), 10)
cov_upper  <- cov_median + 3 * sqrt(cov_median)

tibble(median_pooled_n_reads = cov_median, lower = cov_lower, upper = cov_upper,
       regions_before = nrow(pooled),
       regions_after = sum(between(pooled$n_reads, cov_lower, cov_upper)),
       removed_below = sum(pooled$n_reads < cov_lower),
       removed_above = sum(pooled$n_reads > cov_upper)) %>%
  knitr::kable(digits = 1)

regions <- pooled %>% filter(between(n_reads, cov_lower, cov_upper))
```


# 5. Pooled metrics

There is one row per region, and every value comes from the pooled counts.

- `fire_freq` = `n_fire / n_reads`
- `fp_count_all` = summed `fp_n_elements`: footprints fully inside the region on any spanning read.
  This is the main metric.
- `fp_count_fire` = summed `fp_n_elements_gated`: footprints on FIRE-positive reads
- `fp_count_nonfire` = `fp_count_all - fp_count_fire`
- `fp_per_read` = `fp_count_all / n_reads`
- `fp_per_fire_read_per_kb` = `fp_count_fire / n_fire / (width / 1000)`: a secondary metric. It is
  NA when `n_fire == 0`.

The same metrics on the footprints inside each timepoint's FIRE peaks:

- `fp_peak_count_all` = summed `fp_peak_n_elements`, `fp_peak_count_fire` = summed
  `fp_peak_n_elements_gated`, `fp_peak_count_nonfire` = `fp_peak_count_all - fp_peak_count_fire`
- `fp_peak_per_read` = `fp_peak_count_all / n_reads_peak`. NA when `n_reads_peak == 0`.
- `fp_peak_per_fire_read_per_kb` = `fp_peak_count_fire / fire_read_peak_kb`. NA when
  `fire_read_peak_kb == 0`.

```{r metrics}
regions <- regions %>%
  rename(fp_count_all = fp_n_elements, fp_count_fire = fp_n_elements_gated,
         fp_peak_count_all = fp_peak_n_elements, fp_peak_count_fire = fp_peak_n_elements_gated) %>%
  mutate(fp_count_nonfire = fp_count_all - fp_count_fire,
         fire_freq = n_fire / n_reads,
         fp_per_read = fp_count_all / n_reads,
         fp_per_fire_read_per_kb = if_else(n_fire > 0, fp_count_fire / n_fire / (width / 1000),
                                           NA_real_),
         .after = fp_count_fire) %>%
  mutate(fp_peak_count_nonfire = fp_peak_count_all - fp_peak_count_fire,
         fp_peak_per_read = if_else(n_reads_peak > 0, fp_peak_count_all / n_reads_peak, NA_real_),
         fp_peak_per_fire_read_per_kb = if_else(fire_read_peak_kb > 0,
                                                fp_peak_count_fire / fire_read_peak_kb, NA_real_),
         .after = fp_peak_count_fire)
stopifnot(all(regions$fp_count_nonfire >= 0), all(regions$fp_peak_count_nonfire >= 0))

head(regions)
```


### Relationship between footprint frequency and FIRE frequency in a given FIRE region

Footprints are counted inside each timepoint's own FIRE peaks; `fire_freq` is still the union region's.
Only regions with a FIRE peak at all four timepoints are used here. In the other regions, the
timepoints without a peak add reads but no peak footprints, and those regions are mostly the less
open ones, which would tie the peak counts to `fire_freq`.

 - fp_peak_count_all: total footprints in the peaks on all reads
 - fp_peak_count_fire: footprints on FIRE (open) reads
 - fp_peak_count_nonfire: footprints on non-FIRE (closed) reads
 - fp_peak_per_read: total footprints divided by reads from timepoints that called a peak
 - fp_peak_per_fire_read_per_kb: footprint density on open fibers, per kb of peak
 - rho_peak_bp: correlation with the peak width (`peak_bp`)

 1. More open regions have many more footprints in total
 2. footprints on closed fibers don't change with openess, even though open regions have fewer closed fibers
 3. per open fiber and per kb, footprint density barely rises with openness
 4. Wider FIRE regions tend to have more footprints

```{r correlations}
spearman <- function(x, y) cor(x, y, method = "spearman", use = "complete.obs")

all_peak_regions <- regions %>%
  filter(if_all(all_of(paste0("peak_bp_", timepoints)), ~ .x > 0))
cat(nrow(all_peak_regions), "of", nrow(regions), "regions have a FIRE peak at all four timepoints\n")

fp_metrics <- c("fp_peak_count_all", "fp_peak_count_fire", "fp_peak_count_nonfire",
                "fp_peak_per_read", "fp_peak_per_fire_read_per_kb")
tibble(metric = fp_metrics,
       n = sapply(fp_metrics, function(m) sum(!is.na(all_peak_regions[[m]]))),
       rho_fire_freq = sapply(fp_metrics, function(m)
         spearman(all_peak_regions$fire_freq, all_peak_regions[[m]])),
       rho_peak_bp = sapply(fp_metrics, function(m)
         spearman(all_peak_regions$peak_bp, all_peak_regions[[m]])),
       rho_n_reads = sapply(fp_metrics, function(m)
         spearman(all_peak_regions$n_reads, all_peak_regions[[m]]))) %>%
  knitr::kable(digits = 3)
```


 

# Rank region by footprint and FIRE frequency

The ranks are absolute, from 1 to the number of regions. Ties share the lowest rank
(`min_rank`), and NA values stay NA and sort last.

- `fire_rank` ranks `fire_freq`. 1 = most open.
- `fp_rank` ranks the column named by `fp_rank_col`. 1 = most footprints.

| Pattern | Ranks | Sort by |
|---|---|---|
| Open, footprint-poor | `fire_rank` small, `fp_rank` large | `open_fp_poor_gap = fp_rank - fire_rank`, largest first |
| Open, footprint-rich | both small | `open_fp_rich_sum = fire_rank + fp_rank`, smallest first |
| Closed, footprint-rich | `fire_rank` large, `fp_rank` small | `closed_fp_rich_gap = fire_rank - fp_rank`, largest first |

The sum and the gaps let one rank make up for the other: a region can sort high by being very
extreme on one axis and only moderate on the other. They do not require both ranks to be extreme.

```{r ranks}
add_ranks <- function(regions, fp_col) {
  regions %>%
    mutate(fire_rank = min_rank(desc(fire_freq)),
           fp_rank = min_rank(desc(.data[[fp_col]])),
           open_fp_poor_gap = fp_rank - fire_rank,
           open_fp_rich_sum = fire_rank + fp_rank,
           closed_fp_rich_gap = fire_rank - fp_rank)
}

regions <- add_ranks(regions, fp_rank_col)

head(regions)
```


# 8. Annotation

`tss_distance` is the distance in bp to the nearest Ensembl-canonical TSS (0 = the region
contains it).

```{r annotate_regions}
annotate_regions <- function(regions, tss) {
  reg.gr <- GRanges(regions$chrom, IRanges(regions$start + 1L, regions$end))

  # the TSS BED start is the 0-based TSS
  tss.gr <- GRanges(tss$chrom, IRanges(tss$start + 1L, width = 1))
  nearest <- distanceToNearest(reg.gr, tss.gr)
  tss_distance <- rep(NA_integer_, length(reg.gr))
  tss_distance[queryHits(nearest)] <- mcols(nearest)$distance

  mutate(regions, tss_distance = tss_distance)
}

regions <- annotate_regions(regions, tables$tss)
```


# 9. Output table

`region_scores_pooled.tsv.gz` has one row per kept region: pooled counts, rates, ranks and
`tss_distance`. The per-timepoint information columns come last.

```{r write_table}
regions <- relocate(regions, all_of(info_cols), .after = last_col())
write_tsv(regions, file.path(input_dir, "region_scores_pooled.tsv.gz"))
```

The top regions of each ordering:

```{r top_regions}
# each ordering and whether it sorts largest first
orderings <- c(open_fp_poor_gap = TRUE, open_fp_rich_sum = FALSE, closed_fp_rich_gap = TRUE)

# arrange() always sorts NA last, also under desc()
top_by <- function(df, sort_col, decreasing, n) {
  df %>%
    arrange(if (decreasing) desc(.data[[sort_col]]) else .data[[sort_col]]) %>%
    slice_head(n = n)
}

show_ranked <- function(regions, sort_col, decreasing, top_n) {
  top_by(regions, sort_col, decreasing, top_n) %>%
    select(region_id, fire_rank, fp_rank, all_of(sort_col), fire_freq, n_reads, n_fire,
           fp_count_all, fp_count_fire, fp_count_nonfire, width, tss_distance) %>%
    knitr::kable(digits = 3, caption = sprintf("Top %d regions by %s, %s first", top_n, sort_col,
                                               if (decreasing) "largest" else "smallest"))
}

show_ranked(regions, "open_fp_poor_gap", decreasing = TRUE, top_n_show)
show_ranked(regions, "open_fp_rich_sum", decreasing = FALSE, top_n_show)
show_ranked(regions, "closed_fp_rich_gap", decreasing = TRUE, top_n_show)
```

```{r top-ranked-regions}
ranked <- read_tsv(file.path(input_dir, "region_scores_pooled.tsv.gz"), show_col_types = FALSE) %>%
  mutate(fp_per_nonfire_read_per_kb = if_else(n_reads > n_fire,
           fp_count_nonfire / (n_reads - n_fire) / (width / 1000), NA_real_),
         nonfire_fraction = fp_count_nonfire / fp_count_all)

top_n <- 500

profile <- function(df, label) {
  df %>% summarise(
    set = label, regions = n(),
    fire_freq_median = median(fire_freq),
    width_median = median(width),
    width_iqr = paste(quantile(width, c(.25, .75)), collapse = "-"),
    pct_at_tss = 100 * mean(tss_distance == 0, na.rm = TRUE),
    tss_distance_median = median(tss_distance, na.rm = TRUE),
    fp_per_fire_read_per_kb = median(fp_per_fire_read_per_kb, na.rm = TRUE),
    fp_per_nonfire_read_per_kb = median(fp_per_nonfire_read_per_kb, na.rm = TRUE),
    pct_fp_on_nonfire = 100 * median(nonfire_fraction, na.rm = TRUE))
}

bind_rows(
  profile(ranked, "all regions"),
  lapply(names(orderings), function(o) profile(top_by(ranked, o, orderings[[o]], top_n), o))) %>%
  knitr::kable(digits = 2)
```

`open_fp_rich_sum` sorts smallest first, so a negative rho there means the property is higher at
the top of that ordering.

```{r rank_correlation}
props <- c("width", "tss_distance", "fire_freq", "n_reads",
           "fp_per_fire_read_per_kb", "fp_per_nonfire_read_per_kb")

expand_grid(ordering = names(orderings), property = props) %>%
  rowwise() %>%
  mutate(rho = cor(ranked[[ordering]], ranked[[property]],
                   method = "spearman", use = "complete.obs")) %>%
  ungroup() %>%
  pivot_wider(names_from = property, values_from = rho) %>%
  knitr::kable(digits = 2)
```

# 10. Plots

## FIRE frequency vs footprint counts in the FIRE peaks

Each panel shows the pooled values, one hexbin of regions per panel. The footprints are those
inside each timepoint's FIRE peaks. The y axis is on a log1p scale.

```{r plot_count_hex, fig.width=5.5, fig.height=10}
plot_count_hex <- function(regions, y_col) {
  label <- sprintf("rho = %.2f\nn = %s", spearman(regions$fire_freq, regions[[y_col]]),
                   scales::comma(sum(!is.na(regions[[y_col]]))))
  ggplot(regions, aes(fire_freq, .data[[y_col]])) +
    geom_hex(bins = 60) +
    annotate("text", x = -Inf, y = Inf, label = label, hjust = -0.1, vjust = 1.2, size = 2.8) +
    scale_y_continuous(transform = "log1p", breaks = c(0, 3, 10, 30, 100, 300, 1000)) +
    scale_fill_viridis_c(transform = "log10", name = "regions") +
    labs(x = "FIRE frequency (pooled)", y = paste(y_col, "(log1p scale)")) +
    theme_fiberseq("classic")
}

p_count_hex <- plot_count_hex(regions, "fp_peak_count_all") /
  plot_count_hex(regions, "fp_peak_count_fire") /
  plot_count_hex(regions, "fp_peak_count_nonfire")
save_figure(p_count_hex, file.path(plot_dir, "fire_freq_vs_fp_peak_counts_hex.pdf"),
            width = 5.5, height = 10)
p_count_hex
```

## The same, colored by peak width

Each hex is colored by the median `peak_bp` of its regions.

```{r plot_width_hex, fig.width=5.5, fig.height=3.5}
plot_summary_hex <- function(regions, y_col, z_col, fun, fill_lab) {
  ggplot(regions, aes(fire_freq, .data[[y_col]], z = .data[[z_col]])) +
    stat_summary_hex(fun = fun, bins = 60) +
    scale_y_continuous(transform = "log1p", breaks = c(0, 3, 10, 30, 100, 300, 1000)) +
    scale_fill_viridis_c(name = fill_lab) +
    labs(x = "FIRE frequency (pooled)", y = paste(y_col, "(log1p scale)")) +
    theme_fiberseq("classic")
}

p_width_hex <- plot_summary_hex(regions, "fp_peak_count_all", "peak_bp", "median",
                                "median peak_bp (bp)")
save_figure(p_width_hex,
            file.path(plot_dir, "fire_freq_vs_fp_peak_count_all_median_peak_bp_hex.pdf"),
            width = 5.5, height = 3.5)
p_width_hex
```

## FIRE rank vs footprint rank

Both axes are absolute ranks and are reversed, so the most open and most footprinted regions sit
at the top right.

```{r plot_rank_hex, fig.width=5, fig.height=4.5}
plot_rank_hex <- function(regions, fp_col) {
  ggplot(regions, aes(fire_rank, fp_rank)) +
    geom_hex(bins = 50) +
    scale_x_reverse() +
    scale_y_reverse() +
    scale_fill_viridis_c(name = "regions") +
    coord_equal() +
    labs(x = "fire_rank (1 = most open)", y = paste0("fp_rank of ", fp_col, " (1 = most footprints)")) +
    theme_fiberseq("classic")
}

p_rank_hex <- plot_rank_hex(regions, fp_rank_col)
save_figure(p_rank_hex, file.path(plot_dir, "fire_rank_vs_fp_rank_hex.pdf"),
            width = 5, height = 4.5)
p_rank_hex
```


# 11. Region plots: top regions per ordering

This section plots the top `top_n_region_plots` regions of each ordering with `plot_region_panels()`
from `plotting_functions.r`. Each window is the region +/- `region_pad` bp, and the yellow band
marks the region itself.

- Only reads that span the whole region are shown, which is the read set counted in `n_reads`.
- The accessibility panel is one panel with one m6A-fraction line per timepoint, in the
  `LEIDEN_TIMEPOINT_COLORS` of `plotting_functions.r`. `plot_region_panels()` would draw one panel
  per timepoint, so this panel is drawn with `plot_group_profile()` instead. CpG is not drawn.
- The FIRE peaks panel has one row per timepoint, LPS_0 at the top, each showing that
  timepoint's own peak calls in the window, colored by -log10(FDR). The yellow band is their union,
  the region the reads and counts are taken from. `plot_region_panels()` can only draw the peaks
  of all timepoints on one row, so `plot_timepoint_peaks()` below draws this panel.
- The SMF reads and footprint panels are split by timepoint, LPS_0 at the top. The footprint
  panel shows every FiberHMM TF footprint (score >= 50, 10-80 bp), not only the footprints inside
  FIRE elements that `plot_region_panels()` draws by default. This matches `fp_count_all`, which
  counts footprints on every spanning read.
- The figure is reassembled from these panels and the reads and footprint panels of
  `plot_region_panels()`.
- The subtitle gives the FIRE-positive read count (an element from the FIRE per-read elements file
  covering >= 50% of the region, the `03b_feature_count.py` rule) and the footprint counts for the
  whole region and for inside the peaks.

```{r region_plot_setup}
code_dir <- "/project/spott/cshan/fiber-seq/code/parsing_functions"
# packages the region helpers attached before they moved into parsing_functions
suppressPackageStartupMessages({
  library(data.table); library(dplyr); library(GenomicRanges); library(IRanges)
  library(ggplot2); library(patchwork)
})
source(file.path(code_dir, "parsing_footprints_functions.r"))
source(file.path(code_dir, "plotting_functions.r"), local = TRUE)
extract_script <- file.path(code_dir, "extract_region_result_macrophage.sh")
extract_root   <- file.path(input_dir, "region_plots")

top_regions <- bind_rows(lapply(names(orderings), function(o)
    top_by(regions, o, orderings[[o]], top_n_region_plots) %>%
      mutate(score = o, score_value = .data[[o]], rank = row_number()))) %>%
  mutate(outname = paste(chrom, start, end, sep = "_"),
         window_start = as.integer(start - region_pad),
         window_end   = as.integer(end + region_pad))

top_regions %>%
  select(score, rank, region_id, score_value, fire_freq, n_reads, n_fire, fp_count_all,
         fp_count_fire, fp_count_nonfire, width, tss_distance) %>%
  knitr::kable(digits = 3)
```

Each region and timepoint is extracted once into `region_plots/<chrom_start_end>/<timepoint>/parsed`
by `plot_region_example()`, in the same call that plots it. The extraction is skipped if
`fire.bed` already exists there.

The subtitle compares the plotted read counts with the table's `n_reads` and `n_fire`. They
can differ slightly, because the plot takes each read's span from `fire.bed` and drops reads with no
m6A call in the window.

```{r region_plots, fig.width=10, fig.height=14}
fire_positive_reads <- function(parsed_dirs, region_start, region_end) {
  elements <- bind_rows(lapply(file.path(parsed_dirs, "fire_elements.bed"), function(f)
    read_tsv(f, col_names = c("chrom", "start", "end", "RID"), col_select = 1:4,
             col_types = "ciic")))
  overlap <- pmin(elements$end, region_end) - pmax(elements$start, region_start)
  unique(elements$RID[overlap >= 0.5 * (region_end - region_start)])
}

# one row per timepoint (LPS_0 on top) with that timepoint's own FIRE peak calls, colored like
# the single-row FIRE peaks track of plot_region_panels(); highlight = the union region
plot_timepoint_peaks <- function(parsed_dirs, timepoints, xlims, highlight) {
  peaks <- bind_rows(lapply(seq_along(timepoints), function(i) {
    pk <- read_fire_peaks_region(file.path(parsed_dirs[i], "fire_peaks.bed"))
    if (is.null(pk)) NULL else mutate(pk, y = length(timepoints) + 1 - i)
  }))
  ggplot(peaks, aes(x = start, xend = end, y = y, yend = y, colour = logFDR)) +
    annotate("rect", xmin = highlight[1], xmax = highlight[2], ymin = -Inf, ymax = Inf,
             fill = "yellow", alpha = 0.3) +
    geom_segment(linewidth = 1.5) +
    scale_colour_gradient2(low = "black", mid = "gray", high = "red", midpoint = 1.3,
                           limits = c(0, 5.2), oob = scales::squish, name = "-log10(FDR)") +
    scale_y_continuous(breaks = seq_along(timepoints), labels = rev(timepoints)) +
    coord_cartesian(xlim = xlims, ylim = c(0.5, length(timepoints) + 0.5), expand = FALSE) +
    labs(x = NULL, y = "FIRE peaks") +
    theme_fiberseq("bw", base_size = 11)
}

plot_top_region <- function(row, extract_root, timepoints) {
  parsed_dirs <- file.path(extract_root, row$outname, timepoints, "parsed")

  # draw every TF footprint, not only those inside a FIRE element, and split the
  # spanning reads by timepoint; the pooled one-row FIRE peaks track is dropped,
  # plot_timepoint_peaks() replaces it
  group_reads <- function(res) {
    res$fire_peaks <- NULL
    res$fps_infire <- res$fps
    spans <- res$rids_df[res$rids_df$start <= row$start & res$rids_df$end >= row$end, ]
    labels <- setNames(as.character(spans$sample_name), as.character(spans$RID))
    add_read_groups(res, labels, group_col = "timepoint", group_levels = timepoints,
                    verbose = FALSE)
  }

  panel_options <- list(highlight = c(row$start, row$end),
                        y_labels = c(pileup = "Met. prop.", reads = "SMF reads",
                                     fire_fiberHMM = "FiberHMM TF footprints (all)"))

  out <- plot_region_example(
    region       = data.frame(chr = row$chrom, start = row$window_start, end = row$window_end),
    sample_names = timepoints,
    out_root     = extract_root,
    outname      = row$outname,
    script       = extract_script,
    load_options = list(fiberHMM_feature      = "tf",
                        min_fiberHMM_fp_score = 50,
                        fiberHMM_size_breaks  = c(10, 30, 60, 80),
                        verbose = FALSE),
    group = group_reads,
    panel_options = panel_options)

  # the subtitle counts the plotted (spanning) reads
  res <- out$res
  fire_pos <- fire_positive_reads(parsed_dirs, row$start, row$end)
  subtitle <- sprintf(paste0("%s = %.0f | fire_freq %.2f | %d bp | TSS distance %s bp\n",
                             "spanning reads plotted %d (table n_reads %d), FIRE-positive %d ",
                             "(table n_fire %d)\nfootprints all %d, FIRE %d, non-FIRE %d | ",
                             "in peaks all %d, FIRE %d, non-FIRE %d"),
                      row$score, row$score_value, row$fire_freq, row$width,
                      format(row$tss_distance, big.mark = ","),
                      nrow(res$rids_df), row$n_reads,
                      sum(as.character(res$rids_df$RID) %in% fire_pos), row$n_fire,
                      row$fp_count_all, row$fp_count_fire, row$fp_count_nonfire,
                      row$fp_peak_count_all, row$fp_peak_count_fire, row$fp_peak_count_nonfire)

  # one accessibility panel, one m6A line per timepoint
  xlims <- c(res$region$start, res$region$end)
  accessibility <- plot_group_profile(
    filter(res$pileup, base == "A"), group_col = "timepoint", y_col = "smooth_frac",
    colors = LEIDEN_TIMEPOINT_COLORS,
    highlight = list(start = row$start, end = row$end, fill = "yellow", alpha = 0.3,
                     position = "under"),
    scales = list(scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))),
    theme = theme_fiberseq("bw", base_size = 11),
    labels = list(x = NULL, y = "m6A prop.", colour = "timepoint",
                  title = row$region_id, subtitle = subtitle)) +
    coord_cartesian(xlim = xlims, expand = FALSE)

  peaks <- plot_timepoint_peaks(parsed_dirs, timepoints, xlims = xlims,
                                highlight = c(row$start, row$end))
  panels <- out$plot$panels
  wrap_plots(accessibility, peaks, panels$reads, panels$fire_fiberHMM, ncol = 1,
             heights = c(4, 2.5, 10, 10))
}

for (i in seq_len(nrow(top_regions))) {
  row <- top_regions[i, ]
  p <- plot_top_region(row, extract_root, timepoints)
  save_figure(p, file.path(plot_dir, "region_plots",
                           sprintf("%s_%d_%s.pdf", row$score, row$rank, row$outname)),
              width = 10, height = 14)
  print(p)
}
```
