---
title: "footprints at FIRE regions"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

# FIRE frequency vs absolute footprint counts, timepoints pooled

The four LPS timepoints are treated as one pooled sample. For each region, this notebook relates
FIRE frequency to the absolute number of TF footprints in that region. Regions are then ranked on
three continuous scores:

1. `open_fp_poor_score`: high FIRE frequency, few footprints
2. `open_fp_rich_score`: high FIRE frequency, many footprints
3. `closed_fp_rich_score`: low FIRE frequency, many footprints

All filtering, ranking, scoring, correlations, summaries and plots use the pooled values. There
are no cutoffs or categories; the pooled coverage window is the only filter.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE}

suppressPackageStartupMessages({
  library(GenomicRanges)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(patchwork)
})
source("/project/spott/cshan/fiber-seq/code/parsing_functions/plotting_functions.r", local = TRUE)

input_dir      <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency"
plot_dir       <- file.path(input_dir, "TF_FIRE", "footprints_in_FIRE")
annotation_dir <- "/project/spott/cshan/annotations"
timepoints     <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")
```

Parameters:

- `fp_rank_col`: the footprint count used for `fp_rank` and for the scores. Options are
  `fp_count_all`, `fp_count_fire` and `fp_count_nonfire`.
- `top_n_show`: how many rows the top tables show. It is for display only and filters nothing.
- `top_n_region_plots`: how many top regions per score get a read-level region plot (section 11).
- `region_pad`: how many bp of flank to show on each side of the region in those plots.

```{r parameters}
fp_rank_col <- "fp_count_all"
top_n_show  <- 20
top_n_region_plots <- 2
region_pad  <- 1000
```


# 1. Load the tables

- `feature_count_long.tsv.gz` (03b_feature_count.py): one row per region and timepoint
- `n_reads` = reads spanning the whole region, `n_fire` = those with a FIRE element covering >= 50% of it, `fp_*` = FiberHMM TF footprints
  (score >= 50, 10-80 bp) lying fully inside the region `_gated` columns count footprints on the FIRE-positive reads only.
- the GENCODE v46 Ensembl-canonical TSSs (1 bp BED, start = 0-based TSS)



```{r load_tables}
load_tables <- function(input_dir, annotation_dir) {
  long <- read_tsv(file.path(input_dir, "feature_count_long.tsv.gz"), show_col_types = FALSE) %>%
    mutate(timepoint = factor(timepoint, levels = timepoints))
  tss <- read_tsv(file.path(annotation_dir, "gencodev46_Ensembl_canonical_TSS.bed"),
                  show_col_types = FALSE,
                  col_names = c("chrom", "start", "end", "name", "score", "strand"))

  # format checks: one long row per region x timepoint, n_fire is the FIRE read count
  stopifnot(!anyNA(long$timepoint),
            !anyDuplicated(long[c("region_id", "timepoint")]),
            nrow(long) == n_distinct(long$region_id) * length(timepoints),
            all(long$n_fire == long$fire_n_reads),
            all(tss$end - tss$start == 1))
  list(long = long, tss = tss)
}

tables <- load_tables(input_dir, annotation_dir)
```

| Column | Meaning |
|---|---|
| `region_id` | Unique region name, `chrom:start-end`. Use it to join to `fire_frequency_long`. |
| `chrom` | Chromosome of the region. |
| `start` | Region start (BED coordinates, 0-based). |
| `end` | Region end. |
| `width` | Region length in bp (`end - start`). |
| `timepoint` | LPS timepoint (LPS_0, LPS_5, LPS_10, LPS_15). |
| `n_reads` | Number of reads whose alignment covers 100% of the region. Denominator for the per-read rates. |
| `n_fire` | Number of spanning reads with a FIRE element covering at least 50% of the region (the FIRE-positive reads). Identical to `fire_n_reads`. |
| `fire_n_elements` | Total FIRE elements covering at least 50% of the region on spanning reads. |
| `fire_n_reads` | Spanning reads with at least one such FIRE element. |
| `fp_n_elements` | Total footprints lying fully inside the region on any spanning read, open or not. |
| `fp_n_reads` | Spanning reads with at least one footprint inside the region. |
| `fp_n_elements_gated` | Footprints inside the region on FIRE-positive reads only. Numerator for `fp_per_fire_read`. |
| `fp_n_reads_gated` | FIRE-positive reads with at least one footprint inside the region. Numerator for `fp_freq_given_fire`. |
| `fire_freq` | Fraction of spanning reads that are FIRE-positive (`fire_n_reads / n_reads`). Matches `freq` from script 03; measures how open the region is. |
| `fire_elements_per_read` | FIRE elements per spanning read (`fire_n_elements / n_reads`). |
| `fp_per_read` | Footprints per spanning read (`fp_n_elements / n_reads`), over all fibers. |
| `fp_per_fire_read` | Footprints per FIRE-positive read (`fp_n_elements_gated / n_fire`). Main footprint frequency; can exceed 1. NA when `n_fire = 0`. |
| `fp_freq_given_fire` | Fraction of FIRE-positive reads carrying at least one footprint (`fp_n_reads_gated / n_fire`). Bounded 0-1. NA when `n_fire = 0`. |
| `fp_per_fire_read_per_kb` | `fp_per_fire_read` divided by region width in kb, so wide and narrow regions can be compared. |


# 2. Pool the timepoints

For each region, the counts are summed over the four timepoints. Every pooled rate below is
recomputed from these summed counts; per-timepoint rates are never averaged.

```{r pool_timepoints}
pool_timepoints <- function(long) {
  long %>%
    group_by(region_id, chrom, start, end, width) %>%
    summarise(across(c(n_reads, n_fire, fp_n_elements, fp_n_elements_gated), sum),
              .groups = "drop")
}

pooled <- pool_timepoints(tables$long)
```


# 3. Per-timepoint information columns

The per-timepoint values are attached as wide columns named `<column>_<timepoint>`, plus
`min_timepoint_n_reads` and `n_timepoints_with_reads`. They are carried for reference only and are
not used in filtering, ranking, scoring, correlations or plots.

```{r timepoint_info}
# reference only: these columns never enter a filter, rank, score, correlation or plot
timepoint_info <- function(long) {
  per_timepoint <- long %>%
    select(region_id, timepoint, n_reads, n_fire, fp_n_elements, fp_n_elements_gated, fire_freq) %>%
    pivot_wider(names_from = timepoint, names_sort = TRUE,
                values_from = c(n_reads, n_fire, fp_n_elements, fp_n_elements_gated, fire_freq))
  long %>%
    group_by(region_id) %>%
    summarise(min_timepoint_n_reads = min(n_reads),
              n_timepoints_with_reads = sum(n_reads > 0), .groups = "drop") %>%
    left_join(per_timepoint, by = "region_id")
}

info <- timepoint_info(tables$long)
info_cols <- setdiff(names(info), "region_id")
pooled <- left_join(pooled, info, by = "region_id")
```


# 4. Coverage filter

This is the only filter. It uses the pooled `n_reads`: a region is kept when
max(M - 3 sqrt(M), 10) <= n_reads <= M + 3 sqrt(M), where M is the median pooled `n_reads` over all
regions. The kept regions all have similar total coverage, so their absolute footprint counts can
be compared directly, without dividing by depth.

```{r coverage_filter}
cov_median <- median(pooled$n_reads)
cov_lower  <- max(cov_median - 3 * sqrt(cov_median), 10)
cov_upper  <- cov_median + 3 * sqrt(cov_median)

tibble(median_pooled_n_reads = cov_median, lower = cov_lower, upper = cov_upper,
       regions_before = nrow(pooled),
       regions_after = sum(between(pooled$n_reads, cov_lower, cov_upper)),
       removed_below = sum(pooled$n_reads < cov_lower),
       removed_above = sum(pooled$n_reads > cov_upper)) %>%
  knitr::kable(digits = 1)

regions <- pooled %>% filter(between(n_reads, cov_lower, cov_upper))
```


# 5. Pooled metrics

There is one row per region, and every value comes from the pooled counts.

- `fire_freq` = `n_fire / n_reads`
- `fp_count_all` = summed `fp_n_elements`: footprints fully inside the region on any spanning read.
  This is the main metric.
- `fp_count_fire` = summed `fp_n_elements_gated`: footprints on FIRE-positive reads
- `fp_count_nonfire` = `fp_count_all - fp_count_fire`
- `fp_per_read` = `fp_count_all / n_reads`
- `fp_per_fire_read_per_kb` = `fp_count_fire / n_fire / (width / 1000)`: a secondary metric. It is
  NA when `n_fire == 0`.

```{r metrics}
regions <- regions %>%
  rename(fp_count_all = fp_n_elements, fp_count_fire = fp_n_elements_gated) %>%
  mutate(fp_count_nonfire = fp_count_all - fp_count_fire,
         fire_freq = n_fire / n_reads,
         fp_per_read = fp_count_all / n_reads,
         fp_per_fire_read_per_kb = if_else(n_fire > 0, fp_count_fire / n_fire / (width / 1000),
                                           NA_real_),
         .after = fp_count_fire)
stopifnot(all(regions$fp_count_nonfire >= 0))
```


# 6. FIRE frequency vs footprint counts

Spearman correlations on the pooled values. The first column is each footprint metric against
`fire_freq`. The other two columns check how much each metric follows region width and depth.

`fp_count_fire` is partly tied to `fire_freq` by construction: the more reads are FIRE-positive,
the more reads its footprints are counted on. `fp_count_all` and `fp_count_nonfire` are therefore
the fairer readouts.

```{r correlations}
spearman <- function(x, y) cor(x, y, method = "spearman", use = "complete.obs")

fp_metrics <- c("fp_count_all", "fp_count_fire", "fp_count_nonfire", "fp_per_read",
                "fp_per_fire_read_per_kb")
tibble(metric = fp_metrics,
       n = sapply(fp_metrics, function(m) sum(!is.na(regions[[m]]))),
       rho_fire_freq = sapply(fp_metrics, function(m) spearman(regions$fire_freq, regions[[m]])),
       rho_width = sapply(fp_metrics, function(m) spearman(regions$width, regions[[m]])),
       rho_n_reads = sapply(fp_metrics, function(m) spearman(regions$n_reads, regions[[m]]))) %>%
  knitr::kable(digits = 3)
```

### Relationship between footprint frequency and FIRE frequency in a given FIRE region

1. More open regions have many more footprints in total
2. rise in total footprints is mostly because open regions have more open fibers, not because each open fiber gets more footprints
3. Footprints on closed fibers is not related with FIRE frequency
4. Wider FIRE regions tend to have more footprints


The counts by `fire_freq` bin. The bins are only a description of the relationship; they are not
used for ranking or scoring.

1. Most regions have FIRE peaks are moderately open

```{r counts_by_fire_bin}
regions %>%
  mutate(fire_freq_bin = cut(fire_freq, c(0, 0.1, 0.25, 0.5, 0.75, 1), include.lowest = TRUE)) %>%
  select(fire_freq_bin, fp_count_all, fp_count_fire, fp_count_nonfire) %>%
  pivot_longer(-fire_freq_bin, names_to = "count") %>%
  group_by(count = factor(count, levels = c("fp_count_all", "fp_count_fire", "fp_count_nonfire")),
           fire_freq_bin) %>%
  summarise(regions = n(), median = median(value),
            p25 = quantile(value, 0.25), p75 = quantile(value, 0.75), .groups = "drop") %>%
  knitr::kable(digits = 1)
```


# 7. Ranks and scores

The ranks are pooled percentile ranks from 0 to 1, with ties averaged and NAs kept as NA.

- `fire_rank` ranks `fire_freq` over all kept regions, including the `n_fire == 0` regions
  (`fire_freq = 0`). 1 = most open.
- `fp_rank` ranks the count named by `fp_rank_col`. 1 = most footprints.

Each score is the smaller of two ranks, so a region scores high only when both conditions hold. A
score of 0.9 means the region is in the top 10% on both axes, and being extreme on one axis does
not make up for the other.

- `open_fp_poor_score = pmin(fire_rank, 1 - fp_rank)`: the region is open and has few footprints.
- `open_fp_rich_score = pmin(fire_rank, fp_rank)`: the region is open and has many footprints.
- `closed_fp_rich_score = pmin(1 - fire_rank, fp_rank)`: the region is mostly closed but still has
  many footprints.
- `fp_resid_rank`: the percentile rank of the residual from `lm(log1p(<fp_rank_col>) ~ fire_freq)`.
  1 = more footprints than expected for that openness, 0 = fewer. It measures footprints relative
  to openness rather than requiring a particular level of openness.

```{r scores}
pct_rank <- function(x) {
  (rank(x, ties.method = "average", na.last = "keep") - 1) / (sum(!is.na(x)) - 1)
}

add_scores <- function(regions, fp_col) {
  fit <- lm(log1p(fp) ~ fire_freq, na.action = na.exclude,
            data = tibble(fp = regions[[fp_col]], fire_freq = regions$fire_freq))
  regions %>%
    mutate(fire_rank = pct_rank(fire_freq),
           fp_rank = pct_rank(.data[[fp_col]]),
           open_fp_poor_score = pmin(fire_rank, 1 - fp_rank),
           open_fp_rich_score = pmin(fire_rank, fp_rank),
           closed_fp_rich_score = pmin(1 - fire_rank, fp_rank),
           fp_resid_rank = pct_rank(unname(residuals(fit))))
}

regions <- add_scores(regions, fp_rank_col)
```


# 8. Annotation

`tss_distance` is the distance in bp to the nearest Ensembl-canonical TSS (0 = the region
contains it).

```{r annotate_regions}
annotate_regions <- function(regions, tss) {
  reg.gr <- GRanges(regions$chrom, IRanges(regions$start + 1L, regions$end))

  # the TSS BED start is the 0-based TSS
  tss.gr <- GRanges(tss$chrom, IRanges(tss$start + 1L, width = 1))
  nearest <- distanceToNearest(reg.gr, tss.gr)
  tss_distance <- rep(NA_integer_, length(reg.gr))
  tss_distance[queryHits(nearest)] <- mcols(nearest)$distance

  mutate(regions, tss_distance = tss_distance)
}

regions <- annotate_regions(regions, tables$tss)
```


# 9. Output table

`region_scores_pooled.tsv.gz` has one row per kept region: pooled counts, rates, ranks, scores and
`tss_distance`. The per-timepoint information columns come last.

```{r write_table}
regions <- relocate(regions, all_of(info_cols), .after = last_col())
write_tsv(regions, file.path(input_dir, "region_scores_pooled.tsv.gz"))
```

The top regions by each score:

```{r top_regions}
show_top <- function(regions, score, top_n) {
  regions %>%
    slice_max(.data[[score]], n = top_n, with_ties = FALSE) %>%
    select(region_id, all_of(score), fire_freq, n_reads, n_fire, fp_count_all, fp_count_fire,
           fp_count_nonfire, width, tss_distance) %>%
    knitr::kable(digits = 3, caption = sprintf("Top %d regions by %s", top_n, score))
}

show_top(regions, "open_fp_poor_score", top_n_show)
show_top(regions, "open_fp_rich_score", top_n_show)
show_top(regions, "closed_fp_rich_score", top_n_show)
```

```{r top-ranked-regions}
scores <- read_tsv(file.path(input_dir, "region_scores_pooled.tsv.gz"), show_col_types = FALSE) %>%
  mutate(fp_per_nonfire_read_per_kb = if_else(n_reads > n_fire,
           fp_count_nonfire / (n_reads - n_fire) / (width / 1000), NA_real_),
         nonfire_fraction = fp_count_nonfire / fp_count_all)

score_cols <- c("open_fp_poor_score", "open_fp_rich_score", "closed_fp_rich_score")
top_n <- 500

profile <- function(df, label) {
  df %>% summarise(
    set = label, regions = n(),
    fire_freq_median = median(fire_freq),
    width_median = median(width),
    width_iqr = paste(quantile(width, c(.25, .75)), collapse = "-"),
    pct_at_tss = 100 * mean(tss_distance == 0, na.rm = TRUE),
    tss_distance_median = median(tss_distance, na.rm = TRUE),
    fp_per_fire_read_per_kb = median(fp_per_fire_read_per_kb, na.rm = TRUE),
    fp_per_nonfire_read_per_kb = median(fp_per_nonfire_read_per_kb, na.rm = TRUE),
    pct_fp_on_nonfire = 100 * median(nonfire_fraction, na.rm = TRUE))
}

bind_rows(
  profile(scores, "all regions"),
  lapply(score_cols, function(s) profile(slice_max(scores, .data[[s]], n = top_n,
                                                   with_ties = FALSE), s))) %>%
  knitr::kable(digits = 2)
```


```{r score_correlation}
props <- c("width", "tss_distance", "fire_freq", "n_reads",
           "fp_per_fire_read_per_kb", "fp_per_nonfire_read_per_kb")

expand_grid(score = score_cols, property = props) %>%
  rowwise() %>%
  mutate(rho = cor(scores[[score]], scores[[property]],
                   method = "spearman", use = "complete.obs")) %>%
  ungroup() %>%
  pivot_wider(names_from = property, values_from = rho) %>%
  knitr::kable(digits = 2)
```

# 10. Plots

## FIRE frequency vs footprint counts

Each panel shows the pooled values, one hexbin of regions per panel. The y axis is on a log1p
scale.

```{r plot_count_hex, fig.width=5.5, fig.height=10}
plot_count_hex <- function(regions, y_col) {
  label <- sprintf("rho = %.2f\nn = %s", spearman(regions$fire_freq, regions[[y_col]]),
                   scales::comma(sum(!is.na(regions[[y_col]]))))
  ggplot(regions, aes(fire_freq, .data[[y_col]])) +
    geom_hex(bins = 60) +
    annotate("text", x = -Inf, y = Inf, label = label, hjust = -0.1, vjust = 1.2, size = 2.8) +
    scale_y_continuous(transform = "log1p", breaks = c(0, 3, 10, 30, 100, 300, 1000)) +
    scale_fill_viridis_c(transform = "log10", name = "regions") +
    labs(x = "FIRE frequency (pooled)", y = paste(y_col, "(log1p scale)")) +
    theme_fiberseq("classic")
}

p_count_hex <- plot_count_hex(regions, "fp_count_all") /
  plot_count_hex(regions, "fp_count_fire") /
  plot_count_hex(regions, "fp_count_nonfire")
save_figure(p_count_hex, file.path(plot_dir, "fire_freq_vs_fp_counts_hex.pdf"),
            width = 5.5, height = 10)
p_count_hex
```

## The same, colored by region width

Each hex is colored by the median width of its regions.

```{r plot_width_hex, fig.width=5.5, fig.height=3.5}
plot_summary_hex <- function(regions, y_col, z_col, fun, fill_lab) {
  ggplot(regions, aes(fire_freq, .data[[y_col]], z = .data[[z_col]])) +
    stat_summary_hex(fun = fun, bins = 60) +
    scale_y_continuous(transform = "log1p", breaks = c(0, 3, 10, 30, 100, 300, 1000)) +
    scale_fill_viridis_c(name = fill_lab) +
    labs(x = "FIRE frequency (pooled)", y = paste(y_col, "(log1p scale)")) +
    theme_fiberseq("classic")
}

p_width_hex <- plot_summary_hex(regions, "fp_count_all", "width", "median", "median width (bp)")
save_figure(p_width_hex, file.path(plot_dir, "fire_freq_vs_fp_count_all_median_width_hex.pdf"),
            width = 5.5, height = 3.5)
p_width_hex
```

## The same, colored by score

Each hex is colored by the mean score of its regions. The y axis is the count used for the scores
(`fp_rank_col`).

```{r plot_score_hex, fig.width=14, fig.height=4}
scores <- c("open_fp_poor_score", "open_fp_rich_score", "closed_fp_rich_score")
p_score_hex <- wrap_plots(lapply(scores, function(s)
  plot_summary_hex(regions, fp_rank_col, s, "mean", "mean score") + labs(title = s)), nrow = 1)
save_figure(p_score_hex, file.path(plot_dir, "fire_freq_vs_fp_count_score_hex.pdf"),
            width = 14, height = 4)
p_score_hex
```

## FIRE rank vs footprint rank

```{r plot_rank_hex, fig.width=5, fig.height=4.5}
plot_rank_hex <- function(regions, fp_col) {
  ggplot(regions, aes(fire_rank, fp_rank)) +
    geom_hex(bins = 50) +
    scale_fill_viridis_c(name = "regions") +
    coord_equal() +
    labs(x = "fire_rank (1 = most open)", y = paste0("fp_rank of ", fp_col, " (1 = most footprints)")) +
    theme_fiberseq("classic")
}

p_rank_hex <- plot_rank_hex(regions, fp_rank_col)
save_figure(p_rank_hex, file.path(plot_dir, "fire_rank_vs_fp_rank_hex.pdf"),
            width = 5, height = 4.5)
p_rank_hex
```


# 11. Region plots: top regions per score

This section plots the top `top_n_region_plots` regions by each score with `plot_region_panels()`
from `plotting_functions.r`. Each window is the region +/- `region_pad` bp, and the yellow band
marks the region itself. Reads from the four timepoints are pooled.

- Only reads that span the whole region are shown, which is the read set counted in `n_reads`.
- The reads are split into two groups. FIRE-positive reads have an element from the FIRE per-read
  elements file covering >= 50% of the region (the `03b_feature_count.py` rule). Non-FIRE reads
  do not.
- The footprint panel shows every FiberHMM TF footprint (score >= 50, 10-80 bp), not only the
  footprints inside FIRE elements that `plot_region_panels()` draws by default. This matches
  `fp_count_all`, which counts footprints on every spanning read.
- The FIRE peaks row shows the per-timepoint peak calls in the window. The region is their union.

```{r region_plot_setup}
code_dir <- "/project/spott/cshan/fiber-seq/code/parsing_functions"
# packages the region helpers attached before they moved into parsing_functions
suppressPackageStartupMessages({
  library(data.table); library(dplyr); library(GenomicRanges); library(IRanges)
  library(ggplot2); library(patchwork)
})
source(file.path(code_dir, "parsing_footprints_functions.r"))
source(file.path(code_dir, "plotting_functions.r"), local = TRUE)
extract_script <- file.path(code_dir, "extract_region_result_macrophage.sh")
extract_root   <- file.path(input_dir, "region_plots")

top_regions <- bind_rows(lapply(
  c("open_fp_poor_score", "open_fp_rich_score", "closed_fp_rich_score"), function(s)
    regions %>%
      slice_max(.data[[s]], n = top_n_region_plots, with_ties = FALSE) %>%
      mutate(score = s, score_value = .data[[s]], rank = row_number()))) %>%
  mutate(outname = paste(chrom, start, end, sep = "_"),
         window_start = as.integer(start - region_pad),
         window_end   = as.integer(end + region_pad))

top_regions %>%
  select(score, rank, region_id, score_value, fire_freq, n_reads, n_fire, fp_count_all,
         fp_count_fire, fp_count_nonfire, width, tss_distance) %>%
  knitr::kable(digits = 3)
```

Each region and timepoint is extracted once into `region_plots/<chrom_start_end>/<timepoint>/parsed`
by `plot_region_example()`, in the same call that plots it. The extraction is skipped if
`fire.bed` already exists there.

The subtitle compares the plotted read counts with the table's `n_reads` and `n_fire`. They
can differ slightly, because the plot takes each read's span from `fire.bed` and drops reads with no
m6A call in the window.

```{r region_plots, fig.width=10, fig.height=14}
fire_positive_reads <- function(parsed_dirs, region_start, region_end) {
  elements <- bind_rows(lapply(file.path(parsed_dirs, "fire_elements.bed"), function(f)
    read_tsv(f, col_names = c("chrom", "start", "end", "RID"), col_select = 1:4,
             col_types = "ciic")))
  overlap <- pmin(elements$end, region_end) - pmax(elements$start, region_start)
  unique(elements$RID[overlap >= 0.5 * (region_end - region_start)])
}

plot_top_region <- function(row, extract_root, timepoints) {
  parsed_dirs <- file.path(extract_root, row$outname, timepoints, "parsed")

  # draw every TF footprint, not only those inside a FIRE element, and split the
  # spanning reads into FIRE-positive and non-FIRE
  group_reads <- function(res) {
    res$fps_infire <- res$fps
    spans <- res$rids_df
    spanning <- as.character(spans$RID[spans$start <= row$start & spans$end >= row$end])
    fire_pos <- fire_positive_reads(parsed_dirs, row$start, row$end)
    labels <- setNames(ifelse(spanning %in% fire_pos, "FIRE-positive", "non-FIRE"), spanning)
    add_read_groups(res, labels, group_col = "read_class",
                    group_levels = c("FIRE-positive", "non-FIRE"), verbose = FALSE)
  }

  # the subtitle counts the grouped (spanning) reads
  panel_options <- function(res) {
    read_class <- res$rids_df$read_class
    subtitle <- sprintf(paste0("%s = %.3f | fire_freq %.2f | %d bp | TSS distance %s bp\n",
                               "spanning reads plotted %d (table n_reads %d), FIRE-positive %d ",
                               "(table n_fire %d) | footprints all %d, FIRE %d, non-FIRE %d"),
                        row$score, row$score_value, row$fire_freq, row$width,
                        format(row$tss_distance, big.mark = ","),
                        length(read_class), row$n_reads, sum(read_class == "FIRE-positive"),
                        row$n_fire, row$fp_count_all, row$fp_count_fire, row$fp_count_nonfire)
    list(plot_name = row$region_id,
         subtitle = subtitle,
         highlight = c(row$start, row$end),
         y_labels = c(pileup = "Met. prop.", reads = "SMF reads",
                      fire_fiberHMM = "FiberHMM TF footprints (all)"),
         heights = c(pileup = 6, peaks = 2, fire_peaks = 1,
                     reads = 8, fire_fiberHMM = 8))
  }

  plot_region_example(
    region       = data.frame(chr = row$chrom, start = row$window_start, end = row$window_end),
    sample_names = timepoints,
    out_root     = extract_root,
    outname      = row$outname,
    script       = extract_script,
    load_options = list(fiberHMM_feature      = "tf",
                        min_fiberHMM_fp_score = 50,
                        fiberHMM_size_breaks  = c(10, 30, 60, 80),
                        verbose = FALSE),
    group = group_reads,
    panel_options = panel_options)$plot
}

for (i in seq_len(nrow(top_regions))) {
  row <- top_regions[i, ]
  p <- plot_top_region(row, extract_root, timepoints)
  save_figure(p$combined,
              file.path(plot_dir, "region_plots",
                        sprintf("%s_%d_%s.pdf", row$score, row$rank, row$outname)),
              width = 10, height = 14)
  print(p$combined)
}
```
