---
title: "Fourier transformation on m6a"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    mathjax: null
    pandoc_args: ["--mathml"]
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r load-functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/haplotype_phasing/LCL_phasing.r")
source("/project/spott/cshan/fiber-seq/code/haplotype_phasing/LCL_phased_m6a_input.r")
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
```

```{r report-output}
report_writer <- create_report_plot_writer(
  file.path(getwd(), "fourier_transformation_files", "report-figures")
)
save_report_plot <- report_writer$save
```


**extract reads from a specific genomic to converting it a read by position binary m6a matrix**

  1. extract_ft_region_reads(ft_extracted_file, region): pulls the reads for one window. Given a per-chromosome ft extract file and a GRanges of one region, it fetches the overlapping BED12 records, keeps the longest alignment per read, and expands each read into one row per m6A call. Output columns are chrom, pos, read_start, read_end, RID, score, strand, with 1-based positions
  
  2. extract_ft_read_info(reads): collapses to one row per read: RID, chr, start, end, strand -> which positions each read actually covers
  
  3. get_sparse_met_mat(reads, rids_df, window_start, window_end, base = "A"): builds the read by position matrix. Rows are reads, columns are genomic positions. 1 is an m6A call, a 0 is a covered position with no call, and NA means the read does not span that position
  
  4. filter_met_mat(met_mat): filter for fully spanning reads 
  
  5. assemble_region_met_data(sample_names, chr, start, end, ft_result_dir)
  

**region selection**

  1. select_lcl_top_asfire_regions(asfire_results, top_n = 10L, width_bp = 2000L)



```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(data.table)
library(Rsamtools)
library(cowplot)
```


# Inputs and outputs
```{r settings}
# FFT preprocessing: mean-center every read, then optionally apply a Hann window.
use_hann <- FALSE
preprocessing_label <- if (use_hann) "Mean-centered + Hann window" else "Mean-centered only"

# inputs 
ft_result_dir <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
sample_meta <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv"
sample_table <- read.csv(sample_meta, stringsAsFactors = FALSE)
phasing_root <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples"
existing_phase_dir <- "/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan/phase_summary"
asfire_file <- "/project/spott/kevinluo/Fiber_seq/results/QTL/fireQTL/AS_fire_freq_results/combined_results/combined_AS_fire_freq_merged_peaks_window2000_firefreq0.1_het_only.rds"

# outputs
output_dir <- "/project/spott/cshan/fiber-seq/LCL_project/fourier_transformation"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```

# Region selection 
```{r region-selection}

lcl_selection_file <- "/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan/summary tables/selection_signature.rds"
selected_regions <- if (file.exists(lcl_selection_file)) {
  readRDS(lcl_selection_file)$regions
} else {
  select_lcl_top_asfire_regions(readRDS(asfire_file), top_n = 10L, width_bp = 2000L)
}
if (file.exists(lcl_selection_file)) sample_table <- readRDS(lcl_selection_file)$sample_table
stopifnot(nrow(selected_regions) > 0L, !anyDuplicated(selected_regions$region_id),
          all(selected_regions$analysis_start == selected_regions$start + 1L),
          all(selected_regions$analysis_end == selected_regions$end),
          all(selected_regions$analysis_end - selected_regions$analysis_start + 1L == 2000L))

selected_regions
```

```{r convert-to-GRanges}

gr_input <- as.data.frame(selected_regions)
if ("seqnames" %in% names(gr_input)) {
  stopifnot(identical(as.character(gr_input$seqnames), as.character(gr_input$chr)))
  gr_input$seqnames <- NULL
}
gr <- makeGRangesFromDataFrame(
  gr_input,
  keep.extra.columns = TRUE,
  seqnames.field = "chr",
  start.field = "start",
  end.field = "end",
  strand.field = "strand",
  starts.in.df.are.0based = TRUE
)
names(gr) <- selected_regions$region_id
stopifnot(all(start(gr) == selected_regions$analysis_start),
          all(end(gr) == selected_regions$analysis_end), all(width(gr) == 2000L))

```


# Phased heterozygous input for every LCL region

All calculations in this notebook—FFT power spectra, aggregate summaries,
Fourier features, clustering and plots—use only fully spanning reads from
eligible LCL samples with HP1/HP2, a phased heterozygous focal-SNP genotype
(`0|1` or `1|0`), and a resolved reference/alternate allele. This is the same
read filter used by LCL Manhattan-Leiden and autocorrelation. It runs before
FFT for every selected region, including future regions; no rank is special.
Unphased, unresolved and nonheterozygous reads are excluded, not imputed.

The binary representation is unchanged: one column per genomic base, with
m6A calls = 1 and covered non-calls = 0, including non-A and all-zero columns.
The original Fourier extractor omits reads with no in-window m6A calls before
matrix construction. Matrices remain in memory; per-read inclusion audits and
retained phasing metadata are saved as tables.

```{r build-m6a-binary-matrix, message=FALSE, results="hide"}
region_met_mats <- list()
input_summary <- list()
input_tables <- file.path(output_dir, "tables")
dir.create(input_tables, recursive = TRUE, showWarnings = FALSE)
for (i in seq_len(nrow(selected_regions))) {
  region <- selected_regions[i, , drop = FALSE]
  region_id <- region$region_id
  input <- lcl_build_phased_m6a_input(region, sample_table, ft_result_dir,
    phasing_root, existing_phase_dir)
  stopifnot(all(input$read_meta$haplotype %in% c("HP1", "HP2")),
    all(input$read_meta$focal_genotype %in% c("0|1", "1|0")),
    all(input$read_meta$allele_status == "phased_focal_genotype"))
  region_met_mats[[region_id]] <- list(region_id = region_id, chr = region$chr,
    start = region$analysis_start, end = region$analysis_end,
    met_mat = input$met_mat, read_meta = input$read_meta,
    read_filter_audit = input$read_filter_audit, variants = input$variants,
    focal = input$focal, phasing_sources = input$phasing_sources)
  data.table::fwrite(input$read_filter_audit,
    file.path(input_tables, paste0(region_id, ".input_read_audit.tsv")), sep = "\t")
  data.table::fwrite(input$read_meta,
    file.path(input_tables, paste0(region_id, ".phased_read_metadata.tsv")), sep = "\t")
  input_summary[[region_id]] <- data.frame(region_id = region_id,
    n_full_span_candidates = nrow(input$read_filter_audit),
    n_phased_heterozygous_reads = nrow(input$read_meta),
    n_excluded = sum(!input$read_filter_audit$retained_for_analysis),
    n_bases = ncol(input$met_mat))
}
input_summary <- dplyr::bind_rows(input_summary)
data.table::fwrite(input_summary, file.path(input_tables, "phased_input_summary.tsv"), sep = "\t")
```

```{r phased-input-summary}
knitr::kable(input_summary)
```

# Fourier transformation

Each read is mean-centered before FFT
  - `use_hann` in the global settings to enable or disable the Hann window

For each complex coefficient X, amplitude = |X| and phase = Arg(X). The real
and imaginary components jointly encode both: Re(X) = amplitude * cos(phase)
and Im(X) = amplitude * sin(phase)

Hann window is a smooth weighting function that gradually reduces the signal toward zero at both ends

Test the all posssible cycles across the region
  - k = 0 -> direct current = mean -> removed
  - k = number of cycles, period = length / k 
  - each period gets a complex number

```{r Hann-preprocessing}
hann_window <- function(n) {
  stopifnot(length(n) == 1L, n >= 2L)
  0.5 * (1 - cos(2 * pi * (seq_len(n) - 1) / (n - 1)))
}
```

```{r FFT-set-up}

# ============================================================
# FFT for every read in one binary m6A matrix
#
# Input:
#   met_mat
#     rows    = reads
#     columns = consecutive genomic positions at 1-bp spacing
#     values  = 0/1
#
# Output:
#   raw_power       = read x period power matrix
#   relative_power  = read x period relative-power matrix
#   periods_bp      = period represented by each column
#   frequency       = frequency represented by each column
# ============================================================

get_fft_power_matrix <- function(met_mat, use_hann = TRUE) {

  # ----------------------------------------------------------
  # 1. Check that genomic positions are consecutive
  # ----------------------------------------------------------
  pos <- as.integer(colnames(met_mat))

  stopifnot(all(diff(pos) == 1))

  # Check that there are no missing values
  stopifnot(!anyNA(met_mat))

  # Number of genomic positions
  N <- ncol(met_mat)
  stopifnot(N >= 2L, is.logical(use_hann), length(use_hann) == 1L, !is.na(use_hann))

  # Optional preprocessing window; applied only after mean-centering below.
  if (use_hann) window <- hann_window(N)

  # ----------------------------------------------------------
  # 2. Define all positive Fourier frequencies
  #
  # k = 0 is excluded because it represents the mean of
  # the signal, not a repeating spatial pattern.
  #
  # For N = 2000:
  # k = 1, 2, ..., 1000
  # ----------------------------------------------------------
  k <- 1:floor(N / 2)

  # Frequency in cycles per bp
  # because genomic positions are spaced exactly 1 bp apart
  frequency <- k / N

  # Convert frequency to spatial period in bp
  #
  # period = 1 / frequency = N / k
  # ----------------------------------------------------------
  periods_bp <- 1 / frequency

  # Equivalent:
  # periods_bp <- N / k

  # ----------------------------------------------------------
  # 3. Prepare output matrices
  # ----------------------------------------------------------
  period_names <- paste0(
    "period_",
    round(periods_bp, 3),
    "bp"
  )

  raw_power <- matrix(
    NA_real_,
    nrow = nrow(met_mat),
    ncol = length(k),
    dimnames = list(
      rownames(met_mat),
      period_names
    )
  )

  relative_power <- matrix(
    NA_real_,
    nrow = nrow(met_mat),
    ncol = length(k),
    dimnames = list(
      rownames(met_mat),
      period_names
    )
  )

  # ----------------------------------------------------------
  # 4. Run FFT independently for every read
  # ----------------------------------------------------------
  for (i in seq_len(nrow(met_mat))) {

    # Extract one binary read
    signal <- as.numeric(met_mat[i, ])

    # --------------------------------------------------------
    # Remove mean signal
    # --------------------------------------------------------
    signal <- signal - mean(signal)

    # --------------------------------------------------------
    # Optional Hann preprocessing (no change to power definitions)
    # --------------------------------------------------------
    if (use_hann) {
      signal <- signal * window
    }

    # --------------------------------------------------------
    # Fast Fourier Transform
    # --------------------------------------------------------
    ft <- fft(signal)

    # --------------------------------------------------------
    # Keep positive-frequency Fourier coefficients
    #
    # ft[1] = k = 0
    # ft[2] = k = 1
    # ft[3] = k = 2
    #
    # therefore positive k values correspond to k + 1
    # --------------------------------------------------------
    ft_half <- ft[k + 1]

    # --------------------------------------------------------
    # Calculate periodogram power
    # --------------------------------------------------------
    power <- Mod(ft_half)^2 / N

    # --------------------------------------------------------
    # One-sided spectrum correction
    #
    # The FFT  has symmetric positive
    # and negative frequency components.
    #
    # Because we keep only the positive half, multiply power
    # by 2 to account for the discarded negative-frequency
    # copy.
    #
    # Exception:
    # For even N, the Nyquist frequency (k = N/2) has no
    # separate negative-frequency partner, so do NOT double it.
    # --------------------------------------------------------
    if (N %% 2 == 0) {

      double_idx <- k < N / 2

    } else {

      # For odd N, every retained positive frequency has a
      # negative-frequency partner
      double_idx <- rep(TRUE, length(k))
    }

    power[double_idx] <- 2 * power[double_idx]

    # --------------------------------------------------------
    # Save absolute/raw spectral power
    # --------------------------------------------------------
    raw_power[i, ] <- power

    # --------------------------------------------------------
    # Calculate relative power
    #
    # Divide every frequency by total power in this
    # read so the values sum to 1
    #
    # --------------------------------------------------------
    total_power <- sum(power)

    if (total_power > 0) {

      relative_power[i, ] <-
        power / total_power
    }
  }

  # ----------------------------------------------------------
  # 5. Return results
  # ----------------------------------------------------------
  list(
    raw_power = raw_power,
    relative_power = relative_power,
    periods_bp = periods_bp,
    frequency = frequency
  )
}

```

### run FFT on selected regions

```{r run-FFT}
# ============================================================
# Run FFT for every read in every region
# ============================================================

fft_results <- list()

for (region_id in names(region_met_mats)) {

  message("==============================")
  message("FFT region: ", region_id)
  message("==============================")

  x <- region_met_mats[[region_id]]

  combined_met_mat <- x$met_mat

  # ----------------------------------------------------------
  # Confirm that matrix contains every genomic position
  # ----------------------------------------------------------
  expected_pos <- seq.int(
    x$start,
    x$end
  )

  actual_pos <- as.integer(
    colnames(combined_met_mat)
  )

  stopifnot(
    identical(
      actual_pos,
      expected_pos
    )
  )

  message(
    "  Input: ",
    nrow(combined_met_mat),
    " reads x ",
    ncol(combined_met_mat),
    " genomic positions"
  )

  # ----------------------------------------------------------
  # FFT / periodogram
  # ----------------------------------------------------------
  fft_out <- get_fft_power_matrix(
    combined_met_mat, use_hann = use_hann
  )

  # ----------------------------------------------------------
  # Store result
  # ----------------------------------------------------------
  fft_results[[region_id]] <- list(
    region_id = region_id,
    chr = x$chr,
    start = x$start,
    end = x$end,
    use_hann = use_hann,

    frequency = fft_out$frequency,
    periods_bp = fft_out$periods_bp,

    raw_power = fft_out$raw_power,
    relative_power = fft_out$relative_power,

    read_meta = x$read_meta
  )

  message(
    "  Output: ",
    nrow(fft_out$raw_power),
    " reads x ",
    ncol(fft_out$raw_power),
    " Fourier frequencies"
  )
}

```

# FFT results


### define output directories

```{r FFT-outputs, echo=TRUE, eval=TRUE}

fft_plot_dir <- file.path(output_dir, "fft_plots")

# one directory per output type
periodogram_dir     <- file.path(fft_plot_dir, "periodograms_all_reads")
aggregate_dir       <- file.path(fft_plot_dir, "aggregate_spectra")
heatmap_dir         <- file.path(fft_plot_dir, "power_heatmaps")
dominant_period_dir <- file.path(fft_plot_dir, "dominant_periods")
example_dir         <- file.path(fft_plot_dir, "hann_example")
repeat_length_dir   <- file.path(fft_plot_dir, "chromatin_repeat_length")

for (d in c(periodogram_dir, aggregate_dir, heatmap_dir, dominant_period_dir, example_dir,
            repeat_length_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
```


## 1. Periodogram for every read, for every region
```{r periodogram, results="asis", echo=TRUE, eval=TRUE}
# ============================================================
# 1. Periodogram for ALL reads in each region
# ============================================================

for (region_id in names(fft_results)) {

  message("Plotting periodograms: ", region_id)

  x <- fft_results[[region_id]]

  power_mat <- x$raw_power

  # Lookup between matrix column and actual period
  period_lookup <- data.frame(
    period_name = colnames(power_mat),
    period_bp   = x$periods_bp
  )

  # Convert to long format
  periodogram_long <- as.data.frame(power_mat) %>%
    rownames_to_column("RID") %>%
    pivot_longer(
      cols = -RID,
      names_to = "period_name",
      values_to = "power"
    ) %>%
    left_join(period_lookup, by = "period_name"
    )

  # Plot every read
  p <- ggplot(
    periodogram_long,
    aes(x = period_bp, y = power,group = RID)
  ) +
    geom_line(alpha = 0.08, linewidth = 0.2) +
    scale_x_continuous(breaks = seq(0, 2000, by = 250)) +
    theme_classic() +
    labs(title = region_id,
      subtitle = paste(nrow(power_mat), "individual reads;", preprocessing_label),
      x = "Spatial period (bp)",
      y = "Periodogram power")

  save_report_plot(
    filename = file.path(
      periodogram_dir,
      paste0(region_id, "_all_read_periodograms.pdf")
    ),
    plot = p,
    width = 8,
    height = 5
  )

  rm(periodogram_long)
  invisible(gc())
}
```

## 2. aggregate spectrum across all reads

```{r aggregate-periodgram, results="asis", echo=TRUE, eval=TRUE}
# ============================================================
# 2. Aggregate spectrum per region
# ============================================================

aggregate_spectra <- list()

for (region_id in names(fft_results)) {

  message("Aggregate spectrum: ", region_id)

  x <- fft_results[[region_id]]

  power_mat <- x$relative_power

  aggregate_df <- data.frame(
    region_id = region_id,

    period_bp = x$periods_bp,

    median_power = apply(power_mat,2,median,na.rm = TRUE),
    q25 = apply(power_mat,2,quantile, probs = 0.25, na.rm = TRUE),
    q75 = apply(power_mat,2,quantile, probs = 0.75, na.rm = TRUE))

  aggregate_spectra[[region_id]] <- aggregate_df

  p <- ggplot(
    aggregate_df,
    aes(x = period_bp, y = median_power)) +
    geom_ribbon(
      aes(ymin = q25, ymax = q75),
      alpha = 0.2) +
    geom_line(linewidth = 0.6) +
    scale_x_continuous(breaks = seq(0, 2000, by = 250)) +
    theme_classic() +
    labs(
      title = region_id,
      subtitle = paste(preprocessing_label, "| Median across reads; ribbon = IQR"),
      x = "Spatial period (bp)",
      y = "Median relative power")

  save_report_plot(
    file.path(aggregate_dir, paste0(region_id, "_aggregate_spectrum.pdf")),
    p, width = 8, height = 5)
}

aggregate_spectra <- bind_rows(
  aggregate_spectra
)
```

## 3. Heatmap: all reads x all periods, one plot per region

```{r power-heatmap, results="asis", echo=TRUE, eval=TRUE}
# ============================================================
# 3. Heatmap of relative power: rows = reads, columns = periods
#
# Rows are ordered by each read's dominant period
# ============================================================

heatmap_period_breaks <- seq(0, 2000, by = 250)
heatmap_n_bins <- 400
combined_heatmaps <- list()

for (region_id in names(fft_results)) {

  message("Power heatmap: ", region_id)

  x <- fft_results[[region_id]]
  power_mat <- x$relative_power
  if (nrow(power_mat) == 0) next

  # order reads by dominant period (longest period at the top)
  # Undefined spectra remain in the heatmap as grey rows, at the bottom.
  dominant_k <- apply(power_mat, 1, function(p) {
    if (all(is.na(p))) NA_integer_ else which.max(p)
  })
  row_order <- order(dominant_k, na.last = TRUE)

  # uniform grid of periods (bp, linear); each grid point takes the nearest FFT column
  grid <- seq(min(x$periods_bp), max(x$periods_bp), length.out = heatmap_n_bins)
  nearest_col <- vapply(grid, function(g) which.min(abs(x$periods_bp - g)), integer(1))
  heat_mat <- power_mat[row_order, nearest_col, drop = FALSE]

  heat_long <- data.frame(
    read_rank  = rep(rev(seq_len(nrow(heat_mat))), times = ncol(heat_mat)),
    period_bp = rep(grid, each = nrow(heat_mat)),
    power      = as.vector(heat_mat)
  )
  combined_heatmaps[[region_id]] <- transform(heat_long, region_id = region_id)

  cap <- quantile(heat_long$power, 0.99, na.rm = TRUE)
  if (!is.finite(cap) || cap <= 0) cap <- 1
  heat_long$power_capped <- pmin(heat_long$power, cap)

  breaks <- heatmap_period_breaks[
    heatmap_period_breaks >= min(x$periods_bp) & heatmap_period_breaks <= max(x$periods_bp)]

  p <- ggplot(heat_long, aes(x = period_bp, y = read_rank, fill = power_capped)) +
    geom_raster() +
    scale_fill_viridis_c(name = "Relative power", limits = c(0, cap), na.value = "grey85") +
    scale_x_continuous(breaks = breaks, expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme_classic() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
    labs(
      title = region_id,
      subtitle = paste(nrow(power_mat), "reads;", preprocessing_label,
                       "\nRows ordered by dominant period; color capped at 99th percentile"),
      x = "Spatial period (bp)",
      y = "Reads"
    )

  save_report_plot(
    file.path(heatmap_dir, paste0(region_id, "_power_heatmap.pdf")),
    p, width = 10, height = 7
  )

  rm(heat_long, heat_mat)
  invisible(gc())
}
```

## 4. Combined reads x periods heatmap across all regions



```{r combined-power-heatmap, results="asis", echo=TRUE, eval=TRUE}
combined_heatmap_data <- bind_rows(combined_heatmaps)
combined_heatmap_data$region_id <- factor(
  combined_heatmap_data$region_id, levels = names(fft_results)
)
combined_cap <- quantile(combined_heatmap_data$power, 0.99, na.rm = TRUE)
if (!is.finite(combined_cap) || combined_cap <= 0) combined_cap <- 1
combined_heatmap_data$power_capped <- pmin(combined_heatmap_data$power, combined_cap)

p_combined_heatmap <- ggplot(
  combined_heatmap_data, aes(x = period_bp, y = read_rank, fill = power_capped)
) +
  geom_raster() +
  facet_grid(region_id ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(name = "Relative power", limits = c(0, combined_cap),
                       na.value = "grey85") +
  scale_x_continuous(breaks = heatmap_period_breaks, expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        strip.text.y = element_text(angle = 0, size = 8)) +
  labs(title = "Per-read Fourier spectra across all regions",
       subtitle = paste(preprocessing_label,
                        "\nRows ordered within region by dominant period; color capped at 99th percentile"),
       x = "Spatial period (bp)", y = "Reads grouped by genomic region")
save_report_plot(file.path(heatmap_dir, "all_regions_power_heatmap.pdf"),
       p_combined_heatmap, width = 12, height = 16)
rm(combined_heatmaps, combined_heatmap_data)
invisible(gc())
```

## 5. Dominant periods and repeat-length boxplots across regions

For each read, the dominant period is the Fourier bin with the greatest
per-read spectral power without aggregation

```{r dominant-period, results="asis", echo=TRUE, eval=TRUE}
get_dominant_periods <- function(fft_results) {
  bind_rows(lapply(names(fft_results), function(region_id) {
    x <- fft_results[[region_id]]
    power_mat <- x$raw_power
    stopifnot(ncol(power_mat) > 0L, ncol(power_mat) == length(x$periods_bp))
    idx <- apply(power_mat, 1, function(p) {
      if (all(is.finite(p)) && max(p) > 0) which.max(p) else NA_integer_
    })
    meta_idx <- match(rownames(power_mat), x$read_meta$row_id)
    stopifnot(!anyNA(meta_idx), !anyDuplicated(x$read_meta$row_id))
    meta <- x$read_meta[meta_idx, , drop = FALSE]
    matrix_idx <- cbind(seq_len(nrow(power_mat)), idx)
    data.frame(
      region_id = region_id,
      row_id = rownames(power_mat),
      RID = meta$RID,
      sample_name = meta$sample_name,
      dominant_period_bp = x$periods_bp[idx],
      dominant_spectral_power = power_mat[matrix_idx],
      dominant_relative_power = x$relative_power[matrix_idx],
      dominant_power = x$relative_power[matrix_idx],
      analysis = "complete_spectrum",
      use_hann = x$use_hann,
      stringsAsFactors = FALSE
    )
  })) %>%
    mutate(region_id = factor(region_id, levels = names(fft_results)))
}

dominant_periods <- get_dominant_periods(fft_results)


summarize_dominant_periods <- function(df) {
  df %>% group_by(region_id) %>% summarise(
    n_reads = n(),
    n_defined = sum(!is.na(dominant_period_bp)),
    median_period_bp = median(dominant_period_bp, na.rm = TRUE),
    q25_period_bp = quantile(dominant_period_bp, 0.25, na.rm = TRUE),
    q75_period_bp = quantile(dominant_period_bp, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
}
dominant_summary <- summarize_dominant_periods(dominant_periods)
knitr::kable(dominant_summary)

p_hist <- ggplot(dominant_periods, aes(x = dominant_period_bp)) +
  geom_histogram(bins = 60, fill = "grey30", na.rm = TRUE) +
  scale_x_continuous(breaks = seq(0, 2000, by = 250)) +
  facet_wrap(~ region_id, ncol = 2, scales = "free_y", drop = FALSE) +
  theme_classic() +
  labs(title = "Dominant spatial period per read",
       subtitle = paste(preprocessing_label, "| Complete positive-frequency spectrum"),
       x = "Dominant period (bp)", y = "Number of reads")
save_report_plot(file.path(dominant_period_dir, "dominant_period_histograms_by_region.pdf"),
       p_hist, width = 10, height = 12)

p_points <- ggplot(dominant_periods, aes(x = region_id, y = dominant_period_bp)) +
  geom_point(position = position_jitter(width = 0.25, height = 0, seed = 1),
             size = 0.4, alpha = 0.3, na.rm = TRUE) +
  geom_boxplot(outlier.shape = NA, fill = NA, colour = "firebrick", width = 0.5,
               na.rm = TRUE) +
  scale_y_continuous(breaks = seq(0, 2000, by = 250)) +
  scale_x_discrete(drop = FALSE) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Dominant spatial period per read, all regions",
       subtitle = paste(preprocessing_label, "| Complete positive-frequency spectrum"),
       x = NULL, y = "Dominant period (bp)")
save_report_plot(file.path(dominant_period_dir, "dominant_period_per_read_all_regions.pdf"),
       p_points, width = 10, height = 6)
```

```{r repeat-length-boxplots, results="asis", echo=TRUE, eval=TRUE}
plot_repeat_length <- function(df) {
  ggplot(df, aes(x = region_id, y = dominant_period_bp)) +
    geom_boxplot(fill = "#80B1D3", width = 0.6, outlier.size = 0.6,
                 outlier.alpha = 0.4, na.rm = TRUE) +
    scale_x_discrete(drop = FALSE) +
    scale_y_continuous(breaks = seq(0, 2000, 250)) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = "Dominant Fourier period by region: complete spectrum",
      subtitle = paste(preprocessing_label,
        "\nPer-read maximum across all positive frequencies; long periods need not be nucleosome repeats"),
      x = "Genomic region",
      y = "Dominant period (bp)"
    )
}
p_box_full <- plot_repeat_length(dominant_periods)
save_report_plot(file.path(dominant_period_dir, "dominant_period_boxplot_complete_spectrum.pdf"),
       p_box_full, width = 12, height = 6)
```

## 6. One example region with and without Hann

```{r hann-example, results="asis", echo=TRUE, eval=TRUE}
example_region_id <- names(region_met_mats)[1]
example_mat <- region_met_mats[[example_region_id]]$met_mat
example_spectra <- bind_rows(lapply(c(FALSE, TRUE), function(hann_enabled) {
  res <- get_fft_power_matrix(example_mat, use_hann = hann_enabled)
  data.frame(
    region_id = example_region_id,
    preprocessing = if (hann_enabled) "With Hann" else "Without Hann",
    use_hann = hann_enabled,
    period_bp = res$periods_bp,
    median_power = apply(res$relative_power, 2, median, na.rm = TRUE),
    q25 = apply(res$relative_power, 2, quantile, probs = 0.25, na.rm = TRUE),
    q75 = apply(res$relative_power, 2, quantile, probs = 0.75, na.rm = TRUE)
  )
}))
example_spectra$preprocessing <- factor(example_spectra$preprocessing,
                                      levels = c("Without Hann", "With Hann"))

plot_hann_example <- function() {
  ggplot(example_spectra,
         aes(x = period_bp, y = median_power, colour = preprocessing, fill = preprocessing)) +
    geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
    geom_line(linewidth = 0.6) +
    facet_wrap(~ preprocessing, nrow = 1) +
    scale_colour_manual(values = c("Without Hann" = "#555555", "With Hann" = "#0072B2")) +
    scale_fill_manual(values = c("Without Hann" = "#555555", "With Hann" = "#0072B2")) +
    scale_x_continuous(breaks = seq(0, ncol(example_mat), 250)) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(title = "Complete spectrum",
         subtitle = paste(nrow(example_mat), "reads; mean-centered; median and IQR"),
         x = "Spatial period (bp)", y = "Relative power")
}
p_example_full <- plot_hann_example()
p_hann_example <- cowplot::plot_grid(
  cowplot::ggdraw() + cowplot::draw_label(example_region_id, fontface = "bold"),
  p_example_full, ncol = 1, rel_heights = c(0.12, 1)
)
save_report_plot(file.path(example_dir, paste0(example_region_id, "_with_without_hann.pdf")),
                 p_hann_example, width = 12, height = 5)
ggplot2::ggsave(file.path(example_dir, paste0(example_region_id, "_with_without_hann.png")),
               p_hann_example, width = 12, height = 5, dpi = 160, bg = "white")
```

## 7. Heatmap of median spectral density by repeat length, with peak repeat length

the per-molecule spectral and the fill is the median spectral density across the region's fibers 

```{r spectral-density-heatmap, results="asis", echo=TRUE, eval=TRUE}
spectral_density_type  <- "raw_power"      # one-sided periodogram power; or "relative_power"
spectral_density_label <- if (spectral_density_type == "raw_power") "Spectral density" else "Relative spectral density"

# Per-fiber spectral density at every Fourier bin, all regions.
spectral_density_long <- bind_rows(lapply(names(fft_results), function(region_id) {
  x <- fft_results[[region_id]]
  sd_mat <- x[[spectral_density_type]]
  keep_rows <- rowSums(is.na(sd_mat)) == 0
  sd_mat <- sd_mat[keep_rows, , drop = FALSE]
  data.frame(
    region_id        = region_id,
    row_id           = rep(rownames(sd_mat), times = ncol(sd_mat)),
    repeat_length_bp = rep(x$periods_bp, each = nrow(sd_mat)),
    spectral_density = as.vector(sd_mat),
    stringsAsFactors = FALSE
  )
}))
spectral_density_long$region_id <- factor(spectral_density_long$region_id, levels = names(fft_results))

median_spectral_density <- spectral_density_long %>%
  group_by(region_id, repeat_length_bp) %>%
  summarise(median_density = median(spectral_density), n_fibers = n(), .groups = "drop") %>%
  arrange(region_id, repeat_length_bp)

# tile edges at the midpoints between neighbouring Fourier bins
rl <- sort(unique(median_spectral_density$repeat_length_bp))
rl_edges <- c(rl[1] - (rl[2] - rl[1]) / 2,
              (rl[-1] + rl[-length(rl)]) / 2,
              rl[length(rl)] + (rl[length(rl)] - rl[length(rl) - 1]) / 2)
rl_idx <- match(median_spectral_density$repeat_length_bp, rl)
median_spectral_density$xmin <- rl_edges[rl_idx]
median_spectral_density$xmax <- rl_edges[rl_idx + 1]

# peak repeat length per region = bin with the largest median (ties: longest period)
repeat_length_peaks <- median_spectral_density %>%
  group_by(region_id) %>%
  filter(median_density == max(median_density)) %>%
  slice_max(repeat_length_bp, n = 1, with_ties = FALSE) %>%
  ungroup()

peak_repeat_lengths <- repeat_length_peaks %>%
  select(region_id, repeat_length_bp, median_density) %>%
  arrange(region_id)
knitr::kable(peak_repeat_lengths)


# regions top-to-bottom in selection order
region_levels <- rev(names(fft_results))
median_spectral_density$region_y <- match(as.character(median_spectral_density$region_id), region_levels)
repeat_length_peaks$region_y     <- match(as.character(repeat_length_peaks$region_id), region_levels)

p_sd_heatmap <- ggplot(median_spectral_density) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = region_y - 0.5, ymax = region_y + 0.5,
                fill = median_density)) +
  geom_point(data = repeat_length_peaks,
             aes(x = repeat_length_bp, y = region_y),
             shape = 21, size = 2.4, colour = "white", fill = "black", stroke = 0.7) +
  scale_fill_viridis_c(name = paste("Median", tolower(spectral_density_label))) +
  scale_x_continuous(breaks = seq(0, max(median_spectral_density$repeat_length_bp), by = 250),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = seq_along(region_levels), labels = region_levels, expand = c(0, 0)) +
  theme_classic() +
  labs(title = "Median spectral density by chromatin repeat length",
       subtitle = preprocessing_label,
       x = "1 / m6A spectral frequency (bp)", y = NULL)

save_report_plot(file.path(repeat_length_dir, "median_spectral_density_heatmap.pdf"),
       p_sd_heatmap, width = 11, height = 6)
```

# Fourier amplitude and phase features for  clustering

Fourier bins are defined by k and period -> Re_k1_period_2000bp

1. Calculate mean centered value from the methylation binary vector -> deviation from he read's average m6a
  - mean centered value = original value - mean
  - position with an m6A call = 1 - mean
  - position without m6a call = 0 - mean
  
  
2. Calculate amplitude and phase using complex coefficients from FFT
  - amplitude = how strongly the repeat length is, magnitude of complex number
  - phase = where the repeating pattern is shifted, angle of complex number
  - fourier index k = how many cycles fit across the window = period = length / k
      - k = 1, one cycle
      - k = 11, 11 cyles of 182 bp


3. Normalization
  - each read's features are divided the read overall spectral strength
    1) squares all real and imaginary values
    2) add the values together and takes the square root
    3) divides every real and imaginary value by the same number
    
*Note*: only normalization per read is performed, but no z score normalization is performed across all reads for one feature at a time
    
    
Re and Im are real and imaginary features of each Fourier coefficient for every Fourier period
X_200 (200 bp period) = 3 + 4i
  - Re = 3
  - Im = 4
  
real and imaginary features are converted in to a vector to represent each read

Phase is defined relative to the starting position of the read

```{r Fourier-complex-feature-setup}
get_fft_complex_features <- function(met_mat, use_hann) {
  pos <- as.integer(colnames(met_mat))
  N <- ncol(met_mat)
  stopifnot(N >= 2L, length(pos) == N, !anyNA(pos), all(diff(pos) == 1L),
            !anyNA(met_mat), !is.null(rownames(met_mat)),
            !anyDuplicated(rownames(met_mat)),
            is.logical(use_hann), length(use_hann) == 1L, !is.na(use_hann))

  # define fourier indices
  k <- seq_len(floor(N / 2))
  # convert index to frequency
  frequency <- k / N
  # convert frequency to bp
  periods_bp <- 1 / frequency

  bin_names <- paste0("k", k, "_period_", round(periods_bp, 6), "bp")
  
  # matrix of FFT coefficients
  coefficients <- matrix(NA_complex_, nrow(met_mat), length(k),
                         dimnames = list(rownames(met_mat), bin_names))

  if (use_hann) window <- hann_window(N)
  
  # loop over every read in the methylation matrix
  for (i in seq_len(nrow(met_mat))) {
    # convert per read methylation signal into a vector
    signal <- as.numeric(met_mat[i, ])
    # remove mean methylation sginal
    signal <- signal - mean(signal)
    
    if (use_hann) {
      # positions near the center are preserved, near the edges are downweighted
      signal <- signal * window
    }
    
    # FFT calculation
    coefficients[i, ] <- fft(signal)[k + 1L]
  }

  # amplitude
  amplitude <- Mod(coefficients)
  # phase
  phase_radians <- Arg(coefficients)
  
  # if amplitude is zero, the read contains none of Fourier component, replace phase with NA
  phase_radians[amplitude == 0] <- NA_real_

  # annotation table for clustering features
    # feature, component, k, period
  
  feature_annotation <- data.frame(
    feature_name = as.vector(rbind(paste0("Re_", bin_names), paste0("Im_", bin_names))),
    # for each period, there are two features
      # 1. Re
      # 2. Im
    component = rep(c("Re", "Im"), times = length(k)),
    k = rep(k, each = 2L),
    frequency = rep(frequency, each = 2L),
    period_bp = rep(periods_bp, each = 2L),
    stringsAsFactors = FALSE
  )
  
  
  # create Re/Im clustering matrix
  
  # two features per period
  re_im_features <- matrix(NA_real_, nrow(met_mat), 2L * length(k),
    dimnames = list(rownames(met_mat), feature_annotation$feature_name))
  
  # fill in Re and Im features
  re_im_features[, 2L * seq_along(k) - 1L] <- Re(coefficients)
  re_im_features[, 2L * seq_along(k)] <- Im(coefficients)
  
  # ----------------
  # Normalization
  # ----------------
  
  # squares every amplitude and add them across all frequencies
  coefficient_energy <- rowSums(amplitude^2)
  # square root of the sum -> one number per read for overall Fourier magnitude
  normalization_denominator <- sqrt(coefficient_energy)
  
  valid_energy <- is.finite(normalization_denominator) & normalization_denominator > 0
  normalized_re_im_features <- re_im_features
  normalized_re_im_features[,] <- NA_real_
  
  # normalization for each row
  # divide every Re and Im value in one read by the squared root sum
  normalized_re_im_features[valid_energy, ] <- sweep(
    re_im_features[valid_energy, , drop = FALSE], 1L,
    normalization_denominator[valid_energy], "/"
  )

  # ----------------
  # all outputs
  # ----------------
  # 
  # 1. FFT complex number
  # 2. matrix shape = read x periods
  # 3. Raw re and lm feaures
  # 4. feature annotation
  # 5. normalized re/lm matrix
  
  
  list(
    complex_coefficients = coefficients,
    amplitude = amplitude,
    phase_radians = phase_radians,
    re_im_features = re_im_features,
    feature_annotation = feature_annotation,
    normalized_re_im_features = normalized_re_im_features,
    normalization_denominator = normalization_denominator,
    normalization = "L2 norm of all retained positive-frequency complex coefficients",
    standardized = FALSE,
    frequency = frequency,
    periods_bp = periods_bp,
    use_hann = use_hann,
    phase_origin_bp = pos[1],
    phase_coordinate_order = "ascending reference genomic position"
  )
}
```

# Get Re/lm features from FFT

```{r prepare-Fourier-complex-features}
fourier_complex_results <- list()

for (region_id in names(region_met_mats)) {
  x <- region_met_mats[[region_id]]
  message("Preparing Fourier Re/Im features: ", region_id)
  feature_result <- get_fft_complex_features(
    x$met_mat, use_hann = use_hann
  )
  meta_idx <- match(rownames(x$met_mat), x$read_meta$row_id)
  stopifnot(!anyNA(meta_idx), !anyDuplicated(x$read_meta$row_id))
  feature_result$read_meta <- x$read_meta[meta_idx, , drop = FALSE]
  feature_result$region_id <- region_id
  feature_result$chr <- x$chr
  feature_result$start <- x$start
  feature_result$end <- x$end
  fourier_complex_results[[region_id]] <- feature_result
}

fourier_feature_summary <- bind_rows(lapply(fourier_complex_results, function(x) {
  data.frame(
    region_id = x$region_id,
    n_reads = nrow(x$complex_coefficients),
    n_positive_frequencies = ncol(x$complex_coefficients),
    n_fourier_features = ncol(x$re_im_features),
    n_zero_energy_reads = sum(x$normalization_denominator == 0),
    use_hann = x$use_hann
  )
}))
fourier_feature_summary
```

Inspect the first region, each element is a list of per read matrices

*Fourier bins*: bin k means k cycles across the window, for k from 1 to 1000
  - bin 1 is one cycle per 2,000-bp window. Bin 2 is 1,000 bp, bin 10 is 200 bp, bin 11 is approximately 181.8 bp, and bin 1000 is 2 bp
  - bins above N/2 are mirror images of the ones below, so
  only the first half carries information. Bin 0 would be the signal's mean, which is removed before the transformation
  - all the bins are used for FFT (1 to 1000)
  
```{r example-FFT-features}
x_fft <- fourier_complex_results[[1]]
cat("Region:", x_fft$region_id, "\n")
str(x_fft, max.level = 1, give.attr = FALSE)

n_show_reads <- 4
n_show_bins  <- 5
show_bins <- seq_len(n_show_bins)

cat("\nFourier bins shown (k, period in bp):\n")
print(data.frame(k = show_bins, period_bp = round(x_fft$periods_bp[show_bins], 2)))

cat("\ncomplex_coefficients (reads x bins):\n")
print(x_fft$complex_coefficients[seq_len(n_show_reads), show_bins])

cat("\namplitude = Mod(coefficient):\n")
print(round(x_fft$amplitude[seq_len(n_show_reads), show_bins], 4))

cat("\nphase_radians = Arg(coefficient):\n")
print(round(x_fft$phase_radians[seq_len(n_show_reads), show_bins], 4))

cat("\nre_im_features (Re and Im interleaved; first", 2 * n_show_bins, "columns = first", n_show_bins, "bins):\n")
print(round(x_fft$re_im_features[seq_len(n_show_reads), seq_len(2 * n_show_bins)], 4))

cat("\nnormalized_re_im_features (same columns, divided by each read's L2 norm):\n")
print(round(x_fft$normalized_re_im_features[seq_len(n_show_reads), seq_len(2 * n_show_bins)], 6))

cat("\nnormalization_denominator (per read):\n")
print(round(x_fft$normalization_denominator[seq_len(n_show_reads)], 4))

cat("\nfeature_annotation (one row per Re/Im feature):\n")
print(head(x_fft$feature_annotation, 2 * n_show_bins))

cat("\nread_meta (rows align with the matrices):\n")
print(head(x_fft$read_meta, n_show_reads))
```



# Manhattan / Leiden clustering on Fourier features

```{r Fourier-Leiden-settings}
leiden_code_dir <- "/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan"
fourier_lcl_helpers <- new.env(parent = environment())
sys.source(file.path(leiden_code_dir, "leiden_manhattan_functions.r"), envir = fourier_lcl_helpers)
sys.source("/project/spott/cshan/fiber-seq/code/haplotype_phasing/LCL_phasing.r", envir = fourier_lcl_helpers)
sys.source(file.path(leiden_code_dir, "leiden_manhattan_plots.r"), envir = fourier_lcl_helpers)
local({
  notebook <- file.path(leiden_code_dir, "leiden_LCL.Rmd")
  lines <- readLines(notebook, warn = FALSE)
  start <- which(trimws(lines) == "```{r lcl-output-loader}")
  stopifnot(length(start) == 1L)
  end <- which(seq_along(lines) > start & trimws(lines) == "```")[1L]
  stopifnot(!is.na(end), end > start + 1L)
  eval(parse(text = lines[seq.int(start + 1L, end - 1L)]), envir = fourier_lcl_helpers)
  fourier_lcl_helpers$source_lcl_output_config(notebook, envir = fourier_lcl_helpers)
})
fourier_lcl_helpers$LCL_OUTPUT_ROOT <- normalizePath(output_dir, mustWork = TRUE)
fourier_lcl_helpers$LCL_TOP10_ROOT <- fourier_lcl_helpers$LCL_OUTPUT_ROOT

k_neighbors <- 10L
resolution <- 1
clustering_seed <- 1L
umap_neighbors <- 15L

fourier_clustering_dir <- file.path(output_dir, "manhattan_leiden_clustering")
fourier_sample_table <- sample_table
stopifnot(!anyDuplicated(fourier_sample_table$sample_name),
          k_neighbors >= 1L, k_neighbors == as.integer(k_neighbors),
          is.finite(resolution), resolution > 0, umap_neighbors >= 2L)
for (package in c("igraph", "uwot", "Rsamtools", "GenomicRanges", "cowplot")) {
  if (!requireNamespace(package, quietly = TRUE)) stop("Required package unavailable: ", package)
}
```

```{r Fourier-Leiden-phasing-helpers}

fourier_match_lcl_region <- function(x, selected_regions) {

  regions <- as.data.frame(selected_regions)
  required <- c("region_id", "chr", "start", "analysis_start", "analysis_end",
                "focal_pos", "focal_snp", "ref", "alt", "contributing_samples")
  stopifnot(all(required %in% names(regions)))
  hit <- which(regions$chr == x$chr & regions$analysis_end == x$end &
                 regions$analysis_start == x$start & regions$region_id == x$region_id)
  if (length(hit) != 1L) stop("Cannot uniquely match Fourier region to the LCL focal SNP: ", x$region_id)
  regions[hit, , drop = FALSE]
}
```

```{r run-Fourier-Manhattan-Leiden}
# Inputs were already filtered before any FFT or feature construction.
fourier_leiden_results <- list()
fourier_clustering_audits <- list()

for (region_id in names(fourier_complex_results)) {
  x <- fourier_complex_results[[region_id]]
  message("Fourier Manhattan / Leiden: ", region_id)
  stopifnot(identical(x$use_hann, use_hann), !isTRUE(x$standardized))
  region <- fourier_match_lcl_region(x, selected_regions)
  positions <- as.integer(colnames(region_met_mats[[region_id]]$met_mat))
  stopifnot(identical(positions, seq.int(region$analysis_start, region$analysis_end)),
            length(positions) == 2000L, x$phase_origin_bp == region$analysis_start)
  selected_samples <- lcl_region_samples(region, fourier_sample_table)
  meta <- x$read_meta
  meta$original_RID <- meta$RID
  meta$RID <- meta$row_id                 
  stopifnot(!anyDuplicated(meta$RID), identical(meta$RID, rownames(x$normalized_re_im_features)))
  meta$region_id <- region_id
  eligible <- meta$sample_name %in% selected_samples$sample_name
  stopifnot(all(eligible), all(meta$haplotype %in% c("HP1", "HP2")),
    all(meta$focal_genotype %in% c("0|1", "1|0")),
    all(meta$allele_status == "phased_focal_genotype"))
  original_input <- region_met_mats[[region_id]]
  phased <- list(rids_df = meta, read_filter_audit = original_input$read_filter_audit,
    variants = original_input$variants, focal = original_input$focal)

  assignments <- phased$rids_df
  finite_features <- rowSums(!is.finite(x$normalized_re_im_features[assignments$RID, , drop = FALSE])) == 0L
  assignments <- assignments[finite_features, , drop = FALSE]
  if (nrow(assignments) < 3L || length(unique(assignments$allele_label)) != 2L)
    stop("Insufficient finite Fourier features or missing focal allele: ", region_id)
  features <- x$normalized_re_im_features[assignments$RID, , drop = FALSE]
  stopifnot(ncol(features) == 2L * length(x$frequency))

  knn <- manhattan_knn_graph(features, k_neighbors = k_neighbors, sigma = NULL)
  partition <- leiden_partition(knn$graph, resolution = resolution, seed = clustering_seed)
  membership <- partition$membership[rownames(features)]
  stopifnot(!anyNA(membership))
  labels_by_size <- names(sort(table(membership), decreasing = TRUE))
  assignments$cluster <- factor(paste0("cluster", match(as.character(membership), labels_by_size)),
                                levels = paste0("cluster", seq_along(labels_by_size)))
  assignments$sample_label <- sub("_.*$", "", assignments$sample_name)
  stopifnot(identical(assignments$RID, rownames(features)))

  audit <- meta
  audit$exclusion_reason <- ifelse(eligible, "Unresolved focal genotype / haplotype", "Outside contributing LCLs")
  audit$exclusion_reason[audit$RID %in% phased$rids_df$RID[!finite_features]] <- "Nonfinite Fourier features"
  audit$exclusion_reason[audit$RID %in% assignments$RID] <- "Clustered"
  audit$retained_for_clustering <- audit$exclusion_reason == "Clustered"
  fourier_clustering_audits[[region_id]] <- audit
  distances <- knn$dist
  diag(distances) <- 0                  
  result <- list(region_id = region_id, region = region, assignments = assignments,
    feat_mat = features, feature_annotation = x$feature_annotation,
    distance_matrix = distances, graph = knn$graph,
    knn_edges = igraph::as_data_frame(knn$graph, what = "edges"),
    n_clusters = nlevels(assignments$cluster), read_filter_audit = audit,
    phasing_audit = phased$read_filter_audit, variants = phased$variants, focal = phased$focal,
    phasing_sources = original_input$phasing_sources,
    params = list(k_neighbors = k_neighbors, k_eff = knn$k_eff, resolution = resolution,
      sigma = knn$sigma, seed = clustering_seed, umap_neighbors = umap_neighbors,
      use_hann = x$use_hann, normalization = x$normalization, standardized = FALSE,
      frequency_range = range(x$frequency), phase_origin_bp = x$phase_origin_bp,
      fft_start = x$start, fft_end = x$end, quality = partition$quality))
  result <- fiberseq_display_result(result)
  result$embedding <- fiberseq_umap(result, seed = clustering_seed, n_neighbors = umap_neighbors)
  fourier_leiden_results[[region_id]] <- result
  message("  ", nrow(assignments), "/", nrow(meta), " reads retained; ", result$n_clusters, " clusters")
}
```

```{r Fourier-Leiden-plot-helpers}
fourier_save_plot <- function(plot, region_id, filename, width = 12, height = 7,
                              draw = NULL, label = NULL) {
  directory <- file.path(fourier_clustering_dir, region_id)
  dir.create(directory, recursive = TRUE, showWarnings = FALSE)
  if (is.null(label)) label <- paste(region_id, filename)
  save_report_plot(file.path(directory, paste0(filename, ".pdf")), plot,
                  width = width, height = height, limitsize = FALSE, bg = "white",
                  draw = draw, label = label)
}

fourier_m6a_plot_view <- function(result, original_m6a) {
  view <- result
  a <- result$assignments
  region <- result$region
  samples <- fourier_sample_table[match(unique(a$sample_name), fourier_sample_table$sample_name), , drop = FALSE]
  stopifnot(!anyNA(samples$sample_name), region$analysis_start == region$start + 1L,
            region$width == region$analysis_end - region$analysis_start + 1L)

  paths <- extracted_path(samples, region$chr, "m6a")
  bounds <- dplyr::bind_rows(lapply(seq_len(nrow(samples)), function(i) {
    query <- GenomicRanges::GRanges(region$chr, IRanges::IRanges(region$analysis_start, region$analysis_end))
    bed <- read_ft_bed12(paths[i], query, longest_alignment = TRUE)
    if (!nrow(bed)) return(NULL)
    selected <- a[a$sample_name == samples$sample_name[i], , drop = FALSE]
    matched <- match(bed$RID, selected$original_RID)
    data.frame(RID = selected$RID[matched], start = bed$start + 1L,
               end = bed$end)[!is.na(matched), , drop = FALSE]
  }))
  matched <- match(a$RID, bounds$RID)
  stopifnot(!anyDuplicated(bounds$RID), !anyNA(matched))
  view$assignments$start <- bounds$start[matched]
  view$assignments$end <- bounds$end[matched]
  stopifnot(all(view$assignments$start <= region$analysis_start),
            all(view$assignments$end >= region$analysis_end))
  view <- fiberseq_display_result(view)
  positions <- as.integer(colnames(original_m6a))
  stopifnot(identical(positions, seq.int(region$analysis_start, region$analysis_end)))
  view$met_mat <- original_m6a[a$RID, , drop = FALSE]
  view$site_met_mat <- view$met_mat[, Matrix::colSums(view$met_mat) > 0, drop = FALSE]
  view
}


fourier_save_m6a_heatmap <- function(view, split_alleles = FALSE) {
  filename <- if (split_alleles) "m6a_heatmap_split_by_haplotype" else "m6a_heatmap_by_fourier_cluster"
  heatmap <- plot_genomic_cluster_heatmap(view, view$region,
    sample_palette(view$assignments$sample_name), include_haplotype = TRUE,
    variants = view$variants, show_cluster_profiles = TRUE, split_alleles = split_alleles,
    cluster_label = "Fourier-derived clusters")
  fourier_save_plot(heatmap, view$region_id, filename,
    width = 14, height = 11 + .55 * view$n_clusters,
    draw = function(x) ComplexHeatmap::draw(x, newpage = FALSE),
    label = paste(view$region_id, filename))
}


fourier_matrix_plot <- function(mat, result, matrix_type = c("features", "m6a"), split_alleles = FALSE) {
  matrix_type <- match.arg(matrix_type)
  a <- result$assignments
  a <- a[if (split_alleles) order(a$allele_display, a$cluster, a$sample_name, a$RID)
           else order(a$cluster, a$allele_display, a$sample_name, a$RID), , drop = FALSE]
  a$read_rank <- rev(seq_len(nrow(a)))
  a$panel <- if (split_alleles) as.character(a$allele_display) else "All retained reads"
  values <- as.matrix(mat[a$RID, , drop = FALSE])
  coordinates <- if (matrix_type == "m6a") as.numeric(colnames(values)) else seq_len(ncol(values))
  long <- data.frame(read_rank = rep(a$read_rank, times = ncol(values)),
    coordinate = rep(coordinates, each = nrow(values)), value = as.vector(values),
    panel = rep(a$panel, times = ncol(values)))
  base_theme <- theme_classic(base_size = 10) +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          strip.background = element_blank(), strip.text.y = element_text(angle = 0))
  heat <- ggplot(long, aes(coordinate, read_rank, fill = value)) + geom_raster() +
    facet_grid(panel ~ ., scales = "free_y", space = "free_y") +
    scale_y_continuous(expand = c(0, 0)) + base_theme + labs(y = NULL)
  if (matrix_type == "features") {
    cap <- unname(quantile(abs(values), 0.99))
    if (!is.finite(cap) || cap <= 0) cap <- 1
    tick_k <- unique(round(seq(1, ncol(values) / 2, length.out = 6)))
    heat <- heat + scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-cap, cap), oob = scales::squish, name = "Normalized Re/Im") +
      scale_x_continuous(breaks = 2 * tick_k - 0.5, labels = tick_k, expand = c(0, 0)) +
      labs(x = "Fourier index k (adjacent Re/Im columns; period = window length / k)",
           caption = "Color clipped at the 99th percentile of absolute feature values; clustering uses unclipped features.")
  } else {
    heat <- heat + scale_fill_gradient(low = "white", high = "black", limits = c(0, 1),
      breaks = c(0, 1), name = "m6A call") +
      scale_x_continuous(labels = scales::label_number(accuracy = 1, big.mark = ","), expand = c(0, 0)) +
      geom_vline(xintercept = result$region$focal_pos, color = "#984EA3", linetype = 2, linewidth = 0.4) +
      labs(x = paste0(result$region$chr, " genomic position (bp)"),
           caption = paste("Dashed line:", result$region$focal_snp, "| Original full-window binary m6A calls"))
  }
  colors <- list(cluster = cluster_id_palette(levels(a$cluster)),
    allele_display = fiberseq_category_palette(levels(a$allele_display)))
  bars <- lapply(names(colors), function(column) {
    ggplot(a, aes(x = 1, y = read_rank, fill = .data[[column]])) + geom_tile() +
      facet_grid(panel ~ ., scales = "free_y", space = "free_y") +
      scale_fill_manual(values = colors[[column]], guide = "none") +
      scale_y_continuous(expand = c(0, 0)) + scale_x_continuous(expand = c(0, 0)) +
      base_theme + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
                        strip.text.y = element_blank()) +
      labs(x = if (column == "cluster") "Cluster" else "Allele", y = NULL)
  })
  combined <- cowplot::plot_grid(plotlist = c(bars, list(heat)), nrow = 1,
                                  rel_widths = c(0.055, 0.055, 1), align = "h", axis = "tb")
  title <- cowplot::ggdraw() + cowplot::draw_label(paste0(result$region_id, " | ", nrow(a),
    " reads | Fourier-derived Leiden clusters", if (split_alleles) " | Split by focal allele" else ""), size = 11)
  legends <- cowplot::plot_grid(
    fiberseq_legend(colors$cluster, "Cluster", columns = 6),
    fiberseq_legend(colors$allele_display, "Focal SNP allele", columns = 2), ncol = 1)
  cowplot::plot_grid(title, combined, legends, ncol = 1, rel_heights = c(0.35, 6, 1.15))
}
```

## Prepare data for the clustering figures


```{r Fourier-footprint-helpers}
fourier_footprint_tracks_dir <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks"
fourier_footprint_cache_dir <- file.path(fourier_clustering_dir, "footprint_summary")
fourier_footprint_caches <- fourier_lcl_helpers$cache_footprint_tracks(
  dplyr::bind_rows(lapply(fourier_leiden_results, function(result) result$region)),
  fourier_footprint_tracks_dir, fourier_footprint_cache_dir, workers = 1L)


fourier_footprint_view <- function(result, original_m6a) {
  view <- fourier_m6a_plot_view(result, original_m6a)
  assignments <- view$assignments
  samples <- fourier_sample_table[match(unique(assignments$sample_name),
    fourier_sample_table$sample_name), , drop = FALSE]
  aliases <- assignments
  aliases$RID <- paste(assignments$sample_name, assignments$original_RID, sep = "::")
  stopifnot(!anyDuplicated(aliases$RID))
  nuc <- extract_nucleosomes(samples, view$region, aliases)
  nuc$RID <- assignments$RID[match(nuc$RID, aliases$RID)]
  tf <- region_footprints(fourier_footprint_caches, view$region, assignments)$records
  footprints <- dplyr::bind_rows(nuc, tf)
  stopifnot(!anyNA(footprints$RID), all(footprints$RID %in% assignments$RID))
  signal <- signal_profiles(view, footprints)
  stopifnot(all(is.finite(signal$profiles$fraction)),
    all(signal$profiles$fraction >= 0 & signal$profiles$fraction <= 1))
  list(result = view, signal = signal)
}
```

```{r Fourier-plot-data}
fourier_plot_data <- setNames(vector("list", length(fourier_leiden_results)), names(fourier_leiden_results))
fourier_footprint_audit <- list()
for (region_id in names(fourier_leiden_results)) {
  message("Preparing Fourier figure data: ", region_id)
  prepared <- fourier_footprint_view(fourier_leiden_results[[region_id]],
                                    region_met_mats[[region_id]]$met_mat)
  fourier_plot_data[[region_id]] <- prepared
  view <- prepared$result
  signal <- prepared$signal
  fourier_footprint_audit[[region_id]] <- list(n_reads = nrow(view$assignments),
    n_clusters = view$n_clusters, track_record_counts = table(signal$records$track),
    plot_start = view$region$analysis_start, plot_end = view$region$analysis_end)
}
stopifnot(is.function(fourier_lcl_helpers$save_fiberseq_plots))
```

## 1. Dominant period by cluster and haplotype

```{r dominant_period_by_cluster_and_haplotype, results="asis", echo=TRUE, eval=TRUE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  a <- result$assignments
  x <- fourier_complex_results[[region_id]]
  spectrum <- fft_results[[region_id]]
  stopifnot(identical(spectrum$use_hann, x$use_hann))
  allele_colors <- fiberseq_category_palette(levels(a$allele_display))
  # each read's dominant period from positive frequency spectrum
  power <- spectrum$raw_power[a$RID, , drop = FALSE]
  dominant_k <- apply(power, 1, which.max)
  dominant <- data.frame(a, dominant_period_bp = x$periods_bp[dominant_k])
  dominant_plot <- ggplot(dominant, aes(cluster, dominant_period_bp, fill = allele_display)) +
    geom_boxplot(outlier.size = .6, position = position_dodge(width = .8)) +
    scale_fill_manual(values = allele_colors) + scale_y_continuous(breaks = seq(0, max(x$periods_bp), 250)) +
    theme_classic() + labs(title = region_id, subtitle = "Per-read dominant period; complete positive-frequency spectrum",
                          x = "Fourier-derived Leiden cluster", y = "Dominant period (bp)", fill = "Focal SNP allele")
  fourier_save_plot(dominant_plot, region_id, "dominant_period_by_cluster_and_haplotype", 12, 6)

 
  fourier_leiden_results[[region_id]]$dominant_periods <- dominant
}
```

## 2. Fiber-seq haplotype example

```{r fiberseq_haplotype_example, results="asis", echo=TRUE, eval=TRUE}
for (region_id in names(fourier_leiden_results)) {
  prepared <- fourier_plot_data[[region_id]]
  view <- prepared$result
  signal <- prepared$signal
  sample_colors <- sample_palette(view$assignments$sample_name)
  title <- paste(view$region$annotation, "Fourier-derived Leiden clusters", sep = " | ")
  settings_label <- paste0("k = ", view$params$k_eff, "; resolution = ", view$params$resolution,
                            "; Hann = ", view$params$use_hann)
  example_view <- view
  example_view$region$annotation <- paste(title, settings_label, sep = " | ")
  full_sample_colors <- setNames(unname(sample_colors[sub("_.*$", "", unique(view$assignments$sample_name))]),
    unique(view$assignments$sample_name))
  fourier_lcl_helpers$save_fiberseq_plots(example_view, signal$records,
    file.path(fourier_clustering_dir, region_id), full_sample_colors,
    embedding = view$embedding, example_only = TRUE,
    plot_writer = function(plot, name, width, height) {
      fourier_save_plot(plot, region_id, name, width, height)
    })
}
```

## 3. Individual-read footprints

```{r fig1_read_footprints, results="asis", echo=TRUE, eval=TRUE}
for (region_id in names(fourier_leiden_results)) {
  prepared <- fourier_plot_data[[region_id]]
  view <- prepared$result
  signal <- prepared$signal
  sample_colors <- sample_palette(view$assignments$sample_name)
  title <- paste(view$region$annotation, "Fourier-derived Leiden clusters", sep = " | ")
  settings_label <- paste0("k = ", view$params$k_eff, "; resolution = ", view$params$resolution,
                            "; Hann = ", view$params$use_hann)
  reads_plot <- plot_smf_reads(view, signal$records, sample_colors,
                                  include_haplotype = TRUE) +
    labs(title = title, subtitle = paste("One row per retained molecule;", settings_label))
  fourier_save_plot(reads_plot, region_id, "fig1_read_footprints", 12,
                    max(7, .03 * nrow(view$assignments) + 3.5 + .3 * view$n_clusters))
}
```

## 4. Occupancy by cluster

```{r fig2_occupancy_by_cluster, results="asis", echo=TRUE, eval=TRUE}
for (region_id in names(fourier_leiden_results)) {
  prepared <- fourier_plot_data[[region_id]]
  view <- prepared$result
  signal <- prepared$signal
  sample_colors <- sample_palette(view$assignments$sample_name)
  title <- paste(view$region$annotation, "Fourier-derived Leiden clusters", sep = " | ")
  settings_label <- paste0("k = ", view$params$k_eff, "; resolution = ", view$params$resolution,
                            "; Hann = ", view$params$use_hann)
  occupancy_plot <- plot_signal_profile(signal$profiles, view$region,
    levels(view$assignments$cluster), title = title) +
    labs(subtitle = paste("m6A calls and footprint occupancy;", settings_label))
  fourier_save_plot(occupancy_plot, region_id, "fig2_occupancy_by_cluster", 10,
                    1.4 * view$n_clusters + 2.5)
}
```

## 5. m6A heatmap grouped by Fourier cluster

```{r m6a_heatmap_by_fourier_cluster, results="asis", echo=TRUE, eval=TRUE}
for (region_id in names(fourier_leiden_results)) {
  view <- fourier_plot_data[[region_id]]$result
  fourier_save_m6a_heatmap(view, split_alleles = FALSE)
}
```

## 6. m6A heatmap split by haplotype

```{r m6a_heatmap_split_by_haplotype, results="asis", echo=TRUE, eval=TRUE}
for (region_id in names(fourier_leiden_results)) {
  view <- fourier_plot_data[[region_id]]$result
  fourier_save_m6a_heatmap(view, split_alleles = TRUE)
}
```

clustering summary stats

```{r Fourier-clustering-summary, echo=TRUE, results="asis"}
fourier_clustering_summary <- bind_rows(lapply(fourier_leiden_results, function(r) {
  data.frame(region_id = r$region_id, n_input_reads = nrow(r$read_filter_audit),
    n_clustered_reads = nrow(r$assignments), n_features = ncol(r$feat_mat),
    n_clusters = r$n_clusters, k_neighbors = r$params$k_eff, resolution = r$params$resolution,
    use_hann = r$params$use_hann)
}))
knitr::kable(fourier_clustering_summary)
```

```{r Fourier-knn-graphs, results="asis", echo=TRUE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  graph_plot <- plot_knn_graph(result,
    main = paste(region_id, "Fourier-feature Manhattan KNN graph colored by Leiden cluster", sep = "\n"))
  fourier_save_plot(graph_plot, region_id, "knn_graph", width = 12, height = 9,
    label = paste(region_id, "Fourier-feature Manhattan KNN graph"))
}
```

### UMAP by cluster, sample, and haplotype

```{r umap_cluster_sample_haplotype, results="asis", echo=TRUE, eval=FALSE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  a <- result$assignments
  cluster_colors <- cluster_id_palette(levels(a$cluster))
  allele_colors <- fiberseq_category_palette(levels(a$allele_display))
  short_colors <- sample_palette(unique(a$sample_name))
  sample_colors <- setNames(unname(short_colors[sub("_.*$", "", unique(a$sample_name))]), unique(a$sample_name))
  embedding <- result$embedding
  subtitle <- paste(preprocessing_label, "| Full-spectrum normalized Re/Im | k =", k_neighbors,
                    "| resolution =", resolution, "|", nrow(a), "reads")
  message("Plotting Fourier Leiden: ", region_id)

  overview <- cowplot::plot_grid(
    plot_umap(embedding, "cluster", cluster_colors, "Leiden cluster", show_legend = TRUE),
    plot_umap(embedding, "sample_name", sample_colors, "LCL sample",
      sub("_.*$", "", names(sample_colors)), show_legend = TRUE),
    plot_umap(embedding, "allele_display", allele_colors, "Focal SNP allele", show_legend = TRUE), ncol = 2)
  heading <- cowplot::ggdraw() + cowplot::draw_label(paste(region_id, subtitle, sep = "\n"), size = 11)
  fourier_save_plot(cowplot::plot_grid(heading, overview, ncol = 1, rel_heights = c(.6, 8)),
                   region_id, "umap_cluster_sample_haplotype", 14, 10)
}
```

### UMAP split by haplotype

```{r umap_split_by_haplotype, results="asis", echo=TRUE, eval=FALSE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  a <- result$assignments
  cluster_colors <- cluster_id_palette(levels(a$cluster))
  embedding <- result$embedding
  split_umap <- plot_umap(embedding, "cluster", cluster_colors, region_id, show_legend = TRUE) +
    facet_wrap(~ allele_display, nrow = 1) +
    labs(subtitle = "Same pooled Leiden assignments and UMAP coordinates, split by focal allele")
  fourier_save_plot(split_umap, region_id, "umap_split_by_haplotype", 12, 5)
}
```

### Fourier Re/Im feature heatmap

```{r fourier_re_im_heatmap, results="asis", echo=TRUE, eval=FALSE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  fourier_save_plot(fourier_matrix_plot(result$feat_mat, result, "features"),
                   region_id, "fourier_re_im_heatmap", 14, 9)
}
```

### Median power spectra by cluster

```{r median_power_spectra_by_cluster, results="asis", echo=TRUE, eval=FALSE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  a <- result$assignments
  x <- fourier_complex_results[[region_id]]
  spectrum <- fft_results[[region_id]]
  stopifnot(identical(spectrum$use_hann, x$use_hann))
  cluster_colors <- cluster_id_palette(levels(a$cluster))
  # aggregate per-read relative power
  aggregate <- bind_rows(lapply(levels(a$cluster), function(cl) {
    p <- spectrum$relative_power[a$RID[a$cluster == cl], , drop = FALSE]
    data.frame(cluster = cl, period_bp = spectrum$periods_bp,
      median_power = apply(p, 2, median), q25 = apply(p, 2, quantile, .25),
      q75 = apply(p, 2, quantile, .75), n_reads = nrow(p))
  }))
  aggregate$cluster <- factor(aggregate$cluster, levels = levels(a$cluster))
  spectra_plot <- ggplot(aggregate, aes(period_bp, median_power, color = cluster, fill = cluster)) +
    geom_ribbon(aes(ymin = q25, ymax = q75), alpha = .2, color = NA) + geom_line(linewidth = .5) +
    facet_wrap(~ cluster, ncol = 3) + scale_color_manual(values = cluster_colors) +
    scale_fill_manual(values = cluster_colors) + scale_x_continuous(breaks = seq(0, max(x$periods_bp), 500)) +
    theme_classic() + theme(legend.position = "none") +
    labs(title = region_id, subtitle = "Median relative-power spectrum per Fourier cluster; ribbon = IQR",
         x = "Spatial period (bp)", y = "Relative power")
  fourier_save_plot(spectra_plot, region_id, "median_power_spectra_by_cluster", 12,
                   max(5, 2.8 * ceiling(result$n_clusters / 3)))

  fourier_leiden_results[[region_id]]$aggregate_spectra <- aggregate
}
```

### Fourier phase by haplotype

```{r phase_at_cluster_peak_by_haplotype, results="asis", echo=TRUE, eval=FALSE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  a <- result$assignments
  x <- fourier_complex_results[[region_id]]
  aggregate <- result$aggregate_spectra
  allele_colors <- fiberseq_category_palette(levels(a$allele_display))
  phase_plot_data <- bind_rows(lapply(levels(a$cluster), function(cl) {
    ag <- aggregate[aggregate$cluster == cl, ]
    peak_period <- max(ag$period_bp[ag$median_power == max(ag$median_power)])
    peak_k <- match(peak_period, x$periods_bp)
    reads <- a[a$cluster == cl, , drop = FALSE]
    data.frame(reads, phase_radians = x$phase_radians[reads$RID, peak_k],
      phase_period_bp = peak_period,
      panel = paste0(cl, " | ", round(peak_period, 3), " bp"))
  }))
  phase_plot <- ggplot(phase_plot_data, aes(phase_radians, fill = allele_display)) +
    geom_histogram(breaks = seq(-pi, pi, length.out = 25), position = "identity", alpha = .55, na.rm = TRUE) +
    facet_wrap(~ panel, ncol = 3) + scale_fill_manual(values = allele_colors) +
    scale_x_continuous(limits = c(-pi, pi), breaks = c(-pi, 0, pi), labels = c("-pi", "0", "pi")) +
    theme_classic() + labs(title = region_id,
      subtitle = "Phase at each cluster's median-spectrum peak; the same frequency is used for all reads in a panel",
      caption = "Angles are circular: -pi and pi coincide. Phase is relative to the original Fourier window start; weak components have uncertain phase.",
      x = "Fourier phase (radians)", y = "Reads", fill = "Focal SNP allele")
  fourier_save_plot(phase_plot, region_id, "phase_at_cluster_peak_by_haplotype", 13,
                   max(5, 2.8 * ceiling(result$n_clusters / 3)))

  fourier_leiden_results[[region_id]]$phase_plot_data <- phase_plot_data
}
```

### Cluster composition

```{r cluster_haplotype_sample_composition, results="asis", echo=TRUE, eval=FALSE}
for (region_id in names(fourier_leiden_results)) {
  result <- fourier_leiden_results[[region_id]]
  a <- result$assignments
  allele_colors <- fiberseq_category_palette(levels(a$allele_display))
  short_colors <- sample_palette(unique(a$sample_name))
  composition <- plot_category_composition(result, "allele_display", allele_colors,
    main = "Focal allele composition within each cluster", legend_title = "Focal SNP allele")
  sample_composition <- plot_sample_composition(result, short_colors,
    main = "Sample composition within each Fourier cluster")
  fourier_save_plot(cowplot::plot_grid(composition, sample_composition, ncol = 1),
                   region_id, "cluster_haplotype_sample_composition", 12, 10)
}
```

### Read retention by region

```{r read_retention_by_region, results="asis", echo=TRUE, eval=FALSE}
retention_counts <- bind_rows(fourier_clustering_audits) %>% count(region_id, exclusion_reason)
retention_plot <- ggplot(retention_counts, aes(region_id, n, fill = exclusion_reason)) +
  geom_col() + coord_flip() + theme_classic() +
  labs(title = "Read selection for Fourier-derived Leiden clustering", x = NULL, y = "Reads", fill = NULL)
save_report_plot(file.path(fourier_clustering_dir, "read_retention_by_region.pdf"), retention_plot,
       width = 12, height = 7, bg = "white")
```
