---
title: "LPS time course: DESeq2 likelihood-ratio test"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5, dpi = 150)
```

# File path

Point `counts_dir` at the STAR output folder: it holds `sample_names.txt` (the sample
sheet) and one subdirectory per sample containing `<sample>_ReadsPerGene.out.tab`.

```{r config}
counts_dir <- "/project/spott/cshan/fiber-seq/code/RNA/aligned"
sheet_path <- file.path(counts_dir, "sample_names.txt")
gtf_path   <- "/project/spott/lizarraga/human_references/gencode.v46.annotation.gtf"  
out_dir    <- "/project/spott/cshan/fiber-seq/macrophage_project/RNA-seq/deseq2_lrt_results"
fig_dir    <- "/project/spott/cshan/fiber-seq/macrophage_project/RNA-seq/figures"

strand_col <- 4      # STAR column: 2 = unstranded, 3 = forward, 4 = reverse
fdr_cut    <- 0.05
lfc_cut    <- 1      # log2FC threshold for the core genes responsive to LPS 

dir.create(out_dir, showWarnings = FALSE)
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

# save plots
save_fig <- function(p, name, w, h) {         
  ggsave(file.path(fig_dir, paste0(name, ".png")), p,
         width = w, height = h, dpi = 300)
}
```

```{r libs}
suppressPackageStartupMessages({
  library(DESeq2)
  library(edgeR)     
  library(splines)
  library(ggplot2)
  library(ggrepel)
  library(pheatmap)
})
```

# Sample sheet

18 libraries: 6 timepoints (0, 5, 10, 15, 30, 60 min) x 3 replicates, read from
`sample_names.txt`. Baseline is 0 min

R1/R2/R3 are paired batches present at every timepoint

```{r samples}
samples <- read.delim(sheet_path, header = TRUE, stringsAsFactors = FALSE)
colnames(samples) <- c("sample", "replicate", "time")
samples$sample    <- trimws(samples$sample)
samples$replicate <- trimws(samples$replicate)
samples$time      <- as.numeric(samples$time)
samples <- samples[order(samples$time, samples$replicate), ]
stopifnot(nrow(samples) == 18, !anyDuplicated(samples$sample))
samples$time_f <- factor(samples$time, levels = c(0, 5, 10, 15, 30, 60))
samples$rep    <- factor(samples$replicate)
rownames(samples) <- samples$sample
samples
```

# Read STAR counts

`ReadsPerGene.out.tab` has no header and four columns: gene ID, unstranded, forward,
reverse. The first four rows are alignment summary lines (`N_unmapped`,
`N_multimapping`, `N_noFeature`, `N_ambiguous`) — kept aside as QC, not as genes.

```{r read-counts}
read_star <- function(sample, dir, col) {
  f <- file.path(dir, sample, paste0(sample, "_ReadsPerGene.out.tab"))
  if (!file.exists(f)) stop("missing count file: ", f)
  d <- read.delim(f, header = FALSE, row.names = 1,
                  col.names = c("gene", "unstranded", "fwd", "rev"))
  # get the summary stats 
  list(qc     = d[grepl("^N_", rownames(d)), col - 1],
       counts = d[!grepl("^N_", rownames(d)), col - 1, drop = TRUE],
       genes  = rownames(d)[!grepl("^N_", rownames(d))],
       # for non summary stats, gets the total number of counts
        # stranded
        # fwd
        # reverse
       all3   = colSums(d[!grepl("^N_", rownames(d)), ]))
}

star <- lapply(samples$sample, read_star, dir = counts_dir, col = strand_col)
names(star) <- samples$sample

gene_ids <- star[[1]]$genes
stopifnot(all(vapply(star, function(x) identical(x$genes, gene_ids), logical(1))))

counts <- do.call(cbind, lapply(star, `[[`, "counts"))
rownames(counts) <- gene_ids
colnames(counts) <- samples$sample

cat("genes:", nrow(counts), " samples:", ncol(counts), "\n")
```

## Strandedness check

The library is reverse-stranded if column 4 carries far more reads than column 3.
If this comes out the other way round, change `strand_col` above and re-run.

```{r strand-check}
s3 <- colSums(do.call(rbind, lapply(star, function(x) x$all3)))
names(s3) <- c("unstranded", "fwd", "rev")
print(round(s3 / 1e6, 1))                       # millions of reads per STAR column
cat("\nchosen column:", strand_col,
    "| reverse/forward ratio:", round(s3[["rev"]] / s3[["fwd"]], 1), "\n")
if (s3[["rev"]] < s3[["fwd"]])
  warning("Forward strand dominates - strand_col may be wrong.")
```

## Alignment QC

Takes the qc column in N_unmapped, N_multimapping, N_noFeature, N_ambiguous
```{r qc-table}
qc <- do.call(rbind, lapply(star, function(x) x$qc))
colnames(qc) <- c("N_unmapped", "N_multimapping", "N_noFeature", "N_ambiguous")
qc <- as.data.frame(qc)
# counts how many reads that are 
qc$assigned     <- colSums(counts)
qc$total        <- rowSums(qc[, c(1:4, 5)])

# calculates the percentage of reads that were successfully assigned to genes = 100 x (assigned reads / total reads)
qc$pct_assigned <- round(100 * qc$assigned / qc$total, 1)

# counts how many genes have at least one each in each sample
qc$genes_detect <- colSums(counts > 0)
write.csv(qc, file.path(out_dir, "star_alignment_qc.csv"))
qc[, c("assigned", "pct_assigned", "genes_detect")]
```

# Gene annotation 

Builds a gene annotation table for the genes in your RNA-seq count matrix.keeps only gene-level records, extracts each gene’s gene_id, gene_name, and gene_type from the attribute column, matches those annotations back to your count-matrix gene IDs, and reports how many genes were successfully annotate

chr1    HAVANA  gene    11869   14409   .       +       .       gene_id "ENSG00000290825.1"; gene_type "lncRNA"; gene_name "DDX11L2"; level 2; tag "overlaps_pseudogene";

chr1    HAVANA  transcript      11869   14409   .       +       .       gene_id "ENSG00000290825.1"; transcript_id "ENST00000456328.2"; gene_type "lncRNA"; gene_name "DDX11L2"; transcript_type "lncRNA"; transcript_name "DDX11L2-202"; level 2; transcript_support_level "1"; tag "basic"; tag "Ensembl_canonical"; havana_transcript "OTTHUMT00000362751.1";

chr1    HAVANA  exon    11869   12227   .       +       .       gene_id "ENSG00000290825.1"; transcript_id "ENST00000456328.2"; gene_type "lncRNA"; gene_name "DDX11L2"; transcript_type "lncRNA"; transcript_name "DDX11L2-202"; exon_number 1; exon_id "ENSE00002234944.1"; level 2; transcript_support_level "1"; tag "basic"; tag "Ensembl_canonical"; havana_transcript "OTTHUMT00000362751.1";

```{r annotation}
ann <- data.frame(gene_id = gene_ids, gene_name = NA_character_,
                  gene_type = NA_character_, row.names = gene_ids,
                  stringsAsFactors = FALSE)

if (!is.na(gtf_path) && file.exists(gtf_path)) {
  con <- if (grepl("\\.gz$", gtf_path)) gzfile(gtf_path) else file(gtf_path)
  gtf <- read.delim(con, header = FALSE, comment.char = "#", quote = "",
                    col.names = c("chr","src","feat","start","end","score","strand","frame","attr"))
  gtf <- gtf[gtf$feat == "gene", ]
  grab <- function(x, key) {
    m <- regmatches(x, regexpr(paste0(key, ' "[^"]*"'), x))
    ifelse(length(m) == 0, NA_character_, gsub(paste0(key, ' "|"'), "", m))
  }
  gid  <- vapply(gtf$attr, grab, character(1), key = "gene_id",   USE.NAMES = FALSE)
  gnm  <- vapply(gtf$attr, grab, character(1), key = "gene_name", USE.NAMES = FALSE)
  gtp  <- vapply(gtf$attr, grab, character(1), key = "gene_type", USE.NAMES = FALSE)
  hit  <- match(gene_ids, gid)
  ann$gene_name <- gnm[hit]
  ann$gene_type <- gtp[hit]
  cat("annotated", sum(!is.na(ann$gene_name)), "of", nrow(ann), "genes\n")
} else {
  message("GTF not found - gene_name/gene_type left blank.")
}
write.csv(ann, file.path(out_dir, "gene_annotation.csv"), row.names = FALSE)
```

# Filter low-count genes

`filterByExpr` is given the full design so it keeps genes expressed in *any* timepoint
group — important here, because the interesting genes are off at baseline and on at 60 min.

creates the design matrix: expression ~ replicate + timepoint
- Y=β0 +βrep2 + βrep3+ βtime5 + βtime10 +⋯+ϵ
- beta = baseline

```{r filter}
design_full <- model.matrix(~ rep + time_f, samples)
keep   <- filterByExpr(DGEList(counts, samples = samples), design = design_full)
counts_f <- counts[keep, ]
cat("kept", nrow(counts_f), "of", nrow(counts), "genes\n")
write.csv(counts_f, file.path(out_dir, "counts_filtered.csv"))

head(counts_f)
```

# DESeq2 likelihood-ratio test

The LRT compares the full model (`~ replicate + time`) against a reduced model with the
time term dropped. It answers "does this gene change *anywhere* across the course", which
is the right omnibus question for a time series 

The full model includes both replicate and timepoint effects, while the reduced model includes replicate only. 
The LRT asks whether including time significantly improves the fit, so a significant gene is one whose expression changes at 
at least one timepoint after accounting for replicate differences.

```{r dds-lrt}

# fits both full and reduced models internally for each gene and compares them
dds <- DESeqDataSetFromMatrix(counts_f, colData = samples, design = ~ rep + time_f)
dds <- DESeq(dds, test = "LRT", reduced = ~ rep, quiet = TRUE)

# LRT result comparing the two models
## LRT was performed on original negative binomial count model
res <- results(dds)

# variance stabilizing transformation
## Raw RNA-seq counts have mean-dependent variance: highly expressed genes tend to have much larger variance than low-expression genes. VST transforms the counts so that variance is more comparable across the expression range
vsd  <- vst(dds, blind = TRUE)

# extract transformed expression value as an ordinary matrix
vmat <- assay(vsd)
write.csv(vmat, file.path(out_dir, "vst_matrix.csv"))
summary(res)
```
```{r dds-results}
results(dds)
```


## Per-timepoint fold changes vs baseline

takes the overall DESeq2 LRT result and adds timepoint-specific log2 fold changes versus 0 min so that, for each gene, you know both whether it changes anywhere across the time course and how large/directional the change is at 5, 10, 15, 30, and 60 min


- baseMean is the average normalized expression level of that gene across all 18 samples. It is based on DESeq2 size-factor-normalized counts, not the raw counts. A large baseMean means the gene is generally highly expressed; a small baseMean means it is lowly expressed. It does not tell you whether the gene changes over time

- stat is the likelihood-ratio test statistic for comparing the full model ~ rep + time_f against the reduced model ~ rep. It measures how much better the model fits when the time effect is included

```{r lfc-table}

# foe each time point (not including 0 min), extracts DESeq2 log2 fold change coefficient relative to 0 min
lfc <- sapply(c(5, 10, 15, 30, 60), function(t)
  results(dds, name = paste0("time_f_", t, "_vs_0"))$log2FoldChange)
colnames(lfc) <- paste0("lfc_", c(5, 10, 15, 30, 60), "_vs_0")

# builds results table with annotations, LRT statistics, all fold changes, and the largest absolute fold change across time
# sorts genes by FDR
# counts how many genes are significant and prints the top genes

lrt_tab <- data.frame(
  gene_id   = rownames(res),
  gene_name = ann[rownames(res), "gene_name"],
  gene_type = ann[rownames(res), "gene_type"],
  baseMean  = round(res$baseMean, 2),
  stat = round(res$stat, 3), pvalue = res$pvalue, padj = res$padj,
  round(lfc, 3),
  # calculates the largest absolute fold change for each gene at any timepoint 
  max_abs_lfc = round(apply(abs(lfc), 1, max), 3),
  row.names = rownames(res), stringsAsFactors = FALSE
)
lrt_tab <- lrt_tab[order(lrt_tab$padj), ]
write.csv(lrt_tab, file.path(out_dir, "deseq2_lrt_timecourse.csv"), row.names = FALSE)

# the most statistically significant time-responsive genes
head(lrt_tab)

# sigificant time responsive genes with the largest expression changes
lrt_sig <- lrt_tab[lrt_tab$padj < fdr_cut, ]
lrt_sig <- lrt_sig[order(lrt_sig$max_abs_lfc, decreasing = TRUE), ]

head(lrt_sig[, c("gene_name", "baseMean", "max_abs_lfc", "padj")], 10)

lrt_ <- lrt_tab[order(lrt_tab$max_abs_lfc, decreasing = TRUE), ]

# sum the number of significant has FDR < 0.05
n_sig <- sum(lrt_tab$padj < fdr_cut, na.rm = TRUE)
cat("LRT significant at FDR <", fdr_cut, ":", n_sig, "\n")
cat("  of which |log2FC| >", lfc_cut, ":",
    sum(lrt_tab$padj < fdr_cut & lrt_tab$max_abs_lfc > lfc_cut, na.rm = TRUE), "\n")
head(lrt_tab[, c("gene_name","baseMean","max_abs_lfc","padj")], 15)
```

## continuous-time spline

fits a second DESeq2 time-course model where time is treated as a smooth continuous trajectory using a natural spline with 3 degrees of freedom, rather than as six separate categorical timepoints, does not treat every timepoint as completely independent

compares which genes are significant under the factor-based LRT versus the spline-based LRT

- does any individual time points differ?

- whether the genes follows a significant smooth temporal pattern?

- Does not treat each gene as independent, does adding a smooth time independent trajectory significantly improve the model?

```{r spline}
# Fits the spline model and performs the likelihood-ratio test
# time rescaled to 0..1 so DESeq2 doesn't warn about a large-mean numeric covariate
samples$time_sc <- samples$time / 60
dds_s <- DESeqDataSetFromMatrix(counts_f, colData = samples,
                                design = ~ rep + ns(time_sc, df = 3))
dds_s <- DESeq(dds_s, test = "LRT", reduced = ~ rep, quiet = TRUE)

# Extracts the LRT results from the spline model
res_s <- results(dds_s)

spl <- data.frame(gene_id = rownames(res_s),
                  gene_name = ann[rownames(res_s), "gene_name"],
                  baseMean = round(res_s$baseMean, 2),
                  pvalue = res_s$pvalue, padj = res_s$padj,
                  row.names = rownames(res_s))

write.csv(spl[order(spl$padj), ], file.path(out_dir, "deseq2_lrt_spline_ns3.csv"),
          row.names = FALSE)

set_f <- rownames(lrt_tab)[which(lrt_tab$padj < fdr_cut)]
set_s <- rownames(spl)[which(spl$padj < fdr_cut)]
cat("factor LRT:", length(set_f), "| spline LRT:", length(set_s),
    "| overlap:", length(intersect(set_f, set_s)), "\n")
```

## Venn diagram: factor LRT vs spline LRT

Two circles drawn with `ggforce::geom_circle`; the three counts are factor-only,
overlap, and spline-only significant genes at FDR < `r fdr_cut`.

```{r spline-venn, fig.width=5.5, fig.height=4}
library(ggforce)

n_overlap <- length(intersect(set_f, set_s))
n_f_only  <- length(setdiff(set_f, set_s))
n_s_only  <- length(setdiff(set_s, set_f))

venn_circ <- data.frame(x = c(-0.7, 0.7), y = 0, r = 1.5,
                        test = c("factor LRT", "spline LRT"))
venn_lab  <- data.frame(x = c(-1.5, 0, 1.5), y = 0,
                        n = c(n_f_only, n_overlap, n_s_only))

p_venn <- ggplot() +
  geom_circle(data = venn_circ, aes(x0 = x, y0 = y, r = r, fill = test),
              alpha = 0.35, colour = "grey30", linewidth = 0.3) +
  geom_text(data = venn_lab, aes(x, y, label = n), size = 4.5) +
  annotate("text", x = c(-1.5, 1.5), y = 1.75,
           label = c(paste0("factor LRT (", length(set_f), ")"),
                     paste0("spline LRT (", length(set_s), ")")), size = 3.5) +
  scale_fill_manual(values = c("factor LRT" = "steelblue", "spline LRT" = "goldenrod")) +
  coord_fixed() + theme_void() +
  theme(legend.position = "none") +
  labs(title = paste0("Significant genes at FDR < ", fdr_cut)) +
  theme(plot.title = element_text(hjust = 0.5, size = 11))
save_fig(p_venn, "venn_factor_vs_spline_lrt", 5.5, 4)
p_venn
```

## Top 10 significant genes by fold change, per test

The spline LRT has no per-timepoint fold change (its coefficients are spline basis
terms), so both lists are ranked by `max_abs_lfc` — the largest |log2FC| vs 0 min
across timepoints from the factor model — after filtering by each test's own padj.

```{r spline-top10}
top10_by_lfc <- function(ids, padj_vec, test_name) {
  d <- lrt_tab[ids, c("gene_name", "gene_type", "baseMean",
                      "lfc_30_vs_0", "lfc_60_vs_0", "max_abs_lfc")]
  d$padj <- signif(padj_vec[ids], 3)
  d$test <- test_name
  head(d[order(-d$max_abs_lfc), ], 10)
}

top10 <- rbind(
  top10_by_lfc(set_f, setNames(lrt_tab$padj, rownames(lrt_tab)), "factor LRT"),
  top10_by_lfc(set_s, setNames(spl$padj,     rownames(spl)),     "spline LRT"))
write.csv(top10, file.path(out_dir, "top10_lfc_factor_vs_spline.csv"))
top10
```

# All 15 pairwise timepoint contrasts

pairwise differential-expression comparisons between the six timepoints using DESeq2’s Wald test, then shrinks the log2 fold changes to make effect-size estimates more stable

For every timepoint pair, it saves a full gene-level result table, counts how many significant genes go up or down, counts how many also exceed your fold-change threshold

```{r pairwise}
tps   <- c(0, 5, 10, 15, 30, 60)
# generates every unique pair of timepoints
pairs <- t(combn(tps, 2))

# test specific contrasts between individual timepoints
dds_w <- nbinomWaldTest(dds, quiet = TRUE)

# full gene results for every contrast
pw_list <- list()

# one summary row per contrast
pw_summary <- list()

for (i in seq_len(nrow(pairs))) {
  # gets two timepoints for the current comparison
  a <- pairs[i, 1]; b <- pairs[i, 2]          
  nm <- paste0("t", b, "_vs_t", a)
  # extracts the wald result test for b min - a min 
  ## is the expression at b min different from expression at a min?
  ## positive LFC = higher expression at later timepoint b
  r  <- results(dds_w, 
                contrast = c("time_f", as.character(b), as.character(a)))
  
  # log2 fold change shrinkage
  ## pulls uncertain LFC estimates toward zero
  r  <- lfcShrink(dds_w, 
                  contrast = c("time_f", as.character(b), as.character(a)),
                  res = r, 
                  type = "ashr", quiet = TRUE)
  
  d <- data.frame(gene_id = rownames(r),
                  gene_name = ann[rownames(r), "gene_name"],
                  gene_type = ann[rownames(r), "gene_type"],
                  baseMean = round(r$baseMean, 2),
                  log2FoldChange = round(r$log2FoldChange, 3),
                  lfcSE = round(r$lfcSE, 3),
                  pvalue = r$pvalue, padj = r$padj,
                  contrast = nm,
                  # unscaled timepoints (minutes) for this contrast
                  t_num = b, t_ref = a,
                  row.names = NULL, stringsAsFactors = FALSE)
  
  
  d <- d[order(d$padj), ]
  write.csv(d, file.path(out_dir, paste0("pairwise_", nm, ".csv")), row.names = FALSE)
  pw_list[[nm]] <- d

  # find the rows for sig genes
  sig <- which(d$padj < fdr_cut)
  pw_summary[[nm]] <- data.frame(
    contrast = nm, t_num = b, t_ref = a,
    # counts sig gens with positive LFC
    n_up   = sum(d$log2FoldChange[sig] > 0),
    n_down = sum(d$log2FoldChange[sig] < 0),
    # count all genes sig at FDR threshold, regardless of effect size
    n_total_fdr = length(sig),
    # count sig genes has FDR < 0.05 and lfc > 1
    n_total_lfc1 = sum(d$padj < fdr_cut & abs(d$log2FoldChange) > lfc_cut, na.rm = TRUE))
}
pw_all <- do.call(rbind, pw_list)
write.csv(pw_all, file.path(out_dir, "pairwise_combined_long.csv"), row.names = FALSE)

pw_sum <- do.call(rbind, pw_summary)
write.csv(pw_sum, file.path(out_dir, "pairwise_summary.csv"), row.names = FALSE)
pw_sum

head(pw_all[pw_all$gene_type == "protein_coding", ])
```

# Differentially expressed genes heatmap

```{r plot-pairwise, fig.height=3.6}
p_pw <- ggplot(pw_sum, aes(factor(t_ref, levels = tps), factor(t_num, levels = tps),
                   fill = n_total_fdr)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = ifelse(n_total_fdr > 0, n_total_fdr, "0")), size = 2.8) +
  scale_fill_viridis_c(name = "DE genes", trans = "log1p",
                       breaks = c(0, 10, 100, 1000)) +
  labs(x = "Reference timepoint (min)", y = "Compared timepoint (min)",
       title = "Differentially expressed genes per pairwise contrast",
       subtitle = paste0("DESeq2 Wald, ashr-shrunk, FDR < ", fdr_cut)) +
  theme_minimal(base_size = 9)
save_fig(p_pw, "pairwise_de_counts", 8, 3.6)
p_pw
```

## Volcano plots: each timepoint vs baseline on pairwise Wald test


```{r volcano, fig.width=9, fig.height=5.5}
v0 <- pw_all[pw_all$contrast %in% paste0("t", tps[-1], "_vs_t0"), ]
v0$facet <- factor(v0$contrast, levels = paste0("t", tps[-1], "_vs_t0"),
                   labels = paste0(tps[-1], " min vs 0 min"))
v0$class <- "NS"
v0$class[!is.na(v0$padj) & v0$padj < fdr_cut] <- "FDR"
v0$class[v0$class == "FDR" & abs(v0$log2FoldChange) >= lfc_cut] <- "FDR & LFC"
v0$class <- factor(v0$class, levels = c("NS", "FDR", "FDR & LFC"))

vlab <- do.call(rbind, lapply(split(v0, v0$facet), function(d) {
  d <- d[d$class == "FDR & LFC" & !is.na(d$gene_name), ]
  head(d[order(d$padj), ], 10)
}))

p_volc <- ggplot(v0, aes(log2FoldChange, -log10(padj), colour = class)) +
  geom_point(size = 0.4, alpha = 0.5, na.rm = TRUE) +
  geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = 3, linewidth = 0.3) +
  geom_hline(yintercept = -log10(fdr_cut), linetype = 3, linewidth = 0.3) +
  geom_text_repel(data = vlab, aes(label = gene_name), colour = "black", size = 2.3,
                  max.overlaps = Inf, segment.size = 0.2, box.padding = 0.15) +
  scale_colour_manual(name = NULL,
                      values = c(NS = "grey75", FDR = "steelblue", `FDR & LFC` = "firebrick"),
                      labels = c("NS", paste("padj <", fdr_cut),
                                 paste0("padj < ", fdr_cut, " & |LFC| >= ", lfc_cut))) +
  facet_wrap(~ facet, nrow = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "log2 fold change vs 0 min (ashr-shrunk)", y = "-log10 adjusted p",
       title = "Volcano per timepoint vs baseline") +
  theme_minimal(base_size = 9)
save_fig(p_volc, "volcano_vs_baseline", 9, 5.5)
p_volc
```

# Response gene definition

tiers are
defined on **evidence x effect size**:

* **core** — `padj < 0.05` and `|log2FC| >= 1` at some timepoint
* **subtle** — `padj < 0.05` but `|log2FC| < 1`

Takes the genes that were significant in the overall DESeq2 time-course LRT and divides them into response strength categories based on their largest absolute log2 fold change across 5, 10, 15, 30, and 60 minutes versus 0 minutes

Genes with padj < 0.05 and max_abs_lfc >= 1 are labeled core, meaning they show both statistically significant and relatively large changes; genes with padj < 0.05 but max_abs_lfc < 1 are labeled subtle

```{r response-genes}
lt <- lrt_tab[!is.na(lrt_tab$padj), ]
lt$response_tier <- ifelse(lt$padj >= fdr_cut, "not_called",
                    ifelse(lt$max_abs_lfc >= lfc_cut, "core", "subtle"))
peak_i <- apply(lt[, paste0("lfc_", c(5,10,15,30,60), "_vs_0")], 1,
                function(x) which.max(abs(x)))
lt$peak_timepoint_min <- c(5, 10, 15, 30, 60)[peak_i]
lt$direction <- ifelse(lt$lfc_60_vs_0 > 0, "up", "down")
lt$direction[lt$response_tier == "not_called"] <- ""

# fraction of the 60-min change already reached by 30 min
lt$frac_by_30 <- round(ifelse(abs(lt$lfc_60_vs_0) > 0.1,
                              lt$lfc_30_vs_0 / lt$lfc_60_vs_0, NA), 3)
lt$early_responder <- !is.na(lt$frac_by_30) & lt$frac_by_30 >= 0.5 &
                      lt$response_tier != "not_called"

resp <- lt[lt$response_tier != "not_called", ]
write.csv(resp, file.path(out_dir, "response_genes.csv"), row.names = FALSE)
table(tier = resp$response_tier, direction = resp$direction)
cat("\ntotal response genes:", nrow(resp),
    "| early responders (>=50% of change by 30 min):", sum(resp$early_responder), "\n")



lt_early <- lt[lt$early_responder == TRUE, ]
lt_early <- lt_early[order(lt_early$max_abs_lfc, decreasing = TRUE), ]
lt_early[lt_early$gene_type == "protein_coding", ]
```

## genes silent at baseline

looks for genes that are essentially silent at baseline but strongly induced by 60 minutes. It calculates the mean raw count across the three 0-min replicates and across the three 60-min replicates, keeps genes with mean count ≤ 1 at 0 min and ≥ 20 at 60 min, checks whether each of those genes was also called significant by the overall DESeq2 LRT, sorts them by strongest 60-min expression

```{r zero-baseline}
t0_s  <- samples$sample[samples$time == 0]
t60_s <- samples$sample[samples$time == 60]
silent <- data.frame(
  gene_id   = rownames(counts_f),
  gene_name = ann[rownames(counts_f), "gene_name"],
  mean_t0   = round(rowMeans(counts_f[, t0_s]), 1),
  mean_t60  = round(rowMeans(counts_f[, t60_s]), 1),
  padj      = lrt_tab[rownames(counts_f), "padj"],
  row.names = rownames(counts_f), stringsAsFactors = FALSE
)
silent <- silent[silent$mean_t0 <= 1 & silent$mean_t60 >= 20, ]
silent$called <- !is.na(silent$padj) & silent$padj < fdr_cut
silent <- silent[order(-silent$mean_t60), ]
write.csv(silent, file.path(out_dir, "zero_baseline_induced.csv"), row.names = FALSE)

cat("silent at baseline & induced by 60 min:", nrow(silent),
    "| called by DESeq2 LRT:", sum(silent$called),
    "| MISSED:", sum(!silent$called), "\n\n")
silent[!silent$called, ]
```

# Temporal clustering of genes

clusters the significant genes into temporal expression-pattern groups based on how their VST-transformed expression changes from baseline across the six timepoints. It first averages expression across the three replicates at each timepoint, subtracts the 0-min baseline to get a time-course change profile, scales each gene by its own maximum absolute change so clustering focuses on trajectory shape rather than magnitude, performs Ward hierarchical clustering, then picks the number of clusters k: by default the same tree is cut at every k from 2 to 10 and the k with the highest mean silhouette width (most compact, best-separated clusters) is kept; setting `k_manual` to an integer overrides this and uses that k instead. Each response gene is assigned to one of the k clusters, the assignments are saved, and the average expression-change profile is calculated per cluster

- genes that are significant by LRT are clustered



1. **heatmap** — every response gene x every library. The only view that shows the full
   set, and per-sample columns keep the 30-min replicate desynchronization visible
   instead of averaging it away.
2. **module line plot** — every gene as a faint line plus the module mean, so the module
   shape and its spread are both visible. Direct labels on 100-800 genes per panel are
   unreadable, so only exemplars (top genes per module by LRT padj) are drawn in colour
   and labelled at the line ends; the full membership goes to `cluster_top_genes.csv`.
3. **per-gene trajectories** — top genes individually, per-replicate lines on a linear
   minutes axis (equal-spaced timepoint axes visually exaggerate the early course, where
   nothing happens; linear time puts the 30-60 min burst in proportion).

```{r clusters}
set.seed(1)

# For each timepoint, selects the three replicate samples and calculates the mean VST expression for every gene
tp_mean <- sapply(tps, function(t) rowMeans(vmat[, samples$sample[samples$time == t]]))
colnames(tp_mean) <- tps

# Subtracts each gene's 0-minute expression from every timepoint
delta <- tp_mean - tp_mean[, "0"]

# keep only genes that were called significant response genes by the LRT
d_resp <- delta[rownames(resp), , drop = FALSE]

# Scales each gene by its own largest absolute change
scaled <- d_resp / pmax(apply(abs(d_resp), 1, max), 1e-9)

# calculates Euclidean distance between every pair of scaled gene trajectories
dmat  <- dist(scaled)

# starts with every gene as its own cluster and repeatedly merges clusters so that each merge causes the smallest possible increase in within-cluster variation
hc    <- hclust(dmat, method = "ward.D2")

# choose k by mean silhouette width: cut the same tree at each candidate k and
# keep the k whose clusters are most compact and well separated
ks  <- 2:10
sil <- sapply(ks, function(kk)
  mean(cluster::silhouette(cutree(hc, k = kk), dmat)[, "sil_width"]))
k_sil <- ks[which.max(sil)]

# k_manual: set to an integer (e.g. 5) to override the silhouette choice;
# leave as NA (default) to use the silhouette-selected k
k_manual <- NA
k <- if (is.na(k_manual)) k_sil else k_manual

p_sil <- ggplot(data.frame(k = ks, sil = sil), aes(k, sil)) +
  geom_line(linewidth = 0.4) + geom_point(size = 1.5) +
  geom_point(data = data.frame(k = k_sil, sil = max(sil)), colour = "firebrick", size = 2.5) +
  scale_x_continuous(breaks = ks) +
  labs(x = "number of clusters k", y = "mean silhouette width",
       title = paste0("Silhouette-selected k = ", k_sil,
                      if (!is.na(k_manual)) paste0(" (manual override: k = ", k_manual, ")") else "")) +
  theme_minimal(base_size = 9)
save_fig(p_sil, "cluster_k_silhouette", 5, 3)
print(p_sil)
cat("using k =", k, if (is.na(k_manual)) "(silhouette)" else "(manual)", "\n")

clust <- cutree(hc, k = k)

resp$cluster <- clust[rownames(resp)]
write.csv(resp, file.path(out_dir, "response_genes.csv"), row.names = FALSE)

# Calculates the average unscaled change-from-baseline for all genes in that cluster at every timepoint
cl_prof <- do.call(rbind, lapply(seq_len(k), function(i)
  data.frame(cluster = i, time = tps,
             mean_delta = colMeans(d_resp[clust == i, , drop = FALSE]),
             n = sum(clust == i))))
write.csv(cl_prof, file.path(out_dir, "cluster_profiles.csv"), row.names = FALSE)
table(clust)
```

## Heatmap: all response genes x all libraries

Rows are z-scored VST, grouped by module (dendrogram order within each), columns grouped
by timepoint

```{r cluster-heatmap, fig.width=7, fig.height=7}
row_ord <- rownames(scaled)[hc$order]
row_ord <- row_ord[order(clust[row_ord])]   # stable sort: dendrogram order kept within module
z <- t(scale(t(vmat[row_ord, samples$sample])))
z <- pmax(pmin(z, 2.5), -2.5)               # clip so a few extreme genes don't set the scale

ann_col <- data.frame(time = samples$time_f, row.names = samples$sample)
ann_row <- data.frame(cluster = factor(clust[row_ord]), row.names = row_ord)

ph <- pheatmap(z, cluster_rows = FALSE, cluster_cols = FALSE,
               show_rownames = FALSE, show_colnames = FALSE,
               annotation_col = ann_col, annotation_row = ann_row,
               gaps_col = seq(3, 15, by = 3),
               gaps_row = cumsum(table(clust[row_ord]))[-k],
               color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
               main = paste(nrow(z), "response genes (VST z-score per gene)"))
png(file.path(fig_dir, "response_gene_heatmap.png"),
    width = 7, height = 7, units = "in", res = 300)
grid::grid.newpage(); grid::grid.draw(ph$gtable); dev.off()
```

## Module profiles with exemplar genes

```{r cluster-profiles, fig.width=10, fig.height=3.4}
d_long <- data.frame(
  gene_id = rep(rownames(d_resp), times = ncol(d_resp)),
  time    = rep(tps, each = nrow(d_resp)),
  delta   = as.vector(d_resp),
  cluster = rep(clust[rownames(d_resp)], times = ncol(d_resp)))

n_lab  <- 10
cl_lab <- setNames(paste0("cluster ", seq_len(k), " (n=", table(clust), ")"), seq_len(k))

module_plot <- function(ex, subtitle) {
  ex_end <- data.frame(gene_name = ex$gene_name, cluster = ex$cluster,
                       time = 60, delta = d_resp[ex$gene_id, "60"])
  ggplot(d_long, aes(time, delta, group = gene_id)) +
    geom_hline(yintercept = 0, linetype = 3, linewidth = 0.3) +
    geom_line(colour = "grey70", alpha = 0.2, linewidth = 0.25) +
    geom_line(data = cl_prof, aes(time, mean_delta, colour = factor(cluster), group = cluster),
              inherit.aes = FALSE, linewidth = 1) +
    geom_line(data = d_long[d_long$gene_id %in% ex$gene_id, ],
              aes(colour = factor(cluster)), linewidth = 0.35) +
    geom_text_repel(data = ex_end, aes(time, delta, label = gene_name),
                    inherit.aes = FALSE, size = 2.2, nudge_x = 6, direction = "y",
                    hjust = 0, segment.size = 0.2, box.padding = 0.1,
                    xlim = c(62, 95), max.overlaps = Inf) +
    facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = cl_lab)) +
    scale_x_continuous(breaks = c(0, 30, 60)) +
    coord_cartesian(xlim = c(0, 92), clip = "off") +
    labs(x = "Time after LPS (min)", y = "delta vs 0 min (VST)",
         title = "Temporal response modules", subtitle = subtitle) +
    theme_minimal(base_size = 9) +
    theme(legend.position = "none",
          plot.margin = margin(5.5, 45, 5.5, 5.5))  # room for end-of-line labels
}

ex_padj <- do.call(rbind, lapply(split(resp, resp$cluster), function(d)
  head(d[order(d$padj), ], n_lab)))
p_mod <- module_plot(ex_padj,
  paste0("grey: individual genes | colour: module mean and top ",
         n_lab, " exemplars by LRT padj"))
save_fig(p_mod, "temporal_modules", 10, 3.4)
p_mod
```

## Pathway enrichment per cluster

pathway enrichment separately for each temporal gene-expression cluster

- Are genes from this pathway represented more often in this cluster than expected by chance, given the background universe?

```{r cluster-enrichment}
suppressPackageStartupMessages({ library(fgsea); library(msigdbr) })

gs <- rbind(msigdbr(species = "Homo sapiens", collection = "H"),
            msigdbr(species = "Homo sapiens", collection = "C2",
                    subcollection = "CP:REACTOME"))
pathways <- split(gs$gene_symbol, gs$gs_name)

# use all genes in LRT to test 
universe <- unique(na.omit(lrt_tab$gene_name))

# enrichment analysis in each cluster
enr <- do.call(rbind, lapply(sort(unique(resp$cluster)), function(i) {
  g  <- unique(na.omit(resp$gene_name[resp$cluster == i]))
  
  # over representation analysis
  # Are genes from this pathway represented more often in this cluster than expected by chance, given the background universe?
  fr <- as.data.frame(fora(pathways, genes = g, universe = universe))
  fr$cluster <- i
  fr$genes_in_cluster <- length(g)
  fr
}))
enr$overlapGenes <- vapply(enr$overlapGenes, paste, character(1), collapse = ";")
enr <- enr[order(enr$cluster, enr$padj), ]
write.csv(enr, file.path(out_dir, "cluster_pathway_enrichment.csv"), row.names = FALSE)

enr_sig <- enr[enr$padj < fdr_cut, ]
cat("significant pathways (padj <", fdr_cut, ") per cluster:\n")
print(table(enr_sig$cluster))
# top 10 per cluster
do.call(rbind, lapply(split(enr_sig, enr_sig$cluster), function(d)
  head(d[, c("cluster", "pathway", "overlap", "size", "padj")], 10)))
```

```{r cluster-enrichment-plot, fig.width=11, fig.height=6.5}
top_enr <- do.call(rbind, lapply(split(enr_sig, enr_sig$cluster), function(d)
  head(d[order(d$padj), ], 10)))
top_enr$pathway <- gsub("^HALLMARK_|^REACTOME_", "", top_enr$pathway)
top_enr$pathway <- gsub("_", " ", tolower(top_enr$pathway))
# same pathway can appear in several clusters; order bars by significance within facet
top_enr$path_f <- factor(paste(top_enr$cluster, top_enr$pathway),
                         levels = rev(paste(top_enr$cluster, top_enr$pathway)))

# strip the cluster prefix, then wrap long pathway names onto multiple lines
wrap_lab <- function(x) vapply(sub("^\\S+ ", "", x), function(s)
  paste(strwrap(s, width = 32), collapse = "\n"), character(1), USE.NAMES = FALSE)

p_enr <- ggplot(top_enr, aes(-log10(padj), path_f, fill = overlap / size)) +
  geom_col() +
  geom_vline(xintercept = -log10(fdr_cut), linetype = 3, linewidth = 0.3) +
  scale_y_discrete(labels = wrap_lab) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_viridis_c(name = "gene ratio\n(overlap/set size)") +
  facet_wrap(~ cluster, scales = "free_y", ncol = 2,
             labeller = labeller(cluster = cl_lab)) +
  labs(x = "-log10 adjusted p", y = NULL,
       title = "Top enriched pathways per temporal cluster",
       subtitle = "Hallmark + Reactome, Fisher ORA vs all LRT-tested genes") +
  theme_minimal(base_size = 8.5) +
  theme(axis.text.y = element_text(size = 6.5, lineheight = 0.85))
save_fig(p_enr, "cluster_pathway_enrichment", 11, 6.5)
p_enr
```

temporal expression profile for top 10 genes ranked by LFC

```{r cluster-profiles-lfc, fig.width=10, fig.height=3.4}
ex_lfc <- do.call(rbind, lapply(split(resp, resp$cluster), function(d)
  head(d[order(-d$max_abs_lfc), ], n_lab)))
p_mod_lfc <- module_plot(ex_lfc,
  paste0("grey: individual genes | colour: module mean and top ",
         n_lab, " exemplars by |log2FC|"))
save_fig(p_mod_lfc, "temporal_modules_lfc_labels", 10, 3.4)
p_mod_lfc
```

```{r cluster-top-table}
top_by_cluster <- do.call(rbind, lapply(split(resp, resp$cluster), function(d)
  head(d[order(d$padj), c("gene_name", "cluster", "response_tier", "direction",
                          "peak_timepoint_min", "max_abs_lfc", "padj")], 10)))
write.csv(top_by_cluster, file.path(out_dir, "cluster_top_genes.csv"), row.names = FALSE)
top_by_cluster
```

## Top-gene trajectories

10 most statistically significant response genes by LRT LFC

```{r gene-trajectories, fig.width=9, fig.height=4.5}
top_ids <- rownames(resp)[order(resp$max_abs_lfc, decreasing = TRUE)][1:10]

traj <- do.call(rbind, lapply(top_ids, function(g)
  data.frame(
    gene_name = resp[g, "gene_name"],
    sample = samples$sample,
    time = samples$time,
    rep = samples$rep,
    vst = vmat[g, samples$sample]
  )))

traj$gene_name <- factor(traj$gene_name, levels = unique(traj$gene_name))

p_traj <- ggplot(traj, aes(time, vst, colour = rep)) +
  geom_line(linewidth = 0.4) +
  geom_point(size = 1) +
  facet_wrap(~ gene_name, nrow = 2, scales = "free_y") +
  scale_x_continuous(breaks = c(0, 15, 30, 60)) +
  labs(
    x = "Time after LPS (min)",
    y = "VST expression",
    colour = "replicate",
    title = "Top 10 response genes by maximum |log2FC|"
  ) +
  theme_minimal(base_size = 9)

save_fig(p_traj, "top_gene_trajectories_by_lfc", 9, 4.5)
p_traj
```

# Early window only: 0-15 min

The whole LRT is dominated by the 60-min response

**does anything change between 0 and 15 min?**


```{r early-setup}
early_tps <- c(0, 5, 10, 15)
se <- samples[samples$time %in% early_tps, ]
se$time_f <- factor(se$time, levels = early_tps)
se$rep    <- factor(se$replicate)
se$time15 <- se$time / 15            # time rescaled to 0..1 for the smooth-trend spline

ce     <- counts[, se$sample]
keep_e <- filterByExpr(DGEList(ce, samples = se), design = model.matrix(~ rep + time_f, se))
ce_f   <- ce[keep_e, ]
cat("early window:", ncol(ce_f), "libraries |", nrow(ce_f), "of", nrow(ce), "genes kept\n")
```

## Factor LRT across 0, 5, 10, 15 min

```{r early-lrt}
dds_e <- DESeqDataSetFromMatrix(ce_f, colData = se, design = ~ rep + time_f)
dds_e <- DESeq(dds_e, test = "LRT", reduced = ~ rep, quiet = TRUE)
res_e <- results(dds_e)

lfc_e <- sapply(early_tps[-1], function(t)
  results(dds_e, name = paste0("time_f_", t, "_vs_0"))$log2FoldChange)
colnames(lfc_e) <- paste0("lfc_", early_tps[-1], "_vs_0")

early_tab <- data.frame(
  gene_id   = rownames(res_e),
  gene_name = ann[rownames(res_e), "gene_name"],
  gene_type = ann[rownames(res_e), "gene_type"],
  baseMean  = round(res_e$baseMean, 2),
  pvalue = res_e$pvalue, padj = res_e$padj, round(lfc_e, 3),
  max_abs_lfc = round(apply(abs(lfc_e), 1, max), 3),
  row.names = rownames(res_e), stringsAsFactors = FALSE)
early_tab <- early_tab[order(early_tab$padj), ]
write.csv(early_tab, file.path(out_dir, "early_0_15_lrt.csv"), row.names = FALSE)

cat("factor LRT, FDR <", fdr_cut, ":", sum(early_tab$padj < fdr_cut, na.rm = TRUE), "genes\n")
head(early_tab[, c("gene_name", "baseMean", "lfc_5_vs_0", "lfc_10_vs_0",
                   "lfc_15_vs_0", "padj")], 15)
```

## Smooth-trend LRT

models expression as a nonlinear smooth trajectory of continuous time from 0 to 15 min,
using a natural spline with 2 degrees of freedom

```{r early-trend}
dds_l <- DESeqDataSetFromMatrix(ce_f, colData = se,
                                design = ~ rep + ns(time15, df = 2))
dds_l <- DESeq(dds_l, test = "LRT", reduced = ~ rep, quiet = TRUE)
res_l <- results(dds_l)

# spline coefficients have no single interpretable slope, so report the factor
# model's 15-vs-0 log2FC alongside the smooth-trend test's p-values
trend_tab <- data.frame(
  gene_id   = rownames(res_l),
  gene_name = ann[rownames(res_l), "gene_name"],
  baseMean  = round(res_l$baseMean, 2),
  lfc_15_vs_0 = early_tab[rownames(res_l), "lfc_15_vs_0"],
  pvalue = res_l$pvalue, padj = res_l$padj,
  row.names = rownames(res_l), stringsAsFactors = FALSE)
trend_tab <- trend_tab[order(trend_tab$padj), ]
write.csv(trend_tab, file.path(out_dir, "early_0_15_smooth_trend.csv"), row.names = FALSE)

cat("smooth-trend LRT, FDR <", fdr_cut, ":",
    sum(trend_tab$padj < fdr_cut, na.rm = TRUE), "genes\n")
head(trend_tab[, c("gene_name", "baseMean", "lfc_15_vs_0", "padj")], 15)
```

## Early responders: union of the two tests

```{r early-union}
sig_f <- rownames(early_tab)[which(early_tab$padj  < fdr_cut)]
sig_l <- rownames(trend_tab)[which(trend_tab$padj  < fdr_cut)]
early_ids <- union(sig_f, sig_l)

early_hits <- data.frame(
  early_tab[early_ids, c("gene_name", "gene_type", "baseMean",
                         "lfc_5_vs_0", "lfc_10_vs_0", "lfc_15_vs_0")],
  padj_factor = signif(early_tab[early_ids, "padj"], 3),
  padj_trend  = signif(trend_tab[early_ids, "padj"], 3),
  padj_full_course = signif(lrt_tab[early_ids, "padj"], 3),
  row.names = early_ids)
early_hits <- early_hits[order(pmin(early_hits$padj_factor, early_hits$padj_trend)), ]
write.csv(early_hits, file.path(out_dir, "early_response_genes.csv"), row.names = FALSE)

cat("factor:", length(sig_f), "| trend:", length(sig_l),
    "| union:", length(early_ids), "| both:", length(intersect(sig_f, sig_l)), "\n")
early_hits
```

```{r early-volcano, fig.width=5.5, fig.height=4.5}
ev <- early_tab
ev$class <- ifelse(!is.na(ev$padj) & ev$padj < fdr_cut, "FDR < 0.05", "NS")
p_ev <- ggplot(ev, aes(lfc_15_vs_0, -log10(padj), colour = class)) +
  geom_point(size = 0.5, alpha = 0.5, na.rm = TRUE) +
  geom_hline(yintercept = -log10(fdr_cut), linetype = 3, linewidth = 0.3) +
  geom_text_repel(data = head(ev[order(ev$padj), ], 10), aes(label = gene_name),
                  colour = "black", size = 2.4, max.overlaps = Inf,
                  segment.size = 0.2, box.padding = 0.2) +
  scale_colour_manual(name = NULL, values = c(`FDR < 0.05` = "firebrick", NS = "grey75")) +
  guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "log2 fold change, 15 min vs 0 min", y = "-log10 adjusted p (early LRT)",
       title = "Early window (0-15 min) LRT") +
  theme_minimal(base_size = 9)
save_fig(p_ev, "early_volcano_0_15", 5.5, 4.5)
p_ev
```

## Transcriptional profile of inflammatory response genes at 0-15min

For knwon early response genes, extract their LFC at each time point and identify if the gene is sigificant by LRT

- several canonical targets are already climbing by 15 min at a  fold change but are not called

```{r early-canonical, fig.width=6.5, fig.height=4.5}
canon <- c("TNF","NFKBIA","NFKBIZ","CXCL8","IL1B","CCL4","CCL3","CXCL1","CXCL2","CXCL3",
           "EGR1","FOS","FOSB","JUNB","TNFAIP3","IER3","DUSP1","DUSP2","PTGS2","SOD2",
           "NR4A1","CSF2","CCL20","RGS1")
cn_ids <- rownames(early_tab)[match(canon, early_tab$gene_name)]
cn_ids <- cn_ids[!is.na(cn_ids)]

cn <- do.call(rbind, lapply(early_tps[-1], function(t)
  data.frame(gene_name = early_tab[cn_ids, "gene_name"], time = t,
             lfc = early_tab[cn_ids, paste0("lfc_", t, "_vs_0")],
             padj = early_tab[cn_ids, "padj"])))
cn$gene_name <- factor(cn$gene_name,
                       levels = early_tab[cn_ids, "gene_name"][
                         order(early_tab[cn_ids, "lfc_15_vs_0"])])
cn$called <- ifelse(!is.na(cn$padj) & cn$padj < fdr_cut, "FDR < 0.05", "not called")

p_cn <- ggplot(cn, aes(factor(time), gene_name, fill = lfc)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.1f", lfc)), size = 2.1) +
  geom_point(data = cn[cn$called == "FDR < 0.05" & cn$time == 15, ],
             aes(x = 3.65), shape = 8, size = 1.4, inherit.aes = TRUE, show.legend = FALSE) +
  scale_fill_gradient2(name = "log2FC", low = "navy", mid = "white", high = "firebrick3") +
  coord_cartesian(xlim = c(1, 3.8), clip = "off") +
  labs(x = "Time after LPS (min)", y = NULL,
       title = "Canonical LPS targets at 0-15 min ",
       subtitle = "log2FC vs 0 min | * = called at FDR < 0.05 by the early LRT") +
  theme_minimal(base_size = 9)
save_fig(p_cn, "early_canonical_lfc", 6.5, 4.5)
p_cn
```

```{r early-trajectories, fig.width=9, fig.height=4.5}
show_ids <- unique(c(early_ids, rownames(early_tab)[match(
  c("TNF", "NFKBIA", "DUSP1", "IER3"), early_tab$gene_name)]))
show_ids <- show_ids[!is.na(show_ids)]
etraj <- do.call(rbind, lapply(show_ids, function(g)
  data.frame(gene_name = early_tab[g, "gene_name"], time = se$time, rep = se$rep,
             vst = vmat[g, se$sample])))
etraj$gene_name <- factor(etraj$gene_name, levels = unique(etraj$gene_name))

p_etraj <- ggplot(etraj, aes(time, vst, colour = rep)) +
  geom_line(linewidth = 0.4) + geom_point(size = 1) +
  facet_wrap(~ gene_name, nrow = 2, scales = "free_y") +
  scale_x_continuous(breaks = early_tps) +
  labs(x = "Time after LPS (min)", y = "VST expression", colour = "replicate",
       title = "Early responders, per replicate (0-15 min)") +
  theme_minimal(base_size = 9)
save_fig(p_etraj, "early_trajectories", 9, 4.5)
p_etraj
```




```{r session}
sessionInfo()
```
