---
title: "Tutorial: extract m6A and footprints at an LCL promoter (ALAS1)"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This example uses the ALAS1 promoter, one of the two LCL promoters (PRDX1, ALAS1)
in `code/clustering_methods/Leiden_Manhattan/leiden_LCL.Rmd`, and only functions
from `code/parsing_functions/`:

1. build the read x position m6A matrix with `assemble_region_m6a()`
2. plot it as a heatmap without clustering
3. plot the aggregated accessibility profile (m6A fraction of reads per position)
4. plot the nucleosome, TF footprint and FIRE element tracks of the same reads

Reads from all 31 LCL samples are pooled. Only reads spanning the whole window are kept. There is no phasing (the HP tag is never read) and no clustering: reads are ordered by sample, then by read start


# Setup

```{r functions}
suppressPackageStartupMessages({
  library(dplyr); library(GenomicRanges); library(Rsamtools); library(Matrix)
})
code_dir <- "/project/spott/cshan/fiber-seq/code/parsing_functions"
source(file.path(code_dir, "parsing_footprints_functions.r"))
source(file.path(code_dir, "plotting_functions.r"))
```

```{r settings}
gene     <- "ALAS1"
flank_bp <- 1000L

tss_bed      <- "/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed"
sample_sheet <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv"
fire_root    <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
tf_root      <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/FiberHMM_v2/results/results_v1_trained_model"
tabix_bin    <- "/project/spott/cshan/envs/dimelo/bin/tabix"

nuc_size         <- c(130, 160)  # nucleosome size window (bp), the LCL setting of leiden_LCL.Rmd
min_tf_score     <- 50           # FiberHMM v2 per-footprint score floor
pileup_window_bp <- 10           # sliding window of the smoothed m6A fraction

out_dir <- file.path("/project/spott/cshan/fiber-seq/LCL_project/extract_m6a_tutorial", gene)
```

# 1. Region

The canonical TSS BED stores the TSS as a 1 bp interval `[TSS0, TSS0 + 1)`
(0-based). As in `leiden_LCL.Rmd`, the window is the 2,000 bp
`[TSS0 - 1000, TSS0 + 1000)`: BED `start` / `end`, or 1-based
`analysis_start` / `analysis_end`. The 1-based TSS is `TSS0 + 1`.

ALAS1 is on the + strand, so upstream is on the left. To use PRDX1 instead, set
`gene <- "PRDX1"`; it is on the - strand, so its upstream side is on the right.

```{r region}
tss <- data.table::fread(tss_bed, header = FALSE, data.table = FALSE,
  col.names = c("chr", "start", "end", "annotation", "score", "strand"))
tss <- tss[vapply(strsplit(tss$annotation, ";", fixed = TRUE), `[`, character(1), 3L) == gene, ]
stopifnot(nrow(tss) == 1L, tss$end - tss$start == 1L)

region <- data.frame(region_id = gene, chr = tss$chr,
  start = tss$start - flank_bp, end = tss$start + flank_bp,                # BED, 0-based start
  analysis_start = tss$start - flank_bp + 1L, analysis_end = tss$start + flank_bp,
  tss = tss$start + 1L, strand = tss$strand)
region
```

# 2. Extract the m6A matrix

`assemble_region_m6a()` queries each sample's `ft extract` m6A BED12 with tabix
(`<fire_dir>/extracted_results/m6a_by_chr/<sample>.ft_extracted_m6a.<chr>.bed.gz`),
keeps the longest alignment of each read and drops the sentinel blocks.

- `full_span = TRUE` keeps only reads that cover the whole window, so every entry is
  1 (m6A call) or 0 (no call).
- `positions` = every base of the window, so the matrix has one column per bp.

Read IDs are `<sample>::<read name>`.

```{r extract-m6a}
samples <- read.csv(sample_sheet, stringsAsFactors = FALSE)[, c("sample_name", "fire_dir")]

dat <- assemble_region_m6a(samples, region, full_span = TRUE,
  positions = seq(region$analysis_start, region$analysis_end))
dim(dat$met_mat)
head(dat$rids_df)
dat$qc   # overlapping and full-span reads per sample
```

Read order for all figures: sample sheet order, then read start. No clustering.

```{r read-order}
rids <- dat$rids_df
rids <- rids[order(match(rids$sample_name, samples$sample_name), rids$start, rids$RID), ]
read_order <- rids$RID
```

The same calls as one row per m6A call, the input of `pileup_reads()` and
`plot_region_panels()`:

```{r m6a-calls}
calls <- Matrix::summary(dat$met_mat)
m6a_calls <- data.frame(RID = rownames(dat$met_mat)[calls$i],
  pos = as.integer(colnames(dat$met_mat))[calls$j], base = "A")
```

# 3. m6A heatmap (no clustering)

Rows are reads in `read_order`, columns are bases; black = m6A call. The dashed
line marks the TSS.

```{r heatmap, fig.width=10, fig.height=10}
m6a_mat <- as.matrix(dat$met_mat[read_order, , drop = FALSE])
positions <- as.integer(colnames(m6a_mat))
tss_col <- match(region$tss, positions)
ticks <- unique(round(seq(1, ncol(m6a_mat), length.out = 5L)))

sample_label <- factor(sub("_.*$", "", rids$sample_name),
  levels = sub("_.*$", "", samples$sample_name))
sample_colors <- setNames(grDevices::hcl.colors(nlevels(sample_label), "Dynamic"),
  levels(sample_label))

heatmap <- ComplexHeatmap::Heatmap(m6a_mat, name = "m6A",
  col = c("0" = "white", "1" = "black"),
  heatmap_legend_param = list(at = c(0, 1), labels = c("no m6A call", "m6A")),
  cluster_rows = FALSE, cluster_columns = FALSE,
  show_row_names = FALSE, show_column_names = FALSE,
  use_raster = TRUE, raster_quality = 2,
  left_annotation = ComplexHeatmap::rowAnnotation(sample = sample_label,
    col = list(sample = sample_colors),
    annotation_legend_param = list(sample = list(ncol = 2))),
  bottom_annotation = ComplexHeatmap::HeatmapAnnotation(
    coordinate = ComplexHeatmap::anno_mark(at = ticks,
      labels = format(positions[ticks], big.mark = ","), which = "column",
      side = "bottom", labels_gp = grid::gpar(fontsize = 8)),
    TSS = ComplexHeatmap::anno_mark(at = tss_col, labels = "TSS", which = "column",
      side = "bottom", labels_gp = grid::gpar(fontsize = 8, col = "#D55E00")),
    annotation_name_gp = grid::gpar(fontsize = 8)),
  column_title = sprintf("%s promoter, %s:%d-%d\n%d full-span reads from %d LCL samples, ordered by sample then read start",
    gene, region$chr, region$analysis_start, region$analysis_end,
    nrow(m6a_mat), length(unique(rids$sample_name))),
  column_title_gp = grid::gpar(fontsize = 11))

# the TSS line is drawn on the heatmap body after the heatmap itself
draw_heatmap <- function(x) {
  ComplexHeatmap::draw(x, merge_legend = TRUE)
  ComplexHeatmap::decorate_heatmap_body("m6A", {
    grid::grid.lines(x = rep((tss_col - 0.5) / ncol(m6a_mat), 2), y = c(0, 1),
      gp = grid::gpar(col = "#D55E00", lty = "dashed", lwd = 1.2))
  })
}
save_figure(heatmap, file.path(out_dir, "m6a_heatmap.pdf"), width = 10, height = 10,
  draw = draw_heatmap)
draw_heatmap(heatmap)
```

# 4. Aggregated accessibility profile

`pileup_reads()` gives, at every position with at least one m6A call, the fraction
of reads with a call there (`frac`), and its `pileup_window_bp` sliding-window mean
(`smooth_frac`). All reads span the window, so the denominator is every read.

```{r profile, fig.width=9, fig.height=3.5}
pileup <- pileup_reads(m6a_calls, rids, region$analysis_start, region$analysis_end,
  smooth = TRUE, window_n = pileup_window_bp)

profile_plot <- plot_group_profile(pileup, x_col = "pos", y_col = "smooth_frac",
  layers = list(
    list(geom = "line", mapping = ggplot2::aes(y = frac), colour = "grey75", linewidth = 0.3),
    list(geom = "line", colour = "blue", linewidth = 0.6)),
  markers = list(xintercept = region$tss, linetype = "dashed", colour = "#D55E00"),
  window = c(region$analysis_start, region$analysis_end),
  scales = list(ggplot2::scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))),
  labels = list(x = paste(region$chr, "(1-based)"), y = "m6A fraction of reads",
    title = paste(gene, "promoter: aggregated accessibility"),
    subtitle = sprintf("%d full-span reads; grey = per position, blue = %d bp sliding mean; dashed = TSS",
      nrow(rids), pileup_window_bp)),
  theme = theme_fiberseq("bw", base_size = 10))
save_figure(profile_plot, file.path(out_dir, "aggregate_m6a_profile.pdf"), width = 9, height = 3.5)
profile_plot
```

# 5. Footprint tracks

For the same reads, `load_region()` reads the FIRE elements and
`load_ft_tracks()` reads:

- nucleosomes: `ft extract` `nuc_by_chr` (kept when `nuc_size[1]` to `nuc_size[2]` bp)
- TF footprints: FiberHMM v2 `<sample>.recalled_tf.bed.gz`, score >= `min_tf_score`,
  binned into `FP_SIZE_BINS` (half-open 10-30, 40-60, 60-80 bp; other sizes are
  not drawn)

These functions key reads as `"<sample> <read name>"`.

`load_ft_tracks()` also returns the m6A calls (`$m6a`); this tutorial uses
the matrix from step 2 instead.

```{r load-footprints}
keys <- paste(rids$sample_name, rids$original_RID)
tracks <- load_region(list(chrom = region$chr, start = region$start, end = region$end),
  samples$sample_name, root = NULL, fire_root = fire_root, tabix_bin = tabix_bin,
  span_source = "fire_all")
tracks <- load_ft_tracks(tracks, keys = keys, ft_root = fire_root, tabix_bin = tabix_bin,
  nuc_source = "ft", tf_source = "recalled_tf", tf_root = tf_root,
  min_tf_score = min_tf_score)

# back to the "<sample>::<read name>" IDs of the m6A matrix
nuc_label <- paste0(nuc_size[1], "-", nuc_size[2], " bp nucleosome")
fire <- as.data.frame(tracks$elements)
fire <- fire[fire$key %in% keys, ]
fire <- data.frame(RID = paste(fire$sample_name, fire$RID, sep = "::"),
  start = fire$start, end = fire$end, class = "FIRE")
nucs <- as.data.frame(tracks$nuc)
nucs <- nucs[!is.na(assign_size_class(nucs$end - nucs$start, nuc_size)), ]
nucs <- data.frame(RID = paste(nucs$sample_name, nucs$RID, sep = "::"),
  start = nucs$start, end = nucs$end, class = nuc_label)
tfs <- as.data.frame(tracks$size_fps)
tfs <- data.frame(RID = paste(tfs$sample_name, tfs$RID, sep = "::"),
  start = tfs$start, end = tfs$end, class = factor(tfs$class, levels = FP_SIZE_BINS))

c(FIRE_elements = nrow(fire), nucleosomes = nrow(nucs), TF_footprints = nrow(tfs))
table(tfs$class)
```

`plot_region_panels()` stacks the aggregated profile, the m6A reads and the
footprint reads. It draws whatever is in `res$fps_infire`. Here that is every TF
footprint, not only the ones inside a FIRE element. ggplot puts the first read
level at the bottom, so the levels are reversed to keep the heatmap's top-to-bottom
order.

```{r footprint-tracks, fig.width=10, fig.height=14}
rid_levels <- rev(read_order)
as_read_factor <- function(d) { d$RID <- factor(d$RID, levels = rid_levels); d }
res <- list(
  region     = list(chr = region$chr, start = region$analysis_start, end = region$analysis_end),
  reads      = as_read_factor(m6a_calls),
  rids_df    = as_read_factor(data.frame(RID = rids$RID, start = rids$start, end = rids$end)),
  fire       = as_read_factor(fire),
  nucs       = as_read_factor(nucs),
  fps_infire = as_read_factor(tfs),
  pileup     = pileup,
  size_levels = FP_SIZE_BINS,
  nuc_label  = nuc_label,
  group_col  = NULL)

panels <- plot_region_panels(res,
  plot_name = paste(gene, "promoter"),
  subtitle = sprintf("%s:%d-%d; %d full-span reads, ordered by sample then read start; dashed = TSS",
    region$chr, region$analysis_start, region$analysis_end, nrow(rids)),
  heights = c(pileup = 2, reads = 6, fire_fiberHMM = 6),
  read_options = list(arrows = FALSE),
  y_labels = c(pileup = "m6A fraction", reads = "m6A calls",
               fire_fiberHMM = "FIRE / nucleosomes / TF footprints"),
  x_label = paste(region$chr, "(1-based)"))
footprint_plot <- panels$combined &
  ggplot2::geom_vline(xintercept = region$tss, linetype = "dashed", colour = "#D55E00", linewidth = 0.4)
save_figure(footprint_plot, file.path(out_dir, "footprint_tracks.pdf"), width = 10, height = 14)
footprint_plot
```
