---
title: "Co-accessible cCRE pairs across the LPS timecourse"
author: "Crystal Shan"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Picks ENCODE cCRE pairs where **one partner is a CA-CTCF element** and
co-accessibility is **low at baseline (LPS_0) and rises as the LPS timecourse
progresses** — a weighted upward trend of the fiber-level odds ratio over the
four timepoints — and draws them at single-fiber resolution across all four.

The tables come from `03_coaccess_cres.py`: cCREs overlapping a FIRE
peak, paired within TSS ± 10 kb gene windows at 500 < gap < 20,000 bp, scored on
the fibers that overlap both, with accessibility requiring a single FIRE element
on the same fiber to cover at least 50% of the cCRE
and tested by `fisher.test(table + 1)`. The one structural difference from Kevin
is that **each timepoint is tested by itself** where he pools 17 LCL samples.

The loaders (`load_region()`, `label_reads()`, `load_ft_tracks()`, ...) live in
`parsing_functions/parsing_footprints_functions.r` and the figure is drawn by
`plot_coaccess_pair()` in `parsing_functions/plotting_functions.r`, the same
functions the LCL figures use. Nothing here sources his code.

# Functions

```{r functions}
suppressPackageStartupMessages({
  library(data.table); library(ggplot2); library(patchwork); library(dplyr)
})
source("/project/spott/cshan/fiber-seq/code/parsing_functions/parsing_footprints_functions.r")
source("/project/spott/cshan/fiber-seq/code/parsing_functions/plotting_functions.r", local = TRUE)
```

# Settings

```{r settings}
options(scipen = 999)

root       <- "/project/spott/cshan/fiber-seq/macrophage_project/co-accessibility"
coaccess   <- file.path(root, "coaccess")
uni_dir    <- file.path(root, "universe")
fire_root  <- "/project/spott/lizarraga/pacbio_analysis/macrophage_project/merged_hifi_bams/FIRE"
plot_dir   <- file.path(coaccess, "plots")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

# track locations for plot_coaccess_pair(): fiber spans (root), FIRE elements
# (fire_root), `ft extract` m6A (ft_root, extracted from the unfiltered BAM, so
# display only) and the FiberHMM nucleosome / size-binned footprint calls
# (hmm_root; file names drop the underscore, LPS_0 -> LPS0)
paths <- list(
  root      = root,
  fire_root = fire_root,
  ft_root   = "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir",
  hmm_root  = "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract",
  tabix_bin = "/project/spott/cshan/envs/dimelo/bin/tabix")

samples <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")
t_sel   <- "LPS_15"      # reference table for pair coords/labels/gene anchors

n_top <- 5
flank <- 1000            # bp either side of the pair for the plot window

# selection knobs for the timecourse ranking: pairs with weak/no coupling at
# baseline whose odds ratio RISES over the timecourse (one-sided weighted trend
# on the log-OR over minutes), restricted to pairs where one partner is a
# CA-CTCF cCRE. Only ~2.3K CTCF-partnered pairs are testable (vs ~139K overall)
# so the evidence and fiber floors are relaxed relative to the unrestricted
# screen; the floors still keep every facet legible.
mins            <- c(0, 5, 10, 15)  # LPS timepoints in minutes, aligned to samples
cre_types       <- "CA-CTCF"  # cCRE classes for select_cre_pairs(); NULL allows any type
cre_match       <- "either"   # "either" / "both" members in cre_types (or "pair" + cre_types2)
max_p_trend     <- 0.05     # one-sided weighted trend p (increase)
max_or_baseline <- 1.5      # OR cap at LPS_0: coupling absent/weak at baseline
min_delta_lor   <- log(2)   # logOR(late) - logOR(0): >= 2-fold OR rise
min_dist_sel    <- 200
min_shared_all  <- 15       # n_shared_reads floor in EVERY timepoint
min_co_acc_late <- 5        # co_access floor at the last timepoint

# the read rule the tables were computed under; the figure must match or the
# fibers drawn and the 2x2 will disagree
read_rule <- "any"
fire_overlap_fraction <- 0.5  # fraction of cCRE covered by one same-read FIRE interval

min_locus_sep <- 1e6     # keep the picks on different loci
```

# Candidates

Only pairs anchored to at least one **protein-coding** gene are considered, and
pair labels name the protein-coding genes only. The selection is further
restricted to pairs with a **CA-CTCF partner** (`select_cre_pairs()` with
`cre_types` / `cre_match`, matching the class in each member's `CRE_ID`
exactly; `cre_types <- NULL` allows any cCRE type) — only ~2.3K such pairs are testable across all four
timepoints, which is why the evidence and fiber floors below are looser than an
unrestricted screen would use. Pairs are selected for a
**gain of coupling from baseline**: the odds ratio at LPS_0 must be ≤
`r max_or_baseline` *and* the minimum of the series (coupling starts absent or
weak), and the Haldane log-OR must show a significant one-sided **weighted
upward trend over minutes** (weighted least squares with Woolf weights;
z = slope/SE). Survivors are ranked by the log-OR gain from LPS_0 to LPS_15 —
the visual size of the change. Because the OR conditions on each cCRE's own
marginal accessibility, this is not fooled by both cCREs independently becoming
more accessible after LPS — the *coordination* itself has to grow.

```{r select}
read_coaccess_table <- function(file, select = NULL) {
  required <- c("read_rule", "fire_overlap_fraction")
  header <- names(fread(file, nrows = 0L))
  if (!all(required %in% header)) {
    stop("Missing accessibility-rule metadata in ", file,
         ". Rerun macrophage/run_coaccess.sh with --fire-overlap-fraction 0.5.")
  }
  if (!is.null(select)) select <- unique(c(select, required))
  d <- fread(file, select = select)
  if (anyNA(d$fire_overlap_fraction) ||
      any(!is.finite(d$fire_overlap_fraction)) ||
      any(abs(d$fire_overlap_fraction - fire_overlap_fraction) > 1e-10) ||
      anyNA(d$read_rule) || any(d$read_rule != read_rule)) {
    stop("Saved counting rules disagree with plotting settings in ", file,
         ". Rerun macrophage/run_coaccess.sh with matching read and FIRE rules.")
  }
  d
}

cre_uni <- fread(file.path(uni_dir, "cre_universe.bed"),
                 col.names = c("chrom", "start", "end", "CRE_ID", "CRE_label"))

# per-timepoint 2x2 cells for every pair. Cell identities checked once:
# co_closed == co_inaccess, and the four cells sum to n_shared_reads.
tp_cells <- rbindlist(lapply(seq_along(samples), function(i) {
  d <- read_coaccess_table(file.path(coaccess, paste0(samples[i], "_coaccess_pairs.tsv.gz")),
             select = c("CRE_pair", "co_access", "CRE1_access", "CRE2_access",
                        "co_inaccess", "n_shared_reads", "or_haldane"))
  d[, `:=`(timepoint = samples[i], min_lps = mins[i])]
  d
}))

# one-sided weighted trend of the Haldane log odds ratio on minutes: weighted
# least-squares slope with Woolf weights (+0.5 per cell so empty cells stay
# finite), z = slope / SE, p_trend = P(increase | constant OR). Only pairs
# observed in all four timepoints are tested.
tp_cells[, `:=`(a = co_access + 0.5, b = CRE1_access + 0.5,
                c = CRE2_access + 0.5, d = co_inaccess + 0.5)]
tp_cells[, `:=`(lor = log(a * d / (b * c)), w = 1 / (1/a + 1/b + 1/c + 1/d))]

trend <- tp_cells[, {
  if (.N == length(samples)) {
    xb    <- sum(w * min_lps) / sum(w)
    yb    <- sum(w * lor) / sum(w)
    sxx   <- sum(w * (min_lps - xb)^2)
    slope <- sum(w * (min_lps - xb) * (lor - yb)) / sxx
    z     <- slope / sqrt(1 / sxx)
    .(z = z, p_trend = pnorm(z, lower.tail = FALSE),
      or0 = or_haldane[min_lps == 0],
      delta_logOR = lor[min_lps == max(min_lps)] - lor[min_lps == 0],
      lor_min_at0 = which.min(lor) == which(min_lps == 0),
      or_series = paste(sprintf("%.1f", or_haldane[order(min_lps)]),
                        collapse = "/"),
      min_shared = min(n_shared_reads),
      co_access_late = co_access[min_lps == max(min_lps)])
  }
}, by = CRE_pair]
trend[, fdr_trend := p.adjust(p_trend, "BH")]

# coords/labels/gene anchors from the reference timepoint's table
pairs_dt <- read_coaccess_table(file.path(coaccess, paste0(t_sel, "_coaccess_pairs.tsv.gz")))
cat(nrow(pairs_dt), "distinct cCRE pairs;", nrow(trend), "present in all four timepoints\n")

# restrict to pairs anchored to >= 1 protein-coding gene. transcript_type lives
# only in the gene-anchored stat table (the pair-level dedup drops it), so the
# protein-coding pairs are read off that table, and gene_names is relabelled to
# the protein-coding genes only so titles/filenames name the genes that kept it.
stat_dt <- read_coaccess_table(file.path(coaccess, paste0(t_sel, "_coaccess_stat.tsv.gz")))
pc_map  <- stat_dt[transcript_type == "protein_coding",
                   .(gene_names = paste(sort(unique(gene_name)), collapse = ";")),
                   by = CRE_pair]
pairs_dt <- pairs_dt[CRE_pair %in% pc_map$CRE_pair]
pairs_dt[, gene_names := pc_map[match(CRE_pair, pc_map$CRE_pair), gene_names]]
cat(nrow(pairs_dt), "pairs anchored to a protein-coding gene\n")

pairs_dt <- merge(pairs_dt, trend, by = "CRE_pair")

sig <- select_cre_pairs(pairs_dt, cre_types = cre_types, cre_match = cre_match)[
                p_trend < max_p_trend & or0 <= max_or_baseline & lor_min_at0 &
                delta_logOR >= min_delta_lor & dist > min_dist_sel &
                min_shared >= min_shared_all & co_access_late > min_co_acc_late]

data.table(tested_pairs = nrow(pairs_dt),
           gain_filter = nrow(sig),
           fdr_trend_lt_0.05 = sum(pairs_dt$fdr_trend < 0.05)) |>
  knitr::kable(caption = "co-accessibility gain from baseline across the timecourse")

# rank by the log-OR gain from baseline, then spread the picks across loci
setorder(sig, -delta_logOR)
sel <- sig[0]
for (i in seq_len(nrow(sig))) {
  r <- sig[i]
  clash <- nrow(sel) > 0 &&
    any(sel$chrom == r$chrom & abs(sel$cre1_start - r$cre1_start) < min_locus_sep)
  if (!clash) sel <- rbind(sel, r)
  if (nrow(sel) == n_top) break
}

sel[, pair_id := paste0("pair", seq_len(.N), "_", gsub("[^A-Za-z0-9]+", "-", gene_names),
                        "_", chrom, "_", cre1_start)]

sel[, .(pair_id, gene_names, CRE_pair_labels, dist,
        `OR 0/5/10/15` = or_series, min_shared,
        delta_logOR = round(delta_logOR, 2), z = round(z, 2),
        p_trend = signif(p_trend, 2), fdr_trend = signif(fdr_trend, 2))] |>
  knitr::kable(caption = paste("top", n_top,
                               "CTCF-partnered cCRE pairs gaining co-accessibility after LPS"))
```

`p_trend` is the one-sided weighted trend test on the Haldane-corrected (+0.5)
cells and `fdr_trend` is BH over every pair observed in all four timepoints —
this is a raw-p example screen, not a discovery set, with the cell floors doing
the legibility work. The per-timepoint ORs shown are `or_haldane` from the
tables. As a sanity check: a genuine rising trend implies the ORs are not
homogeneous, so the standard Breslow-Day heterogeneity test on the selected
pairs should also be small:

```{r breslow-day}
bd <- vapply(sel$CRE_pair, function(cp) {
  d <- tp_cells[CRE_pair == cp][match(samples, timepoint)]
  arr <- array(0, dim = c(2, 2, length(samples)))
  for (i in seq_along(samples))
    arr[, , i] <- matrix(c(d$co_access[i], d$CRE1_access[i],
                           d$CRE2_access[i], d$co_inaccess[i]), 2, byrow = TRUE)
  DescTools::BreslowDayTest(arr)$p.value
}, numeric(1))
data.table(pair_id = sel$pair_id, p_trend = signif(sel$p_trend, 2),
           breslow_day_p = signif(bd, 2)) |> knitr::kable()
```

# Read-level figures

One facet per timepoint. Fibers are sorted within each facet by their
configuration at the pair — both accessible, CRE1 only, CRE2 only, neither —
which is the visual counterpart of the 2×2 and what Kevin's
`cluster = "fire_configs"` achieves. The tested pair is shaded on every panel.

Four panels: the **m6A fraction** (fraction of covering fibers methylated at
each A position, computed the way the topic model's empirical profile is), the
cCRE track, the **m6A** raster (the raw footprinting signal, one mark per
methylated adenine), and the **FiberHMM footprints** panel with a class legend
in Kevin's `fire_fiberHMM` colour scheme — **linker** backbone, FiberHMM
**nucleosome** calls (`FiberHMM/extract/firehmm_footprint`), **FIRE elements**,
and the **size-binned footprints** (`firehmm_tf/ft_by_size`, size10-30 /
size40-60 / size60-80) on top, smallest last. m6A comes from the existing
`ft extract` run under `FiberHMM/extract/ft_result_dir`; the FIRE elements,
which are what the 2×2 is actually built from, come from `lizarraga_FIRE`.

```{r plot-fn}
# plot_coaccess_pair() with the macrophage settings: read spans, FiberHMM
# nucleosomes and size-binned footprints, one facet per timepoint. The window is
# the pair +/- flank and the subtitle starts with its coordinates.
plot_one <- function(r, width = 11, height = 16) {
  out <- plot_coaccess_pair(r, samples, paths,
    span_source = "read_spans", nuc_source = "fiberhmm", tf_source = "by_size",
    facet_by = "sample", colors = timepoint_palette(samples), cres = cre_uni,
    flank = flank, read_rule = read_rule, fire_overlap_fraction = fire_overlap_fraction,
    plot_name = paste0(r$gene_names, "  (", r$CRE_pair_labels, ")"),
    subtitle = sprintf("gap %d bp  |  OR(0/5/10/15)=%s  |  trend p=%.2g",
                       r$dist, r$or_series, r$p_trend),
    show_2x2 = FALSE,
    heights = c(pileup = 4, peaks = 1.2, reads = 11, fire_fiberHMM = 13),
    bar_title = paste0(r$gene_names, ": configuration by timepoint"))

  save_figure(out$panels$combined, file.path(plot_dir, paste0(r$pair_id, "_reads.pdf")),
         width = width, height = height, limitsize = FALSE)
  save_figure(out$bars, file.path(plot_dir, paste0(r$pair_id, "_config_bars.pdf")),
         width = 6, height = 4)

  out
}
```

```{r plot-all, fig.width=11, fig.height=16, results="asis"}
out <- list()
for (i in seq_len(nrow(sel))) {
  r <- sel[i]
  cat("\n\n## ", r$gene_names, "  (", r$pair_id, ")\n\n", sep = "")
  out[[r$pair_id]] <- tryCatch(plot_one(r), error = function(e) {
    cat("skipped:", conditionMessage(e), "\n"); NULL })
  if (!is.null(out[[r$pair_id]])) {
    print(out[[r$pair_id]]$panels$combined)
    print(out[[r$pair_id]]$bars)
  }
}
```

# Configuration counts

Read straight off the fibers. These must reproduce the
`co_closed / CRE1_access / CRE2_access / co_access` columns above — the figure and
the table are computed from the same two BEDs under the same rules, so any
disagreement means the plot window or the read rule has drifted.

```{r qc}
if (length(out) > 0 && !all(vapply(out, is.null, logical(1)))) {
  qc <- rbindlist(lapply(names(out), function(nm) {
    if (is.null(out[[nm]])) return(NULL)
    d <- out[[nm]]$labels[!is.na(config)]
    d[, .(n = .N), by = .(sample_name, config)][, pair_id := nm]
  }))
  dcast(qc, pair_id + sample_name ~ config, value.var = "n", fill = 0) |>
    knitr::kable()
} else {
  cat("No figures produced.\n")
}
```
