---
title: "CXCL2 Fiber-seq: TSCAN pseudotime analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
params:
  project_root: "/project/spott/cshan/fiber-seq"
  output_dir: null
  figure_dir: null
  graph_dimensions: 5
---

This notebook contains the complete analysis and all figures, with a separate
named chunk for each stage. Analysis objects and statistics stay in memory;
only PDF figures are exported. The figure chunks use the existing
in-memory results without rerunning TSCAN. The original `PAGA_DT.ipynb` remains
unchanged. There is no DPT comparison or Python execution in this notebook.

## Method and source data

This is a graph-embedding variant of
[Battaglia et al.'s modified `exprmclust2()` workflow](https://github.com/BernsteinLab/targetedNOMe_2022/blob/master/pseudotime/tcell.r).
PCA is replaced by a **weighted Laplacian eigenmap of the existing Manhattan kNN
graph**. The downstream machinery remains the same: `mclust` with
`modelNames = "EVV"`, `G = 2:9`, seed 1234, empirical cluster centers, their
Euclidean minimum-spanning tree, and `TSCANorder()`.

The embedding uses the saved `weight` edge attribute, which is an affinity
(higher means more similar), exactly as exported from the Manhattan graph.
`manhattan_distance` is retained and checked but is not treated as an affinity.
No neighbors, edge weights, site matrix, or read set are rebuilt. All 215 vertices
and every original edge are retained. The binary matrix remains unchanged for
validation, heatmaps, and aggregate m6A profiles.

Five graph dimensions are fixed in advance through `params$graph_dimensions`,
matching the dimension count in the previous PCA analysis. This is an explicit
representation choice, not a dimension count selected using time or Leiden
labels. No PCA, feature scaling, or coordinate standardization follows the graph
embedding. Spectral eigenvalues are not reported as explained m6A variance.

Inspection of `PAGA_DT.ipynb` and the repository's Leiden export code identified
`CXCL2/bin0/k5/tables`: **215 reads by 827 binary m6A sites**. Here `k5` denotes the
number of previous Leiden clusters; the saved neighbor parameter is 10. We load
the existing feature and assignment TSVs directly, preserve RID and site order,
and exclude only `RID`, `sample_name`, `timepoint`, and `cluster` from the matrix.
No reads or sites are filtered, rebuilt, imputed, binned, or substituted.

Experimental minutes and previous Leiden labels are attached only after inference
for evaluation. They do not enter graph embedding, mixture fitting, MST construction, root
selection, or orientation. The inferred direction is arbitrary and is never
reversed to improve its correlation with experimental time.

## Setup

Run the chunks interactively, or run the following from the repository root to
execute the notebook and save **only figures**, without creating an HTML report
or intermediate data files:

```r
local({
  params <- list(project_root = normalizePath("."), output_dir = NULL,
                 figure_dir = NULL, graph_dimensions = 5L)
  code <- knitr::purl(text = readLines("code/pseudotime/TSCAN_CXCL2_pseudotime.Rmd"),
                      output = NULL, documentation = 0, quiet = TRUE)
  grDevices::pdf(NULL, width = 16, height = 10)  # No automatic Rplots.pdf.
  on.exit(grDevices::dev.off(), add = TRUE)
  eval(parse(text = code), envir = environment())
})
```

Required R packages are `mclust`, `igraph`, Bioconductor `TSCAN`, `ggplot2`,
and `knitr`. Dependencies are checked, never installed automatically. Figures
default to `macrophage_project/pseudotime/TSCAN_CXCL2/figures` under the project
root. An explicit `rmarkdown::render()` call additionally creates an HTML report;
the figure-only command above does not.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                     error = FALSE, cache = FALSE, dev = "pdf")
options(width = 110)
```

### Paths and dependencies

```{r paths}
project_root <- normalizePath(params$project_root, mustWork = TRUE)
output_dir <- params$output_dir
if (is.null(output_dir)) {
  output_dir <- file.path(project_root, "macrophage_project/pseudotime/TSCAN_CXCL2")
}
n_bins <- 10L
```

```{r dependencies}
required <- c("mclust", "igraph", "TSCAN", "ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing R packages: ", paste(missing, collapse = ", "),
       ". Install the missing CRAN packages or Bioconductor TSCAN before running")
}
```

## Load the unchanged CXCL2 exports

Numeric genomic column names are preserved with `check.names = FALSE`. Metadata
is aligned by RID and cross-checked against the feature table. The annotation
`feature` field is a site index; its genomic coordinates are checked separately.

```{r input-loader}
read_export <- function(path) {
  if (!file.exists(path)) stop("Missing notebook input: ", path)
  read.delim(path, check.names = FALSE, stringsAsFactors = FALSE,
             quote = "", comment.char = "")
}

load_cxcl2 <- function(project_root) {
  # PAGA_DT.ipynb discovers CXCL2/bin0/k*/tables. Inspection found only k5.
  # The exporter names k<n_clusters>; k5 is NOT the neighbor parameter (10).
  base <- file.path(project_root, "macrophage_project/clustering_TSS_methods",
                    "Leiden_Manhattan/TSS_1000/CXCL2/bin0")
  candidates <- Sys.glob(file.path(base, "k*", "tables"))
  table_dir <- file.path(base, "k5", "tables")
  if (length(candidates) != 1L || !identical(candidates, table_dir)) {
    stop("Expected the notebook's sole CXCL2/bin0/k5/tables directory; found: ",
         paste(candidates, collapse = ", "))
  }
  input_paths <- c(
    features = file.path(table_dir, "feature_matrix_CXCL2.tsv"),
    knn_edges = file.path(table_dir, "knn_edges_CXCL2.tsv"),
    assignments = file.path(table_dir, "read_cluster_assignments_CXCL2.tsv"),
    parameters = file.path(table_dir, "run_parameters_CXCL2.tsv"),
    window_annotation = file.path(table_dir, "window_annotation_CXCL2.tsv")
  )
  tables <- lapply(input_paths, read_export)
  features <- tables$features
  observations <- tables$assignments
  required <- c("RID", "sample_name", "timepoint", "cluster")
  stopifnot(all(required %in% names(features)),
            all(c(required, "chr", "start", "end", "strand") %in% names(observations)))
  if (anyNA(features$RID) || anyNA(observations$RID) ||
      any(!nzchar(features$RID)) || any(!nzchar(observations$RID)) ||
      anyDuplicated(features$RID) || anyDuplicated(observations$RID) ||
      !setequal(features$RID, observations$RID)) {
    stop("Feature and assignment tables must have the same unique, nonmissing RIDs.")
  }
  # Keep the exported matrix row order exactly. Align only metadata by RID.
  original_assignment_order <- observations$RID
  observations <- observations[match(features$RID, observations$RID), , drop = FALSE]
  rownames(observations) <- observations$RID
  for (field in required) {
    if (!identical(as.character(features[[field]]), as.character(observations[[field]]))) {
      stop("Conflicting feature/assignment metadata: ", field)
    }
  }
  # Exactly the notebook's exclusions after RID becomes its index.
  feature_names <- names(features)[!names(features) %in% required]
  if (anyDuplicated(feature_names) || !all(grepl("^[0-9]+$", feature_names))) {
    stop("Expected unique genomic site column names without R name repair.")
  }
  binary_m6a <- as.matrix(features[, feature_names, drop = FALSE])
  if (!is.numeric(binary_m6a) || anyNA(binary_m6a) ||
      !all(binary_m6a %in% c(0, 1))) {
    stop("The existing matrix must contain only 0/1, without imputation or filtering.")
  }
  storage.mode(binary_m6a) <- "double"
  rownames(binary_m6a) <- features$RID
  if (!identical(dim(binary_m6a), c(215L, 827L))) {
    stop("The inspected notebook dataset is 215 reads x 827 sites; input has changed.")
  }
  observations$timepoint <- as.numeric(observations$timepoint)
  stopifnot(!anyNA(observations$timepoint),
            all(observations$timepoint %in% c(0, 5, 10, 15)),
            !anyNA(observations$cluster), length(unique(observations$chr)) == 1L)
  observations$leiden_cluster <- as.character(observations$cluster)
  annotation <- data.frame(feature = feature_names, position = as.numeric(feature_names),
                           chr = unique(observations$chr), stringsAsFactors = FALSE)
  window_annotation <- tables$window_annotation
  stopifnot(nrow(window_annotation) == length(feature_names),
            identical(as.numeric(window_annotation$win_start), annotation$position),
            identical(as.numeric(window_annotation$win_end), annotation$position),
            identical(as.numeric(window_annotation$mid), annotation$position))
  knn_edges <- tables$knn_edges
  required_edges <- c("from", "to", "weight", "manhattan_distance")
  if (!all(required_edges %in% names(knn_edges)) || !nrow(knn_edges) ||
      anyNA(knn_edges[, required_edges])) {
    stop("The original edge table must contain complete endpoints, affinities, and distances.")
  }
  rids <- rownames(binary_m6a)
  if (!all(knn_edges$from %in% rids) || !all(knn_edges$to %in% rids) ||
      any(knn_edges$from == knn_edges$to)) {
    stop("Every graph edge must join distinct members of the same 215-read matrix.")
  }
  edge_from <- match(knn_edges$from, rids)
  edge_to <- match(knn_edges$to, rids)
  edge_keys <- paste(pmin(edge_from, edge_to), pmax(edge_from, edge_to), sep = ":")
  if (anyDuplicated(edge_keys) || any(!is.finite(knn_edges$weight)) ||
      any(knn_edges$weight <= 0) || any(!is.finite(knn_edges$manhattan_distance)) ||
      any(knn_edges$manhattan_distance < 0)) {
    stop("Expected unique undirected edges with positive affinities and nonnegative distances.")
  }
  # Validate the existing edges against the matrix; do not reconstruct neighbors.
  checked_distances <- rowSums(abs(binary_m6a[edge_from, , drop = FALSE] -
                                   binary_m6a[edge_to, , drop = FALSE]))
  stopifnot(all(checked_distances == knn_edges$manhattan_distance))
  list(binary_m6a = binary_m6a, observations = observations, knn_edges = knn_edges,
       feature_annotation = annotation, window_annotation = window_annotation,
       source_parameters = tables$parameters, original_assignment_order = original_assignment_order,
       input_manifest = data.frame(input = names(input_paths), path = unname(input_paths),
                                   md5 = unname(tools::md5sum(input_paths))),
       table_dir = table_dir)
}
```

```{r load-inputs}
input <- load_cxcl2(project_root)
stopifnot(identical(rownames(input$binary_m6a), input$observations$RID))
print(dim(input$binary_m6a))
knitr::kable(input$input_manifest, caption = "Input paths")
```

## Construct the original graph among the 215 reads

Only RID is passed as a vertex attribute. Experimental time, sample labels,
and previous Leiden assignments are absent from the inference graph. The edge
file already contains one row per undirected edge. We do not simplify, symmetrize
by averaging, add self-loops, or drop vertices. Disconnected graphs or isolated
reads stop the analysis instead of silently changing the dataset.

```{r construct-original-knn-graph}
knn_edges <- input$knn_edges
read_ids <- rownames(input$binary_m6a)
g <- igraph::graph_from_data_frame(
  knn_edges, directed = FALSE, vertices = data.frame(name = read_ids))
stopifnot(igraph::vcount(g) == 215L,
          igraph::ecount(g) == nrow(knn_edges),
          identical(igraph::V(g)$name, read_ids),
          identical(igraph::E(g)$weight, knn_edges$weight),
          identical(igraph::E(g)$manhattan_distance, knn_edges$manhattan_distance),
          igraph::is_simple(g))
if (!igraph::is_connected(g)) {
  stop("The original graph is disconnected; no reads or components will be discarded.")
}
cat("Original graph:", igraph::vcount(g), "reads and", igraph::ecount(g), "undirected edges\n")
```

## Derive coordinates from the existing graph

Let $W$ be its unchanged weighted adjacency matrix and $D$ the diagonal matrix
of vertex strengths. We diagonalize the
[symmetric normalized Laplacian](https://r.igraph.org/reference/laplacian_matrix.html),
$L = I-D^{-1/2}WD^{-1/2}$. We discard the single trivial zero-eigenvalue mode
and retain the next five eigenvectors, with the smallest positive eigenvalues.
The read coordinates are $F=D^{-1/2}U$, the degree-normalized Laplacian eigenmap.
This construction is also used by
[`spectral_embedding`](https://github.com/scikit-learn/scikit-learn/blob/main/sklearn/manifold/_spectral_embedding.py).

Each eigenvector's sign is fixed by its largest absolute coordinate; this uses
no metadata. Coordinates have reads as rows and graph dimensions as columns.
All calculations are deterministic for this input and remain in memory.

```{r graph-embedding-helper}
embed_existing_graph <- function(g, n_dimensions = 5L) {
  if (length(n_dimensions) != 1L || !is.finite(n_dimensions) ||
      n_dimensions != as.integer(n_dimensions) || n_dimensions < 2L ||
      n_dimensions >= igraph::vcount(g)) {
    stop("graph_dimensions must be an integer between 2 and n_reads - 1.")
  }
  if (igraph::is_directed(g) || !igraph::is_simple(g) || !igraph::is_connected(g)) {
    stop("Embedding requires the unchanged, connected, simple undirected graph.")
  }
  W <- as.matrix(igraph::as_adjacency_matrix(g, attr = "weight", sparse = FALSE))
  vertex_strength <- rowSums(W)
  stopifnot(identical(rownames(W), igraph::V(g)$name),
            isTRUE(all.equal(W, t(W), tolerance = 1e-14)),
            all(diag(W) == 0), all(is.finite(W)), all(W >= 0))
  if (any(vertex_strength <= 0)) stop("Isolated reads cannot be dropped or embedded silently.")
  inv_sqrt_strength <- 1 / sqrt(vertex_strength)
  L_sym <- diag(nrow(W)) - W * outer(inv_sqrt_strength, inv_sqrt_strength)
  dimnames(L_sym) <- dimnames(W)
  decomposition <- eigen(L_sym, symmetric = TRUE)
  ascending <- order(decomposition$values)
  eigenvalues <- decomposition$values[ascending]
  eigenvectors <- decomposition$vectors[, ascending, drop = FALSE]
  tolerance <- 1e-10
  if (any(eigenvalues < -tolerance) || sum(abs(eigenvalues) <= tolerance) != 1L) {
    stop("Expected one trivial Laplacian mode for this connected graph.")
  }
  selected_modes <- seq.int(2L, n_dimensions + 1L)
  graph_coords <- sweep(eigenvectors[, selected_modes, drop = FALSE],
                        1L, sqrt(vertex_strength), "/")
  for (dimension in seq_len(n_dimensions)) {
    anchor <- which.max(abs(graph_coords[, dimension]))
    if (graph_coords[anchor, dimension] < 0) {
      graph_coords[, dimension] <- -graph_coords[, dimension]
    }
  }
  dimnames(graph_coords) <- list(igraph::V(g)$name, paste0("Graph", seq_len(n_dimensions)))
  # Check generalized eigenvectors: (D-W)F = DF Lambda; F' D F = I.
  D_times_F <- sweep(graph_coords, 1L, vertex_strength, "*")
  residual <- (diag(vertex_strength) - W) %*% graph_coords -
    sweep(D_times_F, 2L, eigenvalues[selected_modes], "*")
  stopifnot(all(is.finite(graph_coords)),
            max(abs(residual)) < 1e-8,
            max(abs(crossprod(graph_coords, D_times_F) - diag(n_dimensions))) < 1e-8,
            max(abs(colSums(D_times_F))) < 1e-8)
  list(graph_coords = graph_coords, adjacency = W, normalized_laplacian = L_sym,
       vertex_strength = setNames(vertex_strength, igraph::V(g)$name),
       spectrum = data.frame(mode = seq_along(eigenvalues), eigenvalue = eigenvalues,
                              retained = seq_along(eigenvalues) %in% selected_modes),
       n_dimensions = as.integer(n_dimensions),
       method = "Weighted Laplacian eigenmap: D^(-1/2) U, excluding the trivial mode")
}
```

```{r graph-derived-coordinates}
graph_embedding <- embed_existing_graph(g, n_dimensions = params$graph_dimensions)
graph_coords <- graph_embedding$graph_coords
stopifnot(identical(rownames(graph_coords), rownames(input$binary_m6a)),
          nrow(graph_coords) == 215L, ncol(graph_coords) == params$graph_dimensions)
knitr::kable(head(graph_embedding$spectrum, 12), digits = 6,
             caption = "Graph Laplacian spectrum (not PCA explained variance)")
cat("Graph coordinate matrix:", nrow(graph_coords), "reads x", ncol(graph_coords), "dimensions\n")
```

## Keep downstream model-based clustering, centers, and MST unchanged

The helper now accepts the read × graph-dimension matrix directly. It performs
no PCA or rescaling. The `mclust` model, candidate cluster counts, seed, empirical
center calculation, Euclidean center distances, MST, and TSCAN object structure
are retained from the preceding implementation. The field `pcareduceres` is
kept solely because `TSCANorder()` requires that API name; it holds graph
coordinates, not principal components.

```{r modified-exprmclust2}
exprmclust2 <- function(graph_coords, clusternum = 2:9, modelNames = "EVV", seed = 1234L) {
  set.seed(seed)
  if (!is.matrix(graph_coords) || !is.numeric(graph_coords) ||
      any(!is.finite(graph_coords)) || ncol(graph_coords) < 2L ||
      is.null(rownames(graph_coords)) || anyDuplicated(rownames(graph_coords))) {
    stop("exprmclust2 needs a finite read x graph-dimension matrix with unique RIDs.")
  }
  pcareduceres <- graph_coords
  clusternum <- clusternum[clusternum > 1L]
  # Keep fit warnings visible: failed candidate fits must be diagnosable.
  # Mclust evaluates a call to mclustBIC in the caller's environment.
  mclustBIC <- mclust::mclustBIC
  res <- mclust::Mclust(pcareduceres, G = clusternum, modelNames = modelNames)
  if (is.null(res) || is.null(res$z) || res$G < 2L || anyNA(res$z)) {
    stop("mclust did not return a valid multi-cluster model.")
  }
  clusterid <- setNames(apply(res$z, 1L, which.max), rownames(pcareduceres))
  if (!setequal(unique(clusterid), seq_len(res$G))) stop("An inferred cluster is empty.")
  clucenter <- do.call(rbind, lapply(seq_len(res$G), function(cid) {
    colMeans(pcareduceres[clusterid == cid, , drop = FALSE])
  }))
  rownames(clucenter) <- as.character(seq_len(res$G))
  colnames(clucenter) <- colnames(pcareduceres)
  center_distances <- as.matrix(stats::dist(clucenter))
  if (any(center_distances[upper.tri(center_distances)] == 0)) {
    stop("Coincident cluster centers: cannot reproduce the reference adjacency graph.")
  }
  gp <- igraph::graph_from_adjacency_matrix(center_distances, mode = "undirected",
                                           weighted = TRUE, diag = FALSE)
  dp_mst <- igraph::mst(gp, weights = igraph::E(gp)$weight)
  stopifnot(igraph::is_connected(dp_mst), igraph::ecount(dp_mst) == res$G - 1L)
  list(pcareduceres = pcareduceres, MSTtree = dp_mst, clusterid = clusterid,
       clucenter = clucenter, mclust = res, center_distances = center_distances,
       graph_ndim = ncol(graph_coords), seed = seed, modelNames = modelNames,
       candidate_clusters = clusternum)
}
```

```{r fit-tscan-model}
# Only graph-derived coordinates cross the model-fitting boundary.
fit <- exprmclust2(graph_coords, clusternum = 2:9, modelNames = "EVV", seed = 1234L)
stopifnot(identical(fit$pcareduceres, graph_coords))
cat("Graph dimensions:", fit$graph_ndim, "\nSelected TSCAN clusters:", fit$mclust$G, "\n")
```

### Model selection

```{r inspect-model-selection}
knitr::kable(data.frame(candidate_clusters = as.integer(rownames(fit$mclust$BIC)),
                        BIC = as.numeric(fit$mclust$BIC)),
             digits = 3, caption = "EVV BIC across candidate cluster counts")
print(table(fit$clusterid))
```

### Cluster centers and minimum-spanning tree

Centers are empirical means of assigned reads in all retained graph dimensions.
Edge weights are Euclidean distances in that space. This cluster-center MST is
distinct from the original read-level Manhattan kNN graph; the latter defines
the embedding and is retained unchanged in memory. The two-dimensional figure
shows the MST projected onto Graph1 and Graph2.

```{r inspect-centers-mst}
knitr::kable(fit$clucenter, digits = 5, caption = "Cluster centers in graph embedding space")
knitr::kable(igraph::as_data_frame(fit$MSTtree), digits = 5, caption = "Cluster-center MST edges")
```

## Infer read-level pseudotime with TSCANorder

We use the default path and orientation from
[`TSCANorder()`](https://github.com/zji90/TSCAN/blob/master/R/TSCANorder.R),
explicitly requesting its read-ID order. Pseudotime is ordinal rank rescaled to
0–1, not elapsed minutes or distance along the MST. No time-based root is used.

For a branched tree, the default TSCAN backbone can leave reads unordered. Such
reads remain in the matrix and metadata with `NA` pseudotime, are counted in the
report, and appear in a separate heatmap block. No artificial global rank is
assigned to them. Correlations and bin summaries use ordered reads only.

```{r tscan-read-order}
# Match Battaglia's default main path; explicitly request a vector because
# TSCAN versions have differed in their default return format.
ordered_rids <- as.character(TSCAN::TSCANorder(fit, orderonly = TRUE))
if (length(ordered_rids) < 2L || anyNA(ordered_rids) || anyDuplicated(ordered_rids) ||
    !all(ordered_rids %in% rownames(input$binary_m6a))) {
  stop("TSCANorder did not return a valid ordering of input RIDs.")
}
# TSCANorder gives ordinal position, not elapsed time or geodesic distance.
# Its default backbone may omit branch reads. Keep them in the complete data.
ranks <- match(rownames(input$binary_m6a), ordered_rids)
pseudotime <- (ranks - 1) / (length(ordered_rids) - 1)
trajectory <- list(fit = fit, ordered_rids = ordered_rids, pseudotime_rank = ranks,
     pseudotime = pseudotime,
     backbone_cluster_order = unique(unname(fit$clusterid[ordered_rids])),
     orientation_note = paste(
       "Unsupervised TSCAN default path and direction, determined by the MST and cluster IDs.",
       "No root or direction was chosen using experimental time or previous Leiden labels.",
       "Pseudotime is TSCAN order rank rescaled to [0,1]; direction is arbitrary.",
       "Reads outside the default ordering retain NA pseudotime."))
```

```{r ordering-audit}
cat("Ordered reads:", length(trajectory$ordered_rids),
    "\nUnordered reads:", sum(is.na(trajectory$pseudotime)), "\n")
cat(trajectory$orientation_note, "\n")
```

## Attach metadata after inference

```{r attach-evaluation-metadata}
obs <- input$observations
fit <- trajectory$fit
obs$tscan_cluster <- as.character(fit$clusterid[obs$RID])
obs$pseudotime_rank <- trajectory$pseudotime_rank
obs$pseudotime <- trajectory$pseudotime
obs$on_backbone <- is.finite(obs$pseudotime)
obs$pseudotime_bin <- as.integer(cut(obs$pseudotime, breaks = seq(0, 1, length.out = n_bins + 1L),
                                    include.lowest = TRUE, labels = FALSE))
obs$Graph1 <- fit$pcareduceres[obs$RID, 1L]
obs$Graph2 <- fit$pcareduceres[obs$RID, 2L]
cluster_levels <- rownames(fit$clucenter)
time_factor <- factor(obs$timepoint, levels = c(0, 5, 10, 15))
cluster_factor <- factor(obs$tscan_cluster, levels = cluster_levels)
bin_factor <- factor(obs$pseudotime_bin, levels = seq_len(n_bins))
leiden_factor <- factor(obs$leiden_cluster, levels = unique(obs$leiden_cluster))
```

## Correlations with experimental minutes

Both correlations are computed on the same ordered reads. Spearman uses
`exact = FALSE` because experimental minutes are tied. Confidence intervals are
reported for Pearson. Read-level p-values treat reads as independent; they do
not establish biological replicate-level significance. A negative coefficient
is retained because trajectory orientation is arbitrary.

```{r correlation-helper}
correlation_table <- function(observations) {
  selected <- observations$on_backbone
  x <- observations$timepoint[selected]
  y <- observations$pseudotime[selected]
  do.call(rbind, lapply(c("spearman", "pearson"), function(method) {
    if (length(x) < 3L || length(unique(x)) < 2L || length(unique(y)) < 2L) {
      return(data.frame(method = method, n_reads = length(x), estimate = NA_real_,
                        p_value = NA_real_, conf_low = NA_real_, conf_high = NA_real_))
    }
    test <- if (method == "spearman") {
      stats::cor.test(x, y, method = method, exact = FALSE)
    } else stats::cor.test(x, y, method = method)
    interval <- if (is.null(test$conf.int)) c(NA_real_, NA_real_) else test$conf.int[1:2]
    data.frame(method = method, n_reads = length(x), estimate = unname(test$estimate),
               p_value = test$p.value, conf_low = interval[1], conf_high = interval[2])
  }))
}
```

```{r time-correlations}
correlations <- correlation_table(obs)
knitr::kable(correlations, digits = 6,
             caption = "Experimental minutes versus inferred read pseudotime")
```

## Composition tables and previous Leiden labels

Counts retain zero combinations. Time composition within a pseudotime bin sums
to one; TSCAN composition is supplied both within experimental time and within
TSCAN cluster. Leiden labels are used only for comparison.

```{r composition-tables}
as_counts <- function(tab) {
  out <- as.data.frame(tab, stringsAsFactors = FALSE)
  names(out)[ncol(out)] <- "n_reads"
  out
}
bin_time <- as_counts(table(pseudotime_bin = bin_factor, timepoint = time_factor))
bin_time$pseudotime_bin <- as.integer(as.character(bin_time$pseudotime_bin))
bin_time$timepoint <- as.numeric(as.character(bin_time$timepoint))
bin_time$fraction <- bin_time$n_reads / ave(bin_time$n_reads, bin_time$pseudotime_bin, FUN = sum)
cluster_time <- as_counts(table(tscan_cluster = cluster_factor, timepoint = time_factor))
cluster_time$timepoint <- as.numeric(as.character(cluster_time$timepoint))
cluster_time$fraction_within_time <- cluster_time$n_reads /
  ave(cluster_time$n_reads, cluster_time$timepoint, FUN = sum)
cluster_time$fraction_within_cluster <- cluster_time$n_reads /
  ave(cluster_time$n_reads, cluster_time$tscan_cluster, FUN = sum)
comparison <- as_counts(table(leiden_cluster = leiden_factor, tscan_cluster = cluster_factor))
comparison$fraction_within_leiden <- comparison$n_reads /
  ave(comparison$n_reads, comparison$leiden_cluster, FUN = sum)
```

## Aggregate original m6A calls across pseudotime bins

Ten equal-width intervals on normalized rank define the bins. Site-level means
use the original binary values. Empty bins, if any, remain missing. No smoothing
or feature standardization is applied to the aggregate signal or heatmap.

```{r aggregate-binary-signal}
aggregate_m6a <- do.call(rbind, lapply(seq_len(n_bins), function(bin) {
  selected <- which(obs$pseudotime_bin == bin)
  if (!length(selected)) return(rep(NA_real_, ncol(input$binary_m6a)))
  colMeans(input$binary_m6a[selected, , drop = FALSE])
}))
dimnames(aggregate_m6a) <- list(as.character(seq_len(n_bins)), colnames(input$binary_m6a))
bin_summary <- do.call(rbind, lapply(seq_len(n_bins), function(bin) {
  selected <- which(obs$pseudotime_bin == bin)
  data.frame(pseudotime_bin = bin, n_reads = length(selected),
             pseudotime_min = if (length(selected)) min(obs$pseudotime[selected]) else NA_real_,
             pseudotime_max = if (length(selected)) max(obs$pseudotime[selected]) else NA_real_,
             mean_m6a = if (length(selected)) mean(input$binary_m6a[selected, , drop = FALSE]) else NA_real_)
}))
```

```{r bin-summary-table}
knitr::kable(bin_summary, digits = 4)
```

## Prepare in-memory figure inputs

All statistics, composition summaries, and plotting coordinates are calculated
here. The figure chunks below use these in-memory objects.

```{r prepare-figure-inputs}
centers <- data.frame(tscan_cluster = cluster_levels, fit$clucenter, check.names = FALSE)
edges <- igraph::as_data_frame(fit$MSTtree, what = "edges")
edges$x <- centers$Graph1[match(edges$from, centers$tscan_cluster)]
edges$y <- centers$Graph2[match(edges$from, centers$tscan_cluster)]
edges$xend <- centers$Graph1[match(edges$to, centers$tscan_cluster)]
edges$yend <- centers$Graph2[match(edges$to, centers$tscan_cluster)]
inclusion <- as_counts(table(timepoint = time_factor, on_backbone = obs$on_backbone))
```

```{r collect-summaries}
summaries <- list(observations = obs, centers = centers, mst_edges = edges,
     graph_spectrum = graph_embedding$spectrum,
     correlations = correlations,
     bin_time_composition = bin_time, cluster_time_composition = cluster_time,
     leiden_tscan_comparison = comparison,
     adjusted_rand_index = mclust::adjustedRandIndex(obs$leiden_cluster, obs$tscan_cluster),
     aggregate_m6a = aggregate_m6a, bin_summary = bin_summary,
     backbone_inclusion_by_time = inclusion,
     heatmap_order = order(obs$pseudotime, seq_len(nrow(obs)), na.last = TRUE))
```

## Keep results and provenance in memory

```{r record-provenance}
provenance <- list(table_dir = input$table_dir, matrix_n_reads = nrow(input$binary_m6a),
  matrix_n_features = ncol(input$binary_m6a),
  n_ordered = length(trajectory$ordered_rids),
  n_unordered = sum(!summaries$observations$on_backbone),
  seed = fit$seed, model = fit$modelNames, graph_dimensions = fit$graph_ndim,
  selected_clusters = fit$mclust$G, candidate_clusters = fit$candidate_clusters,
  graph_n_edges = igraph::ecount(g), embedding_method = graph_embedding$method,
  embedding_weight_attribute = "weight (original affinity)",
  embedding_reference_url = "https://r.igraph.org/reference/laplacian_matrix.html",
  pseudotime_bins = 10L, orientation_note = trajectory$orientation_note,
  reference_url = "https://github.com/BernsteinLab/targetedNOMe_2022/blob/master/pseudotime/tcell.r",
  ordering_reference_url = "https://github.com/zji90/TSCAN/blob/master/R/TSCANorder.R",
  source_notebook = file.path(project_root, "code/pseudotime/PAGA_DT.ipynb"),
  source_notebook_md5 = unname(tools::md5sum(file.path(project_root, "code/pseudotime/PAGA_DT.ipynb"))),
  input_manifest = input$input_manifest,
  package_versions = setNames(vapply(required, function(p) as.character(utils::packageVersion(p)),
                                     character(1)), required),
  created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  session_info = utils::sessionInfo(),
  statistical_note = paste("Correlations use ordered reads only; Spearman uses a tie-aware",
                           "asymptotic test (experimental times repeat). P-values treat reads",
                           "as independent and do not establish biological replicate-level significance.")
)
```

```{r assemble-validate-results}
result <- c(list(schema_version = "2.0", binary_m6a = input$binary_m6a,
                 feature_annotation = input$feature_annotation,
                 window_annotation = input$window_annotation,
                 source_parameters = input$source_parameters,
                 original_assignment_order = input$original_assignment_order,
                 original_knn_graph = g, knn_edges = knn_edges,
                 graph_embedding = graph_embedding, graph_coords = graph_coords,
                 tscan = fit, ordered_rids = trajectory$ordered_rids,
                 backbone_cluster_order = trajectory$backbone_cluster_order,
                 provenance = provenance), summaries)
stopifnot(identical(rownames(result$binary_m6a), result$observations$RID),
          identical(colnames(result$binary_m6a), result$feature_annotation$feature),
          identical(sort(result$heatmap_order), seq_len(nrow(result$binary_m6a))),
          identical(result$observations$RID[order(result$observations$pseudotime, na.last = NA)],
                    result$ordered_rids),
          all(result$observations$pseudotime[result$observations$on_backbone] >= 0),
          all(result$observations$pseudotime[result$observations$on_backbone] <= 1))
```

```{r analysis-summary}
cat(sprintf("CXCL2: %d reads x %d original binary sites\n",
            nrow(input$binary_m6a), ncol(input$binary_m6a)))
cat(sprintf("TSCAN: %d graph dimensions, %d clusters; %d ordered and %d unordered reads\n",
            fit$graph_ndim, fit$mclust$G, provenance$n_ordered, provenance$n_unordered))
print(result$correlations, row.names = FALSE)
cat("Leiden/TSCAN adjusted Rand index:", result$adjusted_rand_index, "\n")
```

All fitted objects, per-read scores, checksums, summary tables, and session
information remain in `result` for interactive inspection. Nothing in this
section writes RDS, TSV, CSV, text, or session-information files.

## Figure generation

The following chunks use `result` from the analysis above. They do not rerun
graph embedding, clustering, trajectory inference, or association tests. Each figure
is displayed here and exported as PDF. The default figure directory is
`figures/` inside the analysis output directory; use `params$figure_dir` to change it.

```{r plotting-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      fig.width = 8, fig.height = 5, dpi = 160)
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop("Plotting requires ggplot2. Install it in your R environment before rendering.")
}
library(ggplot2)
if (!exists("result", inherits = FALSE)) {
  stop("Run the preceding analysis chunks to create the in-memory result object.")
}
required_objects <- c("schema_version", "binary_m6a", "observations",
                      "feature_annotation", "graph_spectrum", "graph_coords", "centers", "mst_edges",
                      "correlations", "bin_time_composition", "cluster_time_composition",
                      "leiden_tscan_comparison", "aggregate_m6a", "bin_summary",
                      "heatmap_order", "provenance")
missing_objects <- setdiff(required_objects, names(result))
if (length(missing_objects)) {
  stop("Incomplete in-memory analysis: missing ", paste(missing_objects, collapse = ", "))
}
if (!identical(result$schema_version, "2.0")) {
  stop("Unsupported analysis schema: ", result$schema_version)
}
require_columns <- function(x, columns, label) {
  missing <- setdiff(columns, names(x))
  if (length(missing)) stop(label, " is missing: ", paste(missing, collapse = ", "))
}
obs <- result$observations
require_columns(obs, c("RID", "timepoint", "leiden_cluster", "tscan_cluster",
                       "pseudotime", "pseudotime_rank", "pseudotime_bin",
                       "on_backbone", "Graph1", "Graph2"), "observations")
require_columns(result$feature_annotation, c("feature", "position", "chr"),
                "feature_annotation")
require_columns(result$graph_spectrum, c("mode", "eigenvalue", "retained"), "graph_spectrum")
require_columns(result$centers, c("tscan_cluster", "Graph1", "Graph2"), "centers")
require_columns(result$mst_edges, c("from", "to", "x", "y", "xend", "yend"), "mst_edges")
require_columns(result$correlations,
                c("method", "n_reads", "estimate", "p_value", "conf_low", "conf_high"),
                "correlations")
require_columns(result$bin_time_composition,
                c("pseudotime_bin", "timepoint", "n_reads", "fraction"),
                "bin_time_composition")
require_columns(result$cluster_time_composition,
                c("tscan_cluster", "timepoint", "n_reads", "fraction_within_time"),
                "cluster_time_composition")
require_columns(result$leiden_tscan_comparison,
                c("leiden_cluster", "tscan_cluster", "n_reads", "fraction_within_leiden"),
                "leiden_tscan_comparison")
require_columns(result$bin_summary,
                c("pseudotime_bin", "n_reads", "pseudotime_min", "pseudotime_max", "mean_m6a"),
                "bin_summary")
required_provenance <- c("table_dir", "matrix_n_reads", "matrix_n_features", "n_ordered",
                         "n_unordered", "seed", "model", "graph_dimensions", "selected_clusters",
                         "orientation_note", "reference_url")
if (length(setdiff(required_provenance, names(result$provenance)))) {
  stop("In-memory provenance is incomplete; rerun the analysis chunks above.")
}
stopifnot(is.matrix(result$binary_m6a),
          identical(rownames(result$binary_m6a), as.character(obs$RID)),
          identical(colnames(result$binary_m6a), as.character(result$feature_annotation$feature)),
          ncol(result$binary_m6a) == nrow(result$feature_annotation),
          all(result$binary_m6a %in% c(0, 1)),
          identical(sort(as.integer(result$heatmap_order)), seq_len(nrow(obs))),
          !anyNA(obs$timepoint), all(obs$timepoint %in% c(0, 5, 10, 15)),
          !anyNA(obs$tscan_cluster), !anyNA(obs$leiden_cluster),
          !anyNA(obs$on_backbone),
          identical(is.finite(obs$pseudotime), obs$on_backbone),
          sum(obs$on_backbone) == result$provenance$n_ordered,
          sum(!obs$on_backbone) == result$provenance$n_unordered,
          is.matrix(result$aggregate_m6a),
          identical(colnames(result$aggregate_m6a), colnames(result$binary_m6a)))
if (any(diff(obs$pseudotime[result$heatmap_order][seq_len(result$provenance$n_ordered)]) < 0)) {
  stop("In-memory heatmap_order is not increasing in inferred pseudotime.")
}
if (!all(obs$on_backbone[result$heatmap_order] ==
         c(rep(TRUE, result$provenance$n_ordered), rep(FALSE, result$provenance$n_unordered)))) {
  stop("In-memory heatmap_order must place unordered reads after backbone reads.")
}
figure_dir <- if (is.null(params$figure_dir)) {
  file.path(output_dir, "figures")
} else params$figure_dir
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
if (!dir.exists(figure_dir)) stop("Could not create figure directory: ", figure_dir)
time_levels <- c("0", "5", "10", "15")
time_colors <- c("0" = "#440154", "5" = "#31688E", "10" = "#35B779", "15" = "#FDE725")
label_order <- function(x) {
  x <- unique(as.character(x))
  numeric_labels <- suppressWarnings(as.numeric(x))
  if (all(is.finite(numeric_labels))) x[order(numeric_labels)] else sort(x)
}
tscan_levels <- label_order(obs$tscan_cluster)
leiden_levels <- label_order(obs$leiden_cluster)
cluster_colors <- setNames(grDevices::hcl.colors(length(tscan_levels), "Dark 3"), tscan_levels)
leiden_colors <- setNames(grDevices::hcl.colors(length(leiden_levels), "Set 2"), leiden_levels)
pseudotime_colors <- grDevices::hcl.colors(256, "Viridis")
obs$experimental_time <- factor(obs$timepoint, levels = time_levels)
obs$tscan_cluster <- factor(obs$tscan_cluster, levels = tscan_levels)
obs$leiden_cluster <- factor(obs$leiden_cluster, levels = leiden_levels)
ordered_obs <- obs[obs$on_backbone, , drop = FALSE]
theme_set(theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(),
                                         plot.title.position = "plot"))
save_plot <- function(plot, name, width = 8, height = 5) {
  ggplot2::ggsave(file.path(figure_dir, paste0(name, ".pdf")), plot,
                 width = width, height = height, bg = "white")
  print(plot)
}
percent_labels <- function(x) paste0(round(x * 100), "%")
backbone_caption <- sprintf("Ordered backbone: %s reads; %s off-backbone reads have no pseudotime.",
                            format(result$provenance$n_ordered, big.mark = ","),
                            format(result$provenance$n_unordered, big.mark = ","))
```

## Analysis provenance

```{r analysis-provenance, echo=FALSE, results='asis'}
provenance_table <- data.frame(
  Item = c("Source table directory", "Original reads", "Original m6A sites",
           "Reads assigned pseudotime", "Off-backbone reads (pseudotime missing)",
           "Random seed", "mclust covariance model", "Graph dimensions used",
           "Selected number of TSCAN clusters", "Orientation", "Reference"),
  Value = c(result$provenance$table_dir, result$provenance$matrix_n_reads,
            result$provenance$matrix_n_features, result$provenance$n_ordered,
            result$provenance$n_unordered, result$provenance$seed,
            result$provenance$model, result$provenance$graph_dimensions,
            result$provenance$selected_clusters, result$provenance$orientation_note,
            result$provenance$reference_url),
  check.names = FALSE)
knitr::kable(provenance_table)
```

## Graph embedding colored by experimental time

All original reads are shown, including reads outside the selected ordering
backbone. Colors are applied after trajectory inference.

```{r graph-embedding-experimental-time}
p <- ggplot(obs, aes(Graph1, Graph2, color = experimental_time)) +
  geom_point(size = 0.85, alpha = 0.45) +
  scale_color_manual(values = time_colors, drop = FALSE, name = "Time (min)") +
  labs(title = "CXCL2: embedding of the original Manhattan kNN graph",
       x = "Graph dimension 1", y = "Graph dimension 2") +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 3)))
save_plot(p, "01_graph_embedding_experimental_time")
```

## TSCAN clusters and minimum-spanning tree

The tree was constructed using the cluster centers in all retained graph
dimensions. This figure projects the centers and edges onto Graph1 and Graph2, so
projected distances do not necessarily equal the distances used for the tree.
Labels identify the model-based TSCAN clusters; every MST edge is shown.

```{r graph-embedding-tscan-mst}
centers <- result$centers
centers$tscan_cluster <- factor(centers$tscan_cluster, levels = tscan_levels)
p <- ggplot(obs, aes(Graph1, Graph2, color = tscan_cluster)) +
  geom_point(size = 0.8, alpha = 0.35) +
  geom_segment(data = result$mst_edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               inherit.aes = FALSE, color = "#262626", linewidth = 0.8) +
  geom_point(data = centers, size = 4, shape = 21, fill = "white", stroke = 1.2) +
  geom_text(data = centers, aes(label = tscan_cluster), color = "black",
            size = 3, show.legend = FALSE) +
  scale_color_manual(values = cluster_colors, drop = FALSE, name = "TSCAN cluster") +
  labs(title = "CXCL2: TSCAN clusters and cluster-center MST",
       x = "Graph dimension 1", y = "Graph dimension 2",
       caption = backbone_caption) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 3)))
save_plot(p, "02_graph_embedding_TSCAN_MST")
```

## Pseudotime and experimental minutes

The boxplots and correlations include only reads assigned finite pseudotime.
The time labels never enter inference or choose the sign of the association.
The computed Spearman and Pearson tests treat each read as an observation; their
p-values do not account for shared sample effects and do not establish a
causal temporal sequence. Missing confidence limits mean that the corresponding
test did not supply an interval.

```{r display-time-correlations, echo=FALSE}
knitr::kable(result$correlations, digits = c(0, 0, 4, 6, 4, 4),
             caption = "Read-level associations with experimental minutes")
```

```{r pseudotime-experimental-time}
p <- ggplot(ordered_obs, aes(experimental_time, pseudotime, fill = experimental_time)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.65) +
  geom_point(position = position_jitter(width = 0.18, height = 0, seed = 1234),
             size = 0.65, alpha = 0.2, color = "#303030") +
  scale_fill_manual(values = time_colors, drop = FALSE, guide = "none") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(title = "CXCL2: inferred pseudotime by experimental time",
       x = "Experimental time (min)", y = "TSCAN pseudotime (normalized rank)",
       caption = backbone_caption)
save_plot(p, "03_pseudotime_experimental_time")
```

## Experimental-time composition across pseudotime bins

Each bar shows the computed fraction of ordered reads from each timepoint within a
pseudotime bin. The bin assignments and counts were computed by the analysis
notebook and are reused unchanged.

```{r time-composition-pseudotime-bins}
bin_composition <- result$bin_time_composition
bin_composition$experimental_time <- factor(bin_composition$timepoint, levels = time_levels)
p <- ggplot(bin_composition, aes(factor(pseudotime_bin), fraction, fill = experimental_time)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = time_colors, drop = FALSE, name = "Time (min)") +
  scale_y_continuous(labels = percent_labels, limits = c(0, 1),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "CXCL2: experimental-time composition along pseudotime",
       x = "Pseudotime bin (early to late in inferred ordering)",
       y = "Fraction of reads within bin", caption = backbone_caption)
save_plot(p, "04_time_composition_pseudotime_bins")
```

```{r display-bin-summary, echo=FALSE}
knitr::kable(result$bin_summary, digits = 4,
             caption = "Pseudotime-bin sizes, rank ranges, and mean binary m6A signal")
```

## TSCAN cluster composition by experimental time

Every original read contributes here, including off-backbone reads. Fractions
sum to one within each experimental timepoint.

```{r tscan-composition-experimental-time}
cluster_composition <- result$cluster_time_composition
cluster_composition$experimental_time <- factor(cluster_composition$timepoint, levels = time_levels)
cluster_composition$tscan_cluster <- factor(cluster_composition$tscan_cluster, levels = tscan_levels)
p <- ggplot(cluster_composition,
            aes(experimental_time, fraction_within_time, fill = tscan_cluster)) +
  geom_col(width = 0.75) +
  scale_fill_manual(values = cluster_colors, drop = FALSE, name = "TSCAN cluster") +
  scale_y_continuous(labels = percent_labels, limits = c(0, 1),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "CXCL2: TSCAN cluster composition by experimental time",
       x = "Experimental time (min)", y = "Fraction of reads within timepoint")
save_plot(p, "05_TSCAN_composition_experimental_time")
```

## Original binary m6A matrix ordered by pseudotime

The matrix below retains every original read and every original binary site
value. The in-memory read permutation sorts finite pseudotime from top to bottom.
Any off-backbone reads form a separately labeled block at the bottom in their
original input order; their pseudotime strip is gray. All annotations use the
same read permutation. Sites retain their original matrix column order; genomic
positions label selected columns, and column spacing is not genomic distance.
The raster uses the complete matrix without sampling or smoothing. At screen
resolution, individual rows or columns may require zooming to inspect.

```{r binary-m6a-pseudotime-heatmap, fig.width=16, fig.height=10}
draw_binary_heatmap <- function() {
  old_par <- par(no.readonly = TRUE)
  on.exit({layout(1); par(old_par)}, add = TRUE)
  index <- result$heatmap_order
  n_reads <- nrow(result$binary_m6a)
  n_sites <- ncol(result$binary_m6a)
  n_ordered <- result$provenance$n_ordered
  heat_obs <- obs[index, , drop = FALSE]
  binary_colors <- c("#F4F4F4", "#283E73")
  pt_index <- 1L + floor(255 * heat_obs$pseudotime)
  pt_strip <- rep("#BDBDBD", n_reads)
  pt_strip[heat_obs$on_backbone] <- pseudotime_colors[pt_index[heat_obs$on_backbone]]
  layout(matrix(seq_len(5), nrow = 1), widths = c(1.3, 0.8, 0.8, 12, 2.8))
  boundary <- function() {
    if (n_ordered > 0 && n_ordered < n_reads) {
      abline(h = n_ordered + 0.5, col = "#C83232", lwd = 1.2)
    }
  }
  strip <- function(colors, title, annotate = FALSE) {
    par(mar = c(5, if (annotate) 5.5 else 0.2, 3.5, 0.4))
    plot.new()
    plot.window(xlim = c(0, 1), ylim = c(n_reads + 0.5, 0.5), xaxs = "i", yaxs = "i")
    rasterImage(as.raster(matrix(colors, ncol = 1)), 0, n_reads + 0.5, 1, 0.5,
                interpolate = FALSE)
    title(main = title, cex.main = 0.9)
    if (annotate) {
      if (n_ordered > 0) {
        axis(2, at = (n_ordered + 1) / 2,
             labels = sprintf("Ordered\n(n=%s)", format(n_ordered, big.mark = ",")),
             las = 1, tick = FALSE, cex.axis = 0.7)
      }
      if (n_ordered < n_reads) {
        axis(2, at = (n_ordered + n_reads + 1) / 2,
             labels = sprintf("Off backbone\n(n=%s)", format(n_reads - n_ordered, big.mark = ",")),
             las = 1, tick = FALSE, cex.axis = 0.7)
      }
    }
    boundary()
    box()
  }
  strip(unname(time_colors[as.character(heat_obs$experimental_time)]), "Time", TRUE)
  strip(unname(cluster_colors[as.character(heat_obs$tscan_cluster)]), "TSCAN")
  strip(pt_strip, "Pseudotime")
  par(mar = c(5, 0.2, 3.5, 0.4))
  plot.new()
  plot.window(xlim = c(0.5, n_sites + 0.5), ylim = c(n_reads + 0.5, 0.5),
              xaxs = "i", yaxs = "i")
  raster_colors <- matrix(binary_colors[result$binary_m6a[index, , drop = FALSE] + 1L],
                           nrow = n_reads, ncol = n_sites)
  rasterImage(as.raster(raster_colors), 0.5, n_reads + 0.5, n_sites + 0.5, 0.5,
              interpolate = FALSE)
  site_ticks <- unique(round(seq(1, n_sites, length.out = min(6, n_sites))))
  axis(1, at = site_ticks,
       labels = format(result$feature_annotation$position[site_ticks],
                       scientific = FALSE, big.mark = ",", trim = TRUE),
       cex.axis = 0.75)
  title(main = "CXCL2: original binary m6A calls", xlab = "Genomic site position (original columns)")
  boundary()
  box()
  par(mar = c(5, 0, 3.5, 0))
  plot.new()
  plot.window(xlim = c(0, 1), ylim = c(0, 1))
  legend(0, 1, legend = paste(time_levels, "min"), fill = time_colors,
         title = "Experimental time", bty = "n", cex = 0.8)
  legend(0, 0.76, legend = names(cluster_colors), fill = cluster_colors,
         title = "TSCAN cluster", bty = "n", cex = 0.75,
         ncol = if (length(cluster_colors) > 7) 2 else 1)
  legend(0, 0.34, legend = c("0: unmarked", "1: marked"), fill = binary_colors,
         title = "Binary m6A", bty = "n", cex = 0.8)
  rasterImage(as.raster(matrix(pseudotime_colors, nrow = 1)),
              0.05, 0.125, 0.9, 0.15, interpolate = FALSE)
  text(c(0.05, 0.475, 0.9), 0.1, c("0", "0.5", "1"), cex = 0.75)
  text(0.475, 0.18, "Pseudotime", cex = 0.8)
  legend(0, 0.07, legend = "Missing: off backbone", fill = "#BDBDBD", bty = "n", cex = 0.75)
}
draw_binary_heatmap()
grDevices::pdf(file.path(figure_dir, "06_binary_m6a_pseudotime_heatmap.pdf"), width = 16, height = 10)
tryCatch(draw_binary_heatmap(), finally = grDevices::dev.off())
```

## Baseline: experimental time and per-read m6A level

This baseline groups reads by **0, 5, 10, then 15 minutes**, from top to bottom.
Within each timepoint, reads are sorted from **low to high fraction of marked
sites**, calculated as `rowMeans()` of the unchanged binary m6A matrix. Because
every read has the same 827 sites, sorting by the number of 1s gives the same
ordering. Ties are resolved by RID. All 215 reads and all original site columns
are retained. The middle panel shows each read's fraction of marked sites;
horizontal lines separate timepoints.

Experimental time is used here only to arrange this baseline figure. This
display order is not pseudotime and does not enter TSCAN inference.

```{r baseline-time-m6a-order}
baseline_metadata <- data.frame(
  RID = rownames(result$binary_m6a),
  timepoint = result$observations$timepoint[
    match(rownames(result$binary_m6a), result$observations$RID)],
  fraction_marked = rowMeans(result$binary_m6a),
  stringsAsFactors = FALSE)
stopifnot(!anyNA(baseline_metadata), !anyDuplicated(baseline_metadata$RID),
          all(baseline_metadata$timepoint %in% c(0, 5, 10, 15)))
baseline_heatmap_order <- with(baseline_metadata,
  order(timepoint, fraction_marked, RID, method = "radix"))
baseline_obs <- baseline_metadata[baseline_heatmap_order, , drop = FALSE]
stopifnot(identical(sort(baseline_heatmap_order), seq_len(nrow(result$binary_m6a))),
          identical(baseline_obs$RID, rownames(result$binary_m6a)[baseline_heatmap_order]),
          all(diff(baseline_obs$timepoint) >= 0),
          all(vapply(split(baseline_obs$fraction_marked, baseline_obs$timepoint),
                     function(x) all(diff(x) >= 0), logical(1))))
```

```{r baseline-time-m6a-heatmap-helper}
draw_baseline_heatmap <- function() {
  old_par <- par(no.readonly = TRUE)
  on.exit({layout(1); par(old_par)}, add = TRUE)
  n_reads <- nrow(result$binary_m6a)
  n_sites <- ncol(result$binary_m6a)
  binary_colors <- c("#F4F4F4", "#283E73")
  group_counts <- table(factor(baseline_obs$timepoint, levels = as.numeric(time_levels)))
  group_ends <- cumsum(as.integer(group_counts))
  group_centers <- group_ends - as.integer(group_counts) / 2 + 0.5
  boundaries <- unique(group_ends[group_ends > 0 & group_ends < n_reads]) + 0.5
  boundary <- function() abline(h = boundaries, col = "#C83232", lwd = 0.8)
  read_window <- function(xlim) {
    plot.new()
    plot.window(xlim = xlim, ylim = c(n_reads + 0.5, 0.5), xaxs = "i", yaxs = "i")
  }
  layout(matrix(seq_len(4), nrow = 1), widths = c(1.8, 2, 12, 1.8))

  par(mar = c(5, 5.5, 3.5, 0.4))
  read_window(c(0, 1))
  rasterImage(as.raster(matrix(unname(time_colors[as.character(baseline_obs$timepoint)]),
                               ncol = 1)),
              0, n_reads + 0.5, 1, 0.5, interpolate = FALSE)
  present <- group_counts > 0
  axis(2, at = group_centers[present],
       labels = sprintf("%s min\n(n=%d)", time_levels[present], group_counts[present]),
       las = 1, tick = FALSE, cex.axis = 0.8)
  title(main = "Time", cex.main = 0.95)
  boundary()
  box()

  par(mar = c(5, 0.3, 3.5, 0.7))
  read_window(c(0, 1))
  abline(v = c(0, 0.5, 1), col = "#E0E0E0", lwd = 0.6)
  points(baseline_obs$fraction_marked, seq_len(n_reads),
         pch = 16, cex = 0.45, col = "#283E73")
  axis(1, at = c(0, 0.5, 1), labels = c("0", "0.5", "1"), cex.axis = 0.75)
  title(main = "Per-read m6A", xlab = "Fraction marked", cex.main = 0.95, cex.lab = 0.85)
  boundary()
  box()

  par(mar = c(5, 0.2, 3.5, 0.4))
  read_window(c(0.5, n_sites + 0.5))
  raster_colors <- matrix(
    binary_colors[result$binary_m6a[baseline_heatmap_order, , drop = FALSE] + 1L],
    nrow = n_reads, ncol = n_sites)
  rasterImage(as.raster(raster_colors), 0.5, n_reads + 0.5, n_sites + 0.5, 0.5,
              interpolate = FALSE)
  site_ticks <- unique(round(seq(1, n_sites, length.out = min(6, n_sites))))
  axis(1, at = site_ticks,
       labels = format(result$feature_annotation$position[site_ticks],
                       scientific = FALSE, big.mark = ",", trim = TRUE),
       cex.axis = 0.75)
  title(main = "CXCL2 baseline: experimental time, then increasing m6A level",
        xlab = "Genomic site position (original columns)")
  boundary()
  box()

  par(mar = c(5, 0, 3.5, 0))
  plot.new()
  plot.window(xlim = c(0, 1), ylim = c(0, 1))
  legend(0, 1, legend = c("0: unmarked", "1: marked"), fill = binary_colors,
         title = "Binary m6A", bty = "n", cex = 0.8)
}
```

```{r baseline-time-m6a-heatmap, fig.width=16, fig.height=10}
draw_baseline_heatmap()
grDevices::pdf(file.path(figure_dir, "06b_binary_m6a_time_level_baseline_heatmap.pdf"),
               width = 16, height = 10)
tryCatch(draw_baseline_heatmap(), finally = grDevices::dev.off())
```

## Aggregate m6A signal across pseudotime bins

Each tile displays the computed mean of the **original binary calls** at a site
within a pseudotime bin. These are fractions of marked reads, not standardized
graph-embedding coordinates. Every original feature is retained. Missing bins, if any, remain
missing; off-backbone reads do not contribute to these aggregates.

```{r aggregate-m6a-pseudotime-bins, fig.width=12, fig.height=5}
aggregate_matrix <- result$aggregate_m6a
aggregate_display <- data.frame(
  pseudotime_bin = rep(as.integer(rownames(aggregate_matrix)), times = ncol(aggregate_matrix)),
  site_index = rep(seq_len(ncol(aggregate_matrix)), each = nrow(aggregate_matrix)),
  fraction_marked = as.vector(aggregate_matrix))
site_ticks <- unique(round(seq(1, ncol(aggregate_matrix), length.out = min(6, ncol(aggregate_matrix)))))
p <- ggplot(aggregate_display, aes(site_index, pseudotime_bin, fill = fraction_marked)) +
  geom_raster() +
  scale_fill_gradientn(colors = pseudotime_colors, limits = c(0, 1),
                        na.value = "#BDBDBD", name = "Fraction\nmarked") +
  scale_x_continuous(breaks = site_ticks,
                     labels = format(result$feature_annotation$position[site_ticks],
                                     scientific = FALSE, big.mark = ",", trim = TRUE),
                     expand = c(0, 0)) +
  scale_y_reverse(breaks = as.integer(rownames(aggregate_matrix)), expand = c(0, 0)) +
  labs(title = "CXCL2: site-level m6A signal across pseudotime bins",
       x = "Genomic site position (original columns)", y = "Pseudotime bin",
       caption = backbone_caption) +
  theme(panel.grid = element_blank())
save_plot(p, "07_aggregate_m6a_pseudotime_bins", width = 12)
```

The following companion panel shows the computed average across all original
binary features in each bin. It summarizes overall marking without smoothing.

```{r aggregate-global-m6a-pseudotime-bins}
p <- ggplot(result$bin_summary, aes(pseudotime_bin, mean_m6a)) +
  geom_line(color = "#31688E", linewidth = 0.7) +
  geom_point(aes(size = n_reads), color = "#31688E") +
  scale_x_continuous(breaks = result$bin_summary$pseudotime_bin) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "CXCL2: overall m6A signal along inferred pseudotime",
       x = "Pseudotime bin", y = "Mean binary m6A signal across original sites",
       size = "Reads per bin", caption = backbone_caption)
save_plot(p, "07b_global_m6a_pseudotime_bins")
```

## Pseudotime distributions for previous Leiden clusters

The existing Leiden labels annotate reads after TSCAN inference. Only ordered
reads enter these distributions; the categories retain their label order
and do not imply a trajectory between Leiden clusters.

```{r pseudotime-previous-leiden-clusters}
p <- ggplot(ordered_obs, aes(leiden_cluster, pseudotime, fill = leiden_cluster)) +
  geom_boxplot(width = 0.65, outlier.shape = NA, alpha = 0.7) +
  geom_point(position = position_jitter(width = 0.18, height = 0, seed = 1234),
             size = 0.65, alpha = 0.2, color = "#303030") +
  scale_x_discrete(drop = FALSE) +
  scale_fill_manual(values = leiden_colors, drop = FALSE, guide = "none") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(title = "CXCL2: TSCAN pseudotime by previous Leiden cluster",
       x = "Previous Leiden cluster", y = "TSCAN pseudotime (normalized rank)",
       caption = backbone_caption)
save_plot(p, "08_pseudotime_previous_Leiden_clusters")
```

## Previous Leiden cluster × TSCAN cluster

This comparison includes every original read. Tile colors represent the computed
fraction within each Leiden cluster; labels show read counts. It compares
two partitions of the same reads and does not assign a temporal sequence.

```{r leiden-tscan-cluster-comparison}
comparison <- result$leiden_tscan_comparison
comparison$leiden_cluster <- factor(comparison$leiden_cluster, levels = leiden_levels)
comparison$tscan_cluster <- factor(comparison$tscan_cluster, levels = tscan_levels)
p <- ggplot(comparison, aes(tscan_cluster, leiden_cluster, fill = fraction_within_leiden)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = n_reads, color = fraction_within_leiden > 0.55), size = 3.6) +
  scale_fill_gradient(low = "#F2F6FA", high = "#174A75", limits = c(0, 1),
                       labels = percent_labels, name = "Fraction within\nLeiden cluster") +
  scale_color_manual(values = c("FALSE" = "#202020", "TRUE" = "white"), guide = "none") +
  scale_x_discrete(drop = FALSE) + scale_y_discrete(drop = FALSE) +
  coord_equal() +
  labs(title = "CXCL2: previous Leiden and TSCAN cluster assignments",
       x = "TSCAN cluster", y = "Previous Leiden cluster",
       caption = "Tile labels: read counts. All original reads are included.") +
  theme(panel.grid = element_blank())
save_plot(p, "09_Leiden_TSCAN_cluster_comparison")
```

## Reproducibility

The input matrix, original metadata, fitted objects, read ordering, and statistics
remain in memory in `result`. The figure chunks only add
display factors, calculate the baseline display order, and reshape in-memory
summaries for plotting. Its plotting-session
information is shown below; the trajectory's software versions are recorded by
the analysis chunks above.

```{r plotting-session-info, echo=FALSE}
sessionInfo()
```
