---
title: "Macrophage TSS autocorrelation by gene expression"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    mathjax: null
params:
  output_dir: "../../../../macrophage_project/auto_correlation/tss"
  windows: ["2000_tss", "upstream_1000_tss", "downstream_1000_tss"]
---

# Analysis and parameters

strand-oriented: negative = upstream, positive = downstream around TSS, so that both strands have the same orientation . For the interval `[s, e)` and the 0 based TSS base `tss` (canonical BED start, 1 bp rows), the genomic window is `[tss + s, tss + e)` on the `+` strand and `[tss - e + 1, tss - s + 1)` on the `-` strand; minus-strand rows are reversed


| Folder | Window (TSS-relative) | Width |
|---|---|---|
| `2000_tss` | `[-1000, +1000)` | 2,000 bp |
| `upstream_1000_tss` | `[-1000, -100)` | 900 bp |
| `downstream_1000_tss` | `[+100, +1000)` | 900 bp |


# Method

**Step 0: expression bins** (`00_prepare_expression_bins.R`)

Protein-coding Ensembl-canonical genes get a mean TPM pooled over the twelve RNA samples (LPS 0, 5, 10, 15 min) and a quartile bin `Q1_low` to `Q4_high`. `not_expressed` genes are excluded. The bins are fixed and shared by all four Fiber-seq samples.

All Python processing stages below (sample, ACF, clustering, tables, footprints, NRL) are functions of one script, `01_process_tss_molecules.py`, run in order in one process per region.

**Stage `sample`: sample molecules**

TSS = canonical BED start, a 1 bp interval (`/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed`). A read is eligible when it fully spans the strand-oriented genomic window of the region. Each physical read is assigned one gene and bin, and each bin contributes 2,500 reads chosen by an independent seeded hash (10,000 in total). Binary rows are built in genomic order and reversed for `-` genes; a check confirms that the column for offset 0 is the genomic TSS base on both strands.

**Stage `acf`: per-read ACF** (parent `../03_compute_autocorrelations.py`, `autocorrelations()`)

$ACF(k) = \sum_t (x_t-\bar x)(x_{t+k}-\bar x) / (N \cdot var(x))$

on the unbinned per-base 0/1 signal at every integer lag 0 … width − 1 (1 bp resolution; no smoothing, binning, tapering or FFT truncation). Zero-variance reads have undefined ACFs.

**Stage `cluster`: joint Leiden clustering** (parent `../04_cluster_autocorrelations.py`, `cluster_profiles()`)

All four expression bins are clustered together: PCA (50 PCs) → correlation-distance kNN (10 neighbors) → Leiden (resolution 0.4, seed 0) → UMAP for display only. Reads that cannot be clustered keep the label `Unclustered` in the tables and are left out of the cluster figures.

**Stage `tables`: plot tables** (parent `../05_summarize_autocorrelations.py`, `repeat_peak()`)

Mean/median ACF per cluster and per expression bin, cluster × bin composition in both directions, cluster summary with the descriptive 140–250 bp positive peak, and the heatmap row order.

**Stage `footprints`: single-molecule footprints**

Match FiberHMM BED12 records by sample, chromosome and read ID; use the exact
sampled read span to resolve multiple records. Unique trimmed records are accepted.
Retain nucleosomes >90 bp and TF footprints <60 bp; clip genomic calls to each
window, then map their included endpoints with the sampler's coordinate helper.
The category matrix retains the original rows and transcriptional orientation;
m6A comes directly from the saved binary matrix. White = no call, dark grey =
nucleosome, orange = TF, purple = m6A (highest precedence). Checks enforce
base-for-base m6A identity and mirrored minus-strand footprint intervals.

**Stage `nrl`: per-molecule NRL**

NRL = lag of the first positive local maximum of the raw parent ACF in 120–300 bp (lags ≥ 60; SAMOSA secondary-peak scan, Abdulhay et al. 2020 eLife), with `scipy.signal.find_peaks` and a prominence threshold **calibrated per region** rather than a fixed peak height: up to 2,000 sampled molecules are drawn (seeded), each molecule's m6A calls are shuffled among the A/T positions of its own strand-oriented reference window (hg38; G/C positions stay 0, the call count is unchanged), the parent ACF of each shuffled molecule is computed, and the threshold is the 95th percentile of the largest positive in-band peak prominence over these null molecules (`tables/nrl_prominence_calibration.tsv`, `tables/nrl_null_prominences.tsv`). Group mean-ACF peaks per expression bin and per cluster are tested against 200 group-sized averages of shuffled ACFs drawn (with replacement) from that group's own molecules, each shuffled five times (`tables/group_peak_tests.tsv`).

**Step 2: per-region figures** (`02_plot_tss_autocorrelation.R`)

Base R; seventeen PDFs per region.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = TRUE, message = TRUE, out.width = "100%")
options(scipen = 999)
code_dir <- normalizePath(".", mustWork = TRUE)
output <- normalizePath(params$output_dir, mustWork = TRUE)
windows <- c("2000_tss", "upstream_1000_tss", "downstream_1000_tss")
unknown <- setdiff(params$windows, windows)
if (length(unknown)) stop("Unknown windows: ", paste(unknown, collapse = ", "))
window_dirs <- setNames(file.path(output, params$windows), params$windows)
```

## Run the pipeline

Submit all three regions as one array job (one task per region). Each task runs
`00_prepare_expression_bins.R` only if the expression table is absent, then
`01_process_tss_molecules.py` and `02_plot_tss_autocorrelation.R`, and publishes
the results atomically. Knitting never runs the pipeline.

```bash
mkdir -p /project/spott/cshan/fiber-seq/macrophage_project/auto_correlation/tss/logs
sbatch /project/spott/cshan/fiber-seq/code/clustering_methods/auto_correlation/tss/03_run_tss_autocorrelation.sh
```

```{r check-results, include=FALSE}
missing <- window_dirs[!file.exists(file.path(window_dirs, "tables", "cluster_summary.tsv"))]
if (length(missing)) stop("No results for: ", paste(names(missing), collapse = ", "),
  ". Submit 03_run_tss_autocorrelation.sh first.")
```

```{r gallery-helper, include=FALSE}
# Embed the published PDFs as PNG previews (fonts and converter as in 06_autocorrelation.Rmd).
pdftopng <- "/software/xpdf-4.05-el8-x86_64/bin/pdftopng"
xpdfrc <- tempfile("xpdfrc-")
sans <- "/usr/share/fonts/liberation-sans/LiberationSans-"
mono <- "/usr/share/fonts/dejavu/DejaVuSansMono"
writeLines(c(
  paste0("fontFile Helvetica ", sans, "Regular.ttf"),
  paste0("fontFile Helvetica-Bold ", sans, "Bold.ttf"),
  paste0("fontFile Helvetica-Oblique ", sans, "Italic.ttf"),
  paste0("fontFile Helvetica-BoldOblique ", sans, "BoldItalic.ttf"),
  paste0("fontFile Times-Roman ", sans, "Regular.ttf"),
  paste0("fontFile Times-Bold ", sans, "Bold.ttf"),
  paste0("fontFile Times-Italic ", sans, "Italic.ttf"),
  paste0("fontFile Times-BoldItalic ", sans, "BoldItalic.ttf"),
  paste0("fontFile Courier ", mono, ".ttf"),
  paste0("fontFile Courier-Bold ", mono, "-Bold.ttf"),
  paste0("fontFile Courier-Oblique ", mono, "-Oblique.ttf"),
  paste0("fontFile Courier-BoldOblique ", mono, "-BoldOblique.ttf"),
  "fontFile Symbol /usr/share/fonts/dejavu/DejaVuSans.ttf",
  "fontFile ZapfDingbats /usr/share/fonts/dejavu/DejaVuSans.ttf"), xpdfrc)

inline_png <- function(path, label, dpi = 120) {
  if (!file.exists(path) || file.info(path)$size == 0) {
    cat("\n\n*Missing figure:* `", path, "`\n\n", sep = "")
    return(invisible(NULL))
  }
  prefix <- tempfile("fig-")
  status <- system2(pdftopng, c("-cfg", shQuote(xpdfrc), "-q", "-r", dpi, "-f", 1, "-l", 1,
    shQuote(path), shQuote(prefix)))
  png <- paste0(prefix, "-000001.png")
  stopifnot(status == 0, file.exists(png))
  on.exit(unlink(png))
  cat(as.character(htmltools::tags$figure(
    htmltools::tags$img(src = base64enc::dataURI(file = png, mime = "image/png"),
      alt = label, loading = "lazy", style = "display:block;max-width:100%;height:auto;"),
    htmltools::tags$figcaption(label))), "\n\n")
  invisible(NULL)
}

# All seventeen per-region figures, grouped into `##` subsections.
figures <- list(
  "Single-molecule footprints" = c("08_single_fiber_footprints" = "Strand-oriented single-molecule footprints, in ACF heatmap order"),
  "ACF heatmap: all clustered molecules" = c("01_acf_heatmap_all" = "ACF heatmap by global Leiden class"),
  "ACF heatmaps by expression bin" = c("03_acf_heatmaps_by_expression" = "ACF heatmaps by expression bin"),
  "Mean ACF by cluster" = c("02_mean_acf_by_cluster" = "Mean ACF by global Leiden class"),
  "Mean ACF by expression bin (from 25 bp)" = c("07_mean_acf_by_expression" = "Mean ACF by expression bin, lags from 25 bp"),
  "Cluster composition across expression bins" = c(
    "04_cluster_fraction_within_expression" = "Cluster fraction within each expression bin",
    "04b_expression_fraction_within_cluster" = "Expression-bin fraction within each cluster"),
  "m6A counts by cluster" = c(
    "05_m6a_by_cluster_and_expression" = "m6A counts by cluster within each expression bin",
    "05b_m6a_by_cluster" = "m6A counts by cluster, all bins"),
  "UMAP" = c(
    "06_umap_by_cluster" = "UMAP by global Leiden class",
    "06b_umap_by_expression" = "UMAP by expression bin",
    "06c_umap_by_m6a" = "UMAP by m6A count"),
  "Per-molecule NRL" = c(
    "08_nrl_metrics_by_expression" = "NRL by expression bin",
    "08b_nrl_metrics_by_cluster" = "NRL by Leiden class",
    "09_nrl_histogram_by_cluster" = "Per-molecule NRL histograms by Leiden class (SAMOSA Fig. 3D style)",
    "10_nrl_peak_fraction" = "Fraction of molecules with a detected NRL peak, by bin and by class"),
  "Null peak prominence and calibrated threshold" = c(
    "11_null_prominence_distribution" = "Largest in-band peak prominence of A/T-shuffled molecules; dashed line = min_prominence"))

show_region <- function(window) {
  if (!window %in% names(window_dirs)) {
    cat("\n\nWindow `", window, "` is not in `params$windows`.\n\n", sep = "")
    return(invisible(NULL))
  }
  for (group in names(figures)) {
    cat("\n\n## ", group, "\n\n", sep = "")
    for (name in names(figures[[group]])) {
      inline_png(file.path(window_dirs[[window]], "plots", paste0(name, ".pdf")),
        paste(window, figures[[group]][[name]], sep = " — "))
    }
  }
}

read_table <- function(path) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE,
  na.strings = c("NA", "NaN", "nan", ""), quote = "")

show_table <- function(name, label, digits = 3, columns = NULL, filter = NULL) {
  for (window in names(window_dirs)) {
    cat("\n\n### ", window, "\n\n", sep = "")
    path <- file.path(window_dirs[[window]], "tables", paste0(name, ".tsv"))
    if (!file.exists(path)) { cat("*Missing table:* `", path, "`\n\n", sep = ""); next }
    tab <- read_table(path)
    if (!is.null(filter)) tab <- filter(tab)
    if (!is.null(columns)) tab <- tab[, intersect(columns, names(tab)), drop = FALSE]
    print(knitr::kable(tab, digits = digits, caption = paste(window, label, sep = " — ")))
  }
}
```

<style>
.main-container { max-width: 1700px; padding: 0 24px; }
table { display: block; overflow-x: auto; }
</style>

# Region 1: 2 kb centred on the TSS (`2000_tss`)

Window `[-1000, +1000)`, 2,000 bp, strand-oriented; lags 0–1999.

```{r region-2000, results='asis'}
show_region("2000_tss")
```

# Region 2: upstream flank, −1000 to −100 (`upstream_1000_tss`)

Window `[-1000, -100)`, 900 bp upstream of the TSS in the transcriptional
direction, stopping 100 bp short of it; lags 0–899.

```{r region-upstream, results='asis'}
show_region("upstream_1000_tss")
```

# Region 3: downstream flank, +100 to +1000 (`downstream_1000_tss`)

Window `[+100, +1000)`, 900 bp downstream of the TSS in the transcriptional
direction, starting 100 bp past it; lags 0–899.

```{r region-downstream, results='asis'}
show_region("downstream_1000_tss")
```

# Result tables

## Eligible pool counts

```{r table-pool, results='asis'}
show_table("eligible_pool_counts", "Unique eligible molecules per expression bin")
```

## Cluster summaries

```{r table-cluster-summary, results='asis'}
show_table("cluster_summary", "Cluster summaries")
```

## Cluster composition

```{r table-composition, results='asis'}
show_table("cluster_composition", "Cluster × expression-bin composition")
```

## Unclustered molecules by expression bin and timepoint

```{r table-unclustered, results='asis'}
show_table("unclustered_counts", "Zero-variance and otherwise unclustered molecules per bin and timepoint")
```

## Per-molecule NRL by expression bin

```{r table-nrl-bin, results='asis'}
show_table("nrl_summary_by_bin", "NRL by expression bin", columns = c("expr_bin", "n", "n_zero_variance",
  "n_unclustered", "n_with_nrl_peak", "fraction_with_nrl_peak", "median_nrl_bp", "iqr_low_nrl_bp",
  "iqr_high_nrl_bp", "mad_nrl_bp", "group_peak_lag_bp", "group_peak_height", "group_peak_prominence",
  "group_peak_pvalue", "min_prominence"))
```

## Per-molecule NRL by cluster

```{r table-nrl-cluster, results='asis'}
show_table("nrl_summary_by_cluster", "NRL by Leiden class", columns = c("cluster", "n", "n_zero_variance",
  "n_unclustered", "n_with_nrl_peak", "fraction_with_nrl_peak", "median_nrl_bp", "iqr_low_nrl_bp",
  "iqr_high_nrl_bp", "mad_nrl_bp", "group_peak_lag_bp", "group_peak_height", "group_peak_prominence",
  "group_peak_pvalue"))
```

## Group mean-ACF peak tests

```{r table-group-tests, results='asis'}
show_table("group_peak_tests", "Observed in-band peak of each group's mean ACF versus group-sized null means", digits = 4)
```

## Prominence calibration

```{r table-calibration, results='asis'}
show_table("nrl_prominence_calibration", "Calibrated min_prominence (A/T-shuffled null) for the fixed window", digits = 4)
```
