---
title: "Macrophage TSS Fiber-seq autocorrelation"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    mathjax: null
params:
  output_dir: "../../../../macrophage_project/auto_correlation/tss"
  windows: ["2000_tss", "upstream_1000_tss", "downstream_1000_tss"]
  test: false
---

# Analysis and parameters

This notebook shows the results of the numbered scripts in this folder, in the
same way `../06_autocorrelation.Rmd` does for the LCL regions. Each TSS region is
one SLURM array task of `06_run_tss_autocorrelation.sh` and writes to its own
folder under `params$output_dir`; the cross-region sliding-window figures are
written by the dependent job `08_plot_sliding_combined.R` to `combined/`.
With `params$test = TRUE` the `test/` subfolders of the chromosome-22 test run
are shown instead.

Offsets are TSS-relative and strand-oriented (transcriptional): negative =
upstream, positive = downstream, end exclusive. For the interval `[s, e)` and the
zero-based TSS base `tss` (canonical BED start + 10), the genomic window is
`[tss + s, tss + e)` on the `+` strand and `[tss - e + 1, tss - s + 1)` on the
`-` strand; minus-strand rows are reversed so that column `j` is offset `s + j`.

| Folder | Window (TSS-relative) | Width | Sliding windows (500 bp, step 100) |
|---|---|---|---|
| `2000_tss` | `[-1000, +1000)` | 2,000 bp | 16 (6 overlap the NDR core `[-100, +100)`) |
| `upstream_1000_tss` | `[-1000, -100)` | 900 bp | 5 |
| `downstream_1000_tss` | `[+100, +1000)` | 900 bp | 5 |

The ACF is invariant to sequence reversal, so per-read ACF values change only
through which reads and bases fall in each window; heatmaps, sliding windows
and other position-based outputs depend on the orientation.

# Method

**Step 0: expression bins** (`00_prepare_expression_bins.R`)

Protein-coding Ensembl-canonical genes get a mean TPM pooled over the twelve RNA samples (LPS 0, 5, 10, 15 min) and a quartile bin `Q1_low` to `Q4_high`. `not_expressed` genes are excluded. The bins are fixed and shared by all four Fiber-seq samples.

**Step 1: sample molecules** (`01_sample_tss_molecules.py`)

TSS = canonical BED start + 10 (`/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed`). A read is eligible when it fully spans the strand-oriented genomic window of the region. Each physical read is assigned one gene and bin, and each bin contributes 2,500 reads chosen by an independent seeded hash (10,000 in total; 100 per bin on chromosome 22 in `--test`). Binary rows are built in genomic order and reversed for `-` genes; a check confirms that the column for offset 0 is the genomic TSS base on both strands.

**Step 2: per-read ACF** (`02_compute_tss_autocorrelations.py`, parent `autocorrelations()`)

$ACF(k) = \sum_t (x_t-\bar x)(x_{t+k}-\bar x) / (N \cdot var(x))$

on the unbinned per-base 0/1 signal at every integer lag 0 … width − 1 (1 bp resolution; no smoothing, binning, tapering or FFT truncation). Zero-variance reads have undefined ACFs.

**Step 3: joint Leiden clustering** (`03_leiden_cluster_autocorrelations.py`, parent `cluster_profiles()`)

All four expression bins are clustered together: PCA (50 PCs) → correlation-distance kNN (10 neighbors) → Leiden (resolution 0.4, seed 0) → UMAP for display only. Reads that cannot be clustered keep the label `Unclustered` in the tables and are left out of the cluster figures.

**Step 4: plot tables** (`04_prepare_plot_tables.py`)

Mean/median ACF per cluster and per expression bin, cluster × bin composition in both directions, cluster summary with the descriptive 140–250 bp positive peak, and the heatmap row order.

**Step 4b: single-molecule footprints** (`04b_extract_tss_footprints.py`)

Match FiberHMM BED12 records by sample, chromosome and read ID; use the exact
sampled read span to resolve multiple records. Unique trimmed records are accepted.
Retain nucleosomes >90 bp and TF footprints <60 bp; clip genomic calls to each
window, then map their included endpoints with the sampler's coordinate helper.
The category matrix retains the original rows and transcriptional orientation;
m6A comes directly from the saved binary matrix. White = no call, dark grey =
nucleosome, orange = TF, purple = m6A (highest precedence). Checks enforce
base-for-base m6A identity and mirrored minus-strand footprint intervals.

**Step 4b: per-molecule NRL and regularity** (`04b_molecule_nrl.py`, `tss_nrl.py`)

NRL = lag of the first positive local maximum of the raw ACF in 120–300 bp (lags ≥ 60; SAMOSA secondary-peak scan, Abdulhay et al. 2020 eLife), with `scipy.signal.find_peaks` and a prominence threshold **calibrated per window length**: the 95th percentile of the largest in-band peak prominence obtained from position-shuffled copies of the same molecules (`tables/nrl_prominence_calibration.tsv`). Regularity follows Baldi et al. 2018 Mol Cell: on the unbiased rescaling `ACF(k)·N/(N−k)` restricted to lags ≤ N/2 (decay metrics only), a damped cosine `A·exp(−k/λ)·cos(2πk/P+φ)+c` gives the decay length λ, and the damping lag is the lag after which the one-period rolling |ACF| envelope stays below 0.05.

**Step 4c: sliding windows** (`04c_sliding_windows.py`)

500 bp windows stepped by 100 bp over each region's own span and molecules (no resampling), parent ACF at lags 0–499, the step-4b metrics per (read, window), the NDR-core overlap flag (Clarkson et al. 2019 NAR), and an A/T reference-sequence control of the same windows.

**Step 5: per-region figures** (`05_plot_tss_autocorrelation.R`); **step 8: combined sliding figures and tests** (`08_plot_sliding_combined.R`)

Base R. Combined panels show read-level statistics per window with bootstrap 95% CIs over genes; upstream versus downstream is an unpaired Wilcoxon test on gene-level medians, mirror windows within the 2 kb span are paired by read, BH-corrected within each family.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = TRUE, message = TRUE, out.width = "100%")
options(scipen = 999)
code_dir <- normalizePath(".", mustWork = TRUE)
output <- normalizePath(params$output_dir, mustWork = TRUE)
windows <- c("2000_tss", "upstream_1000_tss", "downstream_1000_tss")
unknown <- setdiff(params$windows, windows)
if (length(unknown)) stop("Unknown windows: ", paste(unknown, collapse = ", "))
suffix <- if (isTRUE(params$test)) "test" else ""
window_dirs <- setNames(sub("/$", "", file.path(output, params$windows, suffix)), params$windows)
combined_dir <- sub("/$", "", file.path(output, "combined", suffix))
```

## Run the pipeline

Submit all three regions as one array job (one task per region); the lowest
task submits the combined job after the array. Knitting never runs the pipeline.

```bash
mkdir -p /project/spott/cshan/fiber-seq/macrophage_project/auto_correlation/tss/logs
sbatch /project/spott/cshan/fiber-seq/code/clustering_methods/auto_correlation/tss/06_run_tss_autocorrelation.sh
# chromosome-22 test: add --test and knit with params$test = TRUE
```

```{r check-results, include=FALSE}
missing <- window_dirs[!file.exists(file.path(window_dirs, "tables", "cluster_summary.tsv"))]
if (length(missing)) stop("No results for: ", paste(names(missing), collapse = ", "),
  ". Submit 06_run_tss_autocorrelation.sh first.")
have_combined <- file.exists(file.path(combined_dir, "tables", "sliding_tests.tsv"))
```

```{r gallery-helper, include=FALSE}
# Embed the published PDFs as PNG previews (fonts and converter as in 06_autocorrelation.Rmd).
pdftopng <- "/software/xpdf-4.05-el8-x86_64/bin/pdftopng"
xpdfrc <- tempfile("xpdfrc-")
sans <- "/usr/share/fonts/liberation-sans/LiberationSans-"
mono <- "/usr/share/fonts/dejavu/DejaVuSansMono"
writeLines(c(
  paste0("fontFile Helvetica ", sans, "Regular.ttf"),
  paste0("fontFile Helvetica-Bold ", sans, "Bold.ttf"),
  paste0("fontFile Helvetica-Oblique ", sans, "Italic.ttf"),
  paste0("fontFile Helvetica-BoldOblique ", sans, "BoldItalic.ttf"),
  paste0("fontFile Times-Roman ", sans, "Regular.ttf"),
  paste0("fontFile Times-Bold ", sans, "Bold.ttf"),
  paste0("fontFile Times-Italic ", sans, "Italic.ttf"),
  paste0("fontFile Times-BoldItalic ", sans, "BoldItalic.ttf"),
  paste0("fontFile Courier ", mono, ".ttf"),
  paste0("fontFile Courier-Bold ", mono, "-Bold.ttf"),
  paste0("fontFile Courier-Oblique ", mono, "-Oblique.ttf"),
  paste0("fontFile Courier-BoldOblique ", mono, "-BoldOblique.ttf"),
  "fontFile Symbol /usr/share/fonts/dejavu/DejaVuSans.ttf",
  "fontFile ZapfDingbats /usr/share/fonts/dejavu/DejaVuSans.ttf"), xpdfrc)

inline_png <- function(path, label, dpi = 120) {
  if (!file.exists(path) || file.info(path)$size == 0) {
    cat("\n\n*Missing figure:* `", path, "`\n\n", sep = "")
    return(invisible(NULL))
  }
  prefix <- tempfile("fig-")
  status <- system2(pdftopng, c("-cfg", shQuote(xpdfrc), "-q", "-r", dpi, "-f", 1, "-l", 1,
    shQuote(path), shQuote(prefix)))
  png <- paste0(prefix, "-000001.png")
  stopifnot(status == 0, file.exists(png))
  on.exit(unlink(png))
  cat(as.character(htmltools::tags$figure(
    htmltools::tags$img(src = base64enc::dataURI(file = png, mime = "image/png"),
      alt = label, loading = "lazy", style = "display:block;max-width:100%;height:auto;"),
    htmltools::tags$figcaption(label))), "\n\n")
  invisible(NULL)
}

# All sixteen per-region figures, grouped into `##` subsections.
figures <- list(
  "Single-molecule footprints" = c("08_single_fiber_footprints" = "Strand-oriented single-molecule footprints, in ACF heatmap order"),
  "ACF heatmap: all clustered molecules" = c("01_acf_heatmap_all" = "ACF heatmap by global Leiden class"),
  "ACF heatmaps by expression bin" = c("03_acf_heatmaps_by_expression" = "ACF heatmaps by expression bin"),
  "Mean ACF by cluster" = c("02_mean_acf_by_cluster" = "Mean ACF by global Leiden class"),
  "Mean ACF by expression bin (from 25 bp)" = c("07_mean_acf_by_expression" = "Mean ACF by expression bin, lags from 25 bp"),
  "Cluster composition across expression bins" = c(
    "04_cluster_fraction_within_expression" = "Cluster fraction within each expression bin",
    "04b_expression_fraction_within_cluster" = "Expression-bin fraction within each cluster"),
  "m6A counts by cluster" = c(
    "05_m6a_by_cluster_and_expression" = "m6A counts by cluster within each expression bin",
    "05b_m6a_by_cluster" = "m6A counts by cluster, all bins"),
  "UMAP" = c(
    "06_umap_by_cluster" = "UMAP by global Leiden class",
    "06b_umap_by_expression" = "UMAP by expression bin",
    "06c_umap_by_m6a" = "UMAP by m6A count"),
  "Per-molecule NRL and regularity" = c(
    "08_nrl_metrics_by_expression" = "NRL, decay length and damping lag by expression bin",
    "08b_nrl_metrics_by_cluster" = "NRL, decay length and damping lag by Leiden class",
    "09_nrl_histogram_by_cluster" = "Per-molecule NRL histograms by Leiden class (SAMOSA Fig. 3D style)",
    "10_nrl_peak_fraction" = "Fraction of molecules with a detected NRL peak, by bin and by class"))

combined_figures <- list(
  "Median NRL" = c("01_sliding_nrl" = "Median NRL per sliding window",
    "01_sliding_nrl_by_timepoint" = "Median NRL per sliding window, by LPS timepoint"),
  "Fraction with a detected NRL peak" = c("02_sliding_peak_fraction" = "Fraction of molecules with an NRL peak",
    "02_sliding_peak_fraction_by_timepoint" = "Fraction with an NRL peak, by LPS timepoint"),
  "Decay length and damping lag" = c("03_sliding_decay_damping" = "Median decay length and damping lag",
    "03_sliding_decay_damping_by_timepoint" = "Median decay length and damping lag, by LPS timepoint"),
  "m6A density" = c("04_sliding_m6a_density" = "Mean m6A density per sliding window",
    "04_sliding_m6a_density_by_timepoint" = "Mean m6A density, by LPS timepoint"),
  "Mean ACF heatmaps per window" = c("05_sliding_acf_heatmap_2kb" = "Mean ACF (lags 0-499) x window centre, 2 kb span",
    "05b_sliding_acf_heatmap_one_sided" = "Mean ACF (lags 0-499) x window centre, one-sided regions"))

show_region <- function(window) {
  if (!window %in% names(window_dirs)) {
    cat("\n\nWindow `", window, "` is not in `params$windows`.\n\n", sep = "")
    return(invisible(NULL))
  }
  for (group in names(figures)) {
    cat("\n\n## ", group, "\n\n", sep = "")
    for (name in names(figures[[group]])) {
      inline_png(file.path(window_dirs[[window]], "plots", paste0(name, ".pdf")),
        paste(window, figures[[group]][[name]], sep = " — "))
    }
  }
}

show_combined <- function() {
  if (!have_combined) {
    cat("\n\nNo combined results in `", combined_dir, "`; run `08_plot_sliding_combined.R` after all three regions.\n\n", sep = "")
    return(invisible(NULL))
  }
  for (group in names(combined_figures)) {
    cat("\n\n## ", group, "\n\n", sep = "")
    for (name in names(combined_figures[[group]])) {
      inline_png(file.path(combined_dir, "plots", paste0(name, ".pdf")),
        paste("combined", combined_figures[[group]][[name]], sep = " — "))
    }
  }
}

read_table <- function(path) read.delim(path, check.names = FALSE, stringsAsFactors = FALSE,
  na.strings = c("NA", "NaN", "nan", ""), quote = "")

show_table <- function(name, label, digits = 3, columns = NULL, filter = NULL) {
  for (window in names(window_dirs)) {
    cat("\n\n### ", window, "\n\n", sep = "")
    path <- file.path(window_dirs[[window]], "tables", paste0(name, ".tsv"))
    if (!file.exists(path)) { cat("*Missing table:* `", path, "`\n\n", sep = ""); next }
    tab <- read_table(path)
    if (!is.null(filter)) tab <- filter(tab)
    if (!is.null(columns)) tab <- tab[, intersect(columns, names(tab)), drop = FALSE]
    print(knitr::kable(tab, digits = digits, caption = paste(window, label, sep = " — ")))
  }
}
```

<style>
.main-container { max-width: 1700px; padding: 0 24px; }
table { display: block; overflow-x: auto; }
</style>

# Region 1: 2 kb centred on the TSS (`2000_tss`)

Window `[-1000, +1000)`, 2,000 bp, strand-oriented; lags 0–1999.

```{r region-2000, results='asis'}
show_region("2000_tss")
```

# Region 2: upstream flank, −1000 to −100 (`upstream_1000_tss`)

Window `[-1000, -100)`, 900 bp upstream of the TSS in the transcriptional
direction, stopping 100 bp short of it; lags 0–899.

```{r region-upstream, results='asis'}
show_region("upstream_1000_tss")
```

# Region 3: downstream flank, +100 to +1000 (`downstream_1000_tss`)

Window `[+100, +1000)`, 900 bp downstream of the TSS in the transcriptional
direction, starting 100 bp past it; lags 0–899.

```{r region-downstream, results='asis'}
show_region("downstream_1000_tss")
```

# Sliding windows across regions (`combined`)

Window centre relative to the TSS on the x axis (upstream left), one line per
expression bin, bootstrap 95% CI over genes. Pink shading marks 2 kb-span
windows overlapping the NDR core `[-100, +100)`; grey marks the gap between the
one-sided regions. Dashed grey: A/T reference control.

```{r combined, results='asis'}
show_combined()
```

# Result tables

## Eligible pool counts

```{r table-pool, results='asis'}
show_table("eligible_pool_counts", "Unique eligible molecules per expression bin")
```

## Cluster summaries

```{r table-cluster-summary, results='asis'}
show_table("cluster_summary", "Cluster summaries")
```

## Cluster composition

```{r table-composition, results='asis'}
show_table("cluster_composition", "Cluster × expression-bin composition")
```

## Per-molecule NRL by expression bin

```{r table-nrl-bin, results='asis'}
show_table("nrl_summary_by_bin", "NRL and regularity by expression bin", columns = c("expr_bin", "n",
  "n_with_nrl_peak", "fraction_with_nrl_peak", "median_nrl_bp", "iqr_low_nrl_bp", "iqr_high_nrl_bp",
  "median_decay_length_bp", "median_damping_lag_bp", "median_peak_ratio", "fraction_fit_converged", "min_prominence"))
```

## Per-molecule NRL by cluster

```{r table-nrl-cluster, results='asis'}
show_table("nrl_summary_by_cluster", "NRL and regularity by Leiden class", columns = c("cluster", "n",
  "n_with_nrl_peak", "fraction_with_nrl_peak", "median_nrl_bp", "iqr_low_nrl_bp", "iqr_high_nrl_bp",
  "median_decay_length_bp", "median_damping_lag_bp", "median_peak_ratio", "fraction_fit_converged"))
```

## Prominence calibration

```{r table-calibration, results='asis'}
show_table("nrl_prominence_calibration", "Calibrated min_prominence for the fixed window", digits = 4)
show_table("sliding_prominence_calibration", "Calibrated min_prominence for the 500 bp sliding windows", digits = 4)
```

## Sliding-window summaries (all timepoints pooled)

```{r table-sliding, results='asis'}
show_table("sliding_summary", "Per bin × window (timepoint = all)", filter = function(tab) tab[tab$timepoint == "all", ],
  columns = c("expr_bin", "win_start", "win_end", "overlaps_ndr_core", "n", "n_genes", "fraction_with_nrl_peak",
    "median_nrl_bp", "iqr_low_nrl_bp", "iqr_high_nrl_bp", "median_decay_length_bp", "median_damping_lag_bp",
    "mean_m6a_density"))
```

## A/T sequence control

```{r table-at, results='asis'}
show_table("sliding_at_control", "A/T indicator ACF per window (all bins)", filter = function(tab) tab[tab$expr_bin == "all", ])
```

## Sliding-window tests

```{r table-tests, results='asis'}
if (have_combined) {
  tests <- read_table(file.path(combined_dir, "tables", "sliding_tests.tsv"))
  print(knitr::kable(tests, digits = 4, caption = "Upstream vs downstream (unpaired, gene-level medians) and 2 kb mirror windows (paired by read); BH within each family"))
} else cat("\n\nNo combined results available.\n\n")
```
