---
title: "Topic modelling on promoters"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(rtracklayer)
library(data.table)
library(Rsamtools)
library(cowplot)
library(fastTopics)
library(NNLM)
library(gplots)
library(DescTools)   
library(vegan)       
library(stringr)
library(cowplot)
```

# Functions

All functions adapted from Kevin Luo's `process_fiberseq_data.R` and
`topic_model_utils.R`.

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
```

# Settings

```{r settings}
options(scipen = 999)

# Sample metadata, defined inline (was: macrophage_FIRE/extract/sample/sample_table.csv).
# `sample_name` = which sample/file each read comes from; `timepoint` = LPS time (min).
# assemble_region_met_data() tags every pooled read with `sample_name`, so joining
# sample_name -> timepoint labels each read's origin for grouping after topic modelling.
# Add the untreated R replicates as more rows once their BAMs are extracted.
sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

# original:
# ft_result_dir <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

# original:
# output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model"
output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/promoters"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

ncore <- 4
n_topics <- 6
topic_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00",
                  "#CC79A7", "#999999")

# palette for read clusters (from Kevin Luo's plots.R)
colors_25 <- c(
  "dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00",
  "black", "gold1", "skyblue2", "#FB9A99", "palegreen2",
  "#CAB2D6", "#FDBF6F", "gray70", "khaki2", "maroon",
  "orchid1", "deeppink1", "blue1", "steelblue4", "darkturquoise",
  "green1", "yellow4", "yellow3", "darkorange4", "brown"
)

set.seed(123)
```

# m6A + CpG extraction from BAMs

The topic model clusters reads by their per-read m6A pattern, so for every sample
it needs that sample's m6A (and CpG) calls extracted from the BAM and split into
one tabix-indexed bed per chromosome - the layout `assemble_region_met_data()`
reads:
`<ft_result_dir>/<sample>/extracted_results/{m6a,cpg}_by_chr/<sample>.ft_extracted_{m6a,cpg}.<chr>.bed.gz`.

`ft extract` reads each read's alignment span from the BAM, so a position a read
covers but is unmethylated at is scored `0` (protected), not `NA` (missing) - the
reason we use `ft extract` here rather than the FiberHMM `recalled_*` beds, which
store only each modification's call-span.


# Promoter regions

Inflammation-related genes (THP-1). Each region is TSS +/- 1000 bp. The TSS is
the dominant THP-1 CAGE peak per gene; genes without a CAGE peak (IL1A, PTGS2,
CXCL2, IL6, RELB) fall back to the GENCODE v49 annotated TSS. Both BEDs are
keyed on Ensembl gene IDs, so gene symbols are mapped explicitly (IL8 = CXCL8
in GENCODE). `rs2836882` is a motif-disrupting, inflammation-associated SNP
locus expected to be heterozygous in THP-1 - one allele open, one closed; it
has no TSS, so it keeps its original explicit interval.

```{r regions}
# original explicit genomic intervals:
# promoters <- tibble::tribble(
#   ~gene,        ~chr,     ~start,      ~end,
#   "IL1B",       "chr2",   112836323,   112837385,
#   "TNF",        "chr6",   31575265,    31575824,
#   "IL1A",       "chr2",   112793233,   112794969,
#   "PTGS2",      "chr1",   186679985,   186681163,
#   "IL8",        "chr4",   73740006,    73741332,
#   "CXCL2",      "chr4",   74098863,    74099651,
#   "IL6",        "chr7",   22726376,    22727094,
#   "RELB",       "chr19",  45001011,    45002417,
#   "rs2836882",  "chr21",  39094468,    39094712
# )

cage_tss_file    <- "/project/spott/cshan/fiber-seq/macrophage_project/annotations/fantom5_thp1/THP1.consensus.CAGE_peaks.dominantPerGene.bed.gz"
gencode_tss_file <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"
tss_flank <- 1000

# gene symbol -> Ensembl gene ID (both TSS BEDs are keyed on ENSG; IL8 = CXCL8)
genes <- tibble::tribble(
  ~gene,    ~gene_id,
  "IL1B",   "ENSG00000125538",
  "TNF",    "ENSG00000232810",
  "IL1A",   "ENSG00000115008",
  "PTGS2",  "ENSG00000073756",
  "IL8",    "ENSG00000169429",
  "CXCL2",  "ENSG00000081041",
  "IL6",    "ENSG00000136244",
  "RELB",   "ENSG00000104856"
)

# both BEDs are single-base features (end = 1-based TSS position), one row per
# gene; strip the ENSG version suffix so the two sources match
read_tss_bed <- function(file) {
  data.table::fread(file, header = FALSE, select = c(1, 3, 4),
                    col.names = c("chr", "tss", "gene_id")) %>%
    dplyr::mutate(gene_id = sub("\\..*$", "", gene_id))
}

cage_tss    <- read_tss_bed(cage_tss_file)
gencode_tss <- read_tss_bed(gencode_tss_file)

promoters <- genes %>%
  dplyr::left_join(cage_tss,    by = "gene_id") %>%
  dplyr::left_join(gencode_tss, by = "gene_id", suffix = c("_cage", "_gencode")) %>%
  dplyr::mutate(
    tss_source = ifelse(!is.na(tss_cage), "CAGE", "gencode"),
    chr        = dplyr::coalesce(chr_cage, chr_gencode),
    tss        = dplyr::coalesce(tss_cage, tss_gencode),
    start      = tss - tss_flank,
    end        = tss + tss_flank
  ) %>%
  dplyr::select(gene, chr, start, end, tss, tss_source)
stopifnot(!any(is.na(promoters$tss)))

# rs2836882 is a SNP locus, not a gene - no TSS; keep its explicit interval
promoters <- dplyr::bind_rows(
  promoters,
  tibble::tibble(gene = "rs2836882", chr = "chr21", start = 39094468L,
                 end = 39094712L, tss = NA_integer_, tss_source = "manual")
)

# save the resolved regions alongside the per-gene outputs
write.table(promoters, file.path(output_dir, "promoter_table"),
            sep = "\t", quote = FALSE, row.names = FALSE)

promoters
```

# Per-region m6A topic modelling

The m6A methylation flow (assemble -> filter -> fit -> plot -> cluster) is wrapped in `process_region_m6a()` and run once per region in the loop below.
Per-region outputs (RDS fits + PDFs) are written under `output_dir/<outname>/`, where `outname = <gene>` (one folder per gene).


`assemble_region_met_data(sample_names, ...)`: it loops over `sample_names` (the LPS timepoints), tags every read with its `sample_name`,
and `rbind`s them into one pooled `metA_mat` fit by a single topic model - so the four timepoints are **jointly** modelled, not per-timepoint. Each pooled read's origin is `rids_df_filtered$sample_name`; join it to `sample_table` for the timepoint (e.g.`sample_table$timepoint[match(rids_df_filtered$sample_name, sample_table$sample_name)]`) to group reads by origin after topic modelling.

```{r process-region-fn}
process_region_m6a <- function(gene) {
  region <- promoters %>% dplyr::filter(gene == !!gene)
  region_chr <- region$chr
  region_start <- region$start
  region_end <- region$end
  # original:
  # outname <- paste0(gene, "_", region_chr, "_", region_start, "_", region_end)
  outname <- gene
  cat("=== ", gene, ": ", region_chr, ":", region_start, "-", region_end, " ===\n", sep = "")

  dir.create(file.path(output_dir, outname), showWarnings = FALSE, recursive = TRUE)

  ## Assemble m6A data
  region_data <- assemble_region_met_data(sample_names, region_chr, region_start, region_end,
                                          ft_result_dir = ft_result_dir,
                                          verbose = TRUE)
  saveRDS(region_data, file = file.path(output_dir, outname, paste0("region_met_data_", outname, ".rds")))
  print(summary(region_data))

  ## Filter the m6A matrix
  metA_mat <- region_data$metA_mat
  cat("Dimensions of the metA matrix:", nrow(metA_mat), "x", ncol(metA_mat), "\n")
  metA_mat <- filter_met_mat(metA_mat, verbose = TRUE)
  cat("Dimensions of the metA matrix after filtering:", nrow(metA_mat), "x", ncol(metA_mat), "\n")
  keep_idx <- match(rownames(metA_mat), region_data$rids_df$RID)
  rids_df_filtered <- region_data$rids_df[keep_idx, ]

  ## Fit topic models
  ## The multinomial fit is generated and saved for reference only - everything
  ## downstream (plots, clustering, exports) uses the binomial fit.
  fit_multinom_metA <- fit_topic_model(metA_mat, k = n_topics)
  fit_binom_metA    <- fit_binomial_topic_model(metA_mat, k = n_topics, numem = 100)
  saveRDS(fit_multinom_metA, file = file.path(output_dir, outname, paste0("fit_multinom_metA_", outname, ".rds")))
  saveRDS(fit_binom_metA,    file = file.path(output_dir, outname, paste0("fit_binom_metA_",    outname, ".rds")))

  ## Plot topic model
  gg_binom <- plot_topic_model_result(
    fit_binom_metA,
    grouping = rids_df_filtered$sample_name,
    region = list(start = region_start, end = region_end),
    topic_colors = topic_colors,
    color_by = "topic",
    plot = "all",
    title = paste0(gene, " (binomial, m6A)")
  )
  print(gg_binom)
  ggsave(file.path(output_dir, outname, paste0("topic_model_binom_", outname, ".pdf")),
         gg_binom, width = 8, height = 8)

  ## Compare numbers of topics (k = 2-8)
  k_values <- 2:8
  gg_k_list <- lapply(k_values, function(k) {
    set.seed(123)
    fit_k <- fit_binomial_topic_model(metA_mat, k = k, numem = 100)
    structure_plot(fit_k$L, colors = topic_colors[1:k], topics = 1:k, gap = 5,
                   grouping = rids_df_filtered$sample_name) +
      theme_cowplot(font_size = 10) +
      ggtitle(paste0(gene, " (binomial, m6A), k = ", k)) +
      theme(plot.title = element_text(hjust = 0.5),
            axis.text.x  = element_blank(),
            axis.ticks.x = element_blank(),
            axis.title.x = element_blank())
  })
  gg_k_compare <- cowplot::plot_grid(plotlist = gg_k_list, ncol = 1)
  print(gg_k_compare)
  ggsave(file.path(output_dir, outname, paste0("topic_model_k_comparison_", outname, ".pdf")),
         gg_k_compare, width = 8, height = 14)

  ## Cluster reads by topic proportions (k = n_topics)
  fit <- fit_binom_metA   # use the k = n_topics fit for clustering
  set.seed(1)
  pca <- prcomp(fit$L)$x
  clusters <- kmeans(pca, centers = n_topics, iter.max = 100)$cluster
  clusters <- factor(clusters)
  levels(clusters) <- paste0("cluster", 1:n_topics)
  rids_df_filtered$cluster <- clusters
  cluster_colors <- colors_25[1:length(levels(clusters))]
  rids_df_filtered$cluster_color <- factor(rids_df_filtered$cluster)
  levels(rids_df_filtered$cluster_color) <- cluster_colors

  ## Structure plot grouped by cluster
  ## fastTopics' structure_plot()/pca_plot() only accept poisson_nmf_fit and
  ## multinom_topic_model_fit objects, so pass the loadings matrix directly (as the
  ## k-comparison block above does) and draw the PCA plot from `pca` by hand.
  gg_cluster_structure <- structure_plot(fit$L, topics = 1:n_topics, colors = topic_colors, gap = 5,
                                         grouping = clusters)
  print(gg_cluster_structure)
  ggsave(file.path(output_dir, outname, paste0("cluster_structure_", outname, ".pdf")),
         gg_cluster_structure, width = 8, height = 3)

  ## PCA plot colored by cluster
  gg_cluster_pca <- ggplot(data.frame(PC1 = pca[, 1], PC2 = pca[, 2], cluster = clusters),
                           aes(x = PC1, y = PC2, fill = cluster)) +
    geom_point(shape = 21, size = 2, stroke = 0.2) +
    scale_fill_manual(values = cluster_colors) +
    ggtitle(paste0(gene, " (binomial, m6A)")) +
    theme_cowplot(font_size = 10) +
    theme(plot.title = element_text(hjust = 0.5))
  print(gg_cluster_pca)
  ggsave(file.path(output_dir, outname, paste0("cluster_pca_", outname, ".pdf")),
         gg_cluster_pca, width = 5, height = 5)

  ## Heatmap of the m6A matrix ordered by cluster
  rows_idx <- order(clusters)
  metA_mat_clustered <- as.matrix(metA_mat)[rows_idx, ]
  rownames(metA_mat_clustered) <- paste0(rids_df_filtered$cluster, ".",
                                         rids_df_filtered$sample_name)[rows_idx]
  sample_colors <- as.character(rids_df_filtered$cluster_color[rows_idx])
  draw_cluster_heatmap <- function() {
    par(oma = c(1, 1, 1, 3))
    par(mar = c(5, 4, 4, 2))
    gplots::heatmap.2(x = metA_mat_clustered, Rowv = FALSE, Colv = FALSE,
                      dendrogram = "none", scale = "none", col = c("darkgray", "blue"),
                      RowSideColors = sample_colors,
                      margins = c(3, 4),
                      trace = "none", keysize = 0.5, key = FALSE)
  }
  draw_cluster_heatmap()
  pdf(file.path(output_dir, outname, paste0("cluster_heatmap_", outname, ".pdf")),
      width = 10, height = 8)
  draw_cluster_heatmap()
  dev.off()

  ## Average m6A methylation proportion per cluster
  stopifnot(identical(rownames(metA_mat), as.character(rids_df_filtered$RID)))
  met_pos <- as.numeric(sub("^A_", "", colnames(metA_mat)))
  p_list <- lapply(seq_along(levels(rids_df_filtered$cluster)), function(i) {
    cl  <- levels(rids_df_filtered$cluster)[i]
    idx <- which(rids_df_filtered$cluster == cl)
    df  <- data.frame(pos = met_pos,
                      met = Matrix::colMeans(metA_mat[idx, , drop = FALSE]))
    ggplot(df, aes(x = pos, y = met)) +
      geom_col(fill = cluster_colors[i]) +
      ylab("met prop.") + xlab("pos") + ggtitle(cl) +
      theme_cowplot(font_size = 10) +
      theme(plot.title = element_text(hjust = 0.5))
  })
  gg_met_profiles <- cowplot::plot_grid(plotlist = p_list, ncol = 1)
  print(gg_met_profiles)
  ggsave(file.path(output_dir, outname, paste0("cluster_met_profiles_", outname, ".pdf")),
         gg_met_profiles, width = 6, height = 10)

  invisible(NULL)
}
```


# examine the methylation matrix

the original matrix (X) is approximated using two smaller matrices (X = L x F)
- X: reads x genomic positions
- L: reads x topics
- F: genomic positions x topics



**L**: loadings - which topics each read is made of = how much each read belongs to each topic

         Topic 1   Topic 2   Topic 3
read1      0.80      0.15      0.05
read2      0.10      0.75      0.15
read3      0.20      0.20      0.60
read4      0.45      0.40      0.15


**F**: what each topic looks like across genomic positions = topic specific probabilities of observing m6A
 - biological pattern represented by each topic

```{r examine-outputs}
# original:
# ex_metA_rds <- read_rds("/project/spott/cshan/fiber-seq/macrophage_project/topic_model/CXCL2_chr4_74098863_74099651/fit_binom_metA_CXCL2_chr4_74098863_74099651.rds")
ex_metA_rds <- read_rds(file.path(output_dir, "CXCL2", "fit_binom_metA_CXCL2.rds"))

# original:
# ex_region_rds <- read_rds("/project/spott/cshan/fiber-seq/macrophage_project/topic_model/CXCL2_chr4_74098863_74099651/region_met_data_CXCL2_chr4_74098863_74099651.rds")
ex_region_rds <- read_rds(file.path(output_dir, "CXCL2", "region_met_data_CXCL2.rds"))
```

region: all extracted methylation data for the region, pooled across 4 LPS timepoints before any modelling - read each sample's extracted results for the region, tags every read with its sample with its sample name, and binds them into one pooled dataset
```{r region-rds}
ex_region_rds[["reads"]]
```


```{r metA-rds}
head(as.data.frame(ex_metA_rds[["F"]]))

head(as.data.frame(ex_metA_rds[["L"]]))
```

Run the full flow for every region. Inline figures all share this chunk's
`fig.*` size; the per-plot dimensions are preserved in the saved PDFs.

```{r process-all-regions, fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
for (g in promoters$gene) {
  process_region_m6a(g)
}
```

# QC 
- total number of reads 
- number of reads by time point

```{r QC-reads}
# count number of reads
count_num_of_reads <- function(region, gene) {
  
  # extract the read meta table
  rids_df <- region[["rids_df"]]

  # extract region coordinates
  region_chr <- region[["chr"]]
  region_start <- region[["start"]]
  region_end <- region[["end"]]
  
  # count reads at each time point
  reads_by_timepoint <- rids_df %>%
    count(sample_name, name = "n_reads")
  
  # total number of reads in this region
  total_reads <- nrow(rids_df)
  
  # add gene, region and total read information
  reads_by_timepoint %>%
    mutate(
      gene = gene,
      chr = region_chr,
      start = region_start,
      end = region_end,
      total_reads = total_reads
    )
}


# Store the QC results from all genes and regions
qc_reads <- list()

# original:
# base_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model"
base_dir <- output_dir
region_dirs <- list.dirs(base_dir, recursive = FALSE)

for (dir in region_dirs) {
   # folder name
  outname <- basename(dir)
  
  # find region RDS file
  region_file <- list.files(dir, pattern = "^region_met_data.*\\.rds$", full.names = TRUE)
  region <- readRDS(region_file)
  
  # gene name
  gene <- strsplit(outname, "_")[[1]][1]
  
  cat("=====================")
  cat("processing", outname, "\n")
  cat("=====================")
  
  # count and save reads
  qc_reads[[outname]] <- count_num_of_reads(region = region, gene = gene)
}

qc_reads <- bind_rows(qc_reads, .id = "region")

```
```{r plot-QC}
color <- c("LPS_0" = "#0072B2", "LPS_5" = "#E69F00", "LPS_10" = "#009E73", "LPS_15" = "#D55E00")

qcplot <- ggplot(qc_reads,
       aes(x = gene,y = n_reads,fill = sample_name)) +
  geom_col() +
  geom_text(
    aes(label = n_reads),
    position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = color) +
  labs(
    y = "Number of reads",
    fill = "Time point") +
  theme_cowplot()

qcplot

ggsave(filename = file.path(base_dir,"QC_reads_by_timepoint.pdf"))

```

# Recapitulating timepoints from topics

The topic model above pools all four LPS timepoints (0 = control, 5, 10, 15 min) into one `metA_mat` per gene and fits a single model, tagging every pooled read
with its `sample_name` (hence timepoint). 

The question here is whether them**topics recapitulate the timepoints** - i.e. whether any topic is used
predominantly by reads from one timepoint. We answer it per gene by (1) buildingma per-read topic x timepoint table, (2) summarising topic usage by timepoint, and
(3) testing whether that usage is timepoint-dependent.

All three steps reuse the artifacts already written underm`output_dir/<outname>/` by `process_region_m6a()`

For each gene we load the saved region data and binomia fit, recover `sample_name` row-aligned to `fit$L` by matching the filtered matrix's rownames (the RIDs) back into `region_data$rids_df`. Mapping `sample_name -> timepoint` via `sample_table` then labels each read. The result is one data frame per gene with `RID`, `timepoint`, the six topic proportions `L_topic1..L_topic6`, `frac_acc` ), and `dominant_topic = argmax(L)`


**for each gene, does topic membership of individual reads change over time?**
```{r recap-step1-fn}
# Build the per-read topic x timepoint table for one gene from its saved
# region-data + BINOMIAL fit. Rows are aligned to fit$L by construction

######################################
# creates one read level table per gene, then summarizes topic usage at each timepoint
######################################


per_read_topic_timepoint <- function(gene) {
  # original:
  # region <- promoters %>% dplyr::filter(gene == !!gene)
  # outname <- paste0(gene, "_", region$chr, "_", region$start, "_", region$end)
  outname <- gene

  # read the assembled methylation data
  region_data <- readRDS(file.path(output_dir, outname,
                                   paste0("region_met_data_", outname, ".rds")))
  # read the fitted topic model
  fit <- readRDS(file.path(output_dir, outname,
                           paste0("fit_binom_metA_", outname, ".rds")))

  # refilter exactly as process_region_m6a() 
  metA_filt <- filter_met_mat(region_data$metA_mat)
  stopifnot(identical(rownames(metA_filt), rownames(fit$L)))

  # for every read retained in metA, finds its row in the read metadata table
  keep_idx     <- match(rownames(metA_filt), region_data$rids_df$RID)
  # retrieve each read's sample
  sample_name  <- region_data$rids_df$sample_name[keep_idx]
  # map sample to timepoints
  timepoint    <- sample_table$timepoint[match(sample_name, sample_table$sample_name)]

  L <- as.matrix(fit$L)
  colnames(L) <- paste0("L_topic", seq_len(ncol(L)))
  df <- data.frame(
    RID            = rownames(fit$L),
    sample_name    = sample_name,
    timepoint      = factor(timepoint, levels = sort(unique(sample_table$timepoint))),
    L,
    # average m6A fraction acorss positions for each read, and calculated directly from the binary matrix without using the topic model
    frac_acc       = as.numeric(Matrix::rowMeans(metA_filt)),   
    dominant_topic = factor(paste0("topic", max.col(L, ties.method = "first")),
                            levels = paste0("topic", seq_len(ncol(L)))),
    stringsAsFactors = FALSE,
    row.names = NULL
  )

  # verifyf
  stopifnot(nrow(df) == nrow(fit$L))
  stopifnot(!any(is.na(df$timepoint)))
  stopifnot(setequal(levels(droplevels(df$timepoint)), sort(unique(sample_table$timepoint))))

  df
}

# Build for every gene; keep in a named list for the steps below.
per_read_tables <- lapply(promoters$gene, per_read_topic_timepoint)
names(per_read_tables) <- promoters$gene

for (g in promoters$gene) {
  cat("=== ", g, ": ", nrow(per_read_tables[[g]]), " reads ===\n", sep = "")
  print(table(per_read_tables[[g]]$timepoint))
}
```

# examine per read topics

  RID sample_name timepoint          L_topic1
1 m84241_260505_182714_s3/103682934/ccs      LPS_15        15 0.570525927968593
2  m84241_260505_182714_s3/13174534/ccs       LPS_5         5 0.000000009389935
3 m84241_260505_182714_s3/135791181/ccs       LPS_5         5 0.222503692486331
4 m84241_260505_182714_s3/155125361/ccs      LPS_10        10 0.000000009099663
5 m84241_260505_182714_s3/170986656/ccs      LPS_10        10 0.000000009333397
6 m84241_260505_182714_s3/190712068/ccs      LPS_15        15 0.336899111321520
           L_topic2          L_topic3          L_topic4          L_topic5
1 0.000000009587537 0.259254589912311 0.163050429693455 0.000000008945988
2 0.000000009341036 0.000000009331383 0.000000008479193 0.280149259250181
3 0.000000009336609 0.103938640272882 0.000000008822323 0.673557639145690
4 0.000000009587675 0.000000009354673 0.000000010539837 0.498890554216904
5 0.000000009820599 0.016331815727685 0.000000009376147 0.000000010788898
6 0.000000009498098 0.206074672101345 0.000000009019092 0.457026187469548
           L_topic6   frac_acc dominant_topic
1 0.007169033892115 0.24846626         topic1
2 0.719850704208272 0.07975460         topic6
3 0.000000009936166 0.10020450         topic5
4 0.501109407201249 0.08691207         topic6
5 0.983668144953274 0.07668712         topic6
6 0.000000010590397 0.13087935         topic5

```{r per-read-topics}
head(as.data.frame(per_read_tables[["IL1B"]]))
```

## Step 2 - Topic-usage-by-timepoint matrix + plot (per gene)

Two summaries per gene: (a) the **mean topic proportion per timepoint** (a
6 x 4 matrix whose columns - the timepoints - each sum to 1), and (b) the
**dominant-topic counts per timepoint** (a 6 x 4 contingency table). We plot the
mean-proportion summary as a stacked bar (topic composition within each
timepoint) so a timepoint-specific topic shows up as a band that swells at one
timepoint.

Verify: each timepoint column of the mean-proportion matrix sums to 1;
`rowSums()` of the contingency table equal the per-topic read totals.

```{r recap-step2-fn}

####################################
# (a) mean topic proportion per timepoint: topics (rows) x timepoints (cols)
####################################

topic_usage_by_timepoint <- function(df) {
  # find the topic membership columns
  L_cols <- grep("^L_topic", colnames(df), value = TRUE)
  # split reads by time points into different tables
  m <- sapply(split(df[, L_cols, drop = FALSE], df$timepoint),
              # for each timepoint specific table, convert it to a numeric matrix and calculate the mean of every topic column
              function(sub) colMeans(as.matrix(sub)))
  rownames(m) <- sub("^L_", "", L_cols)
  m
}

####################################
# (b) dominant-topic counts per timepoint: topics (rows) x timepoints (cols)
####################################

# create a count table using each read's single dominant topic
dominant_topic_contingency <- function(df) {
  table(topic = df$dominant_topic, timepoint = df$timepoint)
}

# Stacked-bar plot of the mean-proportion matrix (topic composition per timepoint)
plot_topic_usage_by_timepoint <- function(usage_mat, gene) {
  dfp <- as.data.frame(as.table(usage_mat))
  colnames(dfp) <- c("topic", "timepoint", "mean_prop")
  ggplot(dfp, aes(x = timepoint, y = mean_prop, fill = topic)) +
    geom_col() +
    scale_fill_manual(values = topic_colors[seq_len(nrow(usage_mat))]) +
    labs(x = "LPS timepoint (min)", y = "mean topic proportion",
         title = paste0(gene, " - topic composition by timepoint")) +
    theme_cowplot(font_size = 10) +
    theme(plot.title = element_text(hjust = 0.5))
}
```


```{r recap-step2-run, fig.width=6, fig.height=4, message=FALSE, warning=FALSE}
usage_mats   <- list()
contingencies <- list()
for (g in promoters$gene) {
  df <- per_read_tables[[g]]
  usage_mats[[g]]    <- topic_usage_by_timepoint(df)
  contingencies[[g]] <- dominant_topic_contingency(df)

  # verify: columns of the mean-proportion matrix each sum to 1
  stopifnot(all(abs(colSums(usage_mats[[g]]) - 1) < 1e-8))
  # verify: contingency rowSums == per-topic read totals
  stopifnot(all(rowSums(contingencies[[g]]) ==
                as.integer(table(factor(df$dominant_topic,
                                        levels = rownames(contingencies[[g]]))))))

  cat("=== ", g, " : mean topic proportion by timepoint ===\n", sep = "")
  print(round(usage_mats[[g]], 3))
  cat("--- dominant-topic counts by timepoint ---\n")
  print(contingencies[[g]])

  gg <- plot_topic_usage_by_timepoint(usage_mats[[g]], g)
  print(gg)
  # original:
  # outname <- paste0(g, "_", promoters$chr[promoters$gene == g], "_",
  #                   promoters$start[promoters$gene == g], "_",
  #                   promoters$end[promoters$gene == g])
  outname <- g
  ggsave(file.path(output_dir, outname,
                   paste0("topic_usage_by_timepoint_", outname, ".pdf")),
         gg, width = 6, height = 4)
}
```



## Topic trajectories over time (per gene)

The stacked bar above shows topic *composition* at each timepoint but makes it hard
to follow a single topic that rises then falls. This line plot draws each topic's
**mean proportion as a trajectory over the time course** (mean +/- SEM per
timepoint), one line per topic, faceted by gene. Non-monotonic dynamics - a topic
that peaks at 5 min then drops, or dips then rebounds - read directly as the shape of
its line. Because topic proportions sum to 1 within each read the lines are coupled
(one topic rising pushes others down), so read them together, not as six independent
stories.

```{r topic-trajectories, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
traj <- dplyr::bind_rows(lapply(promoters$gene, function(g) {
  per_read_tables[[g]] %>%
    dplyr::select(timepoint, dplyr::starts_with("L_topic")) %>%
    tidyr::pivot_longer(dplyr::starts_with("L_topic"),
                        names_to = "topic", values_to = "prop") %>%
    dplyr::group_by(timepoint, topic) %>%
    dplyr::summarise(mean = mean(prop), sem = sd(prop) / sqrt(dplyr::n()),
                     .groups = "drop") %>%
    dplyr::mutate(gene = g, topic = sub("^L_", "", topic))
}))
traj$tp <- as.numeric(as.character(traj$timepoint))

gg_traj <- ggplot(traj, aes(tp, mean, color = topic, group = topic)) +
  geom_errorbar(aes(ymin = mean - sem, ymax = mean + sem), width = 1, linewidth = 0.4) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  facet_wrap(~ gene, scales = "free_y") +
  scale_color_manual(values = topic_colors) +
  scale_x_continuous(breaks = c(0, 5, 10, 15)) +
  labs(x = "LPS timepoint (min)", y = "mean topic proportion (± SEM)", color = "topic",
       title = "Topic trajectories over the LPS time course (per gene)") +
  theme_cowplot(font_size = 10) +
  theme(legend.position = "bottom")
print(gg_traj)
ggsave(file.path(output_dir, "topic_trajectories_by_gene.pdf"),
       gg_traj, width = 11, height = 8)
```



## Empirical m6A methylation by timepoint (IL1A)

The topic profiles `fit$F` are timepoint-invariant by construction - all four LPS
timepoints are pooled into one `metA_mat` before fitting, so there is a single set of
topic profiles for the whole time course, and the temporal signal lives entirely in the
topic usage (`fit$L`, shown above). To ask directly whether the *observed* methylation
footprint shifts over time, we bypass the model and compute the **empirical mean m6A
fraction at each A position, split by timepoint**, from the same filtered `metA_mat`
(same reads and positions as the fit). One line per timepoint, per base pair (no
binning); the x-axis is genomic position, matching the topic-profile panels so the
empirical profile lines up position-for-position with the fitted topics.

```{r il1a-empirical-m6a-by-timepoint, fig.width=9, fig.height=8, message=FALSE, warning=FALSE}
gene   <- "IL1A"
region <- promoters %>% dplyr::filter(gene == !!gene)
# original:
# outname <- paste0(gene, "_", region$chr, "_", region$start, "_", region$end)
outname <- gene

# same loading + filtering as per_read_topic_timepoint(): reads/positions match the fit
region_data <- readRDS(file.path(output_dir, outname,
                                 paste0("region_met_data_", outname, ".rds")))
metA_filt   <- filter_met_mat(region_data$metA_mat)

keep_idx    <- match(rownames(metA_filt), region_data$rids_df$RID)
sample_name <- region_data$rids_df$sample_name[keep_idx]
timepoint   <- factor(sample_table$timepoint[match(sample_name, sample_table$sample_name)],
                      levels = sort(unique(sample_table$timepoint)))

# mean m6A fraction per A position (columns "A_<genomic_pos>"), per timepoint
prof <- do.call(rbind, lapply(split(seq_len(nrow(metA_filt)), timepoint), function(i)
          Matrix::colMeans(metA_filt[i, , drop = FALSE])))

dfp <- tidyr::pivot_longer(
  data.frame(timepoint = rownames(prof), as.data.frame(as.matrix(prof)), check.names = FALSE),
  cols = -timepoint, names_to = "col", values_to = "mean_m6a")
dfp$pos       <- as.numeric(sub("^A_", "", dfp$col))
dfp$timepoint <- factor(dfp$timepoint, levels = levels(timepoint))

# four contrasting, colourblind-friendly colours, one per timepoint panel
tp_colors <- c("0" = "#0072B2", "5" = "#E69F00", "10" = "#009E73", "15" = "#D55E00")

gg_prof <- ggplot(dfp, aes(pos, mean_m6a, color = timepoint)) +
  geom_line(linewidth = 0.5) +
  facet_wrap(~ timepoint, ncol = 1) +
  scale_x_continuous(limits = c(region$start, region$end), expand = c(0, 0)) +
  scale_color_manual(values = tp_colors) +
  labs(x = paste0(region$chr, " position"), y = "mean m6A fraction",
       color = "LPS timepoint (min)",
       title = paste0(gene, " - per-bp m6A methylation by timepoint")) +
  theme_cowplot(font_size = 10) +
  theme(plot.title = element_text(hjust = 0.5))
print(gg_prof)
ggsave(file.path(output_dir, outname, paste0("empirical_m6a_by_timepoint_", outname, ".pdf")),
       gg_prof, width = 9, height = 8)
```



# Export per-read topic assignments and topic profiles

`process_region_m6a()` computes the k-means read clusters but only uses them for
plotting - they are never written out. This section recovers them from the saved fits
and exports two tables per gene:

- `read_topic_assignments_<outname>.tsv`: one row per read - `RID`, read metadata,
  `cluster` (k-means), `dominant_topic = argmax(L)`, `max_loading`, and the topic
  proportions `k1..k6`.
- `topic_profiles_F_<outname>.tsv`: `fit$F` in long format - one row per
  (A position, topic) with the topic-specific m6A probability `prob`.

Clustering uses the **binomial** fit with `seed = 1` and the same k-means-on-PCA recipe
as `process_region_m6a()`, so the exported `cluster` labels are the ones drawn in the
saved `cluster_*.pdf` files.

```{r export-topic-assignments}
export_topic_tables <- function(gene) {
  # original:
  # region  <- promoters %>% dplyr::filter(gene == !!gene)
  # outname <- paste0(gene, "_", region$chr, "_", region$start, "_", region$end)
  outname <- gene

  region_data <- readRDS(file.path(output_dir, outname,
                                   paste0("region_met_data_", outname, ".rds")))
  fit <- readRDS(file.path(output_dir, outname,
                           paste0("fit_binom_metA_", outname, ".rds")))

  reads  <- read_topic_assignments(fit, region_data$rids_df, seed = 1)
  topics <- topic_profiles_long(fit)

  write.table(reads, file.path(output_dir, outname,
                               paste0("read_topic_assignments_", outname, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(topics, file.path(output_dir, outname,
                                paste0("topic_profiles_F_", outname, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)

  cat("=== ", gene, ": ", nrow(reads), " reads, k = ", ncol(fit$L), " ===\n", sep = "")
  # clusters and dominant topics are not the same labelling - crosstab shows how far apart
  print(table(reads$cluster, reads$dominant_topic))
  cat("\n")

  reads
}

topic_assignment_tables <- lapply(promoters$gene, export_topic_tables)
names(topic_assignment_tables) <- promoters$gene
```

                                  RID sample_name  chr     start       end
1 m84241_260505_182714_s3/103682934/ccs      LPS_15 chr2 112830678 112843009
2  m84241_260505_182714_s3/13174534/ccs       LPS_5 chr2 112826359 112840099
3 m84241_260505_182714_s3/135791181/ccs       LPS_5 chr2 112829972 112839548
4 m84241_260505_182714_s3/155125361/ccs      LPS_10 chr2 112835314 112850315
5 m84241_260505_182714_s3/170986656/ccs      LPS_10 chr2 112829278 112840187
6 m84241_260505_182714_s3/190712068/ccs      LPS_15 chr2 112833049 112838564
  strand  cluster dominant_topic max_loading                k1
1      - cluster5             k1   0.5705259 0.570525927968593
2      + cluster2             k6   0.7198507 0.000000009389935
3      - cluster6             k5   0.6735576 0.222503692486331
4      + cluster2             k6   0.5011094 0.000000009099663
5      + cluster4             k6   0.9836681 0.000000009333397
6      - cluster6             k5   0.4570262 0.336899111321520
                 k2                k3                k4                k5
1 0.000000009587537 0.259254589912311 0.163050429693455 0.000000008945988
2 0.000000009341036 0.000000009331383 0.000000008479193 0.280149259250181
3 0.000000009336609 0.103938640272882 0.000000008822323 0.673557639145690
4 0.000000009587675 0.000000009354673 0.000000010539837 0.498890554216904
5 0.000000009820599 0.016331815727685 0.000000009376147 0.000000010788898
6 0.000000009498098 0.206074672101345 0.000000009019092 0.457026187469548
                 k6
1 0.007169033892115
2 0.719850704208272
3 0.000000009936166
4 0.501109407201249
5 0.983668144953274
6 0.000000010590397

```{r export-topic-assignments-head}
  head(topic_assignment_tables[["IL1B"]])
```


<!-- # Topic modelling on footprints -->

<!-- Same multi-class footprint occupancy as `cluster_by_footprint_class.ipynb` -->
<!-- (merged classes `PPP_unknown`, `PIC`, `FIRE_nucleosome`; full-resolution binary; -->
<!-- strand-aware TSS +/- 300 bp), but reads are clustered with **topic modelling** -->
<!-- (fit a topic model, then k-means on the PCA of the topic proportions `L`) instead -->
<!-- of Ward hierarchical clustering. -->

<!-- Note: at full base-pair resolution, adjacent positions within a footprint are -->
<!-- highly correlated, so topics tend to capture broad coverage blocks rather than -->
<!-- fine structure - expected for contiguous footprints. -->

<!-- ```{r fp-settings} -->
<!-- fp_hw      <- 300 -->
<!-- fp_strands <- c(ENO1 = "-", ZCWPW1 = "-")   # both genes are on the minus strand -->
<!-- fp_classes <- names(FP_CLASS_COLORS) -->
<!-- ``` -->

<!-- ## ENO1 -->

<!-- ### Assemble footprint matrix -->

<!-- ```{r ENO1-fp-assemble} -->
<!-- gene <- "ENO1" -->
<!-- region <- promoters %>% dplyr::filter(gene == !!gene) -->
<!-- region_chr <- region$chr -->
<!-- tss <- region$tss -->
<!-- strand <- fp_strands[[gene]] -->
<!-- outname <- paste0(gene, "_", region_chr, "_", region$start, "_", region$end) -->

<!-- fp_outdir <- file.path(output_dir, outname, "footprints") -->
<!-- dir.create(fp_outdir, showWarnings = FALSE, recursive = TRUE) -->

<!-- fp_data <- assemble_footprint_mat(gene, region_chr, tss, strand, hw = fp_hw, verbose = TRUE) -->
<!-- occ_mat <- fp_data$occ_mat -->
<!-- saveRDS(fp_data, file.path(fp_outdir, paste0("footprint_data_", outname, ".rds"))) -->
<!-- ``` -->

<!-- ### Fit topic models -->

<!-- ```{r ENO1-fp-fit} -->
<!-- fit_multinom_fp <- fit_topic_model(occ_mat, k = n_topics) -->
<!-- fit_binom_fp    <- fit_binomial_topic_model(occ_mat, k = n_topics, numem = 100) -->

<!-- saveRDS(fit_multinom_fp, file.path(fp_outdir, paste0("fit_multinom_fp_", outname, ".rds"))) -->
<!-- saveRDS(fit_binom_fp,    file.path(fp_outdir, paste0("fit_binom_fp_",    outname, ".rds"))) -->
<!-- ``` -->

<!-- ### Plot -->

<!-- ```{r ENO1-fp-plot, fig.width=8, fig.height=8} -->
<!-- plot_fp_topic_model <- function(fit, title) { -->
<!--   gg_struct <- structure_plot(fit$L, colors = topic_colors[1:n_topics], -->
<!--                               topics = 1:n_topics, gap = 5) + -->
<!--     theme_cowplot(font_size = 10) + ggtitle(title) + -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           axis.text.x = element_blank(), axis.ticks.x = element_blank(), -->
<!--           axis.title.x = element_blank()) -->
<!--   gg_factors <- plot_footprint_topic_factors(fit, region = list(start = -fp_hw, end = fp_hw)) -->
<!--   cowplot::plot_grid(gg_struct, gg_factors, ncol = 1, rel_heights = c(1, 2)) -->
<!-- } -->

<!-- gg_multinom <- plot_fp_topic_model(fit_multinom_fp, paste0(gene, " (multinomial, footprints)")) -->
<!-- print(gg_multinom) -->
<!-- ggsave(file.path(fp_outdir, paste0("topic_model_multinom_", outname, ".pdf")), -->
<!--        gg_multinom, width = 8, height = 8) -->

<!-- gg_binom <- plot_fp_topic_model(fit_binom_fp, paste0(gene, " (binomial, footprints)")) -->
<!-- print(gg_binom) -->
<!-- ggsave(file.path(fp_outdir, paste0("topic_model_binom_", outname, ".pdf")), -->
<!--        gg_binom, width = 8, height = 8) -->
<!-- ``` -->

<!-- ### Compare numbers of topics (k = 2-8) -->

<!-- ```{r ENO1-fp-k-comparison, fig.width=8, fig.height=14, message=FALSE, warning=FALSE} -->
<!-- k_values <- 2:8 -->
<!-- gg_k_list <- lapply(k_values, function(k) { -->
<!--   set.seed(123) -->
<!--   fit_k <- fit_binomial_topic_model(occ_mat, k = k, numem = 100) -->
<!--   structure_plot(fit_k$L, colors = topic_colors[1:k], topics = 1:k, gap = 5) + -->
<!--     theme_cowplot(font_size = 10) + -->
<!--     ggtitle(paste0(gene, " (binomial, footprints), k = ", k)) + -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           axis.text.x = element_blank(), axis.ticks.x = element_blank(), -->
<!--           axis.title.x = element_blank()) -->
<!-- }) -->
<!-- gg_k_compare <- cowplot::plot_grid(plotlist = gg_k_list, ncol = 1) -->
<!-- print(gg_k_compare) -->
<!-- ggsave(file.path(fp_outdir, paste0("topic_model_k_comparison_", outname, ".pdf")), -->
<!--        gg_k_compare, width = 8, height = 14) -->
<!-- ``` -->

<!-- ### Cluster reads by topic proportions (k = `r n_topics`) -->

<!-- ```{r ENO1-fp-cluster} -->
<!-- fit <- fit_multinom_fp   # use the k = n_topics fit for clustering -->

<!-- set.seed(1) -->
<!-- pca <- prcomp(fit$L)$x -->
<!-- clusters <- kmeans(pca, centers = n_topics, iter.max = 100)$cluster -->

<!-- clusters <- factor(clusters) -->
<!-- levels(clusters) <- paste0("cluster", 1:n_topics) -->

<!-- cluster_colors <- colors_25[1:length(levels(clusters))] -->
<!-- ``` -->

<!-- Structure plot with reads grouped according to the clusters: -->

<!-- ```{r ENO1-fp-cluster-structure, fig.width=8, fig.height=3, message=FALSE} -->
<!-- gg_cluster_structure <- structure_plot(fit, topics = 1:n_topics, colors = topic_colors, gap = 5, -->
<!--                                        grouping = clusters) -->
<!-- print(gg_cluster_structure) -->
<!-- ggsave(file.path(fp_outdir, paste0("cluster_structure_", outname, ".pdf")), -->
<!--        gg_cluster_structure, width = 8, height = 3) -->
<!-- ``` -->

<!-- PCA plot colored by cluster: -->

<!-- ```{r ENO1-fp-cluster-pca, fig.width=5, fig.height=5, message=FALSE} -->
<!-- pca_plot(fit, fill = clusters) + -->
<!--   scale_fill_manual(values = cluster_colors) -->
<!-- ``` -->

<!-- Heatmap of the footprint occupancy matrix with reads ordered by cluster (color = -->
<!-- footprint class, white = uncovered, row sidebar = cluster): -->

<!-- ```{r ENO1-fp-cluster-heatmap, fig.width=10, fig.height=8} -->
<!-- plot_footprint_cluster_heatmap(occ_mat, clusters, fp_classes, FP_CLASS_COLORS, cluster_colors) -->

<!-- pdf(file.path(fp_outdir, paste0("cluster_heatmap_", outname, ".pdf")), -->
<!--     width = 10, height = 8) -->
<!-- plot_footprint_cluster_heatmap(occ_mat, clusters, fp_classes, FP_CLASS_COLORS, cluster_colors) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- ## ZCWPW1 -->

<!-- ### Assemble footprint matrix -->

<!-- ```{r ZCWPW1-fp-assemble} -->
<!-- gene <- "ZCWPW1" -->
<!-- region <- promoters %>% dplyr::filter(gene == !!gene) -->
<!-- region_chr <- region$chr -->
<!-- tss <- region$tss -->
<!-- strand <- fp_strands[[gene]] -->
<!-- outname <- paste0(gene, "_", region_chr, "_", region$start, "_", region$end) -->

<!-- fp_outdir <- file.path(output_dir, outname, "footprints") -->
<!-- dir.create(fp_outdir, showWarnings = FALSE, recursive = TRUE) -->

<!-- fp_data <- assemble_footprint_mat(gene, region_chr, tss, strand, hw = fp_hw, verbose = TRUE) -->
<!-- occ_mat <- fp_data$occ_mat -->
<!-- saveRDS(fp_data, file.path(fp_outdir, paste0("footprint_data_", outname, ".rds"))) -->
<!-- ``` -->

<!-- ### Fit topic models -->

<!-- ```{r ZCWPW1-fp-fit} -->
<!-- fit_multinom_fp <- fit_topic_model(occ_mat, k = n_topics) -->
<!-- fit_binom_fp    <- fit_binomial_topic_model(occ_mat, k = n_topics, numem = 100) -->

<!-- saveRDS(fit_multinom_fp, file.path(fp_outdir, paste0("fit_multinom_fp_", outname, ".rds"))) -->
<!-- saveRDS(fit_binom_fp,    file.path(fp_outdir, paste0("fit_binom_fp_",    outname, ".rds"))) -->
<!-- ``` -->

<!-- ### Plot -->

<!-- ```{r ZCWPW1-fp-plot, fig.width=8, fig.height=8} -->
<!-- gg_multinom <- plot_fp_topic_model(fit_multinom_fp, paste0(gene, " (multinomial, footprints)")) -->
<!-- print(gg_multinom) -->
<!-- ggsave(file.path(fp_outdir, paste0("topic_model_multinom_", outname, ".pdf")), -->
<!--        gg_multinom, width = 8, height = 8) -->

<!-- gg_binom <- plot_fp_topic_model(fit_binom_fp, paste0(gene, " (binomial, footprints)")) -->
<!-- print(gg_binom) -->
<!-- ggsave(file.path(fp_outdir, paste0("topic_model_binom_", outname, ".pdf")), -->
<!--        gg_binom, width = 8, height = 8) -->
<!-- ``` -->

<!-- ### Compare numbers of topics (k = 2-8) -->

<!-- ```{r ZCWPW1-fp-k-comparison, fig.width=8, fig.height=14, message=FALSE, warning=FALSE} -->
<!-- k_values <- 2:8 -->
<!-- gg_k_list <- lapply(k_values, function(k) { -->
<!--   set.seed(123) -->
<!--   fit_k <- fit_binomial_topic_model(occ_mat, k = k, numem = 100) -->
<!--   structure_plot(fit_k$L, colors = topic_colors[1:k], topics = 1:k, gap = 5) + -->
<!--     theme_cowplot(font_size = 10) + -->
<!--     ggtitle(paste0(gene, " (binomial, footprints), k = ", k)) + -->
<!--     theme(plot.title = element_text(hjust = 0.5), -->
<!--           axis.text.x = element_blank(), axis.ticks.x = element_blank(), -->
<!--           axis.title.x = element_blank()) -->
<!-- }) -->
<!-- gg_k_compare <- cowplot::plot_grid(plotlist = gg_k_list, ncol = 1) -->
<!-- print(gg_k_compare) -->
<!-- ggsave(file.path(fp_outdir, paste0("topic_model_k_comparison_", outname, ".pdf")), -->
<!--        gg_k_compare, width = 8, height = 14) -->
<!-- ``` -->

<!-- ### Cluster reads by topic proportions (k = `r n_topics`) -->

<!-- ```{r ZCWPW1-fp-cluster} -->
<!-- fit <- fit_multinom_fp   # use the k = n_topics fit for clustering -->

<!-- set.seed(1) -->
<!-- pca <- prcomp(fit$L)$x -->
<!-- clusters <- kmeans(pca, centers = n_topics, iter.max = 100)$cluster -->

<!-- clusters <- factor(clusters) -->
<!-- levels(clusters) <- paste0("cluster", 1:n_topics) -->

<!-- cluster_colors <- colors_25[1:length(levels(clusters))] -->
<!-- ``` -->

<!-- Structure plot with reads grouped according to the clusters: -->

<!-- ```{r ZCWPW1-fp-cluster-structure, fig.width=8, fig.height=3, message=FALSE} -->
<!-- gg_cluster_structure <- structure_plot(fit, topics = 1:n_topics, colors = topic_colors, gap = 5, -->
<!--                                        grouping = clusters) -->
<!-- print(gg_cluster_structure) -->
<!-- ggsave(file.path(fp_outdir, paste0("cluster_structure_", outname, ".pdf")), -->
<!--        gg_cluster_structure, width = 8, height = 3) -->
<!-- ``` -->

<!-- PCA plot colored by cluster: -->

<!-- ```{r ZCWPW1-fp-cluster-pca, fig.width=5, fig.height=5, message=FALSE} -->
<!-- pca_plot(fit, fill = clusters) + -->
<!--   scale_fill_manual(values = cluster_colors) -->
<!-- ``` -->

<!-- Heatmap of the footprint occupancy matrix with reads ordered by cluster (color = -->
<!-- footprint class, white = uncovered, row sidebar = cluster): -->

<!-- ```{r ZCWPW1-fp-cluster-heatmap, fig.width=10, fig.height=8} -->
<!-- plot_footprint_cluster_heatmap(occ_mat, clusters, fp_classes, FP_CLASS_COLORS, cluster_colors) -->

<!-- pdf(file.path(fp_outdir, paste0("cluster_heatmap_", outname, ".pdf")), -->
<!--     width = 10, height = 8) -->
<!-- plot_footprint_cluster_heatmap(occ_mat, clusters, fp_classes, FP_CLASS_COLORS, cluster_colors) -->
<!-- dev.off() -->
<!-- ``` -->

<!-- # Footprints grouped by methylation cluster -->

<!-- Reuse the read clusters identified by the **m6A methylation** topic model and plot -->
<!-- the **footprints** of those same reads grouped by their methylation cluster. Reads -->
<!-- are matched between the two analyses by read name (both use PacBio `ccs` read -->
<!-- names). A read is kept only if it has both a methylation cluster and >=1 footprint -->
<!-- in the window; reads without a methylation cluster are dropped. -->

<!-- For each gene: a per-cluster aggregate footprint profile (mean per-class occupancy) -->
<!-- and a per-read footprint class heatmap, both ordered/grouped by the methylation -->
<!-- cluster. Each is drawn in two coordinate systems: **TSS-relative** (strand-oriented, -->
<!-- as in the footprint section above) and **genomic** (absolute coordinates, not -->
<!-- strand-flipped - the same coordinate system as the m6A methylation heatmap, so the -->
<!-- two can be aligned directly). -->

<!-- ```{r fp-by-met-helper} -->
<!-- # Load the saved m6A fit + region data for a gene, recompute the methylation -->
<!-- # clusters, and return a data.frame mapping RID -> methylation cluster. -->
<!-- methylation_clusters_for_gene <- function(outname) { -->
<!--   region_data <- readRDS(file.path(output_dir, outname, paste0("region_met_data_", outname, ".rds"))) -->
<!--   fit_met     <- readRDS(file.path(output_dir, outname, paste0("fit_multinom_metA_", outname, ".rds"))) -->
<!--   metA_filt   <- filter_met_mat(region_data$metA_mat) -->
<!--   met_cl      <- cluster_reads_by_topics(fit_met, n_topics, seed = 1) -->
<!--   data.frame(RID = rownames(metA_filt), met_cluster = met_cl, stringsAsFactors = FALSE) -->
<!-- } -->

<!-- # Draw a footprint heatmap to the screen and save it to PDF (base graphics must -->
<!-- # be re-drawn for each device). -->
<!-- save_fp_heatmap <- function(occ, clusters, file, main) { -->
<!--   plot_footprint_cluster_heatmap(occ, clusters, fp_classes, FP_CLASS_COLORS, cluster_colors, main = main) -->
<!--   pdf(file, width = 10, height = 8) -->
<!--   plot_footprint_cluster_heatmap(occ, clusters, fp_classes, FP_CLASS_COLORS, cluster_colors, main = main) -->
<!--   dev.off() -->
<!-- } -->

<!-- # Draw + save a per-cluster mean-occupancy profile plot. -->
<!-- save_fp_profiles <- function(occ, clusters, file, title, xlab) { -->
<!--   gg <- plot_footprint_cluster_profiles(occ, clusters, fp_classes, FP_CLASS_COLORS, -->
<!--                                         title = title, xlab = xlab) -->
<!--   print(gg) -->
<!--   ggsave(file, gg, width = 8, height = 10) -->
<!-- } -->
<!-- ``` -->

<!-- ## ENO1 -->

<!-- ```{r ENO1-fp-by-met, fig.width=10, fig.height=8} -->
<!-- gene <- "ENO1" -->
<!-- region <- promoters %>% dplyr::filter(gene == !!gene) -->
<!-- outname <- paste0(gene, "_", region$chr, "_", region$start, "_", region$end) -->
<!-- fp_outdir <- file.path(output_dir, outname, "footprints") -->

<!-- met_map <- methylation_clusters_for_gene(outname) -->
<!-- fp_data <- readRDS(file.path(fp_outdir, paste0("footprint_data_", outname, ".rds"))) -->
<!-- occ_mat <- fp_data$occ_mat -->

<!-- # match footprint reads to methylation clusters, keep reads present in both -->
<!-- met_clusters <- met_map$met_cluster[match(rownames(occ_mat), met_map$RID)] -->
<!-- keep <- !is.na(met_clusters) -->
<!-- cat(gene, ": ", sum(keep), " of ", nrow(occ_mat), -->
<!--     " footprint reads have a methylation cluster\n", sep = "") -->

<!-- occ_keep <- occ_mat[keep, , drop = FALSE] -->
<!-- met_keep <- droplevels(met_clusters[keep]) -->
<!-- cluster_colors <- colors_25[1:n_topics] -->

<!-- # genomic-coordinate version (absolute, not strand-flipped; matches the m6A -->
<!-- # methylation heatmap). occ_keep stays TSS-relative (strand-oriented). -->
<!-- occ_genomic <- relabel_occ_to_genomic(occ_keep, region$tss, fp_strands[[gene]]) -->

<!-- # per-cluster aggregate mean-occupancy profiles, in both coordinate systems -->
<!-- save_fp_profiles(occ_keep, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_profiles_tssrel_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (TSS-relative)"), -->
<!--   "Distance to TSS (bp, strand-oriented)") -->

<!-- save_fp_profiles(occ_genomic, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_profiles_genomic_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (genomic coord)"), -->
<!--   paste0(region$chr, " position (bp)")) -->

<!-- # per-read footprint class heatmap, rows grouped by methylation cluster, in both -->
<!-- # coordinate systems. -->
<!-- save_fp_heatmap(occ_keep, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_heatmap_tssrel_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (TSS-relative)")) -->

<!-- save_fp_heatmap(occ_genomic, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_heatmap_genomic_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (genomic coord)")) -->
<!-- ``` -->

<!-- ## ZCWPW1 -->

<!-- ```{r ZCWPW1-fp-by-met, fig.width=10, fig.height=8} -->
<!-- gene <- "ZCWPW1" -->
<!-- region <- promoters %>% dplyr::filter(gene == !!gene) -->
<!-- outname <- paste0(gene, "_", region$chr, "_", region$start, "_", region$end) -->
<!-- fp_outdir <- file.path(output_dir, outname, "footprints") -->

<!-- met_map <- methylation_clusters_for_gene(outname) -->
<!-- fp_data <- readRDS(file.path(fp_outdir, paste0("footprint_data_", outname, ".rds"))) -->
<!-- occ_mat <- fp_data$occ_mat -->

<!-- met_clusters <- met_map$met_cluster[match(rownames(occ_mat), met_map$RID)] -->
<!-- keep <- !is.na(met_clusters) -->
<!-- cat(gene, ": ", sum(keep), " of ", nrow(occ_mat), -->
<!--     " footprint reads have a methylation cluster\n", sep = "") -->

<!-- occ_keep <- occ_mat[keep, , drop = FALSE] -->
<!-- met_keep <- droplevels(met_clusters[keep]) -->
<!-- cluster_colors <- colors_25[1:n_topics] -->

<!-- # genomic-coordinate version (absolute, not strand-flipped; matches the m6A -->
<!-- # methylation heatmap). occ_keep stays TSS-relative (strand-oriented). -->
<!-- occ_genomic <- relabel_occ_to_genomic(occ_keep, region$tss, fp_strands[[gene]]) -->

<!-- # per-cluster aggregate mean-occupancy profiles, in both coordinate systems -->
<!-- save_fp_profiles(occ_keep, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_profiles_tssrel_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (TSS-relative)"), -->
<!--   "Distance to TSS (bp, strand-oriented)") -->

<!-- save_fp_profiles(occ_genomic, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_profiles_genomic_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (genomic coord)"), -->
<!--   paste0(region$chr, " position (bp)")) -->

<!-- # per-read footprint class heatmap, rows grouped by methylation cluster, in both -->
<!-- # coordinate systems. -->
<!-- save_fp_heatmap(occ_keep, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_heatmap_tssrel_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (TSS-relative)")) -->

<!-- save_fp_heatmap(occ_genomic, met_keep, -->
<!--   file.path(fp_outdir, paste0("footprints_by_met_cluster_heatmap_genomic_", outname, ".pdf")), -->
<!--   paste0(gene, " - footprints by m6A cluster (genomic coord)")) -->
<!-- ``` -->
