# footprint profiles around selected genomic features

FiberHMM footprints (LPS timecourse: 0 = control, 5, 10, 15 min) profiled around
THP-1 CAGE TSS, plus nucleosome-to-nucleosome spacing over DHSs (section 6).
Processing lives in `process_nuc_pos.R` (`main_nuc_pos()`, `main_nuc_spacing()`,
or `main()` for both), plotting in `plot_nuc_pos.R` (`main_plots()`); this
document only defines regions and renders figures.


```{r source-functions}
source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/process_nuc_pos.R")
source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/plot_nuc_pos.R")
```


```{r fiberhmm-footprints}
LPS_ft <- fread("/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/firehmm_footprint/LPS0/LPS0_hmm_extracted_footprint_chr1.bed.gz")
```

```{r}
colnames(LPS_ft) <- c("chrom", "chromStart", "chromEnd", "name", "score", "strand",
                  "thickStart", "thickEnd", "itemRgb",
                  "blockCount", "blockSizes", "blockStarts", "score")

LPS_ft <- LPS_ft[, -13]
```

```{r}
# subset for example
LPS_ft_filtered <- LPS_ft[c(1:5),]
head(LPS_ft_filtered)
length(unlist(strsplit(LPS_ft_filtered[[12]][1], ",")))

```

### per read and aggregatefootprint extraction step by step example

```{r extract-fp-function}
# construct footprint midpoint, with one foorprint per row, with its assoicated read ID
extract_footprints_per_read(LPS_ft_filtered)

# convert read span bed12 to Granges
LPS_gr <-  GRanges(
    seqnames = LPS_ft_filtered$chrom,
    ranges   = IRanges(LPS_ft_filtered$chromStart + 1L, LPS_ft_filtered$chromEnd)
  )

LPS_gr
```
```{r overlap-CAGE-TSS}
# Convert FiberHMM read spans from BED coordinates to GRanges coordinates
LPS_gr <- GRanges(
  seqnames = LPS_ft$chrom,
  ranges = IRanges(
    start = LPS_ft$chromStart + 1L,
    end   = LPS_ft$chromEnd
  )
)

# Load the dominant CAGE TSS annotation
CAGE_PATH <- paste0(
  "/project/spott/cshan/fiber-seq/macrophage_project/annotations/",
  "fantom5_thp1/THP1.consensus.CAGE_peaks.dominantPerGene.bed.gz"
)

CAGE_tss <- load_cage_tss(CAGE_PATH)

# Make ±2 kb windows around each CAGE TSS
WINDOW_BP <- 2000L

CAGE_windows <- promoters(
  CAGE_tss,
  upstream = WINDOW_BP,
  downstream = WINDOW_BP + 1L
)

# Find reads whose spans overlap at least one TSS window
hits <- findOverlaps(LPS_gr, CAGE_windows)

keep <- unique(queryHits(hits))

# Keep both the genomic ranges and the original BED12 records
LPS_gr_near_tss <- LPS_gr[keep]
LPS_ft_near_tss <- LPS_ft[keep]

length(LPS_gr_near_tss)
nrow(LPS_ft_near_tss)

LPS_gr_near_tss
LPS_ft_near_tss

# extract footprint midpoints for reads spanning TSS
LPS_ft_tss_filtered <- extract_footprints_per_read(LPS_ft)
LPS_ft_tss_filtered

LPS_ft_tss_filtered_gr <- GRanges(
  seqnames = LPS_ft_tss_filtered$chrom,
  ranges = IRanges(LPS_ft_tss_filtered$fp_mid + 1L, LPS_ft_tss_filtered$fp_mid + 1L),
  strand = "*"
)

# finds every footprint midpoint that lies inside a window
hits <- findOverlaps(LPS_ft_tss_filtered_gr, CAGE_windows)
hits

# find the corresponding footprint and TSS window indices
qi <- queryHits(hits) # extract footprint indices
si <- subjectHits(hits) # corresponding TSS indices


# per read table
################################
rtp <- data.table(
  tss_idx    = si,
  RID        = LPS_ft_tss_filtered$RID[qi],
  fp_mid     = LPS_ft_tss_filtered$fp_mid[qi],
  fp_size    = LPS_ft_tss_filtered$fp_size[qi],
  tss_pos    = start(CAGE_tss)[si],
  tss_strand = as.character(strand(CAGE_tss))[si],
  gene_id    = as.character(mcols(CAGE_tss)$gene_id)[si])

rtp

n_ranks = 3
# add new column with relative position to TSS
rtp[, rel_pos := fifelse(tss_strand == "+", fp_mid - tss_pos, tss_pos - fp_mid)]
rtp[, abs_dist := abs(rel_pos)]
  
# negative positive = upstream
rtp[, direction := fifelse(rel_pos < 0L, "upstream", "downstream")]
setorder(rtp, tss_idx, RID, direction, abs_dist)
  
# for each TSS, read, direction combination, sort by ascending order
rtp[, rank := seq_len(.N), by = .(tss_idx, RID, direction)]
# assign ranking in each group
rtp <- rtp[rank <= n_ranks]
# keep only the nearest 3 footprints in each direction
rtp[, signed_rank := fifelse(direction == "upstream", -rank, rank)]
rtp[]
  

# aggregate table
################################
dt <- data.table(
    tss_idx    = si,
    fp_mid     = LPS_ft_tss_filtered$fp_mid[qi],
    fp_size    = LPS_ft_tss_filtered$fp_size[qi],
    tss_pos    = start(CAGE_tss)[si],
    tss_strand = as.character(strand(CAGE_tss))[si],
    gene_id    = as.character(mcols(CAGE_tss)$gene_id)[si])

dt

dt[, rel_pos := fifelse(tss_strand == "+", fp_mid - tss_pos, tss_pos - fp_mid)]
dt[]
```

### Aggregation across chromosomes for one sample

```{r footprint-meta-profile}
window_bp = 2000
bin_size = 20
# assign each footprint to a bin
rtp_filtered <- rtp[c(1:5),]

bins <- data.table(
  bin = seq(-window_bp, window_bp, by = bin_size)
)
rtp_filtered[, bin := round(rel_pos / bin_size) * bin_size]

meta_raw <- bin_metaprofile(
  hits = rtp,
  window_bp = 2000,
  bin_size = 20,
  normalize = "none"
)

meta_raw

```


## 1. Define genomic regions of interest
```{r regions}
promoters <- tibble::tribble(
  ~gene,        ~chr,     ~start,      ~end,
  "IL1B",       "chr2",   112836323,   112837385,
  "TNF",        "chr6",   31575265,    31575824,
  "IL1A",       "chr2",   112793233,   112794969,
  "PTGS2",      "chr1",   186679985,   186681163,
  "IL8",        "chr4",   73740006,    73741332,
  "CXCL2",      "chr4",   74098863,    74099651,
  "IL6",        "chr7",   22726376,    22727094,
  "RELB",       "chr19",  45001011,    45002417,
  "rs2836882",  "chr21",  39094468,    39094712
)
promoters
```

## 2. Build the THP-1 CAGE consensus TSS annotation

Downloads the three FANTOM5 THP-1 hCAGE CTSS replicates, keeps CTSS present in
all three (tag counts summed), and assigns the nearest gencode gene while keeping
the CAGE strand. See `cage_consensus_thp1.sh`.

```{bash cage_consensus, eval = FALSE}
bash /project/spott/cshan/fiber-seq/code/nucleosome_positioning/cage_consensus_thp1.sh
# -> /project/spott/cshan/fiber-seq/macrophage_project/annotations/fantom5_thp1/THP1.consensus.CAGE_peaks.dominantPerGene.bed.gz
```

## 3. Process footprints around TSS (all timepoints)

Reads the per-chromosome FiberHMM footprint BEDs for LPS0/5/10/15, tallies
strand-aware footprint offsets to CAGE TSS (genome-wide and for the selected
promoters), and finds the nearest upstream/downstream footprint per (TSS, read).
This is the heavy step; run once, then load the TSV outputs below.

```{bash process, eval = FALSE}
module load R/4.4.1
# main_nuc_pos() = the TSS analysis only. Running the script bare instead calls
# main(), which also runs the DHS spacing analysis of section 6.
Rscript -e 'source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/process_nuc_pos.R"); main_nuc_pos()'
# -> /project/spott/cshan/fiber-seq/macrophage_project/nucleosome_positioning/results/tables/*.tsv[.gz]
```

## 4. Figures

```{r figures, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5}
res_dir  <- "/project/spott/cshan/fiber-seq/macrophage_project/nucleosome_positioning/results/tables"
meta_all  <- data.table::fread(file.path(res_dir, "metaprofile_genomewide_by_timepoint.tsv"))
meta_sel  <- data.table::fread(file.path(res_dir, "metaprofile_selected_genes_by_timepoint.tsv"))
meta_comb <- data.table::fread(file.path(res_dir, "metaprofile_combined_selected_by_timepoint.tsv"))
nearest   <- data.table::fread(file.path(res_dir, "nearest_updown_footprint_by_timepoint.tsv.gz"))
shift     <- data.table::fread(file.path(res_dir, "nearest_updown_shift_summary.tsv"))

# Genome-wide metaprofile (footprints filtered to 130-160 bp), timepoints overlaid.
# Aggregated (all midpoints counted once) and per-read.
plot_genomewide_metaprofile(meta_all, "aggregated", "raw")
plot_genomewide_metaprofile(meta_all, "aggregated", "per_tss")
plot_genomewide_metaprofile(meta_all, "per_read", "raw")
plot_genomewide_metaprofile(meta_all, "per_read", "per_read")

# Combined selected genes (all pooled into one metaprofile), timepoints overlaid.
plot_combined_selected_metaprofile(meta_comb, "aggregated", "raw")
plot_combined_selected_metaprofile(meta_comb, "aggregated", "per_tss")
plot_combined_selected_metaprofile(meta_comb, "per_read", "raw")
plot_combined_selected_metaprofile(meta_comb, "per_read", "per_read")

# Selected promoters: one plot per gene, timepoints as side-by-side panels
# (aggregated raw and per-read raw)
for (g in sort(unique(meta_sel$gene))) {
  print(plot_gene_metaprofile(meta_sel, g, "aggregated", "raw"))
  print(plot_gene_metaprofile(meta_sel, g, "per_read", "raw"))
}

# Nearest up/downstream footprint shift relative to TSS
plot_nearest_updown(nearest)
plot_nearest_shift(shift)

# Nearest-footprint size distribution, split by up/downstream
plot_footprint_size_hist(nearest)
```

## 5. Write figure PDFs

```{r save_pdfs, eval = FALSE}
# Regenerates all PDFs (TSS metaprofiles + DHS spacing) under results/plots/.
main_plots()
```

## 6. Nucleosome-to-nucleosome spacing over DHSs (autocorrelation)

Pooled pairwise distances between 130-160 bp nucleosome midpoints on the same
read, within DHS windows (DNase summit +/- 2 kb, all DHSs genome-wide). Peaks at
the nucleosome repeat length and its multiples indicate phased spacing.

Prepare the DHS annotation (ENCFF359IAQ = ENCODE DNase peaks with a summit
column). Build a summit BED, bgzip + tabix.

```{bash dhs_prep, eval = FALSE}
module load htslib/1.16
ANN=/project/spott/cshan/fiber-seq/macrophage_project/nucleosome_positioning/results/annotations
cd $ANN
# already downloaded: ENCFF359IAQ.bed.gz
zcat ENCFF359IAQ.bed.gz | awk 'BEGIN{OFS="\t"} /^#/{next} {print $1, $7-1, $7, $4, $8, "."}' \
  | sort -k1,1 -k2,2n > ENCFF359IAQ.summits.bed
bgzip -f ENCFF359IAQ.summits.bed
tabix -p bed ENCFF359IAQ.summits.bed.gz
```

Processing lives in `process_nuc_pos.R` (`main_nuc_spacing()`), plotting in
`plot_nuc_pos.R` -- both already sourced at the top of this document.

```{bash spacing_process, eval = FALSE}
module load R/4.4.1
Rscript -e 'source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/process_nuc_pos.R"); main_nuc_spacing()'
# -> results/tables/nuc_spacing_over_DHS_by_timepoint.tsv
```

```{r spacing_figures, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5}
spacing <- data.table::fread(file.path(res_dir, "nuc_spacing_over_DHS_by_timepoint.tsv"))
plot_nuc_spacing(spacing, "N_norm")   # depth-normalized (fraction of pairs)
plot_nuc_spacing(spacing, "N")        # raw pair counts
```

## 7. Footprint size distribution (all footprints, NOT size-filtered)

Every other section filters footprints to the 130-160 bp mono-nucleosome window.
This one tallies **all** footprint sizes
```{bash size_process, eval = FALSE}
module load R/4.4.1
Rscript -e 'source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/process_nuc_pos.R"); main_footprint_size_table()'
# -> results/tables/footprint_size_distribution.tsv.gz
# main_footprint_size_table(chroms = "chr21") restricts it to one chromosome.
```

```{r size_figure, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 9}
fp_sizes <- data.table::fread(file.path(res_dir, "footprint_size_distribution.tsv.gz"))

# One panel per timepoint, stacked vertically; shaded band = the 130-160 bp
# mono-nucleosome filter used everywhere else.
plot_footprint_size_distribution(fp_sizes)

# Fractions instead of counts, so the four timepoints are directly comparable
plot_footprint_size_distribution(fp_sizes, normalize = TRUE)
```
