---
title: "Single-molecule nucleosome positioning around TSS by Leiden accessibility cluster"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

FiberHMM nucleosome and TF footprints of single molecules at the
early-response promoters and two custom regions, grouped by their m6A
accessibility cluster from `leiden_manhattan.Rmd` (Leiden + Manhattan, all
four LPS timepoints pooled) and oriented to the GENCODE canonical TSS. Every
region is plotted on exactly the window it was clustered on: canonical TSS
+/- 1000 bp for the promoters, the region itself for the custom regions.
Questions:

1. Do the Leiden accessibility clusters have distinct single-molecule
   nucleosome configurations? (read-level tracks, per-cluster occupancy)
2. Within a cluster, do the -1 / +1 nucleosome positions move across the LPS
   time course? (positional summaries, timepoint bars in the tracks)

Processing functions: `process_nuc_pos.R`; plots: `plot_nuc_pos.R`; cluster
assignments: `clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan.Rmd`
(window_size = 0, Manhattan distance, k = 10, resolution = 1, seed = 1).
Nucleosome positions are never used for clustering: accessibility defines the
states, the footprints describe them.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Libraries and source files

```{r libraries, message=FALSE}
library(data.table)
library(ggplot2)
library(cowplot)

source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/process_nuc_pos.R")
source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/plot_nuc_pos.R")
```

# 2. Settings

```{r settings}
options(scipen = 999)

promoter_table <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical/promoter_regions_gencode_v49.tsv"

# Leiden + Manhattan outputs (leiden_manhattan.Rmd): the run to use is the one
# whose run_parameters match these values and the region's window -- unbinned
# m6A sites, k = 10, resolution = 1, all four timepoints pooled.
leiden_dir    <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/Leiden_Manhattan"
leiden_params <- list(window_size = 0L, k_neighbors = 10, resolution = 1, seed = 1)

out_dir   <- "/project/spott/cshan/fiber-seq/macrophage_project/nucleosome_positioning/results"
table_dir <- file.path(out_dir, "tables")
plot_dir  <- file.path(out_dir, "plots")
for (d in c(table_dir, plot_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)
# the figure files of the knitted html go under results/plots as well, so
# nothing is written next to this Rmd
knitr::opts_chunk$set(fig.path = file.path(plot_dir, "html", ""))

# analysis constants (defined in process_nuc_pos.R)
data.frame(gencode_tss = GENCODE_TSS_PATH, view_half_width = VIEW_HALF_WIDTH,
           profile_bin_bp = PROFILE_BIN_BP)
```

# 3. Regions

The 8 early-response promoters from `make_promoter_table.R` (genes, ENSG and
strand; region_id = gene, IL8 = GENCODE CXCL8) plus the two custom coordinate
regions, as in `leiden_manhattan.Rmd`.

```{r regions}
promoters <- load_promoter_table(promoter_table)
promoters

custom <- data.frame(
  chr = c("chr4", "chr5"),
  start = c(112284963, 150068577),
  end = c(112285236, 150068792)
)
custom$region_id <- paste(custom$chr, custom$start, custom$end, sep = "_")

regions <- rbind(
  data.frame(chr = promoters$chr, start = promoters$start, end = promoters$end,
             region_id = promoters$gene),
  custom
)
regions
```

# 4. GENCODE canonical TSS annotation and windows

Promoters are anchored on the GENCODE v46 Ensembl-canonical TSS of their ENSG
(TSS = 0, upstream negative, mirrored for minus-strand genes); their window
(`view_start`-`view_end`) is the canonical TSS +/- `r VIEW_HALF_WIDTH` bp
clustered in `leiden_manhattan.Rmd`. `tss_offset` is the distance from the
promoter table's own TSS (the v49 gene 5' end) to the canonical TSS in the
direction of transcription. The custom regions are anchored on their centre
and their window is the region itself.

```{r gencode-tss}
gencode_tss <- load_gencode_tss()
head(gencode_tss)

anchors <- build_region_anchors(regions, promoters, gencode_tss)
anchors
fwrite(anchors, file.path(table_dir, "region_anchors.tsv"), sep = "\t")
```

# 5. Load Leiden assignments

One cluster label per read from the pooled-timepoint run of each region
(`<region_id>/bin0/k<n>/read_cluster_assignments_<region_id>.tsv`, the run
whose parameters and window match section 4), with region_id added. Read
spans are kept: they are the denominators of the occupancy profiles and the
baselines of the read tracks.

```{r leiden}
assignments <- load_leiden_assignments(leiden_dir, anchors, leiden_params)
reads <- build_read_table(assignments, anchors)

dcast(reads[, .N, by = .(region_id, timepoint)], region_id ~ timepoint, value.var = "N")
dcast(reads[, .N, by = .(region_id, cluster)], region_id ~ cluster, value.var = "N", fill = 0)
```

# 6. Extract nucleosome footprints

All FiberHMM nucleosome footprints (every block of the footprint BED12, no
size filter) and the three TF footprint size classes for every clustered
read, from each read's own sample BED, RID preserved. One row per footprint,
in genomic (BED) and anchor-relative coordinates.

```{r footprints}
footprints <- extract_region_footprints(reads, anchors)
dcast(footprints[, .N, by = .(region_id, fp_class)], region_id ~ fp_class,
      value.var = "N", fill = 0)
```

# 7. Join footprints to Leiden clusters

```{r join}
footprints <- merge(footprints,
                    reads[, .(region_id, RID, sample_name, timepoint, cluster, gene, ensg)],
                    by = c("region_id", "RID", "sample_name"))
setcolorder(footprints, c("region_id", "gene", "ensg", "RID", "sample_name", "timepoint",
                          "cluster", "fp_class", "fp_start", "fp_end", "fp_mid",
                          "rel_start", "rel_end", "rel_mid", "fp_size"))
head(footprints)

# per-read positional summary (-1 / +1 nucleosome, NFR width, anchor occupied)
read_positions <- read_position_summary(footprints, reads)
head(read_positions)

fwrite(reads,          file.path(table_dir, "reads_by_cluster.tsv"), sep = "\t")
fwrite(footprints,     file.path(table_dir, "footprints_by_read.tsv.gz"), sep = "\t")
fwrite(read_positions, file.path(table_dir, "read_position_summary.tsv"), sep = "\t")
```

# 8. Read-level nucleosome and TF footprint plots (Fig 1)

One row per molecule, footprints at their full width, clipped to the window.
Rows are grouped by Leiden cluster (facets), then timepoint, then the position
of the +1 nucleosome; the bars on the left mark cluster and timepoint.

```{r read-tracks, fig.width=9, fig.height=11}
p_reads <- list()
for (rid in anchors$region_id) {
  m       <- anchors[region_id == rid]
  reads_r <- reads[region_id == rid]
  if (!nrow(reads_r)) next
  p_reads[[rid]] <- plot_read_footprints(footprints[region_id == rid], reads_r,
                                         read_positions[region_id == rid], m)
  print(p_reads[[rid]])
  ggsave(file.path(plot_dir, sprintf("%s_fig1_read_footprints.pdf", rid)), p_reads[[rid]],
         width = 9, height = max(5, 0.03 * nrow(reads_r) + 2.5 + 0.3 * uniqueN(reads_r$cluster)),
         limitsize = FALSE)
}
```

# 9. Nucleosome occupancy per cluster (Fig 2)

Fraction of reads with a nucleosome footprint covering each position
(`r PROFILE_BIN_BP`-bp bins; denominator = reads spanning the position), one
panel per cluster, timepoints pooled. Distinct profiles = the accessibility
states are also distinct nucleosome-positioning states.

```{r occupancy-cluster, fig.width=7.5, fig.height=9}
occ_cluster <- occupancy_profile(footprints, reads, anchors, by = c("region_id", "cluster"))
fwrite(occ_cluster, file.path(table_dir, "occupancy_by_cluster.tsv"), sep = "\t")

for (rid in names(p_reads)) {
  reads_r <- reads[region_id == rid]
  p <- plot_occupancy_by_cluster(occ_cluster[region_id == rid], reads_r, anchors[region_id == rid])
  print(p)
  ggsave(file.path(plot_dir, sprintf("%s_fig2_occupancy_by_cluster.pdf", rid)), p,
         width = 7.5, height = 1.4 * uniqueN(reads_r$cluster) + 1.4)
}
```

# 10. Positional summaries across time

Per read: midpoint of the nearest upstream (-1) and downstream (+1) nucleosome
and the gap between them (NFR width), by cluster and timepoint. A shift of a
few tens of bp within a cluster is easier to read here than from the tracks.

```{r position-summary, fig.width=10, fig.height=6.5}
position_groups <- position_summary_by_group(read_positions)
fwrite(position_groups, file.path(table_dir, "position_summary_by_cluster_timepoint.tsv"),
       sep = "\t")
position_groups

for (rid in names(p_reads)) {
  p <- plot_position_summary(read_positions[region_id == rid], reads[region_id == rid],
                             anchors[region_id == rid])
  print(p)
  ggsave(file.path(plot_dir, sprintf("%s_position_summary.pdf", rid)), p,
         width = 10, height = 6.5)
}
```

# Session info

```{r session-info}
sessionInfo()
```
