---
title: "TF co-occupancy at enhancers"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: pdf_document
---

# TF occupancy at enhancers over time

Analyses 1–3 use hg38 FANTOM5 enhancer intervals and LPS 0, 5, 10 and 15 min.
All interval tables use **0-based, half-open BED coordinates**. An enhancer is
accessible if **more than 50% of its length overlaps one individual MSP block**
at any timepoint. The denominator is the enhancer length; exactly 50% does not
qualify. Adjacent MSPs or MSPs on different fibers are never joined to meet
this criterion. FANTOM5 outer coordinates define each enhancer (not its BED12
CAGE blocks).

Defaults: compare active and inactive enhancers within the accessible universe;
active means overlap with both baseline H3K27ac and ATAC peaks (at least one
replicate per assay). Inactive means lacking either assay. Single-positive
regions are therefore included in inactive. Baseline classes remain fixed across
the time course. Settings below allow a stricter inactive definition or all
FANTOM5 enhancers. Peak overlap defines activity; mean baseline bigWig signal is
reported separately. Signal differences between these classes are descriptive
because the same assays define the classes.

“TF” here means a short protected footprint, not an identified TF protein.
Co-occupancy means at least two distinct <60-bp footprints on the same fiber
within an enhancer. Motif-based TF identities are not inferred from footprint
size alone. Analysis 5 compares co-occupied site pairs in active and inactive enhancers.

```{r setup, include=FALSE}
project_root <- "/project/spott/cshan/fiber-seq"
output_dir <- file.path(project_root, "macrophage_project/enhancer/TF_co-occ")
plot_dir <- file.path(output_dir, "plots")
table_dir <- file.path(output_dir, "tables")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      cache = FALSE, dev = "pdf", fig.path = paste0(plot_dir, "/"),
                      fig.width = 8, fig.height = 5)
options(scipen = 10)
```

```{r packages}
suppressPackageStartupMessages({
  suppressMessages({
    library(data.table)
    library(GenomicRanges)
    library(Rsamtools)
    library(ggplot2)
    library(dplyr)
    library(tidyr)
    library(rtracklayer)
    library(Matrix)
    source(file.path(project_root, "code/topic_model/topic_modelling_functions.r"))
  })
})
```

```{r settings}
extract_root <- file.path(project_root, "macrophage_project/FiberHMM/extract")
ft_result_dir <- file.path(extract_root, "ft_result_dir")
annotation_dir <- file.path(project_root, "macrophage_project/annotations/enhancer_public_data")
fantom_file <- file.path(annotation_dir, "fantom5_enhancer_hg38/F5.hg38.enhancers.bed.gz")
sample_table <- data.table(sample = paste0("LPS_", c(0, 5, 10, 15)),
                           time_min = c(0L, 5L, 10L, 15L))
standard_chromosomes <- paste0("chr", c(1:22, "X", "Y"))
comparison_universe <- "accessible" # MSP-positive FANTOM5 subset; or "all_fantom"
inactive_definition <- "not_active" # fails active criteria; or "neither_signal"
minimum_peak_replicates <- 1L        # Active requires H3K27ac and ATAC peak overlap in at least one replicate per assay
minimum_reads <- 10L                # full-enhancer-span reads at EVERY timepoint, analysis 3
expected_background <- "genomewide" # expected footprint-length source; or "inactive"
set.seed(1)
```

`comparison_universe = "accessible"` restricts the comparison to FANTOM5 enhancers
passing the MSP rule at one or more timepoints. `"all_fantom"` includes all
FANTOM5 enhancers on the selected chromosomes, including those failing the MSP
rule. Accessibility and baseline activity are separate annotations.

`inactive_definition = "not_active"` puts enhancers failing the active criteria
into the inactive group. `"neither_signal"` requires zero baseline H3K27ac peak
support and zero baseline ATAC peak support; partially supported enhancers are
excluded from the comparison. Active always requires both assays to meet
`minimum_peak_replicates`: 1 allows either replicate per assay, while 2 requires
both replicates per assay. Peak support means any interval overlap, and the
classes remain fixed from the 0-min baseline.

`minimum_reads = 10L` requires at least 10 full-enhancer-span fibers at **each of
all four timepoints** for the co-occupancy frequencies and Fisher comparisons in
analysis 3. This keeps a consistent enhancer set across time. It does not filter
the MSP universe, ACF clustering or the footprint-length summaries.

`expected_background = "genomewide"` uses the same timepoint's genome-wide TF
(<60 bp) and nucleosome (>90 bp) calls for expected footprint-length frequencies
and the expected TF:nucleosome ratio. `"inactive"` uses those calls at the selected
inactive enhancers instead. This setting does not change the active-versus-inactive
Fisher comparison.

# Shared input and output functions


```{r save-table-function}
save_table <- function(x, filename) {
  path <- file.path(table_dir, filename)
  data.table::fwrite(x, path, sep = "\t", na = "NA")
  reloaded <- data.table::fread(path)
  print(utils::head(reloaded, 6L))
  invisible(reloaded)
}
```

```{r save-pdf-function}
save_pdf <- function(plot, filename, width = 8, height = 5) {
  stopifnot(grepl("\\.pdf$", filename))
  ggplot2::ggsave(file.path(plot_dir, filename), plot = plot,
                 device = "pdf", width = width, height = height, limitsize = FALSE)
  invisible(plot)
}
```

```{r footprint-functions}
source(file.path(project_root, "code/enhancer/co_occ_functions/run_enhancer_footprints.R"))
```


```{r hmm-file-function}
hmm_file <- function(sample, chr, kind) {
  sample_dir <- gsub("_", "", sample)
  file.path(extract_root, paste0("firehmm_", kind), sample_dir,
            paste0(sample_dir, "_hmm_extracted_", kind, "_", chr, ".bed.gz"))
}
```

# 1. Define the accessible enhancer universe

```{r load-fantom}
enhancers <- fread(fantom_file, header = FALSE, select = 1:4,
                   col.names = c("chr", "start0", "end0", "enhancer_id"))
enhancers <- unique(enhancers[chr %in% standard_chromosomes], by = c("chr", "start0", "end0"))
stopifnot(!anyDuplicated(enhancers$enhancer_id))
enhancers[, midpoint0 := as.integer(floor((start0 + end0) / 2))]
head(enhancers)
```

Submit the four-timepoint accessibility job, then load its tables after it
finishes. Input paths, samples, and chromosomes are configured in the script.

```{r submit-accessible-universe, eval=FALSE}
accessibility_script <- file.path(project_root, "code/enhancer/shell_scripts/find_accessible_enhancers.sh")
accessibility_status <- system2("sbatch", shQuote(accessibility_script))
if (accessibility_status != 0L) stop("Slurm accessibility job submission failed")
```

`sbatch /project/spott/cshan/fiber-seq/code/enhancer/shell_scripts/find_accessible_enhancers.sh`.

```{r load-accessible-universe}
accessibility_paths <- file.path(table_dir,
  c("01_enhancer_accessibility.tsv", "01_accessible_counts.tsv"))
if (!all(file.exists(accessibility_paths)))
  stop("Accessibility tables are missing. Submit the Slurm job and wait for it to finish.")
accessibility_export <- fread(file.path(table_dir, "01_enhancer_accessibility.tsv"))
stopifnot(!anyDuplicated(accessibility_export$enhancer_id),
          all(enhancers$enhancer_id %in% accessibility_export$enhancer_id))
head(accessibility_export)
accessibility <- accessibility_export[match(enhancers$enhancer_id, enhancer_id),
  c("enhancer_id", sample_table$sample, "accessible"), with = FALSE]
enhancers[, accessible := accessibility$accessible]
accessibility_summary <- fread(file.path(table_dir, "01_accessible_counts.tsv"))
head(accessibility_summary)
```

# 2. Chromatin signatures at baseline active and inactive enhancers

```{r baseline-files-function}
baseline_files <- function(directory, suffix) {
  paths <- list.files(directory, pattern = paste0("_0000_.*", suffix, "$"), full.names = TRUE)
  if (length(paths) != 2L) stop("Expected two baseline replicates: ", directory, " / ", suffix)
  paths
}
```

```{r baseline-peak-support-function}
baseline_peak_support <- function(enhancers, files) {
  support <- integer(nrow(enhancers))
  for (path in files) {
    peaks <- fread(path, header = FALSE, select = 1:3,
                   col.names = c("chr", "start0", "end0"))
    support <- support + (countOverlaps(bed_ranges(enhancers), bed_ranges(peaks)) > 0L)
  }
  support
}
```

```{r baseline-signal-function}
baseline_signal <- function(enhancers, files) {
  signal <- matrix(NA_real_, nrow(enhancers), length(files))
  for (j in seq_along(files)) {
    bw <- rtracklayer::BigWigFile(files[j])
    available <- names(GenomeInfoDb::seqlengths(bw))
    for (chrom in intersect(unique(enhancers$chr), available)) {
      idx <- which(enhancers$chr == chrom)
      query <- bed_ranges(enhancers[idx])
      track <- rtracklayer::import(bw, which = reduce(query), as = "GRanges")
      # Uncovered bases on a represented chromosome contribute zero signal;
      # an absent chromosome is NA, not zero.
      hits <- findOverlaps(query, track)
      signal[idx, j] <- 0
      if (length(hits)) {
        q <- queryHits(hits)
        t <- subjectHits(hits)
        bp <- width(pintersect(query[q], track[t]))
        sums <- rowsum(bp * mcols(track)$score[t], q, reorder = FALSE)
        k <- as.integer(rownames(sums))
        signal[idx[k], j] <- as.numeric(sums) / width(query)[k]
      }
    }
  }
  values <- rowMeans(signal, na.rm = TRUE)
  values[is.nan(values)] <- NA_real_
  values
}
```

```{r classify-enhancers}
# Reload saved flags so this chunk also works after rerunning load-fantom.
accessibility_flags <- fread(
  file.path(table_dir, "01_enhancer_accessibility.tsv"),
  select = c("enhancer_id", "accessible")
)
if (anyDuplicated(accessibility_flags$enhancer_id))
  stop("Duplicate enhancer IDs in the accessibility table.")
accessibility_idx <- match(enhancers$enhancer_id, accessibility_flags$enhancer_id)
if (anyNA(accessibility_idx))
  stop("Some enhancer IDs are missing from the accessibility table.")
accessible_values <- as.logical(accessibility_flags$accessible[accessibility_idx])
if (anyNA(accessible_values)) stop("Missing or invalid accessibility flags.")
enhancers[, accessible := accessible_values]

chip_dir <- file.path(annotation_dir, "chip_hg38")
atac_dir <- file.path(annotation_dir, "atac_hg38")
enhancers[, h3k27ac_reps := baseline_peak_support(.SD, baseline_files(chip_dir, "peaks\\.narrowPeak"))]
enhancers[, atac_reps := baseline_peak_support(.SD, baseline_files(atac_dir, "peaks\\.narrowPeak"))]
enhancers[, enhancer_class := fifelse(h3k27ac_reps >= minimum_peak_replicates &
                                       atac_reps >= minimum_peak_replicates, "active", "inactive")]
if (inactive_definition == "neither_signal") {
  enhancers[enhancer_class == "inactive" & (h3k27ac_reps > 0 | atac_reps > 0),
            enhancer_class := "excluded"]
}
stopifnot(comparison_universe %in% c("accessible", "all_fantom"),
          inactive_definition %in% c("not_active", "neither_signal"))
analysis_enhancers <- copy(enhancers[enhancer_class %in% c("active", "inactive") &
                                    (accessible | comparison_universe == "all_fantom")])
if (!nrow(analysis_enhancers)) stop("No enhancers meet the selected universe/class criteria")
analysis_enhancers[, h3k27ac_signal := baseline_signal(.SD, baseline_files(chip_dir, "\\.bw"))]
analysis_enhancers[, atac_signal := baseline_signal(.SD, baseline_files(atac_dir, "\\.bw"))]
save_table(analysis_enhancers[, .(enhancer_id, chr, start0, end0, accessible,
                                 enhancer_class, h3k27ac_reps, atac_reps,
                                 h3k27ac_signal, atac_signal)], "02_enhancer_classes.tsv")
class_summary <- analysis_enhancers[, .(n_enhancers = .N), by = .(accessible, enhancer_class)]
save_table(class_summary, "02_class_counts.tsv")
```

### Sample a balanced enhancer set

Sample 5,000 active and 5,000 inactive enhancers without replacement, using seed
1. The full eligible classes remain in `02_enhancer_classes.tsv`; this chunk
reloads that pool so rerunning it selects the same enhancers. All subsequent
analyses use this sampled set across all four timepoints. The later read-coverage
and ACF quality filters can reduce the number of usable enhancers or fibers.

```{r sample-analysis-enhancers}
enhancers_per_class <- 5000L
enhancer_sampling_seed <- 1L
analysis_pool <- fread(file.path(table_dir, "02_enhancer_classes.tsv"))
stopifnot(!anyDuplicated(analysis_pool$enhancer_id))
available_classes <- analysis_pool[, .(n_enhancers = .N), by = enhancer_class]
if (!setequal(available_classes$enhancer_class, c("active", "inactive")) ||
    any(available_classes$n_enhancers < enhancers_per_class))
  stop("Need at least ", enhancers_per_class, " enhancers in each of the active and inactive classes.")
data.table::setorder(analysis_pool, enhancer_class, enhancer_id)
set.seed(enhancer_sampling_seed)
analysis_enhancers <- analysis_pool[
  , .SD[sample.int(.N, enhancers_per_class, replace = FALSE)], by = enhancer_class]
analysis_enhancers[, midpoint0 := as.integer(floor((start0 + end0) / 2))]
save_table(analysis_enhancers, "02_sampled_enhancer_classes.tsv")
sampled_class_summary <- analysis_enhancers[
  , .(n_enhancers = .N), by = .(accessible, enhancer_class)]
save_table(sampled_class_summary, "02_sampled_class_counts.tsv")
```

### Aggregate FIRE profiles around sampled enhancer midpoints

For each track, compare active and inactive enhancers in separate panels for
LPS 0, 5, 10 and 15 minutes. Use the same sampled enhancers at every timepoint
and the same midpoint-centered, 1,000-bp windows as the ACF analysis: BED offsets
[-500, 500), at 1-bp resolution. Each enhancer contributes equally to its class
mean at each position. Recorded zeros are included; missing bigWig values and
windows extending outside a represented chromosome are excluded, with the
number of contributing enhancers saved for each position.

`all.percent.accessible.bw` contains the percentage of fibers carrying a FIRE
at each position. The FIRE, linker/MSP and nucleosome coverage tracks contain
fiber counts; these raw coverage values also depend on sequencing depth.

```{r fire-profile-settings}
fire_bw_root <- "/project/spott/lizarraga/pacbio_analysis/macrophage_project/merged_hifi_bams/FIRE"
fire_profile_flank <- 500L
fire_profile_tracks <- data.table(
  track = c("percent_accessible", "fire_coverage", "linker_coverage", "nucleosome_coverage"),
  filename = c("all.percent.accessible.bw", "all.fire.coverage.bw",
               "all.linker.coverage.bw", "all.nucleosome.coverage.bw"),
  title = c("FIRE accessibility", "FIRE coverage", "Linker/MSP coverage", "Nucleosome coverage"),
  y_label = c("Mean accessible fibers (%)", rep("Mean coverage (fibers)", 3L))
)
```

```{r aggregate-fire-profile-function}
aggregate_fire_profile <- function(regions, path, flank = 500L) {
  stopifnot(file.exists(path), flank > 0L,
            !anyDuplicated(regions$enhancer_id),
            !anyNA(regions[, .(chr, midpoint0, enhancer_class)]))
  bw <- rtracklayer::BigWigFile(path)
  chromosome_lengths <- GenomeInfoDb::seqlengths(GenomeInfoDb::seqinfo(bw))
  left0 <- regions$midpoint0 - flank
  right0 <- regions$midpoint0 + flank
  chromosome_end <- unname(chromosome_lengths[regions$chr])
  valid <- !is.na(chromosome_end) & left0 >= 0L & right0 <= chromosome_end
  signal <- matrix(NA_real_, nrow(regions), 2L * flank)
  for (chromosome in unique(regions$chr[valid])) {
    idx <- which(valid & regions$chr == chromosome)
    query <- GenomicRanges::GRanges(chromosome,
      IRanges::IRanges(left0[idx] + 1L, right0[idx]))
    # One summary value per genomic base; retain absent values as NA.
    signal[idx, ] <- summary(bw, which = query, size = 2L * flank,
                             type = "mean", defaultValue = NA_real_, as = "matrix")
  }
  signal[!is.finite(signal)] <- NA_real_
  data.table::rbindlist(lapply(c("active", "inactive"), function(class_id) {
    idx <- which(regions$enhancer_class == class_id)
    values <- signal[idx, , drop = FALSE]
    n <- colSums(!is.na(values))
    means <- colMeans(values, na.rm = TRUE)
    means[n == 0L] <- NA_real_
    data.table(enhancer_class = class_id, position_bp = seq.int(-flank, flank - 1L),
               mean_signal = means, n_enhancers = n,
               n_sampled = length(idx), n_windows_in_bounds = sum(valid[idx]))
  }))
}
```

```{r calculate-fire-profiles}
fire_profile_parts <- list()
for (i in seq_len(nrow(fire_profile_tracks))) {
  for (j in seq_len(nrow(sample_table))) {
    bw_path <- file.path(fire_bw_root, sample_table$sample[j], "trackHub-v0.1/bw",
                         fire_profile_tracks$filename[i])
    message("Profiling ", sample_table$sample[j], ": ", fire_profile_tracks$track[i])
    profile <- aggregate_fire_profile(analysis_enhancers, bw_path, fire_profile_flank)
    profile[, `:=`(sample = sample_table$sample[j], time_min = sample_table$time_min[j],
                   track = fire_profile_tracks$track[i])]
    fire_profile_parts[[length(fire_profile_parts) + 1L]] <- profile
  }
}
fire_profiles <- rbindlist(fire_profile_parts)
save_table(fire_profiles, "02_fire_aggregate_profiles.tsv")
```

```{r plot-fire-profiles, fig.width=10, fig.height=7}
for (i in seq_len(nrow(fire_profile_tracks))) {
  plot_data <- copy(fire_profiles[track == fire_profile_tracks$track[i]])
  plot_data[, timepoint := factor(paste0(time_min, " min"),
                                  levels = paste0(sample_table$time_min, " min"))]
  p_fire_profile <- ggplot(plot_data,
    aes(position_bp, mean_signal, color = enhancer_class)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey65") +
    geom_line(linewidth = 0.7, na.rm = TRUE) +
    facet_wrap(~timepoint, ncol = 2) +
    scale_color_manual(values = c(active = "#D55E00", inactive = "#0072B2")) +
    scale_x_continuous(breaks = c(-fire_profile_flank, 0, fire_profile_flank),
                       limits = c(-fire_profile_flank, fire_profile_flank)) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = fire_profile_tracks$title[i],
         x = "Position relative to enhancer midpoint (bp)",
         y = fire_profile_tracks$y_label[i], color = NULL,
         caption = "Mean across sampled enhancers; missing values excluded.") +
    theme_classic() + theme(legend.position = "bottom")
  print(p_fire_profile)
  save_pdf(p_fire_profile, paste0("02_aggregate_", fire_profile_tracks$track[i], ".pdf"),
           width = 10, height = 7)
}
```

```{r baseline-signal-plot}
signal_long <- melt(analysis_enhancers, id.vars = c("enhancer_id", "enhancer_class"),
                    measure.vars = c("h3k27ac_signal", "atac_signal"),
                    variable.name = "assay", value.name = "signal")
head(signal_long)
p_signal <- ggplot(signal_long, aes(enhancer_class, signal, fill = enhancer_class)) +
  geom_boxplot(outlier.shape = NA) + facet_wrap(~assay, scales = "free_y") +
  scale_y_continuous(trans = "log1p") + theme_classic() +
  labs(x = NULL, y = "Mean baseline bigWig signal (log1p scale)") +
  theme(legend.position = "none")
save_pdf(p_signal, "02_baseline_H3K27ac_ATAC.pdf")

p_signal
```


### Share capped fibers across joint active/inactive clustering

ACF and Manhattan use **one saved fiber sample**, prepared once and loaded by
both methods. Sampling is performed separately within each baseline enhancer
class, then the selected active and inactive fibers are combined and clustered
**together**. The default remains **up to 10,000 fibers per class**: up to
20,000 rows in each pooled analysis, with exactly the same ordered fiber IDs in
both methods. Active/inactive labels remain metadata and do not determine the
clusters.

For each class, extract all fibers fully spanning the enhancer midpoint +/-
500 bp across LPS 0, 5, 10 and 15 min. With seed 1, randomly retain at most
**3 fibers per enhancer × timepoint**, then retain at most **10 per enhancer**
using round-robin selection across timepoints. Each round randomizes timepoint
order and skips exhausted timepoints. Draw a uniform sample of up to 10,000
fibers without replacement from that capped class pool. If fewer remain, use
all of them and report the shortfall without relaxing either cap. The caps limit
enhancer contributions; they do not impose equal final timepoint totals.

Each row contains **1,000 consecutive genomic bases (-500 through +499)**,
with 1 for an m6A call and 0 otherwise, including G/C bases. There is no binning,
smoothing, imputation or A/T normalization. A molecule spanning multiple
windows contributes a separate enhancer-specific row, subject to the caps.

```{r enhancer-pooled-job-files}
shared_cap_per_timepoint <- 3L
shared_cap_per_enhancer <- 10L
shared_n_fibers_per_class <- 10000L
shared_sampling_seed <- 1L
pooled_output_suffix <- "_pooled_capped"
manhattan_k_neighbors <- 100L
manhattan_neighbor_method <- "approximate" # or "exact", which takes quadratic time
shared_fibers_file <- Sys.getenv("ENHANCER_SHARED_FIBERS",
  file.path(table_dir, "enhancer_shared_fibers_pooled_capped.rds"))
acf_results_file <- Sys.getenv("ENHANCER_ACF_OUTPUT",
  file.path(table_dir, "enhancer_acf_results_pooled_capped.rds"))
manhattan_results_file <- Sys.getenv("ENHANCER_MANHATTAN_OUTPUT",
  file.path(table_dir, "enhancer_manhattan_results_pooled_capped.rds"))
pooled_submission_script <- file.path(project_root,
  "code/enhancer/shell_scripts/submit_enhancer_pooled_jobs.sh")
pooled_job_id_file <- file.path(table_dir, "enhancer_pooled_job_ids.tsv")
pooled_jobs <- if (file.exists(pooled_job_id_file)) fread(pooled_job_id_file) else
  data.table(stage = c("prepare", "manhattan", "acf"), job_id = NA_character_)
pooled_jobs
```

The submission helper launches one preparation job, then one pooled Manhattan
job and one pooled ACF job, each depending on successful preparation. The method
jobs never resample fibers. Preparation writes the shared RDS, an ordered
`shared_fiber_manifest_pooled_capped.tsv`, sampling diagnostics and extraction
QC. Each method records that RDS's checksum and sample ID alongside its results.
All new analysis outputs use `_pooled_capped` filenames.

Submission is manual (`eval=FALSE`): run the following chunk **once** when new
results are needed, and wait for the recorded jobs before loading results.
Knitting does not submit jobs.

```{r submit-enhancer-pooled-jobs, eval=FALSE}
stopifnot(manhattan_neighbor_method %in% c("approximate", "exact"))
pooled_submission <- system2("bash", shQuote(pooled_submission_script),
  stdout = TRUE, stderr = TRUE, env = c(
    paste0("ENHANCER_CAP_PER_TIMEPOINT=", shared_cap_per_timepoint),
    paste0("ENHANCER_CAP_PER_ENHANCER=", shared_cap_per_enhancer),
    paste0("ENHANCER_N_FIBERS_PER_CLASS=", shared_n_fibers_per_class),
    paste0("ENHANCER_SAMPLING_SEED=", shared_sampling_seed),
    paste0("ENHANCER_SHARED_FIBERS=", shQuote(shared_fibers_file)),
    paste0("ENHANCER_ACF_OUTPUT=", shQuote(acf_results_file)),
    paste0("ENHANCER_MANHATTAN_OUTPUT=", shQuote(manhattan_results_file)),
    paste0("ENHANCER_MANHATTAN_NEIGHBORS=", manhattan_neighbor_method)))
cat(pooled_submission, sep = "\n")
if (!is.null(attr(pooled_submission, "status")))
  stop("Pooled enhancer job submission failed; inspect the submission log and job record.")
pooled_jobs <- fread(pooled_job_id_file)
pooled_jobs
```

```{r load-shared-enhancer-fibers}
shared_sampling_helpers <- new.env(parent = globalenv())
sys.source(file.path(project_root, "code/enhancer/shared_functions/enhancer_shared_sampling.R"),
           envir = shared_sampling_helpers)
if (!file.exists(shared_fibers_file))
  stop("Shared fibers are missing. Wait for the pooled preparation job to finish.")
shared_enhancer_fibers <- shared_sampling_helpers$load_shared_enhancer_fibers(shared_fibers_file)
pooled_analysis_enhancers <- as.data.table(shared_enhancer_fibers$inputs$regions)
pooled_samples <- as.character(shared_enhancer_fibers$inputs$samples)
shared_fiber_sampling <- copy(as.data.table(shared_enhancer_fibers$fiber_sampling))
shared_sampling_diagnostics <- copy(as.data.table(shared_enhancer_fibers$sampling_diagnostics))
shared_fiber_sampling
shared_sampling_diagnostics
```

`pooled_analysis_enhancers` and the saved sampling parameters describe the
actual preparation input. They are independent of the current interactive
`analysis_enhancers`; loading completed results does not resample or change the
inputs for other analyses. The preparation TSVs report fibers and distinct
enhancers before capping, after each cap and after final selection, including
counts by timepoint and minimum/median/maximum fibers per contributing enhancer.
Enhancers contributing zero fibers at a stage are excluded from that stage's
per-enhancer distribution.

```{r validate-pooled-enhancer-results-function}
validate_pooled_enhancer_results <- function(saved, shared, method) {
  if (!identical(names(saved$results), "pooled"))
    stop(method, " results must contain one joint active/inactive clustering.")
  if (!identical(saved$shared_sample$md5, shared$md5) ||
      !identical(saved$shared_sample$sample_id, shared$sample_id))
    stop(method, " results were generated from a different shared fiber file.")
  sampling_parameters <- c("cap_per_timepoint", "cap_per_enhancer",
    "n_fibers_per_class", "sampling_seed", "sampling_method", "position_bp", "signal")
  if (!all(sampling_parameters %in% names(saved$parameters)) ||
      !identical(saved$parameters[sampling_parameters], shared$parameters[sampling_parameters]) ||
      !identical(saved$inputs$regions, shared$inputs$regions) ||
      !identical(saved$inputs$samples, shared$inputs$samples))
    stop(method, " input snapshot or sampling parameters do not match shared preparation.")
  assignments <- as.data.table(saved$results$pooled$assignments)
  metadata <- as.data.table(shared$metadata)
  for (field in c("RID", "enhancer_id", "enhancer_class", "sample_name")) {
    if (!field %in% names(assignments) ||
        !identical(as.character(assignments[[field]]), as.character(metadata[[field]])))
      stop(method, " does not preserve the complete ordered shared sample: ", field)
  }
  stopifnot(nrow(assignments) == nrow(shared$mat), !anyDuplicated(assignments$RID))
  invisible(TRUE)
}
```

### Pool single-fiber autocorrelation profiles across both enhancer classes

The ACF worker imports `03_compute_autocorrelations.py` and
`04_cluster_autocorrelations.py` directly from
`code/clustering_methods/auto_correlation` through `reticulate`. Those files and
the numerical method remain unchanged. For each shared binary fiber, the first
function computes the mean-centered, variance-normalized linear ACF:

\[
\mathrm{ACF}(k) =
\frac{\sum_{t=0}^{999-k}(x_t-\bar{x})(x_{t+k}-\bar{x})}
     {\sum_{t=0}^{999}(x_t-\bar{x})^2},\qquad k=0,\ldots,999.
\]

All 1,000 lag features remain at **1-bp resolution**. Constant fibers, including
all-zero fibers, have undefined ACF. They remain in the shared sample, assignments
and QC as unclustered rows, and are excluded from the ACF clustering computation.
They are not replaced with other fibers. Lag 0 remains a clustering feature and
becomes constant zero when PCA centers the columns.

The reference clustering uses centered, unscaled PCA (up to 50 PCs),
correlation-distance neighbors on the PCA scores (10 neighbors), the UMAP
connectivity graph and Leiden at resolution 0.4, with clustering seed 0. UMAP
coordinates are for display. ACF calculation is batched; PCA/neighbors holds the
valid pooled sample in memory. The clustering seed is distinct from the shared
sampling seed of 1. Active and inactive fibers enter **one** clustering, so an
ACF cluster number has the same meaning in both class panels.

```{r load-enhancer-acf-results}
if (!file.exists(acf_results_file))
  stop("Pooled ACF results are missing. Wait for the ACF job to finish.")
acf_saved <- readRDS(acf_results_file)
validate_pooled_enhancer_results(acf_saved, shared_enhancer_fibers, "ACF")
stopifnot(acf_saved$parameters$n_features == 1000L,
          acf_saved$parameters$n_pcs == 50L,
          acf_saved$parameters$n_neighbors == 10L,
          acf_saved$parameters$resolution == 0.4,
          acf_saved$parameters$seed == 0L)
enhancer_acf_results <- acf_saved$results
acf_shared_sample <- acf_saved$shared_sample
acf_assignments <- as.data.table(enhancer_acf_results$pooled$assignments)
acf_assignments[, .(n_fibers = .N, n_enhancers = uniqueN(enhancer_id)),
                by = .(enhancer_class, sample_name, acf_status)]
enhancer_acf_results$pooled$cluster_diagnostics
```

The batch job saves tables and PDFs automatically. This chunk regenerates those
outputs from the saved results without extracting, sampling or clustering again.
Profiles show the pooled cluster means, and composition panels retain each fiber's
true enhancer class and timepoint. Unclustered fibers remain in the composition
denominators. UMAP plots color the joint embedding by cluster, timepoint and
active/inactive class. ACF curves show every positive integer lag without binning
or smoothing; the saved profile tables also contain lag 0.

```{r enhancer-acf-pdfs, results='hide'}
acf_plot_helpers <- new.env(parent = globalenv())
sys.source(file.path(project_root, "code/enhancer/acf/enhancer_acf_plots.R"),
           envir = acf_plot_helpers)
acf_plot_outputs <- acf_plot_helpers$save_enhancer_acf_outputs(
  enhancer_acf_results$pooled, plot_dir, table_dir,
  samples = pooled_samples, suffix = pooled_output_suffix)
```

```{r enhancer-acf-summary-tables}
acf_cluster_counts <- fread(file.path(table_dir,
  paste0("acf_cluster_counts", pooled_output_suffix, ".tsv")))
acf_qc_summary <- fread(file.path(table_dir,
  paste0("acf_qc", pooled_output_suffix, ".tsv")))
acf_qc_summary
head(acf_cluster_counts)
```

### Pool unbinned single-fiber patterns for Manhattan–Leiden clustering

Manhattan clusters the **same saved binary rows**, combining active and inactive
fibers once. The representation, Manhattan metric, neighbor construction, edge
weighting and Leiden settings remain unchanged. All-zero and other constant
fibers remain eligible because Manhattan distance is defined for them.

For two fibers, Manhattan distance counts the midpoint-relative bases where
binary calls differ. The graph follows
`code/clustering_methods/Leiden_Manhattan/marcophage_leiden_manhattan.Rmd`, with
**100 neighbors per fiber**, edge weight `exp(-distance / sigma)`, where `sigma`
is the mean directed neighbor distance (1 if all distances are zero), and an
undirected union graph. Leiden uses the original `leiden_partition()` function,
the modularity objective, resolution 1, seed 1 and iterations to convergence.
Clusters are numbered largest first across the pooled sample. Active/inactive
panels refer to those same joint cluster labels. Manhattan and ACF cluster
numbers do not have a correspondence between methods.

The default uses **approximate Manhattan neighbors**, with the selected edge
distances recomputed exactly from the original 1-bp patterns. Approximate search
can select different neighbors from exact search, including ties among identical
patterns. The optional exact mode computes every pairwise distance in blocks
with the reference's input-order tie handling and still takes quadratic time.
Both modes use the shared sample. Fewer than three pooled fibers would be
reported as unclustered.

```{r load-enhancer-manhattan-results}
if (!file.exists(manhattan_results_file))
  stop("Pooled Manhattan results are missing. Wait for the Manhattan job to finish.")
manhattan_saved <- readRDS(manhattan_results_file)
validate_pooled_enhancer_results(manhattan_saved, shared_enhancer_fibers, "Manhattan")
stopifnot(manhattan_saved$parameters$k_neighbors == manhattan_k_neighbors,
          identical(manhattan_saved$parameters$neighbor_method, manhattan_neighbor_method))
enhancer_manhattan_results <- manhattan_saved$results
manhattan_shared_sample <- manhattan_saved$shared_sample
manhattan_assignments <- as.data.table(enhancer_manhattan_results$pooled$assignments)
# ACF keeps undefined/constant fibers as unclustered rows, so this checks the
# entire selected population rather than only the ACF-valid clustering subset.
manhattan_acf_same_fibers <- identical(as.character(manhattan_assignments$RID),
                                      as.character(acf_assignments$RID))
stopifnot(manhattan_acf_same_fibers,
          identical(manhattan_shared_sample$md5, acf_shared_sample$md5),
          identical(manhattan_shared_sample$sample_id, acf_shared_sample$sample_id))
message("Verified identical shared input for ACF and Manhattan: ",
        nrow(manhattan_assignments), " ordered fibers; checksum ",
        manhattan_shared_sample$md5)
manhattan_assignments[, .(n_fibers = .N, n_enhancers = uniqueN(enhancer_id)),
                      by = .(enhancer_class, sample_name, clustering_status)]
```

Both methods export cluster diagnostics: fiber count, number of contributing
enhancers, largest single-enhancer fraction and timepoint counts. Clusters are
flagged if any enhancer contributes **more than 10%** of their fibers or if
**fewer than 50 distinct enhancers** contribute. These diagnostic flags do not
remove clusters or fibers. Sampling diagnostics preserve active/inactive class
labels; `pooled` describes the fit, not a replacement for enhancer class.

```{r enhancer-manhattan-summary-tables}
manhattan_sampling <- copy(as.data.table(enhancer_manhattan_results$pooled$fiber_sampling))
manhattan_sampling_diagnostics <- copy(as.data.table(
  enhancer_manhattan_results$pooled$sampling_diagnostics))
manhattan_cluster_diagnostics <- copy(as.data.table(
  enhancer_manhattan_results$pooled$cluster_diagnostics))
manhattan_sampling
manhattan_sampling_diagnostics
manhattan_cluster_diagnostics
```

The m6A heatmap includes every selected fiber, grouped by the joint Manhattan
cluster. Footprint profiles use all selected fibers and retain FiberHMM
nucleosomes **>90 bp** and TF footprints **<60 bp**, without score or
FIRE-containment filters. Single-fiber PDFs stack **all resulting clusters in
one figure**, with labeled blocks showing displayed/total fiber counts and
cluster, timepoint and enhancer-class color bars. Fibers are ordered by timepoint
within each cluster. There is no display subsampling, and the new Leiden fit
determines the number of clusters. All tracks align to the enhancer midpoint at
1-bp resolution; profile curves are unsmoothed. Composition plots show the
fraction assigned to each joint cluster within each enhancer class and timepoint.

The batch job saves these PDFs automatically. Run this chunk to regenerate them
from the saved pooled result without repeating extraction or clustering.

```{r enhancer-manhattan-pdfs, results='hide'}
suppressPackageStartupMessages({library(cowplot); library(grid); library(ComplexHeatmap)})
manhattan_plot_helpers <- new.env(parent = globalenv())
sys.source(file.path(project_root,
  "code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r"),
  envir = manhattan_plot_helpers)
sys.source(file.path(project_root, "code/enhancer/manhattan/enhancer_manhattan_plots.R"),
           envir = manhattan_plot_helpers)
manhattan_plot_helpers$save_enhancer_manhattan_pdfs(
  enhancer_manhattan_results$pooled, "pooled", plot_dir, table_dir,
  expected_k = manhattan_k_neighbors, suffix = "_capped")
```

### Extract all TF footprints shorter than 60 bp

This shell script expands every block in the complete FiberHMM TF BED12 files,
retains lengths 1–59 bp, and writes sorted, bgzipped BED6 files with tabix indexes.
The BED6 columns match `split_footprints_by_size.sh`:
`chrom`, `start`, `end`, `read_name`, `size`, `strand`.

Outputs use the existing TF folder layout:
`firehmm_tf/ft_by_size/size0-60/LPS0/LPS0_tf_size0-60_chr1.bed.gz`
(and the corresponding chromosomes/timepoints). These contain the same short
TF calls selected from the parent files by the counting functions below.
The R chunk saves the standalone script in `code/enhancer/shell_scripts`.
The submission chunk is manual so knitting does not submit another extraction job.
This minimal script runs as a four-task Slurm array, skips completed BED/index
pairs, and writes task-specific logs in `code/enhancer/logs`.

```{r function-write-ft-less-60bp-script}
write_ft_less_60bp_script <- function(path) {
  script <- r"---(#!/bin/bash
#SBATCH --job-name=extract_tf_footprints_lt60bp
#SBATCH --account=pi-spott
#SBATCH --partition=caslake
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem=32G
#SBATCH --time=12:00:00
#SBATCH --array=1-4
#SBATCH --output=/project/spott/cshan/fiber-seq/code/enhancer/logs/out_%A_%a.out
#SBATCH --error=/project/spott/cshan/fiber-seq/code/enhancer/logs/err_%A_%a.err

set -euo pipefail
shopt -s failglob
samples=(LPS0 LPS5 LPS10 LPS15)
[[ ${SLURM_ARRAY_TASK_ID:-} =~ ^[1-4]$ ]] || { echo "Submit with sbatch" >&2; exit 1; }
sample=${samples[SLURM_ARRAY_TASK_ID-1]}
root=${TF_EXTRACT_ROOT:-/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract}
bin=/project/spott/cshan/envs/dimelo/bin
outdir="$root/firehmm_tf/ft_by_size/size0-60/$sample"
mkdir -p "$outdir"
workdir=$(mktemp -d "$outdir/.tf_lt60.XXXXXX")
trap 'rm -rf -- "$workdir"' EXIT

for input in "$root/firehmm_tf/$sample/${sample}_hmm_extracted_tf_"*.bed.gz; do
    name=${input##*/}
    output="$outdir/${name/_hmm_extracted_tf_/_tf_size0-60_}"
    [[ -s "$output" && -s "$output.tbi" ]] && continue

    gzip -cd "$input" | awk '
      BEGIN { FS = OFS = "\t" }
      {
        if (NF < 12) exit 1
        sub(/,$/, "", $11); sub(/,$/, "", $12)
        n = split($11, sizes, ","); m = split($12, offsets, ",")
        if ($10 > 0 && (n != $10 || m != $10)) exit 1
        for (i = 1; i <= $10; i++)
          if (sizes[i] > 0 && sizes[i] < 60) {
            start = $2 + offsets[i]
            print $1, start, start + sizes[i], $4, sizes[i], $6
          }
      }' | LC_ALL=C sort -k1,1 -k2,2n -k3,3n -S 2G -T "$workdir" \
         | "$bin/bgzip" -@ "${SLURM_CPUS_PER_TASK:-4}" -c > "$workdir/calls.bed.gz"
    "$bin/tabix" -p bed "$workdir/calls.bed.gz"
    mv "$workdir/calls.bed.gz" "$output"
    mv "$workdir/calls.bed.gz.tbi" "$output.tbi"
    echo "Wrote: $output"
done
)---"
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  writeLines(script, path)
  Sys.chmod(path, "0755")
  invisible(path)
}
```

```{r save-ft-less-60bp-script}
ft_less_60bp_script <- file.path(project_root, "code/enhancer/shell_scripts/extract_ft_less_60bp.sh")
write_ft_less_60bp_script(ft_less_60bp_script)
```

```{r submit-ft-less-60bp-extraction, eval=FALSE}
# One Slurm array task per timepoint: LPS0, LPS5, LPS10 and LPS15.
system2("sbatch", shQuote(ft_less_60bp_script))
```

### Count enhancer footprints

The existing size bins omit footprints below 10 bp and from 30–39 bp. Therefore,
the counts below use their complete parent FiberHMM BED12 files and apply the
requested strict cutoffs: TF footprints <60 bp and nucleosome footprints >90 bp.
The 60–90-bp interval is excluded, including footprints exactly 60 or 90 bp.
Every FiberHMM block is a real call; these files have no `ft extract` sentinels.

A footprint is assigned to an enhancer when its midpoint lies inside that
enhancer. This avoids favoring long footprints simply because they overlap more
bases. Only reads spanning the entire enhancer contribute. Read denominators
come from the matching, unfiltered m6A extraction and include reads with zero TF
footprints. Two distinct short footprints on one enhancer–read observation are
called co-occupied; these length-defined footprints do not identify named TFs.


The enhancer counts and background scan run in a single Slurm job covering all
four timepoints. It requests 4 CPUs, 64G memory and 30 hours on `caslake`.
Both the notebook and worker use the functions in `co_occ_functions/run_enhancer_footprints.R`.

Run the submission chunk manually after defining `analysis_enhancers` and the
settings above. It saves a separate input snapshot for each submission, preserving
the current enhancer selection, timepoints, chromosome list and background choice.
Knitting does not submit a job. After Slurm reports completion, run the loading
chunk to restore counts and continue with the analyses below.

```{r footprint-job-files}
source(file.path(project_root, "code/enhancer/co_occ_functions/run_enhancer_footprints.R"))
footprint_script <- file.path(project_root, "code/enhancer/shell_scripts/run_enhancer_footprints.sh")
footprint_results_file <- file.path(table_dir, "enhancer_footprint_results.rds")
footprint_job_id_file <- file.path(table_dir, "footprint_job_id.txt")
```

```{r submit-enhancer-footprints, eval=FALSE}
footprint_config_file <- tempfile("enhancer_footprint_inputs_", tmpdir = table_dir, fileext = ".rds")
saveRDS(list(
  inputs = footprint_job_inputs(analysis_enhancers, sample_table, standard_chromosomes,
                                extract_root, ft_result_dir, expected_background),
  result_file = footprint_results_file,
  counts_file = file.path(table_dir, "enhancer_footprint_counts.tsv")
), footprint_config_file)
footprint_submission <- system2("sbatch", c("--parsable", shQuote(footprint_script),
                                            shQuote(footprint_config_file)), stdout = TRUE)
if (!is.null(attr(footprint_submission, "status")) || length(footprint_submission) != 1L ||
    !grepl("^[0-9]+(;.*)?$", footprint_submission))
  stop("Slurm enhancer footprint job submission failed")
footprint_job_id <- sub(";.*$", "", footprint_submission)
writeLines(footprint_job_id, footprint_job_id_file)
message("Submitted footprint job ", footprint_job_id, ". Wait for completion before loading.")
```

```{r count-footprints}
if (!file.exists(footprint_results_file))
  stop("Footprint results are missing. Run submit-enhancer-footprints and wait for completion.")
footprint_result <- readRDS(footprint_results_file)
current_footprint_inputs <- footprint_job_inputs(
  analysis_enhancers, sample_table, standard_chromosomes,
  extract_root, ft_result_dir, expected_background)
if (!identical(footprint_result$inputs, current_footprint_inputs))
  stop("Saved footprint results do not match current enhancers/settings. Submit a new footprint job.")
if (file.exists(footprint_job_id_file) &&
    !identical(footprint_result$job_id, readLines(footprint_job_id_file, n = 1L)))
  stop("The latest footprint job has not published its results. Check its Slurm status.")
footprint_counts <- footprint_result$footprint_counts
expected_length_counts <- footprint_result$expected_length_counts
head(footprint_counts)
```


```{r tf-nucleosome-enrichment-summary}
footprint_comparison <- footprint_counts[, .(n_tf = sum(n_tf), n_nuc = sum(n_nuc),
                                            n_reads = sum(n_reads)),
                                         by = .(sample, time_min, enhancer_class)]
# Bins <60 are TF; bins >=90 contain only calls strictly >90 from the loader.
expected_categories <- expected_length_counts[, .(
  expected_tf = sum(n[length_bin < 60L]), expected_nuc = sum(n[length_bin >= 90L])), by = sample]
footprint_comparison <- merge(footprint_comparison, expected_categories, by = "sample")
footprint_comparison[, tf_fraction := fifelse(n_tf + n_nuc > 0, n_tf / (n_tf + n_nuc), NA_real_)]
footprint_comparison[, tf_to_nuc := fifelse(n_nuc > 0, n_tf / n_nuc, NA_real_)]
footprint_comparison[, expected_tf_to_nuc := fifelse(expected_nuc > 0,
                                                    expected_tf / expected_nuc, NA_real_)]
footprint_comparison[, enrichment := fifelse(expected_tf_to_nuc > 0,
                                             tf_to_nuc / expected_tf_to_nuc, NA_real_)]
save_table(footprint_comparison[, .(sample, enhancer_class, n_tf, n_nuc,
                                   tf_fraction, tf_to_nuc, expected_tf_to_nuc, enrichment)],
           "02_TF_to_nucleosome_enrichment.tsv")
p_enrichment <- ggplot(footprint_comparison,
  aes(time_min, enrichment, color = enhancer_class, group = enhancer_class)) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey60") +
  geom_line(na.rm = TRUE) + geom_point(na.rm = TRUE) + theme_classic() +
  scale_x_continuous(breaks = sample_table$time_min) +
  labs(x = "LPS (min)", y = "TF:nucleosome ratio / expected ratio", color = NULL)
save_pdf(p_enrichment, "02_TF_to_nucleosome_enrichment.pdf")
```

# 3. TF co-occupancy enrichment at active enhancers

For each timepoint, the Fisher table contains active/inactive rows and
co-occupied/not-co-occupied columns. A co-occupied observation is a full-span
fiber with at least two short TF footprints assigned to the enhancer. Enhancers
must have at least `minimum_reads` full-span fibers **at every timepoint** to
enter this comparison, keeping the same enhancer set across the time course.
All enhancer/timepoint counts, including lower coverage and zero calls, remain
in the count table. Missing coverage is not evidence of no co-occupancy.

The test is one-sided (`alternative = "greater"`) for enrichment at active
enhancers, uses integer counts without pseudocounts, and applies BH correction
across the four timepoint tests. The odds ratio is Fisher's conditional estimate.
The method follows the count-table approach in `code/fire_frequency`, implemented
here in R. Observations are enhancer–fiber pairs; a fiber can cover multiple
enhancers. P-values are therefore nominal molecule-level comparisons, not
replicate-level inference. Enhancer length and locus composition can also affect
these pooled comparisons; per-enhancer frequencies are plotted alongside them.

```{r fisher-cooccupancy-function}
fisher_cooccupancy <- function(x) {
  active <- x[enhancer_class == "active"]
  inactive <- x[enhancer_class == "inactive"]
  a <- sum(active$n_cooccupied)
  b <- sum(active$n_reads - active$n_cooccupied)
  c <- sum(inactive$n_cooccupied)
  d <- sum(inactive$n_reads - inactive$n_cooccupied)
  table <- matrix(c(a, b, c, d), nrow = 2L, byrow = TRUE,
                   dimnames = list(c("active", "inactive"), c("cooccupied", "not_cooccupied")))
  result <- data.table(active_co = a, active_other = b, inactive_co = c, inactive_other = d,
                       odds_ratio = NA_real_, p_value = NA_real_, status = "not_testable")
  if (all(rowSums(table) > 0) && all(colSums(table) > 0)) {
    fit <- fisher.test(table, alternative = "greater")
    result[, `:=`(odds_ratio = unname(fit$estimate), p_value = fit$p.value, status = "tested")]
  }
  result
}
```

```{r cooccupancy-fisher-tests}
eligible_enhancers <- footprint_counts[, .(
  keep = .N == nrow(sample_table) && all(n_reads >= minimum_reads)), by = enhancer_id][keep == TRUE, enhancer_id]
coocc_counts <- copy(footprint_counts[enhancer_id %in% eligible_enhancers])
coocc_counts[, cooccupied_fraction := n_cooccupied / n_reads]
head(coocc_counts)
coocc_tests <- rbindlist(lapply(sample_table$sample, function(s) {
  result <- fisher_cooccupancy(coocc_counts[sample == s])
  result[, sample := s]
  result
}))
coocc_tests[, q_value := p.adjust(p_value, method = "BH")]
save_table(coocc_tests[, .(sample, active_co, active_other, inactive_co, inactive_other,
                          odds_ratio, p_value, q_value, status)], "03_cooccupancy_fisher.tsv")
coocc_frequency <- coocc_counts[, .(n_enhancers = .N, n_reads = sum(n_reads),
                                    n_cooccupied = sum(n_cooccupied)),
                                 by = .(sample, time_min, enhancer_class)]
coocc_frequency[, frequency := n_cooccupied / n_reads]
save_table(coocc_frequency, "03_cooccupancy_frequency.tsv")
```

```{r cooccupancy-plots}
if (nrow(coocc_frequency)) {
  p_coocc <- ggplot(coocc_frequency, aes(time_min, frequency,
                                        color = enhancer_class, group = enhancer_class)) +
    geom_line() + geom_point() + scale_x_continuous(breaks = sample_table$time_min) +
    scale_y_continuous(limits = c(0, 1)) + theme_classic() +
    labs(x = "LPS (min)", y = "Fraction of fibers with >=2 TF footprints", color = NULL)
  save_pdf(p_coocc, "03_cooccupancy_timecourse.pdf")
  p_locus <- ggplot(coocc_counts,
                    aes(factor(time_min, levels = sample_table$time_min), cooccupied_fraction,
                        fill = enhancer_class)) +
    geom_boxplot(outlier.shape = NA) + theme_classic() +
    labs(x = "LPS (min)", y = "Per-enhancer co-occupied fiber fraction", fill = NULL)
  save_pdf(p_locus, "03_cooccupancy_per_enhancer.pdf")
}
```

### Example fibers at an active enhancer

The example is the active enhancer with the largest total number of co-occupied
fibers across time points. BED12 m6A blocks are read from each displayed
fiber's selected alignment, with the extraction sentinel blocks removed. Gray
lines are spanning fibers, blue marks are m6A, magenta segments are complete
FiberHMM TF calls shorter than 60 bp, and dark-blue segments are nucleosome
footprints longer than 90 bp. A top panel summarizes m6A calls on the displayed
fibers (per-base call fraction among covering fibers, averaged over +/-5 bp);
it is not an adenine-normalized methylation estimate. A genomic track marks the
FANTOM5 enhancer and its boundaries. All panels share the enhancer +/-100 bp
window. Fibers are ordered by their number of TF footprints with midpoints inside
the enhancer and capped at 100 per time point for display.
This selected example is descriptive; all eligible fibers enter the tests above.

```{r function-load-enhancer-example}
# Parse one chosen BED12 alignment per displayed fiber. Grouping several
# alignments by read ID before parsing can recycle incompatible block vectors.
enhancer_example_m6a_blocks <- function(raw, reads) {
  alignments <- data.table(read_id = as.character(raw[[4L]]),
    start0 = as.numeric(raw[[2L]]), end0 = as.numeric(raw[[3L]]),
    record_index = seq_len(nrow(raw)))
  chosen <- merge(alignments, reads[, .(read_id, start0, end0)],
    by = c("read_id", "start0", "end0"), sort = FALSE)
  setorder(chosen, record_index)
  chosen <- unique(chosen, by = "read_id")
  stopifnot(nrow(chosen) == nrow(reads))
  empty <- data.table(read_id = character(), start0 = numeric(), end0 = numeric())
  blocks <- lapply(chosen$record_index, function(index) {
    sizes <- as.numeric(strsplit(as.character(raw[[11L]][index]), ",", fixed = TRUE)[[1L]])
    offsets <- as.numeric(strsplit(as.character(raw[[12L]][index]), ",", fixed = TRUE)[[1L]])
    count <- as.integer(raw[[10L]][index])
    stopifnot(length(sizes) == count, length(offsets) == count,
              !anyNA(sizes), !anyNA(offsets))
    keep <- setdiff(seq_len(count), c(1L, count))
    keep <- keep[sizes[keep] > 0]
    if (!length(keep)) return(empty)
    starts <- as.numeric(raw[[2L]][index]) + offsets[keep]
    data.table(read_id = as.character(raw[[4L]][index]),
               start0 = starts, end0 = starts + sizes[keep])
  })
  unique(rbindlist(c(list(empty), blocks)))
}

load_enhancer_example <- function(enhancer, samples, max_reads = 100L) {
  chromosome <- enhancer$chr[[1]]
  lo <- enhancer$start0[[1]]
  hi <- enhancer$end0[[1]]
  window <- c(max(0L, lo - 100L), hi + 100L)
  query <- GenomicRanges::GRanges(chromosome,
    IRanges::IRanges(window[1] + 1L, window[2]))
  spans <- m6a <- footprints <- nucleosomes <- read_counts <- vector("list", nrow(samples))
  for (i in seq_len(nrow(samples))) {
    sample_name <- samples$sample[i]
    short_name <- gsub("_", "", sample_name, fixed = TRUE)
    methylation_path <- file.path(ft_result_dir, sample_name, "extracted_results",
      "m6a_by_chr", sprintf("%s.ft_extracted_m6a.%s.bed.gz", sample_name, chromosome))
    raw <- tabix_bed(methylation_path, query)
    if (!nrow(raw)) next
    reads <- unique(raw[, .(read_id = as.character(V4),
      start0 = as.numeric(V2), end0 = as.numeric(V3))])
    reads[, span_width := end0 - start0]
    data.table::setorder(reads, read_id, -span_width, start0, end0)
    reads <- reads[!duplicated(read_id)][start0 <= lo & end0 >= hi]
    if (!nrow(reads)) next
    tf_path <- file.path(extract_root, "firehmm_tf", short_name,
      sprintf("%s_hmm_extracted_tf_%s.bed.gz", short_name, chromosome))
    tf <- expand_hmm_blocks(tabix_bed(tf_path, query))
    tf <- unique(tf[length < 60L & length > 0L & read_id %in% reads$read_id &
      start0 < window[2] & end0 > window[1]])
    counts <- tf[(start0 + end0) / 2 >= lo & (start0 + end0) / 2 < hi,
      .(n_tf = .N), by = read_id]
    reads <- merge(reads, counts, by = "read_id", all.x = TRUE)
    reads[is.na(n_tf), n_tf := 0L]
    data.table::setorder(reads, -n_tf, read_id)
    read_counts[[i]] <- data.table(sample = sample_name, total = nrow(reads),
                                   displayed = min(nrow(reads), max_reads))
    reads <- head(reads, max_reads)
    reads[, `:=`(sample = sample_name, row = rev(seq_len(.N)))]
    reads[, key := paste(sample, read_id)]
    m <- enhancer_example_m6a_blocks(raw, reads)
    m <- m[start0 < window[2] & end0 > window[1]]
    m <- merge(m, reads[, .(read_id, sample, row)], by = "read_id")
    tf <- merge(tf, reads[, .(read_id, sample, row)], by = "read_id")
    nuc_path <- file.path(extract_root, "firehmm_footprint", short_name,
      sprintf("%s_hmm_extracted_footprint_%s.bed.gz", short_name, chromosome))
    nuc <- expand_hmm_blocks(tabix_bed(nuc_path, query))
    nuc <- unique(nuc[length > 90L & read_id %in% reads$read_id &
      start0 < window[2] & end0 > window[1]],
      by = c("chr", "read_id", "start0", "end0"))
    nuc <- merge(nuc, reads[, .(read_id, sample, row)], by = "read_id")
    nucleosomes[[i]] <- nuc
    spans[[i]] <- reads
    m6a[[i]] <- m
    footprints[[i]] <- tf
  }
  sp <- data.table::rbindlist(spans, fill = TRUE)
  if (!nrow(sp)) stop("No spanning fibers found for the example enhancer")
  list(spans = sp, m6a = data.table::rbindlist(m6a, fill = TRUE),
    tf = data.table::rbindlist(footprints, fill = TRUE),
    nuc = data.table::rbindlist(nucleosomes, fill = TRUE),
    read_counts = data.table::rbindlist(read_counts, fill = TRUE),
    max_reads = max_reads, samples = samples, window = window, enhancer = enhancer)
}
```

```{r function-plot-enhancer-example}
# Same displayed molecules in the aggregate and single-fiber panels. Using
# coverage at each base avoids treating uncovered flanks as unmethylated.
enhancer_example_pileup <- function(example) {
  positions <- seq.int(example$window[1], example$window[2] - 1L)
  rbindlist(lapply(example$samples$sample, function(sample_name) {
    reads <- example$spans[sample == sample_name]
    calls <- example$m6a
    if (nrow(calls)) calls <- calls[sample == sample_name]
    coverage <- vapply(positions, function(pos)
      sum(reads$start0 <= pos & reads$end0 > pos), integer(1))
    modified <- vapply(positions, function(pos) {
      if (!nrow(calls)) return(0L)
      covering_ids <- reads$read_id[reads$start0 <= pos & reads$end0 > pos]
      uniqueN(calls$read_id[calls$start0 <= pos & calls$end0 > pos &
                             calls$read_id %in% covering_ids])
    }, integer(1))
    stopifnot(all(modified <= coverage))
    fraction <- ifelse(coverage > 0L, modified / coverage, NA_real_)
    smoothed <- vapply(seq_along(positions), function(i) {
      if (!is.finite(fraction[i])) return(NA_real_)
      mean(fraction[max(1L, i - 5L):min(length(positions), i + 5L)], na.rm = TRUE)
    }, numeric(1))
    data.table(sample = sample_name, position = positions, coverage = coverage,
               modified = modified, fraction = fraction, smooth_fraction = smoothed)
  }))
}

plot_enhancer_example <- function(example) {
  sp <- data.table::copy(example$spans)
  m <- data.table::copy(example$m6a)
  tf <- data.table::copy(example$tf)
  nuc <- data.table::copy(example$nuc)
  samples <- example$samples$sample
  sp[, sample := factor(sample, levels = samples)]
  if (nrow(m)) m[, sample := factor(sample, levels = samples)]
  if (nrow(tf)) tf[, sample := factor(sample, levels = samples)]
  if (nrow(nuc)) nuc[, sample := factor(sample, levels = samples)]
  count_index <- match(samples, example$read_counts$sample)
  displayed <- example$read_counts$displayed[count_index]
  total <- example$read_counts$total[count_index]
  displayed[is.na(displayed)] <- 0L
  total[is.na(total)] <- 0L
  labels <- setNames(sprintf("%s min\n%d / %d fibers",
    example$samples$time_min, displayed, total), samples)
  time_colors <- setNames(rep(c("#0072B2", "#009E73", "#E69F00", "#D55E00"),
                              length.out = length(samples)), samples)
  xscale <- function() ggplot2::scale_x_continuous(
    breaks = pretty(example$window, n = 5),
    labels = function(x) format(x, scientific = FALSE, big.mark = ",", trim = TRUE),
    expand = ggplot2::expansion(mult = 0))
  shared_theme <- ggplot2::theme_bw(base_size = 10) +
    ggplot2::theme(panel.grid = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(fill = "grey95"),
      strip.text.y = ggplot2::element_text(angle = 0),
      legend.position = "bottom")
  highlight <- function() ggplot2::annotate("rect",
    xmin = example$enhancer$start0, xmax = example$enhancer$end0,
    ymin = -Inf, ymax = Inf, fill = "#44d6d8", alpha = 0.2)
  genomic_axis <- paste0(example$enhancer$chr, " position (bp; hg38)")
  pileup <- enhancer_example_pileup(example)
  pileup[, sample := factor(sample, levels = samples)]
  p_summary <- ggplot2::ggplot(pileup, ggplot2::aes(position, smooth_fraction)) +
    highlight() + ggplot2::geom_area(ggplot2::aes(fill = sample), alpha = 0.8,
                                    na.rm = TRUE) +
    ggplot2::geom_line(linewidth = 0.25, na.rm = TRUE) +
    ggplot2::facet_grid(sample ~ ., drop = FALSE,
      labeller = ggplot2::labeller(sample = labels)) +
    ggplot2::scale_fill_manual(values = time_colors, guide = "none") +
    ggplot2::scale_y_continuous(limits = c(0, NA)) + xscale() +
    ggplot2::coord_cartesian(xlim = example$window, expand = FALSE) + shared_theme +
    ggplot2::labs(title = "Aggregate m6A on displayed fibers",
      subtitle = "Per-base call fraction among covering fibers; +/-5 bp mean",
      x = NULL, y = "m6A call fraction")
  p_region <- ggplot2::ggplot() +
    ggplot2::annotate("segment", x = example$window[1], xend = example$window[2],
      y = 1, yend = 1, colour = "grey65", linewidth = 0.5) +
    ggplot2::annotate("rect", xmin = example$enhancer$start0,
      xmax = example$enhancer$end0, ymin = 0.75, ymax = 1.25,
      fill = "#44d6d8", colour = "#16888A") +
    ggplot2::annotate("text", x = mean(c(example$enhancer$start0, example$enhancer$end0)),
      y = 1.55, label = example$enhancer$enhancer_id, size = 3.2) +
    ggplot2::scale_y_continuous(breaks = 1, labels = "FANTOM5 enhancer",
                               limits = c(0.5, 1.9)) + xscale() +
    ggplot2::coord_cartesian(xlim = example$window, expand = FALSE) + shared_theme +
    ggplot2::labs(x = genomic_axis, y = NULL) +
    ggplot2::theme(axis.ticks.y = ggplot2::element_blank())
  base <- ggplot2::ggplot(sp) + highlight() +
    ggplot2::geom_segment(ggplot2::aes(x = start0, xend = end0,
      y = row, yend = row), colour = "grey70", linewidth = 0.10) +
    ggplot2::facet_grid(sample ~ ., scales = "free_y", drop = FALSE,
      labeller = ggplot2::labeller(sample = labels)) + xscale() +
    ggplot2::coord_cartesian(xlim = example$window, expand = FALSE) + shared_theme +
    ggplot2::theme(axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank()) +
    ggplot2::labs(x = genomic_axis, y = "Spanning fibers")
  p_m6a <- base + ggplot2::labs(title = "Single-fiber m6A")
  if (nrow(m)) p_m6a <- p_m6a + ggplot2::geom_segment(data = m,
    ggplot2::aes(x = start0, xend = end0, y = row, yend = row),
    colour = "blue", linewidth = 0.22)
  p_tf <- base + ggplot2::labs(title = "Single-fiber TF and nucleosome footprints")
  # Full footprint lengths determine eligibility; clipping happens only at display.
  # Draw TF calls last so short footprints remain visible above nucleosomes.
  if (nrow(nuc)) p_tf <- p_tf + ggplot2::geom_segment(data = nuc,
    ggplot2::aes(x = start0, xend = end0, y = row, yend = row,
                 colour = "Nucleosome (>90 bp)"), linewidth = 0.18)
  if (nrow(tf)) p_tf <- p_tf + ggplot2::geom_segment(data = tf,
    ggplot2::aes(x = start0, xend = end0, y = row, yend = row,
                 colour = "TF (<60 bp)"), linewidth = 0.24)
  p_tf <- p_tf + ggplot2::scale_colour_manual(name = NULL,
    values = c("Nucleosome (>90 bp)" = "darkblue", "TF (<60 bp)" = "magenta"),
    limits = c("Nucleosome (>90 bp)", "TF (<60 bp)"), drop = FALSE) +
    ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(linewidth = 1)))
  patchwork::wrap_plots(p_summary, p_region, p_m6a, p_tf, ncol = 1,
                        heights = c(3, 0.8, 5, 5)) +
    patchwork::plot_annotation(title = example$enhancer$enhancer_id,
      subtitle = sprintf(paste0("Enhancer shaded; up to %d fibers/timepoint, ordered by TF count; ",
                                "labels show displayed / total spanning fibers"), example$max_reads))
}
```

```{r plot-example-enhancer, fig.width=12, fig.height=16}
example_rank <- footprint_counts[enhancer_class == "active",
  .(n_cooccupied = sum(n_cooccupied)), by = enhancer_id]
data.table::setorder(example_rank, -n_cooccupied, enhancer_id)
if (nrow(example_rank) && example_rank$n_cooccupied[1] > 0L) {
  example_enhancer <- analysis_enhancers[enhancer_id == example_rank$enhancer_id[1]]
  print(example_enhancer[, .(enhancer_id, chr, start0, end0)])
  example_fibers <- load_enhancer_example(example_enhancer, sample_table)
  example_plot <- plot_enhancer_example(example_fibers)
  print(example_plot)
  save_pdf(example_plot, "active_enhancer_example_fibers.pdf", width = 12, height = 16)
}
```


# 5. Co-occupied site pairs in active versus inactive enhancers

Both enhancer classes use the same site discovery and full-enhancer-span reads.
Enhancers must pass Analysis 3's coverage filter at every timepoint. Sites are
50-bp groups of pooled short-footprint midpoints with at least five observations;
a fiber binds a site when a TF footprint (<60 bp) midpoint is within 25 bp of
its fixed center. Site windows may overlap, allowing one footprint to support
both sites. Discovery and testing reuse the same fibers: these are exploratory
associations, not evidence of a molecular interaction or biological replication.

Each pair has four states: both bound, site 1 only, site 2 only, neither bound.
"Neither" means no qualifying TF call at these sites, not absence of nucleosomes.
Normalized co-binding is observed both-bound fraction divided by the product of
the two marginal bound fractions. Two-sided Fisher tests detect enrichment and
depletion; BH correction pools both enhancer classes within each timepoint.
The three classes are enriched, depleted, and no significant association.
Untestable pairs remain labeled untestable and do not enter these three classes.

For the active/inactive comparison, the unit is one enhancer per timepoint,
including enhancers with fewer than two detected sites (zero pairs). The primary
outcome is the expected number of distinct co-occupied pairs detectable in
`min_pair_reads` fibers, using exact sampling without replacement:
`sum(1 - choose(N - n_both, depth) / choose(N, depth))`. This standardizes read
coverage without drawing a random subset. Raw distinct-pair counts, pair events
per fiber, and the fraction of possible pairs observed are also reported.
Site discovery remains depth-dependent; an adjusted sensitivity analysis includes
read coverage, enhancer length, and possible-pair count. These adjustments answer
a conditional question and do not establish a causal effect of enhancer activity.

One-sided Wilcoxon tests ask whether active values tend to be higher. A second
analysis fits enhancer-level `log1p(outcome)` with the covariates above, using HC3
robust standard errors and a one-sided active coefficient test. Tests are run
separately per timepoint; BH adjustment is across timepoints within each metric
and test method. Loci, not biological replicates, provide the observations.

```{r site-pair-settings}
site_window_bp <- 50L
min_site_observations <- 5L
min_pair_reads <- 10L
cobinding_alpha <- 0.01
cobinding_significance <- "q_value"  # BH-adjusted threshold; "p_value" for raw p
```

```{r site-pair-input-functions}
source(file.path(project_root, "code/enhancer/co_occ_functions/enhancer_site_pair_functions.R"),
       local = TRUE)
```

```{r site-pair-statistic-functions}
# Pair counting, two-sided tests, summaries and class comparisons are defined
# together in enhancer_site_pair_functions.R above.
```

The complete analysis can also run as a batch job. After it finishes, use the
optional loading chunk instead of recalculating pairs interactively.

```{r submit-site-pair-comparison, eval=FALSE}
site_pair_script <- file.path(project_root,"code/enhancer/shell_scripts/run_enhancer_site_pairs.sh")
site_pair_job <- system2("sbatch",c("--parsable",shQuote(site_pair_script)),stdout=TRUE)
if (!is.null(attr(site_pair_job,"status")) || length(site_pair_job)!=1L ||
    !grepl("^[0-9]+(;.*)?$",site_pair_job)) stop("Site-pair job submission failed")
writeLines(sub(";.*$","",site_pair_job),file.path(table_dir,"site_pair_job_id.txt"))
```

```{r load-site-pair-comparison, eval=FALSE}
site_pair_result <- readRDS(file.path(table_dir,"enhancer_site_pair_results.rds"))
stopifnot(identical(site_pair_result$inputs,footprint_result$inputs),
  identical(site_pair_result$parameters,list(site_window_bp=site_window_bp,
    min_site_observations=min_site_observations,min_pair_reads=min_pair_reads,
    cobinding_alpha=cobinding_alpha,cobinding_significance=cobinding_significance)))
site_pair_cooccupancy <- site_pair_result$pairs
site_pair_enhancer_summary <- site_pair_result$enhancers
site_pair_comparisons <- site_pair_result$comparisons
```

```{r calculate-site-pair-cooccupancy}
site_pair_result <- calculate_enhancer_site_pairs(
  analysis_enhancers, sample_table, footprint_counts, footprint_result,
  eligible_enhancers, site_window_bp, min_site_observations, min_pair_reads,
  cobinding_alpha, cobinding_significance)
site_pair_cooccupancy <- site_pair_result$pairs
site_pair_enhancer_summary <- site_pair_result$enhancers
save_table(site_pair_cooccupancy, "05_site_pair_cooccupancy.tsv")
save_table(site_pair_enhancer_summary, "05_cobinding_per_enhancer.tsv")
save_table(site_pair_result$sites, "05_cobinding_site_inventory.tsv")
```

```{r site-pair-active-inactive-tests}
site_pair_comparisons <- compare_enhancer_pair_counts(site_pair_enhancer_summary)
save_table(site_pair_comparisons, "05_active_inactive_cobinding_tests.tsv")
```

### Co-binding panels

A: hexagon counts of site pairs, normalized co-binding against -log10(two-sided
Fisher p). The red dotted line is raw p=0.01; class assignments use the configured
threshold, BH q<0.01 by default. A log1p x-axis retains zero ratios and the full tail.
B: coverage-standardized distinct pairs per enhancer, active versus inactive;
raw distinct-pair counts and the total counts in the three association classes
are saved as additional panels. C: site-center distances by co-binding class.
D: four state-occupancy fractions by co-binding class. C/D are descriptive and
use site pairs as observations; significance of the enhancer-class comparison
comes from the enhancer-level tests, not the number of plotted pairs.

```{r site-pair-cooccupancy-plots}
source(file.path(project_root, "code/enhancer/co_occ_functions/enhancer_site_pair_plots.R"),
       local = TRUE)
save_enhancer_cobinding_plots(site_pair_result, site_pair_comparisons, plot_dir)
```
