---
title: "TF co-occupancy at enhancers"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: pdf_document
---

# TF occupancy at enhancers over time

Analyses 1–3 use hg38 FANTOM5 enhancer intervals and LPS 0, 5, 10 and 15 min.
All interval tables use **0-based, half-open BED coordinates**. An enhancer is
accessible if **more than 50% of its length overlaps one individual MSP block**
at any timepoint. The denominator is the enhancer length; exactly 50% does not
qualify. Adjacent MSPs or MSPs on different fibers are never joined to meet
this criterion. FANTOM5 outer coordinates define each enhancer (not its BED12
CAGE blocks).

Defaults: compare active and inactive enhancers within the accessible universe;
active means overlap with both baseline H3K27ac and ATAC peaks (at least one
replicate per assay). Inactive means lacking either assay. Single-positive
regions are therefore included in inactive. Baseline classes remain fixed across
the time course. Settings below allow a stricter inactive definition or all
FANTOM5 enhancers. Peak overlap defines activity; mean baseline bigWig signal is
reported separately. Signal differences between these classes are descriptive
because the same assays define the classes.

“TF” here means a short protected footprint, not an identified TF protein.
Co-occupancy means at least two distinct <60-bp footprints on the same fiber
within an enhancer. Motif-based TF identities are not inferred from footprint
size alone. Analyses 4–5 are not included.

```{r setup, include=FALSE}
project_root <- "/project/spott/cshan/fiber-seq"
output_dir <- file.path(project_root, "macrophage_project/enhancer/TF_co-occ")
plot_dir <- file.path(output_dir, "plots")
table_dir <- file.path(output_dir, "tables")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      cache = FALSE, dev = "pdf", fig.path = paste0(plot_dir, "/"),
                      fig.width = 8, fig.height = 5)
options(scipen = 10)
```

```{r packages}
suppressPackageStartupMessages({
  suppressMessages({
    library(data.table)
    library(GenomicRanges)
    library(Rsamtools)
    library(ggplot2)
    library(dplyr)
    library(tidyr)
    library(rtracklayer)
    library(Matrix)
    source(file.path(project_root, "code/topic_model/topic_modelling_functions.r"))
  })
})
```

```{r settings}
extract_root <- file.path(project_root, "macrophage_project/FiberHMM/extract")
ft_result_dir <- file.path(extract_root, "ft_result_dir")
annotation_dir <- file.path(project_root, "macrophage_project/annotations/enhancer_public_data")
fantom_file <- file.path(annotation_dir, "fantom5_enhancer_hg38/F5.hg38.enhancers.bed.gz")
sample_table <- data.table(sample = paste0("LPS_", c(0, 5, 10, 15)),
                           time_min = c(0L, 5L, 10L, 15L))
standard_chromosomes <- paste0("chr", c(1:22, "X", "Y"))
comparison_universe <- "accessible" # MSP-positive FANTOM5 subset; or "all_fantom"
inactive_definition <- "not_active" # fails active criteria; or "neither_signal"
minimum_peak_replicates <- 1L        # Active requires H3K27ac and ATAC peak overlap in at least one replicate per assay
minimum_reads <- 10L                # full-enhancer-span reads at EVERY timepoint, analysis 3
expected_background <- "genomewide" # expected footprint-length source; or "inactive"
set.seed(1)
```

`comparison_universe = "accessible"` restricts the comparison to FANTOM5 enhancers
passing the MSP rule at one or more timepoints. `"all_fantom"` includes all
FANTOM5 enhancers on the selected chromosomes, including those failing the MSP
rule. Accessibility and baseline activity are separate annotations.

`inactive_definition = "not_active"` puts enhancers failing the active criteria
into the inactive group. `"neither_signal"` requires zero baseline H3K27ac peak
support and zero baseline ATAC peak support; partially supported enhancers are
excluded from the comparison. Active always requires both assays to meet
`minimum_peak_replicates`: 1 allows either replicate per assay, while 2 requires
both replicates per assay. Peak support means any interval overlap, and the
classes remain fixed from the 0-min baseline.

`minimum_reads = 10L` requires at least 10 full-enhancer-span fibers at **each of
all four timepoints** for the co-occupancy frequencies and Fisher comparisons in
analysis 3. This keeps a consistent enhancer set across time. It does not filter
the MSP universe, ACF clustering or the footprint-length summaries.

`expected_background = "genomewide"` uses the same timepoint's genome-wide TF
(<60 bp) and nucleosome (>90 bp) calls for expected footprint-length frequencies
and the expected TF:nucleosome ratio. `"inactive"` uses those calls at the selected
inactive enhancers instead. This setting does not change the active-versus-inactive
Fisher comparison.

# Shared input and output functions


```{r save-table-function}
save_table <- function(x, filename) {
  path <- file.path(table_dir, filename)
  data.table::fwrite(x, path, sep = "\t", na = "NA")
  reloaded <- data.table::fread(path)
  print(utils::head(reloaded, 6L))
  invisible(reloaded)
}
```

```{r save-pdf-function}
save_pdf <- function(plot, filename, width = 8, height = 5) {
  stopifnot(grepl("\\.pdf$", filename))
  ggplot2::ggsave(file.path(plot_dir, filename), plot = plot,
                 device = "pdf", width = width, height = height, limitsize = FALSE)
  invisible(plot)
}
```

```{r bed-ranges-function}
bed_ranges <- function(x) {
  stopifnot(all(x$start0 >= 0), all(x$end0 > x$start0))
  GenomicRanges::GRanges(x$chr, IRanges::IRanges(x$start0 + 1L, x$end0))
}
```

```{r tabix-bed-function}
tabix_bed <- function(path, regions) {
  if (!file.exists(path) || !file.exists(paste0(path, ".tbi")))
    stop("Missing BED or tabix index: ", path)
  if (!length(regions)) return(data.table())
  tf <- Rsamtools::TabixFile(path)
  open(tf)
  on.exit(close(tf))
  # A long fiber may overlap several query regions: remove duplicate lines.
  lines <- unique(unlist(Rsamtools::scanTabix(tf, param = reduce(regions)),
                         use.names = FALSE))
  if (!length(lines)) return(data.table())
  data.table::fread(text = paste(lines, collapse = "\n"), header = FALSE,
                    sep = "\t", showProgress = FALSE)
}
```

```{r expand-hmm-blocks-function}
expand_hmm_blocks <- function(raw) {
  empty <- data.table(chr = character(), start0 = integer(), end0 = integer(),
                      read_id = character(), strand = character(), length = integer())
  if (!nrow(raw)) return(empty)
  if (ncol(raw) < 12L) stop("Expected FiberHMM BED12 or BED12+ input")
  sizes <- strsplit(sub(",$", "", as.character(raw[[11L]])), ",", fixed = TRUE)
  offsets <- strsplit(sub(",$", "", as.character(raw[[12L]])), ",", fixed = TRUE)
  counts <- as.integer(raw[[10L]])
  if (anyNA(counts) || any(lengths(sizes) != counts) || any(lengths(offsets) != counts))
    stop("Invalid FiberHMM block counts")
  parent <- rep(seq_len(nrow(raw)), counts)
  size <- as.integer(unlist(sizes, use.names = FALSE))
  offset <- as.integer(unlist(offsets, use.names = FALSE))
  start0 <- as.integer(raw[[2L]][parent]) + offset
  if (anyNA(size) || anyNA(offset) || any(size <= 0L) || any(offset < 0L) ||
      any(start0 + size > as.integer(raw[[3L]][parent]))) stop("Invalid FiberHMM blocks")
  unique(data.table(chr = as.character(raw[[1L]][parent]), start0 = start0,
                    end0 = start0 + size, read_id = as.character(raw[[4L]][parent]),
                    strand = as.character(raw[[6L]][parent]), length = size))
}
```

```{r hmm-file-function}
hmm_file <- function(sample, chr, kind) {
  sample_dir <- gsub("_", "", sample)
  file.path(extract_root, paste0("firehmm_", kind), sample_dir,
            paste0(sample_dir, "_hmm_extracted_", kind, "_", chr, ".bed.gz"))
}
```

# 1. Define the accessible enhancer universe

```{r load-fantom}
enhancers <- fread(fantom_file, header = FALSE, select = 1:4,
                   col.names = c("chr", "start0", "end0", "enhancer_id"))
enhancers <- unique(enhancers[chr %in% standard_chromosomes], by = c("chr", "start0", "end0"))
stopifnot(!anyDuplicated(enhancers$enhancer_id))
enhancers[, midpoint0 := as.integer(floor((start0 + end0) / 2))]
head(enhancers)
```

Submit the four-timepoint accessibility job, then load its tables after it
finishes. Input paths, samples, and chromosomes are configured in the script.

```{r submit-accessible-universe, eval=FALSE}
accessibility_script <- file.path(project_root, "code/enhancer/find_accessible_enhancers.sh")
accessibility_status <- system2("sbatch", shQuote(accessibility_script))
if (accessibility_status != 0L) stop("Slurm accessibility job submission failed")
```

`sbatch /project/spott/cshan/fiber-seq/code/enhancer/find_accessible_enhancers.sh`.

```{r load-accessible-universe}
accessibility_paths <- file.path(table_dir,
  c("01_enhancer_accessibility.tsv", "01_accessible_counts.tsv"))
if (!all(file.exists(accessibility_paths)))
  stop("Accessibility tables are missing. Submit the Slurm job and wait for it to finish.")
accessibility_export <- fread(file.path(table_dir, "01_enhancer_accessibility.tsv"))
stopifnot(!anyDuplicated(accessibility_export$enhancer_id),
          all(enhancers$enhancer_id %in% accessibility_export$enhancer_id))
head(accessibility_export)
accessibility <- accessibility_export[match(enhancers$enhancer_id, enhancer_id),
  c("enhancer_id", sample_table$sample, "accessible"), with = FALSE]
enhancers[, accessible := accessibility$accessible]
accessibility_summary <- fread(file.path(table_dir, "01_accessible_counts.tsv"))
head(accessibility_summary)
```

# 2. Chromatin signatures at baseline active and inactive enhancers

```{r baseline-files-function}
baseline_files <- function(directory, suffix) {
  paths <- list.files(directory, pattern = paste0("_0000_.*", suffix, "$"), full.names = TRUE)
  if (length(paths) != 2L) stop("Expected two baseline replicates: ", directory, " / ", suffix)
  paths
}
```

```{r baseline-peak-support-function}
baseline_peak_support <- function(enhancers, files) {
  support <- integer(nrow(enhancers))
  for (path in files) {
    peaks <- fread(path, header = FALSE, select = 1:3,
                   col.names = c("chr", "start0", "end0"))
    support <- support + (countOverlaps(bed_ranges(enhancers), bed_ranges(peaks)) > 0L)
  }
  support
}
```

```{r baseline-signal-function}
baseline_signal <- function(enhancers, files) {
  signal <- matrix(NA_real_, nrow(enhancers), length(files))
  for (j in seq_along(files)) {
    bw <- rtracklayer::BigWigFile(files[j])
    available <- names(GenomeInfoDb::seqlengths(bw))
    for (chrom in intersect(unique(enhancers$chr), available)) {
      idx <- which(enhancers$chr == chrom)
      query <- bed_ranges(enhancers[idx])
      track <- rtracklayer::import(bw, which = reduce(query), as = "GRanges")
      # Uncovered bases on a represented chromosome contribute zero signal;
      # an absent chromosome is NA, not zero.
      hits <- findOverlaps(query, track)
      signal[idx, j] <- 0
      if (length(hits)) {
        q <- queryHits(hits)
        t <- subjectHits(hits)
        bp <- width(pintersect(query[q], track[t]))
        sums <- rowsum(bp * mcols(track)$score[t], q, reorder = FALSE)
        k <- as.integer(rownames(sums))
        signal[idx[k], j] <- as.numeric(sums) / width(query)[k]
      }
    }
  }
  values <- rowMeans(signal, na.rm = TRUE)
  values[is.nan(values)] <- NA_real_
  values
}
```

```{r classify-enhancers}
# Reload saved flags so this chunk also works after rerunning load-fantom.
accessibility_flags <- fread(
  file.path(table_dir, "01_enhancer_accessibility.tsv"),
  select = c("enhancer_id", "accessible")
)
if (anyDuplicated(accessibility_flags$enhancer_id))
  stop("Duplicate enhancer IDs in the accessibility table.")
accessibility_idx <- match(enhancers$enhancer_id, accessibility_flags$enhancer_id)
if (anyNA(accessibility_idx))
  stop("Some enhancer IDs are missing from the accessibility table.")
accessible_values <- as.logical(accessibility_flags$accessible[accessibility_idx])
if (anyNA(accessible_values)) stop("Missing or invalid accessibility flags.")
enhancers[, accessible := accessible_values]

chip_dir <- file.path(annotation_dir, "chip_hg38")
atac_dir <- file.path(annotation_dir, "atac_hg38")
enhancers[, h3k27ac_reps := baseline_peak_support(.SD, baseline_files(chip_dir, "peaks\\.narrowPeak"))]
enhancers[, atac_reps := baseline_peak_support(.SD, baseline_files(atac_dir, "peaks\\.narrowPeak"))]
enhancers[, enhancer_class := fifelse(h3k27ac_reps >= minimum_peak_replicates &
                                       atac_reps >= minimum_peak_replicates, "active", "inactive")]
if (inactive_definition == "neither_signal") {
  enhancers[enhancer_class == "inactive" & (h3k27ac_reps > 0 | atac_reps > 0),
            enhancer_class := "excluded"]
}
stopifnot(comparison_universe %in% c("accessible", "all_fantom"),
          inactive_definition %in% c("not_active", "neither_signal"))
analysis_enhancers <- copy(enhancers[enhancer_class %in% c("active", "inactive") &
                                    (accessible | comparison_universe == "all_fantom")])
if (!nrow(analysis_enhancers)) stop("No enhancers meet the selected universe/class criteria")
analysis_enhancers[, h3k27ac_signal := baseline_signal(.SD, baseline_files(chip_dir, "\\.bw"))]
analysis_enhancers[, atac_signal := baseline_signal(.SD, baseline_files(atac_dir, "\\.bw"))]
save_table(analysis_enhancers[, .(enhancer_id, chr, start0, end0, accessible,
                                 enhancer_class, h3k27ac_reps, atac_reps,
                                 h3k27ac_signal, atac_signal)], "02_enhancer_classes.tsv")
class_summary <- analysis_enhancers[, .(n_enhancers = .N), by = .(accessible, enhancer_class)]
save_table(class_summary, "02_class_counts.tsv")
```

### Sample a balanced enhancer set

Sample 5,000 active and 5,000 inactive enhancers without replacement, using seed
1. The full eligible classes remain in `02_enhancer_classes.tsv`; this chunk
reloads that pool so rerunning it selects the same enhancers. All subsequent
analyses use this sampled set across all four timepoints. The later read-coverage
and ACF quality filters can reduce the number of usable enhancers or fibers.

```{r sample-analysis-enhancers}
enhancers_per_class <- 5000L
enhancer_sampling_seed <- 1L
analysis_pool <- fread(file.path(table_dir, "02_enhancer_classes.tsv"))
stopifnot(!anyDuplicated(analysis_pool$enhancer_id))
available_classes <- analysis_pool[, .(n_enhancers = .N), by = enhancer_class]
if (!setequal(available_classes$enhancer_class, c("active", "inactive")) ||
    any(available_classes$n_enhancers < enhancers_per_class))
  stop("Need at least ", enhancers_per_class, " enhancers in each of the active and inactive classes.")
data.table::setorder(analysis_pool, enhancer_class, enhancer_id)
set.seed(enhancer_sampling_seed)
analysis_enhancers <- analysis_pool[
  , .SD[sample.int(.N, enhancers_per_class, replace = FALSE)], by = enhancer_class]
analysis_enhancers[, midpoint0 := as.integer(floor((start0 + end0) / 2))]
save_table(analysis_enhancers, "02_sampled_enhancer_classes.tsv")
sampled_class_summary <- analysis_enhancers[
  , .(n_enhancers = .N), by = .(accessible, enhancer_class)]
save_table(sampled_class_summary, "02_sampled_class_counts.tsv")
```

### Aggregate FIRE profiles around sampled enhancer midpoints

For each track, compare active and inactive enhancers in separate panels for
LPS 0, 5, 10 and 15 minutes. Use the same sampled enhancers at every timepoint
and the same midpoint-centered, 1,000-bp windows as the ACF analysis: BED offsets
[-500, 500), at 1-bp resolution. Each enhancer contributes equally to its class
mean at each position. Recorded zeros are included; missing bigWig values and
windows extending outside a represented chromosome are excluded, with the
number of contributing enhancers saved for each position.

`all.percent.accessible.bw` contains the percentage of fibers carrying a FIRE
at each position. The FIRE, linker/MSP and nucleosome coverage tracks contain
fiber counts; these raw coverage values also depend on sequencing depth.

```{r fire-profile-settings}
fire_bw_root <- "/project/spott/lizarraga/pacbio_analysis/macrophage_project/merged_hifi_bams/FIRE"
fire_profile_flank <- 500L
fire_profile_tracks <- data.table(
  track = c("percent_accessible", "fire_coverage", "linker_coverage", "nucleosome_coverage"),
  filename = c("all.percent.accessible.bw", "all.fire.coverage.bw",
               "all.linker.coverage.bw", "all.nucleosome.coverage.bw"),
  title = c("FIRE accessibility", "FIRE coverage", "Linker/MSP coverage", "Nucleosome coverage"),
  y_label = c("Mean accessible fibers (%)", rep("Mean coverage (fibers)", 3L))
)
```

```{r aggregate-fire-profile-function}
aggregate_fire_profile <- function(regions, path, flank = 500L) {
  stopifnot(file.exists(path), flank > 0L,
            !anyDuplicated(regions$enhancer_id),
            !anyNA(regions[, .(chr, midpoint0, enhancer_class)]))
  bw <- rtracklayer::BigWigFile(path)
  chromosome_lengths <- GenomeInfoDb::seqlengths(GenomeInfoDb::seqinfo(bw))
  left0 <- regions$midpoint0 - flank
  right0 <- regions$midpoint0 + flank
  chromosome_end <- unname(chromosome_lengths[regions$chr])
  valid <- !is.na(chromosome_end) & left0 >= 0L & right0 <= chromosome_end
  signal <- matrix(NA_real_, nrow(regions), 2L * flank)
  for (chromosome in unique(regions$chr[valid])) {
    idx <- which(valid & regions$chr == chromosome)
    query <- GenomicRanges::GRanges(chromosome,
      IRanges::IRanges(left0[idx] + 1L, right0[idx]))
    # One summary value per genomic base; retain absent values as NA.
    signal[idx, ] <- summary(bw, which = query, size = 2L * flank,
                             type = "mean", defaultValue = NA_real_, as = "matrix")
  }
  signal[!is.finite(signal)] <- NA_real_
  data.table::rbindlist(lapply(c("active", "inactive"), function(class_id) {
    idx <- which(regions$enhancer_class == class_id)
    values <- signal[idx, , drop = FALSE]
    n <- colSums(!is.na(values))
    means <- colMeans(values, na.rm = TRUE)
    means[n == 0L] <- NA_real_
    data.table(enhancer_class = class_id, position_bp = seq.int(-flank, flank - 1L),
               mean_signal = means, n_enhancers = n,
               n_sampled = length(idx), n_windows_in_bounds = sum(valid[idx]))
  }))
}
```

```{r calculate-fire-profiles}
fire_profile_parts <- list()
for (i in seq_len(nrow(fire_profile_tracks))) {
  for (j in seq_len(nrow(sample_table))) {
    bw_path <- file.path(fire_bw_root, sample_table$sample[j], "trackHub-v0.1/bw",
                         fire_profile_tracks$filename[i])
    message("Profiling ", sample_table$sample[j], ": ", fire_profile_tracks$track[i])
    profile <- aggregate_fire_profile(analysis_enhancers, bw_path, fire_profile_flank)
    profile[, `:=`(sample = sample_table$sample[j], time_min = sample_table$time_min[j],
                   track = fire_profile_tracks$track[i])]
    fire_profile_parts[[length(fire_profile_parts) + 1L]] <- profile
  }
}
fire_profiles <- rbindlist(fire_profile_parts)
save_table(fire_profiles, "02_fire_aggregate_profiles.tsv")
```

```{r plot-fire-profiles, fig.width=10, fig.height=7}
for (i in seq_len(nrow(fire_profile_tracks))) {
  plot_data <- copy(fire_profiles[track == fire_profile_tracks$track[i]])
  plot_data[, timepoint := factor(paste0(time_min, " min"),
                                  levels = paste0(sample_table$time_min, " min"))]
  p_fire_profile <- ggplot(plot_data,
    aes(position_bp, mean_signal, color = enhancer_class)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey65") +
    geom_line(linewidth = 0.7, na.rm = TRUE) +
    facet_wrap(~timepoint, ncol = 2) +
    scale_color_manual(values = c(active = "#D55E00", inactive = "#0072B2")) +
    scale_x_continuous(breaks = c(-fire_profile_flank, 0, fire_profile_flank),
                       limits = c(-fire_profile_flank, fire_profile_flank)) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(title = fire_profile_tracks$title[i],
         x = "Position relative to enhancer midpoint (bp)",
         y = fire_profile_tracks$y_label[i], color = NULL,
         caption = "Mean across sampled enhancers; missing values excluded.") +
    theme_classic() + theme(legend.position = "bottom")
  print(p_fire_profile)
  save_pdf(p_fire_profile, paste0("02_aggregate_", fire_profile_tracks$track[i], ".pdf"),
           width = 10, height = 7)
}
```

```{r baseline-signal-plot}
signal_long <- melt(analysis_enhancers, id.vars = c("enhancer_id", "enhancer_class"),
                    measure.vars = c("h3k27ac_signal", "atac_signal"),
                    variable.name = "assay", value.name = "signal")
head(signal_long)
p_signal <- ggplot(signal_long, aes(enhancer_class, signal, fill = enhancer_class)) +
  geom_boxplot(outlier.shape = NA) + facet_wrap(~assay, scales = "free_y") +
  scale_y_continuous(trans = "log1p") + theme_classic() +
  labs(x = NULL, y = "Mean baseline bigWig signal (log1p scale)") +
  theme(legend.position = "none")
save_pdf(p_signal, "02_baseline_H3K27ac_ATAC.pdf")

p_signal
```


### Pool unbinned single-fiber autocorrelation profiles within each enhancer class

Each enhancer contributes all molecules fully spanning its midpoint +/- 500 bp,
pooled across the four timepoints within its baseline active/inactive class. Each
class pool is then uniformly sampled to **10,000 fibers without replacement**
(seed 1); if fewer are available, all are used. This is the **same fiber sample**
used by the Manhattan–Leiden analysis below: both jobs extract the pooled class in
the same order and apply the same seed, and the loading chunk for the Manhattan
results checks that the two fiber sets are identical. Each
row contains **1,000 consecutive genomic bases**, with 1 for an m6A call and 0 for
no call. This follows the binary input used by the autocorrelation workflow:
G/C bases are also zero. There is no A/T normalization, binning, smoothing or
imputation. A molecule spanning multiple enhancer windows contributes separately
to each window, and enhancers with more full-span fibers contribute more rows.

The batch worker imports `03_compute_autocorrelations.py` and
`04_cluster_autocorrelations.py` directly from `code/clustering_methods/auto_correlation`
through `reticulate`. Those files are read without modification. For each fiber,
the first function computes the mean-centered, variance-normalized linear ACF:

\[
\mathrm{ACF}(k) =
\frac{\sum_{t=0}^{999-k}(x_t-\bar{x})(x_{t+k}-\bar{x})}
     {\sum_{t=0}^{999}(x_t-\bar{x})^2},\qquad k=0,\ldots,999.
\]

All 1,000 lag features are retained at **1-bp resolution**. Constant fibers
(including all-zero fibers) remain in the QC counts; their ACF is undefined, so
they are excluded from clustering. Lag 0 is retained for clustering and becomes
a constant-zero feature when PCA centers the columns.

The reference clustering function uses centered, unscaled PCA (up to 50 PCs),
correlation-distance neighbors on the PCA scores (10 neighbors), the UMAP
connectivity graph and Leiden at resolution 0.4, with seed 0. UMAP coordinates
are for display. Each enhancer class is clustered independently; cluster numbers
have no correspondence between active and inactive classes. All valid fibers
among the 10,000 sampled per class enter clustering. ACF calculation is batched; the reference
PCA/neighbors step still holds the sampled class in memory.

```{r acf-job-files}
acf_script <- file.path(project_root, "code/enhancer/run_enhancer_acf.sh")
acf_results_file <- file.path(table_dir, "enhancer_acf_results.rds")
acf_max_fibers <- 10000L  # same pooled-class fiber sample as the Manhattan job
acf_sampling_seed <- 1L
```

The batch job reads `02_sampled_enhancer_classes.tsv`, pools all four timepoints
within each class and samples 10,000 fibers per class. Its settings are in
`code/enhancer/run_enhancer_acf.R`.


```{r submit-enhancer-acf, eval=FALSE}
acf_status <- system2("sbatch", shQuote(acf_script))
if (acf_status != 0L) stop("Slurm enhancer ACF job submission failed")
```

```{r load-enhancer-acf-results}
if (!file.exists(acf_results_file))
  stop("ACF results are missing. Submit the Slurm job and wait for it to finish.")
acf_saved <- readRDS(acf_results_file)
current_acf_regions <- as.data.frame(analysis_enhancers[, .(
  enhancer_id = as.character(enhancer_id), enhancer_class = as.character(enhancer_class),
  chr = as.character(chr), midpoint0 = as.integer(midpoint0))])
current_acf_regions <- current_acf_regions[order(current_acf_regions$enhancer_id), , drop = FALSE]
rownames(current_acf_regions) <- NULL
if (!identical(acf_saved$inputs$regions, current_acf_regions) ||
    !identical(acf_saved$inputs$samples, as.character(sample_table$sample)) ||
    !identical(acf_saved$parameters$max_fibers, acf_max_fibers) ||
    !identical(acf_saved$parameters$sampling_seed, acf_sampling_seed))
  stop("Saved ACF results do not match the current sampled enhancers/timepoints/fiber sampling. Resubmit the job.")
enhancer_acf_results <- acf_saved$results
stopifnot(setequal(names(enhancer_acf_results), unique(analysis_enhancers$enhancer_class)))
rm(acf_saved, current_acf_regions)
```

```{r acf-qc-summary-function}
summarize_acf_qc <- function(results) {
  rbindlist(lapply(names(results), function(class_id) {
    result <- results[[class_id]]
    extraction <- copy(as.data.table(result$extraction_qc))
    assignments <- as.data.table(result$assignments)
    a <- extraction[, .(n_enhancers = uniqueN(enhancer_id),
                        n_reads = sum(n_reads)), by = status]
    a[, stage := "extraction"]
    b <- assignments[, .(n_enhancers = uniqueN(enhancer_id), n_reads = .N),
                       by = .(status = acf_status)]
    b[, stage := "autocorrelation_clustering"]
    out <- rbindlist(list(a, b), use.names = TRUE)
    out[, enhancer_class := class_id]
    out[, .(enhancer_class, stage, status, n_enhancers, n_reads)]
  }), use.names = TRUE)
}
```

```{r acf-cluster-counts-function}
count_acf_clusters <- function(results) {
  counts <- rbindlist(lapply(names(results), function(class_id) {
    dt <- copy(as.data.table(results[[class_id]]$assignments))
    dt[, cluster := as.character(cluster)]
    dt[is.na(cluster), cluster := "unclustered"]
    dt[, .(n_reads = .N, n_enhancers = uniqueN(enhancer_id)),
       by = .(enhancer_class, sample_name, cluster)]
  }), use.names = TRUE)
  if (nrow(counts))
    counts[, fraction := n_reads / sum(n_reads), by = .(enhancer_class, sample_name)]
  else counts[, fraction := numeric()]
  counts
}
```

```{r acf-profile-table-function}
acf_profile_table <- function(results, component = "acf_profiles") {
  out <- rbindlist(lapply(names(results), function(class_id) {
    mat <- results[[class_id]][[component]]
    if (is.null(mat) || !nrow(mat)) return(NULL)
    data.table(enhancer_class = class_id,
               cluster = rep(rownames(mat), each = ncol(mat)),
               position_bp = rep(as.integer(colnames(mat)), nrow(mat)),
               mean_value = as.vector(t(mat)))
  }), use.names = TRUE)
  if (!ncol(out)) out <- data.table(enhancer_class = character(), cluster = character(),
                                   position_bp = integer(), mean_value = numeric())
  out
}
```

```{r acf-profile-plot-function}
plot_acf_profiles <- function(profiles, class_id, profile_type = "acf") {
  dt <- copy(profiles[enhancer_class == class_id])
  if (profile_type == "acf") dt <- dt[position_bp > 0L]
  cluster_levels <- unique(dt$cluster)
  cluster_levels <- cluster_levels[order(as.integer(sub("cluster", "", cluster_levels)))]
  dt[, cluster := factor(cluster, levels = cluster_levels)]
  p <- ggplot(dt, aes(position_bp, mean_value)) +
    geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.25) +
    geom_line(linewidth = 0.3, colour = "#176B87") +
    facet_wrap(~cluster, ncol = 1) + theme_bw(base_size = 11) +
    labs(title = paste(class_id, "enhancers"),
         x = if (profile_type == "acf") "Lag (bp)" else "Position relative to midpoint (bp)",
         y = if (profile_type == "acf") "Mean autocorrelation" else "Fraction of fibers with m6A")
  if (profile_type != "acf") p <- p + geom_vline(xintercept = 0, linetype = 2,
                                                colour = "grey45", linewidth = 0.3)
  p
}
```

```{r acf-composition-plot-function}
plot_acf_composition <- function(counts, class_id, samples) {
  dt <- copy(counts[enhancer_class == class_id])
  cluster_levels <- setdiff(unique(dt$cluster), "unclustered")
  cluster_levels <- cluster_levels[order(as.integer(sub("cluster", "", cluster_levels)))]
  if ("unclustered" %in% dt$cluster) cluster_levels <- c(cluster_levels, "unclustered")
  dt[, cluster := factor(cluster, levels = cluster_levels)]
  dt[, sample_name := factor(sample_name, levels = samples$sample)]
  ggplot(dt, aes(sample_name, fraction, fill = cluster)) +
    geom_col(width = 0.8) +
    scale_x_discrete(drop = FALSE, labels = setNames(samples$time_min, samples$sample)) +
    scale_y_continuous(limits = c(0, 1), labels = function(x) paste0(100 * x, "%")) +
    theme_bw(base_size = 11) +
    labs(title = paste(class_id, "enhancers"), x = "Time after LPS (min)",
         y = "Fraction of extracted fibers", fill = "Cluster")
}
```

```{r acf-umap-plot-function}
plot_acf_umap <- function(assignments, class_id, samples, colour_by = "cluster") {
  dt <- copy(as.data.table(assignments))[acf_status == "clustered" &
                                         is.finite(UMAP1) & is.finite(UMAP2)]
  if (!nrow(dt)) return(NULL)
  if (colour_by == "sample_name") {
    dt[, plot_group := factor(sample_name, levels = samples$sample,
                               labels = paste0(samples$time_min, " min"))]
  } else {
    groups <- unique(as.character(dt$cluster))
    groups <- groups[order(as.integer(sub("cluster", "", groups)))]
    dt[, plot_group := factor(cluster, levels = groups)]
  }
  ggplot(dt, aes(UMAP1, UMAP2, colour = plot_group)) +
    geom_point(size = 0.35, alpha = 0.45, stroke = 0) +
    guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
    coord_equal() + theme_bw(base_size = 11) +
    labs(title = paste(class_id, "enhancers"),
         colour = if (colour_by == "sample_name") "Time after LPS" else "Cluster")
}
```

```{r enhancer-acf-summary-tables}
acf_fiber_sampling <- rbindlist(lapply(names(enhancer_acf_results), function(class_id) {
  x <- as.data.table(enhancer_acf_results[[class_id]]$fiber_sampling)
  x[, enhancer_class := class_id]
  x
}))
save_table(acf_fiber_sampling, "acf_fiber_sampling.tsv")
acf_qc_summary <- summarize_acf_qc(enhancer_acf_results)
save_table(acf_qc_summary, "acf_qc.tsv")
acf_cluster_counts <- count_acf_clusters(enhancer_acf_results)
save_table(acf_cluster_counts, "acf_cluster_counts.tsv")
acf_cluster_profiles <- acf_profile_table(enhancer_acf_results)
acf_profiles_export <- copy(acf_cluster_profiles)
setnames(acf_profiles_export, c("position_bp", "mean_value"), c("lag_bp", "mean_acf"))
save_table(acf_profiles_export, "acf_cluster_profiles.tsv")
m6a_base_profiles <- acf_profile_table(enhancer_acf_results, "m6a_profiles")
head(m6a_base_profiles)
```

The ACF curves show every positive integer lag without binning or smoothing;
lag zero is retained in the table. The m6A curves show each genomic base, with
the proportion of fibers carrying an m6A call at that midpoint-relative base.
Composition percentages include all sampled fibers in each timepoint's
denominator; fibers excluded from clustering appear as **unclustered**, with
the reasons summarized in the QC table. Extraction QC counts all eligible
fibers; clustering QC and plots describe only the 10,000 sampled fibers. UMAP plots contain only fibers with
computed coordinates. Cluster labels are specific to each enhancer class.

```{r enhancer-acf-pdfs, results='hide'}
for (class_id in names(enhancer_acf_results)) {
  if (any(acf_cluster_profiles$enhancer_class == class_id)) {
    n_clusters <- uniqueN(acf_cluster_profiles[enhancer_class == class_id, cluster])
    save_pdf(plot_acf_profiles(acf_cluster_profiles, class_id),
             paste0("acf_cluster_profiles_", class_id, ".pdf"),
             width = 8, height = max(4, 1.5 * n_clusters + 1))
    save_pdf(plot_acf_profiles(m6a_base_profiles, class_id, "m6a"),
             paste0("acf_m6a_base_cluster_profiles_", class_id, ".pdf"),
             width = 8, height = max(4, 1.5 * n_clusters + 1))
  }
  if (any(acf_cluster_counts$enhancer_class == class_id))
    save_pdf(plot_acf_composition(acf_cluster_counts, class_id, sample_table),
             paste0("acf_cluster_composition_", class_id, ".pdf"))
  for (colour_by in c("cluster", "sample_name")) {
    p <- plot_acf_umap(enhancer_acf_results[[class_id]]$assignments,
                       class_id, sample_table, colour_by)
    if (!is.null(p)) save_pdf(p, paste0("acf_umap_", colour_by, "_", class_id, ".pdf"))
  }
}
```

### Pool unbinned single-fiber patterns for Manhattan–Leiden clustering

This is a second clustering analysis of the **raw binary m6A patterns** in the
same midpoint +/- 500-bp windows. Each class pools eligible full-span fibers from its
5,000 sampled enhancers and all four timepoints, then uniformly samples
**10,000 fibers without replacement** (seed 1) before neighbor search. If fewer
than 10,000 are available, all are used. Sampling is across the pooled class,
without forcing equal numbers per enhancer or timepoint. Fiber samples match
ACF only when both jobs use the same extraction order, enhancer set and seed. Active and inactive classes are
clustered independently. The shared extractor retains **1,000 individual bases
(-500 through +499)**, with 1 for a call and 0 for no call, including G/C bases.
There is no binning, smoothing, PCA, imputation or A/T normalization. A fiber spanning multiple enhancers contributes once per
window; enhancers with more fibers contribute more rows. All-zero and other
constant fibers are included because Manhattan distance remains defined.

For two fibers, Manhattan distance is the number of midpoint-relative bases
where their binary calls differ. The graph method follows
`code/clustering_methods/Leiden_Manhattan/marcophage_leiden_manhattan.Rmd`,
with the requested neighbor count changed to **100**:
100 neighbors per fiber, edge weight `exp(-distance / sigma)`, with `sigma` equal
to the mean directed neighbor distance (1 if all distances are zero), and an
undirected union graph. Leiden calls the original `leiden_partition()` function
with the modularity objective, resolution 1, seed 1 and iterations to convergence.
Clusters are numbered largest first within each class; their labels do not
correspond across classes or to the ACF clusters.

The default uses **approximate Manhattan neighbors** to avoid the
reference's full pairwise distance matrix. All selected edge distances are
recomputed exactly from the original 1-bp patterns. This can select different
neighbors/clusters from the reference, including different ties among identical
patterns. Set `manhattan_neighbor_method <- "exact"` below to use every pairwise
distance with the reference's input-order tie handling. Exact search works in
memory-limited blocks but still takes quadratic time. Both modes use the same sampled
fibers; classes with fewer than three fibers are reported as unclustered.

```{r manhattan-job-files}
manhattan_script <- file.path(project_root, "code/enhancer/run_enhancer_manhattan.sh")
manhattan_classes <- c("active", "inactive")
manhattan_max_fibers <- 10000L
manhattan_k_neighbors <- 100L
manhattan_results_files <- setNames(file.path(table_dir,
  paste0("enhancer_manhattan_", manhattan_classes, "_results.rds")), manhattan_classes)
manhattan_neighbor_method <- "approximate" # or "exact" (much slower for the pooled set)
manhattan_job_id_file <- file.path(table_dir, "manhattan_array_job_id.txt")
if (file.exists(manhattan_job_id_file))
  manhattan_job_id <- readLines(manhattan_job_id_file, n = 1L)
```


```{r submit-enhancer-manhattan, eval=FALSE}
stopifnot(manhattan_neighbor_method %in% c("approximate", "exact"))
manhattan_submission <- system2("sbatch", c("--parsable",
  shQuote(paste0("--export=ALL,ENHANCER_MANHATTAN_NEIGHBORS=", manhattan_neighbor_method)),
  shQuote(manhattan_script)), stdout = TRUE)
if (!is.null(attr(manhattan_submission, "status")) ||
    length(manhattan_submission) != 1L || !grepl("^[0-9]+(;.*)?$", manhattan_submission))
  stop("Slurm enhancer Manhattan job submission failed")
manhattan_job_id <- sub(";.*$", "", manhattan_submission)
writeLines(manhattan_job_id, manhattan_job_id_file)
message("Submitted Manhattan array ", manhattan_job_id, ". Wait for both tasks before loading.")
```

```{r load-enhancer-manhattan-results}
missing_manhattan <- manhattan_results_files[!file.exists(manhattan_results_files)]
if (length(missing_manhattan))
  stop("Wait for both Manhattan array tasks. Missing: ", paste(basename(missing_manhattan), collapse = ", "))
manhattan_saved <- lapply(manhattan_results_files, readRDS)
for (class_id in manhattan_classes) {
  saved <- manhattan_saved[[class_id]]
  # The completed job's input snapshot defines this analysis, independently of
  # any later resampling/reclassification in the interactive session.
  regions <- as.data.table(saved$inputs$regions)
  assignments <- as.data.table(saved$results[[class_id]]$assignments)
  stopifnot(all(c("enhancer_id", "enhancer_class", "chr", "midpoint0") %in% names(regions)),
            nrow(regions) > 0L, !anyNA(regions), !anyDuplicated(regions$enhancer_id),
            all(regions$enhancer_class == class_id),
            all(assignments$enhancer_id %in% regions$enhancer_id),
            all(assignments$enhancer_class == class_id),
            all(assignments$sample_name %in% saved$inputs$samples),
            !anyDuplicated(assignments$RID),
            nrow(assignments) == saved$results[[class_id]]$fiber_sampling$n_selected)
  if (exists("analysis_enhancers", inherits = FALSE)) {
    current_ids <- analysis_enhancers$enhancer_id[analysis_enhancers$enhancer_class == class_id]
    if (!setequal(regions$enhancer_id, current_ids))
      message("Manhattan ", class_id, ": using ", nrow(regions),
              " enhancers recorded in the completed job; the current R-session sample differs.")
  }
  if (!identical(saved$parameters$k_neighbors, manhattan_k_neighbors))
    stop("Saved Manhattan ", class_id, " results use k=", saved$parameters$k_neighbors,
         "; submit the array for k=", manhattan_k_neighbors, ".")
  if (!identical(saved$inputs$samples, as.character(sample_table$sample)) ||
      !identical(saved$parameters$neighbor_method, manhattan_neighbor_method) ||
      !identical(saved$parameters$max_fibers, manhattan_max_fibers) ||
      !identical(saved$parameters$sampling_seed, 1L))
    stop("Saved Manhattan ", class_id, " results do not match current timepoints/method/fiber sampling.")
  stopifnot(identical(saved$enhancer_class, class_id), identical(names(saved$results), class_id),
            identical(saved$parameters$position_bp, -500:499),
            identical(saved$parameters$signal, "raw_binary_m6a"))
}
manhattan_array_ids <- vapply(manhattan_saved, function(x) x$array_job_id, character(1))
if (length(unique(manhattan_array_ids)) != 1L)
  stop("Class results come from different arrays. Wait for both tasks of the current array.")
if (exists("manhattan_job_id", inherits = FALSE) &&
    !all(manhattan_array_ids == manhattan_job_id))
  warning("Loading validated results from completed array ", unique(manhattan_array_ids),
          "; the latest submission record is ", manhattan_job_id,
          ". A newer submission does not replace results until it completes.", call. = FALSE)
enhancer_manhattan_results <- setNames(lapply(manhattan_classes,
  function(class_id) manhattan_saved[[class_id]]$results[[class_id]]), manhattan_classes)
manhattan_analysis_enhancers <- rbindlist(lapply(manhattan_saved, function(x) x$inputs$regions))
# Standalone Manhattan plots remain valid when ACF used a different sample.
manhattan_acf_same_fibers <- NULL
if (exists("enhancer_acf_results", inherits = FALSE)) {
  manhattan_acf_same_fibers <- setNames(vapply(manhattan_classes, function(class_id)
    setequal(enhancer_manhattan_results[[class_id]]$assignments$RID,
             enhancer_acf_results[[class_id]]$assignments$RID), logical(1)), manhattan_classes)
  if (!all(manhattan_acf_same_fibers))
    warning("ACF and Manhattan used different fiber samples. Standalone Manhattan plots are valid; do not treat the samples as paired.", call. = FALSE)
}
rm(manhattan_saved, manhattan_array_ids, regions, assignments, saved, missing_manhattan)
```

`manhattan_analysis_enhancers` is the enhancer input snapshot from the completed
Manhattan jobs. It is independent of the current `analysis_enhancers`; loading
results does not resample or change inputs for other analyses.

The loaded `enhancer_manhattan_results` contains per-fiber assignments with
original read IDs, enhancer IDs, classes, timepoints and coordinates;
`m6a_profiles` contains the cluster means at each base. `fiber_sampling` records
the eligible and selected fiber counts. Assignments contain only sampled fibers. `extraction_qc` records
enhancers without usable full-span fibers, and `info` records the graph settings
and any exact repairs of incomplete approximate neighbor lists. These R results
include the sampled sparse 1-bp feature matrix and selected-fiber footprint
records for plotting; the neighbor graph is omitted. Extraction QC counts
all eligible fibers; clustering QC and plots describe only sampled fibers.

```{r enhancer-manhattan-summary-tables}
manhattan_sampling <- rbindlist(lapply(names(enhancer_manhattan_results), function(class_id) {
  x <- as.data.table(enhancer_manhattan_results[[class_id]]$fiber_sampling)
  x[, enhancer_class := class_id]
  x
}))
save_table(manhattan_sampling, "manhattan_fiber_sampling.tsv")
manhattan_qc <- rbindlist(lapply(names(enhancer_manhattan_results), function(class_id) {
  result <- enhancer_manhattan_results[[class_id]]
  extraction <- as.data.table(result$extraction_qc)[,
    .(n_enhancers = uniqueN(enhancer_id), n_reads = sum(n_reads)), by = status]
  extraction[, stage := "extraction"]
  clustered <- as.data.table(result$assignments)[,
    .(n_enhancers = uniqueN(enhancer_id), n_reads = .N), by = .(status = clustering_status)]
  clustered[, stage := "manhattan_clustering"]
  out <- rbindlist(list(extraction, clustered), use.names = TRUE)
  out[, enhancer_class := class_id]
  out[, .(enhancer_class, stage, status, n_enhancers, n_reads)]
}))
save_table(manhattan_qc, "manhattan_qc.tsv")
# The summary/plot helpers above also accept these per-fiber cluster assignments.
manhattan_cluster_counts <- count_acf_clusters(enhancer_manhattan_results)
save_table(manhattan_cluster_counts, "manhattan_cluster_counts.tsv")
manhattan_m6a_profiles <- acf_profile_table(enhancer_manhattan_results, "m6a_profiles")
manhattan_profiles_export <- copy(manhattan_m6a_profiles)
setnames(manhattan_profiles_export, "mean_value", "mean_m6a")
save_table(manhattan_profiles_export, "manhattan_m6a_base_cluster_profiles.tsv")
head(manhattan_cluster_counts)
```

Profiles show unsmoothed means at each base. Composition plots show the fiber
proportions among sampled fibers within each timepoint, using clusters fitted to all timepoints
pooled within that class. These are fiber-weighted summaries, so equal enhancer
counts do not imply equal numbers of fibers across classes or timepoints.

The m6A heatmap includes every sampled fiber, grouped by Manhattan cluster and
annotated by timepoint. Footprint profiles use all sampled fibers and count
per-base occupancy of FiberHMM nucleosomes >90 bp and TF footprints <60 bp.
Single-fiber PDFs have one page per cluster, with up to 100 randomly selected
fibers for legibility (display only; seed 1), split by timepoint. All tracks align
to the enhancer midpoint at 1-bp resolution. **k=100 is the neighbor count**;
profile curves are unsmoothed. Existing k=10 results must be reclustered.
The array also saves these PDFs automatically; rerun this chunk to regenerate
them from the saved result without extracting data or clustering again.

```{r enhancer-manhattan-pdfs, results='hide'}
suppressPackageStartupMessages({library(cowplot); library(grid); library(ComplexHeatmap)})
manhattan_plot_helpers <- new.env(parent = globalenv())
sys.source(file.path(project_root,
  "code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r"), envir = manhattan_plot_helpers)
sys.source(file.path(project_root, "code/enhancer/enhancer_manhattan_plots.R"),
           envir = manhattan_plot_helpers)
for (class_id in names(enhancer_manhattan_results))
  manhattan_plot_helpers$save_enhancer_manhattan_pdfs(
    enhancer_manhattan_results[[class_id]], class_id, plot_dir, table_dir,
    expected_k = manhattan_k_neighbors)
```

### Extract all TF footprints shorter than 60 bp

This shell script expands every block in the complete FiberHMM TF BED12 files,
retains lengths 1–59 bp, and writes sorted, bgzipped BED6 files with tabix indexes.
The BED6 columns match `split_footprints_by_size.sh`:
`chrom`, `start`, `end`, `read_name`, `size`, `strand`.

Outputs use the existing TF folder layout:
`firehmm_tf/ft_by_size/size0-60/LPS0/LPS0_tf_size0-60_chr1.bed.gz`
(and the corresponding chromosomes/timepoints). These contain the same short
TF calls selected from the parent files by the counting functions below.
The R chunk saves the standalone script in `FiberHMM/extract/code`.
The submission chunk is manual so knitting does not submit another extraction job.
No log files are saved.

```{r function-write-tf-lt60-script}
write_tf_lt60_script <- function(path) {
  script <- r"---(#!/bin/bash
#SBATCH --job-name=extract_tf_lt60
#SBATCH --account=pi-spott
#SBATCH --partition=caslake
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=4
#SBATCH --mem=32G
#SBATCH --time=12:00:00
#SBATCH --array=1-4
#SBATCH --output=/dev/null
#SBATCH --error=/dev/null

# Expand the complete per-chromosome FiberHMM TF BED12+ files into BED6.
# Keep every real block with 0 < length < 60 bp, including 1-9 and 30-39 bp.
# FiberHMM has no first/last sentinel blocks; every block is inspected.
#
# Input : firehmm_tf/<LPSn>/<LPSn>_hmm_extracted_tf_<chrom>.bed.gz
# Output: firehmm_tf/ft_by_size/size0-60/<LPSn>/
#             <LPSn>_tf_size0-60_<chrom>.bed.gz (+ .tbi)
# BED6  : chrom  start  end  read_name  size  strand
# Coordinates are 0-based, half-open. No plots, logs or uncompressed BEDs are saved.
#
# Usage:
#   sbatch extract_tf_lt60.sh          # four array tasks, one per timepoint
#   bash extract_tf_lt60.sh LPS0       # one timepoint locally
#   bash extract_tf_lt60.sh all        # all four timepoints locally
#   bash extract_tf_lt60.sh LPS0 chrM  # one chromosome (small validation run)
# Optional environment overrides: TF_EXTRACT_ROOT, TF_OUTPUT_ROOT,
# TF_SORT_TMP, TF_SORT_MEMORY, TF_THREADS, TF_BGZIP, TF_TABIX.
# Existing complete BED/index pairs are skipped; incomplete pairs are rebuilt.

set -euo pipefail
shopt -s nullglob

CONDS=(LPS0 LPS5 LPS10 LPS15)
EXTRACT_ROOT=${TF_EXTRACT_ROOT:-/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract}
OUT_ROOT=${TF_OUTPUT_ROOT:-${EXTRACT_ROOT}/firehmm_tf/ft_by_size/size0-60}
SORT_TMP=${TF_SORT_TMP:-${TMPDIR:-/scratch/midway3/cshan/tmp}}
SORT_MEMORY=${TF_SORT_MEMORY:-2G}
NCORE=${TF_THREADS:-${SLURM_CPUS_PER_TASK:-4}}
BGZIP=${TF_BGZIP:-/project/spott/cshan/envs/dimelo/bin/bgzip}
TABIX=${TF_TABIX:-/project/spott/cshan/envs/dimelo/bin/tabix}
work_dir=""

die() { echo "ERROR: $*" >&2; exit 1; }
cleanup() { if [[ -n "$work_dir" && -d "$work_dir" ]]; then rm -rf -- "$work_dir"; fi; }
trap cleanup EXIT
trap 'exit 130' INT
trap 'exit 143' TERM

[[ $# -le 2 ]] || die "Usage: bash $0 [LPS0|LPS5|LPS10|LPS15|all] [chromosome]"
if [[ -n "${SLURM_ARRAY_TASK_ID:-}" ]]; then
    [[ $# -eq 0 ]] || die "Submit the four-task Slurm array without positional arguments"
    [[ "$SLURM_ARRAY_TASK_ID" =~ ^[1-4]$ ]] || die "Array task ID must be 1-4"
    LABELS=("${CONDS[$((SLURM_ARRAY_TASK_ID - 1))]}")
else
    label=${1:-all}
    label=${label/_/}
    case "$label" in
        all) LABELS=("${CONDS[@]}") ;;
        LPS0|LPS5|LPS10|LPS15) LABELS=("$label") ;;
        *) die "Unknown timepoint: $label" ;;
    esac
fi
CHROMOSOME=${2:-}
[[ "$NCORE" =~ ^[1-9][0-9]*$ ]] || die "TF_THREADS must be a positive integer"
[[ -x "$BGZIP" && -x "$TABIX" ]] || die "bgzip or tabix executable is missing"
mkdir -p "$SORT_TMP"

for LABEL in "${LABELS[@]}"; do
    IN_DIR="${EXTRACT_ROOT}/firehmm_tf/${LABEL}"
    OUT_DIR="${OUT_ROOT}/${LABEL}"
    [[ -d "$IN_DIR" ]] || die "Input directory not found: $IN_DIR"
    if [[ -n "$CHROMOSOME" ]]; then
        inputs=("${IN_DIR}/${LABEL}_hmm_extracted_tf_${CHROMOSOME}.bed.gz")
    else
        inputs=("${IN_DIR}/${LABEL}_hmm_extracted_tf_"*.bed.gz)
    fi
    [[ ${#inputs[@]} -gt 0 ]] || die "No TF BED12 inputs found in $IN_DIR"
    mkdir -p "$OUT_DIR"

    for in_gz in "${inputs[@]}"; do
        [[ -s "$in_gz" ]] || die "Input missing or empty: $in_gz"
        base=${in_gz##*/}
        chrom=${base#"${LABEL}_hmm_extracted_tf_"}
        chrom=${chrom%.bed.gz}
        out="${OUT_DIR}/${LABEL}_tf_size0-60_${chrom}.bed.gz"
        if [[ -s "$out" && -s "${out}.tbi" ]]; then
            echo "Already extracted: $out"
            continue
        fi
        work_dir=$(mktemp -d "${OUT_DIR}/.tf_lt60_${chrom}.XXXXXX")
        partial="${work_dir}/footprints.bed.gz"

        # All stages belong to this shell pipeline, so pipefail catches failures
        # in decompression, validation, sorting and compression before publishing.
        if ! gzip -cd -- "$in_gz" | awk -v expected_chr="$chrom" '
        BEGIN { FS = OFS = "\t" }
        function fail(message) {
            printf "Invalid BED12 row %d: %s\n", NR, message > "/dev/stderr"
            exit 2
        }
        {
            if (NF < 12) fail("expected at least 12 columns")
            if ($1 != expected_chr) fail("chromosome does not match input filename")
            if ($2 !~ /^[0-9]+$/ || $3 !~ /^[0-9]+$/ || $3 <= $2)
                fail("invalid outer coordinates")
            if ($10 !~ /^[0-9]+$/) fail("invalid blockCount")
            if ($10 == 0) {
                if (($11 != "" && $11 != ".") || ($12 != "" && $12 != "."))
                    fail("nonempty blocks with blockCount zero")
                next
            }
            sub(/,$/, "", $11); sub(/,$/, "", $12)
            ns = split($11, sizes, ","); nt = split($12, offsets, ",")
            if (ns != $10 || nt != $10) fail("block counts disagree")
            for (i = 1; i <= ns; i++) {
                if (sizes[i] !~ /^[0-9]+$/ || sizes[i] <= 0 || offsets[i] !~ /^[0-9]+$/)
                    fail("invalid block size or offset")
                size = sizes[i] + 0
                start = $2 + offsets[i]
                if (start + size > $3) fail("block extends outside outer interval")
                if (size < 60) print $1, start, start + size, $4, size, $6
            }
        }' | LC_ALL=C sort -k1,1 -k2,2n -k3,3n -S "$SORT_MEMORY" -T "$SORT_TMP" \
            | "$BGZIP" -@ "$NCORE" -c > "$partial"; then
            die "Extraction failed for $LABEL $chrom"
        fi
        "$TABIX" -f -p bed "$partial"
        mv -- "$partial" "$out"
        mv -- "${partial}.tbi" "${out}.tbi"
        rmdir -- "$work_dir"
        work_dir=""
        echo "Wrote: $out"
    done
done)---"
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  writeLines(script, path, useBytes = TRUE)
  Sys.chmod(path, mode = "0755")
  invisible(normalizePath(path, mustWork = TRUE))
}
```

```{r save-tf-lt60-script}
tf_lt60_script <- file.path(extract_root, "code/extract_tf_lt60.sh")
write_tf_lt60_script(tf_lt60_script)
```

```{r submit-tf-lt60-extraction, eval=FALSE}
# One Slurm array task per timepoint: LPS0, LPS5, LPS10 and LPS15.
system2("sbatch", shQuote(tf_lt60_script))
# For an interactive single-chromosome run instead:
# system2("bash", c(shQuote(tf_lt60_script), "LPS0", "chrM"))
```

### Count enhancer footprints

The existing size bins omit footprints below 10 bp and from 30–39 bp. Therefore,
the counts below use their complete parent FiberHMM BED12 files and apply the
requested strict cutoffs: TF footprints <60 bp and nucleosome footprints >90 bp.
The 60–90-bp interval is excluded, including footprints exactly 60 or 90 bp.
Every FiberHMM block is a real call; these files have no `ft extract` sentinels.

A footprint is assigned to an enhancer when its midpoint lies inside that
enhancer. This avoids favoring long footprints simply because they overlap more
bases. Only reads spanning the entire enhancer contribute. Read denominators
come from the matching, unfiltered m6A extraction and include reads with zero TF
footprints. Two distinct short footprints on one enhancer–read observation are
called co-occupied; these length-defined footprints do not identify named TFs.

```{r function-footprint-length-bin}
footprint_length_bin <- function(length_bp) {
  # Lower bin edges: 0 = 1–9 bp, 10 = 10–19 bp, ..., 500 = >=500 bp.
  pmin(500L, as.integer(length_bp %/% 10L) * 10L)
}
```

```{r function-stream-footprint-lengths}
stream_footprint_lengths <- function(path, footprint_type, chunk_lines = 5000L) {
  stopifnot(footprint_type %in% c("TF", "nucleosome"), chunk_lines > 0L)
  if (!file.exists(path)) stop("Missing footprint input: ", path)
  con <- gzfile(path, open = "rt")
  on.exit(close(con))
  totals <- numeric(51L)
  repeat {
    lines <- readLines(con, n = chunk_lines, warn = FALSE)
    if (!length(lines)) break
    raw <- data.table::fread(text = paste(lines, collapse = "\n"),
                            header = FALSE, sep = "\t", select = 11L,
                            colClasses = "character", showProgress = FALSE)
    sizes <- as.integer(unlist(strsplit(raw[[1L]], ",", fixed = TRUE),
                               use.names = FALSE))
    if (anyNA(sizes)) stop("Invalid footprint block size in ", path)
    sizes <- if (footprint_type == "TF") sizes[sizes > 0L & sizes < 60L] else
      sizes[sizes > 90L]
    if (length(sizes)) {
      bins <- footprint_length_bin(sizes)
      totals <- totals + tabulate(bins %/% 10L + 1L, nbins = 51L)
    }
  }
  data.table::data.table(length_bin = seq.int(0L, 500L, 10L), n = totals)
}
```

```{r function-expected-footprint-lengths}
expected_footprint_lengths <- function(sample_table, chromosomes, extract_root) {
  result <- vector("list", nrow(sample_table))
  for (s in seq_len(nrow(sample_table))) {
    sample_name <- sample_table$sample[s]
    compact_name <- gsub("_", "", sample_name, fixed = TRUE)
    pieces <- list()
    for (chromosome in chromosomes) {
      for (kind in c("tf", "footprint")) {
        path <- file.path(extract_root, paste0("firehmm_", kind), compact_name,
                          paste0(compact_name, "_hmm_extracted_", kind, "_",
                                 chromosome, ".bed.gz"))
        pieces[[length(pieces) + 1L]] <- stream_footprint_lengths(
          path, if (kind == "tf") "TF" else "nucleosome")
      }
    }
    combined <- data.table::rbindlist(pieces)[, .(n = sum(n)), by = length_bin]
    combined[, `:=`(sample = sample_name, time_min = sample_table$time_min[s])]
    result[[s]] <- combined[, .(sample, time_min, length_bin, n)]
  }
  data.table::rbindlist(result)
}
```

```{r function-count-enhancer-footprints}
count_enhancer_footprints <- function(analysis_enhancers, sample_table,
                                      extract_root, ft_result_dir,
                                      batch_size = 100L) {
  stopifnot(!anyDuplicated(analysis_enhancers$enhancer_id), batch_size > 0L)
  enhancer_parts <- list()
  length_parts <- list()
  for (s in seq_len(nrow(sample_table))) {
    sample_name <- sample_table$sample[s]
    compact_name <- gsub("_", "", sample_name, fixed = TRUE)
    for (chromosome in unique(analysis_enhancers$chr)) {
      enh <- data.table::copy(analysis_enhancers[chr == chromosome])
      m6a_path <- file.path(ft_result_dir, sample_name, "extracted_results",
                           "m6a_by_chr", paste0(sample_name,
                             ".ft_extracted_m6a.", chromosome, ".bed.gz"))
      tf_path <- file.path(extract_root, "firehmm_tf", compact_name,
                          paste0(compact_name, "_hmm_extracted_tf_",
                                 chromosome, ".bed.gz"))
      nuc_path <- file.path(extract_root, "firehmm_footprint", compact_name,
                           paste0(compact_name, "_hmm_extracted_footprint_",
                                  chromosome, ".bed.gz"))
      for (first in seq.int(1L, nrow(enh), by = batch_size)) {
        last <- min(nrow(enh), first + batch_size - 1L)
        current <- enh[first:last]
        enh_gr <- bed_ranges(current)
        query_gr <- GenomicRanges::reduce(enh_gr)
        raw_spans <- tabix_bed(m6a_path, query_gr)
        coverage <- data.table::data.table(enhancer_id = character(),
                                          read_id = character())
        if (nrow(raw_spans)) {
          spans <- unique(data.table::data.table(
            chr = as.character(raw_spans[[1L]]), start0 = as.integer(raw_spans[[2L]]),
            end0 = as.integer(raw_spans[[3L]]), read_id = as.character(raw_spans[[4L]])))
          spans[, alignment_width := end0 - start0]
          data.table::setorderv(spans, c("read_id", "alignment_width"), c(1L, -1L))
          spans <- unique(spans, by = "read_id")
          hits <- GenomicRanges::findOverlaps(enh_gr, bed_ranges(spans),
                                               type = "within", ignore.strand = TRUE)
          coverage <- unique(data.table::data.table(
            enhancer_id = current$enhancer_id[S4Vectors::queryHits(hits)],
            read_id = spans$read_id[S4Vectors::subjectHits(hits)]))
        }
        observed <- list()
        for (kind in c("TF", "nucleosome")) {
          raw <- tabix_bed(if (kind == "TF") tf_path else nuc_path, query_gr)
          fp <- expand_hmm_blocks(raw)
          fp <- if (kind == "TF") fp[length > 0L & length < 60L] else fp[length > 90L]
          if (!nrow(fp) || !nrow(coverage)) next
          fp <- unique(fp, by = c("chr", "read_id", "start0", "end0"))
          # Convert BED midpoint to a 1-based point; length remains unclipped.
          midpoint <- floor((fp$start0 + fp$end0) / 2) + 1L
          mid_gr <- GenomicRanges::GRanges(fp$chr, IRanges::IRanges(midpoint, midpoint))
          hits <- GenomicRanges::findOverlaps(mid_gr, enh_gr, ignore.strand = TRUE)
          if (!length(hits)) next
          assigned <- data.table::data.table(
            enhancer_id = current$enhancer_id[S4Vectors::subjectHits(hits)],
            read_id = fp$read_id[S4Vectors::queryHits(hits)],
            length = fp$length[S4Vectors::queryHits(hits)])
          assigned <- merge(assigned, coverage, by = c("enhancer_id", "read_id"),
                            all = FALSE, sort = FALSE)
          if (!nrow(assigned)) next
          assigned[, footprint_type := kind]
          observed[[kind]] <- assigned
        }
        summary <- current[, .(enhancer_id, enhancer_class)]
        summary[, `:=`(n_reads = 0L, n_cooccupied = 0L, n_tf = 0L, n_nuc = 0L)]
        if (nrow(coverage)) {
          covered_counts <- coverage[, .(n_reads = .N), by = enhancer_id]
          summary[covered_counts, n_reads := i.n_reads, on = "enhancer_id"]
        }
        if (length(observed)) {
          calls <- data.table::rbindlist(observed)
          tf_counts <- calls[footprint_type == "TF", .N, by = .(enhancer_id, read_id)]
          if (nrow(tf_counts)) {
            tf_summary <- tf_counts[, .(n_tf = sum(N), n_cooccupied = sum(N >= 2L)),
                                     by = enhancer_id]
            summary[tf_summary, `:=`(n_tf = i.n_tf, n_cooccupied = i.n_cooccupied),
                    on = "enhancer_id"]
          }
          nuc_counts <- calls[footprint_type == "nucleosome", .(n_nuc = .N),
                              by = enhancer_id]
          if (nrow(nuc_counts)) summary[nuc_counts, n_nuc := i.n_nuc, on = "enhancer_id"]
          calls <- merge(calls, current[, .(enhancer_id, enhancer_class)],
                         by = "enhancer_id", all.x = TRUE, sort = FALSE)
          calls[, length_bin := footprint_length_bin(length)]
          lengths <- calls[, .(n = .N), by = .(enhancer_class, length_bin)]
          lengths[, `:=`(sample = sample_name, time_min = sample_table$time_min[s])]
          length_parts[[length(length_parts) + 1L]] <- lengths
        }
        summary[, `:=`(sample = sample_name, time_min = sample_table$time_min[s])]
        stopifnot(all(summary$n_cooccupied <= summary$n_reads),
                  all(summary$n_tf >= 2L * summary$n_cooccupied))
        enhancer_parts[[length(enhancer_parts) + 1L]] <- summary[
          , .(enhancer_id, sample, time_min, enhancer_class,
              n_reads, n_cooccupied, n_tf, n_nuc)]
      }
    }
  }
  counts <- data.table::rbindlist(enhancer_parts)
  if (length(length_parts)) {
    lengths <- data.table::rbindlist(length_parts)[
      , .(n = sum(n)), by = .(sample, time_min, enhancer_class, length_bin)]
  } else {
    lengths <- data.table::data.table(sample = character(), time_min = numeric(),
                                     enhancer_class = character(),
                                     length_bin = integer(), n = numeric())
  }
  list(footprint_counts = counts, length_counts = lengths)
}
```

```{r count-footprints}
footprint_result <- count_enhancer_footprints(
  analysis_enhancers, sample_table, extract_root, ft_result_dir)
footprint_counts <- footprint_result$footprint_counts
length_counts <- footprint_result$length_counts
save_table(footprint_counts, "enhancer_footprint_counts.tsv")

# The expected length distribution is descriptive: genome-wide called TF
# footprints <60 bp plus nucleosome footprints >90 bp on analyzed chromosomes.
# It is not a genomic-position shuffle or an estimate of random TF binding.
stopifnot(expected_background %in% c("genomewide", "inactive"))
if (expected_background == "genomewide") {
  expected_length_counts <- expected_footprint_lengths(
    sample_table, standard_chromosomes, extract_root)
} else {
  expected_length_counts <- merge(
    CJ(sample = sample_table$sample, length_bin = seq.int(0L, 500L, 10L)),
    length_counts[enhancer_class == "inactive", .(sample, length_bin, n)],
    by = c("sample", "length_bin"), all.x = TRUE)
  expected_length_counts[is.na(n), n := 0]
  expected_length_counts[, time_min := sample_table$time_min[match(sample, sample_table$sample)]]
}
head(length_counts)
head(expected_length_counts)
```

### Compare short footprints and nucleosome footprints

Frequencies below are normalized within each class and timepoint over the two
requested footprint categories, <60 bp and >90 bp; lengths 60–90 bp are excluded.
Genome-wide expected frequencies use the same call sources and chromosomes but
are not restricted to enhancer-spanning fibers. This is an empirical reference,
not a random-binding null. The final length bin pools all footprints >=500 bp.
The alternate `inactive` background uses inactive-enhancer footprints at the same
timepoint. Zero denominators are shown as missing, never as zero enrichment.

```{r footprint-length-distributions}
length_grid <- CJ(sample = sample_table$sample,
                  enhancer_class = c("active", "inactive"),
                  length_bin = seq.int(0L, 500L, 10L))
observed_lengths <- merge(length_grid, length_counts,
                          by = c("sample", "enhancer_class", "length_bin"), all.x = TRUE)
observed_lengths[is.na(n), n := 0]
observed_lengths[, time_min := sample_table$time_min[match(sample, sample_table$sample)]]
observed_lengths[, observed_frequency := if (sum(n) > 0) n / sum(n) else NA_real_,
                  by = .(sample, enhancer_class)]
expected_length_counts[, expected_frequency := if (sum(n) > 0) n / sum(n) else NA_real_,
                       by = sample]
length_comparison <- merge(observed_lengths,
  expected_length_counts[, .(sample, length_bin, expected_n = n, expected_frequency)],
  by = c("sample", "length_bin"), all.x = TRUE)
save_table(length_comparison[, .(sample, enhancer_class, length_bin, observed_n = n,
                                 expected_n, observed_frequency, expected_frequency)],
           "02_footprint_length_distribution.tsv")
length_plot_data <- melt(length_comparison,
  id.vars = c("time_min", "enhancer_class", "length_bin"),
  measure.vars = c("observed_frequency", "expected_frequency"),
  variable.name = "distribution", value.name = "frequency")
head(length_plot_data)
p_length <- ggplot(length_plot_data, aes(length_bin, frequency, color = distribution)) +
  geom_line(na.rm = TRUE) + facet_grid(enhancer_class ~ time_min) + theme_classic() +
  labs(x = "Footprint length (10-bp bin start; 500 = >=500 bp)",
       y = "Normalized frequency", color = NULL,
       subtitle = paste("Expected background:", expected_background))
save_pdf(p_length, "02_footprint_length_observed_expected.pdf", width = 12, height = 6)
```

```{r tf-nucleosome-enrichment-summary}
footprint_comparison <- footprint_counts[, .(n_tf = sum(n_tf), n_nuc = sum(n_nuc),
                                            n_reads = sum(n_reads)),
                                         by = .(sample, time_min, enhancer_class)]
# Bins <60 are TF; bins >=90 contain only calls strictly >90 from the loader.
expected_categories <- expected_length_counts[, .(
  expected_tf = sum(n[length_bin < 60L]), expected_nuc = sum(n[length_bin >= 90L])), by = sample]
footprint_comparison <- merge(footprint_comparison, expected_categories, by = "sample")
footprint_comparison[, tf_fraction := fifelse(n_tf + n_nuc > 0, n_tf / (n_tf + n_nuc), NA_real_)]
footprint_comparison[, tf_to_nuc := fifelse(n_nuc > 0, n_tf / n_nuc, NA_real_)]
footprint_comparison[, expected_tf_to_nuc := fifelse(expected_nuc > 0,
                                                    expected_tf / expected_nuc, NA_real_)]
footprint_comparison[, enrichment := fifelse(expected_tf_to_nuc > 0,
                                             tf_to_nuc / expected_tf_to_nuc, NA_real_)]
save_table(footprint_comparison[, .(sample, enhancer_class, n_tf, n_nuc,
                                   tf_fraction, tf_to_nuc, expected_tf_to_nuc, enrichment)],
           "02_TF_to_nucleosome_enrichment.tsv")
p_enrichment <- ggplot(footprint_comparison,
  aes(time_min, enrichment, color = enhancer_class, group = enhancer_class)) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey60") +
  geom_line(na.rm = TRUE) + geom_point(na.rm = TRUE) + theme_classic() +
  scale_x_continuous(breaks = sample_table$time_min) +
  labs(x = "LPS (min)", y = "TF:nucleosome ratio / expected ratio", color = NULL)
save_pdf(p_enrichment, "02_TF_to_nucleosome_enrichment.pdf")
```

# 3. TF co-occupancy enrichment at active enhancers

For each timepoint, the Fisher table contains active/inactive rows and
co-occupied/not-co-occupied columns. A co-occupied observation is a full-span
fiber with at least two short TF footprints assigned to the enhancer. Enhancers
must have at least `minimum_reads` full-span fibers **at every timepoint** to
enter this comparison, keeping the same enhancer set across the time course.
All enhancer/timepoint counts, including lower coverage and zero calls, remain
in the count table. Missing coverage is not evidence of no co-occupancy.

The R test is one-sided (`alternative = "greater"`) for enrichment at active
enhancers, uses integer counts without pseudocounts, and applies BH correction
across the four timepoint tests. The odds ratio is Fisher's conditional estimate.
The method follows the count-table approach in `code/fire_frequency`, implemented
here in R. Observations are enhancer–fiber pairs; a fiber can cover multiple
enhancers. P-values are therefore nominal molecule-level comparisons, not
replicate-level inference. Enhancer length and locus composition can also affect
these pooled comparisons; per-enhancer frequencies are plotted alongside them.

```{r fisher-cooccupancy-function}
fisher_cooccupancy <- function(x) {
  active <- x[enhancer_class == "active"]
  inactive <- x[enhancer_class == "inactive"]
  a <- sum(active$n_cooccupied)
  b <- sum(active$n_reads - active$n_cooccupied)
  c <- sum(inactive$n_cooccupied)
  d <- sum(inactive$n_reads - inactive$n_cooccupied)
  table <- matrix(c(a, b, c, d), nrow = 2L, byrow = TRUE,
                   dimnames = list(c("active", "inactive"), c("cooccupied", "not_cooccupied")))
  result <- data.table(active_co = a, active_other = b, inactive_co = c, inactive_other = d,
                       odds_ratio = NA_real_, p_value = NA_real_, status = "not_testable")
  if (all(rowSums(table) > 0) && all(colSums(table) > 0)) {
    fit <- fisher.test(table, alternative = "greater")
    result[, `:=`(odds_ratio = unname(fit$estimate), p_value = fit$p.value, status = "tested")]
  }
  result
}
```

```{r cooccupancy-fisher-tests}
eligible_enhancers <- footprint_counts[, .(
  keep = .N == nrow(sample_table) && all(n_reads >= minimum_reads)), by = enhancer_id][keep == TRUE, enhancer_id]
coocc_counts <- copy(footprint_counts[enhancer_id %in% eligible_enhancers])
coocc_counts[, cooccupied_fraction := n_cooccupied / n_reads]
head(coocc_counts)
coocc_tests <- rbindlist(lapply(sample_table$sample, function(s) {
  result <- fisher_cooccupancy(coocc_counts[sample == s])
  result[, sample := s]
  result
}))
coocc_tests[, q_value := p.adjust(p_value, method = "BH")]
save_table(coocc_tests[, .(sample, active_co, active_other, inactive_co, inactive_other,
                          odds_ratio, p_value, q_value, status)], "03_cooccupancy_fisher.tsv")
coocc_frequency <- coocc_counts[, .(n_enhancers = .N, n_reads = sum(n_reads),
                                    n_cooccupied = sum(n_cooccupied)),
                                 by = .(sample, time_min, enhancer_class)]
coocc_frequency[, frequency := n_cooccupied / n_reads]
save_table(coocc_frequency, "03_cooccupancy_frequency.tsv")
```

```{r cooccupancy-plots}
if (nrow(coocc_frequency)) {
  p_coocc <- ggplot(coocc_frequency, aes(time_min, frequency,
                                        color = enhancer_class, group = enhancer_class)) +
    geom_line() + geom_point() + scale_x_continuous(breaks = sample_table$time_min) +
    scale_y_continuous(limits = c(0, 1)) + theme_classic() +
    labs(x = "LPS (min)", y = "Fraction of fibers with >=2 TF footprints", color = NULL)
  save_pdf(p_coocc, "03_cooccupancy_timecourse.pdf")
  p_locus <- ggplot(coocc_counts,
                    aes(factor(time_min, levels = sample_table$time_min), cooccupied_fraction,
                        fill = enhancer_class)) +
    geom_boxplot(outlier.shape = NA) + theme_classic() +
    labs(x = "LPS (min)", y = "Per-enhancer co-occupied fiber fraction", fill = NULL)
  save_pdf(p_locus, "03_cooccupancy_per_enhancer.pdf")
}
```

### Example fibers at an active enhancer

The example is the active enhancer with the largest total number of co-occupied
fibers across time points. The panels reuse the BED12 m6A reader from
`coaccess_plot_functions.R`; the display is adapted to a single enhancer. Gray
lines are spanning fibers, blue marks are m6A, and magenta segments are complete
FiberHMM TF calls shorter than 60 bp. Fibers are ordered by their number of TF
footprints with midpoints inside the enhancer and capped at 100 per time point for display.
This selected example is descriptive; all eligible fibers enter the tests above.

```{r source-read-plot-helper}
read_plot_helpers <- new.env(parent = globalenv())
sys.source(file.path(project_root, "code/co-accessibility/coaccess_plot_functions.R"),
           envir = read_plot_helpers)
```

```{r function-load-enhancer-example}
load_enhancer_example <- function(enhancer, samples, max_reads = 100L) {
  chromosome <- enhancer$chr[[1]]
  lo <- enhancer$start0[[1]]
  hi <- enhancer$end0[[1]]
  window <- c(max(0L, lo - 100L), hi + 100L)
  query <- GenomicRanges::GRanges(chromosome,
    IRanges::IRanges(window[1] + 1L, window[2]))
  spans <- m6a <- footprints <- vector("list", nrow(samples))
  for (i in seq_len(nrow(samples))) {
    sample_name <- samples$sample[i]
    short_name <- gsub("_", "", sample_name, fixed = TRUE)
    methylation_path <- file.path(ft_result_dir, sample_name, "extracted_results",
      "m6a_by_chr", sprintf("%s.ft_extracted_m6a.%s.bed.gz", sample_name, chromosome))
    raw <- tabix_bed(methylation_path, query)
    if (!nrow(raw)) next
    reads <- unique(raw[, .(read_id = as.character(V4),
      start0 = as.numeric(V2), end0 = as.numeric(V3))])
    reads[, span_width := end0 - start0]
    data.table::setorder(reads, read_id, -span_width, start0, end0)
    reads <- reads[!duplicated(read_id)][start0 <= lo & end0 >= hi]
    if (!nrow(reads)) next
    tf_path <- file.path(extract_root, "firehmm_tf", short_name,
      sprintf("%s_hmm_extracted_tf_%s.bed.gz", short_name, chromosome))
    tf <- expand_hmm_blocks(tabix_bed(tf_path, query))
    tf <- unique(tf[length < 60L & length > 0L & read_id %in% reads$read_id &
      start0 < window[2] & end0 > window[1]])
    counts <- tf[(start0 + end0) / 2 >= lo & (start0 + end0) / 2 < hi,
      .(n_tf = .N), by = read_id]
    reads <- merge(reads, counts, by = "read_id", all.x = TRUE)
    reads[is.na(n_tf), n_tf := 0L]
    data.table::setorder(reads, -n_tf, read_id)
    reads <- head(reads, max_reads)
    reads[, `:=`(sample = sample_name, row = rev(seq_len(.N)))]
    reads[, key := paste(sample, read_id)]
    m <- read_plot_helpers$read_bed12_blocks(methylation_path, chromosome,
      window[1], window[2], sample_name, keys = reads$key, sentinels = TRUE)
    if (nrow(m)) {
      m <- merge(m[, .(read_id = RID, start0 = start, end0 = end)],
        reads[, .(read_id, sample, row)], by = "read_id")
    }
    tf <- merge(tf, reads[, .(read_id, sample, row)], by = "read_id")
    spans[[i]] <- reads
    m6a[[i]] <- m
    footprints[[i]] <- tf
  }
  sp <- data.table::rbindlist(spans, fill = TRUE)
  if (!nrow(sp)) stop("No spanning fibers found for the example enhancer")
  list(spans = sp, m6a = data.table::rbindlist(m6a, fill = TRUE),
    tf = data.table::rbindlist(footprints, fill = TRUE),
    samples = samples, window = window, enhancer = enhancer)
}
```

```{r function-plot-enhancer-example}
plot_enhancer_example <- function(example) {
  sp <- data.table::copy(example$spans)
  m <- data.table::copy(example$m6a)
  tf <- data.table::copy(example$tf)
  samples <- example$samples$sample
  sp[, sample := factor(sample, levels = samples)]
  if (nrow(m)) m[, sample := factor(sample, levels = samples)]
  if (nrow(tf)) tf[, sample := factor(sample, levels = samples)]
  labels <- setNames(paste0(example$samples$time_min, " min"), samples)
  base <- ggplot2::ggplot(sp) +
    ggplot2::annotate("rect", xmin = example$enhancer$start0,
      xmax = example$enhancer$end0, ymin = -Inf, ymax = Inf,
      fill = "#44d6d8", alpha = 0.2) +
    ggplot2::geom_segment(ggplot2::aes(x = start0, xend = end0,
      y = row, yend = row), colour = "grey70", linewidth = 0.5) +
    ggplot2::facet_grid(sample ~ ., scales = "free_y", drop = FALSE,
      labeller = ggplot2::labeller(sample = labels)) +
    ggplot2::coord_cartesian(xlim = example$window) +
    ggplot2::theme_bw(base_size = 10) +
    ggplot2::theme(axis.text.y = ggplot2::element_blank(),
      axis.ticks.y = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank()) +
    ggplot2::labs(x = example$enhancer$chr, y = "Spanning fibers")
  p_m6a <- base + ggplot2::labs(title = "m6A")
  if (nrow(m)) p_m6a <- p_m6a + ggplot2::geom_segment(data = m,
    ggplot2::aes(x = start0, xend = end0, y = row, yend = row),
    colour = "blue", linewidth = 0.7)
  p_tf <- base + ggplot2::labs(title = "TF footprints (<60 bp)")
  if (nrow(tf)) p_tf <- p_tf + ggplot2::geom_segment(data = tf,
    ggplot2::aes(x = start0, xend = end0, y = row, yend = row),
    colour = "magenta", linewidth = 1)
  patchwork::wrap_plots(p_m6a, p_tf, ncol = 2) +
    patchwork::plot_annotation(title = example$enhancer$enhancer_id,
      subtitle = "Enhancer shaded; at most 100 fibers/time point, ordered by TF count")
}
```

```{r plot-example-enhancer, fig.width=12, fig.height=10}
example_rank <- footprint_counts[enhancer_class == "active",
  .(n_cooccupied = sum(n_cooccupied)), by = enhancer_id]
data.table::setorder(example_rank, -n_cooccupied, enhancer_id)
if (nrow(example_rank) && example_rank$n_cooccupied[1] > 0L) {
  example_enhancer <- analysis_enhancers[enhancer_id == example_rank$enhancer_id[1]]
  print(example_enhancer[, .(enhancer_id, chr, start0, end0)])
  example_fibers <- load_enhancer_example(example_enhancer, sample_table)
  example_plot <- plot_enhancer_example(example_fibers)
  save_pdf(example_plot, "active_enhancer_example_fibers.pdf", width = 12, height = 10)
}
```
