---
title: "Hamming + hierarchical clustering of single molecules at early-response promoters"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

Replicates the single-molecule cluster-based phasing of Bellini, Do & Skok
(2025), "Phasing single-molecule nano-NOMe-seq reveals chromatin state
heterogeneity in the context of transcription and long-range interactions"
(bioRxiv 2025.09.08.674887), on THP-1 Fiber-seq m6A instead of nano-NOMe GpC
methylation. The procedure parameters are the paper's: 50-bp bins; reads
< 10 bp or with < 10 methylated sites removed; fully methylated /
unmethylated reads pre-assigned; sliding windows of 60 valid bins with
>= 20-bin overlap; >= 30 valid bins per read per window; masked Hamming
distance; hierarchical clustering to t clusters per window (the paper's
average linkage replaced by Ward - see the deviations below); global
reassignment of every read to its closest cluster profile. The paper's
Hungarian stitching of adjacent windows is commented out in
`phasing_functions.r` - every run here is a single window, whose cluster
means serve as the profiles directly. See `phasing_functions.r` for the
step-by-step implementation.

Deviations from the paper (switches in the Settings chunk):

* Data type: Fiber-seq m6A (1 = m6A / accessible, 0 = covered but
  unmodified, NA = not covered) instead of GpC methylation.
* Bin-size sweep (`bin_sizes`): each promoter is phased once per entry -
  a bin size in bp (the paper: 50), or 0 for no binning, i.e. clustering
  directly on the 0/1/NA m6A site matrix. The paper's window geometry
  (60 valid bins, >= 20 overlapping bins, >= 30 valid bins per read, all at
  50-bp bins) is held fixed in genomic bp (3000 / 1000 / 1500 bp) and
  converted to bins per size, so bin size 50 reproduces the paper's numbers
  exactly and the other sizes scale to the same genomic spans; the bin-0
  (no-bin) run uses one window over all sites and >= 30 valid sites per
  read. Outputs go to `<gene>/bin<size>/k<t>/` (t = the number of clusters
  used for that run), each split into `results/` and `plots/`.
* Binary bins (`binarize_bins`, on): a read's bin value is 1 if the
  read has any m6A call in the bin, instead of the mean site call.
  Fractional m6A bins are unimodal near zero, which compresses all masked
  Hamming distances and collapses the clustering into one cluster plus
  outliers; binary bins are bimodal like the paper's GpC data. (No effect
  on the no-bin run, whose site calls are already 0/1.)
* Full-span reads only at the promoters (`full_span_only`, on): the topic
  model's filter_met_mat row filter, so both analyses cluster the same read
  set. Not applied at the wide locus (see below).
* Ward linkage (`hclust` "ward.D2") instead of the paper's average linkage.
  On this data average linkage peels off outlier reads, so any cut of the
  tree gave one cluster holding ~90% of the reads plus tiny shards (worst
  at fine bin sizes); Ward favors compact, more balanced clusters. Both the
  t-selection tree and the per-window clustering use Ward, so they cut the
  same tree.
* t: the paper predefines t but does not state it for its Pol II analysis
  (its Sox2 locus analyses used 9 and 7). With `choose_t_by_dispersion`
  (on), t is selected per gene and bin setting from the within-cluster
  Hamming dispersion curve (`choose_t_dispersion()`: for every cut
  t = 1..10 of the Ward tree, the size-weighted average
  within-cluster masked Hamming distance; t = the elbow of this monotone
  non-increasing curve, i.e. the candidate maximizing its second
  difference). This replaces two earlier criteria kept in
  `phasing_functions.r` for reference: average silhouette width (monotone
  decreasing in t at every one of these loci, so always t = 2) and
  recursive bootstrap stability (no split was stable, so always t = 1).
  Switched off, the manually set `t_clusters` (currently 6) is used
  everywhere.

The paper's Pol II analysis clusters single molecules in 2-kb regions centered
on ChIP-seq-validated, Pol II-bound TSSs and annotates the clusters as Pol II
states (elongating / active with +1 nucleosome / paused / inactive), ordering
by a Pol II pausing index. Here the 2-kb regions are the early-response
promoter windows (TSS +/- 1000 bp, dominant THP-1 CAGE peak or GENCODE v49
TSS), and the Pol II validation/ordering uses THP-1 RNAPII (phospho-S5)
ChIP-seq (GSE32324; control and 2 h LPS, 2 reps each) lifted hg19 -> hg38.
rs2836882 is excluded: its 244-bp interval has at most 5 bins, so no read can
ever reach the paper's >= 30-valid-bin threshold.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(rtracklayer)
library(data.table)
library(Rsamtools)
library(cowplot)
library(gplots)
library(clue)
```

# Functions

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/phasing_functions.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Number of clusters for the 2-kb TSS regions.
# choose_t_by_dispersion = TRUE: t is selected per gene and bin setting from
# the within-cluster Hamming dispersion curve (choose_t_dispersion() inside
# phase_reads(): for every cut t = 1..10 of the Ward tree, the
# size-weighted average within-cluster masked Hamming distance, read-weight-
# averaged across windows; the curve is monotone non-increasing, so t is the
# elbow - the candidate maximizing the second difference W(t-1)-2W(t)+W(t+1)).
# Replaces two earlier criteria kept in phasing_functions.r for reference:
# silhouette (choose_t(), monotone decreasing at these loci -> always t = 2)
# and recursive bootstrap stability (choose_t_stability(), no stable split
# -> always t = 1). FALSE: the manually set t_clusters below is used
# everywhere (the paper's predefined-t behavior; its Sox2 locus analyses
# used 9 and 7, its Pol II t is unstated).
choose_t_by_dispersion <- TRUE
t_clusters <- 6   # used only when choose_t_by_dispersion = FALSE

# Optional deviations from the paper, both on for now:
# - full_span_only: keep only reads covering every m6A site in the region
#   (filter_met_mat's row filter - the topic model's read set). Applied to
#   the 2-kb promoters only; at the 20-kb wide locus it would discard nearly
#   every read and defeat the stitching demo.
# - binarize_bins: a read's 50-bp bin value is 1 if the read has any m6A call
#   in the bin (instead of the mean site call), making bins bimodal like the
#   paper's GpC data.
# - use_polII_chip: quantify the GSE32324 RNAPII ChIP and order genes by
#   pausing index, as the paper does. Off for now; with it off the ChIP table
#   is never read and no PI ordering is applied.
full_span_only <- TRUE
binarize_bins  <- TRUE
use_polII_chip <- FALSE

# Bin-size sweep: one full phasing run per promoter per entry. Positive =
# bin size in bp; 0 = no binning (cluster on the m6A site matrix directly).
# Output folders: <gene>/bin<size>/k<t>/ (t = the number of clusters used
# for that run), each with results/ and plots/ inside.
bin_sizes <- c(0, 10, 20, 30, 50)

# The paper's sliding-window geometry (60 valid bins, >= 20 overlapping bins,
# >= 30 valid bins per read) is defined at 50-bp bins; keep it fixed in
# genomic bp and convert per bin size, so bin size 50 gives exactly 60/20/30
# and every other size scales to the same genomic spans. For the bin-0
# (no-bin) run the thresholds count m6A site columns instead: a single window
# over all sites, >= 30 valid sites per read (trivial for full-span reads).
window_bp    <- 3000   # 60 bins x 50 bp
overlap_bp   <- 1000   # 20 bins x 50 bp
min_valid_bp <- 1500   # 30 bins x 50 bp

bin_label <- function(b) paste0("bin", b)
phase_params_for_bin <- function(b) {
  if (b == 0) {
    list(window_valid_bins = Inf, min_overlap_bins = 20,
         min_valid_bins_per_read = 30)
  } else {
    list(window_valid_bins       = ceiling(window_bp / b),
         min_overlap_bins        = ceiling(overlap_bp / b),
         min_valid_bins_per_read = ceiling(min_valid_bp / b))
  }
}

set.seed(123)
```

# Promoter regions

TSS +/- 1000 bp (2 kb, the paper's Pol II-analysis geometry). The table is
built from the GENCODE v49 annotation by `make_promoter_table.R`: TSS =
strand-aware gene 5' end on the primary chromosomes (IL8 = CXCL8 in GENCODE),
plus the gene span used for the ChIP pausing-index gene bodies.

```{bash make-regions, eval=FALSE}
/software/R-4.4.1-el8-x86_64/bin/Rscript \
  /project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/make_promoter_table.R
```

```{r regions}
promoters <- read.delim(file.path(output_dir, "promoter_regions_gencode_v49.tsv"))
promoters
```

# PolII ChIP-seq: lifted read counts and pausing index

`quantify_polII_chip.R` lifts the GSE32324 hg19 read coordinates to hg38
(conservative liftOver, as in `liftover_tracks.R`) and counts overlaps with
the promoter (TSS +/- 1 kb), pause (TSS -50..+300, strand-aware) and gene-body
(TSS +300..TES) windows; pausing index PI = pause density / body density.

```{bash chip-quantify, eval=FALSE}
<!-- /software/R-4.4.1-el8-x86_64/bin/Rscript \ -->
<!--   /project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/quantify_polII_chip.R -->


<!-- sbatch /project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/run_quantify_polII_chip.sh -->

```

Genes are ordered by LPS PI, mirroring the paper's PI-based
ordering/annotation of the Pol II heatmaps. `use_polII_chip`
(currently `FALSE`), so by default this section is skipped 

```{r chip-pausing-index}
# chip_counts_file <- file.path(output_dir, "PolII_chip_tss_counts.tsv")
# 
# if (!use_polII_chip) {
#   cat("use_polII_chip = FALSE; skipping the PolII ChIP table and the PI ordering.\n")
#   chip_table <- NULL
# } else if (!file.exists(chip_counts_file)) {
#   cat("PolII_chip_tss_counts.tsv missing - run quantify_polII_chip.R first; skipping ChIP table.\n")
#   chip_table <- NULL
# } else {
#   chip_counts <- read.delim(chip_counts_file)
#   chip_table <- chip_counts %>%
#     dplyr::mutate(condition = sub("_rep[12]$", "", sample)) %>%
#     dplyr::group_by(gene, condition) %>%
#     dplyr::summarise(promoter_rpm_per_kb = mean(promoter_rpm_per_kb),
#                      PI = mean(PI), .groups = "drop") %>%
#     tidyr::pivot_wider(names_from = condition,
#                        values_from = c(promoter_rpm_per_kb, PI)) %>%
#     dplyr::arrange(dplyr::desc(PI_LPS))
# 
#   write.table(chip_table, file.path(output_dir, "PolII_chip_pausing_index.tsv"),
#               sep = "\t", quote = FALSE, row.names = FALSE)
#   chip_table
# }
```

# m6A read x position matrix per promoter

Pooled across the four LPS timepoints, tagging each read with its sample of
origin (as in the topic-model notebook); m6A only. `get_sparse_met_mat()`
gives 1 = m6A call, 0 = covered/unmodified, NA = not covered. The full-span
filter and the bin binarization happen downstream inside `phase_reads()`, so
the site-level matrix assembled here keeps every read and the raw calls.

```{r assemble-function}
assemble_region_m6a <- function(sample_names, region_chr, region_start, region_end,
                                ft_result_dir, verbose = TRUE) {
  region_gr <- GRanges(region_chr, IRanges(region_start, region_end))
  reads_list <- lapply(sample_names, function(sample_name) {
    extracted_file <- file.path(
      ft_result_dir, sample_name, "extracted_results", "m6a_by_chr",
      paste0(sample_name, ".ft_extracted_m6a.", region_chr, ".bed.gz"))
    sample_reads <- extract_ft_region_reads(extracted_file, region_gr,
                                            keep_pos_in_region_only = TRUE,
                                            verbose = verbose)
    if (nrow(sample_reads) == 0) return(NULL)
    dplyr::mutate(sample_reads, sample_name = sample_name, .before = 1)
  })
  reads <- dplyr::bind_rows(reads_list)
  if (nrow(reads) == 0) stop("no reads in region")
  rids_df <- extract_ft_read_info(reads, keep_columns = c("sample_name", "score"))
  met_mat <- get_sparse_met_mat(reads, rids_df,
                                window_start = region_start,
                                window_end = region_end, base = "A")
  list(reads = reads, rids_df = rids_df, met_mat = met_mat)
}
```

# Phasing at each promoter

Per gene and per `bin_sizes` entry: `phase_reads()` with t selected at the
elbow of the within-cluster Hamming dispersion curve when
`choose_t_by_dispersion` is on (`choose_t_dispersion()`; the printed
"t = ... selected at the within-cluster dispersion elbow" line records the
choice) or the manually set `t_clusters` when it is off, the
full-span read filter (reads not covering every m6A site in the 2-kb
window are dropped, matching the topic model's read set) and binary bins
of the given size - or the raw m6A site matrix for the no-bin run. The window
geometry scales with bin size (see Settings) and its 3000 bp always covers
the whole 2-kb promoter, so clustering here is one Ward cut with
no stitching at every bin size; the no-bin run is a single window over all
sites by construction. The read x position matrix is assembled once per gene
and reused across bin sizes. Outputs per gene x bin setting, under
`<gene>/bin<size>/k<t>/` (t = the number of clusters used for that run):
`results/` holds the phasing
object (.rds), per-read cluster assignments, global cluster profiles, site
coverage and - when `choose_t_by_dispersion` is on - the dispersion curve
behind the t choice (one row per candidate t: the size-weighted average
within-cluster Hamming distance, with the chosen elbow flagged) (.tsv);
`plots/` holds the bin
heatmap (skipped for the no-bin
run, where it would just duplicate the site heatmap), m6A-site heatmap,
cluster profiles, cluster composition by timepoint, and per-cluster m6A
methylation proportion at each site (`cluster_met_profiles`, the topic
model's plot of the same name, so the two clusterings can be compared panel
for panel). Heatmap rows are grouped by final cluster (side colors) and
ordered by read start within cluster.

```{r phase-promoters, fig.width=8, fig.height=8}
phase_results <- list()
for (i in seq_len(nrow(promoters))) {
  gene <- promoters$gene[i]
  cat("\n=====", gene, "=====\n")

  dat <- assemble_region_m6a(sample_names, promoters$chr[i],
                             promoters$start[i], promoters$end[i],
                             ft_result_dir, verbose = FALSE)
  cat(gene, ": ", nrow(dat$rids_df), " reads x ", ncol(dat$met_mat), " m6A sites\n", sep = "")

  for (bs in bin_sizes) {
    lab      <- bin_label(bs)
    bin_desc <- if (bs == 0) "m6A sites, no bins" else paste0(bs, "-bp bins")
    cat("--- ", gene, ", ", bin_desc, " ---\n", sep = "")

    pars <- phase_params_for_bin(bs)
    res <- phase_reads(dat$met_mat, dat$rids_df,
                       t = if (choose_t_by_dispersion) NULL else t_clusters,
                       bin_size = if (bs == 0) NA else bs,
                       window_valid_bins = pars$window_valid_bins,
                       min_overlap_bins = pars$min_overlap_bins,
                       min_valid_bins_per_read = pars$min_valid_bins_per_read,
                       full_span_only = full_span_only,
                       binarize_bins = binarize_bins)
    res$assignments$sample_name <- factor(res$assignments$sample_name,
                                          levels = sample_names)
    phase_results[[gene]][[lab]] <- res

    # k subfolder = the number of clusters used for this run (silhouette-
    # selected or manual), so runs ending at different t sit side by side
    res_dir  <- file.path(output_dir, gene, lab, paste0("k", res$params$t), "results")
    plot_dir <- file.path(output_dir, gene, lab, paste0("k", res$params$t), "plots")
    for (d in c(res_dir, plot_dir))
      if (!dir.exists(d)) dir.create(d, recursive = TRUE)

    saveRDS(res, file.path(res_dir, paste0("phasing_", gene, ".rds")))
    write.table(res$assignments,
                file.path(res_dir, paste0("read_cluster_assignments_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(cluster = rownames(res$profiles),
                           res$profiles, check.names = FALSE),
                file.path(res_dir, paste0("global_cluster_profiles_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$site_cov,
                file.path(res_dir, paste0("site_coverage_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    if (!is.null(res$t_selection)) {
      write.table(data.frame(t = res$t_selection$t_candidates,
                             within_dispersion = res$t_selection$score,
                             chosen = res$t_selection$t_candidates == res$t_selection$t),
                  file.path(res_dir, paste0("dispersion_t_selection_", gene, ".tsv")),
                  sep = "\t", quote = FALSE, row.names = FALSE)
    }

    if (bs != 0) {   # for the bin-0 (no-bin) run this would duplicate the site heatmap
      pdf(file.path(plot_dir, paste0("bin_heatmap_", gene, ".pdf")), width = 8, height = 8)
      plot_phasing_bin_heatmap(res, zmax = if (binarize_bins) 1 else 0.5,
                               main = paste0(gene, " (", bin_desc, ")"))
      dev.off()
    }

    pdf(file.path(plot_dir, paste0("site_heatmap_", gene, ".pdf")), width = 8, height = 8)
    plot_phasing_site_heatmap(res, dat$met_mat, main = paste0(gene, " (m6A sites)"))
    dev.off()

    gg <- plot_global_profiles(res, tss = promoters$tss[i],
                               main = paste0(gene, " (", bin_desc, ")"))
    ggsave(file.path(plot_dir, paste0("global_profiles_", gene, ".pdf")), gg,
           width = 7, height = 1.4 * nrow(res$profiles) + 1)

    gg <- plot_cluster_composition(res, main = paste0(gene, " (", bin_desc, ")"))
    ggsave(file.path(plot_dir, paste0("cluster_composition_", gene, ".pdf")), gg,
           width = 6, height = 4)

    gg <- plot_cluster_met_profiles(res, dat$met_mat, tss = promoters$tss[i],
                                    main = paste0(gene, " (", bin_desc, ")"))
    n_panels <- length(unique(res$assignments$final_cluster[
      grepl("^cluster|^fully", res$assignments$final_cluster)]))
    ggsave(file.path(plot_dir, paste0("cluster_met_profiles_", gene, ".pdf")), gg,
           width = 6, height = 1.4 * n_panels + 1)
  }
}
```

# Cluster sizes across genes

Reads per final cluster, gene and bin setting (`t_selected` is the t used
for that gene x bin setting - dispersion-elbow-chosen when
`choose_t_by_dispersion` is on, else the manual `t_clusters`; `removed_*`
columns count reads that failed the informative-read filters; reads dropped
by the full-span filter are not in the table, as in the topic model).

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(phase_results), function(g) {
  dplyr::bind_rows(lapply(names(phase_results[[g]]), function(lab) {
    df <- phase_results[[g]][[lab]]$assignments
    df %>% dplyr::count(final_cluster) %>%
      dplyr::mutate(gene = g, bins = lab,
                    t_selected = phase_results[[g]][[lab]]$params$t,
                    .before = 1)
  }))
}))
write.table(cluster_summary, file.path(output_dir, "cluster_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = final_cluster, values_from = n)
```

<!-- Wide-locus demo commented out with the stitching (stitch_windows() /
global_cluster_profiles() in phasing_functions.r are commented out, and
phase_reads() now stops on multi-window input); re-enable both together.

# Wide-locus demo: IL1B TSS +/- 10 kb

A 2-kb promoter gives at most ~40 valid bins, i.e. a single window, so the
sliding-window + Hungarian machinery degenerates. This 20-kb locus (~400
bins, 10 overlapping windows) exercises the full stitching procedure.
Unlike the promoter analysis it keeps partial reads - no full-span filter,
since partial reads are what the stitching machinery is for, and requiring
~10-kb HiFi reads to span 20 kb would discard nearly all of them - and t is
selected at the elbow of the within-cluster Hamming dispersion curve
(`choose_t_dispersion()`; per-window curves are read-weight-averaged because
the Hungarian stitching needs one t in every window). Binary bins are used
as in the promoter analysis. The chunk also runs the paper's held-out
long-read validation (>= 50 kb reads; with HiFi Fiber-seq read lengths the
expected report is "not applicable").
-->


```{r wide-demo, fig.width=10, fig.height=8}
# wide_i <- match("IL1B", promoters$gene)
# wide <- list(gene = "IL1B", chr = promoters$chr[wide_i],
#              start = promoters$tss[wide_i] - 10000,
#              end   = promoters$tss[wide_i] + 10000,
#              tss   = promoters$tss[wide_i])
# wide_dir <- file.path(output_dir, "IL1B_wide_20kb")
# if (!dir.exists(wide_dir)) dir.create(wide_dir, recursive = TRUE)
# 
# dat_wide <- assemble_region_m6a(sample_names, wide$chr, wide$start, wide$end,
#                                 ft_result_dir, verbose = FALSE)
# cat("IL1B wide: ", nrow(dat_wide$rids_df), " reads x ",
#     ncol(dat_wide$met_mat), " m6A sites\n", sep = "")
# 
# res_wide <- phase_reads(dat_wide$met_mat, dat_wide$rids_df,
#                         binarize_bins = binarize_bins)
# res_wide$assignments$sample_name <- factor(res_wide$assignments$sample_name,
#                                            levels = sample_names)
# cat("windows clustered:", length(res_wide$win_res), "\n")
# 
# saveRDS(res_wide, file.path(wide_dir, "phasing_IL1B_wide.rds"))
# write.table(res_wide$assignments,
#             file.path(wide_dir, "read_cluster_assignments_IL1B_wide.tsv"),
#             sep = "\t", quote = FALSE, row.names = FALSE)
# 
# pdf(file.path(wide_dir, "bin_heatmap_IL1B_wide.pdf"), width = 10, height = 8)
# plot_phasing_bin_heatmap(res_wide, zmax = if (binarize_bins) 1 else 0.5,
#                          main = "IL1B TSS +/- 10 kb (50-bp bins)")
# dev.off()
# 
# gg <- plot_global_profiles(res_wide, tss = wide$tss, main = "IL1B TSS +/- 10 kb")
# ggsave(file.path(wide_dir, "global_profiles_IL1B_wide.pdf"), gg,
#        width = 10, height = 1.4 * nrow(res_wide$profiles) + 1)
# 
# # held-out long-read validation (paper: reads >= 50 kb)
# holdout <- holdout_long_read_accuracy(dat_wide$met_mat, dat_wide$rids_df,
#                                       binarize_bins = binarize_bins)
# if (!is.null(holdout)) {
#   write.table(holdout, file.path(wide_dir, "holdout_long_read_accuracy.tsv"),
#               sep = "\t", quote = FALSE, row.names = FALSE)
#   cat("mean held-out long-read accuracy:", mean(holdout$accuracy), "\n")
# }
```

# Session info

```{r session-info}
sessionInfo()
```
