---
title: "Hamming + hierarchical clustering of single molecules at early-response promoters"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

Replicates the single-molecule cluster-based phasing of Bellini, Do & Skok
(2025), "Phasing single-molecule nano-NOMe-seq reveals chromatin state
heterogeneity in the context of transcription and long-range interactions"
(bioRxiv 2025.09.08.674887), on THP-1 Fiber-seq m6A instead of nano-NOMe GpC
methylation. The procedure parameters are the paper's: 50-bp bins; reads
< 10 bp or with < 10 methylated sites removed; fully methylated /
unmethylated reads pre-assigned; sliding windows of 60 valid bins with
>= 20-bin overlap; >= 30 valid bins per read per window; masked Hamming
distance; average-linkage clustering to t clusters per window; Hungarian
stitching of adjacent windows; global reassignment of every read to its
closest stitched profile. See `phasing_functions.r` for the step-by-step
implementation.

Deviations from the paper (switches in the Settings chunk):

* Data type: Fiber-seq m6A (1 = m6A / accessible, 0 = covered but
  unmodified, NA = not covered) instead of GpC methylation.
* Binary 50-bp bins (`binarize_bins`, on): a read's bin value is 1 if the
  read has any m6A call in the bin, instead of the mean site call.
  Fractional m6A bins are unimodal near zero, which compresses all masked
  Hamming distances and collapses the clustering into one cluster plus
  outliers; binary bins are bimodal like the paper's GpC data.
* Full-span reads only at the promoters (`full_span_only`, on): the topic
  model's filter_met_mat row filter, so both analyses cluster the same read
  set. Not applied at the wide locus (see below).
* t: the paper predefines t but does not state it for its Pol II analysis
  (its Sox2 locus analyses used 9 and 7). The promoters use a predefined
  t = 6; the wide-locus demo instead selects t by average silhouette width
  (`choose_t()`).

The paper's Pol II analysis clusters single molecules in 2-kb regions centered
on ChIP-seq-validated, Pol II-bound TSSs and annotates the clusters as Pol II
states (elongating / active with +1 nucleosome / paused / inactive), ordering
by a Pol II pausing index. Here the 2-kb regions are the early-response
promoter windows (TSS +/- 1000 bp, dominant THP-1 CAGE peak or GENCODE v49
TSS), and the Pol II validation/ordering uses THP-1 RNAPII (phospho-S5)
ChIP-seq (GSE32324; control and 2 h LPS, 2 reps each) lifted hg19 -> hg38.
rs2836882 is excluded: its 244-bp interval has at most 5 bins, so no read can
ever reach the paper's >= 30-valid-bin threshold.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(rtracklayer)
library(data.table)
library(Rsamtools)
library(cowplot)
library(gplots)
library(clue)
```

# Functions

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/phasing_functions.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Number of clusters for the 2-kb TSS regions: predefined t = 6 (the paper's
# predefined-t behavior; its Sox2 locus analyses used 9 and 7, its Pol II t is
# unstated). Data-driven selection (choose_t(), used by the wide demo below)
# picks t = 2 at every one of these loci.
t_clusters <- 6

# Optional deviations from the paper, both on for now:
# - full_span_only: keep only reads covering every m6A site in the region
#   (filter_met_mat's row filter - the topic model's read set). Applied to
#   the 2-kb promoters only; at the 20-kb wide locus it would discard nearly
#   every read and defeat the stitching demo.
# - binarize_bins: a read's 50-bp bin value is 1 if the read has any m6A call
#   in the bin (instead of the mean site call), making bins bimodal like the
#   paper's GpC data.
# - use_polII_chip: quantify the GSE32324 RNAPII ChIP and order genes by
#   pausing index, as the paper does. Off for now; with it off the ChIP table
#   is never read and no PI ordering is applied.
full_span_only <- TRUE
binarize_bins  <- TRUE
use_polII_chip <- FALSE

set.seed(123)
```

# Promoter regions

TSS +/- 1000 bp (2 kb, the paper's Pol II-analysis geometry). The table is
built from the GENCODE v49 annotation by `make_promoter_table.R`: TSS =
strand-aware gene 5' end on the primary chromosomes (IL8 = CXCL8 in GENCODE),
plus the gene span used for the ChIP pausing-index gene bodies.

```{bash make-regions, eval=FALSE}
/software/R-4.4.1-el8-x86_64/bin/Rscript \
  /project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/make_promoter_table.R
```

```{r regions}
promoters <- read.delim(file.path(output_dir, "promoter_regions_gencode_v49.tsv"))
promoters
```

# PolII ChIP-seq: lifted read counts and pausing index

`quantify_polII_chip.R` lifts the GSE32324 hg19 read coordinates to hg38
(conservative liftOver, as in `liftover_tracks.R`) and counts overlaps with
the promoter (TSS +/- 1 kb), pause (TSS -50..+300, strand-aware) and gene-body
(TSS +300..TES) windows; pausing index PI = pause density / body density.

```{bash chip-quantify, eval=FALSE}
<!-- /software/R-4.4.1-el8-x86_64/bin/Rscript \ -->
<!--   /project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/quantify_polII_chip.R -->


<!-- sbatch /project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/run_quantify_polII_chip.sh -->

```

Genes are ordered by LPS PI, mirroring the paper's PI-based
ordering/annotation of the Pol II heatmaps. `use_polII_chip`
(currently `FALSE`), so by default this section is skipped 

```{r chip-pausing-index}
# chip_counts_file <- file.path(output_dir, "PolII_chip_tss_counts.tsv")
# 
# if (!use_polII_chip) {
#   cat("use_polII_chip = FALSE; skipping the PolII ChIP table and the PI ordering.\n")
#   chip_table <- NULL
# } else if (!file.exists(chip_counts_file)) {
#   cat("PolII_chip_tss_counts.tsv missing - run quantify_polII_chip.R first; skipping ChIP table.\n")
#   chip_table <- NULL
# } else {
#   chip_counts <- read.delim(chip_counts_file)
#   chip_table <- chip_counts %>%
#     dplyr::mutate(condition = sub("_rep[12]$", "", sample)) %>%
#     dplyr::group_by(gene, condition) %>%
#     dplyr::summarise(promoter_rpm_per_kb = mean(promoter_rpm_per_kb),
#                      PI = mean(PI), .groups = "drop") %>%
#     tidyr::pivot_wider(names_from = condition,
#                        values_from = c(promoter_rpm_per_kb, PI)) %>%
#     dplyr::arrange(dplyr::desc(PI_LPS))
# 
#   write.table(chip_table, file.path(output_dir, "PolII_chip_pausing_index.tsv"),
#               sep = "\t", quote = FALSE, row.names = FALSE)
#   chip_table
# }
```

# m6A read x position matrix per promoter

Pooled across the four LPS timepoints, tagging each read with its sample of
origin (as in the topic-model notebook); m6A only. `get_sparse_met_mat()`
gives 1 = m6A call, 0 = covered/unmodified, NA = not covered. The full-span
filter and the bin binarization happen downstream inside `phase_reads()`, so
the site-level matrix assembled here keeps every read and the raw calls.

```{r assemble-function}
assemble_region_m6a <- function(sample_names, region_chr, region_start, region_end,
                                ft_result_dir, verbose = TRUE) {
  region_gr <- GRanges(region_chr, IRanges(region_start, region_end))
  reads_list <- lapply(sample_names, function(sample_name) {
    extracted_file <- file.path(
      ft_result_dir, sample_name, "extracted_results", "m6a_by_chr",
      paste0(sample_name, ".ft_extracted_m6a.", region_chr, ".bed.gz"))
    sample_reads <- extract_ft_region_reads(extracted_file, region_gr,
                                            keep_pos_in_region_only = TRUE,
                                            verbose = verbose)
    if (nrow(sample_reads) == 0) return(NULL)
    dplyr::mutate(sample_reads, sample_name = sample_name, .before = 1)
  })
  reads <- dplyr::bind_rows(reads_list)
  if (nrow(reads) == 0) stop("no reads in region")
  rids_df <- extract_ft_read_info(reads, keep_columns = c("sample_name", "score"))
  met_mat <- get_sparse_met_mat(reads, rids_df,
                                window_start = region_start,
                                window_end = region_end, base = "A")
  list(reads = reads, rids_df = rids_df, met_mat = met_mat)
}
```

# Phasing at each promoter

Per gene: `phase_reads()` with predefined t = 6, the full-span read filter
(reads not covering every m6A site in the 2-kb window are dropped, matching
the topic model's read set) and binary 50-bp bins. A 2-kb promoter has ~40
valid bins, i.e. a single window, so clustering here is one average-linkage
cut with no stitching. Outputs per gene: the phasing object (.rds), per-read
cluster assignments, global cluster profiles and site coverage (.tsv), and
five plots - 50-bp-bin heatmap, m6A-site heatmap, cluster profiles, cluster
composition by timepoint, and per-cluster m6A methylation proportion at each
site (`cluster_met_profiles`, the topic model's plot of the same name, so the
two clusterings can be compared panel for panel). Heatmap rows are grouped by
final cluster (side colors) and ordered by read start within cluster.

```{r phase-promoters, fig.width=8, fig.height=8}
phase_results <- list()
for (i in seq_len(nrow(promoters))) {
  gene <- promoters$gene[i]
  cat("\n=====", gene, "=====\n")
  gene_dir <- file.path(output_dir, gene)
  if (!dir.exists(gene_dir)) dir.create(gene_dir, recursive = TRUE)

  dat <- assemble_region_m6a(sample_names, promoters$chr[i],
                             promoters$start[i], promoters$end[i],
                             ft_result_dir, verbose = FALSE)
  cat(gene, ": ", nrow(dat$rids_df), " reads x ", ncol(dat$met_mat), " m6A sites\n", sep = "")

  res <- phase_reads(dat$met_mat, dat$rids_df, t = t_clusters,
                     full_span_only = full_span_only,
                     binarize_bins = binarize_bins)
  res$assignments$sample_name <- factor(res$assignments$sample_name,
                                        levels = sample_names)
  phase_results[[gene]] <- res

  saveRDS(res, file.path(gene_dir, paste0("phasing_", gene, ".rds")))
  write.table(res$assignments,
              file.path(gene_dir, paste0("read_cluster_assignments_", gene, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(cluster = rownames(res$profiles),
                         res$profiles, check.names = FALSE),
              file.path(gene_dir, paste0("global_cluster_profiles_", gene, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(res$site_cov,
              file.path(gene_dir, paste0("site_coverage_", gene, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)

  pdf(file.path(gene_dir, paste0("bin_heatmap_", gene, ".pdf")), width = 8, height = 8)
  plot_phasing_bin_heatmap(res, zmax = if (binarize_bins) 1 else 0.5,
                           main = paste0(gene, " (50-bp bins)"))
  dev.off()

  pdf(file.path(gene_dir, paste0("site_heatmap_", gene, ".pdf")), width = 8, height = 8)
  plot_phasing_site_heatmap(res, dat$met_mat, main = paste0(gene, " (m6A sites)"))
  dev.off()

  gg <- plot_global_profiles(res, tss = promoters$tss[i], main = gene)
  ggsave(file.path(gene_dir, paste0("global_profiles_", gene, ".pdf")), gg,
         width = 7, height = 1.4 * nrow(res$profiles) + 1)

  gg <- plot_cluster_composition(res, main = gene)
  ggsave(file.path(gene_dir, paste0("cluster_composition_", gene, ".pdf")), gg,
         width = 6, height = 4)

  gg <- plot_cluster_met_profiles(res, dat$met_mat, tss = promoters$tss[i],
                                  main = gene)
  n_panels <- length(unique(res$assignments$final_cluster[
    grepl("^cluster|^fully", res$assignments$final_cluster)]))
  ggsave(file.path(gene_dir, paste0("cluster_met_profiles_", gene, ".pdf")), gg,
         width = 6, height = 1.4 * n_panels + 1)
}
```

# Cluster sizes across genes

Reads per final cluster and gene (t = 6 everywhere; `removed_*` columns
count reads that failed the informative-read filters; reads dropped by the
full-span filter are not in the table, as in the topic model).

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(phase_results), function(g) {
  df <- phase_results[[g]]$assignments
  df %>% dplyr::count(final_cluster) %>% dplyr::mutate(gene = g, .before = 1)
}))
write.table(cluster_summary, file.path(output_dir, "cluster_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = final_cluster, values_from = n)
```

# Wide-locus demo: IL1B TSS +/- 10 kb

A 2-kb promoter gives at most ~40 valid bins, i.e. a single window, so the
sliding-window + Hungarian machinery degenerates. This 20-kb locus (~400
bins, 10 overlapping windows) exercises the full stitching procedure.
Unlike the promoter analysis it keeps partial reads - no full-span filter,
since partial reads are what the stitching machinery is for, and requiring
~10-kb HiFi reads to span 20 kb would discard nearly all of them - and t is
selected by silhouette (`choose_t()`, currently t = 2). Binary bins are used
as in the promoter analysis. The chunk also runs the paper's held-out
long-read validation (>= 50 kb reads; with HiFi Fiber-seq read lengths the
expected report is "not applicable").

```{r wide-demo, fig.width=10, fig.height=8}
# wide_i <- match("IL1B", promoters$gene)
# wide <- list(gene = "IL1B", chr = promoters$chr[wide_i],
#              start = promoters$tss[wide_i] - 10000,
#              end   = promoters$tss[wide_i] + 10000,
#              tss   = promoters$tss[wide_i])
# wide_dir <- file.path(output_dir, "IL1B_wide_20kb")
# if (!dir.exists(wide_dir)) dir.create(wide_dir, recursive = TRUE)
# 
# dat_wide <- assemble_region_m6a(sample_names, wide$chr, wide$start, wide$end,
#                                 ft_result_dir, verbose = FALSE)
# cat("IL1B wide: ", nrow(dat_wide$rids_df), " reads x ",
#     ncol(dat_wide$met_mat), " m6A sites\n", sep = "")
# 
# res_wide <- phase_reads(dat_wide$met_mat, dat_wide$rids_df,
#                         binarize_bins = binarize_bins)
# res_wide$assignments$sample_name <- factor(res_wide$assignments$sample_name,
#                                            levels = sample_names)
# cat("windows clustered:", length(res_wide$win_res), "\n")
# 
# saveRDS(res_wide, file.path(wide_dir, "phasing_IL1B_wide.rds"))
# write.table(res_wide$assignments,
#             file.path(wide_dir, "read_cluster_assignments_IL1B_wide.tsv"),
#             sep = "\t", quote = FALSE, row.names = FALSE)
# 
# pdf(file.path(wide_dir, "bin_heatmap_IL1B_wide.pdf"), width = 10, height = 8)
# plot_phasing_bin_heatmap(res_wide, zmax = if (binarize_bins) 1 else 0.5,
#                          main = "IL1B TSS +/- 10 kb (50-bp bins)")
# dev.off()
# 
# gg <- plot_global_profiles(res_wide, tss = wide$tss, main = "IL1B TSS +/- 10 kb")
# ggsave(file.path(wide_dir, "global_profiles_IL1B_wide.pdf"), gg,
#        width = 10, height = 1.4 * nrow(res_wide$profiles) + 1)
# 
# # held-out long-read validation (paper: reads >= 50 kb)
# holdout <- holdout_long_read_accuracy(dat_wide$met_mat, dat_wide$rids_df,
#                                       binarize_bins = binarize_bins)
# if (!is.null(holdout)) {
#   write.table(holdout, file.path(wide_dir, "holdout_long_read_accuracy.tsv"),
#               sep = "\t", quote = FALSE, row.names = FALSE)
#   cat("mean held-out long-read accuracy:", mean(holdout$accuracy), "\n")
# }
```

# Session info

```{r session-info}
sessionInfo()
```
