---
title: "Testing topic usage over the LPS time course"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Note:
I previously ran the topic modelling on the following regions:
IL1B promoter region chr2:112836323-112837385
TNF promoter region chr6:31575265-31575824
IL1A upstream intergenic region chr2:112793233-112794969
PTGS2 promoter chr1:186679985-186681163
IL8 promoter chr4:73740006-73741332
CXCL2 promoter chr4:74098863-74099651
IL6 promoter chr7:22726376-22727094 
RELB promoter chr19:45001011-45002417
rs2836882 region chr21:39094468-39094712
 (this is a region with a SNP thatdisrupts a binding motif and is involved in inflamation. 
 THP1 should be heterozygous and we should be able to see one allele open and one closed)

 And this returned significant result for 3 topcis for CXCL2, RELB, and IL1A. However, after centering on +/- 1000 bp of the TSS, the results are no longer significant.


Purely statistical test of the topic trajectories from
`1_m6a_promoter_topic_modelling.Rmd`: **does each topic's usage change
significantly with time?**

Inputs are the per-gene `read_topic_assignments_<outname>.tsv` tables exported
by notebook 1 (one row per read; `k1..k6` are the topic proportions from the
binomial `fit$L`; `sample_name` gives the LPS timepoint).

Each gene gets exactly k tests, one per topic: for every gene x topic
(9 x 6 = 54 tests) the unit of analysis is one read, the response is that
read's topic proportion, and the predictor is the timepoint treated as
**categorical** (0 min = reference level):

$$p_i = \beta_0 + \beta_5 I(t_i = 5) + \beta_{10} I(t_i = 10) + \beta_{15} I(t_i = 15) + \epsilon_i$$

- $\beta_0$ = mean loading at 0 min; $\beta_5$, $\beta_{10}$, $\beta_{15}$ =
  change in mean loading at 5/10/15 min relative to 0 min.
- The per-topic test is the model F-test: do the timepoint means differ at
  all? Because time is categorical this does not require the trajectory to be
  a straight line (a topic that peaks at 5 min and returns to baseline is
  detected).
- BH-FDR **within each gene** (across its 6 topics): each gene is treated as
  its own hypothesis family - "does topic usage change at this gene?" - so a
  p-value competes only against the other 5 topics of the same gene. Note this
  does not control false discoveries across the 9 genes jointly.

The read-level test treats reads as independent draws from their timepoint,
but reads from the same library share culture, prep, and run, so they are
correlated (pseudoreplication). The **pseudobulk** sections at the end re-run
the same test at the replicate level: each timepoint was sequenced as 3
barcoded replicates (R1-R3), reads are aggregated to per-library mean
proportions (12 libraries = 4 timepoints x 3 replicates), and the error term
becomes between-replicate variation.

# Settings

Same sample table, regions, and `output_dir` as notebook 1, so the `outname`
paths resolve to the exported TSVs.

```{r settings}
options(scipen = 999)

sample_table <- data.frame(
  sample_name = c("LPS_0", "LPS_5", "LPS_10", "LPS_15"),
  timepoint   = c(0L, 5L, 10L, 15L)
)

# original:
# output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model"
output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/promoters"

# original (explicit genomic intervals):
# promoters <- data.frame(
#   gene  = c("IL1B", "TNF", "IL1A", "PTGS2", "IL8", "CXCL2", "IL6", "RELB", "rs2836882"),
#   chr   = c("chr2", "chr6", "chr2", "chr1", "chr4", "chr4", "chr7", "chr19", "chr21"),
#   start = c(112836323, 31575265, 112793233, 186679985, 73740006, 74098863, 22726376, 45001011, 39094468),
#   end   = c(112837385, 31575824, 112794969, 186681163, 73741332, 74099651, 22727094, 45002417, 39094712)
# )

# TSS +/- 1000 bp, matching the TSS-centered regions built in
# 1_m6a_promoter_topic_modelling.Rmd (dominant THP-1 CAGE peak per gene:
# IL1B/TNF/IL8; GENCODE v49 TSS fallback: IL1A/PTGS2/CXCL2/IL6/RELB).
# rs2836882 is a SNP locus with no TSS - it keeps its explicit interval.
promoters <- data.frame(
  gene  = c("IL1B", "TNF", "IL1A", "PTGS2", "IL8", "CXCL2", "IL6", "RELB", "rs2836882"),
  chr   = c("chr2", "chr6", "chr2", "chr1", "chr4", "chr4", "chr7", "chr19", "chr21"),
  start = c(112835779, 31574568, 112783493, 186679922, 73739569, 74098196, 22724884, 45000461, 39094468),
  end   = c(112837779, 31576568, 112785493, 186681922, 73741569, 74100196, 22726884, 45002461, 39094712)
)
```

# Load per-read topic tables

One table per gene from `read_topic_assignments_<outname>.tsv`; `sample_name`
is mapped to numeric time via `sample_table`.

```{r topic-assignments}
# check read_topic_assignments_<outname>.tsv files 

topic_assignments <- lapply(seq_len(nrow(promoters)), function(i) {
  # original:
  # outname <- paste0(promoters$gene[i], "_", promoters$chr[i], "_",
  #                   promoters$start[i], "_", promoters$end[i])
  outname <- promoters$gene[i]
  file.path(output_dir, outname,
            paste0("read_topic_assignments_", outname, ".tsv"))
})

cat(unlist(topic_assignments), sep = "\n")

```

```{r check-files}
example_file <- topic_assignments[[1]]
head(read.delim(example_file), n = 3)
```

                                 RID sample_name  chr     start       end
1 m84241_260505_182714_s3/103682934/ccs      LPS_15 chr2 112830678 112843009
2  m84241_260505_182714_s3/13174534/ccs       LPS_5 chr2 112826359 112840099
3 m84241_260505_182714_s3/135791181/ccs       LPS_5 chr2 112829972 112839548
  strand  cluster dominant_topic max_loading                k1
1      - cluster5             k1   0.5705259 0.570525927968593
2      + cluster2             k6   0.7198507 0.000000009389935
3      - cluster6             k5   0.6735576 0.222503692486331
                 k2                k3                k4                k5
1 0.000000009587537 0.259254589912311 0.163050429693455 0.000000008945988
2 0.000000009341036 0.000000009331383 0.000000008479193 0.280149259250181
3 0.000000009336609 0.103938640272882 0.000000008822323 0.673557639145690
                 k6
1 0.007169033892115
2 0.719850704208272
3 0.000000009936166


 IL1B       TNF      IL1A     PTGS2       IL8     CXCL2       IL6      RELB 
      175       100       192       329       207       215       161       145 
rs2836882 
      216 

```{r load-tables}
per_read_tables <- lapply(seq_len(nrow(promoters)), function(i) {
  # original:
  # outname <- paste0(promoters$gene[i], "_", promoters$chr[i], "_",
  #                   promoters$start[i], "_", promoters$end[i])
  outname <- promoters$gene[i]
  df <- read.delim(file.path(output_dir, outname,
                             paste0("read_topic_assignments_", outname, ".tsv")))
  df$time <- sample_table$timepoint[match(df$sample_name, sample_table$sample_name)]
  stopifnot(!any(is.na(df$time)))
  df
})
names(per_read_tables) <- promoters$gene

sapply(per_read_tables, nrow)
```

# Per-topic linear models of topic proportion vs categorical time

```{r topic-time-lm}
##################
# fit linear regression of topic proportion vs categorical time for each gene x topic
##################

# loops over genes and topics, fit lm(prop ~ time) for each topic's read-level proportions

topic_time_tests <- do.call(rbind, lapply(promoters$gene, function(g) {
  # pulls out the per-read table for this gene, then loops over its 6 topics
  df <- per_read_tables[[g]]
  # identifies the topic columns by regex, then loops over them
  L_cols <- grep("^k[0-9]+$", colnames(df), value = TRUE)

  # a second loop over the six topic columns within this gene
  do.call(rbind, lapply(L_cols, function(cl) {
    # one row per read: topic proportion + categorical time
    # two columns: prop = topic proportion, time = factor(0/5/10/15)
    d <- data.frame(prop = df[[cl]],
                    time = factor(df$time, levels = c(0, 5, 10, 15)))
    
    # fit the categorical-time model and extract the coefficients
    fit <- lm(prop ~ time, data = d)
    est <- coef(fit)
    data.frame(gene = g, topic = sub("^k", "topic", cl), n_reads = nrow(d),
               mean_0min  = est[["(Intercept)"]],
               beta_5min  = est[["time5"]],
               beta_10min = est[["time10"]],
               beta_15min = est[["time15"]],
               # p_F = does the topic change somewhere across the 4 timepoints? (F-test of the lm)
               p_F        = anova(fit)$`Pr(>F)`[1])
  }))
}))

##################
# multiple testing correction
##################
# BH within each gene: adjust each gene's 6 F-test p-values as one family

## ave() groups by gene, then applies p.adjust() to each gene's p-values
### apply a function within groups, then return the results in the same positions as the original row
topic_time_tests$fdr_within_gene <- ave(topic_time_tests$p_F,
                                        topic_time_tests$gene,
                                        FUN = function(p) p.adjust(p, method = "BH"))

# verify compositional coupling: within each gene the 0-min means sum to 1
# across topics, and each timepoint's betas sum to 0 across topics
stopifnot(all(abs(tapply(topic_time_tests$mean_0min,
                         topic_time_tests$gene, sum) - 1) < 1e-10))
for (b in c("beta_5min", "beta_10min", "beta_15min")) {
  stopifnot(all(abs(tapply(topic_time_tests[[b]],
                           topic_time_tests$gene, sum)) < 1e-10))
}

write.table(topic_time_tests, file.path(output_dir, "topic_time_lm_results.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)


head(topic_time_tests[, c("gene", "topic", "n_reads", "mean_0min",
                          "beta_5min", "beta_10min", "beta_15min",
                          "p_F", "fdr_within_gene")], n = 3)
```

  gene  topic n_reads  mean_0min   beta_5min   beta_10min  beta_15min       p_F
1 IL1B topic1     175 0.10962695 -0.02813547 -0.044041817  0.01062016 0.3211720
2 IL1B topic2     175 0.15523414 -0.01584559  0.004705654 -0.03260006 0.8949180
3 IL1B topic3     175 0.07050958 -0.02343586  0.015932075  0.04110897 0.2963964
  fdr_within_gene
1       0.4817580
2       0.9386838
3       0.4817580

# Results

All 54 tests sorted by the F-test p-value, then the hits at within-gene FDR < 0.05


  gene  topic mean_0min beta_5min beta_10min beta_15min fdr_within_gene
 CXCL2 topic4    0.1142    0.0739    -0.0415    -0.0453        0.000423
  RELB topic5    0.1308    0.0149     0.1212     0.2516        0.008786
  IL1A topic4    0.0702   -0.0133     0.0771     0.0462        0.019013

  
```{r results}
print(topic_time_tests[order(topic_time_tests$p_F), ],
      digits = 3, row.names = FALSE)

cat("\n== significant at within-gene FDR < 0.05 ==\n")
sig <- topic_time_tests[topic_time_tests$fdr_within_gene < 0.05, ]
print(sig[order(sig$p_F),
          c("gene", "topic", "mean_0min", "beta_5min", "beta_10min",
            "beta_15min", "fdr_within_gene")],
      digits = 3, row.names = FALSE)
```

# Plots: raw reads + estimated means, and mean trajectories with 95% CI

Two views of the categorical-time model, per gene (one panel per topic):

1. **Raw reads + estimated mean**: every read's topic proportion, overlaid with the model-estimated mean at each timepoint.
   Shows how much read-to-read spread sits under each mean.
2. **Mean trajectory + 95% CI**: the estimated means at 0/5/10/15 min
   connected, with the model's 95% confidence band
   (`predict(..., interval = "confidence")`).

Topics significant at within-gene FDR < 0.05 in the read-level test (the
`sig` table above) are flagged with `***` in their panel title.

Note: the CIs come from the read-level model, so they inherit its assumption
that reads are independent


```{r topic-time-plots, fig.width=9, fig.height=6, message=FALSE, warning=FALSE}
library(ggplot2)

topic_colors <- c(topic1 = "#E69F00", topic2 = "#56B4E9", topic3 = "#009E73",
                  topic4 = "#F0E442", topic5 = "#0072B2", topic6 = "#D55E00")

for (g in promoters$gene) {
  df <- per_read_tables[[g]]
  L_cols <- grep("^k[0-9]+$", colnames(df), value = TRUE)

  # per-read proportions in long format
  long <- do.call(rbind, lapply(L_cols, function(cl)
    data.frame(topic = sub("^k", "topic", cl), time = df$time, prop = df[[cl]])))

  # estimated timepoint means + 95% CI from the categorical-time model
  nd <- data.frame(time = factor(c(0, 5, 10, 15), levels = c(0, 5, 10, 15)))
  means <- do.call(rbind, lapply(L_cols, function(cl) {
    fit <- lm(prop ~ time,
              data = data.frame(prop = df[[cl]],
                                time = factor(df$time, levels = c(0, 5, 10, 15))))
    ci <- predict(fit, nd, interval = "confidence")
    data.frame(topic = sub("^k", "topic", cl), time = c(0, 5, 10, 15),
               mean = ci[, "fit"], lwr = ci[, "lwr"], upr = ci[, "upr"])
  }))

  # label topics significant at within-gene FDR < 0.05 (from `sig`) with ***
  sig_topics <- sig$topic[sig$gene == g]
  lab <- function(t) ifelse(t %in% sig_topics, paste0(t, " ***"), t)
  topic_levels    <- sub("^k", "topic", L_cols)
  long$topic_lab  <- factor(lab(long$topic),  levels = lab(topic_levels))
  means$topic_lab <- factor(lab(means$topic), levels = lab(topic_levels))

  # original:
  # region  <- promoters[promoters$gene == g, ]
  # outname <- paste0(g, "_", region$chr, "_", region$start, "_", region$end)
  outname <- g

  # 1. raw reads + estimated mean
  gg1 <- ggplot() +
    # plot each read as a point
    geom_jitter(data = long, aes(time, prop), width = 0.7, height = 0,
                size = 0.4, alpha = 0.35, color = "grey30") +
    geom_line(data = means, aes(time, mean, color = topic), linewidth = 0.8) +
    # point of estimated mean at each timepoint
    geom_point(data = means, aes(time, mean, color = topic), size = 1.8) +
    facet_wrap(~ topic_lab) +
    scale_color_manual(values = topic_colors, guide = "none") +
    scale_x_continuous(breaks = c(0, 5, 10, 15)) +
    labs(x = "LPS timepoint (min)", y = "topic proportion",
         title = paste0(g, " - per-read topic proportions + estimated means")) +
    theme_classic(base_size = 10) +
    theme(plot.title = element_text(hjust = 0.5))
  print(gg1)
  ggsave(file.path(output_dir, outname,
                   paste0("topic_reads_mean_", outname, ".pdf")),
         gg1, width = 9, height = 6)

  # 2. mean trajectory + 95% CI
  gg2 <- ggplot(means, aes(time, mean, color = topic, fill = topic)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    facet_wrap(~ topic_lab) +
    scale_color_manual(values = topic_colors, guide = "none") +
    scale_fill_manual(values = topic_colors, guide = "none") +
    scale_x_continuous(breaks = c(0, 5, 10, 15)) +
    labs(x = "LPS timepoint (min)", y = "mean topic proportion",
         title = paste0(g, " - estimated mean trajectory (95% CI)")) +
    theme_classic(base_size = 10) +
    theme(plot.title = element_text(hjust = 0.5))
  print(gg2)
  ggsave(file.path(output_dir, outname,
                   paste0("topic_mean_ci_", outname, ".pdf")),
         gg2, width = 9, height = 6)
}
```

# Replicate assignments

Each timepoint's library was sequenced as 3 barcoded replicates (R1-R3). The
`rid_replicate` maps under
`<ft_result_dir>/<sample>/extracted_results/replicate_by_chr/` assign every
read to its replicate. Here every read in the per-gene tables gets its
replicate label; all reads must be assigned (verified). The genome-wide maps
are large, so each one is subset to the RIDs we actually use right after
loading.

```{r replicate-maps}
ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

# RIDs we need, per chromosome (keeps the genome-wide maps out of memory)
needed_rids <- lapply(split(promoters$gene, promoters$chr), function(gs)
  unique(unlist(lapply(per_read_tables[gs], function(df) df$RID))))

# rid -> replicate lookup per (sample, chr), subset to the needed RIDs
rep_maps <- list()
for (s in sample_table$sample_name) {
  for (chr in names(needed_rids)) {
    f <- file.path(ft_result_dir, s, "extracted_results", "replicate_by_chr",
                   paste0(s, ".rid_replicate.", chr, ".tsv.gz"))
    m <- read.delim(f, colClasses = c("character", "character"))
    m <- m[m$RID %in% needed_rids[[chr]], ]
    rep_maps[[paste(s, chr, sep = "_")]] <- setNames(m$replicate, m$RID)
  }
}

# attach the replicate label to every read
per_read_tables <- lapply(promoters$gene, function(g) {
  df  <- per_read_tables[[g]]
  chr <- promoters$chr[promoters$gene == g]
  df$replicate <- NA_character_
  for (s in sample_table$sample_name) {
    idx <- df$sample_name == s
    df$replicate[idx] <- unname(rep_maps[[paste(s, chr, sep = "_")]][df$RID[idx]])
  }
  stopifnot(!any(is.na(df$replicate)))
  df
})
names(per_read_tables) <- promoters$gene

# reads per timepoint x replicate library, per gene
for (g in promoters$gene) {
  cat("=== ", g, " ===\n", sep = "")
  print(table(time      = per_read_tables[[g]]$time,
              replicate = per_read_tables[[g]]$replicate))
}

head(per_read_tables$IL1B[, c("RID", "sample_name", "time", "replicate")], n = 3)
```

                                    RID sample_name time replicate
1 m84241_260505_182714_s3/103682934/ccs      LPS_15   15        R2
2  m84241_260505_182714_s3/13174534/ccs       LPS_5    5        R2
3 m84241_260505_182714_s3/135791181/ccs       LPS_5    5        R3

# Pseudobulk: aggregate to replicate level

The read-level test above uses read-to-read scatter as its error term, which
ignores that reads from the same library are correlated. Here each
(timepoint x replicate) library is collapsed to one observation per topic -
the **mean topic proportion across its reads** - giving 12 observations per
gene x topic. The same categorical-time model and F-test are then fit on
those means, with 3 and 8 degrees of freedom: a timepoint effect now has to
stand out against between-replicate variation, and must be consistent across
R1-R3 to be significant. Same within-gene BH as before.


```{r pseudobulk}
pseudobulk_tests <- do.call(rbind, lapply(promoters$gene, function(g) {
  df <- per_read_tables[[g]]
  L_cols <- grep("^k[0-9]+$", colnames(df), value = TRUE)

  # one row per (timepoint, replicate) library: mean topic proportions
  ## aggregate() groups by time and replicate, then takes the mean of each topic column
  ## for each time x replicate group, calculate the mean of topic proportions across all reads in that library
  pb <- aggregate(df[, L_cols],
                  by = list(time = df$time, replicate = df$replicate),
                  FUN = mean)

  do.call(rbind, lapply(L_cols, function(cl) {
    d <- data.frame(prop = pb[[cl]],
                    time = factor(pb$time, levels = c(0, 5, 10, 15)))
    fit <- lm(prop ~ time, data = d)
    est <- coef(fit)
    data.frame(gene = g, topic = sub("^k", "topic", cl),
               n_libraries = nrow(d),
               mean_0min  = est[["(Intercept)"]],
               beta_5min  = est[["time5"]],
               beta_10min = est[["time10"]],
               beta_15min = est[["time15"]],
               p_F        = anova(fit)$`Pr(>F)`[1])
  }))
}))

pseudobulk_tests$fdr_within_gene <- ave(pseudobulk_tests$p_F,
                                        pseudobulk_tests$gene,
                                        FUN = function(p) p.adjust(p, method = "BH"))

# same compositional coupling holds at the library level
stopifnot(all(abs(tapply(pseudobulk_tests$mean_0min,
                         pseudobulk_tests$gene, sum) - 1) < 1e-10))
for (b in c("beta_5min", "beta_10min", "beta_15min")) {
  stopifnot(all(abs(tapply(pseudobulk_tests[[b]],
                           pseudobulk_tests$gene, sum)) < 1e-10))
}

write.table(pseudobulk_tests,
            file.path(output_dir, "topic_time_pseudobulk_results.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

head(pseudobulk_tests[, c("gene", "topic", "n_libraries", "mean_0min",
                          "beta_5min", "beta_10min", "beta_15min",
                          "p_F", "fdr_within_gene")], n = 3)
```


  gene  topic n_libraries mean_0min    beta_5min  beta_10min  beta_15min
1 IL1B topic1          12 0.1379681 -0.013478263 -0.01446101 -0.03871273
2 IL1B topic2          12 0.1528722 -0.072098702 -0.06137230 -0.02076232
3 IL1B topic3          12 0.1114026 -0.001162152  0.09894804  0.01878489
         p_F fdr_within_gene
1 0.87506195       0.9996089
2 0.08920973       0.3678950
3 0.19421843       0.3884369


# Pseudobulk results

All 54 tests sorted by the F-test p-value, the hits at within-gene
FDR < 0.05, and a side-by-side of the read-level vs pseudobulk p-values for
the read-level hits.

 gene  topic mean_0min beta_5min beta_10min beta_15min fdr_within_gene
  IL8 topic5     0.117   -0.0319     0.0863     0.0211          0.0111
```{r pseudobulk-results}
print(pseudobulk_tests[order(pseudobulk_tests$p_F), ],
      digits = 3, row.names = FALSE)

cat("\n== significant at within-gene FDR < 0.05 ==\n")
sig_pb <- pseudobulk_tests[pseudobulk_tests$fdr_within_gene < 0.05, ]
print(sig_pb[order(sig_pb$p_F),
             c("gene", "topic", "mean_0min", "beta_5min", "beta_10min",
               "beta_15min", "fdr_within_gene")],
      digits = 3, row.names = FALSE)

cat("\n== read-level hits: read-level vs pseudobulk ==\n")
cmp <- merge(sig[, c("gene", "topic", "p_F", "fdr_within_gene")],
             pseudobulk_tests[, c("gene", "topic", "p_F", "fdr_within_gene")],
             by = c("gene", "topic"), suffixes = c("_read", "_pseudobulk"))
print(cmp[order(cmp$p_F_read), ], digits = 3, row.names = FALSE)
```
  gene  topic  p_F_read fdr_within_gene_read p_F_pseudobulk
 CXCL2 topic4 0.0000705             0.000423         0.0302
  RELB topic5 0.0014644             0.008786         0.0422
  IL1A topic4 0.0031689             0.019013         0.0174
 fdr_within_gene_pseudobulk
                     0.0905
                     0.1266
                     0.0523

# IL8 pseudobulk trajectory

Replicate-level view of the one pseudobulk hit. Each panel shows the 12
library means (one point per timepoint x replicate, shaped by replicate) with
the pseudobulk model's estimated mean trajectory and its 95% CI - the
replicate-aware counterpart of the read-level CI plot above. The topic
significant at within-gene FDR < 0.05 in the pseudobulk test (`sig_pb`) is
flagged with `***`.

```{r il8-pseudobulk-trajectory, fig.width=9, fig.height=6, message=FALSE, warning=FALSE}
g  <- "IL8"
df <- per_read_tables[[g]]
L_cols <- grep("^k[0-9]+$", colnames(df), value = TRUE)

# per-library mean proportions (same aggregation as the pseudobulk test)
pb <- aggregate(df[, L_cols],
                by = list(time = df$time, replicate = df$replicate),
                FUN = mean)

pb_long <- do.call(rbind, lapply(L_cols, function(cl)
  data.frame(topic = sub("^k", "topic", cl), time = pb$time,
             replicate = pb$replicate, prop = pb[[cl]])))

# pseudobulk model means + 95% CI per topic (fit on the 12 library means)
nd <- data.frame(time = factor(c(0, 5, 10, 15), levels = c(0, 5, 10, 15)))
means_pb <- do.call(rbind, lapply(L_cols, function(cl) {
  fit <- lm(prop ~ time,
            data = data.frame(prop = pb[[cl]],
                              time = factor(pb$time, levels = c(0, 5, 10, 15))))
  ci <- predict(fit, nd, interval = "confidence")
  data.frame(topic = sub("^k", "topic", cl), time = c(0, 5, 10, 15),
             mean = ci[, "fit"], lwr = ci[, "lwr"], upr = ci[, "upr"])
}))

# label topics significant at within-gene FDR < 0.05 in the pseudobulk test
sig_topics <- sig_pb$topic[sig_pb$gene == g]
lab <- function(t) ifelse(t %in% sig_topics, paste0(t, " ***"), t)
topic_levels       <- sub("^k", "topic", L_cols)
pb_long$topic_lab  <- factor(lab(pb_long$topic),  levels = lab(topic_levels))
means_pb$topic_lab <- factor(lab(means_pb$topic), levels = lab(topic_levels))

# original:
# region  <- promoters[promoters$gene == g, ]
# outname <- paste0(g, "_", region$chr, "_", region$start, "_", region$end)
outname <- g

gg_pb <- ggplot() +
  geom_ribbon(data = means_pb, aes(time, ymin = lwr, ymax = upr, fill = topic),
              alpha = 0.2) +
  geom_line(data = means_pb, aes(time, mean, color = topic), linewidth = 0.8) +
  geom_point(data = means_pb, aes(time, mean, color = topic), size = 1.8) +
  geom_point(data = pb_long, aes(time, prop, shape = replicate),
             size = 1.8, color = "grey20") +
  facet_wrap(~ topic_lab) +
  scale_color_manual(values = topic_colors, guide = "none") +
  scale_fill_manual(values = topic_colors, guide = "none") +
  scale_x_continuous(breaks = c(0, 5, 10, 15)) +
  labs(x = "LPS timepoint (min)", y = "mean topic proportion",
       title = paste0(g, " - pseudobulk trajectory (library means, 95% CI)"),
       shape = "replicate") +
  theme_classic(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
print(gg_pb)
ggsave(file.path(output_dir, outname,
                 paste0("topic_pseudobulk_traj_", outname, ".pdf")),
       gg_pb, width = 9, height = 6)
```

