---
title: "Genome-wide LCL co-accessibility: top five pairs"
author: "Crystal Shan"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  root: "/project/spott/cshan/fiber-seq/LCL_project/co-accessibility"
  sample_sheet: "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv"
  fire_root: "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
  tf_root: "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/FiberHMM_v2/results/results_v1_trained_model"
  tabix: "/project/spott/cshan/envs/dimelo/bin/tabix"
  code_root: "/project/spott/cshan/fiber-seq/code"
  reference_fasta: "/project/spott/reference/human/GRCh38.p14/hg38.fa"
  tss_bed: "/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed"
  genome_build: "GRCh38"
  summarize_lengths: true
  triangle_tile_bp: null
  locus_flank_bp: 1000
  footprint_flank_bp: 200
  footprint_max_length_bp: 400
  footprint_bin_bp: 5
  tf_format: "bed15_fiberhmm_tf"
  make_plots: true
  max_plot_reads: 120
  seed: 20261009
  allow_partial_results: false
---

This notebook reads the completed **genome-wide, pooled 31-sample LCL analysis**
from `LCL_coaccess_cres.py`. Run `LCL_fire_universe.sh`, `LCL_read_spans.sh`, and
`LCL_run_coaccess.sh` first. Donor labels are metadata only; there are no donor
strata, haplotype tests, or timepoints.

The plotting workflow is written directly below in separate chunks. Source the
shared parsing and plotting functions, then run the chunks in order, or knit this
notebook. `LCL_plot_coaccess.sh` is an optional SLURM launcher for the same notebook.

All 31 LCL samples form one pooled group. The additional figures are cCRE and
tiled FIRE correlation triangles around the current top-five pairs, and footprint
midpoint–length plots anchored at **each member cCRE's midpoint**. There is no
motif orientation or MNase input. Length and coverage summaries are descriptive:
this notebook adds no QC stage, sample exclusions, or quality-score thresholds.

Candidate cCREs overlap the union of all 31 samples' FIRE peaks. Pairs share a
canonical TSS ±10 kb window and have **500 < gap < 20,000 bp**. A shared fiber
overlaps both cCREs by at least one base. It is accessible at a cCRE when **one
FIRE element on that same sample/read overlaps at least 50% of the cCRE length**.
Several smaller FIRE elements do not combine to meet the threshold. The fraction
is relative to cCRE length, not to FIRE-element length.

The four configuration counts are pooled across samples before testing. The
primary FDR comes from the macrophage-compatible **two-sided Fisher test after
adding one pseudocount to each pooled cell**, corrected across all tested unique
pairs genome-wide. Positive pairs require `fdr < 0.05`, `OR > 1`, and observed
joint accessibility greater than its pooled independence expectation. This
notebook selects the top five by FDR, decreasing OR, then decreasing shared-fiber
coverage, using pair ID to resolve ties. It does not recalculate FDR after selection.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                     fig.width = 11, fig.height = 6, cache = FALSE)
```

```{r packages-and-functions}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
})
source(file.path(params$code_root, "parsing_functions", "parsing_footprints_functions.r"), local = TRUE)
source(file.path(params$code_root, "parsing_functions", "plotting_functions.r"), local = TRUE)
options(scipen = 999)
```

```{r settings}
root <- params$root
coaccess_dir <- file.path(root, "coaccess")
universe_dir <- file.path(root, "universe")
plot_dir <- file.path(coaccess_dir, "plots", "top5")
read_rule <- "any"
fire_overlap_fraction <- 0.5
fdr_cutoff <- 0.05
max_plot_reads <- as.integer(params$max_plot_reads)
stopifnot(length(max_plot_reads) == 1L, !is.na(max_plot_reads), max_plot_reads > 0L)
paths <- list(root = root, fire_root = params$fire_root,
              ft_root = params$fire_root, tf_root = params$tf_root,
              tabix_bin = params$tabix, fire_ver = "v0.1")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
summary_dir <- file.path(coaccess_dir, "plots", "summaries")
dir.create(summary_dir, recursive = TRUE, showWarnings = FALSE)
stopifnot(params$locus_flank_bp >= 0, params$footprint_flank_bp > 0,
          params$footprint_max_length_bp > 0, params$footprint_bin_bp > 0)
```

# Input provenance

```{r validate-all-31-samples}
sample_meta <- fread(params$sample_sheet)
manifest <- fread(file.path(universe_dir, "sample_manifest.tsv"))
stopifnot(all(c("sample_name", "cell_line") %in% names(sample_meta)),
          all(c("sample_name", "donor") %in% names(manifest)),
          nrow(sample_meta) == 31L, nrow(manifest) == 31L,
          !anyDuplicated(sample_meta$sample_name), !anyDuplicated(manifest$sample_name),
          setequal(sample_meta$sample_name, manifest$sample_name))
sample_meta[, donor := as.character(cell_line)]
samples <- as.character(manifest$sample_name)
stopifnot(identical(as.character(manifest$donor),
                    sample_meta$donor[match(samples, sample_meta$sample_name)]))
knitr::kable(sample_meta[, .(sample_name, donor)])
```

```{r validate-completed-run}
# This marker is published last and removed at the start of a rerun. Do not
# combine old pair tables with an unfinished run or a changed input manifest.
run_info_file <- file.path(coaccess_dir, "run_info.json")
if (!file.exists(run_info_file))
  stop("No completed-run marker: ", run_info_file,
       ". Finish LCL_run_coaccess.sh before rendering.")
run_info <- jsonlite::fromJSON(run_info_file)
stopifnot(isTRUE(run_info$dataset == "LCL"), isTRUE(run_info$n_samples == 31L),
          isTRUE(run_info$read_rule == read_rule),
          isTRUE(run_info$fire_overlap_fraction == fire_overlap_fraction),
          isFALSE(run_info$donor_used_in_testing),
          length(run_info$sample_names) == 31L,
          !anyDuplicated(run_info$sample_names),
          setequal(run_info$sample_names, samples),
          length(run_info$analysis_scope) == 1L, !is.na(run_info$analysis_scope),
          length(run_info$tested_pairs) == 1L, !is.na(run_info$tested_pairs))
if (!isTRUE(params$allow_partial_results) && run_info$analysis_scope != "genome-wide")
  stop("Completed run covers ", run_info$analysis_scope, "; genome-wide results are required.")

run_manifest <- fread(file.path(coaccess_dir, "sample_manifest.tsv"))
provenance_columns <- c("sample_name", "donor", "fire_dir", "cram_path", "elements_path", "spans_path")
stopifnot(all(provenance_columns %in% names(run_manifest)),
          all(provenance_columns %in% names(manifest)),
          nrow(run_manifest) == 31L, !anyDuplicated(run_manifest$sample_name),
          setequal(run_manifest$sample_name, samples))
run_manifest <- run_manifest[match(samples, sample_name)]
for (column in provenance_columns)
  if (!identical(as.character(run_manifest[[column]]), as.character(manifest[[column]])))
    stop("Universe and completed-run manifests differ in ", column, "; rerun the analysis.")
same_paths <- function(a, b)
  identical(normalizePath(a, mustWork = FALSE), normalizePath(b, mustWork = FALSE))
stopifnot(same_paths(run_manifest$spans_path,
  file.path(root, samples, paste0(samples, ".read_spans.bed.gz"))),
  same_paths(run_manifest$elements_path,
    file.path(paths$fire_root, samples, "additional-outputs-v0.1", "fire-peaks",
              paste0(samples, "-v0.1-fire-elements.bed.gz"))))
```

```{r read-genome-wide-pair-results}
pairs_file <- file.path(coaccess_dir, "LCL_coaccess_pairs.tsv.gz")
pairs_dt <- fread(pairs_file)
required_columns <- c("CRE_pair", "CRE1", "CRE2", "chrom", "cre1_start", "cre1_end",
  "cre2_start", "cre2_end", "gene_names", "CRE_pair_labels", "dist", "n_shared_reads",
  "co_access", "co_closed", "CRE1_access", "CRE2_access", "OR", "fdr", "fdr_raw",
  "expected_co_access", "observed_expected_ratio", "excess_fraction", "read_rule",
  "fire_overlap_fraction", "dataset", "analysis_scope")
if (!all(required_columns %in% names(pairs_dt)))
  stop("Missing result columns: ", paste(setdiff(required_columns, names(pairs_dt)), collapse = ", "))
stopifnot(!anyDuplicated(pairs_dt$CRE_pair),
          nrow(pairs_dt) == run_info$tested_pairs,
          all(pairs_dt$dataset == "LCL"),
          all(pairs_dt$analysis_scope == run_info$analysis_scope),
          all(pairs_dt$read_rule == read_rule),
          all(pairs_dt$fire_overlap_fraction == fire_overlap_fraction),
          all(pairs_dt$dist > 500 & pairs_dt$dist < 20000),
          all(pairs_dt$n_shared_reads > 0),
          all(pairs_dt$n_shared_reads == pairs_dt$co_access + pairs_dt$co_closed +
                pairs_dt$CRE1_access + pairs_dt$CRE2_access))
if (!isTRUE(params$allow_partial_results) && any(pairs_dt$analysis_scope != "genome-wide"))
  stop("Partial chromosome results supplied. Run the genome-wide analysis before rendering.")
expected_from_counts <- with(pairs_dt,
  (co_access + CRE1_access) * (co_access + CRE2_access) / n_shared_reads)
stopifnot(isTRUE(all.equal(as.numeric(pairs_dt$expected_co_access), as.numeric(expected_from_counts),
                          tolerance = 1e-8, check.attributes = FALSE)))
cat(nrow(pairs_dt), "unique tested pairs loaded.\n")
```

```{r read-ccre-and-gene-annotations}
cre_universe <- fread(file.path(universe_dir, "cre_universe.bed"), header = FALSE,
  col.names = c("chrom", "start", "end", "CRE_ID", "CRE_label"))
tss_bed <- params$tss_bed
tss <- fread(tss_bed, header = FALSE, select = 1:6,
  col.names = c("chrom", "start", "end", "annotation", "score", "strand"))
tss[, gene_name := tstrsplit(annotation, ";", fixed = TRUE, keep = 3L)[[1L]]]
```

# Descriptive read-length and coverage summaries

The existing FIRE report's `fiber_length` frequency table supplies sequencing
fiber lengths for its reported population. Only that table is read; no QC is run.
The primary BED files separately supply **reference-alignment span lengths** for
the analysis population. These lengths may differ because of clipping, indels,
and the input populations. Pooled quantiles and N50 are calculated from combined
length counts, never by averaging sample medians or N50s.

```{r length-summary-functions}
summarize_length_histogram <- function(histogram) {
  h <- copy(as.data.table(histogram))[, .(n = sum(as.numeric(n))), by = length_bp]
  h <- h[n > 0]
  if (!nrow(h)) return(data.table(n_reads = 0, total_bp = 0, mean_bp = NA_real_,
    median_bp = NA_real_, p90_bp = NA_real_, n50_bp = NA_real_, max_bp = NA_real_))
  stopifnot(all(is.finite(h$length_bp)), all(h$length_bp > 0), all(is.finite(h$n)))
  setorder(h, length_bp)
  total <- sum(h$n)
  cumulative <- cumsum(h$n)
  quantile_at <- function(p) {
    index <- 1 + (total - 1) * p
    lower <- h$length_bp[which(cumulative >= floor(index))[1]]
    upper <- h$length_bp[which(cumulative >= ceiling(index))[1]]
    lower + (index - floor(index)) * (upper - lower)
  }
  bases <- sum(h$length_bp * h$n)
  descending <- h[order(-length_bp)]
  n50 <- descending$length_bp[which(cumsum(descending$length_bp * descending$n) >= bases / 2)[1]]
  data.table(n_reads = total, total_bp = bases, mean_bp = bases / total,
    median_bp = quantile_at(.5), p90_bp = quantile_at(.9), n50_bp = n50,
    max_bp = max(h$length_bp))
}
```

```{r calculate-read-length-summaries}
length_histograms <- data.table(sample_name = character(), length_type = character(),
  length_bp = numeric(), n = numeric())
length_sources <- data.table()
length_summary <- data.table()
if (isTRUE(params$summarize_lengths)) {
  length_sources <- rbindlist(lapply(seq_along(samples), function(i) {
    s <- samples[i]
    data.table(sample_name = s,
      length_type = c("FIRE_report_fiber_length", "primary_reference_span"),
      path = c(file.path(paths$fire_root, s, paste0(s, "-", paths$fire_ver, "-qc.tbl.gz")),
        run_manifest$spans_path[i]))
  }))
  length_sources[, `:=`(available = file.exists(path),
    size = file.info(path)$size, mtime = as.numeric(file.info(path)$mtime))]
  cache_file <- file.path(summary_dir, "length_histograms.rds")
  cache <- if (file.exists(cache_file)) readRDS(cache_file) else NULL
  if (!is.null(cache) && identical(cache$sources, length_sources) && identical(cache$version, 1L)) {
    length_histograms <- cache$histograms
  } else {
    pieces <- vector("list", nrow(length_sources))
    for (i in seq_len(nrow(length_sources))) {
      input <- length_sources[i]
      if (!input$available) next
      if (input$length_type == "FIRE_report_fiber_length") {
        report <- fread(input$path, showProgress = FALSE)
        h <- report[statistic == "fiber_length",
          .(length_bp = as.numeric(value), n = as.numeric(count))]
      } else {
        spans_for_lengths <- fread(input$path, header = FALSE, select = c(2L, 3L),
          col.names = c("start", "end"), showProgress = FALSE)
        h <- spans_for_lengths[, .(n = .N), by = .(length_bp = end - start)]
      }
      h[, `:=`(sample_name = input$sample_name, length_type = input$length_type)]
      pieces[[i]] <- h
    }
    length_histograms <- rbindlist(pieces, use.names = TRUE, fill = TRUE)
    saveRDS(list(version = 1L, sources = copy(length_sources), histograms = length_histograms), cache_file)
  }
  if (nrow(length_histograms)) {
    length_summary <- length_histograms[, summarize_length_histogram(.SD),
      by = .(sample_name, length_type), .SDcols = c("length_bp", "n")]
    pooled <- length_histograms[, summarize_length_histogram(.SD),
      by = length_type, .SDcols = c("length_bp", "n")]
    pooled[, sample_name := "pooled"]
    length_summary <- rbindlist(list(length_summary, pooled), use.names = TRUE)
    length_summary[, donor := sample_meta$donor[match(sample_name, sample_meta$sample_name)]]
    fwrite(length_histograms, file.path(summary_dir, "LCL_read_length_histograms.tsv.gz"), sep = "\t")
    fwrite(length_summary, file.path(summary_dir, "LCL_read_length_summary.tsv"), sep = "\t")
    knitr::kable(length_summary, digits = 1)
  }
  fwrite(length_sources, file.path(summary_dir, "LCL_read_length_sources.tsv"), sep = "\t")
  if (any(!length_sources$available))
    cat("Length reports unavailable for", sum(!length_sources$available),
      "inputs; see the source table. No lengths are inferred for missing files.\n")
}
```

```{r genome-wide-pair-coverage-summary}
# Existing pair counts are descriptive here; no coverage filter or new test.
distance_summary <- if (nrow(pairs_dt)) pairs_dt[, .(n_pairs = .N, min_shared = min(n_shared_reads),
  median_shared = median(n_shared_reads), p90_shared = as.numeric(quantile(n_shared_reads, .9)),
  max_shared = max(n_shared_reads)), by = .(gap_bin_start_bp = floor(dist / 1000) * 1000)] else
  data.table(gap_bin_start_bp = numeric(), n_pairs = integer(), min_shared = numeric(),
    median_shared = numeric(), p90_shared = numeric(), max_shared = numeric())
setorder(distance_summary, gap_bin_start_bp)
fwrite(distance_summary, file.path(summary_dir, "LCL_pair_coverage_by_distance.tsv"), sep = "\t")
if (nrow(distance_summary)) {
  coverage_plot <- ggplot(distance_summary, aes(gap_bin_start_bp / 1000, median_shared)) +
    geom_line() + geom_point() + theme_bw() +
    labs(x = "cCRE gap bin start (kb; 1 kb bins)", y = "Median jointly observed fibers",
      title = "Observed support within the existing genome-wide pair search",
      subtitle = "Descriptive only; no added coverage threshold or change to the tested distance range")
  print(coverage_plot)
  ggsave(file.path(summary_dir, "LCL_pair_coverage_by_distance.pdf"), coverage_plot, width = 8, height = 5)
}
```

# Top five positive associations

```{r select-top-five}
eligible <- pairs_dt[!is.na(fdr) & fdr < fdr_cutoff & !is.na(OR) & OR > 1 &
                       co_access > expected_co_access]
setorderv(eligible, c("fdr", "OR", "n_shared_reads", "CRE_pair"), c(1L, -1L, -1L, 1L))
selected <- head(eligible, 5L)
selected[, pair_id := sprintf("LCL_top%02d", seq_len(.N))]
fwrite(selected, file.path(plot_dir, "LCL_top5_pairs.tsv"), sep = "\t")
cat(nrow(eligible), "pairs pass the positive-association criteria; selected",
    nrow(selected), "for plotting.\n")
if (nrow(selected) < 5L)
  cat("Fewer than five pairs qualify; only qualifying pairs will be shown.\n")
knitr::kable(selected[, .(pair_id, gene_names, CRE_pair, dist, n_shared_reads,
  co_access, expected_co_access, observed_expected_ratio, OR, fdr, fdr_raw)], digits = 4)
```

```{r observed-versus-expected, fig.height=5}
if (nrow(selected)) {
  oe <- melt(selected[, .(pair_id, observed = co_access, expected = expected_co_access)],
             id.vars = "pair_id", variable.name = "count_type", value.name = "fibers")
  oe_plot <- ggplot(oe, aes(pair_id, fibers, fill = count_type)) +
    geom_col(position = "dodge") + theme_bw() +
    scale_fill_manual(values = c(observed = "#B2182B", expected = "grey70")) +
    labs(x = NULL, y = "Jointly accessible fibers", fill = NULL,
         title = "Observed and expected co-accessibility",
         subtitle = "Expectation uses pooled margins among the same jointly covered fibers")
  print(oe_plot)
  ggsave(file.path(plot_dir, "LCL_top5_observed_expected.pdf"), oe_plot, width = 9, height = 5)
}
```

# Fiber plots and count checks

The following code reloads **all 31 samples** from the same primary-read-span and
FIRE-element files used for counting. The complete reloaded 2×2 must match the
saved result; a discrepancy stops rendering. The raster shows at most
`r max_plot_reads` randomly selected shared fibers per pair; the m6A profile uses
those displayed fibers. Configuration bars, count checks, and statistical
annotations always use all shared fibers.

m6A, nucleosome, and TF-footprint tracks are display annotations restricted to
those same read identities. Missing optional display tracks produce warnings;
they never change the FIRE accessibility call. Donor identifiers are attached to
the exported read labels and sample contributions without entering a test.

```{r validate-plot-inputs}
if (isTRUE(params$make_plots) && nrow(selected)) {
  spans <- file.path(root, samples, paste0(samples, ".read_spans.bed.gz"))
  elements <- file.path(paths$fire_root, samples, "additional-outputs-v0.1", "fire-peaks",
                         paste0(samples, "-v0.1-fire-elements.bed.gz"))
  needed <- c(paths$tabix_bin, spans, paste0(spans, ".tbi"), elements, paste0(elements, ".tbi"))
  if (any(!file.exists(needed)))
    stop("Missing plotting inputs: ", paste(needed[!file.exists(needed)], collapse = "\n"))
}
```

## Load fibers and assign accessibility

`load_region()`, `label_reads()`, and `coaccess_2x2()` come from
`parsing_footprints_functions.r`. Read keys include the sample name so that equal
read names from different samples cannot be joined. Keep these complete labels
for count verification, configuration bars, and metadata exports.

```{r load-pair-fibers}
pair_data <- list()
if (isTRUE(params$make_plots) && nrow(selected)) {
  for (i in seq_len(nrow(selected))) {
    pair <- selected[i]
    region <- list(chrom = pair$chrom,
      start = max(0, min(pair$cre1_start, pair$cre2_start) - params$locus_flank_bp),
      end = max(pair$cre1_end, pair$cre2_end) + params$locus_flank_bp)
    cre1 <- list(start = pair$cre1_start, end = pair$cre1_end)
    cre2 <- list(start = pair$cre2_start, end = pair$cre2_end)
    res <- load_region(region, samples, paths$root, paths$fire_root,
      fire_ver = paths$fire_ver, tabix_bin = paths$tabix_bin,
      span_source = "read_spans")
    labels <- label_reads(res, cre1, cre2, read_rule = read_rule,
      fire_overlap_fraction = fire_overlap_fraction)
    pair_data[[pair$pair_id]] <- list(pair = pair, region = region,
      cre1 = cre1, cre2 = cre2, res = res, labels = labels,
      counts = coaccess_2x2(labels))
  }
}
```

```{r verify-reloaded-counts}
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  actual <- unlist(entry$counts[1, c("neither", "cre1_only", "cre2_only", "both", "n_shared")],
                   use.names = FALSE)
  expected <- unlist(pair[, .(co_closed, CRE1_access, CRE2_access, co_access, n_shared_reads)],
                     use.names = FALSE)
  if (!identical(as.numeric(actual), as.numeric(expected)))
    stop("Reloaded counts differ from the analysis table for ", pair$CRE_pair,
         "; reloaded: ", paste(actual, collapse = ","),
         "; saved: ", paste(expected, collapse = ","))
}
```

```{r export-fiber-metadata}
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  read_labels <- copy(entry$labels)[shared == TRUE]
  read_labels[, donor := sample_meta$donor[match(sample_name, sample_meta$sample_name)]]
  read_labels[, `:=`(pair_id = pair$pair_id, CRE_pair = pair$CRE_pair)]
  fwrite(read_labels, file.path(plot_dir, paste0(pair$pair_id, "_read_labels.tsv.gz")), sep = "\t")
  contributions <- read_labels[, .(n_shared = .N, co_access = sum(acc1 & acc2),
    CRE1_access = sum(acc1 & !acc2), CRE2_access = sum(!acc1 & acc2),
    co_closed = sum(!acc1 & !acc2)), by = .(pair_id, sample_name, donor)]
  fwrite(contributions, file.path(plot_dir, paste0(pair$pair_id, "_sample_counts.tsv")), sep = "\t")
}
```

## Prepare the displayed fibers and annotation tracks

Sample up to `r max_plot_reads` shared fibers with a fixed seed. Load m6A,
ft nucleosomes, and FiberHMM TF footprints only for these display keys using
`load_ft_tracks()` from the parsing functions. Copies of the display tables are
labelled `pooled`; the original sample/read keys and complete labels are retained.

```{r prepare-display-fibers}
display_data <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  drawn_labels <- copy(entry$labels)[shared == TRUE]
  set.seed(as.integer(params$seed))
  if (nrow(drawn_labels) > max_plot_reads) {
    drawn_keys <- sample(drawn_labels$key, max_plot_reads)
    drawn_labels <- drawn_labels[key %in% drawn_keys]
  }

  display_res <- load_ft_tracks(copy(entry$res), keys = drawn_labels$key,
    ft_root = paths$ft_root, tabix_bin = paths$tabix_bin,
    nuc_source = "ft", tf_source = "recalled_tf", tf_root = paths$tf_root)
  for (track in c("spans", "elements", "m6a", "nuc", "size_fps")) {
    if (!is.null(display_res[[track]]) && nrow(display_res[[track]]))
      display_res[[track]][, sample_name := "pooled"]
  }
  display_res$samples <- "pooled"
  drawn_labels[, sample_name := "pooled"]

  region <- entry$region
  cres_here <- cre_universe[chrom == region$chrom & end > region$start & start < region$end]
  region_res <- coaccess_region_result(display_res, entry$cre1, entry$cre2,
    drawn_labels, cres = cres_here)

  # Keep the cCRE class order used by the shared plotting functions.
  pk <- as.data.frame(cres_here)
  cls <- factor(pk$CRE_label, levels = c(intersect(CCRE_CLASSES, pk$CRE_label),
    setdiff(unique(pk$CRE_label), CCRE_CLASSES)))
  pk <- pk[order(cls, pk$start), , drop = FALSE]
  peaks_df <- data.frame(chr = pk$chrom, start = pk$start, end = pk$end,
    group = as.character(pk$CRE_label))
  display_data[[id]] <- list(region_res = region_res, peaks_df = peaks_df,
    n_drawn = nrow(drawn_labels))
}
```

## Draw the fiber panels

`coaccess_region_result()` above adapts the display data for
`plot_region_panels()` from `plotting_functions.r`. Panel choices, colours,
relative heights, titles, and the highlighted pair are set explicitly here.

```{r draw-fiber-panels}
fiber_plots <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  region <- entry$region
  counts <- entry$counts
  display <- display_data[[id]]
  region_res <- display$region_res

  subtitle <- sprintf(paste0("%s:%d-%d | gap %d bp | O %.0f / E %.2f | ",
    "OR %.3g | genome-wide FDR %.3g"), region$chrom, region$start, region$end,
    pair$dist, pair$co_access, pair$expected_co_access, pair$OR, pair$fdr)
  subtitle <- paste0(subtitle, "\n", sprintf(paste0(
    "%d fibers span both cCREs: both %d | CRE1 only %d | CRE2 only %d | neither %d",
    " | Fisher OR (table + 1) = %.2f, p = %.2g"), counts$n_shared,
    counts$both, counts$cre1_only, counts$cre2_only, counts$neither,
    counts$fisher_or, counts$fisher_p))
  if (display$n_drawn < counts$n_shared)
    subtitle <- paste0(subtitle, sprintf("\ndrawn: %d of %d fibers (seed %d)",
      display$n_drawn, counts$n_shared, as.integer(params$seed)))

  raster_height <- max(2, 0.025 * display$n_drawn + 0.3)
  heights <- c(pileup = 1.5, peaks = 0.3 + 0.3 * length(unique(display$peaks_df$group)),
    reads = raster_height, fire_fiberHMM = raster_height)
  read_theme <- list(axis.ticks.y = element_blank(), panel.grid.major.y = element_blank())
  fiber_plots[[id]] <- plot_region_panels(region_res,
    plot_name = paste(id, pair$gene_names, pair$CRE_pair_labels, sep = " | "),
    subtitle = subtitle, peaks_df = display$peaks_df, highlight = region_res$highlight,
    linewidth = 1.1, expand = TRUE, heights = heights,
    profile_options = list(colour_by = "group", colors = cluster_palette("pooled"),
      y_col = "frac", linewidth = 0.4, facet_scales = "fixed",
      facet_space = "fixed", require_data = TRUE),
    read_options = list(tile_width = 8, tile_height = 0.85,
      backbone_col = BACKBONE_COL, arrows = FALSE,
      base_colors = c(A = M6A_COL), require_data = TRUE),
    peaks_options = list(variant = "intervals"),
    footprint_options = list(draw_order = c("linker", "nucleosome", "FIRE element",
      rev(region_res$size_levels)), backbone = FALSE, always = TRUE,
      legend_title = NULL, legend_drop = FALSE, legend_linewidth = 3),
    theme_options = list(base = "bw", base_size = 11,
      pileup = list(strip.text.y = element_text(angle = 0)),
      reads = read_theme, fire_fiberHMM = read_theme),
    y_labels = c(pileup = "m6A fraction", reads = "m6A",
      fire_fiberHMM = "ft nucleosomes + FIRE + FiberHMM TF"),
    panel_names = c(peaks = "cres", pileup = "m6a_prop", reads = "m6a", fire_fiberHMM = "reads"))
}
```

```{r add-canonical-tss-panel}
for (id in names(pair_data)) {
  region <- pair_data[[id]]$region
  genes_here <- tss[chrom == region$chrom & end > region$start & start < region$end]
  if (nrow(genes_here)) {
    gene_plot <- ggplot(genes_here, aes(start, gene_name)) +
      geom_point(aes(shape = strand), size = 2) + theme_bw() +
      coord_cartesian(xlim = c(region$start, region$end), expand = TRUE) +
      labs(x = NULL, y = "Canonical TSS", shape = "Strand")
    fiber_plots[[id]]$combined <- gene_plot / fiber_plots[[id]]$combined +
      plot_layout(heights = c(1, 12))
  }
}
```

## Draw configuration bars using all shared fibers

`configuration_proportion_inputs()` and `plot_stacked_proportion()` come from
`plotting_functions.r`. Each sample retains its original label, and an additional
pooled bar combines all samples. Counts above the bars use all shared fibers.

```{r draw-configuration-bars}
configuration_plots <- list()
for (id in names(pair_data)) {
  labels <- copy(pair_data[[id]]$labels)[shared == TRUE]
  pooled_labels <- copy(labels)[, sample_name := "pooled"]
  bar_levels <- c("pooled", samples)
  config_inputs <- configuration_proportion_inputs(rbind(pooled_labels, labels), bar_levels)
  configuration_plots[[id]] <- plot_stacked_proportion(config_inputs$rows,
    "sample_name", "config", CONFIG_COLS,
    position = "fill", width = .7, legend = list(drop = FALSE),
    totals = list(data = config_inputs$totals, x = "sample_name", y = 1.04,
      label = "lab", style = list(size = 3)),
    scales = list(scale_y_continuous(labels = scales::percent, limits = c(0, 1.08),
      breaks = c(0, .25, .5, .75, 1))),
    labels = list(x = NULL, y = "fibers spanning both cCREs", fill = "configuration",
      title = paste(id, "all shared fibers; sample labels are metadata")),
    theme = theme_fiberseq("bw", base_size = 11,
      overrides = list(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))))
}
```

## Save and display the figures

```{r save-pair-figures}
for (id in names(pair_data)) {
  ggsave(file.path(plot_dir, paste0(id, "_reads.pdf")),
    fiber_plots[[id]]$combined, width = 12, height = 16, limitsize = FALSE)
  ggsave(file.path(plot_dir, paste0(id, "_config_bars.pdf")),
    configuration_plots[[id]], width = 13, height = 5)
}
```

```{r plot-top-five, results="asis", fig.width=12, fig.height=16}
if (length(pair_data)) {
  for (id in names(pair_data)) {
    pair <- pair_data[[id]]$pair
    cat("\n\n## ", pair$pair_id, ": ", pair$gene_names, "\n\n", sep = "")
    print(fiber_plots[[id]]$combined)
    print(configuration_plots[[id]])
  }
} else {
  cat("No fiber plots requested or no qualifying pairs.\n")
}
```

```{r verified-configuration-counts}
if (length(pair_data)) {
  verified <- rbindlist(lapply(pair_data, function(entry) entry$counts), idcol = "pair_id")
  fwrite(verified, file.path(plot_dir, "LCL_top5_verified_counts.tsv"), sep = "\t")
  knitr::kable(verified)
  cat("Every displayed pair reproduces the pooled analysis counts exactly.\n")
}
```

# Correlation triangles at the top-five pairs

Each cCRE triangle includes the FIRE-overlapping cCREs fully inside that pair's
plotting window. Rows are all primary fibers overlapping the window, before
raster downsampling. An unobserved cCRE is `NA`; an observed cCRE lacking a
qualifying FIRE is zero. Pearson correlation of binary cCRE states is the phi
coefficient. There is no additional minimum-coverage or accessibility filter.
Cells with fewer than two observations or a constant state are undefined and
shown in grey. The accompanying triangle gives the exact pairwise read counts.

The companion tiled triangle uses the union of FIRE intervals as a **continuous
fraction of each tile**, rather than the binary cCRE rule. A tile is observed
only when its entire interval is within a primary read span; this defines its
denominator, without changing the cCRE tests. Internal alignment deletions are
not resolved by the BED span inputs. All correlations describe the pooled LCL
group; donor is metadata only. The single highlighted arc is the selected cCRE
pair and carries its saved genome-wide statistics, not a test of the tile cells.

```{r triangle-matrix-functions}
pairwise_triangle_values <- function(states, features) {
  n <- crossprod(1L * !is.na(states))
  r <- suppressWarnings(cor(states, use = "pairwise.complete.obs", method = "pearson"))
  r[!is.finite(r) | n < 2] <- NA_real_
  indexes <- which(upper.tri(n), arr.ind = TRUE)
  cells <- data.table(i = indexes[, 1], j = indexes[, 2],
    feature1 = colnames(states)[indexes[, 1]], feature2 = colnames(states)[indexes[, 2]],
    correlation = r[indexes], n_shared = n[indexes])
  list(states = states, features = features, correlation = r, coverage = n, cells = cells)
}

# Draw a 45-degree triangle in feature-index space. The coordinate map below
# supplies the actual intervals; unequal cCRE spacing is not genomic distance.
triangle_panel <- function(cells, n_features, value, title, labels) {
  if (!nrow(cells)) return(ggplot() + theme_void() + labs(title = title,
    subtitle = "Fewer than two features; no off-diagonal correlation"))
  polygons <- rbindlist(lapply(seq_len(nrow(cells)), function(k) {
    a <- cells$i[k] + c(-.5, .5, .5, -.5)
    b <- cells$j[k] + c(-.5, -.5, .5, .5)
    data.table(cell = k, x = (a + b) / 2, y = (b - a) / 2,
      value = cells[[value]][k])
  }))
  p <- ggplot(polygons, aes(x, y, group = cell, fill = value)) +
    geom_polygon(colour = "white", linewidth = .15) +
    coord_fixed(xlim = c(.5, n_features + .5), ylim = c(0, n_features / 2), expand = FALSE) +
    scale_x_continuous(breaks = seq_len(n_features), labels = labels) +
    theme_bw() + theme(panel.grid = element_blank(), axis.text.y = element_blank(),
      axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(title = title, x = NULL, y = NULL)
  if (value == "correlation")
    p <- p + scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-1, 1), na.value = "grey75", name = "Pearson r")
  else p <- p + scale_fill_viridis_c(name = "Shared fibers", na.value = "grey75")
  if (n_features <= 12 && value == "n_shared")
    p <- p + geom_text(data = cells, aes(x = (i + j) / 2, y = (j - i) / 2,
      label = n_shared), inherit.aes = FALSE, size = 3)
  p
}
```

```{r build-ccre-triangles}
ccre_triangles <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  region <- entry$region
  features <- copy(cre_universe[chrom == region$chrom & start >= region$start & end <= region$end])
  setorder(features, start, end, CRE_ID)
  features[, feature_label := paste0("C", seq_len(.N))]
  states <- matrix(NA_real_, nrow(entry$res$spans), nrow(features),
    dimnames = list(entry$res$spans$key, features$CRE_ID))
  for (j in seq_len(nrow(features))) {
    cre <- as.list(features[j, .(start, end)])
    labels <- label_reads(entry$res, cre, cre, read_rule = read_rule,
      fire_overlap_fraction = fire_overlap_fraction)
    states[labels$shared, j] <- as.integer(labels$acc1[labels$shared])
  }
  triangle <- pairwise_triangle_values(states, features)
  a <- match(entry$pair$CRE1, features$CRE_ID)
  b <- match(entry$pair$CRE2, features$CRE_ID)
  stopifnot(!is.na(a), !is.na(b), triangle$coverage[a, b] == entry$counts$n_shared)
  ccre_triangles[[id]] <- triangle
  saveRDS(list(states = states, features = features,
    read_metadata = entry$res$spans[, .(key, sample_name,
      donor = sample_meta$donor[match(sample_name, sample_meta$sample_name)])]),
    file.path(plot_dir, paste0(id, "_ccre_states.rds")))
  fwrite(triangle$cells, file.path(plot_dir, paste0(id, "_ccre_correlations.tsv")), sep = "\t")
  fwrite(features, file.path(plot_dir, paste0(id, "_ccre_coordinate_map.tsv")), sep = "\t")
}
```

```{r draw-ccre-triangles, fig.width=12, fig.height=14}
for (id in names(ccre_triangles)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  triangle <- ccre_triangles[[id]]
  features <- triangle$features
  feature_labels <- paste(features$feature_label, features$CRE_label)
  correlation_plot <- triangle_panel(triangle$cells, nrow(features), "correlation",
    paste(id, "pooled LCL cCRE accessibility correlation"), feature_labels)
  count_plot <- triangle_panel(triangle$cells, nrow(features), "n_shared",
    "Exact joint-read support", feature_labels)
  marginal <- data.table(index = seq_len(nrow(features)),
    accessibility = colMeans(triangle$states, na.rm = TRUE),
    n_observed = colSums(!is.na(triangle$states)))
  marginal_plot <- ggplot(marginal, aes(index, accessibility)) + geom_col(fill = "grey45") +
    scale_x_continuous(breaks = marginal$index, labels = features$feature_label) +
    scale_y_continuous(limits = c(0, 1)) + theme_bw() +
    labs(x = "cCRE order (see coordinate map)", y = "Accessible fraction")
  a <- match(pair$CRE1, features$CRE_ID)
  b <- match(pair$CRE2, features$CRE_ID)
  arc <- data.table(x = seq(min(a, b), max(a, b), length.out = 101),
    y = sin(seq(0, pi, length.out = 101)))
  arc_plot <- ggplot(arc, aes(x, y)) + geom_line(colour = "#B2182B", linewidth = .8) +
    scale_x_continuous(limits = c(.5, nrow(features) + .5),
      breaks = seq_len(nrow(features)), labels = features$feature_label) +
    theme_void() + labs(subtitle = sprintf(
      "%s—%s | n=%d | OR(table+1)=%.3g | genome-wide FDR=%.3g",
      features$feature_label[a], features$feature_label[b], pair$n_shared_reads, pair$OR, pair$fdr))
  combined <- (correlation_plot / count_plot / arc_plot / marginal_plot) +
    plot_layout(heights = c(4, 4, 1, 1.5))
  print(combined)
  knitr::kable(features[, .(feature_label, CRE_ID, CRE_label, chrom, start, end)]) |> print()
  ggsave(file.path(plot_dir, paste0(id, "_ccre_triangle.pdf")), combined,
    width = 12, height = 14, limitsize = FALSE)
}
```

```{r build-tiled-fire-triangles}
tile_triangles <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  region <- entry$region
  sp <- entry$res$spans
  el <- entry$res$elements
  typical_span <- median(sp$end - sp$start)
  if (nrow(length_summary)) {
    pooled_span <- length_summary[sample_name == "pooled" & length_type == "primary_reference_span", median_bp]
    if (length(pooled_span) && is.finite(pooled_span)) typical_span <- pooled_span
  }
  # Descriptive resolution heuristic: about 20 tiles per typical span, with
  # at most about 80 tiles per locus. This is not a coverage/power threshold.
  target <- max(100, typical_span / 20, (region$end - region$start) / 80)
  candidates <- c(100, 250, 500, 1000, 2000, 5000)
  auto_width <- candidates[which.min(abs(log(candidates / target)))]
  tile_bp <- if (is.null(params$triangle_tile_bp)) auto_width else as.integer(params$triangle_tile_bp)
  stopifnot(length(tile_bp) == 1L, is.finite(tile_bp), tile_bp > 0)
  starts <- seq(region$start, region$end - 1, by = tile_bp)
  tiles <- data.table(chrom = region$chrom, start = starts,
    end = pmin(starts + tile_bp, region$end), CRE_ID = paste0("T", seq_along(starts)))
  values <- matrix(NA_real_, nrow(sp), nrow(tiles), dimnames = list(sp$key, tiles$CRE_ID))
  for (j in seq_len(nrow(tiles))) {
    left <- tiles$start[j]; right <- tiles$end[j]
    observed <- sp$start <= left & sp$end >= right
    values[observed, j] <- 0
    if (nrow(el)) {
      hits <- copy(el[start < right & end > left & key %in% sp$key[observed]])
      if (nrow(hits)) {
        fractions <- hits[, .(fraction = sum(IRanges::width(IRanges::reduce(
          IRanges::IRanges(pmax(start, left) + 1, pmin(end, right))))) / (right - left)), by = key]
        values[match(fractions$key, sp$key), j] <- fractions$fraction
      }
    }
  }
  tile_triangles[[id]] <- pairwise_triangle_values(values, tiles)
  tile_triangles[[id]]$tile_bp <- tile_bp
  saveRDS(tile_triangles[[id]], file.path(plot_dir, paste0(id, "_tile_states.rds")))
  fwrite(tile_triangles[[id]]$cells, file.path(plot_dir, paste0(id, "_tile_correlations.tsv")), sep = "\t")
  fwrite(tiles, file.path(plot_dir, paste0(id, "_tile_coordinate_map.tsv")), sep = "\t")
}
```

```{r draw-tiled-fire-triangles, fig.width=12, fig.height=12}
for (id in names(tile_triangles)) {
  triangle <- tile_triangles[[id]]
  tiles <- triangle$features
  labels <- format((tiles$start + tiles$end) / 2000, trim = TRUE)
  correlation_plot <- triangle_panel(triangle$cells, nrow(tiles), "correlation",
    sprintf("%s | %d bp tiles | pooled FIRE-fraction correlation", id, triangle$tile_bp), labels)
  count_plot <- triangle_panel(triangle$cells, nrow(tiles), "n_shared",
    "Fibers fully covering both tiles (tile midpoint in kb)", labels)
  combined <- correlation_plot / count_plot
  print(combined)
  ggsave(file.path(plot_dir, paste0(id, "_tile_triangle.pdf")), combined, width = 12, height = 12)
}
```

```{r top-five-statistics-and-coverage}
pair_statistics <- list()
coverage_by_sample <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  c <- entry$counts
  contingency <- matrix(c(c$neither, c$cre2_only, c$cre1_only, c$both), 2, byrow = TRUE)
  raw <- fisher.test(contingency)
  corrected <- fisher.test(contingency + 1)
  pair_statistics[[id]] <- data.table(pair_id = id, CRE_pair = pair$CRE_pair,
    neither = c$neither, CRE1_only = c$cre1_only, CRE2_only = c$cre2_only,
    both = c$both, n_shared = c$n_shared,
    observed_both = c$both, expected_both = pair$expected_co_access,
    raw_fisher_conditional_OR = unname(raw$estimate),
    raw_OR_CI_low = raw$conf.int[1], raw_OR_CI_high = raw$conf.int[2],
    raw_fisher_p = raw$p.value, raw_genomewide_FDR = pair$fdr_raw,
    corrected_crossproduct_OR = pair$OR,
    corrected_fisher_conditional_OR = unname(corrected$estimate),
    corrected_fisher_p = corrected$p.value, corrected_genomewide_FDR = pair$fdr,
    phi = ccre_triangles[[id]]$correlation[pair$CRE1, pair$CRE2])
  sp <- entry$res$spans
  sample_counts <- entry$labels[, .(overlap_locus = .N,
    jointly_observed = sum(shared)), by = sample_name]
  sample_counts <- merge(data.table(sample_name = samples), sample_counts, all.x = TRUE)
  for (column in c("overlap_locus", "jointly_observed"))
    set(sample_counts, which(is.na(sample_counts[[column]])), column, 0L)
  sample_counts[, `:=`(pair_id = id,
    donor = sample_meta$donor[match(sample_name, sample_meta$sample_name)])]
  coverage_by_sample[[id]] <- sample_counts
}
if (length(pair_statistics)) {
  statistics_table <- rbindlist(pair_statistics)
  fwrite(statistics_table, file.path(plot_dir, "LCL_top5_pair_statistics.tsv"), sep = "\t")
  fwrite(rbindlist(coverage_by_sample), file.path(plot_dir, "LCL_top5_coverage_by_sample.tsv"), sep = "\t")
  knitr::kable(statistics_table, digits = 4)
}
```

# Footprint midpoint–length plots at the same top-five pairs

Each member cCRE's **midpoint** is an anchor, in reference-genome orientation.
Shared cCREs are loaded once and linked back to every selected pair. These are
footprint midpoint–length density plots, not MNase fragment V-plots: a V shape
or secondary vertex is not assumed, and separate footprints are never joined.
The selected cCREs may contain several binding sites; their midpoints are not
necessarily TF binding centres. CpG calls do not define accessibility here.

For each anchor, the denominator is all primary fibers covering the displayed
window plus half the displayed maximum footprint length on each side. This
ensures every displayed complete footprint could have been observed. It is an
observability definition, not a QC or score filter. Densities use all eligible
fibers, not the 120-read raster subset. TF calls and nucleosome calls retain
separate densities and available-input denominators. All called TF lengths and
scores are retained; the legacy raster's selected TF size bins are not used.
Missing files are labelled unavailable, rather than counted as zero-footprint
fibers. Longer footprints remain in the exported event table.

```{r define-top-five-footprint-anchors}
anchor_links <- data.table()
anchors <- data.table()
if (length(pair_data)) {
  anchor_links <- rbindlist(lapply(names(pair_data), function(id) {
    pair <- pair_data[[id]]$pair
    data.table(pair_id = id, side = c("CRE1", "CRE2"), chrom = pair$chrom,
      CRE_ID = c(pair$CRE1, pair$CRE2), start = c(pair$cre1_start, pair$cre2_start),
      end = c(pair$cre1_end, pair$cre2_end))
  }))
  anchors <- unique(anchor_links, by = c("chrom", "start", "end", "CRE_ID"))
  anchors[, `:=`(anchor_id = sprintf("anchor%02d", seq_len(.N)),
    centre = (start + end) / 2, origin_pair = pair_id)]
  anchor_links <- merge(anchor_links, anchors[, .(chrom, start, end, CRE_ID, anchor_id, centre)],
    by = c("chrom", "start", "end", "CRE_ID"), sort = FALSE)
  fwrite(anchors, file.path(plot_dir, "LCL_top5_footprint_anchors.tsv"), sep = "\t")
  fwrite(anchor_links, file.path(plot_dir, "LCL_top5_pair_anchor_links.tsv"), sep = "\t")
}
```

```{r footprint-block-loader}
load_anchor_footprints <- function(anchor, spans, region) {
  sources <- c("FiberHMM_TF", "ft_nucleosome")
  events <- list(); availability <- list()
  query <- GenomicRanges::GRanges(region$chrom, IRanges::IRanges(region$start + 1, region$end))
  for (sample in samples) {
    sample_spans <- spans[sample_name == sample]
    for (track_source in sources) {
      path <- if (track_source == "FiberHMM_TF")
        file.path(paths$tf_root, sample, paste0(sample, ".recalled_tf.bed.gz")) else
        file.path(paths$ft_root, sample, "extracted_results", "nuc_by_chr",
          paste0(sample, ".ft_extracted_nuc.", region$chrom, ".bed.gz"))
      available <- file.exists(path) && file.exists(paste0(path, ".tbi"))
      availability[[length(availability) + 1L]] <- data.table(anchor_id = anchor$anchor_id,
        sample_name = sample, source = track_source, path = path, available = available,
        n_window_spanning = nrow(sample_spans),
        n_eligible = if (available) nrow(sample_spans) else 0L)
      if (!available || !nrow(sample_spans)) next
      bed <- read_tabix_region(path, query)
      if (!nrow(bed)) next
      bed <- bed[paste(sample, bed[[4]]) %in% sample_spans$key, , drop = FALSE]
      if (!nrow(bed)) next
      format <- if (track_source == "FiberHMM_TF") params$tf_format else "bed12_fibertools"
      blocks <- as.data.table(convert_ft_bed12_to_bed6(bed, format = format,
        keep_block_scores = track_source == "FiberHMM_TF", source = path))
      if (!nrow(blocks)) next
      blocks[, key := paste(sample, RID)]
      blocks <- merge(blocks, sample_spans[, .(key, span_start = start, span_end = end)], by = "key")
      # Match the primary alignment used for the denominator, not a different
      # alignment of the same RID. There is no new footprint-score cutoff.
      blocks <- blocks[end > start & start >= span_start & end <= span_end]
      if (!nrow(blocks)) next
      blocks[, `:=`(anchor_id = anchor$anchor_id, sample_name = sample, source = track_source,
        length_bp = end - start, midpoint = (start + end) / 2,
        offset_bp = (start + end) / 2 - anchor$centre)]
      if (!"block_score" %in% names(blocks)) blocks[, block_score := NA_real_]
      blocks <- unique(blocks[, .(anchor_id, sample_name, source, key, RID,
        start, end, length_bp, midpoint, offset_bp, block_score)],
        by = c("anchor_id", "source", "key", "start", "end"))
      events[[length(events) + 1L]] <- blocks
    }
  }
  empty_events <- data.table(anchor_id = character(), sample_name = character(), source = character(),
    key_id = character(), RID = character(), start = numeric(), end = numeric(),
    length_bp = numeric(), midpoint = numeric(), offset_bp = numeric(), block_score = numeric())
  setnames(empty_events, "key_id", "key")
  list(events = if (length(events)) rbindlist(events) else empty_events,
    availability = rbindlist(availability))
}
```

```{r load-all-footprints-at-anchors}
footprint_data <- list()
if (nrow(anchors)) {
  for (i in seq_len(nrow(anchors))) {
    anchor <- anchors[i]
    padding <- params$footprint_flank_bp + params$footprint_max_length_bp / 2
    region <- list(chrom = anchor$chrom,
      start = max(0, floor(anchor$centre - padding)), end = ceiling(anchor$centre + padding))
    res <- pair_data[[anchor$origin_pair]]$res
    if (region$start < res$region$start || region$end > res$region$end)
      res <- load_region(region, samples, paths$root, paths$fire_root,
        fire_ver = paths$fire_ver, tabix_bin = paths$tabix_bin, span_source = "read_spans")
    eligible_spans <- copy(res$spans[start <= region$start & end >= region$end])
    footprint_data[[anchor$anchor_id]] <- load_anchor_footprints(anchor, eligible_spans, region)
    footprint_data[[anchor$anchor_id]]$region <- region
  }
}
footprint_events <- rbindlist(lapply(footprint_data, function(x) x$events), fill = TRUE)
footprint_availability <- rbindlist(lapply(footprint_data, function(x) x$availability), fill = TRUE)
if (length(footprint_data)) {
  footprint_events[, donor := sample_meta$donor[match(sample_name, sample_meta$sample_name)]]
  footprint_availability[, donor := sample_meta$donor[match(sample_name, sample_meta$sample_name)]]
  fwrite(footprint_events, file.path(plot_dir, "LCL_top5_footprint_events.tsv.gz"), sep = "\t")
  fwrite(footprint_availability, file.path(plot_dir, "LCL_top5_footprint_input_coverage.tsv"), sep = "\t")
}
```

```{r bin-footprint-midpoints-and-lengths}
footprint_bins <- list()
flank <- params$footprint_flank_bp
max_length <- params$footprint_max_length_bp
bin_bp <- params$footprint_bin_bp
stopifnot((2 * flank) %% bin_bp == 0, max_length %% bin_bp == 0)
for (id in names(footprint_data)) {
  input <- footprint_data[[id]]
  support <- input$availability[, .(n_eligible = sum(n_eligible),
    n_available_samples = sum(available), n_missing_samples = sum(!available)), by = source]
  events <- copy(input$events[offset_bp >= -flank & offset_bp < flank & length_bp <= max_length])
  events[, `:=`(x_bin = floor((offset_bp + flank) / bin_bp),
    y_bin = pmin(floor(length_bp / bin_bp), max_length / bin_bp - 1))]
  counts <- events[, .(n_footprints = .N), by = .(source, x_bin, y_bin)]
  grid <- CJ(source = c("FiberHMM_TF", "ft_nucleosome"),
    x_bin = seq_len(2 * flank / bin_bp) - 1, y_bin = seq_len(max_length / bin_bp) - 1)
  grid <- merge(grid, counts, by = c("source", "x_bin", "y_bin"), all.x = TRUE)
  grid[is.na(n_footprints), n_footprints := 0L]
  grid <- merge(grid, support, by = "source", all.x = TRUE)
  grid[, `:=`(anchor_id = id, offset_bp = -flank + (x_bin + .5) * bin_bp,
    length_bp = (y_bin + .5) * bin_bp,
    calls_per_100_fibers = fifelse(n_eligible > 0, 100 * n_footprints / n_eligible, NA_real_))]
  footprint_bins[[id]] <- grid
}
if (length(footprint_bins))
  fwrite(rbindlist(footprint_bins), file.path(plot_dir, "LCL_top5_footprint_density_bins.tsv.gz"), sep = "\t")
```

```{r draw-footprint-midpoint-length-plots, fig.width=12, fig.height=10}
for (pair_id in names(pair_data)) {
  # Use an explicit external variable to avoid data.table's column precedence.
  current_pair <- pair_id
  links <- anchor_links[pair_id == current_pair]
  grid <- merge(rbindlist(footprint_bins[links$anchor_id]),
    links[, .(anchor_id, side, CRE_ID, centre)], by = "anchor_id")
  grid[, anchor_label := paste(side, CRE_ID, paste0("midpoint ", centre))]
  grid[, source_label := source]
  support_labels <- unique(grid[, .(anchor_label, source_label, n_eligible, n_available_samples)])
  density_plot <- ggplot(grid, aes(offset_bp, length_bp, fill = calls_per_100_fibers)) +
    geom_tile(width = bin_bp, height = bin_bp) +
    geom_vline(xintercept = 0, colour = "grey50", linewidth = .25) +
    geom_text(data = support_labels, aes(x = -flank + bin_bp, y = max_length - bin_bp,
      label = paste0("n=", n_eligible, "; files ", n_available_samples, "/", length(samples))),
      inherit.aes = FALSE, hjust = 0, vjust = 1, size = 3) +
    facet_grid(source_label ~ anchor_label) +
    scale_fill_viridis_c(name = "Calls / 100\neligible fibers", na.value = "grey80") +
    coord_cartesian(xlim = c(-flank, flank), ylim = c(0, max_length), expand = FALSE) +
    theme_bw() + labs(x = "Footprint midpoint relative to cCRE midpoint (bp)",
      y = "Called footprint length (bp)", title = paste(current_pair, "pooled LCL footprints"),
      subtitle = "cCRE midpoint anchors; all eligible fibers; TF and nucleosome calls shown separately")
  print(density_plot)
  ggsave(file.path(plot_dir, paste0(current_pair, "_footprint_midpoint_length.pdf")),
    density_plot, width = 12, height = 10)

  # Fixed length slices permit comparison to the reference figures without
  # treating the protected intervals as MNase fragments.
  short_plot <- density_plot + coord_cartesian(xlim = c(-flank, flank), ylim = c(0, 100), expand = FALSE)
  nuc_plot <- density_plot + coord_cartesian(xlim = c(-flank, flank), ylim = c(100, 200), expand = FALSE)
  ggsave(file.path(plot_dir, paste0(current_pair, "_footprint_length_slices.pdf")),
    short_plot / nuc_plot, width = 12, height = 14)
}
```

```{r footprint-length-side-profiles, fig.width=10, fig.height=6}
side_profiles <- list()
for (id in names(footprint_data)) {
  input <- footprint_data[[id]]
  support <- input$availability[, .(n_eligible = sum(n_eligible)), by = source]
  # The central +/- one histogram bin is explicit, rather than claiming
  # 2 bp boundary resolution from the footprint calls.
  central <- input$events[abs(offset_bp) <= bin_bp & length_bp <= max_length]
  counts <- central[, .(n_footprints = .N),
    by = .(source, length_bin = pmin(floor(length_bp / bin_bp), max_length / bin_bp - 1))]
  grid <- CJ(source = c("FiberHMM_TF", "ft_nucleosome"), length_bin = seq_len(max_length / bin_bp) - 1)
  grid <- merge(grid, counts, by = c("source", "length_bin"), all.x = TRUE)
  grid[is.na(n_footprints), n_footprints := 0L]
  grid <- merge(grid, support, by = "source")
  grid[, `:=`(anchor_id = id, length_bp = (length_bin + .5) * bin_bp,
    calls_per_100_fibers = fifelse(n_eligible > 0, 100 * n_footprints / n_eligible, NA_real_))]
  side_profiles[[id]] <- grid
}
if (length(side_profiles)) {
  profiles <- merge(rbindlist(side_profiles), anchors[, .(anchor_id, CRE_ID)], by = "anchor_id")
  p <- ggplot(profiles, aes(calls_per_100_fibers, length_bp, colour = source)) +
    geom_path() + facet_wrap(~CRE_ID) + theme_bw() +
    labs(x = "Calls / 100 eligible fibers", y = "Footprint length (bp)",
      title = sprintf("Length profiles within +/- %g bp of each cCRE midpoint", bin_bp))
  print(p)
  ggsave(file.path(plot_dir, "LCL_top5_footprint_length_profiles.pdf"), p, width = 12, height = 8)
  fwrite(profiles, file.path(plot_dir, "LCL_top5_footprint_length_profiles.tsv"), sep = "\t")
}
```

```{r anchor-sequence-resolution-summary}
# Descriptive sequence context only: reference A/T spacing is a potential
# methylation-site spacing across both strands, not measured boundary precision.
sequence_resolution <- list()
if (nrow(anchors) && file.exists(params$reference_fasta) &&
    file.exists(paste0(params$reference_fasta, ".fai"))) {
  for (i in seq_len(nrow(anchors))) {
    anchor <- anchors[i]
    left <- max(0, floor(anchor$centre - params$footprint_flank_bp))
    right <- ceiling(anchor$centre + params$footprint_flank_bp)
    region <- GenomicRanges::GRanges(anchor$chrom, IRanges::IRanges(left + 1, right))
    sequence <- as.character(Rsamtools::scanFa(params$reference_fasta, param = region)[[1]])
    bases <- strsplit(toupper(sequence), "", fixed = TRUE)[[1]]
    sites <- which(bases %in% c("A", "T"))
    gaps <- diff(sites)
    sequence_resolution[[anchor$anchor_id]] <- data.table(anchor_id = anchor$anchor_id,
      genome_build = params$genome_build, chrom = anchor$chrom, start = left, end = right,
      n_potential_AT_sites = length(sites),
      median_AT_gap_bp = if (length(gaps)) median(gaps) else NA_real_,
      max_AT_gap_bp = if (length(gaps)) max(gaps) else NA_real_,
      plot_bin_bp = bin_bp)
  }
  sequence_table <- rbindlist(sequence_resolution)
  fwrite(sequence_table, file.path(plot_dir, "LCL_top5_anchor_sequence_spacing.tsv"), sep = "\t")
  knitr::kable(sequence_table)
}
```

A broad or secondary density peak describes the called intervals at these cCRE
anchors. It is not evidence by itself for a co-bound particle or cooperative
binding. There is no MNase arm projection or artificial joining of footprints.
The existing 2×2 test measures cCRE co-accessibility, not TF co-occupancy.

```{r figure-parameter-record}
jsonlite::write_json(list(params = params, n_samples = length(samples),
  group = "pooled_LCL", donor_used_in_testing = FALSE, added_qc_filters = FALSE,
  ccre_read_rule = read_rule, ccre_fire_overlap_fraction = fire_overlap_fraction,
  tile_observability = "entire tile within primary reference span",
  footprint_anchor = "each selected cCRE midpoint; reference orientation",
  footprint_denominator = "primary fibers covering flank + half displayed maximum length; available input files",
  triangle_tile_bp = lapply(tile_triangles, function(x) x$tile_bp)),
  file.path(plot_dir, "LCL_figure_parameters.json"), auto_unbox = TRUE, pretty = TRUE, null = "null")
```

```{r session-information}
sessionInfo()
```
