---
title: "Genome-wide LCL co-accessibility: top five pairs"
author: "Crystal Shan"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  root: "/project/spott/cshan/fiber-seq/LCL_project/co-accessibility"
  sample_sheet: "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv"
  fire_root: "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
  tf_root: "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/FiberHMM_v2/results/results_v1_trained_model"
  tabix: "/project/spott/cshan/envs/dimelo/bin/tabix"
  code_root: "/project/spott/cshan/fiber-seq/code"
  tss_bed: "/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed"
  locus_flank_bp: 1000
  make_plots: true
  max_plot_reads: 120
  seed: 20261009
  allow_partial_results: false
---

This notebook reads the completed **genome-wide, pooled 31-sample LCL analysis**
from `LCL_coaccess_cres.py`. Run `LCL_fire_universe.sh`, `LCL_read_spans.sh`, and
`LCL_run_coaccess.sh` first. Donor labels are metadata only; there are no donor
strata, haplotype tests, or timepoints.

The plotting workflow is written directly below in separate chunks. Source the
shared parsing and plotting functions, then run the chunks in order, or knit this
notebook. `LCL_plot_coaccess.sh` is an optional SLURM launcher for the same notebook.

Candidate cCREs overlap the union of all 31 samples' FIRE peaks. Pairs share a
canonical TSS ±10 kb window and have **500 < gap < 20,000 bp**. A shared fiber
has aligned bases at both cCREs (at least one base each). It is accessible at a
cCRE when **one FIRE element on that same sample/read, intersected with the
read's aligned blocks, overlaps at least 50% of the cCRE length**. CIGAR deletions
and skipped reference regions are unobserved, not closed. Aligned pieces of one
element can be summed; separate FIRE elements cannot. The fraction is relative
to cCRE length, not to FIRE-element length.

The four configuration counts are pooled across samples before testing. The
primary FDR comes from the macrophage-compatible **two-sided Fisher test after
adding one pseudocount to each pooled cell**, corrected across all tested unique
pairs genome-wide. Positive pairs require `fdr < 0.05`, `OR > 1`, and observed
joint accessibility greater than its pooled independence expectation. This
notebook selects the top five by FDR, decreasing OR, then decreasing shared-fiber
coverage, using pair ID to resolve ties. It does not recalculate FDR after selection.

# Setup

```{r setup, include=FALSE}
# Use a portable locale for tabix and other child processes on compute nodes.
Sys.setenv(LC_ALL = "C")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                     fig.width = 11, fig.height = 6, cache = FALSE)
```

```{r packages-and-functions}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
})
source(file.path(params$code_root, "parsing_functions", "parsing_footprints_functions.r"), local = TRUE)
source(file.path(params$code_root, "parsing_functions", "plotting_functions.r"), local = TRUE)
options(scipen = 999)
```

```{r settings}
root <- params$root
coaccess_dir <- file.path(root, "coaccess")
universe_dir <- file.path(root, "universe")
plot_dir <- file.path(coaccess_dir, "plots", "top5")
read_rule <- "any"
fire_overlap_fraction <- 0.5
fdr_cutoff <- 0.05
max_plot_reads <- as.integer(params$max_plot_reads)
stopifnot(length(max_plot_reads) == 1L, !is.na(max_plot_reads), max_plot_reads > 0L)
paths <- list(root = root, fire_root = params$fire_root,
              ft_root = params$fire_root, tf_root = params$tf_root,
              tabix_bin = params$tabix, fire_ver = "v0.1")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
stopifnot(params$locus_flank_bp >= 0)
```

# Input provenance

```{r validate-all-31-samples}
sample_meta <- fread(params$sample_sheet)
manifest <- fread(file.path(universe_dir, "sample_manifest.tsv"))
stopifnot(all(c("sample_name", "cell_line") %in% names(sample_meta)),
          all(c("sample_name", "donor") %in% names(manifest)),
          nrow(sample_meta) == 31L, nrow(manifest) == 31L,
          !anyDuplicated(sample_meta$sample_name), !anyDuplicated(manifest$sample_name),
          setequal(sample_meta$sample_name, manifest$sample_name))
sample_meta[, donor := as.character(cell_line)]
samples <- as.character(manifest$sample_name)
stopifnot(identical(as.character(manifest$donor),
                    sample_meta$donor[match(samples, sample_meta$sample_name)]))
knitr::kable(sample_meta[, .(sample_name, donor)])
```

```{r validate-completed-run}
# Do not combine old pair tables with a changed input manifest.
run_manifest <- fread(file.path(coaccess_dir, "sample_manifest.tsv"))
provenance_columns <- c("sample_name", "donor", "fire_dir", "cram_path", "elements_path", "spans_path")
stopifnot(all(provenance_columns %in% names(run_manifest)),
          all(provenance_columns %in% names(manifest)),
          nrow(run_manifest) == 31L, !anyDuplicated(run_manifest$sample_name),
          setequal(run_manifest$sample_name, samples))
run_manifest <- run_manifest[match(samples, sample_name)]
for (column in provenance_columns)
  if (!identical(as.character(run_manifest[[column]]), as.character(manifest[[column]])))
    stop("Universe and completed-run manifests differ in ", column, "; rerun the analysis.")
same_paths <- function(a, b)
  identical(normalizePath(a, mustWork = FALSE), normalizePath(b, mustWork = FALSE))
stopifnot("blocks_path" %in% names(run_manifest),
  same_paths(run_manifest$blocks_path,
  file.path(root, samples, paste0(samples, ".aligned_blocks.bed.gz"))),
  same_paths(run_manifest$elements_path,
    file.path(paths$fire_root, samples, "additional-outputs-v0.1", "fire-peaks",
              paste0(samples, "-v0.1-fire-elements.bed.gz"))))
```

```{r read-genome-wide-pair-results}
pairs_file <- file.path(coaccess_dir, "LCL_coaccess_pairs.tsv.gz")
pairs_dt <- fread(pairs_file)
required_columns <- c("CRE_pair", "CRE1", "CRE2", "chrom", "cre1_start", "cre1_end",
  "cre2_start", "cre2_end", "gene_names", "CRE_pair_labels", "dist", "n_shared_reads",
  "co_access", "co_closed", "CRE1_access", "CRE2_access", "OR", "fdr", "fdr_raw",
  "expected_co_access", "observed_expected_ratio", "excess_fraction", "read_rule",
  "fire_overlap_fraction", "dataset", "analysis_scope", "coverage_mode")
if (!all(required_columns %in% names(pairs_dt)))
  stop("Missing result columns: ", paste(setdiff(required_columns, names(pairs_dt)), collapse = ", "))
stopifnot(!anyDuplicated(pairs_dt$CRE_pair),
          all(pairs_dt$dataset == "LCL"),
          all(pairs_dt$read_rule == read_rule),
          all(pairs_dt$coverage_mode == "aligned_blocks_v1"),
          all(pairs_dt$fire_overlap_fraction == fire_overlap_fraction),
          all(pairs_dt$dist > 500 & pairs_dt$dist < 20000),
          all(pairs_dt$n_shared_reads > 0),
          all(pairs_dt$n_shared_reads == pairs_dt$co_access + pairs_dt$co_closed +
                pairs_dt$CRE1_access + pairs_dt$CRE2_access))
if (!isTRUE(params$allow_partial_results) && any(pairs_dt$analysis_scope != "genome-wide"))
  stop("Partial chromosome results supplied. Run the genome-wide analysis before rendering.")
expected_from_counts <- with(pairs_dt,
  (co_access + CRE1_access) * (co_access + CRE2_access) / n_shared_reads)
stopifnot(isTRUE(all.equal(as.numeric(pairs_dt$expected_co_access), as.numeric(expected_from_counts),
                          tolerance = 1e-8, check.attributes = FALSE)))
cat(nrow(pairs_dt), "unique tested pairs loaded.\n")
```

```{r read-ccre-and-gene-annotations}
cre_universe <- fread(file.path(universe_dir, "cre_universe.bed"), header = FALSE,
  col.names = c("chrom", "start", "end", "CRE_ID", "CRE_label"))
tss_bed <- params$tss_bed
tss <- fread(tss_bed, header = FALSE, select = 1:6,
  col.names = c("chrom", "start", "end", "annotation", "score", "strand"))
tss[, gene_name := tstrsplit(annotation, ";", fixed = TRUE, keep = 3L)[[1L]]]
```

# Top five co-accessible pairs

```{r select-top-five}
eligible <- pairs_dt[!is.na(fdr) & fdr < fdr_cutoff & !is.na(OR) & OR > 1 &
                       co_access > expected_co_access]
setorderv(eligible, c("fdr", "OR", "n_shared_reads", "CRE_pair"), c(1L, -1L, -1L, 1L))
selected <- head(eligible, 5L)
selected[, pair_id := sprintf("LCL_top%02d", seq_len(.N))]
fwrite(selected, file.path(plot_dir, "LCL_top5_pairs.tsv"), sep = "\t")
cat(nrow(eligible), "pairs pass the positive-association criteria; selected",
    nrow(selected), "for plotting.\n")
if (nrow(selected) < 5L)
  cat("Fewer than five pairs qualify; only qualifying pairs will be shown.\n")
knitr::kable(selected[, .(pair_id, gene_names, CRE_pair, dist, n_shared_reads,
  co_access, expected_co_access, observed_expected_ratio, OR, fdr, fdr_raw)], digits = 4)
```

```{r observed-versus-expected, fig.height=5}
if (nrow(selected)) {
  oe <- melt(selected[, .(pair_id, observed = co_access, expected = expected_co_access)],
             id.vars = "pair_id", variable.name = "count_type", value.name = "fibers")
  oe_plot <- ggplot(oe, aes(pair_id, fibers, fill = count_type)) +
    geom_col(position = "dodge") + theme_bw() +
    scale_fill_manual(values = c(observed = "#B2182B", expected = "grey70")) +
    labs(x = NULL, y = "Jointly accessible fibers", fill = NULL,
         title = "Observed and expected co-accessibility",
         subtitle = "Expectation uses pooled margins among the same jointly covered fibers")
  print(oe_plot)
  ggsave(file.path(plot_dir, "LCL_top5_observed_expected.pdf"), oe_plot, width = 9, height = 5)
}
```

# Fiber plots and count checks

The following code reloads **all 31 samples** from the same primary aligned-block and
FIRE-element files used for counting. The complete reloaded 2×2 must match the
saved result; a discrepancy stops rendering. The raster shows at most
`r max_plot_reads` randomly selected shared fibers per pair; the m6A profile uses
those displayed fibers. Configuration bars, count checks, and statistical
annotations always use all shared fibers.

m6A, nucleosome, and TF-footprint tracks are display annotations restricted to
those same read identities. Missing optional display tracks produce warnings;
they never change the FIRE accessibility call. Donor identifiers are attached to
the exported read labels and sample contributions without entering a test.

```{r validate-plot-inputs}
if (isTRUE(params$make_plots) && nrow(selected)) {
  spans <- file.path(root, samples, paste0(samples, ".aligned_blocks.bed.gz"))
  elements <- file.path(paths$fire_root, samples, "additional-outputs-v0.1", "fire-peaks",
                         paste0(samples, "-v0.1-fire-elements.bed.gz"))
  needed <- c(paths$tabix_bin, spans, paste0(spans, ".tbi"), elements, paste0(elements, ".tbi"))
  if (any(!file.exists(needed)))
    stop("Missing plotting inputs: ", paste(needed[!file.exists(needed)], collapse = "\n"))
  for (i in seq_along(samples)) {
    marker <- paste0(spans[i], ".source.tsv")
    if (!file.exists(marker) || !identical(readLines(marker, n = 2L),
        c(run_manifest$cram_path[i], "aligned_blocks_v1")))
      stop("Missing/current aligned-block provenance required for ", samples[i])
    if (file.info(spans[i])$mtime > file.info(pairs_file)$mtime ||
        file.info(elements[i])$mtime > file.info(pairs_file)$mtime)
      stop("Plot inputs changed after counting; rerun the analysis for ", samples[i])
  }
}
```

## Load fibers and assign accessibility

`load_region()`, `label_reads()`, and `coaccess_2x2()` come from
`parsing_footprints_functions.r`. Read keys include the sample name so that equal
read names from different samples cannot be joined. Keep these complete labels
for count verification, configuration bars, and metadata exports.

```{r load-pair-fibers}
pair_data <- list()
if (isTRUE(params$make_plots) && nrow(selected)) {
  for (i in seq_len(nrow(selected))) {
    pair <- selected[i]
    region <- list(chrom = pair$chrom,
      start = max(0, min(pair$cre1_start, pair$cre2_start) - params$locus_flank_bp),
      end = max(pair$cre1_end, pair$cre2_end) + params$locus_flank_bp)
    cre1 <- list(start = pair$cre1_start, end = pair$cre1_end)
    cre2 <- list(start = pair$cre2_start, end = pair$cre2_end)
    res <- load_region(region, samples, paths$root, paths$fire_root,
      fire_ver = paths$fire_ver, tabix_bin = paths$tabix_bin,
      span_source = "aligned_blocks")
    labels <- label_reads(res, cre1, cre2, read_rule = read_rule,
      fire_overlap_fraction = fire_overlap_fraction)
    pair_data[[pair$pair_id]] <- list(pair = pair, region = region,
      cre1 = cre1, cre2 = cre2, res = res, labels = labels,
      counts = coaccess_2x2(labels))
  }
}
```

```{r verify-reloaded-counts}
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  actual <- unlist(entry$counts[1, c("neither", "cre1_only", "cre2_only", "both", "n_shared")],
                   use.names = FALSE)
  expected <- unlist(pair[, .(co_closed, CRE1_access, CRE2_access, co_access, n_shared_reads)],
                     use.names = FALSE)
  if (!identical(as.numeric(actual), as.numeric(expected)))
    stop("Reloaded counts differ from the analysis table for ", pair$CRE_pair,
         "; reloaded: ", paste(actual, collapse = ","),
         "; saved: ", paste(expected, collapse = ","))
}
```

```{r export-fiber-metadata}
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  read_labels <- copy(entry$labels)[shared == TRUE]
  read_labels[, donor := sample_meta$donor[match(sample_name, sample_meta$sample_name)]]
  read_labels[, `:=`(pair_id = pair$pair_id, CRE_pair = pair$CRE_pair)]
  fwrite(read_labels, file.path(plot_dir, paste0(pair$pair_id, "_read_labels.tsv.gz")), sep = "\t")
  contributions <- read_labels[, .(n_shared = .N, co_access = sum(acc1 & acc2),
    CRE1_access = sum(acc1 & !acc2), CRE2_access = sum(!acc1 & acc2),
    co_closed = sum(!acc1 & !acc2)), by = .(pair_id, sample_name, donor)]
  fwrite(contributions, file.path(plot_dir, paste0(pair$pair_id, "_sample_counts.tsv")), sep = "\t")
}
```

## Prepare the displayed fibers and annotation tracks

Sample up to `r max_plot_reads` shared fibers with a fixed seed. Load m6A,
ft nucleosomes, and FiberHMM TF footprints only for these display keys using
`load_ft_tracks()` from the parsing functions. Copies of the display tables are
labelled `pooled`; the original sample/read keys and complete labels are retained.
Read backbones, FIRE, nucleosome and TF intervals are broken at alignment gaps;
the m6A denominator counts only fibers with an aligned base at that position.

```{r prepare-display-fibers}
display_data <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  drawn_labels <- copy(entry$labels)[shared == TRUE]
  set.seed(as.integer(params$seed))
  if (nrow(drawn_labels) > max_plot_reads) {
    drawn_keys <- sample(drawn_labels$key, max_plot_reads)
    drawn_labels <- drawn_labels[key %in% drawn_keys]
  }

  display_res <- load_ft_tracks(copy(entry$res), keys = drawn_labels$key,
    ft_root = paths$ft_root, tabix_bin = paths$tabix_bin,
    nuc_source = "ft", tf_source = "recalled_tf", tf_root = paths$tf_root)
  for (track in c("spans", "blocks", "elements", "m6a", "nuc", "size_fps")) {
    if (!is.null(display_res[[track]]) && nrow(display_res[[track]]))
      display_res[[track]][, sample_name := "pooled"]
  }
  display_res$samples <- "pooled"
  drawn_labels[, sample_name := "pooled"]

  region <- entry$region
  cres_here <- cre_universe[chrom == region$chrom & end > region$start & start < region$end]
  region_res <- coaccess_region_result(display_res, entry$cre1, entry$cre2,
    drawn_labels, cres = cres_here)

  # Keep the cCRE class order used by the shared plotting functions.
  pk <- as.data.frame(cres_here)
  cls <- factor(pk$CRE_label, levels = c(intersect(CCRE_CLASSES, pk$CRE_label),
    setdiff(unique(pk$CRE_label), CCRE_CLASSES)))
  pk <- pk[order(cls, pk$start), , drop = FALSE]
  peaks_df <- data.frame(chr = pk$chrom, start = pk$start, end = pk$end,
    group = as.character(pk$CRE_label))
  display_data[[id]] <- list(region_res = region_res, peaks_df = peaks_df,
    n_drawn = nrow(drawn_labels))
}
```

## Draw the fiber panels

`coaccess_region_result()` above adapts the display data for
`plot_region_panels()` from `plotting_functions.r`. Panel choices, colours,
relative heights, titles, and the highlighted pair are set explicitly here.

```{r draw-fiber-panels}
fiber_plots <- list()
for (id in names(pair_data)) {
  entry <- pair_data[[id]]
  pair <- entry$pair
  region <- entry$region
  counts <- entry$counts
  display <- display_data[[id]]
  region_res <- display$region_res

  subtitle <- sprintf(paste0("%s:%d-%d | gap %d bp | O %.0f / E %.2f | ",
    "OR %.3g | genome-wide FDR %.3g"), region$chrom, region$start, region$end,
    pair$dist, pair$co_access, pair$expected_co_access, pair$OR, pair$fdr)
  subtitle <- paste0(subtitle, "\n", sprintf(paste0(
    "%d fibers have aligned bases at both cCREs: both %d | CRE1 only %d | CRE2 only %d | neither %d",
    " | Fisher OR (table + 1) = %.2f, p = %.2g"), counts$n_shared,
    counts$both, counts$cre1_only, counts$cre2_only, counts$neither,
    counts$fisher_or, counts$fisher_p))
  if (display$n_drawn < counts$n_shared)
    subtitle <- paste0(subtitle, sprintf("\ndrawn: %d of %d fibers (seed %d)",
      display$n_drawn, counts$n_shared, as.integer(params$seed)))

  raster_height <- max(2, 0.025 * display$n_drawn + 0.3)
  heights <- c(pileup = 1.5, peaks = 0.3 + 0.3 * length(unique(display$peaks_df$group)),
    reads = raster_height, fire_fiberHMM = raster_height)
  read_theme <- list(axis.ticks.y = element_blank(), panel.grid.major.y = element_blank())
  fiber_plots[[id]] <- plot_region_panels(region_res,
    plot_name = paste(id, pair$gene_names, pair$CRE_pair_labels, sep = " | "),
    subtitle = subtitle, peaks_df = display$peaks_df, highlight = region_res$highlight,
    linewidth = 1.1, expand = TRUE, heights = heights,
    profile_options = list(colour_by = "group", colors = cluster_palette("pooled"),
      y_col = "frac", linewidth = 0.4, facet_scales = "fixed",
      facet_space = "fixed", require_data = TRUE),
    read_options = list(tile_width = 8, tile_height = 0.85,
      backbone_col = BACKBONE_COL, arrows = FALSE,
      base_colors = c(A = M6A_COL), require_data = TRUE),
    peaks_options = list(variant = "intervals"),
    footprint_options = list(draw_order = c("linker", "nucleosome", "FIRE element",
      rev(region_res$size_levels)), backbone = FALSE, always = TRUE,
      legend_title = NULL, legend_drop = FALSE, legend_linewidth = 3),
    theme_options = list(base = "bw", base_size = 11,
      pileup = list(strip.text.y = element_text(angle = 0)),
      reads = read_theme, fire_fiberHMM = read_theme),
    y_labels = c(pileup = "m6A fraction", reads = "m6A",
      fire_fiberHMM = "ft nucleosomes + FIRE + FiberHMM TF"),
    panel_names = c(peaks = "cres", pileup = "m6a_prop", reads = "m6a", fire_fiberHMM = "reads"))
}
```

```{r add-canonical-tss-panel}
for (id in names(pair_data)) {
  region <- pair_data[[id]]$region
  genes_here <- tss[chrom == region$chrom & end > region$start & start < region$end]
  if (nrow(genes_here)) {
    gene_plot <- ggplot(genes_here, aes(start, gene_name)) +
      geom_point(aes(shape = strand), size = 2) + theme_bw() +
      coord_cartesian(xlim = c(region$start, region$end), expand = TRUE) +
      labs(x = NULL, y = "Canonical TSS", shape = "Strand")
    fiber_plots[[id]]$combined <- gene_plot / fiber_plots[[id]]$combined +
      plot_layout(heights = c(1, 12))
  }
}
```

## Draw configuration bars using all shared fibers

`configuration_proportion_inputs()` and `plot_stacked_proportion()` come from
`plotting_functions.r`. Each sample retains its original label, and an additional
pooled bar combines all samples. Counts above the bars use all shared fibers.

```{r draw-configuration-bars}
configuration_plots <- list()
for (id in names(pair_data)) {
  labels <- copy(pair_data[[id]]$labels)[shared == TRUE]
  pooled_labels <- copy(labels)[, sample_name := "pooled"]
  bar_levels <- c("pooled", samples)
  config_inputs <- configuration_proportion_inputs(rbind(pooled_labels, labels), bar_levels)
  configuration_plots[[id]] <- plot_stacked_proportion(config_inputs$rows,
    "sample_name", "config", CONFIG_COLS,
    position = "fill", width = .7, legend = list(drop = FALSE),
    totals = list(data = config_inputs$totals, x = "sample_name", y = 1.04,
      label = "lab", style = list(size = 3)),
    scales = list(scale_y_continuous(labels = scales::percent, limits = c(0, 1.08),
      breaks = c(0, .25, .5, .75, 1))),
    labels = list(x = NULL, y = "fibers spanning both cCREs", fill = "configuration",
      title = paste(id, "all shared fibers; sample labels are metadata")),
    theme = theme_fiberseq("bw", base_size = 11,
      overrides = list(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))))
}
```

## Save and display the figures

```{r save-pair-figures}
for (id in names(pair_data)) {
  ggsave(file.path(plot_dir, paste0(id, "_reads.pdf")),
    fiber_plots[[id]]$combined, width = 12, height = 16, limitsize = FALSE)
  ggsave(file.path(plot_dir, paste0(id, "_config_bars.pdf")),
    configuration_plots[[id]], width = 13, height = 5)
}
```

```{r plot-top-five, results="asis", fig.width=12, fig.height=16}
if (length(pair_data)) {
  for (id in names(pair_data)) {
    pair <- pair_data[[id]]$pair
    cat("\n\n## ", pair$pair_id, ": ", pair$gene_names, "\n\n", sep = "")
    print(fiber_plots[[id]]$combined)
    print(configuration_plots[[id]])
  }
} else {
  cat("No fiber plots requested or no qualifying pairs.\n")
}
```

```{r verified-configuration-counts}
if (length(pair_data)) {
  verified <- rbindlist(lapply(pair_data, function(entry) entry$counts), idcol = "pair_id")
  fwrite(verified, file.path(plot_dir, "LCL_top5_verified_counts.tsv"), sep = "\t")
  knitr::kable(verified)
  cat("Every displayed pair reproduces the pooled analysis counts exactly.\n")
}
```

```{r session-information}
sessionInfo()
```
