---
title: "Read-level region plots: early response gene promoters"
author: "Crystal Shan"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Stacked read-level panels for the nine early-response promoter regions, with the SMF
reads split by the topic-model read clusters.

Inputs are whatever `run_extract_regions.sh` wrote under `extract_root`; nothing here
reads a BAM or queries tabix.

# Functions

```{r functions}
code_dir <- "/project/spott/cshan/fiber-seq/code/topic_model/read_region_extraction"
source(file.path(code_dir, "region_data_utils.R"))
source(file.path(code_dir, "region_plot_functions.R"))
```

# Settings

```{r settings}
options(scipen = 999)

# original:
# extract_root <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/region_plots/early_repsonse_genes"
# topic_model_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model"
# plot_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/region_plots/read_region_extraction/results/plots/region_examples"
extract_root <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/region_plots/early_repsonse_genes/promoters"
topic_model_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/promoters"
plot_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/topic_model/region_plots/read_region_extraction/results/plots/region_examples/promoters"
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

sample_names <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")

# FiberHMM `tf` calls are already floored at score 50 by fiberhmm-extract.
fiberHMM_feature      <- "tf"
min_fiberHMM_fp_score <- 50
fiberHMM_size_breaks  <- c(10, 30, 60, 80)
fiberHMM_cols         <- c("purple", "magenta", "#41c4e1")
```

# Regions

Same nine regions as the topic model
(`code/topic_model/1_m6a_promoter_topic_modelling.Rmd`).

```{r regions}
# original (explicit genomic intervals):
# promoters <- tibble::tribble(
#   ~gene,        ~chr,     ~start,      ~end,
#   "IL1B",       "chr2",   112836323,   112837385,
#   "TNF",        "chr6",   31575265,    31575824,
#   "IL1A",       "chr2",   112793233,   112794969,
#   "PTGS2",      "chr1",   186679985,   186681163,
#   "IL8",        "chr4",   73740006,    73741332,
#   "CXCL2",      "chr4",   74098863,    74099651,
#   "IL6",        "chr7",   22726376,    22727094,
#   "RELB",       "chr19",  45001011,    45002417,
#   "rs2836882",  "chr21",  39094468,    39094712
# )

# TSS +/- 1000 bp, matching the TSS-centered regions in
# code/topic_model/1_m6a_promoter_topic_modelling.Rmd and run_extract_regions.sh
# (dominant THP-1 CAGE peak: IL1B/TNF/IL8; GENCODE v49 TSS fallback:
# IL1A/PTGS2/CXCL2/IL6/RELB). rs2836882 is a SNP locus with no TSS - it keeps
# its explicit interval.
promoters <- tibble::tribble(
  ~gene,        ~chr,     ~start,      ~end,
  "IL1B",       "chr2",   112835779,   112837779,
  "TNF",        "chr6",   31574568,    31576568,
  "IL1A",       "chr2",   112783493,   112785493,
  "PTGS2",      "chr1",   186679922,   186681922,
  "IL8",        "chr4",   73739569,    73741569,
  "CXCL2",      "chr4",   74098196,    74100196,
  "IL6",        "chr7",   22724884,    22726884,
  "RELB",       "chr19",  45000461,    45002461,
  "rs2836882",  "chr21",  39094468,    39094712
)

# original:
# promoters$outname <- with(promoters, paste0(gene, "_", chr, "_", start, "_", end))
promoters$outname <- promoters$gene
promoters
```

# Plot one region

`plot_promoter_region()` wires the three steps together. Set `use_clusters = FALSE` to
get the same panels without the topic-model split.

```{r plot-region-fn}
plot_promoter_region <- function(gene, use_clusters = TRUE, peaks_df = NULL,
                                 width = 10, height = 14) {
  region  <- promoters %>% dplyr::filter(gene == !!gene)
  outname <- region$outname
  cat("=== ", gene, ": ", region$chr, ":", region$start, "-", region$end, " ===\n", sep = "")

  result_dirs <- file.path(extract_root, outname, sample_names, "parsed")

  res <- load_region_results(
    region      = region[, c("chr", "start", "end")],
    sample_names = sample_names,
    result_dirs  = result_dirs,
    fiberHMM_feature      = fiberHMM_feature,
    min_fiberHMM_fp_score = min_fiberHMM_fp_score,
    fiberHMM_size_breaks  = fiberHMM_size_breaks)

  if (use_clusters) {
    assignments_file <- file.path(topic_model_dir, outname,
                                  paste0("read_topic_assignments_", outname, ".tsv"))
    res <- add_topic_clusters(res, assignments_file, assignment_col = "cluster")
  }

  p <- plot_region_panels(res,
                          plot_name = gene,
                          peaks_df = peaks_df,
                          fiberHMM_cols = fiberHMM_cols)

  suffix <- if (use_clusters) "_by_cluster" else ""
  ggsave(file.path(plot_dir, paste0("region_", outname, suffix, ".pdf")),
         p$combined, width = width, height = height, limitsize = FALSE)

  invisible(list(res = res, plot = p))
}
```

# All regions

```{r plot-all, fig.width=10, fig.height=14}
out <- list()
for (g in promoters$gene) {
  out[[g]] <- plot_promoter_region(g, use_clusters = TRUE)
  print(out[[g]]$plot$combined)
}
```

# QC: reads per cluster and timepoint

Compare against the counts in `read_topic_assignments_<outname>.tsv`; the difference is
the reads dropped because `ft fire` produced no segmentation for them.

```{r qc-counts}
qc <- lapply(names(out), function(g) {
  rids <- out[[g]]$res$rids_df
  as.data.frame(table(cluster = rids$cluster, sample_name = rids$sample_name)) %>%
    dplyr::mutate(gene = g, .before = 1)
}) %>% dplyr::bind_rows()

qc %>%
  tidyr::pivot_wider(names_from = sample_name, values_from = Freq) %>%
  knitr::kable()
```

```{r qc-footprint-sizes}
# every drawn footprint must fall inside the outer size breaks
sizes <- lapply(names(out), function(g) {
  fps <- out[[g]]$res$fps_infire
  fps <- fps[!is.na(fps$class), ]
  data.frame(gene = g, n = nrow(fps),
             min_size = min(fps$size), max_size = max(fps$size))
}) %>% dplyr::bind_rows()
sizes
stopifnot(all(sizes$min_size >= min(fiberHMM_size_breaks)),
          all(sizes$max_size <= max(fiberHMM_size_breaks)))
```
