---
title: "LCL Fiber-seq autocorrelation"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    mathjax: null
params:
  output_dir: "../../../LCL_project/auto_correlation/resolution04"
  regions: null
---

# Analysis and parameters

### Input


**Step 1: Original fibertools m6A BED12 records**

The IDs, coordinates, block counts, and block values below come from the
original BED file. All four records have chromosome `chr2` and RGB `128,0,128`.
BED starts are 0-based; BED ends are exclusive. `read_start` and `read_end`
are BED12 columns 7 and 8. Block lists are abbreviated with `…` for display;
`blockCount` gives their full length. Complete, unmodified BED12 lines are
included immediately below the table.

| Read | RID | BED start | BED end | read_start | read_end | Score | Strand | blockCount | blockSizes  | blockStarts  |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | m84241_250908_180102_s3/160175746/ccs | 119282125 | 119298507 | 119282125 | 119298507 | 8 | - | 144 | 0,1,1,1,1, …, 1,1,1 | 0,37,452,618,1019, …, 15880,16060,16381 |
| B | m84241_250515_002859_s3/157422976/ccs | 119280507 | 119293438 | 119280507 | 119293438 | 7 | - | 248 | 0,1,1,1,1, …, 1,1,1 | 0,309,310,312,462, …, 12680,12873,12930 |
| C | m84241_250908_180102_s3/258540726/ccs | 119289060 | 119298334 | 119289060 | 119298334 | 18 | - | 266 | 0,1,1,1,1, …, 1,1,1 | 0,293,299,301,309, …, 9078,9080,9273 |
| D | m84241_250515_002859_s3/165545798/ccs | 119291372 | 119296368 | 119291372 | 119296368 | 32 | + | 120 | 0,1,1,1,1, …, 1,1,1 | 0,117,122,184,277, …, 4815,4821,4995 |

The first and last blocks are fibertools blocks and are removed.
In these real records, the first block has size **0** and the last has size
**1**. Internal m6A-call blocks have size **1**.

**Step 2: Expand blocks and keep calls inside the analysis window**

`extract_ft_region_reads()` retains the longest alignment per RID, expands
BED12 blocks, removes the sentinel blocks, and keeps calls within
119290056–119292055. A call's 1-based coordinate is its block end:

`call position = BED start + blockStarts entry + blockSizes entry`

For example, read A's call at 119291048 is `119282125 + 8922 + 1`.
Strand does not reverse the genomic column order.


| Read | Block number (1-based) | Block offset | Block size | Call position |
| --- | --- | --- | --- | --- |
| A | 82 | 7937 | 1 | 119290063 |
| A | 93 | 8922 | 1 | 119291048 |
| A | 94 | 8924 | 1 | 119291050 |
| A | 95 | 8926 | 1 | 119291052 |
| A | 96 | 8929 | 1 | 119291055 |
| B | 182 | 9563 | 1 | 119290071 |
| C | 43 | 1086 | 1 | 119290147 |
| C | 73 | 1987 | 1 | 119291048 |
| C | 74 | 1989 | 1 | 119291050 |
| C | 75 | 1991 | 1 | 119291052 |
| C | 76 | 1994 | 1 | 119291055 |
| C | 77 | 1998 | 1 | 119291059 |
| D | 2 | 117 | 1 | 119291490 |


**Step 3: Extract one metadata row per read**

`extract_ft_read_info(reads)` uses `BED12 column 7 + 1` as the inclusive
read start and column 8 as the inclusive read end

| Read | Chromosome | Start (1-based) | End (inclusive) | Strand | In-window m6A calls |
| --- | --- | --- | --- | --- | --- |
| A | chr2 | 119282126 | 119298507 | - | 26 |
| B | chr2 | 119280508 | 119293438 | - | 50 |
| C | chr2 | 119289061 | 119298334 | - | 81 |
| D | chr2 | 119291373 | 119296368 | + | 15 |

**Step 4: Construct the read-by-position matrix**


1. Initialize entries to `NA`.
2. Set positions covered by a read to `0`.
3. Set observed m6A-call positions to `1`.

`0` means covered without a reported m6A call; `NA` means uncovered.
Every genomic base is a column, including bases with no calls in any read.
The following values were checked with the current R extraction and matrix
functions against the original indexed BED file.

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |
| D | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA |

D starts at 119291373, so all 12 displayed bases are uncovered (`NA`).
It covers only part of the full analysis window.

**Step 5: Filter coverage and phasing**

| Read | Spans full 2,000 bp? | Haplotype | Phase set | Focal genotype | Resolved allele | Outcome |
| --- | --- | --- | --- | --- | --- | --- |
| A | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| B | Yes | HP2 | 119099734 | 0\|1 | rs12470189: A | Keep |
| C | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| D | No | Not evaluated | — | — | — | Remove before phasing |

`allele_status = phased_focal_genotype` and pass the phased
heterozygote filter. Requires at least three retained
reads, both focal alleles, and at least two columns with observed m6A calls.


**Step 6: Final m6A matrix**

`lcl_build_phased_m6a_input()` builds all 2,000 genomic columns for the
retained reads

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |


The defaults are a **2,000-bp window and all 2,000 non-negative ACF lags**.
The analysis runner accepts `--window-size` and `--n-features`; omitting
`--n-features` retains all lags. Knitting uses the saved run parameters.

# Method


## Calculate autocorrelation

`ACF(k) = sum((x[t]-mean(x))*(x[t+k]-mean(x))) / (N * var(x))`

for each lag k (distance apart for a given comparison), it compares the centered signal with itself shifted by k bp. Positions with positive deviations after subtracting the mean give positive products. The differences are summed across all overlapping positions (positions that still line up after shifting by the lag), and divided by the read's total variance (variation in m6a pattern across all positions after subtracting the mean, used for normalization)

Autocorrelation calculates a ACF value measures the linear relationship and correlation coefficient between a time serious and a lagged version of itself at various time steps, specifically in Fiber-seq, it answers the question of how similar is the methylation pattern to itself after shifting it by k base pairs?

    - ACF(0) = 1, the pattern is compared to itself, so it matches perfectly
    - ACF(k) > 0, positions k bp apart tend to have the same state 
    - ACF(k) < 0, positions k bp apart tend to have opposite state
    - ACG(k) = 0, little systematic relationship at that spacing


Previous analysis using Fourier transformation does not give base pair resolution, where the resolution/period length is dependent on the frequency (k = 1,2,3…) and the length of the region, autocorrelation bypass this issue because:

  - FFT has uniform frequency resolution but not uniform period resolution -> how much of this molecule can be described by oscillations having the specific frequencies?
  - Autocorrelation samples uniformly in genomic distance
  - Autocorrelation represents the information on a distance axis, whereas the FFT represents it on an evenly spaced frequency axis

Constant molecules have undefined ACF and are excluded from clustering

```{r setup, include=TRUE, purl=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
  out.width = "100%", fig.width = 12, fig.height = 8, dpi = 300,
  fig.retina = 1, dev = "svglite")
options(scipen = 999)
output <- normalizePath(params$output_dir, mustWork = TRUE)
region_dirs <- sort(list.dirs(file.path(output, "outputs"), recursive = FALSE, full.names = TRUE))
region_dirs <- region_dirs[grepl("^rank[0-9]+_", basename(region_dirs))]
if (!is.null(params$regions)) {
  missing <- setdiff(params$regions, basename(region_dirs))
  if (length(missing)) stop("Regions not found: ", paste(missing, collapse = ", "))
  region_dirs <- region_dirs[basename(region_dirs) %in% params$regions]
}
if (!length(region_dirs)) stop("No region results found under ", output)
```



```{r empty-footprints-function, include=TRUE}
empty_footprints <- function() {
  data.frame(RID = character(), original_RID = character(), chr = character(),
    start = integer(), end = integer(), size = integer(), track = character())
}
```

```{r prepare-nucleosomes-function, include=TRUE}
prepare_nucleosomes <- function(region, assignments, ft_root) {
  output <- list()
  query <- GenomicRanges::GRanges(region$chr,
    IRanges::IRanges(region$analysis_start, region$analysis_end))
  for (sample in unique(assignments$sample_name)) {
    path <- file.path(ft_root, sample, "extracted_results", "nuc_by_chr",
      paste0(sample, ".ft_extracted_nuc.", region$chr, ".bed.gz"))
    if (!file.exists(path) || !file.exists(paste0(path, ".tbi")))
      stop("Missing indexed nucleosome BED: ", path)
    lines <- unlist(Rsamtools::scanTabix(Rsamtools::TabixFile(path), param = query),
      use.names = FALSE)
    if (!length(lines)) next
    bed <- utils::read.table(text = lines, sep = "\t", header = FALSE,
      quote = "", comment.char = "", stringsAsFactors = FALSE)
    stopifnot(ncol(bed) == 12L)
    names(bed) <- c("chr", "start", "end", "RID", "score", "strand",
      "read_start", "read_end", "rgb", "blockCount", "blockSizes", "blockStarts")
    bed <- bed[order(bed$RID, -(bed$end - bed$start)), , drop = FALSE]
    bed <- bed[!duplicated(bed$RID), , drop = FALSE]
    bed <- bed[bed$RID %in% assignments$original_RID[assignments$sample_name == sample], , drop = FALSE]
    for (i in seq_len(nrow(bed))) {
      sizes <- as.integer(strsplit(sub(",$", "", bed$blockSizes[i]), ",", fixed = TRUE)[[1]])
      offsets <- as.integer(strsplit(sub(",$", "", bed$blockStarts[i]), ",", fixed = TRUE)[[1]])
      stopifnot(length(sizes) == bed$blockCount[i], length(offsets) == length(sizes),
        !anyNA(sizes), !anyNA(offsets))
      # Fibertools' first and last blocks are sentinels, not footprints.
      keep <- if (length(sizes) > 2L) seq.int(2L, length(sizes) - 1L) else integer()
      starts <- bed$start[i] + offsets[keep]
      ends <- starts + sizes[keep]
      valid <- sizes[keep] >= 130L & sizes[keep] <= 160L &
        starts < region$end & ends > region$start
      if (!any(valid)) next
      output[[length(output) + 1L]] <- data.frame(
        RID = paste(sample, bed$RID[i], sep = "::"), original_RID = bed$RID[i],
        chr = region$chr, start = starts[valid], end = ends[valid],
        size = sizes[keep][valid], track = "ft_nuc_130-160bp")
    }
  }
  if (length(output)) do.call(rbind, output) else empty_footprints()
}
```

```{r prepare-fiberhmm-function, include=TRUE}
prepare_fiberhmm_track <- function(region, assignments, size, project, tracks_root) {
  filename <- paste0("combined_", region$chr, "_", size, "bp_fps.bed.gz")
  cache <- file.path(project, "LCL_project/Leiden_manhattan/footprint_summary",
    paste0(filename, ".rds"))
  source <- file.path(tracks_root, region$chr, filename)
  cached <- if (file.exists(cache)) readRDS(cache) else NULL
  bounds <- if (!is.null(cached)) cached$signature$regions else NULL
  covered <- !is.null(bounds) && any(bounds$chr == region$chr &
    bounds$start <= region$start & bounds$end >= region$end)
  if (covered) {
    records <- cached$records
    used <- cache
  } else {
    if (!file.exists(source)) stop("Missing footprint source: ", source)
    program <- sprintf('BEGIN {FS=OFS="\t"} $1 == "%s" && $2 < %.0f && $3 > %.0f {print}',
      region$chr, region$end, region$start)
    command <- paste("gzip -cd", shQuote(source), "| awk", shQuote(program))
    lines <- system2("/bin/bash", c("-o", "pipefail", "-c", shQuote(command)), stdout = TRUE)
    if (!is.null(attr(lines, "status")) && attr(lines, "status") != 0L)
      stop("Footprint extraction failed: ", source)
    records <- if (length(lines)) utils::read.table(text = lines, sep = "\t",
      header = FALSE, quote = "", comment.char = "", stringsAsFactors = FALSE) else
      data.frame(chr = character(), start = integer(), end = integer(), original_RID = character())
    stopifnot(ncol(records) == 4L)
    names(records) <- c("chr", "start", "end", "original_RID")
    used <- source
  }
  records <- records[records$chr == region$chr & records$start < region$end &
    records$end > region$start, , drop = FALSE]
  records <- merge(records, assignments[, c("original_RID", "RID")],
    by = "original_RID", sort = FALSE)
  records$size <- records$end - records$start
  records$track <- rep(paste0("FiberHMM_", size, "bp"), nrow(records))
  list(records = records[, names(empty_footprints()), drop = FALSE],
    source = data.frame(track = paste0("FiberHMM_", size, "bp"), source = used,
      n_records = nrow(records)))
}
```

```{r prepare-report-data-function, include=TRUE}
prepare_report_data <- function(payload, project, destination,
    ft_root = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
    tracks_root = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks") {
  stopifnot(payload$schema_version == 1L, is.data.frame(payload$region),
    nrow(payload$region) == 1L, is.data.frame(payload$records),
    identical(payload$row_ids, payload$records$row_id), !anyDuplicated(payload$row_ids))
  payload$acf <- autocor_payload_matrix(payload$acf, nrow(payload$records))
  stopifnot(nrow(payload$acf) == nrow(payload$records),
    ncol(payload$acf) == payload$params$n_features)
  a <- payload$records[payload$records$status == "clustered", , drop = FALSE]
  a$original_RID <- a$RID
  a$RID <- paste(a$sample_name, a$original_RID, sep = "::")
  stopifnot(!anyDuplicated(a$RID))
  payload$footprints <- empty_footprints()
  payload$footprint_tracks <- character()
  payload$footprint_sources <- data.frame(track = character(), source = character(), n_records = integer())
  if (nrow(a)) {
    nucleosomes <- prepare_nucleosomes(payload$region, a, ft_root)
    tracks <- lapply(c("10-30", "40-60", "60-80"), function(size)
      prepare_fiberhmm_track(payload$region, a, size, project, tracks_root))
    payload$footprints <- do.call(rbind, c(list(nucleosomes), lapply(tracks, `[[`, "records")))
    payload$footprint_tracks <- c("ft_nuc_130-160bp",
      paste0("FiberHMM_", c("10-30", "40-60", "60-80"), "bp"))
    payload$footprint_sources <- rbind(data.frame(track = "ft_nuc_130-160bp",
      source = ft_root, n_records = nrow(nucleosomes)), do.call(rbind, lapply(tracks, `[[`, "source")))
  }
  payload$footprints_available <- length(payload$footprint_tracks) > 0L
  dir.create(dirname(destination), recursive = TRUE, showWarnings = FALSE)
  temporary <- tempfile("report-data-", tmpdir = dirname(destination))
  on.exit(unlink(temporary), add = TRUE)
  saveRDS(payload, temporary, compress = "gzip")
  if (!file.rename(temporary, destination)) stop("Could not save report data: ", destination)
  message("Saved portable report data: ", destination)
  invisible(destination)
}
```

```{r read-report-data-function, include=TRUE}
read_report_data <- function(folder) {
  path <- file.path(folder, "report_data.rds")
  if (!file.exists(path)) stop("Missing analysis data for direct plotting: ", path)
  payload <- readRDS(path)
  stopifnot(identical(as.integer(payload$schema_version), 1L),
    nrow(payload$region) == 1L, payload$region$region_id == basename(folder),
    identical(payload$row_ids, payload$records$row_id),
    is.matrix(payload$acf), nrow(payload$acf) == nrow(payload$records))
  rows <- summarize_repeats(payload$acf, payload$records, payload$region, payload$tables)
  list(folder = folder, region_id = payload$region$region_id, data = payload,
    repeats = rows, fiberseq = autocor_fiberseq_result(payload))
}
```

```{r report-fiberseq-figure-function, include=TRUE}
report_fiberseq_figure <- function(entry, draw) {
  if (is.null(entry$fiberseq)) return(ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = "No clustered reads in this region") +
    ggplot2::labs(title = entry$region_id) + ggplot2::theme_void())
  draw(entry$fiberseq)
}
```

```{r print-report-table-function, include=TRUE}
print_report_table <- function(table, caption) {
  if (is.null(table) || !is.data.frame(table) || !nrow(table)) {
    cat(caption, ": no rows.\n\n", sep = "")
  } else {
    cat(knitr::kable(table, caption = caption, row.names = FALSE, digits = 4,
      format = "html", escape = TRUE), sep = "\n")
    cat("\n\n")
  }
  invisible(NULL)
}
```

```{r report-empty-plot-function, include=TRUE}
report_empty_plot <- function(message, title = NULL) {
  ggplot2::ggplot() + ggplot2::annotate("text", x = 0, y = 0, label = message) +
    ggplot2::labs(title = title) + ggplot2::theme_void()
}
```

```{r report-allele-labels-function, include=TRUE}
report_allele_labels <- function(payload) {
  c(REF = paste0("REF (", payload$region$ref, ")"), ALT = paste0("ALT (", payload$region$alt, ")"))
}
```

```{r report-figure-frame-function, include=TRUE}
report_figure_frame <- function(plot, payload, caption = NULL) {
  title <- paste0(payload$region$region_id, "\n", payload$region$focal_snp, ": ",
    payload$region$ref, " / ", payload$region$alt)
  heading <- cowplot::ggdraw() + cowplot::draw_label(title, size = 11)
  if (is.null(caption)) return(cowplot::plot_grid(heading, plot, ncol = 1, rel_heights = c(.12, 1)))
  footer <- cowplot::ggdraw() + cowplot::draw_label(caption, size = 9)
  cowplot::plot_grid(heading, plot, footer, ncol = 1, rel_heights = c(.12, 1, .13))
}
```

```{r report-cluster-acf-function, include=TRUE}
plot_report_cluster_acf <- function(payload) {
  d <- payload$tables$cluster_acf_summary
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("No clustered reads", payload$region$region_id))
  d <- d[d$lag_bp > 0, , drop = FALSE]
  if (!nrow(d)) return(report_empty_plot("No nonzero ACF lags available", payload$region$region_id))
  groups <- as.character(sort(unique(as.integer(as.character(d$cluster)))))
  d$cluster <- factor(d$cluster, levels = groups)
  colors <- cluster_id_palette(groups)
  counts <- d$n_reads[match(groups, d$cluster)]
  curves <- ggplot2::ggplot(d, ggplot2::aes(lag_bp, mean_acf, color = cluster)) +
    ggplot2::geom_hline(yintercept = 0, color = "grey60", linewidth = .3) +
    ggplot2::geom_line(linewidth = .4) +
    ggplot2::scale_color_manual(values = colors, labels = paste0("C", groups, " (n=", counts, ")")) +
    ggplot2::labs(title = "Cluster averages; no smoothing", x = "Lag (bp)", y = "Mean ACF", color = "Cluster") +
    ggplot2::theme_bw()
  bound <- max(c(abs(d$mean_acf[is.finite(d$mean_acf)]), .01))
  heatmap <- ggplot2::ggplot(d, ggplot2::aes(lag_bp, cluster, fill = mean_acf)) +
    ggplot2::geom_tile(width = 1, height = 1) +
    ggplot2::scale_y_discrete(limits = rev(groups), labels = paste0("C", rev(groups))) +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-bound, bound), name = "Mean ACF") +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::labs(x = "Lag (bp)", y = "Cluster") + ggplot2::theme_bw()
  report_figure_frame(cowplot::plot_grid(curves, heatmap, ncol = 1), payload)
}
```

```{r report-sample-composition-function, include=TRUE}
plot_report_sample_composition <- function(payload) {
  d <- payload$tables$sample_composition
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("No clustered reads", payload$region$region_id))
  groups <- as.character(sort(unique(as.integer(as.character(d$cluster)))))
  d$cluster <- factor(d$cluster, levels = groups)
  ggplot2::ggplot(d, ggplot2::aes(sample_name, fraction_within_sample, fill = cluster)) +
    ggplot2::geom_col(position = ggplot2::position_stack(reverse = TRUE)) +
    ggplot2::scale_fill_manual(values = cluster_id_palette(groups)) +
    ggplot2::scale_y_continuous(limits = c(0, 1), expand = ggplot2::expansion(mult = c(0, .01))) +
    ggplot2::labs(title = payload$region$region_id, x = "Sample", y = "Fraction of clustered molecules", fill = "Cluster") +
    ggplot2::theme_bw() + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = .5))
}
```

```{r report-allele-coverage-function, include=TRUE}
plot_report_allele_coverage <- function(payload) {
  d <- payload$tables$allele_coverage
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("Allele coverage unavailable", payload$region$region_id))
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  columns <- c(n_valid = "Valid ACF reads", n_clustered = "Clustered reads")
  plots <- lapply(names(columns), function(column) {
    values <- d[[column]]
    maximum <- max(c(values[is.finite(values)], 1))
    ggplot2::ggplot(d, ggplot2::aes(allele_group, .data[[column]], fill = allele_group)) +
      ggplot2::geom_col() + ggplot2::geom_text(ggplot2::aes(label = .data[[column]]), vjust = -.4) +
      ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), guide = "none") +
      ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = report_allele_labels(payload), drop = FALSE) +
      ggplot2::scale_y_continuous(limits = c(0, maximum * 1.18), expand = c(0, 0)) +
      ggplot2::labs(title = columns[[column]], x = NULL, y = "Unique molecules") + ggplot2::theme_bw()
  })
  report_figure_frame(cowplot::plot_grid(plotlist = plots, nrow = 1), payload)
}
```

```{r report-allele-acf-function, include=TRUE}
plot_report_allele_acf <- function(payload) {
  d <- payload$tables$allele_acf_summary
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("Allele ACF summaries unavailable", payload$region$region_id))
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  maximum <- max(d$lag_bp)
  coverage <- payload$tables$allele_coverage
  missing <- c("REF", "ALT")[!c("REF", "ALT") %in% as.character(coverage$allele_group[coverage$n_valid > 0])]
  bounds <- list(c(1, maximum), c(0, 500), c(140, 250))
  headings <- c("All retained lags (lag 0 omitted)", "Lags 0–500 bp", "Lags 140–250 bp")
  plots <- lapply(seq_along(bounds), function(i) {
    lower <- bounds[[i]][1]; upper <- max(lower + 1, bounds[[i]][2])
    values <- d[d$lag_bp >= lower & d$lag_bp <= upper, , drop = FALSE]
    values$mean_acf <- as.numeric(values$mean_acf)
    note <- if (maximum < lower) "Lag range not available" else if (length(missing))
      paste("No valid ACF reads:", paste(missing, collapse = ", ")) else NULL
    ggplot2::ggplot(values, ggplot2::aes(lag_bp, mean_acf, color = allele_group)) +
      ggplot2::geom_line(linewidth = .4, na.rm = TRUE) +
      ggplot2::scale_color_manual(values = c(REF = "#2878B5", ALT = "#D55E00"),
        limits = c("REF", "ALT"), labels = report_allele_labels(payload), drop = FALSE) +
      ggplot2::scale_x_continuous(limits = c(lower, upper)) +
      ggplot2::labs(title = headings[i], subtitle = note, x = "Lag (bp)", y = "Mean ACF", color = "Focal allele") +
      ggplot2::theme_bw()
  })
  panels <- cowplot::plot_grid(plots[[1]], cowplot::plot_grid(plots[[2]], plots[[3]], nrow = 1), ncol = 1)
  report_figure_frame(panels, payload, paste0("Available lags: 0–", maximum,
    " bp. Each panel uses the same pooled REF/ALT means; vertical scales are independent."))
}
```

```{r report-allele-proportions-function, include=TRUE}
plot_report_allele_proportions <- function(payload) {
  d <- payload$tables$allele_cluster_proportions
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("No clustered molecules", payload$region$region_id))
  d$cluster <- factor(d$cluster, levels = as.character(sort(unique(as.integer(as.character(d$cluster))))))
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  d$fraction <- as.numeric(d$fraction)
  coverage <- payload$tables$allele_coverage
  missing <- c("REF", "ALT")[!c("REF", "ALT") %in% as.character(coverage$allele_group[coverage$n_clustered > 0])]
  p <- ggplot2::ggplot(d, ggplot2::aes(cluster, fraction, fill = allele_group)) +
    ggplot2::geom_col(position = "dodge", na.rm = TRUE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"),
      limits = c("REF", "ALT"), labels = report_allele_labels(payload), drop = FALSE) +
    ggplot2::scale_y_continuous(limits = c(0, 1.05)) +
    ggplot2::labs(x = "ACF cluster", y = "Fraction of clustered molecules within allele", fill = "Focal allele",
      subtitle = if (length(missing)) paste("No clustered reads:", paste(missing, collapse = ", "))) +
    ggplot2::theme_bw()
  report_figure_frame(p, payload)
}
```

```{r report-allele-peak-distribution-function, include=TRUE}
plot_report_peak_distribution <- function(audit, payload, column, ylabel) {
  audit$allele_group <- factor(audit$allele_group, levels = c("REF", "ALT"))
  audit[[column]] <- as.numeric(audit[[column]])
  values <- audit[is.finite(audit[[column]]), , drop = FALSE]
  missing <- setdiff(c("REF", "ALT"), as.character(unique(values$allele_group)))
  note_y <- if (column == "positive_local_peak_140_250_bp") 195 else if (nrow(values))
    mean(range(values[[column]])) else .5
  note <- data.frame(allele_group = factor(missing, levels = c("REF", "ALT")),
    value = rep(note_y, length(missing)))
  p <- ggplot2::ggplot(values, ggplot2::aes(allele_group, .data[[column]], fill = allele_group)) +
    ggplot2::geom_boxplot(width = .45, alpha = .65, na.rm = TRUE) +
    ggplot2::geom_text(data = note, ggplot2::aes(allele_group, value), label = "No peaks", inherit.aes = FALSE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), guide = "none") +
    ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = report_allele_labels(payload), drop = FALSE) +
    ggplot2::labs(x = NULL, y = ylabel) + ggplot2::theme_bw()
  if (column == "positive_local_peak_140_250_bp") p <- p + ggplot2::scale_y_continuous(limits = c(135, 255))
  p
}
```

```{r report-allele-peaks-function, include=TRUE}
plot_report_allele_peaks <- function(payload) {
  d <- payload$tables$allele_peak_features
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("Allele peak summaries unavailable", payload$region$region_id))
  if (!all(d$peak_band_complete %in% TRUE)) {
    plots <- rep(list(report_empty_plot("140–250 bp peak range\nnot fully available")), 3L)
    return(report_figure_frame(cowplot::plot_grid(plotlist = plots, nrow = 1), payload))
  }
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  d$peak_fraction <- as.numeric(d$peak_fraction)
  d$value_label <- ifelse(is.finite(d$peak_fraction), sprintf("%.2f", d$peak_fraction), "")
  missing <- d[d$n_valid == 0, , drop = FALSE]
  fraction <- ggplot2::ggplot(d, ggplot2::aes(allele_group, peak_fraction, fill = allele_group)) +
    ggplot2::geom_col(na.rm = TRUE) + ggplot2::geom_text(ggplot2::aes(label = value_label), vjust = -.5, na.rm = TRUE) +
    ggplot2::geom_text(data = missing, ggplot2::aes(allele_group), y = .5, label = "No valid reads", inherit.aes = FALSE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), guide = "none") +
    ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = report_allele_labels(payload), drop = FALSE) +
    ggplot2::scale_y_continuous(limits = c(0, 1.15)) +
    ggplot2::labs(x = NULL, y = "Fraction with a qualifying peak") + ggplot2::theme_bw()
  audit <- payload$tables$allele_read_audit
  audit <- audit[audit$allele_analysis_included & audit$acf_valid, , drop = FALSE]
  lag <- plot_report_peak_distribution(audit, payload, "positive_local_peak_140_250_bp", "Molecule peak lag (bp)")
  height <- plot_report_peak_distribution(audit, payload, "peak_acf", "Molecule peak ACF height")
  report_figure_frame(cowplot::plot_grid(fraction, lag, height, nrow = 1), payload,
    "Peak fraction: all valid reads. Distributions: peak-bearing reads only; boxes show median and quartiles.")
}
```

```{r report-table-plot-function, include=TRUE}
plot_report_table <- function(payload, kind) {
  switch(kind,
    cluster_acf = plot_report_cluster_acf(payload),
    sample_composition = plot_report_sample_composition(payload),
    allele_coverage = plot_report_allele_coverage(payload),
    allele_acf = plot_report_allele_acf(payload),
    allele_proportions = plot_report_allele_proportions(payload),
    allele_peaks = plot_report_allele_peaks(payload),
    stop("Unknown summary figure: ", kind))
}
```

```{r plot-acf-umap-function, include=TRUE}
plot_acf_umap <- function(payload) {
  a <- payload$records[payload$records$status == "clustered", , drop = FALSE]
  if (!nrow(a)) return(ggplot2::ggplot() + ggplot2::annotate("text", x = 0, y = 0,
    label = "No clustered reads") + ggplot2::labs(title = payload$region$region_id) +
    ggplot2::theme_void())
  clusters <- ggplot2::ggplot(a, ggplot2::aes(umap1, umap2, color = factor(cluster))) +
    ggplot2::geom_point(size = 1) + ggplot2::labs(title = "ACF clusters", x = "UMAP 1",
      y = "UMAP 2", color = "Cluster") + ggplot2::theme_bw()
  methylation <- ggplot2::ggplot(a, ggplot2::aes(umap1, umap2, color = m6a_call_fraction)) +
    ggplot2::geom_point(size = 1) + ggplot2::scale_color_viridis_c() +
    ggplot2::labs(title = payload$region$region_id, x = "UMAP 1", y = "UMAP 2",
      color = "m6A call fraction") + ggplot2::theme_bw()
  cowplot::plot_grid(clusters, methylation, ncol = 2)
}
```

```{r function-fiberseq-annotation-bar, include=TRUE}
fiberseq_annotation_bar <- function(reads, anchor, width, offset, colors) {
  data.frame(cluster = reads$cluster, row = reads$row,
    left = anchor$left - offset * width, right = anchor$left - (offset - 0.025) * width,
    fill = unname(colors))
}
```

```{r function-fiberseq-composition-table, include=TRUE}
fiberseq_composition_table <- function(assignments, group, counts, column) {
  tab <- as.data.frame(table(cluster = group, category = assignments[[column]]), stringsAsFactors = FALSE)
  names(tab)[3] <- "n_reads"
  tab$cluster <- factor(tab$cluster, levels = levels(group))
  tab$cluster_reads <- counts$n_reads[match(tab$cluster, counts$cluster)]
  tab$fraction <- tab$n_reads / tab$cluster_reads
  stopifnot(all(abs(tapply(tab$fraction, tab$cluster, sum) - 1) < 1e-10))
  tab
}
```

```{r function-fiberseq-composition-bar, include=TRUE}
fiberseq_composition_bar <- function(tab, cluster, n, colors, heading) {
  tab <- tab[tab$cluster == cluster, , drop = FALSE]
  tab$category <- factor(tab$category, levels = names(colors))
  ggplot2::ggplot(tab, ggplot2::aes(x = 1, y = fraction, fill = category)) +
    ggplot2::geom_col(width = .6, position = ggplot2::position_stack(reverse = TRUE)) +
    ggplot2::coord_flip() + ggplot2::scale_fill_manual(values = colors, drop = FALSE) +
    ggplot2::scale_y_continuous(limits = c(0, 1), breaks = c(0, .5, 1), labels = scales::percent,
      expand = ggplot2::expansion(mult = 0)) +
    ggplot2::scale_x_continuous(breaks = NULL) +
    ggplot2::labs(title = heading, subtitle = paste0("n = ", n), x = NULL, y = "Within cluster") +
    ggplot2::theme_bw(base_size = 9) + ggplot2::theme(legend.position = "none",
      panel.grid = ggplot2::element_blank(), plot.margin = ggplot2::margin(5.5, 16, 5.5, 5.5))
}
```

```{r function-autocor-payload-matrix, include=TRUE}
autocor_payload_matrix <- function(values, n_reads) {
  if (is.matrix(values) || is.data.frame(values)) {
    result <- as.matrix(values)
  } else if (is.list(values)) {
    rows <- lapply(values, function(row) {
      vapply(row, function(value) if (is.null(value)) NA_real_ else as.numeric(value), numeric(1))
    })
    if (length(unique(lengths(rows))) > 1L) stop("ACF rows have inconsistent lengths")
    result <- do.call(rbind, rows)
  } else {
    result <- matrix(values, nrow = n_reads, byrow = TRUE)
  }
  if (is.null(result) || nrow(result) != n_reads) stop("ACF and saved molecule IDs do not align")
  storage.mode(result) <- "double"
  result
}
```

```{r function-autocor-fiberseq-result, include=TRUE}
autocor_fiberseq_result <- function(payload) {
  region <- as.data.frame(payload$region, stringsAsFactors = FALSE)
  stopifnot(nrow(region) == 1L, region$width == region$analysis_end - region$analysis_start + 1L)
  records <- as.data.frame(payload$records, stringsAsFactors = FALSE)
  a <- records[records$status == "clustered" & !is.na(records$status), , drop = FALSE]
  if (!nrow(a)) return(NULL)
  stopifnot(all(a$region_id == region$region_id), !anyDuplicated(a$row_id),
    all(a$haplotype %in% c("HP1", "HP2")), all(a$focal_genotype %in% c("0|1", "1|0")),
    all(a$allele_status == "phased_focal_genotype"))
  a$original_RID <- a$RID
  a$RID <- paste(a$sample_name, a$original_RID, sep = "::")
  a$start <- a$read_start
  a$end <- a$read_end
  a$chr <- region$chr
  a$sample_label <- sub("_.*$", "", a$sample_name)
  a$cluster <- factor(a$cluster, levels = as.character(sort(unique(as.integer(a$cluster)))))
  a$heatmap_included <- TRUE
  ids <- unlist(payload$row_ids, use.names = FALSE)
  matched <- match(a$row_id, ids)
  stopifnot(!anyDuplicated(ids), !anyNA(matched), !anyDuplicated(a$RID), !anyNA(a$cluster))
  offsets <- if (is.matrix(payload$call_offsets)) asplit(payload$call_offsets, 1L) else payload$call_offsets
  stopifnot(length(offsets) == length(ids))
  offsets <- lapply(offsets[matched], function(row) as.integer(unlist(row, use.names = FALSE)))
  all_offsets <- unlist(offsets, use.names = FALSE)
  stopifnot(!anyNA(all_offsets), all(all_offsets >= 0L & all_offsets < region$width),
    !any(vapply(offsets, anyDuplicated, integer(1)) > 0L))
  met <- Matrix::sparseMatrix(i = rep(seq_len(nrow(a)), lengths(offsets)),
    j = all_offsets + 1L, x = rep(1, length(all_offsets)), dims = c(nrow(a), region$width),
    dimnames = list(a$RID, as.character(seq.int(region$analysis_start, region$analysis_end))))
  acf <- autocor_payload_matrix(payload$acf, length(ids))[matched, , drop = FALSE]
  stopifnot(ncol(acf) >= 1L, ncol(acf) <= region$width,
    all(is.finite(acf)), all(abs(acf[, 1] - 1) < 1e-8))
  rownames(acf) <- a$RID
  edges <- as.data.frame(payload$edges, stringsAsFactors = FALSE)
  if (!nrow(edges)) edges <- data.frame(source = character(), target = character(), weight = numeric())
  edges$source <- a$RID[match(edges$source, a$row_id)]
  edges$target <- a$RID[match(edges$target, a$row_id)]
  stopifnot(!anyNA(edges), all(is.finite(edges$weight)), all(edges$weight > 0))
  graph <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = a$RID)
  footprints <- payload$footprints
  if (is.null(footprints) || !length(footprints)) footprints <- data.frame(RID = character(),
    original_RID = character(), chr = character(), start = integer(), end = integer(),
    size = integer(), track = character())
  footprints <- as.data.frame(footprints, stringsAsFactors = FALSE)
  footprints <- footprints[footprints$RID %in% a$RID, , drop = FALSE]
  tracks <- if (is.null(payload$footprint_tracks)) unique(footprints$track) else unlist(payload$footprint_tracks)
  focal <- data.frame(chr = region$chr, pos = region$focal_pos, variant_id = region$focal_snp,
    ref = region$ref, alt = region$alt)
  result <- list(assignments = a, region = region, site_met_mat = met, acf = acf, graph = graph,
    n_clusters = nlevels(a$cluster), variants = focal, focal = focal,
    params = if (is.null(payload$params)) list(seed = 0L) else payload$params,
    footprints = footprints, footprint_tracks = intersect(LCL_FOOTPRINT_TRACKS, tracks))
  fiberseq_display_result(result)
}
```

```{r function-autocor-acf-heatmap, include=TRUE}
autocor_acf_heatmap <- function(result) {
  a <- result$assignments
  ordering <- order(a$cluster, a$start, a$RID)
  display <- result$acf[a$RID[ordering], -1, drop = FALSE]
  if (!ncol(display)) stop("The ACF heatmap requires at least one nonzero lag")
  limit <- max(as.numeric(stats::quantile(abs(display), .99)), .01)
  ticks <- unique(round(seq(1, ncol(display), length.out = min(5L, ncol(display)))))
  ComplexHeatmap::Heatmap(display, name = "ACF",
    col = circlize::colorRamp2(c(-limit, 0, limit), c("#2166ac", "white", "#b2182b")),
    cluster_rows = FALSE, cluster_columns = FALSE, cluster_row_slices = FALSE,
    row_split = a$cluster[ordering], show_row_names = FALSE, show_column_names = FALSE,
    use_raster = TRUE, raster_quality = 4, raster_resize_mat = FALSE,
    raster_device = "png", raster_by_magick = FALSE,
    column_title = "Autocorrelation: lag (bp)",
    bottom_annotation = ComplexHeatmap::HeatmapAnnotation(lag_bp = ComplexHeatmap::anno_mark(
      at = ticks, labels = as.character(ticks), which = "column", side = "bottom")),
    width = grid::unit(65, "mm"))
}
```

```{r function-autocor-cluster-proportions, include=TRUE}
autocor_cluster_proportions <- function(result) {
  groups <- levels(result$assignments$cluster)
  counts <- data.frame(cluster = factor(groups, levels = groups),
    n_reads = as.integer(table(result$assignments$cluster)))
  counts$proportion <- counts$n_reads / sum(counts$n_reads)
  counts$label <- sprintf("%s\nn=%d (%.1f%%)", counts$cluster, counts$n_reads, 100 * counts$proportion)
  colors <- cluster_id_palette(groups)
  rgb <- grDevices::col2rgb(colors) / 255
  text_colors <- setNames(ifelse(colSums(rgb * c(.299, .587, .114)) < .5, "white", "black"), names(colors))
  ggplot2::ggplot(counts, ggplot2::aes(x = "", y = proportion, fill = cluster)) +
    ggplot2::geom_col(width = 1, color = "white", position = ggplot2::position_stack(reverse = TRUE)) +
    ggplot2::geom_text(ggplot2::aes(label = ifelse(proportion >= .03, sprintf("%.1f%%", 100 * proportion), ""), color = cluster),
      position = ggplot2::position_stack(vjust = .5, reverse = TRUE), size = 3) +
    ggplot2::scale_color_manual(values = text_colors, guide = "none") +
    ggplot2::coord_polar(theta = "y") + ggplot2::scale_fill_manual(values = colors, labels = counts$label) +
    ggplot2::labs(title = paste0("Cluster proportions: ", nrow(result$assignments), " reads"), fill = NULL) +
    ggplot2::theme_void() + ggplot2::theme(legend.position = "right", legend.text = ggplot2::element_text(size = 8))
}
```

```{r function-autocor-umap-overview, include=TRUE}
autocor_umap_overview <- function(result) {
  embedding <- result$assignments
  embedding$UMAP1 <- embedding$umap1
  embedding$UMAP2 <- embedding$umap2
  stopifnot(all(is.finite(embedding$UMAP1)), all(is.finite(embedding$UMAP2)))
  short_colors <- sample_palette(embedding$sample_name)
  samples <- unique(embedding$sample_name)
  sample_colors <- setNames(unname(short_colors[sub("_.*$", "", samples)]), samples)
  groups <- levels(embedding$cluster)
  cluster_colors <- cluster_id_palette(groups)
  allele_colors <- fiberseq_category_palette(result$allele_display_levels)
  cluster_labels <- paste0(groups, " (n=", as.integer(table(embedding$cluster)), ")")
  cluster_plot <- plot_umap(embedding, "cluster", cluster_colors, "UMAP: cluster")
  sample_plot <- plot_umap(embedding, "sample_name", sample_colors, "UMAP: LCL sample")
  allele_plot <- plot_umap(embedding, "allele_display", allele_colors, "UMAP: focal SNP allele")
  legends <- list(fiberseq_legend(cluster_colors, "Cluster", 5, cluster_labels),
    fiberseq_legend(sample_colors, "LCL sample", 6, sub("_.*$", "", names(sample_colors))),
    fiberseq_legend(allele_colors, "Focal SNP allele", 3))
  heights <- vapply(legends, fiberseq_legend_height, numeric(1))
  heading <- cowplot::ggdraw() + cowplot::draw_label(result$region$annotation, size = 12)
  plot <- cowplot::plot_grid(plotlist = c(list(heading,
    cowplot::plot_grid(cluster_plot, sample_plot, allele_plot, autocor_cluster_proportions(result), ncol = 2)), legends),
    ncol = 1, rel_heights = c(.45, 8, heights))
  list(plot = plot, height = 8.45 + sum(heights))
}
```

```{r function-autocor-profile-composition, include=TRUE}
autocor_profile_composition <- function(result) {
  short_colors <- sample_palette(result$assignments$sample_name)
  samples <- unique(result$assignments$sample_name)
  sample_colors <- setNames(unname(short_colors[sub("_.*$", "", samples)]), samples)
  tables <- fiberseq_tables(result, result$footprints)
  plot_fiberseq_profiles(tables, result$region, result$markers,
    cluster_id_palette(tables$groups), sample_colors,
    fiberseq_category_palette(result$allele_display_levels), result$region$annotation)
}
```

```{r function-autocor-haplotype-example, include=TRUE}
autocor_haplotype_example <- function(result) {
  overview <- autocor_umap_overview(result)
  profiles <- autocor_profile_composition(result)
  cowplot::plot_grid(overview$plot, profiles$plot, ncol = 1,
    rel_heights = c(overview$height, profiles$height))
}
```

```{r function-fiberseq-colors, include=TRUE}
LCL_CLUSTER_COLORS <- c("dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00", "black", "gold1",
  "skyblue2", "#FB9A99", "palegreen2", "#CAB2D6", "#FDBF6F", "gray70", "khaki2",
  "maroon", "orchid1", "deeppink1", "blue1", "steelblue4", "darkturquoise",
  "green1", "yellow4", "yellow3", "darkorange4", "brown")
LCL_HAPLOTYPE_COLORS <- c(HP1 = "#ADD8E6", HP2 = "#FFF2AE", unphased = "#999999", pooled = "#BBBBBB")
LCL_TRACK_COLORS <- c(m6A = "#800080", "ft_nuc_130-160bp" = "#4d4d4d",
  "FiberHMM_10-30bp" = "#fdae6b", "FiberHMM_40-60bp" = "#f16913", "FiberHMM_60-80bp" = "#a63603")
LCL_TRACK_LABELS <- c(m6A = "m6A", "ft_nuc_130-160bp" = "nucleosome footprint (130-160 bp)",
  "FiberHMM_10-30bp" = "TF footprint 10-30 bp", "FiberHMM_40-60bp" = "TF footprint 40-60 bp",
  "FiberHMM_60-80bp" = "TF footprint 60-80 bp")
LCL_FOOTPRINT_TRACKS <- c("ft_nuc_130-160bp", "FiberHMM_10-30bp", "FiberHMM_40-60bp", "FiberHMM_60-80bp")
```

```{r function-autocor-fiberseq-packages, include=TRUE}
autocor_fiberseq_packages <- function() {
  packages <- c("Matrix", "ggplot2", "ComplexHeatmap", "circlize", "igraph", "cowplot", "dplyr", "scales", "svglite", "png")
  missing <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) stop("Install these R packages before rendering: ", paste(missing, collapse = ", "))
  invisible(packages)
}
```

```{r function-cluster-id-palette, include=TRUE}
cluster_id_palette <- function(levels) {
  if (!length(levels)) return(setNames(character(), character()))
  levels <- as.character(levels)
  index <- suppressWarnings(as.integer(sub("^cluster", "", levels)))
  if (all(grepl("^[0-9]+$", levels))) index <- index + 1L
  if (anyNA(index) || any(index < 1L)) index <- seq_along(levels)
  colors <- if (max(index) <= length(LCL_CLUSTER_COLORS)) LCL_CLUSTER_COLORS[index] else
    grDevices::colorRampPalette(LCL_CLUSTER_COLORS)(max(index))[index]
  setNames(colors, levels)
}
```

```{r function-feature-colors, include=TRUE}
feature_colors <- function(tracks) {
  stopifnot(all(tracks %in% names(LCL_TRACK_COLORS)))
  LCL_TRACK_COLORS[tracks]
}
```

```{r function-plot-smf-reads, include=TRUE}
plot_smf_reads <- function(result, records, sample_colors, tracks = LCL_FOOTPRINT_TRACKS,
                                include_haplotype = FALSE) {
  tracks <- intersect(tracks, result$footprint_tracks)
  records <- records[records$track %in% tracks, , drop = FALSE]
  prepared <- prepare_read_tracks(result, records, sample_colors)
  reads <- prepared$reads
  features <- prepared$features[prepared$features$track %in% tracks, , drop = FALSE]
  anchor <- prepared$anchor
  width <- max(1, anchor$right - anchor$left)
  feature_colors <- feature_colors(tracks)
  # Draw nucleosomes first and TF footprints on top, as in TNF Fig 1.
  features <- features[order(match(features$track, tracks)), ]
  features$ymin <- features$row - 0.45
  features$ymax <- features$row + 0.45
  features$fill <- unname(feature_colors[features$track])
  cluster_colors <- cluster_id_palette(levels(reads$cluster))
  bars <- rbind(fiberseq_annotation_bar(reads, anchor, width, 0.075, cluster_colors[as.character(reads$cluster)]),
    fiberseq_annotation_bar(reads, anchor, width, 0.040, sample_colors[reads$sample_label]))
  allele_colors <- if (identical(result$region$region_type, "top_asfire_het"))
    fiberseq_category_palette(result$allele_display_levels) else LCL_HAPLOTYPE_COLORS
  allele <- if (identical(result$region$region_type, "top_asfire_het")) as.character(reads$allele_display) else reads$haplotype
  if (include_haplotype) bars <- rbind(bars, fiberseq_annotation_bar(reads, anchor, width, 0.110, allele_colors[allele]))
  legend_colors <- c(feature_colors, sample_colors,
    if (include_haplotype) allele_colors)
  legend_labels <- c(unname(LCL_TRACK_LABELS[tracks]), names(sample_colors),
    if (include_haplotype) names(allele_colors))
  counts <- table(reads$cluster)
  cluster_labels <- setNames(paste0(names(counts), " (n=", as.integer(counts), ")"), names(counts))
  ticks <- pretty(c(anchor$left, anchor$right), n = 6)
  ggplot2::ggplot() +
    ggplot2::geom_segment(data = reads, ggplot2::aes(x = left, xend = right, y = row, yend = row),
      color = "grey80", linewidth = 0.25) +
    ggplot2::geom_rect(data = features,
      ggplot2::aes(xmin = left, xmax = right, ymin = ymin, ymax = ymax, fill = fill)) +
    ggplot2::geom_rect(data = bars,
      ggplot2::aes(xmin = left, xmax = right, ymin = row - 0.5, ymax = row + 0.5, fill = fill)) +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.3) +
    ggplot2::scale_fill_identity(guide = "legend", name = NULL,
      breaks = unname(legend_colors), labels = legend_labels) +
    ggplot2::scale_x_continuous(limits = c(anchor$left - (if (include_haplotype) 0.12 * width else 0.085 * width),
      anchor$right + 0.5), breaks = ticks[ticks >= anchor$left & ticks <= anchor$right], expand = c(0, 0)) +
    ggplot2::scale_y_reverse(expand = ggplot2::expansion(add = 0.6)) +
    ggplot2::facet_grid(cluster ~ ., scales = "free_y", space = "free_y",
      labeller = ggplot2::as_labeller(cluster_labels)) +
    ggplot2::labs(x = anchor$x_label, y = "SMF reads",
      title = paste(c(result$region$annotation, result$group_id), collapse = " | "),
      subtitle = "One row per retained molecule; saved autocorrelation-defined Leiden clusters") +
    ggplot2::guides(fill = ggplot2::guide_legend(ncol = 4, byrow = TRUE)) +
    cowplot::theme_cowplot(font_size = 10) +
    ggplot2::theme(axis.text.y = ggplot2::element_blank(), axis.ticks.y = ggplot2::element_blank(),
      axis.line.y = ggplot2::element_blank(), strip.text.y = ggplot2::element_text(angle = 0),
      panel.spacing.y = grid::unit(2, "mm"), legend.position = "bottom",
      legend.text = ggplot2::element_text(size = 8), legend.key.size = grid::unit(3, "mm"))
}
```

```{r function-plot-signal-profile, include=TRUE}
plot_signal_profile <- function(profiles, region, clusters,
                                     title = region$annotation) {
  profiles <- profiles[profiles$track %in% names(LCL_TRACK_COLORS), , drop = FALSE]
  anchor <- plot_anchor(region)
  profiles$relative_pos <- anchor$direction * (profiles$pos - anchor$anchor)
  profiles$cluster <- factor(profiles$cluster, levels = clusters)
  profiles$track <- factor(profiles$track, levels = names(LCL_TRACK_COLORS)[names(LCL_TRACK_COLORS) %in% profiles$track])
  profiles <- profiles[order(profiles$cluster, profiles$track, profiles$relative_pos), ]
  colors <- feature_colors(levels(profiles$track))
  counts <- unique(profiles[, c("cluster", "n_reads")])
  labels <- setNames(paste0(counts$cluster, " (n=", counts$n_reads, ")"), counts$cluster)
  plot <- ggplot2::ggplot(profiles,
    ggplot2::aes(relative_pos, fraction, fill = track, color = track, group = track))
  plot <- plot + ggplot2::geom_ribbon(
    data = profiles[profiles$track == "ft_nuc_130-160bp", , drop = FALSE],
    ggplot2::aes(ymin = 0, ymax = fraction), fill = "grey60", color = NA, alpha = .18) +
    ggplot2::geom_line(linewidth = 0.5)
  plot + ggplot2::facet_wrap(~cluster, ncol = 1, labeller = ggplot2::as_labeller(labels)) +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "grey30", linewidth = 0.4) +
    ggplot2::scale_fill_manual(values = colors, labels = LCL_TRACK_LABELS) +
    ggplot2::scale_color_manual(values = colors, labels = LCL_TRACK_LABELS) +
    ggplot2::scale_x_continuous(limits = c(anchor$left - 0.5, anchor$right + 0.5), expand = c(0, 0)) +
    ggplot2::scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1),
      expand = ggplot2::expansion(mult = c(0, 0.02))) +
    ggplot2::labs(x = anchor$x_label, y = "Fraction of cluster reads", title = title,
      subtitle = "m6A calls and footprint occupancy; retained heterozygous reads", fill = "Plot track", color = "Plot track") +
    cowplot::theme_cowplot(font_size = 10) + cowplot::panel_border() +
    ggplot2::guides(fill = ggplot2::guide_legend(ncol = 4), color = ggplot2::guide_legend(ncol = 4)) +
    ggplot2::theme(legend.position = "bottom", legend.text = ggplot2::element_text(size = 8),
      plot.margin = ggplot2::margin(5.5, 16, 5.5, 5.5))
}
```

```{r function-sample-palette, include=TRUE}
sample_palette <- function(sample_names) {
  labels <- sub("_.*$", "", sample_names)
  labels <- unique(labels[order(as.integer(sub("^AL-?([0-9]+).*$", "\\1", labels)))])
  setNames(grDevices::hcl.colors(length(labels), "Dynamic"), labels)
}
```

```{r function-plot-genomic-cluster-heatmap, include=TRUE}
plot_genomic_cluster_heatmap <- function(res, region, sample_colors,
                                     include_haplotype = FALSE, variants = NULL,
                                     show_cluster_profiles = FALSE, split_alleles = FALSE,
                                     cluster_label = "Autocorrelation-defined Leiden clusters") {
  assignments <- res$assignments
  if (split_alleles) stopifnot(include_haplotype, !is.null(assignments$allele_display))
  ordering <- if (split_alleles) order(assignments$allele_display, assignments$cluster, assignments$start, assignments$RID) else
    order(assignments$cluster, assignments$start, assignments$RID)
  assignments <- assignments[ordering, , drop = FALSE]
  sample <- factor(sub("_.*$", "", assignments$sample_name), levels = names(sample_colors))
  annotation <- ComplexHeatmap::rowAnnotation(
    cluster = assignments$cluster, sample = sample,
    col = list(cluster = cluster_id_palette(levels(assignments$cluster)), sample = sample_colors))
  if (include_haplotype && identical(region$region_type, "top_asfire_het")) {
    annotation <- ComplexHeatmap::rowAnnotation(
      cluster = assignments$cluster, sample = sample, allele = assignments$allele_display,
      col = list(cluster = cluster_id_palette(levels(assignments$cluster)), sample = sample_colors,
        allele = fiberseq_category_palette(res$allele_display_levels)))
  } else if (include_haplotype) {
    annotation <- ComplexHeatmap::rowAnnotation(
      cluster = assignments$cluster, sample = sample, haplotype = assignments$haplotype,
      col = list(cluster = cluster_id_palette(levels(assignments$cluster)), sample = sample_colors,
        haplotype = LCL_HAPLOTYPE_COLORS))
  }
  met <- matrix(0L, nrow(assignments), region$width,
                  dimnames = list(assignments$RID, seq.int(region$analysis_start, region$analysis_end)))
  met[, match(colnames(res$site_met_mat), colnames(met))] <- as.matrix(res$site_met_mat[assignments$RID, , drop = FALSE])
  display <- met
  colors <- c("0" = "white", "1" = "black")
  legend <- list(at = c(0, 1), labels = c("no m6A call", "m6A"))
  legend_name <- "m6A"
  top <- NULL
  if (show_cluster_profiles) {
    cluster_colors <- cluster_id_palette(levels(assignments$cluster))
    observed <- match(colnames(res$site_met_mat), colnames(met))
    profile_annotations <- lapply(levels(assignments$cluster), function(cluster) {
      selected <- assignments$cluster == cluster
      ComplexHeatmap::anno_lines(
        stats::approx(observed, colMeans(met[selected, observed, drop = FALSE]),
          xout = seq_len(ncol(met)), rule = 1)$y,
        ylim = c(0, 1), gp = grid::gpar(col = cluster_colors[[cluster]], lwd = 0.7),
        axis_param = list(at = c(0, 0.5, 1), labels = c("0", ".5", "1")),
        height = grid::unit(12, "mm"))
    })
    names(profile_annotations) <- paste0(levels(assignments$cluster), " m6A")
    top <- do.call(ComplexHeatmap::HeatmapAnnotation, c(profile_annotations,
      list(annotation_name_gp = grid::gpar(fontsize = 8), gap = grid::unit(1.5, "mm"))))
  }
  ticks <- unique(round(seq(1, ncol(display), length.out = 5L)))
  bottom_parts <- list(coordinate = ComplexHeatmap::anno_mark(at = ticks,
    labels = format(as.integer(colnames(display)[ticks]), scientific = FALSE, trim = TRUE),
    which = "column", side = "bottom", labels_gp = grid::gpar(fontsize = 8)))
  if (identical(region$region_type, "promoter") && !is.null(region$tss) &&
      !is.na(region$tss) && region$tss >= region$analysis_start && region$tss <= region$analysis_end) {
    bottom_parts$TSS <- ComplexHeatmap::anno_mark(
      at = region$tss - region$analysis_start + 1L,
      labels = paste0("TSS: ", region$chr, ":", region$tss),
      which = "column", side = "bottom", labels_gp = grid::gpar(fontsize = 8, col = "#D55E00"))
  }
  snps <- if (!is.null(region$focal_snp)) data.frame(pos = region$focal_pos,
    label = paste0(region$focal_snp, " ", region$ref, ">", region$alt)) else window_snps(variants, region)
  if (nrow(snps)) {
    bottom_parts$SNP <- ComplexHeatmap::anno_mark(
      at = snps$pos - region$analysis_start + 1L, labels = snps$label,
      which = "column", side = "bottom", labels_gp = grid::gpar(fontsize = 7),
      link_gp = grid::gpar(col = "#984EA3"))
  }
  bottom <- do.call(ComplexHeatmap::HeatmapAnnotation, c(bottom_parts,
    list(annotation_name_gp = grid::gpar(fontsize = 8))))
  ComplexHeatmap::Heatmap(
    display, name = legend_name, col = colors, heatmap_legend_param = legend,
    cluster_rows = FALSE, cluster_columns = FALSE, cluster_row_slices = FALSE,
    row_split = if (split_alleles) data.frame(allele = assignments$allele_display,
      cluster = assignments$cluster) else assignments$cluster,
    row_gap = grid::unit(if (split_alleles) 3 else 0.6, "mm"),
    row_title_rot = 0, row_title_gp = grid::gpar(fontsize = 8),
    show_row_names = FALSE, show_column_names = FALSE,
    use_raster = TRUE, raster_quality = 4, raster_resize_mat = FALSE,
    raster_device = "png", raster_by_magick = FALSE,
    left_annotation = annotation,
    top_annotation = top, bottom_annotation = bottom,
    column_title = paste(c(region$annotation,
      if (include_haplotype) paste0(cluster_label, "; focal allele annotated per sample"),
      if (split_alleles) "Rows split by focal allele, then cluster",
      if (show_cluster_profiles) "Top: m6A call fraction per cluster (all full-span reads)",
      if (include_haplotype && !nrow(snps)) "No phased heterozygous SNP in this window",
      paste0(region$chr, ":", region$analysis_start, "-", region$analysis_end)), collapse = "\n"),
    column_title_gp = grid::gpar(fontsize = 11))
}
```

```{r function-fiberseq-category-palette, include=TRUE}
fiberseq_category_palette <- function(categories) {
  categories <- sort(unique(as.character(categories)))
  categories <- c(setdiff(categories, c("unphased", "Unknown allele")),
    intersect(c("unphased", "Unknown allele"), categories))
  colors <- setNames(grDevices::hcl.colors(length(categories), "Dynamic"), categories)
  focal <- categories[grepl("^rs[0-9]+: [ACGT]$", categories)]
  colors[focal] <- c("#0072B2", "#D55E00", "#CC79A7", "#009E73")[seq_along(focal)]
  unknown <- intersect(c("unphased", "Unknown allele"), categories)
  colors[unknown] <- "#555555"
  colors
}
```

```{r function-fiberseq-legend, include=TRUE}
fiberseq_legend <- function(colors, title, columns = 4L, labels = names(colors)) {
  d <- data.frame(category = factor(names(colors), levels = names(colors)), x = 1, y = 1)
  p <- ggplot2::ggplot(d, ggplot2::aes(x, y, color = category)) +
    ggplot2::geom_point() + ggplot2::scale_color_manual(values = colors, labels = labels, drop = FALSE) +
    ggplot2::guides(color = ggplot2::guide_legend(title = title, title.position = "top", ncol = columns, byrow = TRUE,
      override.aes = list(size = 3))) + ggplot2::theme_void() +
    ggplot2::theme(legend.position = "bottom", legend.text = ggplot2::element_text(size = 8))
  cowplot::get_legend(p)
}
```

```{r function-fiberseq-legend-height, include=TRUE}
fiberseq_legend_height <- function(legend) {
  grid::convertHeight(sum(legend$heights), "in", valueOnly = TRUE) + .2
}
```

```{r function-plot-fiberseq-profiles, include=TRUE}
plot_fiberseq_profiles <- function(tables, region, markers, cluster_colors, sample_colors, allele_colors,
                                  title = region$annotation) {
  allele_heading <- if (identical(region$region_type, "top_asfire_het"))
    "Focal SNP allele" else "Allele / local haplotype"
  rows <- lapply(tables$groups, function(cluster) {
    d <- tables$profiles[tables$profiles$cluster == cluster, ]
    n <- tables$counts$n_reads[match(cluster, tables$counts$cluster)]
    nuc <- d[d$track == "ft_nuc_130-160bp", , drop = FALSE]
    met <- d[d$track == "m6A", , drop = FALSE]
    p <- ggplot2::ggplot(d, ggplot2::aes(pos, fraction)) +
      ggplot2::geom_ribbon(data = nuc, ggplot2::aes(ymin = 0, ymax = fraction),
        fill = "#808080", alpha = .18, color = NA) +
      ggplot2::geom_line(data = nuc, color = "#666666", linewidth = .55) +
      ggplot2::geom_line(data = met, color = cluster_colors[[cluster]], linewidth = .45) +
      ggplot2::scale_x_continuous(limits = c(region$analysis_start, region$analysis_end),
        labels = function(x) format(x, scientific = FALSE, trim = TRUE), expand = ggplot2::expansion(mult = 0)) +
      ggplot2::scale_y_continuous(limits = c(0, 1), breaks = c(0, .5, 1)) +
      ggplot2::labs(title = paste0(cluster, " (n = ", n, ")"), x = paste0(region$chr, " (1-based bp)"), y = "Read fraction") +
      ggplot2::theme_bw(base_size = 9) + ggplot2::theme(legend.position = "none",
        plot.title = ggplot2::element_text(color = cluster_colors[[cluster]], face = "bold"))
    if (nrow(markers)) {
      p <- p +
        ggplot2::geom_vline(data = markers, ggplot2::aes(xintercept = pos),
          inherit.aes = FALSE, color = "#666666", linetype = "dotted", linewidth = .35)
    }
    cowplot::plot_grid(p, fiberseq_composition_bar(tables$sample, cluster, n, sample_colors, "LCL sample"),
      fiberseq_composition_bar(tables$allele, cluster, n, allele_colors, allele_heading), nrow = 1,
      rel_widths = c(2.7, 1, 1.5), align = "h", axis = "tb")
  })
  marker_text <- if (nrow(markers)) paste(paste0(markers$label, ": ", markers$pos),
    collapse = "; ") else "No focal annotation resolved"
  footprint_caption <- if ("ft_nuc_130-160bp" %in% tables$profiles$track)
    "grey line / light grey fill: 130-160 bp nucleosome occupancy" else "nucleosome footprints unavailable"
  header <- cowplot::ggdraw() + cowplot::draw_label(paste0(title,
    "\nCluster color: m6A calls; ", footprint_caption, "\n",
    paste(strwrap(marker_text, width = 155), collapse = "\n")), size = 10)
  sample_legend <- fiberseq_legend(sample_colors, "LCL sample", 6, sub("_.*$", "", names(sample_colors)))
  allele_legend <- fiberseq_legend(allele_colors, allele_heading, 3)
  heights <- c(.85, rep(1.35, length(rows)), fiberseq_legend_height(sample_legend), fiberseq_legend_height(allele_legend))
  plot <- cowplot::plot_grid(plotlist = c(list(header), rows, list(sample_legend, allele_legend)),
                            ncol = 1, rel_heights = heights)
  list(plot = plot, height = sum(heights))
}
```

```{r function-plot-umap, include=TRUE}
plot_umap <- function(embedding, column, colors, title, labels = names(colors),
                           show_legend = FALSE) {
  ggplot2::ggplot(embedding, ggplot2::aes(UMAP1, UMAP2, color = .data[[column]])) +
    ggplot2::geom_point(size = .9, alpha = .75) +
    ggplot2::scale_color_manual(values = colors, labels = labels, breaks = names(colors), drop = FALSE) +
    ggplot2::coord_equal() + ggplot2::labs(title = title, color = NULL) +
    ggplot2::theme_bw(base_size = 10) +
    ggplot2::theme(legend.position = if (show_legend) "right" else "none")
}
```

```{r function-plot-knn-graph, include=TRUE}
plot_knn_graph <- function(result, main = NULL, seed = NULL, vertex_size = 1.5,
                           edge_width = 0.3, edge_alpha = 0.2, color_by = c("cluster", "allele")) {
  color_by <- match.arg(color_by)
  graph <- result$graph
  assignments <- result$assignments
  if (!igraph::is_igraph(graph)) stop("The clustering result must contain its saved KNN graph")
  stopifnot(all(c("RID", "cluster") %in% names(assignments)),
    !igraph::is_directed(graph), igraph::vcount(graph) > 0L,
    !anyNA(assignments$RID), !anyDuplicated(assignments$RID), !anyNA(assignments$cluster))
  read_ids <- igraph::V(graph)$name
  stopifnot(length(read_ids) == igraph::vcount(graph), !anyNA(read_ids),
    !anyDuplicated(read_ids), setequal(read_ids, as.character(assignments$RID)))
  assignments <- assignments[match(read_ids, assignments$RID), , drop = FALSE]
  clusters <- if (is.factor(assignments$cluster)) levels(droplevels(assignments$cluster)) else
    unique(as.character(assignments$cluster))
  colors <- cluster_id_palette(clusters)
  weights <- igraph::edge_attr(graph, "weight")
  if (length(weights)) stopifnot(all(is.finite(weights)), all(weights > 0))
  if (is.null(seed)) seed <- if (is.null(result$params$seed)) 1L else result$params$seed
  stopifnot(length(seed) == 1L, is.finite(seed))
  had_seed <- exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  if (had_seed) previous_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
  on.exit({
    if (had_seed) {
      assign(".Random.seed", previous_seed, envir = .GlobalEnv)
    } else if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
      rm(".Random.seed", envir = .GlobalEnv)
    }
  }, add = TRUE)
  set.seed(seed)
  coordinates <- igraph::layout_with_fr(graph, weights = weights)
  nodes <- data.frame(RID = read_ids, layout_x = coordinates[, 1], layout_y = coordinates[, 2],
    cluster = factor(assignments$cluster, levels = clusters))
  nodes$color_group <- nodes$cluster
  color_levels <- clusters
  legend_title <- "Leiden cluster"
  if (color_by == "allele") {
    summary <- knn_allele_summary(result)
    color_levels <- summary$allele_levels
    nodes$color_group <- factor(assignments$allele_label, levels = color_levels)
    colors <- fiberseq_category_palette(color_levels)
    legend_title <- "Focal SNP allele"
  }
  endpoints <- igraph::as_edgelist(graph, names = FALSE)
  edges <- data.frame(from_x = coordinates[endpoints[, 1], 1],
    from_y = coordinates[endpoints[, 1], 2], to_x = coordinates[endpoints[, 2], 1],
    to_y = coordinates[endpoints[, 2], 2])
  counts <- table(nodes$color_group)
  labels <- paste0(color_levels, " (n=", as.integer(counts[color_levels]), ")")
  if (is.null(main)) main <- paste(c(result$region$annotation,
    paste("ACF correlation-neighbor graph colored by", if (color_by == "allele") "focal SNP allele" else "Leiden cluster")),
    collapse = "\n")
  neighbors <- if (is.null(result$params$k_eff)) result$params$k_neighbors else result$params$k_eff
  settings <- c(paste(nrow(nodes), "reads"), paste(nrow(edges), "edges"),
    if (!is.null(neighbors)) paste0("k = ", neighbors),
    if (!is.null(result$params$resolution)) paste0("resolution = ", result$params$resolution))
  ggplot2::ggplot() +
    ggplot2::geom_segment(data = edges,
      ggplot2::aes(x = from_x, y = from_y, xend = to_x, yend = to_y),
      color = "grey60", linewidth = edge_width, alpha = edge_alpha) +
    ggplot2::geom_point(data = nodes, ggplot2::aes(layout_x, layout_y, color = color_group),
      size = vertex_size) +
    ggplot2::scale_color_manual(values = colors, breaks = color_levels, labels = labels, drop = FALSE) +
    ggplot2::coord_equal() +
    ggplot2::labs(title = main, subtitle = paste(settings, collapse = " | "), color = legend_title) +
    ggplot2::theme_void(base_size = 11) +
    ggplot2::theme(legend.position = "right", plot.title = ggplot2::element_text(hjust = 0.5),
      plot.subtitle = ggplot2::element_text(hjust = 0.5),
      plot.margin = ggplot2::margin(12, 12, 12, 12))
}
```

```{r function-knn-allele-summary, include=TRUE}
knn_allele_summary <- function(result) {
  graph <- result$graph
  assignments <- result$assignments
  focal <- result$focal
  stopifnot(igraph::is_igraph(graph), !igraph::is_directed(graph), igraph::is_simple(graph),
    nrow(assignments) > 1L, !is.null(focal), nrow(focal) == 1L,
    all(c("RID", "cluster", "allele_label", "allele_status") %in% names(assignments)),
    !anyNA(assignments$RID), !anyDuplicated(assignments$RID), !anyNA(assignments$cluster),
    !anyNA(assignments$allele_label), all(assignments$allele_status == "phased_focal_genotype"))
  alleles <- sort(paste0(focal$variant_id, ": ", c(focal$ref, focal$alt)))
  stopifnot(length(alleles) == 2L, length(unique(alleles)) == 2L,
    all(grepl("^rs[0-9]+: [ACGT]$", alleles)), all(assignments$allele_label %in% alleles))
  read_ids <- igraph::V(graph)$name
  stopifnot(length(read_ids) == igraph::vcount(graph), !anyNA(read_ids),
    !anyDuplicated(read_ids), setequal(read_ids, as.character(assignments$RID)))
  assignments <- assignments[match(read_ids, assignments$RID), , drop = FALSE]
  clusters <- if (is.factor(assignments$cluster)) levels(droplevels(assignments$cluster)) else
    unique(as.character(assignments$cluster))
  stopifnot(!"All reads" %in% clusters)
  allele <- factor(assignments$allele_label, levels = alleles)
  composition <- as.data.frame(table(group = factor(assignments$cluster, levels = clusters),
    allele = allele), stringsAsFactors = FALSE)
  names(composition)[3] <- "n_reads"
  overall <- data.frame(group = "All reads", allele = alleles, n_reads = as.integer(table(allele)))
  composition <- dplyr::bind_rows(overall, composition)
  composition$group <- factor(composition$group, levels = c("All reads", clusters))
  composition$allele <- factor(composition$allele, levels = alleles)
  composition$total_reads <- ave(composition$n_reads, composition$group, FUN = sum)
  composition$fraction <- composition$n_reads / composition$total_reads
  endpoints <- igraph::as_edgelist(graph, names = FALSE)
  source_nodes <- c(endpoints[, 1], endpoints[, 2])
  target_nodes <- c(endpoints[, 2], endpoints[, 1])
  degree <- tabulate(source_nodes, nbins = length(read_ids))
  same <- tabulate(source_nodes[allele[source_nodes] == allele[target_nodes]], nbins = length(read_ids))
  reads <- data.frame(RID = read_ids, cluster = assignments$cluster, allele = allele,
    degree = degree, same_allele_neighbors = same, opposite_allele_neighbors = degree - same)
  reads$same_fraction <- ifelse(degree > 0L, same / degree, NA_real_)
  reads$opposite_fraction <- ifelse(degree > 0L, (degree - same) / degree, NA_real_)
  connectivity <- dplyr::bind_rows(lapply(alleles, function(label) {
    selected <- reads$allele == label
    connected <- selected & reads$degree > 0L
    data.frame(allele = label, n_reads = sum(selected), n_reads_with_neighbors = sum(connected),
      n_isolated_reads = sum(selected & reads$degree == 0L),
      same_fraction = if (any(connected)) mean(reads$same_fraction[connected]) else NA_real_,
      opposite_fraction = if (any(connected)) mean(reads$opposite_fraction[connected]) else NA_real_,
      expected_same_fraction = if (any(selected)) (sum(selected) - 1L) / (nrow(reads) - 1L) else NA_real_)
  }))
  connectivity$allele <- factor(connectivity$allele, levels = alleles)
  list(allele_levels = alleles, composition = composition, reads = reads, connectivity = connectivity)
}
```

```{r function-cluster-site-profiles, include=TRUE}
cluster_site_profiles <- function(res, met_mat) {
  M  <- as.matrix(met_mat)
  df <- res$assignments
  df <- df[df$RID %in% rownames(M), ]
  do.call(rbind, lapply(levels(df$cluster), function(cl) {
    rids <- df$RID[df$cluster == cl]
    met  <- colMeans(M[rids, , drop = FALSE], na.rm = TRUE)
    met[is.nan(met)] <- NA
    data.frame(cluster = cl, pos = as.numeric(colnames(M)), met = met,
               n_reads = length(rids), row.names = NULL)
  }))
}
```

```{r function-window-snps, include=TRUE}
window_snps <- function(variants, region) {
  empty <- data.frame(pos = integer(), label = character())
  if (is.null(variants) || !nrow(variants)) return(empty)
  snps <- variants[variants$chr == region$chr & variants$pos >= region$analysis_start &
    variants$pos <= region$analysis_end & nchar(variants$ref) == 1L &
    grepl("^[ACGT](,[ACGT])*$", variants$alt), , drop = FALSE]
  if (!nrow(snps)) return(empty)
  snps$id <- ifelse(is.na(snps$variant_id) | snps$variant_id == ".", "SNP", snps$variant_id)
  snps$label <- paste0(snps$id, " ", snps$pos, " ", snps$ref, ">", snps$alt)
  labels <- tapply(snps$label, snps$pos, function(x) paste(unique(x), collapse = "; "))
  data.frame(pos = as.integer(names(labels)), label = unname(labels))
}
```

```{r function-focal-markers, include=TRUE}
focal_markers <- function(result) {
  focal <- result$focal
  if (!is.null(focal) && nrow(focal))
    return(data.frame(pos = focal$pos, end = focal$pos,
      label = paste0(focal$variant_id, " ", focal$ref, ">", focal$alt)))
  region <- result$region
  if (!is.null(region$focal_pos) && !is.na(region$focal_pos))
    return(data.frame(pos = region$focal_pos, end = region$focal_pos, label = region$focal_snp))
  if (!is.null(region$tss) && !is.na(region$tss))
    return(data.frame(pos = region$tss, end = region$tss, label = paste(region$gene, "TSS")))
  data.frame(pos = integer(), end = integer(), label = character())
}
```

```{r function-fiberseq-display-result, include=TRUE}
fiberseq_display_result <- function(result) {
  result$markers <- focal_markers(result)
  a <- result$assignments
  a$allele_display <- a$allele_label
  anchored <- !is.null(result$region$focal_pos) && !is.na(result$region$focal_pos)
  if (anchored) {
    known <- a$allele_status == "phased_focal_genotype"
    strict <- identical(result$region$region_type, "top_asfire_het")
    if (strict && (!all(known) || !all(a$focal_genotype %in% c("0|1", "1|0"))))
      stop("Top AS-FIRE plots require only resolved phased heterozygous reads")
    a$allele_display[!known] <- "Unknown allele"
    bases <- a$allele_label[known]
    v <- result$variants
    if (!is.null(v) && nrow(v)) {
      v <- v[v$chr == result$region$chr & v$pos == result$region$focal_pos, , drop = FALSE]
      bases <- c(bases, paste0(if (is.null(result$region$focal_snp)) result$region$region_id else result$region$focal_snp, ": ",
        unique(c(v$ref, unlist(strsplit(v$alt, ",", fixed = TRUE))))))
    }
    bases <- sort(unique(bases[grepl("^rs[0-9]+: [ACGT]$", bases)]))
    if (strict) {
      bases <- sort(paste0(result$region$focal_snp, ": ", c(result$region$ref, result$region$alt)))
      stopifnot(length(unique(bases)) == 2L, all(a$allele_label %in% bases))
    }
    result$allele_display_levels <- if (strict) bases else c(bases, "Unknown allele")
  } else {
    result$allele_display_levels <- if (is.null(result$allele_levels))
      sort(unique(a$allele_label)) else result$allele_levels
  }
  a$allele_display <- factor(a$allele_display, levels = result$allele_display_levels)
  stopifnot(!anyNA(a$allele_display))
  result$assignments <- a
  result
}
```

```{r function-fiberseq-tables, include=TRUE}
fiberseq_tables <- function(result, footprints) {
  result <- fiberseq_display_result(result)
  a <- result$assignments
  groups <- if (is.factor(a$cluster)) levels(droplevels(a$cluster)) else unique(a$cluster)
  group <- factor(a$cluster, levels = groups)
  counts <- data.frame(cluster = groups, n_reads = as.integer(table(group)), total_reads = nrow(a))
  counts$proportion <- counts$n_reads / counts$total_reads
  view <- result
  view$assignments$cluster <- group
  nuc <- footprint_profiles(footprints, intersect("ft_nuc_130-160bp", result$footprint_tracks),
    view$assignments, view$region)
  met <- cluster_site_profiles(view, view$site_met_mat)
  profiles <- dplyr::bind_rows(data.frame(cluster = met$cluster, pos = met$pos,
    track = "m6A", fraction = met$met, n_reads = met$n_reads), nuc)
  profiles$cluster <- factor(profiles$cluster, levels = groups)
  profiles$chr <- result$region$chr
  profiles$coordinate_system <- "1-based inclusive"
  stopifnot(all(is.finite(profiles$fraction)), all(profiles$fraction >= 0 & profiles$fraction <= 1))
  list(counts = counts, sample = fiberseq_composition_table(a, group, counts, "sample_name"),
       allele = fiberseq_composition_table(a, group, counts, "allele_display"),
       profiles = profiles, groups = groups)
}
```

```{r function-m6a-intervals, include=TRUE}
m6a_intervals <- function(result) {
  sites <- Matrix::summary(as(result$site_met_mat, "dgCMatrix"))
  sites <- sites[sites$x > 0, , drop = FALSE]
  positions <- as.integer(colnames(result$site_met_mat)[sites$j])
  identifiers <- rownames(result$site_met_mat)[sites$i]
  data.frame(RID = identifiers,
    original_RID = result$assignments$original_RID[match(identifiers, result$assignments$RID)],
    chr = rep(result$region$chr, length(identifiers)), start = positions - 1L,
    end = positions, size = rep(1L, length(identifiers)), track = rep("m6A", length(identifiers)))
}
```

```{r function-plot-anchor, include=TRUE}
plot_anchor <- function(region) {
  promoter <- identical(region$region_type, "promoter") && !is.null(region$tss) && !is.na(region$tss)
  focal <- !is.null(region$focal_pos) && !is.na(region$focal_pos)
  anchor <- if (focal) region$focal_pos else if (promoter) region$tss else floor(mean(c(region$analysis_start, region$analysis_end)))
  direction <- if (promoter && region$strand == "-") -1L else 1L
  bounds <- sort(direction * (c(region$analysis_start, region$analysis_end) - anchor))
  list(anchor = anchor, direction = direction, left = bounds[1], right = bounds[2],
    x_label = if (focal) paste0("Position relative to ", region$focal_snp, " (bp)") else if (promoter) "Position relative to canonical TSS (bp)" else "Position relative to region centre (bp)")
}
```

```{r function-prepare-read-tracks, include=TRUE}
prepare_read_tracks <- function(result, records, sample_colors) {
  anchor <- plot_anchor(result$region)
  reads <- result$assignments
  reads <- reads[order(reads$cluster, reads$start, reads$RID), , drop = FALSE]
  reads$row <- as.integer(ave(seq_len(nrow(reads)), reads$cluster, FUN = seq_along))
  reads$sample_label <- sub("_.*$", "", reads$sample_name)
  stopifnot(all(reads$sample_label %in% names(sample_colors)))
  if (!"haplotype" %in% names(reads)) reads$haplotype <- "pooled"
  relative_start <- anchor$direction * (reads$start - anchor$anchor)
  relative_end <- anchor$direction * (reads$end - anchor$anchor)
  reads$left <- pmax(pmin(relative_start, relative_end), anchor$left)
  reads$right <- pmin(pmax(relative_start, relative_end), anchor$right)
  matched <- match(records$RID, reads$RID)
  records <- records[!is.na(matched), , drop = FALSE]
  matched <- matched[!is.na(matched)]
  records$row <- reads$row[matched]
  records$cluster <- reads$cluster[matched]
  relative_start <- anchor$direction * (records$start + 1L - anchor$anchor)
  relative_end <- anchor$direction * (records$end - anchor$anchor)
  records$left <- pmax(pmin(relative_start, relative_end) - 0.5, anchor$left - 0.5)
  records$right <- pmin(pmax(relative_start, relative_end) + 0.5, anchor$right + 0.5)
  list(reads = reads, features = records, anchor = anchor, region = result$region)
}
```

```{r function-occupancy-matrix, include=TRUE}
occupancy_matrix <- function(records, assignments, region) {
  occupancy <- matrix(0L, nrow(assignments), region$width,
                       dimnames = list(assignments$RID, seq.int(region$analysis_start, region$analysis_end)))
  if (!nrow(records)) return(occupancy)
  for (record_index in seq_len(nrow(records))) {
    read_index <- match(records$RID[record_index], assignments$RID)
    if (is.na(read_index)) next
    left <- max(records$start[record_index], region$start) - region$start + 1L
    right <- min(records$end[record_index], region$end) - region$start
    if (left <= right) occupancy[read_index, seq.int(left, right)] <- 1L
  }
  occupancy
}
```

```{r function-footprint-profiles, include=TRUE}
footprint_profiles <- function(records, tracks, assignments, region) {
  dplyr::bind_rows(lapply(tracks, function(track) {
    occupancy <- occupancy_matrix(records[records$track == track, ], assignments, region)
    dplyr::bind_rows(lapply(levels(assignments$cluster), function(cluster) {
      selected <- assignments$cluster == cluster
      data.frame(cluster = cluster, pos = as.integer(colnames(occupancy)), track = track,
                 fraction = colMeans(occupancy[selected, , drop = FALSE]), n_reads = sum(selected))
    }))
  }))
}
```

```{r function-signal-profiles, include=TRUE}
signal_profiles <- function(result, footprints) {
  footprints <- footprints[footprints$track %in% LCL_FOOTPRINT_TRACKS, , drop = FALSE]
  records <- dplyr::bind_rows(m6a_intervals(result), footprints)
  fp <- footprint_profiles(footprints, result$footprint_tracks, result$assignments, result$region)
  met <- cluster_site_profiles(result, result$site_met_mat)
  met <- data.frame(cluster = met$cluster, pos = met$pos, track = "m6A",
    fraction = met$met, n_reads = met$n_reads)
  list(records = records, tracks = c("m6A", result$footprint_tracks),
       profiles = dplyr::bind_rows(met, fp))
}
```

```{r function-acf-local-peaks, include=TRUE}
acf_local_peaks <- function(profile, prominence = 0.01, distance = 20L) {
  # Plateau peaks use the midpoint rounded down; endpoints are never peaks.
  profile <- as.numeric(profile)
  n <- length(profile)
  if (n < 3L || any(!is.finite(profile))) return(integer())
  peaks <- integer()
  i <- 2L
  while (i < n) {
    if (profile[i] > profile[i - 1L]) {
      right <- i
      while (right < n && profile[right + 1L] == profile[i]) right <- right + 1L
      if (right < n && profile[right + 1L] < profile[i]) {
        peaks <- c(peaks, as.integer(floor((i + right) / 2)))
      }
      i <- right + 1L
    } else {
      i <- i + 1L
    }
  }
  if (!length(peaks)) return(integer())
  # As in scipy.signal.find_peaks, distance filtering precedes prominence.
  # Give the rightmost peak priority when heights tie.
  keep <- rep(TRUE, length(peaks))
  priority <- order(profile[peaks], peaks, decreasing = TRUE)
  for (j in priority) {
    if (keep[j]) {
      nearby <- which(abs(peaks - peaks[j]) < ceiling(distance))
      keep[setdiff(nearby, j)] <- FALSE
    }
  }
  peaks <- peaks[keep]
  prominence_values <- numeric(length(peaks))
  for (j in seq_along(peaks)) {
    peak <- peaks[j]
    height <- profile[peak]
    higher_left <- which(profile[seq_len(peak - 1L)] > height)
    left <- if (length(higher_left)) tail(higher_left, 1L) else 1L
    higher_right <- which(profile[seq.int(peak + 1L, n)] > height)
    right <- if (length(higher_right)) peak + higher_right[1L] else n
    prominence_values[j] <- height - max(
      min(profile[seq.int(left, peak)]), min(profile[seq.int(peak, right)])
    )
  }
  peaks[prominence_values >= prominence]
}
```

```{r function-repeat-candidates, include=TRUE}
repeat_candidates <- function(profile) {
  # Returned positions are lags in bp (lag 0 is R vector position 1).
  profile <- as.numeric(profile)
  missing <- list(primary = NA_real_, higher = integer())
  if (length(profile) <= 251L || any(!is.finite(profile))) return(missing)
  positions <- acf_local_peaks(profile, prominence = 0.01, distance = 20L)
  positions <- positions[profile[positions] > 0]
  lags <- positions - 1L
  band <- positions[lags >= 120L & lags <= 250L]
  if (!length(band)) return(missing)
  primary <- band[which.max(profile[band])] - 1L
  higher <- integer()
  last_multiple <- (length(profile) - 1L) %/% primary
  if (last_multiple >= 2L) {
    for (multiple in seq.int(2L, last_multiple)) {
      nearby <- positions[abs(lags - multiple * primary) <= 20L]
      if (length(nearby)) higher <- c(higher, nearby[which.max(profile[nearby])] - 1L)
    }
  }
  list(primary = primary, higher = higher)
}
```

```{r function-summarize-repeats, include=TRUE}
summarize_repeats <- function(profiles, records, region, allele_tables) {
  profiles <- as.matrix(profiles)
  audit <- allele_tables$allele_read_audit
  needed <- c("row_id", "allele_analysis_included", "acf_valid", "allele_group")
  if (is.null(audit) || !all(needed %in% names(audit))) {
    stop("The allele read audit is missing required columns.")
  }
  if (nrow(profiles) != nrow(records) || nrow(audit) != nrow(records) ||
      !identical(as.character(records$row_id), as.character(audit$row_id))) {
    stop("ACFs, records and allele audit must have matching rows in the same order.")
  }
  if (!ncol(profiles)) stop("The ACF matrix must contain at least lag 0.")
  included <- as.logical(audit$allele_analysis_included)
  valid <- as.logical(audit$acf_valid)
  if (anyNA(included) || anyNA(valid)) stop("Audit inclusion and ACF validity must be defined.")
  results <- vector("list", 2L)
  for (a in seq_along(results)) {
    allele <- c("REF", "ALT")[a]
    allele_mask <- !is.na(audit$allele_group) & audit$allele_group == allele
    selected <- included & allele_mask
    indices <- which(selected & valid)
    values <- profiles[indices, , drop = FALSE]
    if (length(indices) && any(!is.finite(values))) {
      stop("The audit marks a non-finite ACF profile as valid.")
    }
    median <- if (length(indices)) apply(values, 2L, stats::median) else rep(NA_real_, ncol(profiles))
    candidates <- repeat_candidates(median)
    per_fiber <- vapply(seq_len(nrow(values)), function(i) repeat_candidates(values[i, ])$primary, numeric(1))
    ordering <- order(per_fiber, seq_along(per_fiber), na.last = TRUE)
    results[[a]] <- list(
      region_id = as.character(region$region_id), allele = allele,
      base = as.character(if (allele == "REF") region$ref else region$alt),
      median = median, primary = candidates$primary, higher = candidates$higher,
      peak_lags = per_fiber, indices = indices[ordering], sorted_peaks = per_fiber[ordering],
      n_valid = length(indices), n_invalid = sum(selected & !valid),
      n_unique = sum(selected), n_duplicates = sum(!included & allele_mask)
    )
  }
  results
}
```

```{r function-repeat-summary-table, include=TRUE}
repeat_summary_table <- function(rows) {
  do.call(rbind, lapply(rows, function(row) {
    complete <- length(row$median) >= 252L
    n_peak <- sum(is.finite(row$peak_lags))
    data.frame(
      region_id = row$region_id, allele = row$allele, base = row$base,
      unique_fibers = row$n_unique, valid_acf = row$n_valid,
      undefined_acf = row$n_invalid, duplicates_removed = row$n_duplicates,
      peak_band_complete = complete,
      fibers_with_peak = if (complete) n_peak else NA_integer_,
      fibers_without_peak = if (complete) row$n_valid - n_peak else NA_integer_,
      peak_fraction = if (complete && row$n_valid > 0L) n_peak / row$n_valid else NA_real_,
      median_curve_primary_bp = row$primary,
      observed_higher_order_bp = paste(row$higher, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }))
}
```

```{r function-repeat-counts-caption, include=TRUE}
repeat_counts_caption <- function(rows) {
  paste(vapply(rows, function(row) {
    sprintf("%s: %s valid, %s undefined ACF, %s duplicates removed",
            row$allele, row$n_valid, row$n_invalid, row$n_duplicates)
  }, character(1)), collapse = "; ")
}
```

```{r function-plot-haplotype-median-acf, include=TRUE}
plot_haplotype_median_acf <- function(rows, region, lower = 0, upper = NULL) {
  max_lag <- max(vapply(rows, function(row) length(row$median), integer(1))) - 1L
  if (is.null(upper)) upper <- max_lag
  curves <- do.call(rbind, lapply(rows, function(row) {
    data.frame(allele = row$allele, lag = seq_along(row$median) - 1L, acf = row$median)
  }))
  # Keep tables intact when a lag range contains just one row or no peaks.
  curves <- subset(curves, lag >= lower & lag <= upper)
  primary <- do.call(rbind, lapply(rows, function(row) {
    if (!is.finite(row$primary)) return(NULL)
    data.frame(allele = row$allele, lag = row$primary, acf = row$median[row$primary + 1L])
  }))
  higher <- do.call(rbind, lapply(rows, function(row) {
    if (!length(row$higher)) return(NULL)
    data.frame(allele = row$allele, lag = row$higher, acf = row$median[row$higher + 1L])
  }))
  annotations <- do.call(rbind, lapply(rows, function(row) {
    message <- if (is.finite(row$primary)) {
      sprintf("n = %s; primary = %s bp", row$n_valid, row$primary)
    } else if (!row$n_valid) {
      "No valid ACF fibers"
    } else if (max_lag < 251L) {
      "Search band incomplete"
    } else {
      sprintf("n = %s; no qualifying primary peak", row$n_valid)
    }
    data.frame(allele = row$allele, label = message)
  }))
  allele_levels <- c("REF", "ALT")
  curves$allele <- factor(curves$allele, levels = allele_levels)
  annotations$allele <- factor(annotations$allele, levels = allele_levels)
  if (!is.null(primary)) primary$allele <- factor(primary$allele, levels = allele_levels)
  if (!is.null(higher)) higher$allele <- factor(higher$allele, levels = allele_levels)
  p <- ggplot2::ggplot(curves, ggplot2::aes(lag, acf, color = allele)) +
    ggplot2::annotate("rect", xmin = 120, xmax = 250, ymin = -Inf, ymax = Inf,
                      fill = "#61a788", alpha = 0.13) +
    ggplot2::geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
    ggplot2::geom_line(linewidth = 0.45, na.rm = TRUE) +
    ggplot2::geom_text(data = annotations, ggplot2::aes(x = Inf, y = Inf, label = label),
                       inherit.aes = FALSE, hjust = 1.03, vjust = 1.3, size = 3) +
    ggplot2::facet_grid(allele ~ ., drop = FALSE) +
    ggplot2::scale_color_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), drop = FALSE) +
    ggplot2::coord_cartesian(xlim = c(lower, max(lower + 1, upper))) +
    ggplot2::labs(title = paste0(region$region_id, ": phased-allele median ACF"),
      subtitle = paste0("REF ", region$ref, " / ALT ", region$alt),
      x = "Lag (bp)", y = "Median raw ACF",
      caption = paste("Primary search: 120–250 bp (shading); stars: primary peaks; open circles: observed peaks near multiples.",
        "Positive maxima: prominence ≥0.01, spacing ≥20 bp; multiple tolerance ±20 bp. No smoothing.",
        repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw() + ggplot2::theme(legend.position = "none", plot.caption = ggplot2::element_text(hjust = 0))
  if (!is.null(primary)) {
    primary <- subset(primary, lag >= lower & lag <= upper)
    p <- p + ggplot2::geom_point(data = primary, shape = 8, color = "#B8860B", size = 3.5)
  }
  if (!is.null(higher)) {
    higher <- subset(higher, lag >= lower & lag <= upper)
    p <- p + ggplot2::geom_point(data = higher, shape = 1, size = 2)
  }
  p
}
```

```{r function-plot-haplotype-fiber-acf-heatmap, include=TRUE}
plot_haplotype_fiber_acf_heatmap <- function(profiles, rows, region, upper = NULL) {
  profiles <- as.matrix(profiles)
  max_lag <- ncol(profiles) - 1L
  if (is.null(upper)) upper <- max_lag
  upper <- min(as.integer(upper), max_lag)
  ordered <- unlist(lapply(rows, function(row) row$indices), use.names = FALSE)
  peaks <- unlist(lapply(rows, function(row) row$sorted_peaks), use.names = FALSE)
  title <- paste0(region$region_id, ": individual-fiber ACF")
  if (!length(ordered) || upper < 1L) {
    return(ggplot2::ggplot() + ggplot2::annotate("text", x = 0, y = 0,
      label = if (!length(ordered)) "No valid ACF fibers" else "No nonzero ACF lags available") +
      ggplot2::labs(title = title, caption = repeat_counts_caption(rows)) + ggplot2::theme_void())
  }
  all_values <- profiles[ordered, -1L, drop = FALSE]
  limit <- max(as.numeric(stats::quantile(abs(all_values), 0.99, names = FALSE)), 0.01)
  values <- all_values[, seq_len(upper), drop = FALSE]
  tiles <- data.frame(lag = rep(seq_len(upper), times = length(ordered)),
                      fiber = rep(seq_along(ordered), each = upper), acf = as.vector(t(values)))
  dots <- data.frame(lag = peaks, fiber = seq_along(ordered))
  dots <- dots[is.finite(dots$lag) & dots$lag <= upper, , drop = FALSE]
  sizes <- vapply(rows, function(row) row$n_valid, numeric(1))
  ends <- cumsum(sizes)
  centers <- ends - (sizes - 1) / 2
  labels <- vapply(rows, function(row) sprintf("%s (n=%s)", row$allele, row$n_valid), character(1))
  p <- ggplot2::ggplot(tiles, ggplot2::aes(lag, fiber, fill = acf)) +
    ggplot2::geom_raster() +
    ggplot2::geom_point(data = dots, ggplot2::aes(lag, fiber), inherit.aes = FALSE, size = 0.35) +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-limit, limit), oob = scales::squish, na.value = "#dddddd",
      name = "Raw ACF") +
    ggplot2::scale_y_reverse(breaks = centers[sizes > 0], labels = labels[sizes > 0], expand = c(0, 0)) +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::labs(title = title, subtitle = paste0("REF ", region$ref, " / ALT ", region$alt),
      x = "Lag (bp); lag 0 omitted", y = "Fibers: allele, then primary peak lag",
      caption = paste("All valid unique fibers, independent of cluster assignment. No-peak fibers sort last within each allele.",
        "Dots: primary peaks. Color limits clip the largest 1% of absolute ACF values for display.",
        repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw() + ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0))
  boundaries <- ends[ends > 0 & ends < length(ordered)] + 0.5
  if (length(boundaries)) p <- p + ggplot2::geom_hline(yintercept = boundaries, linewidth = 0.4)
  p
}
```

```{r function-plot-haplotype-repeat-boxplot, include=TRUE}
plot_haplotype_repeat_boxplot <- function(rows, region) {
  values <- do.call(rbind, lapply(rows, function(row) {
    peaks <- row$peak_lags[is.finite(row$peak_lags)]
    data.frame(allele = rep(row$allele, length(peaks)), peak = peaks)
  }))
  values$allele <- factor(values$allele, levels = c("REF", "ALT"))
  labels <- setNames(vapply(rows, function(row) {
    sprintf("%s\n%s/%s with peak", row$allele, sum(is.finite(row$peak_lags)), row$n_valid)
  }, character(1)), vapply(rows, function(row) row$allele, character(1)))
  empty <- data.frame(allele = vapply(rows, function(row) row$allele, character(1)),
                      has_peak = vapply(rows, function(row) any(is.finite(row$peak_lags)), logical(1)))
  empty <- empty[!empty$has_peak, , drop = FALSE]
  ggplot2::ggplot(values, ggplot2::aes(allele, peak, fill = allele)) +
    ggplot2::geom_boxplot(width = 0.5, outlier.size = 1) +
    ggplot2::geom_text(data = empty, ggplot2::aes(x = allele, y = 185, label = "No estimates"),
                       inherit.aes = FALSE, size = 3) +
    ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = labels, drop = FALSE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), drop = FALSE) +
    ggplot2::coord_cartesian(ylim = c(115, 255)) +
    ggplot2::labs(title = paste0(region$region_id, ": per-fiber repeat-length candidates"),
      x = NULL, y = "Per-fiber candidate repeat length (bp)",
      caption = paste("Boxes include only fibers with a qualifying positive local maximum in 120–250 bp; no smoothing.",
        "No qualifying peak means no estimate.", repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw() + ggplot2::theme(legend.position = "none", plot.caption = ggplot2::element_text(hjust = 0))
}
```

```{r function-plot-haplotype-repeat-histogram, include=TRUE}
plot_haplotype_repeat_histogram <- function(rows, region) {
  # Use all valid fibers as the denominator, including those without a peak.
  breaks <- seq(119.5, 254.5, by = 5)
  bins <- do.call(rbind, lapply(rows, function(row) {
    peaks <- row$peak_lags[is.finite(row$peak_lags)]
    counts <- graphics::hist(peaks, breaks = breaks, plot = FALSE, right = TRUE)$counts
    data.frame(allele = row$allele, xmin = head(breaks, -1L), xmax = tail(breaks, -1L),
      fraction = if (row$n_valid) counts / row$n_valid else rep(0, length(counts)))
  }))
  bins$allele <- factor(bins$allele, levels = c("REF", "ALT"))
  ggplot2::ggplot(bins, ggplot2::aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = fraction, color = allele)) +
    ggplot2::geom_rect(fill = NA, linewidth = 0.6) +
    ggplot2::scale_color_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), drop = FALSE) +
    ggplot2::coord_cartesian(xlim = c(119, 251)) +
    ggplot2::labs(title = paste0(region$region_id, ": repeat-length candidate frequencies"),
      x = "Per-fiber primary peak lag (bp)", y = "Fraction of all valid fibers per 5-bp bin", color = "Allele",
      caption = paste("Strongest qualifying positive local maximum in 120–250 bp; no smoothing.",
        "Histogram denominators include all valid fibers, including those without an estimate.",
        repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw() + ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0))
}
```

```{r function-plot-locus-heatmap, include=TRUE}
plot_locus_heatmap <- function(rows, region = NULL) {
  region_ids <- unique(vapply(rows, function(row) as.character(row$region_id), character(1)))
  if (length(region_ids) != 1L) stop("Plot one region at a time.")
  lags <- 120:250
  tiles <- do.call(rbind, lapply(seq_along(rows), function(i) {
    row <- rows[[i]]
    values <- rep(NA_real_, length(lags))
    available <- lags < length(row$median)
    values[available] <- row$median[lags[available] + 1L]
    data.frame(lag = lags, allele_row = i, acf = values)
  }))
  limit <- max(c(abs(tiles$acf[is.finite(tiles$acf)]), 0.01))
  stars <- data.frame(lag = vapply(rows, function(row) row$primary, numeric(1)), allele_row = seq_along(rows))
  stars <- stars[is.finite(stars$lag), , drop = FALSE]
  labels <- vapply(rows, function(row) sprintf("%s (%s), n=%s", row$allele, row$base, row$n_valid), character(1))
  ggplot2::ggplot(tiles, ggplot2::aes(lag, allele_row, fill = acf)) +
    ggplot2::geom_tile(width = 1, height = 1) +
    ggplot2::geom_point(data = stars, ggplot2::aes(lag, allele_row), inherit.aes = FALSE,
                       shape = 8, color = "#B8860B", size = 4) +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
      limits = c(-limit, limit), na.value = "#dddddd", name = "Median raw ACF") +
    ggplot2::scale_y_reverse(breaks = seq_along(rows), labels = labels, expand = c(0, 0)) +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::labs(title = paste0(region_ids, ": phased-allele median autocorrelation"),
      x = "Candidate repeat length / ACF lag (bp)", y = "Phased allele",
      caption = paste("120–250 bp; common color scale within this region. Stars: qualifying primary peaks. Gray: unavailable.",
        "Every unique retained fiber with a valid ACF contributes, including unclustered fibers.", sep = "\n")) +
    ggplot2::theme_bw() + ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0))
}
```

```{r function-plot-haplotype-repeats, include=TRUE}
plot_haplotype_repeats <- function(region, profiles, rows) {
  # Return figures for the notebook; rendering and filenames belong to knitr.
  list(
    median_acf = plot_haplotype_median_acf(rows, region),
    median_acf_zoom = plot_haplotype_median_acf(rows, region, lower = 100, upper = 500),
    fiber_heatmap = plot_haplotype_fiber_acf_heatmap(profiles, rows, region),
    fiber_heatmap_zoom = plot_haplotype_fiber_acf_heatmap(profiles, rows, region, upper = 500),
    repeat_boxplot = plot_haplotype_repeat_boxplot(rows, region),
    repeat_histogram = plot_haplotype_repeat_histogram(rows, region),
    allele_median_heatmap = plot_locus_heatmap(rows, region)
  )
}
```

```{r function-compact-locus-rows, include=TRUE}
compact_locus_rows <- function(rows) {
  lapply(rows, function(row) row[c("region_id", "allele", "base", "median", "primary", "n_valid")])
}
```

```{r load-region-results, include=TRUE, purl=TRUE}
missing_data <- region_dirs[!file.exists(file.path(region_dirs, "report_data.rds"))]
if (length(missing_data)) stop(
  "Direct plotting requires report_data.rds for each region. Missing: ",
  paste(basename(missing_data), collapse = ", "),
  ". Run 01_run_autocorrelation.py to save numerical inputs, or set output_dir to a prepared run.")
autocor_fiberseq_packages()
region_results <- lapply(region_dirs, read_report_data)
```

# Autocorrelation figures

## Allele-average ACF

```{r figure-allele-acf, results='asis', fig.height=10, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_acf")
  print(figure)
}
```

## Median ACF by focal allele

```{r figure-haplotype-median, results='asis', fig.height=8, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_median_acf(entry$repeats, entry$data$region)
  print(figure)
}
```

## Median ACF: nucleosome-scale view

```{r figure-haplotype-median-zoom, results='asis', fig.height=8, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_median_acf(entry$repeats, entry$data$region, lower = 100, upper = 500)
  print(figure)
}
```

## Allele median ACF: 120–250 bp

```{r figure-region-repeat-heatmap, results='asis', fig.height=5, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_locus_heatmap(entry$repeats, entry$data$region)
  print(figure)
}
```

## Individual-fiber ACF by allele

```{r figure-fiber-acf, results='asis', fig.height=9, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_fiber_acf_heatmap(entry$data$acf, entry$repeats, entry$data$region)
  print(figure)
}
```

## Individual-fiber ACF: first 500 bp

```{r figure-fiber-acf-zoom, results='asis', fig.height=9, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_fiber_acf_heatmap(entry$data$acf, entry$repeats, entry$data$region, upper = 500)
  print(figure)
}
```

## Per-fiber repeat-length candidates

```{r figure-repeat-boxplot, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_repeat_boxplot(entry$repeats, entry$data$region)
  print(figure)
}
```

## Per-fiber repeat-length frequencies

```{r figure-repeat-histogram, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_repeat_histogram(entry$repeats, entry$data$region)
  print(figure)
}
```

## Allele positive-peak fractions

```{r figure-allele-peak-features, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_peaks")
  print(figure)
}
```

## Cluster-average ACF

```{r figure-cluster-acf, results='asis', fig.height=7, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "cluster_acf")
  print(figure)
}
```

## ACF and genomic m6A heatmaps

```{r figure-acf-m6a-heatmap, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) autocor_acf_heatmap(result) + plot_genomic_cluster_heatmap(result, result$region, sample_palette(result$assignments$sample_name), include_haplotype = TRUE, show_cluster_profiles = TRUE))
  if (inherits(figure, c("Heatmap", "HeatmapList"))) {
    ComplexHeatmap::draw(figure, newpage = TRUE)
  } else {
    print(figure)
  }
}
```

## ACF UMAP and m6A call fraction

```{r figure-acf-umap, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_acf_umap(entry$data)
  print(figure)
}
```

## Saved ACF neighbor graph

```{r figure-knn-graph, results='asis', fig.height=9, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, plot_knn_graph)
  print(figure)
}
```

# Fiber-seq figures

## Read footprints

```{r figure-read-footprints, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_smf_reads(result, result$footprints, sample_palette(result$assignments$sample_name), include_haplotype = TRUE))
  print(figure)
}
```

## Genomic m6A heatmap

```{r figure-m6a-heatmap, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_genomic_cluster_heatmap(result, result$region, sample_palette(result$assignments$sample_name), include_haplotype = TRUE, show_cluster_profiles = TRUE))
  if (inherits(figure, c("Heatmap", "HeatmapList"))) {
    ComplexHeatmap::draw(figure, newpage = TRUE)
  } else {
    print(figure)
  }
}
```

## Footprint occupancy by cluster

```{r figure-occupancy, results='asis', fig.height=10, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_signal_profile(signal_profiles(result, result$footprints)$profiles, result$region, levels(result$assignments$cluster)))
  print(figure)
}
```

## Signal profiles and composition

```{r figure-fiberseq-composition, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) autocor_profile_composition(result)$plot)
  print(figure)
}
```

## Cluster proportions

```{r figure-cluster-proportions, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, autocor_cluster_proportions)
  print(figure)
}
```

## UMAP and allele overview

```{r figure-umap-overview, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) autocor_umap_overview(result)$plot)
  print(figure)
}
```

## Combined haplotype overview

```{r figure-haplotype-overview, results='asis', fig.height=18, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, autocor_haplotype_example)
  print(figure)
}
```

## Sample composition

```{r figure-sample-composition, results='asis', fig.height=7, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "sample_composition")
  print(figure)
}
```

## Allele coverage

```{r figure-allele-coverage, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_coverage")
  print(figure)
}
```

## Allele cluster proportions

```{r figure-allele-proportions, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_proportions")
  print(figure)
}
```

# Result tables

## Saved analysis parameters

```{r table-parameters, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(as.data.frame(entry$data$params), "Saved analysis parameters")
}
```

## Genomic region

```{r table-regions, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$region, "Genomic region")
}
```

## Allele coverage

```{r table-coverage, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$allele_coverage, "Allele coverage")
}
```

## Cluster summaries

```{r table-cluster-summary, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$cluster_summary, "Cluster summaries")
}
```

## Allele cluster proportions

```{r table-allele-clusters, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$allele_cluster_proportions, "Allele cluster proportions")
}
```

## Positive local peaks: 140–250 bp

```{r table-allele-peaks, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$allele_peak_features, "Positive local peaks: 140–250 bp")
}
```

## Repeat-length candidates: 120–250 bp

```{r table-repeat-summary, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(repeat_summary_table(entry$repeats), "Repeat-length candidates: 120–250 bp")
}
```

## Sample composition

```{r table-sample-composition, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$sample_composition, "Sample composition")
}
```

