---
title: "LCL Fiber-seq autocorrelation"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    mathjax: null
params:
  output_dir: "../../../LCL_project/auto_correlation/resolution04"
  regions: null
---

# Analysis and parameters

### Input


**Step 1: load m6A BED12 table**


| Read | RID | BED start | BED end | read_start | read_end | Score | Strand | blockCount | blockSizes  | blockStarts  |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | m84241_250908_180102_s3/160175746/ccs | 119282125 | 119298507 | 119282125 | 119298507 | 8 | - | 144 | 0,1,1,1,1, …, 1,1,1 | 0,37,452,618,1019, …, 15880,16060,16381 |
| B | m84241_250515_002859_s3/157422976/ccs | 119280507 | 119293438 | 119280507 | 119293438 | 7 | - | 248 | 0,1,1,1,1, …, 1,1,1 | 0,309,310,312,462, …, 12680,12873,12930 |
| C | m84241_250908_180102_s3/258540726/ccs | 119289060 | 119298334 | 119289060 | 119298334 | 18 | - | 266 | 0,1,1,1,1, …, 1,1,1 | 0,293,299,301,309, …, 9078,9080,9273 |
| D | m84241_250515_002859_s3/165545798/ccs | 119291372 | 119296368 | 119291372 | 119296368 | 32 | + | 120 | 0,1,1,1,1, …, 1,1,1 | 0,117,122,184,277, …, 4815,4821,4995 |

The first and last blocks and are removed


**Step 2: Expand blocks and keep calls inside the analysis window**

`extract_ft_region_reads()` retains the longest alignment per RID, expands BED12 blocks, removes the first and last blocks, and keeps calls within selected window

`m6a position = BED start + blockStarts + blockSizes`

For example, read A's call at 119291048 is `119282125 + 8922 + 1`.


| Read | Block number (1-based) | Block offset | Block size | Call position |
| --- | --- | --- | --- | --- |
| A | 82 | 7937 | 1 | 119290063 |
| A | 93 | 8922 | 1 | 119291048 |
| A | 94 | 8924 | 1 | 119291050 |
| A | 95 | 8926 | 1 | 119291052 |
| A | 96 | 8929 | 1 | 119291055 |
| B | 182 | 9563 | 1 | 119290071 |
| C | 43 | 1086 | 1 | 119290147 |
| C | 73 | 1987 | 1 | 119291048 |
| C | 74 | 1989 | 1 | 119291050 |
| C | 75 | 1991 | 1 | 119291052 |
| C | 76 | 1994 | 1 | 119291055 |
| C | 77 | 1998 | 1 | 119291059 |
| D | 2 | 117 | 1 | 119291490 |


**Step 3: Extract one metadata row per read**

`extract_ft_read_info(reads)` 

| Read | Chromosome | Start (1-based) | End (inclusive) | Strand | In-window m6A calls |
| --- | --- | --- | --- | --- | --- |
| A | chr2 | 119282126 | 119298507 | - | 26 |
| B | chr2 | 119280508 | 119293438 | - | 50 |
| C | chr2 | 119289061 | 119298334 | - | 81 |
| D | chr2 | 119291373 | 119296368 | + | 15 |

**Step 4: Construct the read-by-position matrix**


1. Initialize entries to `NA`.
2. Set positions covered by a read to `0`.
3. Set observed m6A-call positions to `1`.

`0` means covered without a reported m6A call; `NA` means uncovered.
Every genomic base is a column, including bases with no calls in any read.

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |


**Step 5: Filter coverage and phasing**

| Read | Spans full 2,000 bp? | Haplotype | Phase set | Focal genotype | Resolved allele | Outcome |
| --- | --- | --- | --- | --- | --- | --- |
| A | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| B | Yes | HP2 | 119099734 | 0\|1 | rs12470189: A | Keep |
| C | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| D | No | Not evaluated | — | — | — | Remove before phasing |

`allele_status = phased_focal_genotype` and pass the phased
heterozygote filter. Requires at least three retained
reads, both focal alleles, and at least two columns with observed m6A calls.


**Step 6: Final m6A matrix**

`lcl_build_phased_m6a_input()` builds all 2,000 genomic columns for the
retained reads

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |


The defaults are a **2,000-bp window and all 2,000 non-negative ACF lags**

The analysis accepts `--window-size` and `--n-features`


# Method


## Calculate autocorrelation

`ACF(k) = sum((x[t]-mean(x))*(x[t+k]-mean(x))) / (N * var(x))`

for each lag k (distance apart for a given comparison), it compares the centered signal with itself shifted by k bp. Positions with positive deviations after subtracting the mean give positive products. The differences are summed across all overlapping positions (positions that still line up after shifting by the lag), and divided by the read's total variance (variation in m6a pattern across all positions after subtracting the mean, used for normalization)

Autocorrelation calculates a ACF value measures the linear relationship and correlation coefficient between a time serious and a lagged version of itself at various time steps, specifically in Fiber-seq, it answers the question of how similar is the methylation pattern to itself after shifting it by k base pairs?

    - ACF(0) = 1, the pattern is compared to itself, so it matches perfectly
    - ACF(k) > 0, positions k bp apart tend to have the same state 
    - ACF(k) < 0, positions k bp apart tend to have opposite state
    - ACG(k) = 0, little systematic relationship at that spacing


Previous analysis using Fourier transformation does not give base pair resolution, where the resolution/period length is dependent on the frequency (k = 1,2,3…) and the length of the region, autocorrelation bypass this issue because:

  - FFT has uniform frequency resolution but not uniform period resolution -> how much of this molecule can be described by oscillations having the specific frequencies?
  - Autocorrelation samples uniformly in genomic distance
  - Autocorrelation represents the information on a distance axis, whereas the FFT represents it on an evenly spaced frequency axis

Constant molecules have undefined ACF and are excluded from clustering

<!-- set knitr options and find the region folders. Parameters: figures 12 x 8 in, 300 dpi, SVG; run folder = params$output_dir; regions = every rank* folder under outputs/, or only those in params$regions -->
```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
  out.width = "100%", fig.width = 12, fig.height = 8, dpi = 300,
  fig.retina = 1, dev = "svglite")
options(scipen = 999)
output <- normalizePath(params$output_dir, mustWork = TRUE)
region_dirs <- sort(list.dirs(file.path(output, "outputs"), recursive = FALSE, full.names = TRUE))
region_dirs <- region_dirs[grepl("^rank[0-9]+_", basename(region_dirs))]
if (!is.null(params$regions)) {
  missing <- setdiff(params$regions, basename(region_dirs))
  if (length(missing)) stop("Regions not found: ", paste(missing, collapse = ", "))
  region_dirs <- region_dirs[basename(region_dirs) %in% params$regions]
}
if (!length(region_dirs)) stop("No region results found under ", output)
```


<!-- load the shared Leiden_Manhattan plotting functions (read plot, m6A heatmap, signal profiles, UMAP, kNN graph, colors) so the figures use the same code, layout and fonts as the Leiden_Manhattan report -->
```{r source-leiden-plots, include=FALSE}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r", local = TRUE)
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r", local = TRUE)
```

RID = sample + read id
```{r empty-footprints-function, include=TRUE}
empty_footprints <- function() {
  data.frame(RID = character(), original_RID = character(), chr = character(),
    start = integer(), end = integer(), size = integer(), track = character())
}
```



##### example code to extract the nucleoosme footprints 
More on the output format at: https://fiberseq.github.io/fibertools/extracting/extract.html

```{r example-nuc-bed12, include=TRUE}
suppressPackageStartupMessages({
  library(readr)
  library(tidyverse)
})

nuc_bed12_ex <- read_tsv("/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL19_bc2138_18861/extracted_results/nuc_by_chr/AL19_bc2138_18861.ft_extracted_nuc.chr10.bed.gz", col_names = FALSE)

nuc_bed12_ex <- nuc_bed12_ex[1:1000,]

# Preserve this example's historical trailing-comma behavior: its old unnest
# created a blank block before removing the read's first and last entries.
# Represent blanks as temporary zero-size blocks for the shared converter.
nuc_bed12_parse <- as.data.frame(nuc_bed12_ex)
nuc_trailing <- grepl(",$", nuc_bed12_parse[[11L]])
stopifnot(identical(nuc_trailing, grepl(",$", nuc_bed12_parse[[12L]])))
nuc_bed12_parse[[10L]] <- as.integer(nuc_bed12_parse[[10L]]) + as.integer(nuc_trailing)
for (column in c(11L, 12L)) {
  nuc_bed12_parse[[column]][nuc_trailing] <-
    paste0(nuc_bed12_parse[[column]][nuc_trailing], "0")
}
nuc_blocks <- convert_ft_bed12_to_bed6(nuc_bed12_parse)
if (!nrow(nuc_blocks)) {
  nuc_blocks <- data.frame(chr = character(), start = numeric(), end = numeric(),
    RID = character(), strand = character())
}

# Recover the example's alignment metadata and input order. The shared converter
# groups blocks by read ID; each read's first/last entries are sentinel blocks.
# Keep the metadata in that same order before returning to the input row order.
nuc_block_rows <- rep(seq_len(nrow(nuc_bed12_ex)), nuc_bed12_parse[[10L]])
nuc_block_reads <- nuc_bed12_ex[[4L]][nuc_block_rows]
nuc_block_order <- which(duplicated(nuc_block_reads) &
  duplicated(nuc_block_reads, fromLast = TRUE))
# Follow the converter's actual group order, independent of the R locale.
nuc_block_order <- nuc_block_order[order(match(nuc_block_reads[nuc_block_order],
  unique(nuc_blocks$RID)))]
stopifnot(nrow(nuc_blocks) == length(nuc_block_order),
  identical(as.character(nuc_blocks$RID), as.character(nuc_block_reads[nuc_block_order])))
nuc_alignment_rows <- nuc_block_rows[nuc_block_order]
processed_bed12 <- tibble::tibble(
  chr = nuc_blocks$chr,
  start = nuc_bed12_ex[[2L]][nuc_alignment_rows],
  end = nuc_bed12_ex[[3L]][nuc_alignment_rows],
  read_id = nuc_blocks$RID,
  strand = nuc_blocks$strand,
  nuc_start = nuc_blocks$start + 1,
  nuc_end = nuc_blocks$end + 1,
  nuc_size = as.integer(nuc_blocks$end - nuc_blocks$start))
# Repeated read IDs can leave an earlier alignment's blank block inside the
# read group. Restore the missing coordinates and size produced by old unnest.
nuc_blank_blocks <- cumsum(nuc_bed12_parse[[10L]])[nuc_trailing]
nuc_missing <- nuc_block_order %in% nuc_blank_blocks
processed_bed12[nuc_missing, c("nuc_start", "nuc_end", "nuc_size")] <- NA
processed_bed12 <- processed_bed12[order(nuc_block_order), ]
head(processed_bed12, 20)
  
```

#### build nucleosome footprint track

builds the nucleosome footprint track for each clustered read in the autocorrelation

  - Keeps footprints that are 130–160 bp

```{r prepare-nucleosomes-function, include=TRUE}
prepare_nucleosomes <- function(region, assignments, ft_root) {
  output <- list()
  query <- GenomicRanges::GRanges(region$chr,
    IRanges::IRanges(region$analysis_start, region$analysis_end))
  for (sample in unique(assignments$sample_name)) {
    path <- file.path(ft_root, sample, "extracted_results", "nuc_by_chr",
      paste0(sample, ".ft_extracted_nuc.", region$chr, ".bed.gz"))
    if (!file.exists(path) || !file.exists(paste0(path, ".tbi")))
      stop("Missing indexed nucleosome BED: ", path)
    lines <- unlist(Rsamtools::scanTabix(Rsamtools::TabixFile(path), param = query),
      use.names = FALSE)
    if (!length(lines)) next
    bed <- utils::read.table(text = lines, sep = "\t", header = FALSE,
      quote = "", comment.char = "", stringsAsFactors = FALSE)
    stopifnot(ncol(bed) == 12L)
    names(bed) <- c("chr", "start", "end", "RID", "score", "strand",
      "read_start", "read_end", "rgb", "blockCount", "blockSizes", "blockStarts")
    bed <- bed[order(bed$RID, -(bed$end - bed$start)), , drop = FALSE]
    bed <- bed[!duplicated(bed$RID), , drop = FALSE]
    bed <- bed[bed$RID %in% assignments$original_RID[assignments$sample_name == sample], , drop = FALSE]
    for (i in seq_len(nrow(bed))) {
      sizes <- as.integer(strsplit(sub(",$", "", bed$blockSizes[i]), ",", fixed = TRUE)[[1]])
      offsets <- as.integer(strsplit(sub(",$", "", bed$blockStarts[i]), ",", fixed = TRUE)[[1]])
      stopifnot(length(sizes) == bed$blockCount[i], length(offsets) == length(sizes),
        !anyNA(sizes), !anyNA(offsets))
      keep <- if (length(sizes) > 2L) seq.int(2L, length(sizes) - 1L) else integer()
      starts <- bed$start[i] + offsets[keep]
      ends <- starts + sizes[keep]
      valid <- sizes[keep] >= 130L & sizes[keep] <= 160L &
        starts < region$end & ends > region$start
      if (!any(valid)) next
      output[[length(output) + 1L]] <- data.frame(
        RID = paste(sample, bed$RID[i], sep = "::"), original_RID = bed$RID[i],
        chr = region$chr, start = starts[valid], end = ends[valid],
        
        size = sizes[keep][valid], track = "ft_nuc_130-160bp")
    }
  }
  if (length(output)) do.call(rbind, output) else empty_footprints()
}
```


#### build footprint tracks from FiberHMM

10–30, 40–60, and 60–80 bp size footprints

```{r prepare-fiberhmm-function, include=TRUE}
prepare_fiberhmm_track <- function(region, assignments, size, project, tracks_root) {
  filename <- paste0("combined_", region$chr, "_", size, "bp_fps.bed.gz")
  cache <- file.path(project, "LCL_project/Leiden_manhattan/footprint_summary",
    paste0(filename, ".rds"))
  source <- file.path(tracks_root, region$chr, filename)
  cached <- if (file.exists(cache)) readRDS(cache) else NULL
  bounds <- if (!is.null(cached)) cached$signature$regions else NULL
  covered <- !is.null(bounds) && any(bounds$chr == region$chr &
    bounds$start <= region$start & bounds$end >= region$end)
  if (covered) {
    records <- cached$records
    used <- cache
  } else {
    if (!file.exists(source)) stop("Missing footprint source: ", source)
    program <- sprintf('BEGIN {FS=OFS="\t"} $1 == "%s" && $2 < %.0f && $3 > %.0f {print}',
      region$chr, region$end, region$start)
    command <- paste("gzip -cd", shQuote(source), "| awk", shQuote(program))
    lines <- system2("/bin/bash", c("-o", "pipefail", "-c", shQuote(command)), stdout = TRUE)
    if (!is.null(attr(lines, "status")) && attr(lines, "status") != 0L)
      stop("Footprint extraction failed: ", source)
    records <- if (length(lines)) utils::read.table(text = lines, sep = "\t",
      header = FALSE, quote = "", comment.char = "", stringsAsFactors = FALSE) else
      data.frame(chr = character(), start = integer(), end = integer(), original_RID = character())
    stopifnot(ncol(records) == 4L)
    names(records) <- c("chr", "start", "end", "original_RID")
    used <- source
  }
  records <- records[records$chr == region$chr & records$start < region$end &
    records$end > region$start, , drop = FALSE]
  records <- merge(records, assignments[, c("original_RID", "RID")],
    by = "original_RID", sort = FALSE)
  records$size <- records$end - records$start
  records$track <- rep(paste0("FiberHMM_", size, "bp"), nrow(records))
  list(records = records[, names(empty_footprints()), drop = FALSE],
    source = data.frame(track = paste0("FiberHMM_", size, "bp"), source = used,
      n_records = nrow(records)))
}
```

run autocorrelation, and get nucleeosomes and footprints
```{r prepare-report-data-function, include=FALSE}
prepare_report_data <- function(region_data, project, destination,
    ft_root = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
    tracks_root = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks") {
  stopifnot(region_data$schema_version == 1L, is.data.frame(region_data$region),
    nrow(region_data$region) == 1L, is.data.frame(region_data$records),
    identical(region_data$row_ids, region_data$records$row_id), !anyDuplicated(region_data$row_ids))
  region_data$acf <- autocor_payload_matrix(region_data$acf, nrow(region_data$records))
  stopifnot(nrow(region_data$acf) == nrow(region_data$records),
    ncol(region_data$acf) == region_data$params$n_features)
  a <- region_data$records[region_data$records$status == "clustered", , drop = FALSE]
  a$original_RID <- a$RID
  a$RID <- paste(a$sample_name, a$original_RID, sep = "::")
  stopifnot(!anyDuplicated(a$RID))
  region_data$footprints <- empty_footprints()
  region_data$footprint_tracks <- character()
  region_data$footprint_sources <- data.frame(track = character(), source = character(), n_records = integer())
  if (nrow(a)) {
    nucleosomes <- prepare_nucleosomes(region_data$region, a, ft_root)
    tracks <- lapply(c("10-30", "40-60", "60-80"), function(size)
      prepare_fiberhmm_track(region_data$region, a, size, project, tracks_root))
    region_data$footprints <- do.call(rbind, c(list(nucleosomes), lapply(tracks, `[[`, "records")))
    region_data$footprint_tracks <- c("ft_nuc_130-160bp",
      paste0("FiberHMM_", c("10-30", "40-60", "60-80"), "bp"))
    region_data$footprint_sources <- rbind(data.frame(track = "ft_nuc_130-160bp",
      source = ft_root, n_records = nrow(nucleosomes)), do.call(rbind, lapply(tracks, `[[`, "source")))
  }
  region_data$footprints_available <- length(region_data$footprint_tracks) > 0L
  dir.create(dirname(destination), recursive = TRUE, showWarnings = FALSE)
  temporary <- tempfile("report-data-", tmpdir = dirname(destination))
  on.exit(unlink(temporary), add = TRUE)
  saveRDS(region_data, temporary, compress = "gzip")
  if (!file.rename(temporary, destination)) stop("Could not save report data: ", destination)
  message("Saved portable report data: ", destination)
  invisible(destination)
}
```

<!-- load one region's report_data.rds, check that reads and ACF rows line up, number the ACF clusters from 1 as Leiden_Manhattan does (saved Scanpy cluster 0, the largest, becomes cluster 1) in the reads and every summary table, compute the repeat-length candidates and build the plotting object -->
```{r read-report-data-function, include=FALSE}
read_report_data <- function(folder) {
  path <- file.path(folder, "report_data.rds")
  if (!file.exists(path)) stop("Missing analysis data for direct plotting: ", path)
  region_data <- readRDS(path)
  stopifnot(identical(as.integer(region_data$schema_version), 1L),
    nrow(region_data$region) == 1L, region_data$region$region_id == basename(folder),
    identical(region_data$row_ids, region_data$records$row_id),
    is.matrix(region_data$acf), nrow(region_data$acf) == nrow(region_data$records))
  renumber <- function(table) {
    if (is.data.frame(table) && "cluster" %in% names(table)) {
      id <- suppressWarnings(as.integer(as.character(table$cluster)))
      table$cluster <- ifelse(is.na(id), as.character(table$cluster), as.character(id + 1L))
    }
    table
  }
  region_data$records <- renumber(region_data$records)
  region_data$tables <- lapply(region_data$tables, renumber)
  rows <- summarize_repeats(region_data$acf, region_data$records, region_data$region, region_data$tables)
  list(folder = folder, region_id = region_data$region$region_id, data = region_data,
    repeats = rows, fiberseq = autocor_fiberseq_result(region_data))
}
```

<!-- draw a Fiber-seq figure for one region, or a 'No clustered reads' placeholder when the region has no clustered reads -->
```{r report-fiberseq-figure-function, include=FALSE}
report_fiberseq_figure <- function(entry, draw) {
  if (is.null(entry$fiberseq)) return(ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0, y = 0, label = "No clustered reads in this region") +
    ggplot2::labs(title = entry$region_id) + ggplot2::theme_void())
  draw(entry$fiberseq)
}
```

<!-- print a table as HTML, or 'no rows' when it is empty. Parameters: 4 digits -->
```{r print-report-table-function, include=FALSE}
print_report_table <- function(table, caption) {
  if (is.null(table) || !is.data.frame(table) || !nrow(table)) {
    cat(caption, ": no rows.\n\n", sep = "")
  } else {
    cat(knitr::kable(table, caption = caption, row.names = FALSE, digits = 4,
      format = "html", escape = TRUE), sep = "\n")
    cat("\n\n")
  }
  invisible(NULL)
}
```

<!-- blank plot that shows only a message -->
```{r report-empty-plot-function, include=FALSE}
report_empty_plot <- function(message, title = NULL) {
  ggplot2::ggplot() + ggplot2::annotate("text", x = 0, y = 0, label = message) +
    ggplot2::labs(title = title) + ggplot2::theme_void()
}
```

<!-- axis labels for the two alleles: REF (base) and ALT (base) -->
```{r report-allele-labels-function, include=FALSE}
report_allele_labels <- function(region_data) {
  c(REF = paste0("REF (", region_data$region$ref, ")"), ALT = paste0("ALT (", region_data$region$alt, ")"))
}
```

<!-- add the region / focal SNP title and an optional caption around a figure -->
```{r report-figure-frame-function, include=FALSE}
report_figure_frame <- function(plot, region_data, caption = NULL) {
  title <- paste0(region_data$region$region_id, "\n", region_data$region$focal_snp, ": ",
    region_data$region$ref, " / ", region_data$region$alt)
  heading <- cowplot::ggdraw() + cowplot::draw_label(title, size = 11)
  if (is.null(caption)) return(cowplot::plot_grid(heading, plot, ncol = 1, rel_heights = c(.12, 1)))
  footer <- cowplot::ggdraw() + cowplot::draw_label(caption, size = 9)
  cowplot::plot_grid(heading, plot, footer, ncol = 1, rel_heights = c(.12, 1, .13))
}
```

<!-- plot the mean ACF of each cluster as curves and as a heatmap. Parameters: lag 0 omitted, no smoothing; lower/upper set the lag range (default all nonzero lags) -->
```{r report-cluster-acf-function, include=FALSE}
plot_report_cluster_acf <- function(region_data, lower = 1, upper = Inf, title = "Cluster averages; no smoothing") {
  d <- region_data$tables$cluster_acf_summary
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("No clustered reads", region_data$region$region_id))
  d <- d[d$lag_bp >= lower & d$lag_bp <= upper, , drop = FALSE]
  if (!nrow(d)) return(report_empty_plot("No nonzero ACF lags available", region_data$region$region_id))
  groups <- as.character(sort(unique(as.integer(as.character(d$cluster)))))
  d$cluster <- factor(d$cluster, levels = groups)
  colors <- cluster_id_palette(groups)
  counts <- d$n_reads[match(groups, d$cluster)]
  curves <- ggplot2::ggplot(d, ggplot2::aes(lag_bp, mean_acf, color = cluster)) +
    ggplot2::geom_hline(yintercept = 0, color = "grey60", linewidth = .3) +
    ggplot2::geom_line(linewidth = .4) +
    ggplot2::scale_color_manual(values = colors, labels = paste0("C", groups, " (n=", counts, ")")) +
    ggplot2::labs(title = title, x = "Lag (bp)", y = "Mean ACF", color = "Cluster") +
    ggplot2::theme_bw(base_size = 10)
  bound <- max(c(abs(d$mean_acf[is.finite(d$mean_acf)]), .01))
  heatmap <- ggplot2::ggplot(d, ggplot2::aes(lag_bp, cluster, fill = mean_acf)) +
    ggplot2::geom_tile(width = 1, height = 1) +
    ggplot2::scale_y_discrete(limits = rev(groups), labels = paste0("C", rev(groups))) +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-bound, bound), name = "Mean ACF") +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::labs(x = "Lag (bp)", y = "Cluster") + ggplot2::theme_bw(base_size = 10)
  report_figure_frame(cowplot::plot_grid(curves, heatmap, ncol = 1), region_data)
}
```

<!-- stacked bars: fraction of each sample's clustered reads in each cluster -->
```{r report-sample-composition-function, include=FALSE}
plot_report_sample_composition <- function(region_data) {
  d <- region_data$tables$sample_composition
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("No clustered reads", region_data$region$region_id))
  groups <- as.character(sort(unique(as.integer(as.character(d$cluster)))))
  d$cluster <- factor(d$cluster, levels = groups)
  ggplot2::ggplot(d, ggplot2::aes(sample_name, fraction_within_sample, fill = cluster)) +
    ggplot2::geom_col(position = ggplot2::position_stack(reverse = TRUE)) +
    ggplot2::scale_fill_manual(values = cluster_id_palette(groups)) +
    ggplot2::scale_y_continuous(limits = c(0, 1), expand = ggplot2::expansion(mult = c(0, .01))) +
    ggplot2::labs(title = region_data$region$region_id, x = "Sample", y = "Fraction of clustered molecules", fill = "Cluster") +
    ggplot2::theme_bw(base_size = 10) + ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = .5))
}
```

<!-- bar plots of the number of valid ACF reads and clustered reads for REF and ALT -->
```{r report-allele-coverage-function, include=FALSE}
plot_report_allele_coverage <- function(region_data) {
  d <- region_data$tables$allele_coverage
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("Allele coverage unavailable", region_data$region$region_id))
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  columns <- c(n_valid = "Valid ACF reads", n_clustered = "Clustered reads")
  plots <- lapply(names(columns), function(column) {
    values <- d[[column]]
    maximum <- max(c(values[is.finite(values)], 1))
    ggplot2::ggplot(d, ggplot2::aes(allele_group, .data[[column]], fill = allele_group)) +
      ggplot2::geom_col() + ggplot2::geom_text(ggplot2::aes(label = .data[[column]]), vjust = -.4) +
      ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), guide = "none") +
      ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = report_allele_labels(region_data), drop = FALSE) +
      ggplot2::scale_y_continuous(limits = c(0, maximum * 1.18), expand = c(0, 0)) +
      ggplot2::labs(title = columns[[column]], x = NULL, y = "Unique molecules") + ggplot2::theme_bw(base_size = 10)
  })
  report_figure_frame(cowplot::plot_grid(plotlist = plots, nrow = 1), region_data)
}
```

<!-- plot the mean ACF of REF and ALT reads in three lag ranges. Parameters: all lags (lag 0 omitted), 0–500 bp, 140–250 bp; giving lower/upper draws that single lag range instead -->
```{r report-allele-acf-function, include=FALSE}
plot_report_allele_acf <- function(region_data, lower = NULL, upper = NULL) {
  d <- region_data$tables$allele_acf_summary
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("Allele ACF summaries unavailable", region_data$region$region_id))
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  maximum <- max(d$lag_bp)
  coverage <- region_data$tables$allele_coverage
  missing <- c("REF", "ALT")[!c("REF", "ALT") %in% as.character(coverage$allele_group[coverage$n_valid > 0])]
  bounds <- list(c(1, maximum), c(0, 500), c(140, 250))
  headings <- c("All retained lags (lag 0 omitted)", "Lags 0–500 bp", "Lags 140–250 bp")
  if (!is.null(lower)) {
    bounds <- list(c(lower, upper))
    headings <- paste0("Lags ", lower, "–", upper, " bp")
  }
  plots <- lapply(seq_along(bounds), function(i) {
    lower <- bounds[[i]][1]; upper <- max(lower + 1, bounds[[i]][2])
    values <- d[d$lag_bp >= lower & d$lag_bp <= upper, , drop = FALSE]
    values$mean_acf <- as.numeric(values$mean_acf)
    note <- if (maximum < lower) "Lag range not available" else if (length(missing))
      paste("No valid ACF reads:", paste(missing, collapse = ", ")) else NULL
    ggplot2::ggplot(values, ggplot2::aes(lag_bp, mean_acf, color = allele_group)) +
      ggplot2::geom_line(linewidth = .4, na.rm = TRUE) +
      ggplot2::scale_color_manual(values = c(REF = "#2878B5", ALT = "#D55E00"),
        limits = c("REF", "ALT"), labels = report_allele_labels(region_data), drop = FALSE) +
      ggplot2::scale_x_continuous(limits = c(lower, upper)) +
      ggplot2::labs(title = headings[i], subtitle = note, x = "Lag (bp)", y = "Mean ACF", color = "Focal allele") +
      ggplot2::theme_bw(base_size = 10)
  })
  panels <- if (length(plots) == 1L) plots[[1]] else
    cowplot::plot_grid(plots[[1]], cowplot::plot_grid(plots[[2]], plots[[3]], nrow = 1), ncol = 1)
  report_figure_frame(panels, region_data, paste0("Available lags: 0–", maximum,
    " bp. Each panel uses the same pooled REF/ALT means; vertical scales are independent."))
}
```

<!-- bar plot: fraction of each allele's clustered reads in each ACF cluster -->
```{r report-allele-proportions-function, include=FALSE}
plot_report_allele_proportions <- function(region_data) {
  d <- region_data$tables$allele_cluster_proportions
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("No clustered molecules", region_data$region$region_id))
  d$cluster <- factor(d$cluster, levels = as.character(sort(unique(as.integer(as.character(d$cluster))))))
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  d$fraction <- as.numeric(d$fraction)
  coverage <- region_data$tables$allele_coverage
  missing <- c("REF", "ALT")[!c("REF", "ALT") %in% as.character(coverage$allele_group[coverage$n_clustered > 0])]
  p <- ggplot2::ggplot(d, ggplot2::aes(cluster, fraction, fill = allele_group)) +
    ggplot2::geom_col(position = "dodge", na.rm = TRUE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"),
      limits = c("REF", "ALT"), labels = report_allele_labels(region_data), drop = FALSE) +
    ggplot2::scale_y_continuous(limits = c(0, 1.05)) +
    ggplot2::labs(x = "ACF cluster", y = "Fraction of clustered molecules within allele", fill = "Focal allele",
      subtitle = if (length(missing)) paste("No clustered reads:", paste(missing, collapse = ", "))) +
    ggplot2::theme_bw(base_size = 10)
  report_figure_frame(p, region_data)
}
```

<!-- boxplot by allele of one per-read peak value (peak lag or peak ACF height). Parameters: lag axis 135–255 bp -->
```{r report-allele-peak-distribution-function, include=FALSE}
plot_report_peak_distribution <- function(audit, region_data, column, ylabel) {
  audit$allele_group <- factor(audit$allele_group, levels = c("REF", "ALT"))
  audit[[column]] <- as.numeric(audit[[column]])
  values <- audit[is.finite(audit[[column]]), , drop = FALSE]
  missing <- setdiff(c("REF", "ALT"), as.character(unique(values$allele_group)))
  note_y <- if (column == "positive_local_peak_140_250_bp") 195 else if (nrow(values))
    mean(range(values[[column]])) else .5
  note <- data.frame(allele_group = factor(missing, levels = c("REF", "ALT")),
    value = rep(note_y, length(missing)))
  p <- ggplot2::ggplot(values, ggplot2::aes(allele_group, .data[[column]], fill = allele_group)) +
    ggplot2::geom_boxplot(width = .45, alpha = .65, na.rm = TRUE) +
    ggplot2::geom_text(data = note, ggplot2::aes(allele_group, value), label = "No peaks", inherit.aes = FALSE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), guide = "none") +
    ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = report_allele_labels(region_data), drop = FALSE) +
    ggplot2::labs(x = NULL, y = ylabel) + ggplot2::theme_bw(base_size = 10)
  if (column == "positive_local_peak_140_250_bp") p <- p + ggplot2::scale_y_continuous(limits = c(135, 255))
  p
}
```

<!-- plot the fraction of reads with a positive local ACF peak and the peak lag / height distributions by allele. Parameters: peak band 140–250 bp, needs at least 252 lags -->
```{r report-allele-peaks-function, include=FALSE}
plot_report_allele_peaks <- function(region_data) {
  d <- region_data$tables$allele_peak_features
  if (!is.data.frame(d) || !nrow(d)) return(report_empty_plot("Allele peak summaries unavailable", region_data$region$region_id))
  if (!all(d$peak_band_complete %in% TRUE)) {
    plots <- rep(list(report_empty_plot("140–250 bp peak range\nnot fully available")), 3L)
    return(report_figure_frame(cowplot::plot_grid(plotlist = plots, nrow = 1), region_data))
  }
  d$allele_group <- factor(d$allele_group, levels = c("REF", "ALT"))
  d$peak_fraction <- as.numeric(d$peak_fraction)
  d$value_label <- ifelse(is.finite(d$peak_fraction), sprintf("%.2f", d$peak_fraction), "")
  missing <- d[d$n_valid == 0, , drop = FALSE]
  fraction <- ggplot2::ggplot(d, ggplot2::aes(allele_group, peak_fraction, fill = allele_group)) +
    ggplot2::geom_col(na.rm = TRUE) + ggplot2::geom_text(ggplot2::aes(label = value_label), vjust = -.5, na.rm = TRUE) +
    ggplot2::geom_text(data = missing, ggplot2::aes(allele_group), y = .5, label = "No valid reads", inherit.aes = FALSE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), guide = "none") +
    ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = report_allele_labels(region_data), drop = FALSE) +
    ggplot2::scale_y_continuous(limits = c(0, 1.15)) +
    ggplot2::labs(x = NULL, y = "Fraction with a qualifying peak") + ggplot2::theme_bw(base_size = 10)
  audit <- region_data$tables$allele_read_audit
  audit <- audit[audit$allele_analysis_included & audit$acf_valid, , drop = FALSE]
  lag <- plot_report_peak_distribution(audit, region_data, "positive_local_peak_140_250_bp", "Molecule peak lag (bp)")
  height <- plot_report_peak_distribution(audit, region_data, "peak_acf", "Molecule peak ACF height")
  report_figure_frame(cowplot::plot_grid(fraction, lag, height, nrow = 1), region_data,
    "Peak fraction: all valid reads. Distributions: peak-bearing reads only; boxes show median and quartiles.")
}
```

<!-- pick a summary figure by name: cluster_acf, sample_composition, allele_coverage, allele_acf, allele_proportions, allele_peaks -->
```{r report-table-plot-function, include=FALSE}
plot_report_table <- function(region_data, kind) {
  switch(kind,
    cluster_acf = plot_report_cluster_acf(region_data),
    sample_composition = plot_report_sample_composition(region_data),
    allele_coverage = plot_report_allele_coverage(region_data),
    allele_acf = plot_report_allele_acf(region_data),
    allele_proportions = plot_report_allele_proportions(region_data),
    allele_peaks = plot_report_allele_peaks(region_data),
    stop("Unknown summary figure: ", kind))
}
```

<!-- UMAP of clustered reads colored by ACF cluster and by m6A call fraction -->
```{r plot-acf-umap-function, include=FALSE}
plot_acf_umap <- function(region_data) {
  a <- region_data$records[region_data$records$status == "clustered", , drop = FALSE]
  if (!nrow(a)) return(ggplot2::ggplot() + ggplot2::annotate("text", x = 0, y = 0,
    label = "No clustered reads") + ggplot2::labs(title = region_data$region$region_id) +
    ggplot2::theme_void())
  clusters <- ggplot2::ggplot(a, ggplot2::aes(umap1, umap2, color = factor(cluster))) +
    ggplot2::geom_point(size = 1) + ggplot2::labs(title = "ACF clusters", x = "UMAP 1",
      y = "UMAP 2", color = "Cluster") + ggplot2::theme_bw(base_size = 10)
  methylation <- ggplot2::ggplot(a, ggplot2::aes(umap1, umap2, color = m6a_call_fraction)) +
    ggplot2::geom_point(size = 1) + ggplot2::scale_color_viridis_c() +
    ggplot2::labs(title = region_data$region$region_id, x = "UMAP 1", y = "UMAP 2",
      color = "m6A call fraction") + ggplot2::theme_bw(base_size = 10)
  cowplot::plot_grid(clusters, methylation, ncol = 2)
}
```

<!-- convert the saved ACF values into a numeric reads x lags matrix; missing values become NA -->
```{r function-autocor-payload-matrix, include=FALSE}
autocor_payload_matrix <- function(values, n_reads) {
  if (is.matrix(values) || is.data.frame(values)) {
    result <- as.matrix(values)
  } else if (is.list(values)) {
    rows <- lapply(values, function(row) {
      vapply(row, function(value) if (is.null(value)) NA_real_ else as.numeric(value), numeric(1))
    })
    if (length(unique(lengths(rows))) > 1L) stop("ACF rows have inconsistent lengths")
    result <- do.call(rbind, rows)
  } else {
    result <- matrix(values, nrow = n_reads, byrow = TRUE)
  }
  if (is.null(result) || nrow(result) != n_reads) stop("ACF and saved molecule IDs do not align")
  storage.mode(result) <- "double"
  result
}
```

<!-- build the plotting object for the clustered reads: sparse m6A matrix, ACF matrix, kNN graph, footprints and focal SNP. Parameters: clustered phased heterozygous reads only (HP1/HP2, genotype 0|1 or 1|0); ACF at lag 0 must equal 1 -->
```{r function-autocor-fiberseq-result, include=FALSE}
autocor_fiberseq_result <- function(region_data) {
  region <- as.data.frame(region_data$region, stringsAsFactors = FALSE)
  stopifnot(nrow(region) == 1L, region$width == region$analysis_end - region$analysis_start + 1L)
  records <- as.data.frame(region_data$records, stringsAsFactors = FALSE)
  a <- records[records$status == "clustered" & !is.na(records$status), , drop = FALSE]
  if (!nrow(a)) return(NULL)
  stopifnot(all(a$region_id == region$region_id), !anyDuplicated(a$row_id),
    all(a$haplotype %in% c("HP1", "HP2")), all(a$focal_genotype %in% c("0|1", "1|0")),
    all(a$allele_status == "phased_focal_genotype"))
  a$original_RID <- a$RID
  a$RID <- paste(a$sample_name, a$original_RID, sep = "::")
  a$start <- a$read_start
  a$end <- a$read_end
  a$chr <- region$chr
  a$sample_label <- sub("_.*$", "", a$sample_name)
  # Cluster names as in Leiden_Manhattan: cluster1, cluster2, ...
  a$cluster <- factor(paste0("cluster", a$cluster),
    levels = paste0("cluster", sort(unique(as.integer(a$cluster)))))
  a$heatmap_included <- TRUE
  ids <- unlist(region_data$row_ids, use.names = FALSE)
  matched <- match(a$row_id, ids)
  stopifnot(!anyDuplicated(ids), !anyNA(matched), !anyDuplicated(a$RID), !anyNA(a$cluster))
  offsets <- if (is.matrix(region_data$call_offsets)) asplit(region_data$call_offsets, 1L) else region_data$call_offsets
  stopifnot(length(offsets) == length(ids))
  offsets <- lapply(offsets[matched], function(row) as.integer(unlist(row, use.names = FALSE)))
  all_offsets <- unlist(offsets, use.names = FALSE)
  stopifnot(!anyNA(all_offsets), all(all_offsets >= 0L & all_offsets < region$width),
    !any(vapply(offsets, anyDuplicated, integer(1)) > 0L))
  met <- Matrix::sparseMatrix(i = rep(seq_len(nrow(a)), lengths(offsets)),
    j = all_offsets + 1L, x = rep(1, length(all_offsets)), dims = c(nrow(a), region$width),
    dimnames = list(a$RID, as.character(seq.int(region$analysis_start, region$analysis_end))))
  # As in Leiden_Manhattan: columns are the m6A sites observed in any retained read.
  observed <- sort(unique(as.integer(unlist(if (is.matrix(region_data$call_offsets))
    asplit(region_data$call_offsets, 1L) else region_data$call_offsets, use.names = FALSE)))) + 1L
  met <- met[, observed, drop = FALSE]
  acf <- autocor_payload_matrix(region_data$acf, length(ids))[matched, , drop = FALSE]
  stopifnot(ncol(acf) >= 1L, ncol(acf) <= region$width,
    all(is.finite(acf)), all(abs(acf[, 1] - 1) < 1e-8))
  rownames(acf) <- a$RID
  edges <- as.data.frame(region_data$edges, stringsAsFactors = FALSE)
  if (!nrow(edges)) edges <- data.frame(source = character(), target = character(), weight = numeric())
  edges$source <- a$RID[match(edges$source, a$row_id)]
  edges$target <- a$RID[match(edges$target, a$row_id)]
  stopifnot(!anyNA(edges), all(is.finite(edges$weight)), all(edges$weight > 0))
  graph <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = a$RID)
  footprints <- region_data$footprints
  if (is.null(footprints) || !length(footprints)) footprints <- data.frame(RID = character(),
    original_RID = character(), chr = character(), start = integer(), end = integer(),
    size = integer(), track = character())
  footprints <- as.data.frame(footprints, stringsAsFactors = FALSE)
  footprints <- footprints[footprints$RID %in% a$RID, , drop = FALSE]
  tracks <- if (is.null(region_data$footprint_tracks)) unique(footprints$track) else unlist(region_data$footprint_tracks)
  focal <- data.frame(chr = region$chr, pos = region$focal_pos, variant_id = region$focal_snp,
    ref = region$ref, alt = region$alt)
  result <- list(assignments = a, region = region, site_met_mat = met, acf = acf, graph = graph,
    n_clusters = nlevels(a$cluster), variants = focal, focal = focal,
    params = if (is.null(region_data$params)) list(seed = 0L) else region_data$params,
    footprints = footprints, footprint_tracks = intersect(LCL_FOOTPRINT_TRACKS, tracks))
  fiberseq_display_result(result)
}
```

<!-- heatmap of each read's ACF, rows split by cluster and ordered by read start. Parameters: lag 0 omitted, color limits at the 99th percentile of |ACF| -->
```{r function-autocor-acf-heatmap, include=FALSE}
autocor_acf_heatmap <- function(result) {
  a <- result$assignments
  ordering <- order(a$cluster, a$start, a$RID)
  display <- result$acf[a$RID[ordering], -1, drop = FALSE]
  if (!ncol(display)) stop("The ACF heatmap requires at least one nonzero lag")
  limit <- max(as.numeric(stats::quantile(abs(display), .99)), .01)
  ticks <- unique(round(seq(1, ncol(display), length.out = min(5L, ncol(display)))))
  ComplexHeatmap::Heatmap(display, name = "ACF",
    col = circlize::colorRamp2(c(-limit, 0, limit), c("#2166ac", "white", "#b2182b")),
    cluster_rows = FALSE, cluster_columns = FALSE, cluster_row_slices = FALSE,
    row_split = a$cluster[ordering], show_row_names = FALSE, show_column_names = FALSE,
    use_raster = TRUE, raster_quality = 4, raster_resize_mat = FALSE,
    raster_device = "png", raster_by_magick = FALSE,
    column_title = "Autocorrelation: lag (bp)",
    bottom_annotation = ComplexHeatmap::HeatmapAnnotation(lag_bp = ComplexHeatmap::anno_mark(
      at = ticks, labels = as.character(ticks), which = "column", side = "bottom")),
    width = grid::unit(65, "mm"))
}
```

<!-- combine UMAPs colored by cluster, LCL sample and focal SNP allele with the cluster pie chart -->
```{r function-autocor-umap-overview, include=FALSE}
autocor_umap_overview <- function(result) {
  embedding <- result$assignments
  embedding$UMAP1 <- embedding$umap1
  embedding$UMAP2 <- embedding$umap2
  stopifnot(all(is.finite(embedding$UMAP1)), all(is.finite(embedding$UMAP2)))
  short_colors <- sample_palette(embedding$sample_name)
  samples <- unique(embedding$sample_name)
  sample_colors <- setNames(unname(short_colors[sub("_.*$", "", samples)]), samples)
  groups <- levels(embedding$cluster)
  cluster_colors <- cluster_id_palette(groups)
  allele_colors <- fiberseq_category_palette(result$allele_display_levels)
  cluster_labels <- paste0(groups, " (n=", as.integer(table(embedding$cluster)), ")")
  cluster_plot <- plot_umap(embedding, "cluster", cluster_colors, "UMAP: cluster")
  sample_plot <- plot_umap(embedding, "sample_name", sample_colors, "UMAP: LCL sample")
  allele_plot <- plot_umap(embedding, "allele_display", allele_colors, "UMAP: Focal SNP allele")
  legends <- list(fiberseq_legend(cluster_colors, "Cluster", 5, cluster_labels),
    fiberseq_legend(sample_colors, "LCL sample", 6, sub("_.*$", "", names(sample_colors))),
    fiberseq_legend(allele_colors, "Focal SNP allele", 3))
  heights <- vapply(legends, fiberseq_legend_height, numeric(1))
  heading <- cowplot::ggdraw() + cowplot::draw_label(result$region$annotation, size = 12)
  plot <- cowplot::plot_grid(plotlist = c(list(heading,
    cowplot::plot_grid(cluster_plot, sample_plot, allele_plot, plot_cluster_pie(result), ncol = 2)), legends),
    ncol = 1, rel_heights = c(.45, 8, heights))
  list(plot = plot, height = 8.45 + sum(heights))
}
```

<!-- per-cluster m6A and nucleosome profiles with sample and allele composition bars -->
```{r function-autocor-profile-composition, include=FALSE}
autocor_profile_composition <- function(result) {
  short_colors <- sample_palette(result$assignments$sample_name)
  samples <- unique(result$assignments$sample_name)
  sample_colors <- setNames(unname(short_colors[sub("_.*$", "", samples)]), samples)
  tables <- fiberseq_tables(result, result$footprints)
  plot_fiberseq_profiles(tables, result$region, result$markers,
    cluster_id_palette(tables$groups), sample_colors,
    fiberseq_category_palette(result$allele_display_levels), result$region$annotation)
}
```

<!-- stack the UMAP overview on top of the per-cluster profiles -->
```{r function-autocor-haplotype-example, include=FALSE}
autocor_haplotype_example <- function(result) {
  overview <- autocor_umap_overview(result)
  profiles <- autocor_profile_composition(result)
  cowplot::plot_grid(overview$plot, profiles$plot, ncol = 1,
    rel_heights = c(overview$height, profiles$height))
}
```

<!-- check that the R packages needed for plotting are installed -->
```{r function-autocor-fiberseq-packages, include=FALSE}
autocor_fiberseq_packages <- function() {
  packages <- c("Matrix", "ggplot2", "ComplexHeatmap", "circlize", "igraph", "cowplot", "dplyr", "scales", "svglite", "png")
  missing <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing)) stop("Install these R packages before rendering: ", paste(missing, collapse = ", "))
  invisible(packages)
}
```

<!-- find local maxima of one ACF curve. Parameters: prominence >= 0.01, peaks at least 20 bp apart; spacing filter applied before prominence, as in scipy.signal.find_peaks -->
```{r function-acf-local-peaks, include=FALSE}
acf_local_peaks <- function(profile, prominence = 0.01, distance = 20L) {

  profile <- as.numeric(profile)
  n <- length(profile)
  if (n < 3L || any(!is.finite(profile))) return(integer())
  peaks <- integer()
  i <- 2L
  while (i < n) {
    if (profile[i] > profile[i - 1L]) {
      right <- i
      while (right < n && profile[right + 1L] == profile[i]) right <- right + 1L
      if (right < n && profile[right + 1L] < profile[i]) {
        peaks <- c(peaks, as.integer(floor((i + right) / 2)))
      }
      i <- right + 1L
    } else {
      i <- i + 1L
    }
  }
  if (!length(peaks)) return(integer())
  # As in scipy.signal.find_peaks, distance filtering precedes prominence.
  # Give the rightmost peak priority when heights tie.
  keep <- rep(TRUE, length(peaks))
  priority <- order(profile[peaks], peaks, decreasing = TRUE)
  for (j in priority) {
    if (keep[j]) {
      nearby <- which(abs(peaks - peaks[j]) < ceiling(distance))
      keep[setdiff(nearby, j)] <- FALSE
    }
  }
  peaks <- peaks[keep]
  prominence_values <- numeric(length(peaks))
  for (j in seq_along(peaks)) {
    peak <- peaks[j]
    height <- profile[peak]
    higher_left <- which(profile[seq_len(peak - 1L)] > height)
    left <- if (length(higher_left)) tail(higher_left, 1L) else 1L
    higher_right <- which(profile[seq.int(peak + 1L, n)] > height)
    right <- if (length(higher_right)) peak + higher_right[1L] else n
    prominence_values[j] <- height - max(
      min(profile[seq.int(left, peak)]), min(profile[seq.int(peak, right)])
    )
  }
  peaks[prominence_values >= prominence]
}
```

<!-- estimate the repeat length of one ACF curve: primary = highest positive peak, higher order = peaks near its multiples. Parameters: primary band 120–250 bp, multiples within ±20 bp, needs at least 252 lags -->
```{r function-repeat-candidates, include=FALSE}
repeat_candidates <- function(profile) {

  profile <- as.numeric(profile)
  missing <- list(primary = NA_real_, higher = integer())
  if (length(profile) <= 251L || any(!is.finite(profile))) return(missing)
  positions <- acf_local_peaks(profile, prominence = 0.01, distance = 20L)
  positions <- positions[profile[positions] > 0]
  lags <- positions - 1L
  band <- positions[lags >= 120L & lags <= 250L]
  if (!length(band)) return(missing)
  primary <- band[which.max(profile[band])] - 1L
  higher <- integer()
  last_multiple <- (length(profile) - 1L) %/% primary
  if (last_multiple >= 2L) {
    for (multiple in seq.int(2L, last_multiple)) {
      nearby <- positions[abs(lags - multiple * primary) <= 20L]
      if (length(nearby)) higher <- c(higher, nearby[which.max(profile[nearby])] - 1L)
    }
  }
  list(primary = primary, higher = higher)
}
```

<!-- for REF and ALT: median ACF across valid reads, its repeat candidates, and each read's primary peak; reads sorted by peak lag -->
```{r function-summarize-repeats, include=FALSE}
summarize_repeats <- function(profiles, records, region, allele_tables) {
  profiles <- as.matrix(profiles)
  audit <- allele_tables$allele_read_audit
  needed <- c("row_id", "allele_analysis_included", "acf_valid", "allele_group")
  if (is.null(audit) || !all(needed %in% names(audit))) {
    stop("The allele read audit is missing required columns.")
  }
  if (nrow(profiles) != nrow(records) || nrow(audit) != nrow(records) ||
      !identical(as.character(records$row_id), as.character(audit$row_id))) {
    stop("ACFs, records and allele audit must have matching rows in the same order.")
  }
  if (!ncol(profiles)) stop("The ACF matrix must contain at least lag 0.")
  included <- as.logical(audit$allele_analysis_included)
  valid <- as.logical(audit$acf_valid)
  if (anyNA(included) || anyNA(valid)) stop("Audit inclusion and ACF validity must be defined.")
  results <- vector("list", 2L)
  for (a in seq_along(results)) {
    allele <- c("REF", "ALT")[a]
    allele_mask <- !is.na(audit$allele_group) & audit$allele_group == allele
    selected <- included & allele_mask
    indices <- which(selected & valid)
    values <- profiles[indices, , drop = FALSE]
    if (length(indices) && any(!is.finite(values))) {
      stop("The audit marks a non-finite ACF profile as valid.")
    }
    median <- if (length(indices)) apply(values, 2L, stats::median) else rep(NA_real_, ncol(profiles))
    candidates <- repeat_candidates(median)
    per_fiber <- vapply(seq_len(nrow(values)), function(i) repeat_candidates(values[i, ])$primary, numeric(1))
    ordering <- order(per_fiber, seq_along(per_fiber), na.last = TRUE)
    results[[a]] <- list(
      region_id = as.character(region$region_id), allele = allele,
      base = as.character(if (allele == "REF") region$ref else region$alt),
      median = median, primary = candidates$primary, higher = candidates$higher,
      peak_lags = per_fiber, indices = indices[ordering], sorted_peaks = per_fiber[ordering],
      n_valid = length(indices), n_invalid = sum(selected & !valid),
      n_unique = sum(selected), n_duplicates = sum(!included & allele_mask)
    )
  }
  results
}
```

<!-- one row per allele: read counts, number and fraction of reads with a peak, primary and higher-order repeat lengths -->
```{r function-repeat-summary-table, include=FALSE}
repeat_summary_table <- function(rows) {
  do.call(rbind, lapply(rows, function(row) {
    complete <- length(row$median) >= 252L
    n_peak <- sum(is.finite(row$peak_lags))
    data.frame(
      region_id = row$region_id, allele = row$allele, base = row$base,
      unique_fibers = row$n_unique, valid_acf = row$n_valid,
      undefined_acf = row$n_invalid, duplicates_removed = row$n_duplicates,
      peak_band_complete = complete,
      fibers_with_peak = if (complete) n_peak else NA_integer_,
      fibers_without_peak = if (complete) row$n_valid - n_peak else NA_integer_,
      peak_fraction = if (complete && row$n_valid > 0L) n_peak / row$n_valid else NA_real_,
      median_curve_primary_bp = row$primary,
      observed_higher_order_bp = paste(row$higher, collapse = ", "),
      stringsAsFactors = FALSE
    )
  }))
}
```

<!-- caption text with valid, undefined-ACF and duplicate read counts per allele -->
```{r function-repeat-counts-caption, include=FALSE}
repeat_counts_caption <- function(rows) {
  paste(vapply(rows, function(row) {
    sprintf("%s: %s valid, %s undefined ACF, %s duplicates removed",
            row$allele, row$n_valid, row$n_invalid, row$n_duplicates)
  }, character(1)), collapse = "; ")
}
```

<!-- plot the median ACF of REF and ALT reads with the detected peaks. Parameters: 120–250 bp search band shaded; stars = primary peak, circles = higher-order peaks; lower/upper set the lag range (default all lags) -->
```{r function-plot-haplotype-median-acf, include=FALSE}
plot_haplotype_median_acf <- function(rows, region, lower = 0, upper = NULL) {
  max_lag <- max(vapply(rows, function(row) length(row$median), integer(1))) - 1L
  if (is.null(upper)) upper <- max_lag
  curves <- do.call(rbind, lapply(rows, function(row) {
    data.frame(allele = row$allele, lag = seq_along(row$median) - 1L, acf = row$median)
  }))
  # Keep tables intact when a lag range contains just one row or no peaks.
  curves <- subset(curves, lag >= lower & lag <= upper)
  primary <- do.call(rbind, lapply(rows, function(row) {
    if (!is.finite(row$primary)) return(NULL)
    data.frame(allele = row$allele, lag = row$primary, acf = row$median[row$primary + 1L])
  }))
  higher <- do.call(rbind, lapply(rows, function(row) {
    if (!length(row$higher)) return(NULL)
    data.frame(allele = row$allele, lag = row$higher, acf = row$median[row$higher + 1L])
  }))
  annotations <- do.call(rbind, lapply(rows, function(row) {
    message <- if (is.finite(row$primary)) {
      sprintf("n = %s; primary = %s bp", row$n_valid, row$primary)
    } else if (!row$n_valid) {
      "No valid ACF fibers"
    } else if (max_lag < 251L) {
      "Search band incomplete"
    } else {
      sprintf("n = %s; no qualifying primary peak", row$n_valid)
    }
    data.frame(allele = row$allele, label = message)
  }))
  allele_levels <- c("REF", "ALT")
  curves$allele <- factor(curves$allele, levels = allele_levels)
  annotations$allele <- factor(annotations$allele, levels = allele_levels)
  if (!is.null(primary)) primary$allele <- factor(primary$allele, levels = allele_levels)
  if (!is.null(higher)) higher$allele <- factor(higher$allele, levels = allele_levels)
  p <- ggplot2::ggplot(curves, ggplot2::aes(lag, acf, color = allele)) +
    ggplot2::annotate("rect", xmin = 120, xmax = 250, ymin = -Inf, ymax = Inf,
                      fill = "#61a788", alpha = 0.13) +
    ggplot2::geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
    ggplot2::geom_line(linewidth = 0.45, na.rm = TRUE) +
    ggplot2::geom_text(data = annotations, ggplot2::aes(x = Inf, y = Inf, label = label),
                       inherit.aes = FALSE, hjust = 1.03, vjust = 1.3, size = 3) +
    ggplot2::facet_grid(allele ~ ., drop = FALSE) +
    ggplot2::scale_color_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), drop = FALSE) +
    ggplot2::coord_cartesian(xlim = c(lower, max(lower + 1, upper))) +
    ggplot2::labs(title = paste0(region$region_id, ": phased-allele median ACF"),
      subtitle = paste0("REF ", region$ref, " / ALT ", region$alt),
      x = "Lag (bp)", y = "Median raw ACF",
      caption = paste("Primary search: 120–250 bp (shading); stars: primary peaks; open circles: observed peaks near multiples.",
        "Positive maxima: prominence ≥0.01, spacing ≥20 bp; multiple tolerance ±20 bp. No smoothing.",
        repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "none", plot.caption = ggplot2::element_text(hjust = 0))
  if (!is.null(primary)) {
    primary <- subset(primary, lag >= lower & lag <= upper)
    p <- p + ggplot2::geom_point(data = primary, shape = 8, color = "#B8860B", size = 3.5)
  }
  if (!is.null(higher)) {
    higher <- subset(higher, lag >= lower & lag <= upper)
    p <- p + ggplot2::geom_point(data = higher, shape = 1, size = 2)
  }
  p
}
```

<!-- heatmap of every valid read's ACF, ordered by allele and then primary peak lag; dots = primary peaks. Parameters: lag 0 omitted, color limits at the 99th percentile of |ACF|, upper = last lag shown (default all) -->
```{r function-plot-haplotype-fiber-acf-heatmap, include=FALSE}
plot_haplotype_fiber_acf_heatmap <- function(profiles, rows, region, upper = NULL) {
  profiles <- as.matrix(profiles)
  max_lag <- ncol(profiles) - 1L
  if (is.null(upper)) upper <- max_lag
  upper <- min(as.integer(upper), max_lag)
  ordered <- unlist(lapply(rows, function(row) row$indices), use.names = FALSE)
  peaks <- unlist(lapply(rows, function(row) row$sorted_peaks), use.names = FALSE)
  title <- paste0(region$region_id, ": individual-fiber ACF")
  if (!length(ordered) || upper < 1L) {
    return(ggplot2::ggplot() + ggplot2::annotate("text", x = 0, y = 0,
      label = if (!length(ordered)) "No valid ACF fibers" else "No nonzero ACF lags available") +
      ggplot2::labs(title = title, caption = repeat_counts_caption(rows)) + ggplot2::theme_void())
  }
  all_values <- profiles[ordered, -1L, drop = FALSE]
  limit <- max(as.numeric(stats::quantile(abs(all_values), 0.99, names = FALSE)), 0.01)
  values <- all_values[, seq_len(upper), drop = FALSE]
  tiles <- data.frame(lag = rep(seq_len(upper), times = length(ordered)),
                      fiber = rep(seq_along(ordered), each = upper), acf = as.vector(t(values)))
  dots <- data.frame(lag = peaks, fiber = seq_along(ordered))
  dots <- dots[is.finite(dots$lag) & dots$lag <= upper, , drop = FALSE]
  sizes <- vapply(rows, function(row) row$n_valid, numeric(1))
  ends <- cumsum(sizes)
  centers <- ends - (sizes - 1) / 2
  labels <- vapply(rows, function(row) sprintf("%s (n=%s)", row$allele, row$n_valid), character(1))
  p <- ggplot2::ggplot(tiles, ggplot2::aes(lag, fiber, fill = acf)) +
    ggplot2::geom_raster() +
    ggplot2::geom_point(data = dots, ggplot2::aes(lag, fiber), inherit.aes = FALSE, size = 0.35) +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
      midpoint = 0, limits = c(-limit, limit), oob = scales::squish, na.value = "#dddddd",
      name = "Raw ACF") +
    ggplot2::scale_y_reverse(breaks = centers[sizes > 0], labels = labels[sizes > 0], expand = c(0, 0)) +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::labs(title = title, subtitle = paste0("REF ", region$ref, " / ALT ", region$alt),
      x = "Lag (bp); lag 0 omitted", y = "Fibers: allele, then primary peak lag",
      caption = paste("All valid unique fibers, independent of cluster assignment. No-peak fibers sort last within each allele.",
        "Dots: primary peaks. Color limits clip the largest 1% of absolute ACF values for display.",
        repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw(base_size = 10) + ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0))
  boundaries <- ends[ends > 0 & ends < length(ordered)] + 0.5
  if (length(boundaries)) p <- p + ggplot2::geom_hline(yintercept = boundaries, linewidth = 0.4)
  p
}
```

<!-- boxplot of per-read repeat lengths by allele. Parameters: only reads with a peak in 120–250 bp; y axis 115–255 bp -->
```{r function-plot-haplotype-repeat-boxplot, include=FALSE}
plot_haplotype_repeat_boxplot <- function(rows, region) {
  values <- do.call(rbind, lapply(rows, function(row) {
    peaks <- row$peak_lags[is.finite(row$peak_lags)]
    data.frame(allele = rep(row$allele, length(peaks)), peak = peaks)
  }))
  values$allele <- factor(values$allele, levels = c("REF", "ALT"))
  labels <- setNames(vapply(rows, function(row) {
    sprintf("%s\n%s/%s with peak", row$allele, sum(is.finite(row$peak_lags)), row$n_valid)
  }, character(1)), vapply(rows, function(row) row$allele, character(1)))
  empty <- data.frame(allele = vapply(rows, function(row) row$allele, character(1)),
                      has_peak = vapply(rows, function(row) any(is.finite(row$peak_lags)), logical(1)))
  empty <- empty[!empty$has_peak, , drop = FALSE]
  ggplot2::ggplot(values, ggplot2::aes(allele, peak, fill = allele)) +
    ggplot2::geom_boxplot(width = 0.5, outlier.size = 1) +
    ggplot2::geom_text(data = empty, ggplot2::aes(x = allele, y = 185, label = "No estimates"),
                       inherit.aes = FALSE, size = 3) +
    ggplot2::scale_x_discrete(limits = c("REF", "ALT"), labels = labels, drop = FALSE) +
    ggplot2::scale_fill_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), drop = FALSE) +
    ggplot2::coord_cartesian(ylim = c(115, 255)) +
    ggplot2::labs(title = paste0(region$region_id, ": per-fiber repeat-length candidates"),
      x = NULL, y = "Per-fiber candidate repeat length (bp)",
      caption = paste("Boxes include only fibers with a qualifying positive local maximum in 120–250 bp; no smoothing.",
        "No qualifying peak means no estimate.", repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw(base_size = 10) + ggplot2::theme(legend.position = "none", plot.caption = ggplot2::element_text(hjust = 0))
}
```

<!-- histogram of per-read repeat lengths by allele. Parameters: 5-bp bins over 120–250 bp; fractions are out of all valid reads -->
```{r function-plot-haplotype-repeat-histogram, include=FALSE}
plot_haplotype_repeat_histogram <- function(rows, region) {
  # Use all valid fibers as the denominator, including those without a peak.
  breaks <- seq(119.5, 254.5, by = 5)
  bins <- do.call(rbind, lapply(rows, function(row) {
    peaks <- row$peak_lags[is.finite(row$peak_lags)]
    counts <- graphics::hist(peaks, breaks = breaks, plot = FALSE, right = TRUE)$counts
    data.frame(allele = row$allele, xmin = head(breaks, -1L), xmax = tail(breaks, -1L),
      fraction = if (row$n_valid) counts / row$n_valid else rep(0, length(counts)))
  }))
  bins$allele <- factor(bins$allele, levels = c("REF", "ALT"))
  ggplot2::ggplot(bins, ggplot2::aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = fraction, color = allele)) +
    ggplot2::geom_rect(fill = NA, linewidth = 0.6) +
    ggplot2::scale_color_manual(values = c(REF = "#2878B5", ALT = "#D55E00"), drop = FALSE) +
    ggplot2::coord_cartesian(xlim = c(119, 251)) +
    ggplot2::labs(title = paste0(region$region_id, ": repeat-length candidate frequencies"),
      x = "Per-fiber primary peak lag (bp)", y = "Fraction of all valid fibers per 5-bp bin", color = "Allele",
      caption = paste("Strongest qualifying positive local maximum in 120–250 bp; no smoothing.",
        "Histogram denominators include all valid fibers, including those without an estimate.",
        repeat_counts_caption(rows), sep = "\n")) +
    ggplot2::theme_bw(base_size = 10) + ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0))
}
```

<!-- heatmap of the REF and ALT median ACF; stars = primary peaks. Parameters: lags 120–250 bp -->
```{r function-plot-locus-heatmap, include=FALSE}
plot_locus_heatmap <- function(rows, region = NULL) {
  region_ids <- unique(vapply(rows, function(row) as.character(row$region_id), character(1)))
  if (length(region_ids) != 1L) stop("Plot one region at a time.")
  lags <- 120:250
  tiles <- do.call(rbind, lapply(seq_along(rows), function(i) {
    row <- rows[[i]]
    values <- rep(NA_real_, length(lags))
    available <- lags < length(row$median)
    values[available] <- row$median[lags[available] + 1L]
    data.frame(lag = lags, allele_row = i, acf = values)
  }))
  limit <- max(c(abs(tiles$acf[is.finite(tiles$acf)]), 0.01))
  stars <- data.frame(lag = vapply(rows, function(row) row$primary, numeric(1)), allele_row = seq_along(rows))
  stars <- stars[is.finite(stars$lag), , drop = FALSE]
  labels <- vapply(rows, function(row) sprintf("%s (%s), n=%s", row$allele, row$base, row$n_valid), character(1))
  ggplot2::ggplot(tiles, ggplot2::aes(lag, allele_row, fill = acf)) +
    ggplot2::geom_tile(width = 1, height = 1) +
    ggplot2::geom_point(data = stars, ggplot2::aes(lag, allele_row), inherit.aes = FALSE,
                       shape = 8, color = "#B8860B", size = 4) +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0,
      limits = c(-limit, limit), na.value = "#dddddd", name = "Median raw ACF") +
    ggplot2::scale_y_reverse(breaks = seq_along(rows), labels = labels, expand = c(0, 0)) +
    ggplot2::scale_x_continuous(expand = c(0, 0)) +
    ggplot2::labs(title = paste0(region_ids, ": phased-allele median autocorrelation"),
      x = "Candidate repeat length / ACF lag (bp)", y = "Phased allele",
      caption = paste("120–250 bp; common color scale within this region. Stars: qualifying primary peaks. Gray: unavailable.",
        "Every unique retained fiber with a valid ACF contributes, including unclustered fibers.", sep = "\n")) +
    ggplot2::theme_bw(base_size = 10) + ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0))
}
```

<!-- make all seven allele repeat figures for one region. Parameters: zoomed median ACF 100–500 bp, zoomed read heatmap 0–500 bp -->
```{r function-plot-haplotype-repeats, include=FALSE}
plot_haplotype_repeats <- function(region, profiles, rows) {
  # Return figures for the notebook; rendering and filenames belong to knitr.
  list(
    median_acf = plot_haplotype_median_acf(rows, region),
    median_acf_zoom = plot_haplotype_median_acf(rows, region, lower = 100, upper = 500),
    fiber_heatmap = plot_haplotype_fiber_acf_heatmap(profiles, rows, region),
    fiber_heatmap_zoom = plot_haplotype_fiber_acf_heatmap(profiles, rows, region, upper = 500),
    repeat_boxplot = plot_haplotype_repeat_boxplot(rows, region),
    repeat_histogram = plot_haplotype_repeat_histogram(rows, region),
    allele_median_heatmap = plot_locus_heatmap(rows, region)
  )
}
```

<!-- keep only the fields needed for the locus summary (region, allele, base, median ACF, primary peak, n valid) -->
```{r function-compact-locus-rows, include=FALSE}
compact_locus_rows <- function(rows) {
  lapply(rows, function(row) row[c("region_id", "allele", "base", "median", "primary", "n_valid")])
}
```

<!-- check that every region has report_data.rds, check the plotting packages, and load all regions -->
```{r load-region-results, include=FALSE, purl=TRUE}
missing_data <- region_dirs[!file.exists(file.path(region_dirs, "report_data.rds"))]
if (length(missing_data)) stop(
  "Direct plotting requires report_data.rds for each region. Missing: ",
  paste(basename(missing_data), collapse = ", "),
  ". Run 01_run_autocorrelation.py to save numerical inputs, or set output_dir to a prepared run.")
autocor_fiberseq_packages()
region_results <- lapply(region_dirs, read_report_data)
```

<style>
.main-container { max-width: 1700px; padding: 0 24px; }
table { display: block; overflow-x: auto; }
</style>

<!-- draw every figure the way leiden_LCL.Rmd does: save as PDF at a fixed size, convert to PNG with Liberation Sans fonts, embed in the report. The PNG is temporary and only embedded in the HTML; figure files are written as PDF only (report_pdf_figure with file = path keeps the PDF), into the region's autocorrelation/ subfolder for the ACF clusters and leiden/ for the m6A Leiden clusters. Parameters: 120 dpi; fonts and converter shared with leiden_LCL.Rmd via report_helpers.r -->
```{r gallery-helper, include=FALSE}
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/report_helpers.r", local = TRUE)

report_pdf_figure <- function(figure, width, height, label, file = NULL) {
  # file = NULL: temporary PDF for the embedded preview only; otherwise the PDF is kept at that path.
  pdf <- if (is.null(file)) tempfile("figure-", fileext = ".pdf") else file
  if (is.null(file)) on.exit(unlink(pdf)) else dir.create(dirname(file), recursive = TRUE, showWarnings = FALSE)
  if (inherits(figure, c("Heatmap", "HeatmapList"))) {
    grDevices::pdf(pdf, width = width, height = height)
    tryCatch(ComplexHeatmap::draw(figure, newpage = FALSE), finally = grDevices::dev.off())
  } else {
    ggplot2::ggsave(pdf, figure, width = width, height = height, limitsize = FALSE, bg = "white")
  }
  inline_png(pdf, label)
}

# Figure sizes used by leiden_LCL.Rmd; n = clustered reads, k = clusters.
fiberseq_figure_size <- function(entry, kind) {
  n <- if (is.null(entry$fiberseq)) 0L else nrow(entry$fiberseq$assignments)
  k <- if (is.null(entry$fiberseq)) 1L else entry$fiberseq$n_clusters
  switch(kind,
    reads = c(12, max(7, 0.03 * n + 3.5 + 0.3 * k)),
    heatmap = c(14, 11 + 0.55 * k),
    acf_heatmap = c(17, 11 + 0.55 * k),
    occupancy = c(10, 1.4 * k + 2.5),
    knn = c(12, 9),
    proportions = c(8, 5),
    stop("Unknown figure size: ", kind))
}
```

# Autocorrelation figures

## Allele-average ACF

```{r figure-allele-acf, echo=FALSE, results='asis', fig.height=10, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_acf")
  report_pdf_figure(figure, 12, 10, paste(entry$region_id, "Allele-average ACF", sep = " — "))
}
```

## Median ACF by focal allele

```{r figure-haplotype-median, echo=FALSE, results='asis', fig.height=8, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_median_acf(entry$repeats, entry$data$region)
  report_pdf_figure(figure, 12, 8, paste(entry$region_id, "Median ACF by focal allele", sep = " — "))
}
```

## Median ACF: nucleosome-scale view

```{r figure-haplotype-median-zoom, echo=FALSE, results='asis', fig.height=8, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_median_acf(entry$repeats, entry$data$region, lower = 100, upper = 500)
  report_pdf_figure(figure, 12, 8, paste(entry$region_id, "Median ACF: nucleosome-scale view", sep = " — "))
}
```

## Allele median ACF: 120–250 bp

```{r figure-region-repeat-heatmap, echo=FALSE, results='asis', fig.height=5, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_locus_heatmap(entry$repeats, entry$data$region)
  report_pdf_figure(figure, 12, 5, paste(entry$region_id, "Allele median ACF: 120–250 bp", sep = " — "))
}
```

## Individual-fiber ACF by allele

```{r figure-fiber-acf, echo=FALSE, results='asis', fig.height=9, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_fiber_acf_heatmap(entry$data$acf, entry$repeats, entry$data$region)
  report_pdf_figure(figure, 12, 9, paste(entry$region_id, "Individual-fiber ACF by allele", sep = " — "))
}
```

## Individual-fiber ACF: first 500 bp

```{r figure-fiber-acf-zoom, echo=FALSE, results='asis', fig.height=9, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_fiber_acf_heatmap(entry$data$acf, entry$repeats, entry$data$region, upper = 500)
  report_pdf_figure(figure, 12, 9, paste(entry$region_id, "Individual-fiber ACF: first 500 bp", sep = " — "))
}
```

## Per-fiber repeat-length candidates

```{r figure-repeat-boxplot, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_repeat_boxplot(entry$repeats, entry$data$region)
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "Per-fiber repeat-length candidates", sep = " — "))
}
```

## Per-fiber repeat-length frequencies

```{r figure-repeat-histogram, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_repeat_histogram(entry$repeats, entry$data$region)
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "Per-fiber repeat-length frequencies", sep = " — "))
}
```

## Allele positive-peak fractions

```{r figure-allele-peak-features, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_peaks")
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "Allele positive-peak fractions", sep = " — "))
}
```

## Cluster-average ACF

```{r figure-cluster-acf, echo=FALSE, results='asis', fig.height=7, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "cluster_acf")
  report_pdf_figure(figure, 12, 7, paste(entry$region_id, "Cluster-average ACF", sep = " — "), file.path(entry$folder, "autocorrelation", "cluster_average_acf.pdf"))
}
```

## ACF and genomic m6A heatmaps

```{r figure-acf-m6a-heatmap, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) autocor_acf_heatmap(result) + plot_genomic_cluster_heatmap(result, result$region, sample_palette(result$assignments$sample_name), include_haplotype = TRUE, variants = result$variants, show_cluster_profiles = TRUE, cluster_label = "Autocorrelation-defined Leiden clusters"))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "acf_heatmap")[1], fiberseq_figure_size(entry, "acf_heatmap")[2], paste(entry$region_id, "ACF and genomic m6A heatmaps", sep = " — "), file.path(entry$folder, "autocorrelation", "acf_and_m6a_heatmaps.pdf"))
}
```

## ACF UMAP and m6A call fraction

```{r figure-acf-umap, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_acf_umap(entry$data)
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "ACF UMAP and m6A call fraction", sep = " — "), file.path(entry$folder, "autocorrelation", "umap.pdf"))
}
```

## Saved ACF neighbor graph

```{r figure-knn-graph, echo=FALSE, results='asis', fig.height=9, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_knn_graph(result, main = paste(result$region$annotation, "ACF correlation-neighbor graph colored by Leiden cluster", sep = "\n")))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "knn")[1], fiberseq_figure_size(entry, "knn")[2], paste(entry$region_id, "ACF neighbor graph colored by Leiden cluster", sep = " — "), file.path(entry$folder, "autocorrelation", "knn_graph.pdf"))
}
```

# Fiber-seq figures

## Read footprints

```{r figure-read-footprints, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_smf_reads(result, signal_profiles(result, result$footprints)$records, sample_palette(result$assignments$sample_name), include_haplotype = TRUE) + ggplot2::labs(subtitle = "One row per retained molecule; saved autocorrelation-defined Leiden clusters"))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "reads")[1], fiberseq_figure_size(entry, "reads")[2], paste(entry$region_id, "Read-level footprints", sep = " — "), file.path(entry$folder, "autocorrelation", "fig1_read_footprints.pdf"))
}
```

## Genomic m6A heatmap

```{r figure-m6a-heatmap, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_genomic_cluster_heatmap(result, result$region, sample_palette(result$assignments$sample_name), include_haplotype = TRUE, variants = result$variants, show_cluster_profiles = TRUE, cluster_label = "Autocorrelation-defined Leiden clusters"))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "heatmap")[1], fiberseq_figure_size(entry, "heatmap")[2], paste(entry$region_id, "m6A and footprint heatmaps", sep = " — "), file.path(entry$folder, "autocorrelation", "heatmap_m6a_footprints.pdf"))
}
```

## Footprint occupancy by cluster

```{r figure-occupancy, echo=FALSE, results='asis', fig.height=10, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_signal_profile(signal_profiles(result, result$footprints)$profiles, result$region, levels(result$assignments$cluster)))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "occupancy")[1], fiberseq_figure_size(entry, "occupancy")[2], paste(entry$region_id, "Occupancy profiles by cluster", sep = " — "), file.path(entry$folder, "autocorrelation", "fig2_occupancy_by_cluster.pdf"))
}
```

## Signal profiles and composition

```{r figure-fiberseq-composition, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  panel <- if (is.null(entry$fiberseq)) NULL else autocor_profile_composition(entry$fiberseq)
  figure <- if (is.null(panel)) report_fiberseq_figure(entry, identity) else panel$plot
  report_pdf_figure(figure, 16, if (is.null(panel)) 7 else panel$height, paste(entry$region_id, "Signal profiles and composition", sep = " — "), file.path(entry$folder, "autocorrelation", "fiberseq_profiles_composition.pdf"))
}
```

## Cluster proportions

```{r figure-cluster-proportions, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, plot_cluster_pie)
  report_pdf_figure(figure, fiberseq_figure_size(entry, "proportions")[1], fiberseq_figure_size(entry, "proportions")[2], paste(entry$region_id, "Cluster proportions", sep = " — "), file.path(entry$folder, "autocorrelation", "fiberseq_cluster_proportions.pdf"))
}
```

## UMAP and allele overview

```{r figure-umap-overview, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  panel <- if (is.null(entry$fiberseq)) NULL else autocor_umap_overview(entry$fiberseq)
  figure <- if (is.null(panel)) report_fiberseq_figure(entry, identity) else panel$plot
  report_pdf_figure(figure, 16, if (is.null(panel)) 7 else panel$height, paste(entry$region_id, "UMAP and allele overview", sep = " — "), file.path(entry$folder, "autocorrelation", "fiberseq_umap_overview.pdf"))
}
```

## Combined haplotype overview

```{r figure-haplotype-overview, echo=FALSE, results='asis', fig.height=18, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  overview <- if (is.null(entry$fiberseq)) NULL else autocor_umap_overview(entry$fiberseq)
  profiles <- if (is.null(entry$fiberseq)) NULL else autocor_profile_composition(entry$fiberseq)
  figure <- report_fiberseq_figure(entry, autocor_haplotype_example)
  report_pdf_figure(figure, 16, if (is.null(overview)) 7 else overview$height + profiles$height, paste(entry$region_id, "Haplotype overview", sep = " — "), file.path(entry$folder, "autocorrelation", "fiberseq_haplotype_example.pdf"))
}
```

## Sample composition

```{r figure-sample-composition, echo=FALSE, results='asis', fig.height=7, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "sample_composition")
  report_pdf_figure(figure, 12, 7, paste(entry$region_id, "Sample composition", sep = " — "), file.path(entry$folder, "autocorrelation", "sample_composition.pdf"))
}
```

## Allele coverage

```{r figure-allele-coverage, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_coverage")
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "Allele coverage", sep = " — "))
}
```

## Allele cluster proportions

```{r figure-allele-proportions, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_table(entry$data, "allele_proportions")
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "Allele cluster proportions", sep = " — "), file.path(entry$folder, "autocorrelation", "allele_cluster_proportions.pdf"))
}
```

# Result tables

```{r table-parameters, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(as.data.frame(entry$data$params), "Saved analysis parameters")
}
```

## Genomic region

```{r table-regions, echo=FALSE, results='asis', purl=FALSE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$region, "Genomic region")
}
```

## Allele coverage

```{r table-coverage, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$allele_coverage, "Allele coverage")
}
```

## Cluster summaries

```{r table-cluster-summary, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$cluster_summary, "Cluster summaries")
}
```

## Allele cluster proportions

```{r table-allele-clusters, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$allele_cluster_proportions, "Allele cluster proportions")
}
```

## Positive local peaks: 140–250 bp

```{r table-allele-peaks, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$allele_peak_features, "Positive local peaks: 140–250 bp")
}
```

## Repeat-length candidates: 120–250 bp

```{r table-repeat-summary, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(repeat_summary_table(entry$repeats), "Repeat-length candidates: 120–250 bp")
}
```

## Sample composition

```{r table-sample-composition, echo=FALSE, results='asis', purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  print_report_table(entry$data$tables$sample_composition, "Sample composition")
}
```

# Leiden clusters directly on binary m6A

The same reads are clustered directly on the binary m6A matrix with the Leiden_Manhattan code, the autocorrelation is not used to form these clusters

  - `leiden_manhattan_cluster()` with the settings of `leiden_LCL.Rmd`: every m6A site is one feature (no windowing), Manhattan distance, k = 10, Leiden resolution 1, seed 1
  - cluster1 is the largest cluster
  - the ACF of each read is then averaged within each m6A cluster

<!-- load the Leiden_Manhattan clustering functions (Manhattan kNN graph, exponential kernel, Leiden) so the m6A clusters come from the same code as the Leiden_Manhattan report -->
```{r source-leiden-functions, include=FALSE}
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_functions.r", local = TRUE)
```

<!-- cluster one region's reads on the binary m6A matrix and swap the m6A clusters, Manhattan kNN graph, Manhattan UMAP and per-cluster mean ACF into a copy of the region's plotting object, so the existing figure functions draw them. Parameters: as in leiden_LCL.Rmd — window_size 0 (each m6A site is a feature), k = 10, resolution 1, seed 1 -->
```{r function-autocor-m6a-leiden, include=FALSE}
autocor_m6a_leiden <- function(entry) {
  result <- entry$fiberseq
  if (is.null(result)) return(entry)
  clustering <- leiden_manhattan_cluster(result$site_met_mat, result$assignments,
    result$region$analysis_start, result$region$analysis_end, window_size = 0L,
    k_neighbors = 10L, resolution = 1, seed = 1L, verbose = FALSE)
  stopifnot(identical(clustering$assignments$RID, result$assignments$RID))
  result$assignments$cluster <- clustering$assignments$cluster
  result[c("feat_mat", "graph", "n_clusters", "params")] <- clustering[c("feat_mat", "graph", "n_clusters", "params")]
  embedding <- fiberseq_umap(result, seed = 1L)
  result$assignments$umap1 <- embedding$UMAP1
  result$assignments$umap2 <- embedding$UMAP2
  entry$fiberseq <- result
  # Same layout as the saved cluster_acf_summary table, for plot_report_cluster_acf().
  entry$data$tables$cluster_acf_summary <- do.call(rbind, lapply(levels(result$assignments$cluster), function(cluster) {
    selected <- result$assignments$cluster == cluster
    data.frame(cluster = sub("^cluster", "", cluster), lag_bp = seq_len(ncol(result$acf)) - 1L,
      mean_acf = colMeans(result$acf[selected, , drop = FALSE]), n_reads = sum(selected))
  }))
  entry
}
```

<!-- cluster every region on binary m6A -->
```{r m6a-leiden-results, include=FALSE, purl=TRUE}
m6a_results <- lapply(region_results, autocor_m6a_leiden)
```

## Cluster-average ACF

```{r figure-m6a-leiden-cluster-acf, echo=FALSE, results='asis', fig.height=7, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_cluster_acf(entry$data, title = "m6A-defined Leiden cluster averages; no smoothing")
  report_pdf_figure(figure, 12, 7, paste(entry$region_id, "m6A Leiden clusters: cluster-average ACF", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_cluster_average_acf.pdf"))
}
```

## Cluster-average ACF: 25–750 bp

```{r figure-m6a-leiden-cluster-acf-range, echo=FALSE, results='asis', fig.height=7, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_cluster_acf(entry$data, lower = 25, upper = 750, title = "m6A-defined Leiden cluster averages; lags 25–750 bp; no smoothing")
  report_pdf_figure(figure, 12, 7, paste(entry$region_id, "m6A Leiden clusters: cluster-average ACF, 25–750 bp", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_cluster_average_acf_25_750bp.pdf"))
}
```

## ACF and genomic m6A heatmaps

```{r figure-m6a-leiden-acf-m6a-heatmap, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) autocor_acf_heatmap(result) + plot_genomic_cluster_heatmap(result, result$region, sample_palette(result$assignments$sample_name), include_haplotype = TRUE, variants = result$variants, show_cluster_profiles = TRUE, cluster_label = "m6A-defined Leiden clusters"))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "acf_heatmap")[1], fiberseq_figure_size(entry, "acf_heatmap")[2], paste(entry$region_id, "m6A Leiden clusters: ACF and genomic m6A heatmaps", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_acf_and_m6a_heatmaps.pdf"))
}
```

## Read footprints

```{r figure-m6a-leiden-read-footprints, echo=FALSE, results='asis', fig.height=12, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_smf_reads(result, signal_profiles(result, result$footprints)$records, sample_palette(result$assignments$sample_name), include_haplotype = TRUE) + ggplot2::labs(subtitle = "One row per retained molecule; m6A-defined Leiden clusters"))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "reads")[1], fiberseq_figure_size(entry, "reads")[2], paste(entry$region_id, "m6A Leiden clusters: read-level footprints", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_read_footprints.pdf"))
}
```

## Footprint occupancy by cluster

```{r figure-m6a-leiden-occupancy, echo=FALSE, results='asis', fig.height=10, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, function(result) plot_signal_profile(signal_profiles(result, result$footprints)$profiles, result$region, levels(result$assignments$cluster)))
  report_pdf_figure(figure, fiberseq_figure_size(entry, "occupancy")[1], fiberseq_figure_size(entry, "occupancy")[2], paste(entry$region_id, "m6A Leiden clusters: occupancy profiles by cluster", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_occupancy_by_cluster.pdf"))
}
```

## Combined haplotype overview

```{r figure-m6a-leiden-haplotype-overview, echo=FALSE, results='asis', fig.height=18, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  overview <- if (is.null(entry$fiberseq)) NULL else autocor_umap_overview(entry$fiberseq)
  profiles <- if (is.null(entry$fiberseq)) NULL else autocor_profile_composition(entry$fiberseq)
  figure <- report_fiberseq_figure(entry, autocor_haplotype_example)
  report_pdf_figure(figure, 16, if (is.null(overview)) 7 else overview$height + profiles$height, paste(entry$region_id, "m6A Leiden clusters: haplotype overview (Manhattan UMAP)", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_haplotype_example.pdf"))
}
```

## Manhattan neighbor graph

```{r figure-m6a-leiden-knn-graph, echo=FALSE, results='asis', fig.height=9, purl=TRUE}
for (entry in m6a_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- report_fiberseq_figure(entry, plot_knn_graph)
  report_pdf_figure(figure, fiberseq_figure_size(entry, "knn")[1], fiberseq_figure_size(entry, "knn")[2], paste(entry$region_id, "m6A Leiden clusters: Manhattan neighbor graph colored by Leiden cluster", sep = " — "), file.path(entry$folder, "leiden", "m6a_leiden_knn_graph.pdf"))
}
```

# Autocorrelation curves: 25–750 bp

The existing ACF curve figures redrawn for lags 25–750 bp. The vertical scale is fit to these lags only, so the short-lag ACF near lag 0 no longer sets the scale

## Allele-average ACF: 25–750 bp

```{r figure-allele-acf-range, echo=FALSE, results='asis', fig.height=6, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_allele_acf(entry$data, lower = 25, upper = 750)
  report_pdf_figure(figure, 12, 6, paste(entry$region_id, "Allele-average ACF: 25–750 bp", sep = " — "), file.path(entry$folder, "autocorrelation", "allele_average_acf_25_750bp.pdf"))
}
```

## Median ACF by focal allele: 25–750 bp

```{r figure-haplotype-median-range, echo=FALSE, results='asis', fig.height=8, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_haplotype_median_acf(entry$repeats, entry$data$region, lower = 25, upper = 750)
  report_pdf_figure(figure, 12, 8, paste(entry$region_id, "Median ACF by focal allele: 25–750 bp", sep = " — "), file.path(entry$folder, "autocorrelation", "haplotype_median_acf_25_750bp.pdf"))
}
```

## Cluster-average ACF: 25–750 bp

```{r figure-cluster-acf-range, echo=FALSE, results='asis', fig.height=7, purl=TRUE}
for (entry in region_results) {
  cat("\n\n### ", entry$region_id, "\n\n", sep = "")
  figure <- plot_report_cluster_acf(entry$data, lower = 25, upper = 750, title = "Cluster averages; lags 25–750 bp; no smoothing")
  report_pdf_figure(figure, 12, 7, paste(entry$region_id, "Cluster-average ACF: 25–750 bp", sep = " — "), file.path(entry$folder, "autocorrelation", "cluster_average_acf_25_750bp.pdf"))
}
```
