---
title: "LCL Fiber-seq autocorrelation"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = "autocorrelation.html", output_dir = "/project/spott/cshan/fiber-seq/LCL_project/auto_correlation/resolution04/outputs") })
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    self_contained: true
params:
  run_analysis: false
  window_size: 2000
  n_features: null
  n_pcs: 50
  n_neighbors: 10
  resolution: 0.4
  seed: 0
  regions: null
  output_dir: "/project/spott/cshan/fiber-seq/LCL_project/auto_correlation/resolution04"
  python: "/project/spott/cshan/envs/Jupyter-notebook/bin/python"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, out.width = "100%")
project <- "/project/spott/cshan/fiber-seq"
scripts <- file.path(project, "code/clustering_methods/auto_correlation")
output <- normalizePath(params$output_dir, mustWork = FALSE)
if (isTRUE(params$run_analysis)) {
  args <- c(shQuote(file.path(scripts, "01_run_autocorrelation.py")),
    "--window-size", params$window_size, "--n-pcs", params$n_pcs,
    "--n-neighbors", params$n_neighbors, "--resolution", params$resolution,
    "--seed", params$seed,
    "--output-dir", shQuote(output))
  if (!is.null(params$n_features)) args <- c(args, "--n-features", params$n_features)
  for (rid in params$regions) args <- c(args, "--region", shQuote(rid))
  status <- system2(params$python, args)
  if (status != 0L) stop("Autocorrelation pipeline failed; inspect its diagnostic logs")
}
region_dirs <- list.dirs(file.path(output, "outputs"), recursive = FALSE, full.names = TRUE)
region_dirs <- sort(region_dirs[grepl("^rank[0-9]+_", basename(region_dirs))])
if (!is.null(params$regions)) region_dirs <- region_dirs[basename(region_dirs) %in% params$regions]
if (!length(region_dirs)) stop("No region figures found; run 01_run_autocorrelation.py first")
```

# Analysis and parameters

### Input

**Real example setup**

These are actual records from sample `AL16_bc2114_18508`, read from:

`/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL16_bc2114_18508/extracted_results/m6a_by_chr/AL16_bc2114_18508.ft_extracted_m6a.chr2.bed.gz`

The selected region is `rank01_rs12470189_chr2_119290945_119291178`.
Its analysis window is **chr2:119290056–119292055** (2,000 bases, 1-based,
inclusive), centered on **rs12470189 at chr2:119291056, REF G / ALT A**.
This historical example uses the region definitions in the selection RDS.

Four real reads are labeled A–D below. A–C are retained reads representing
both focal alleles; D illustrates a read that does not span the full window.
They are an explanatory subset of the regional cohort. Matrix tables display
only **chr2:119291048–119291059**, 12 consecutive columns from the full
2,000-column matrix. Coverage filtering and call offsets always use the full
analysis window

**Step 1: Original fibertools m6A BED12 records**

The IDs, coordinates, block counts, and block values below come from the
original BED file. All four records have chromosome `chr2` and RGB `128,0,128`.
BED starts are 0-based; BED ends are exclusive. `read_start` and `read_end`
are BED12 columns 7 and 8. Block lists are abbreviated with `…` for display;
`blockCount` gives their full length. Complete, unmodified BED12 lines are
included immediately below the table.

| Read | RID | BED start | BED end | read_start | read_end | Score | Strand | blockCount | blockSizes  | blockStarts  |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | m84241_250908_180102_s3/160175746/ccs | 119282125 | 119298507 | 119282125 | 119298507 | 8 | - | 144 | 0,1,1,1,1, …, 1,1,1 | 0,37,452,618,1019, …, 15880,16060,16381 |
| B | m84241_250515_002859_s3/157422976/ccs | 119280507 | 119293438 | 119280507 | 119293438 | 7 | - | 248 | 0,1,1,1,1, …, 1,1,1 | 0,309,310,312,462, …, 12680,12873,12930 |
| C | m84241_250908_180102_s3/258540726/ccs | 119289060 | 119298334 | 119289060 | 119298334 | 18 | - | 266 | 0,1,1,1,1, …, 1,1,1 | 0,293,299,301,309, …, 9078,9080,9273 |
| D | m84241_250515_002859_s3/165545798/ccs | 119291372 | 119296368 | 119291372 | 119296368 | 32 | + | 120 | 0,1,1,1,1, …, 1,1,1 | 0,117,122,184,277, …, 4815,4821,4995 |


The first and last blocks are fibertools sentinel blocks and are removed.
In these real records, the first block has size **0** and the last has size
**1**. Internal m6A-call blocks have size **1**.

**Step 2: Expand blocks and keep calls inside the analysis window**

`extract_ft_region_reads()` retains the longest alignment per RID, expands
BED12 blocks, removes the sentinel blocks, and keeps calls within
119290056–119292055. A call's 1-based coordinate is its block end:

`call position = BED start + blockStarts entry + blockSizes entry`

For example, read A's call at 119291048 is `119282125 + 8922 + 1`.
Strand does not reverse the genomic column order.

The table includes each read's first in-window call and every call in the
12-base display excerpt. Other calls remain in the full matrix.

| Read | Block number (1-based) | Block offset | Block size | Call position |
| --- | --- | --- | --- | --- |
| A | 82 | 7937 | 1 | 119290063 |
| A | 93 | 8922 | 1 | 119291048 |
| A | 94 | 8924 | 1 | 119291050 |
| A | 95 | 8926 | 1 | 119291052 |
| A | 96 | 8929 | 1 | 119291055 |
| B | 182 | 9563 | 1 | 119290071 |
| C | 43 | 1086 | 1 | 119290147 |
| C | 73 | 1987 | 1 | 119291048 |
| C | 74 | 1989 | 1 | 119291050 |
| C | 75 | 1991 | 1 | 119291052 |
| C | 76 | 1994 | 1 | 119291055 |
| C | 77 | 1998 | 1 | 119291059 |
| D | 2 | 117 | 1 | 119291490 |


**Step 3: Extract one metadata row per read**

`extract_ft_read_info(reads)` uses `BED12 column 7 + 1` as the inclusive
read start and column 8 as the inclusive read end. Read labels refer to the
exact RIDs in Step 1.

| Read | Chromosome | Start (1-based) | End (inclusive) | Strand | In-window m6A calls |
| --- | --- | --- | --- | --- | --- |
| A | chr2 | 119282126 | 119298507 | - | 26 |
| B | chr2 | 119280508 | 119293438 | - | 50 |
| C | chr2 | 119289061 | 119298334 | - | 81 |
| D | chr2 | 119291373 | 119296368 | + | 15 |

**Step 4: Construct the read-by-position matrix**

The actual call is
`get_sparse_met_mat(..., window_start = 119290056, window_end = 119292055, all_met_pos = 119290056:119292055, base = "A")`.

1. Initialize entries to `NA`.
2. Set positions covered by a read to `0`.
3. Set observed m6A-call positions to `1`.

`0` means covered without a reported m6A call; `NA` means uncovered.
Every genomic base is a column, including bases with no calls in any read.
The following values were checked with the current R extraction and matrix
functions against the original indexed BED file.

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |
| D | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA |

D starts at 119291373, so all 12 displayed bases are uncovered (`NA`).
It covers only part of the full analysis window.

**Step 5: Filter coverage and phasing**

`rowSums(is.na(mat)) == 0` removes D before phasing. The sample prefix is
then added to each retained RID as `AL16_bc2114_18508__<RID>`.

The phasing values below are historical results matched by sample and original
RID. Current runs hold the extraction audit and retained assignments in memory.
No haplotype or genotype is inferred from the m6A calls. D is absent from
that audit because the audit begins after full-span filtering.

| Read | Spans full 2,000 bp? | Haplotype | Phase set | Focal genotype | Resolved allele | Outcome |
| --- | --- | --- | --- | --- | --- | --- |
| A | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| B | Yes | HP2 | 119099734 | 0\|1 | rs12470189: A | Keep |
| C | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| D | No | Not evaluated | — | — | — | Remove before phasing |

A–C have `allele_status = phased_focal_genotype` and pass the phased
heterozygote filter. The current filter requires at least three retained
reads, both focal alleles, and at least two columns with observed m6A calls.
The regional cohort, rather than just the displayed excerpt, is filtered.

**Step 6: Final m6A matrix**

The shared phasing filter temporarily removes all-zero columns.
`lcl_build_phased_m6a_input()` restores all 2,000 genomic columns for the
retained reads. The same 12-column excerpt after filtering is:

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |

Read labels A–C stand for the sample-prefixed RIDs described above.
Positions 119291049, 119291051, 119291053, 119291054, and 119291056–119291058
are all zero in these three rows and remain in the matrix. These zeros do
not imply that the positions are zero across every other regional read.
B's displayed values are all zero, but B has 50 calls elsewhere in the full




The notebook parameters configure a new run only. Saved PDFs and diagnostic logs
show the actual results; no saved manifest or table is required to view this gallery.

The defaults are a **2,000-bp window and all 2,000 nonnegative ACF lags**.
Set `window_size` to change the window, centered at the same focal SNP;
set `n_features` to retain the first N lags, including lag 0. `null` retains
all lags. For an even window, the inclusive start is SNP position minus half
the width; the end is start plus width minus one. The worked tables above are historical examples, not summaries of the current run.

Set `run_analysis: true` to launch the pipeline with the notebook parameters;
choose a new, empty `output_dir` inside `LCL_project/auto_correlation`.
Leave it false to render existing results.

# Method

Following [SAMOSA-ChAAT](https://github.com/RamaniLab/SAMOSA-ChAAT), each
nonconstant molecule is mean-centered and its linear autocorrelation is
normalized by `N * population_variance`, with a fixed denominator at every lag.
Lag 0 equals one. Explicit lag indexing avoids the reference's even-length
`mode='same'` slicing offset. Constant molecules have undefined ACF and are
excluded from clustering.

All retained ACF features enter centered, unscaled PCA. Correlation-distance
neighbors and weighted Leiden cluster molecules separately in each region.
UMAP displays this same ACF representation. Cluster numbers are local to each
region. The 140–250-bp positive local peak summary is descriptive and is not
a tested periodicity or nucleosome repeat-length estimate.

# Pooled REF/ALT summaries

HP1/HP2 are mapped through each read's phased focal genotype to REF or ALT.
All unique molecules are pooled by this focal allele, without sample grouping,
cell-line weighting, or per-sample coverage thresholds. Sample identifiers
remain in the in-memory read audit but are not displayed in allele plots.

Cluster fractions use clustered reads within each allele as the denominator.
Mean ACF and peak fractions use all valid ACF reads within each allele, including
valid unclustered reads. The ACF figure has three panels: all retained lags,
0–500 bp, and 500–2000 bp. Each shows the same two mean curves, REF and ALT,
using available lags and independent vertical scales. Each unique
molecule receives equal weight. There are no difference curves or uncertainty bands.

The peak figure shows the fraction with a positive local peak in 140–250 bp
and the distributions of peak lag and height among peak-bearing molecules.
Boxes show medians and quartiles, whiskers extend to the most extreme values
within 1.5 interquartile ranges, and points beyond them are individual outliers.
The complete peak search requires at least 252 retained ACF features.
No peak is a measured zero in the peak fraction; absent coverage or an incomplete
search is undefined. Undefined values are omitted or explained in the plot.
These summaries are descriptive; no hypothesis tests are performed.

The older Fiber-seq plots retain their existing annotations and grouping.

# Haplotype repeat-length views

The new haplotype plots use median raw ACFs across all valid unique retained
fibers, independently of cluster membership. REF/ALT are mapped through phased
focal genotypes; sample-local HP1/HP2 labels are not pooled directly. The strongest
positive local peak in 120–250 bp is shown as a candidate ACF-based repeat length.
Qualifying peaks require prominence ≥0.01 ACF units and spacing ≥20 bp; observed
higher-order peaks must lie within ±20 bp of integer multiples. No peak means
no estimate. These are descriptive candidates, not validated nucleosome spacings.
The legacy 140–250-bp peak plots retain their original definition.

```{r cross-locus-figure, results='asis'}
path <- file.path(output, "outputs/locus_haplotype_median_acf.pdf")
if (file.exists(path)) cat("[All loci × REF/ALT median ACF heatmap](locus_haplotype_median_acf.pdf)\n\n")
```

# Figures

```{r region-figures, results='asis'}
for (folder in region_dirs) {
  rid <- basename(folder)
  cat("\n\n## ", rid, "\n\n", sep = "")
  pdfs <- sort(list.files(folder, pattern = "\\.pdf$", full.names = FALSE))
  if (!length(pdfs)) cat("No completed PDF figures found for this region.\n\n")
  for (name in pdfs)
    cat(sprintf("- [%s](%s/%s)\n", name, rid, name))
}
```
