---
title: "LCL Fiber-seq autocorrelation"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = "autocorrelation.html", output_dir = "/project/spott/cshan/fiber-seq/LCL_project/auto_correlation/resolution04/outputs") })
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    self_contained: true
params:
  run_analysis: false
  window_size: 2000
  n_features: null
  n_pcs: 50
  n_neighbors: 10
  resolution: 0.4
  seed: 0
  regions: null
  output_dir: "/project/spott/cshan/fiber-seq/LCL_project/auto_correlation/resolution04"
  python: "/project/spott/cshan/envs/Jupyter-notebook/bin/python"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, out.width = "100%")
project <- "/project/spott/cshan/fiber-seq"
scripts <- file.path(project, "code/clustering_methods/auto_correlation")
output <- normalizePath(params$output_dir, mustWork = FALSE)
if (isTRUE(params$run_analysis)) {
  args <- c(shQuote(file.path(scripts, "01_run_autocorrelation.py")),
    "--window-size", params$window_size, "--n-pcs", params$n_pcs,
    "--n-neighbors", params$n_neighbors, "--resolution", params$resolution,
    "--seed", params$seed,
    "--output-dir", shQuote(output))
  if (!is.null(params$n_features)) args <- c(args, "--n-features", params$n_features)
  for (rid in params$regions) args <- c(args, "--region", shQuote(rid))
  status <- system2(params$python, args)
  if (status != 0L) stop("Autocorrelation pipeline failed; inspect its diagnostic logs")
}
region_dirs <- list.dirs(file.path(output, "outputs"), recursive = FALSE, full.names = TRUE)
region_dirs <- sort(region_dirs[grepl("^rank[0-9]+_", basename(region_dirs))])
if (!is.null(params$regions)) region_dirs <- region_dirs[basename(region_dirs) %in% params$regions]
if (!length(region_dirs)) stop("No region figures found; run 01_run_autocorrelation.py first")
```

# Analysis and parameters

### Input


**Step 1: Original fibertools m6A BED12 records**

The IDs, coordinates, block counts, and block values below come from the
original BED file. All four records have chromosome `chr2` and RGB `128,0,128`.
BED starts are 0-based; BED ends are exclusive. `read_start` and `read_end`
are BED12 columns 7 and 8. Block lists are abbreviated with `…` for display;
`blockCount` gives their full length. Complete, unmodified BED12 lines are
included immediately below the table.

| Read | RID | BED start | BED end | read_start | read_end | Score | Strand | blockCount | blockSizes  | blockStarts  |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | m84241_250908_180102_s3/160175746/ccs | 119282125 | 119298507 | 119282125 | 119298507 | 8 | - | 144 | 0,1,1,1,1, …, 1,1,1 | 0,37,452,618,1019, …, 15880,16060,16381 |
| B | m84241_250515_002859_s3/157422976/ccs | 119280507 | 119293438 | 119280507 | 119293438 | 7 | - | 248 | 0,1,1,1,1, …, 1,1,1 | 0,309,310,312,462, …, 12680,12873,12930 |
| C | m84241_250908_180102_s3/258540726/ccs | 119289060 | 119298334 | 119289060 | 119298334 | 18 | - | 266 | 0,1,1,1,1, …, 1,1,1 | 0,293,299,301,309, …, 9078,9080,9273 |
| D | m84241_250515_002859_s3/165545798/ccs | 119291372 | 119296368 | 119291372 | 119296368 | 32 | + | 120 | 0,1,1,1,1, …, 1,1,1 | 0,117,122,184,277, …, 4815,4821,4995 |

The first and last blocks are fibertools blocks and are removed.
In these real records, the first block has size **0** and the last has size
**1**. Internal m6A-call blocks have size **1**.

**Step 2: Expand blocks and keep calls inside the analysis window**

`extract_ft_region_reads()` retains the longest alignment per RID, expands
BED12 blocks, removes the sentinel blocks, and keeps calls within
119290056–119292055. A call's 1-based coordinate is its block end:

`call position = BED start + blockStarts entry + blockSizes entry`

For example, read A's call at 119291048 is `119282125 + 8922 + 1`.
Strand does not reverse the genomic column order.


| Read | Block number (1-based) | Block offset | Block size | Call position |
| --- | --- | --- | --- | --- |
| A | 82 | 7937 | 1 | 119290063 |
| A | 93 | 8922 | 1 | 119291048 |
| A | 94 | 8924 | 1 | 119291050 |
| A | 95 | 8926 | 1 | 119291052 |
| A | 96 | 8929 | 1 | 119291055 |
| B | 182 | 9563 | 1 | 119290071 |
| C | 43 | 1086 | 1 | 119290147 |
| C | 73 | 1987 | 1 | 119291048 |
| C | 74 | 1989 | 1 | 119291050 |
| C | 75 | 1991 | 1 | 119291052 |
| C | 76 | 1994 | 1 | 119291055 |
| C | 77 | 1998 | 1 | 119291059 |
| D | 2 | 117 | 1 | 119291490 |


**Step 3: Extract one metadata row per read**

`extract_ft_read_info(reads)` uses `BED12 column 7 + 1` as the inclusive
read start and column 8 as the inclusive read end

| Read | Chromosome | Start (1-based) | End (inclusive) | Strand | In-window m6A calls |
| --- | --- | --- | --- | --- | --- |
| A | chr2 | 119282126 | 119298507 | - | 26 |
| B | chr2 | 119280508 | 119293438 | - | 50 |
| C | chr2 | 119289061 | 119298334 | - | 81 |
| D | chr2 | 119291373 | 119296368 | + | 15 |

**Step 4: Construct the read-by-position matrix**


1. Initialize entries to `NA`.
2. Set positions covered by a read to `0`.
3. Set observed m6A-call positions to `1`.

`0` means covered without a reported m6A call; `NA` means uncovered.
Every genomic base is a column, including bases with no calls in any read.
The following values were checked with the current R extraction and matrix
functions against the original indexed BED file.

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |
| D | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA | NA |

D starts at 119291373, so all 12 displayed bases are uncovered (`NA`).
It covers only part of the full analysis window.

**Step 5: Filter coverage and phasing**

| Read | Spans full 2,000 bp? | Haplotype | Phase set | Focal genotype | Resolved allele | Outcome |
| --- | --- | --- | --- | --- | --- | --- |
| A | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| B | Yes | HP2 | 119099734 | 0\|1 | rs12470189: A | Keep |
| C | Yes | HP1 | 119099734 | 0\|1 | rs12470189: G | Keep |
| D | No | Not evaluated | — | — | — | Remove before phasing |

`allele_status = phased_focal_genotype` and pass the phased
heterozygote filter. Requires at least three retained
reads, both focal alleles, and at least two columns with observed m6A calls.


**Step 6: Final m6A matrix**

`lcl_build_phased_m6a_input()` builds all 2,000 genomic columns for the
retained reads

| Read | 119291048 | 119291049 | 119291050 | 119291051 | 119291052 | 119291053 | 119291054 | 119291055 | 119291056 | 119291057 | 119291058 | 119291059 |
| --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- | --- |
| A | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |
| B | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| C | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 1 | 0 | 0 | 0 | 1 |


The defaults are a **2,000-bp window and all 2,000 non-negative ACF lags**.
Set `window_size` to change the window, centered at the same focal SNP;
set `n_features` to retain the first N lags, including lag 0. `null` retains
all lags.

# Method


## Calculate autocorrelation

$ACF(k) = sum((x[t]-mean(x))*(x[t+k]-mean(x))) / (N * var(x))$

for each lag k (distance apart for a given comparison), it compares the centered signal with itself shifted by k bp. Positions with positive deviations after subtracting the mean give positive products. The differences are summed across all overlapping positions (positions that still line up after shifting by the lag), and divided by the read's total variance (variation in m6a pattern across all positions after subtracting the mean, used for normalization)

Autocorrelation calculates a ACF value measures the linear relationship and correlation coefficient between a time serious and a lagged version of itself at various time steps, specifically in Fiber-seq, it answers the question of how similar is the methylation pattern to itself after shifting it by k base pairs?

    - ACF(0) = 1, the pattern is compared to itself, so it matches perfectly
    - ACF(k) > 0, positions k bp apart tend to have the same state 
    - ACF(k) < 0, positions k bp apart tend to have opposite state
    - ACG(k) = 0, little systematic relationship at that spacing


Previous analysis using Fourier transformation does not give base pair resolution, where the resolution/period length is dependent on the frequency (k = 1,2,3…) and the length of the region, autocorrelation bypass this issue because:

  - FFT has uniform frequency resolution but not uniform period resolution -> how much of this molecule can be described by oscillations having the specific frequencies?
  - Autocorrelation samples uniformly in genomic distance
  - Autocorrelation represents the information on a distance axis, whereas the FFT represents it on an evenly spaced frequency axis

Constant molecules have undefined ACF and are excluded from clustering



```{r cross-locus-figure, results='asis'}
path <- file.path(output, "outputs/locus_haplotype_median_acf.pdf")
if (file.exists(path)) cat("[All loci × REF/ALT median ACF heatmap](locus_haplotype_median_acf.pdf)\n\n")
```

# Figures

```{r region-figures, results='asis'}
for (folder in region_dirs) {
  rid <- basename(folder)
  cat("\n\n## ", rid, "\n\n", sep = "")
  pdfs <- sort(list.files(folder, pattern = "\\.pdf$", full.names = FALSE))
  if (!length(pdfs)) cat("No completed PDF figures found for this region.\n\n")
  for (name in pdfs)
    cat(sprintf("- [%s](%s/%s)\n", name, rid, name))
}
```
