Define accessibility and footprints (https://doi.org/10.1101/2025.10.28.685222)

I want to first look at PolII footprints 
  - 40 - 60 bp
  - overlapping signal from PRO-seq
  
PIC footprints
  - 20 - 40 bp overlapping a TATA box
  - 60 - 80 bp overlapping the TSS defined by CAGE-seq
  
TF footprints
  - < 90 bp, not overlapping PRO-seq signal
  



```{bash, eval = FALSE}
#!/bin/bash
#SBATCH --job-name=fire_metagene
#SBATCH --time=24:00:00
#SBATCH --mem=100G
#SBATCH --cpus-per-task=10
#SBATCH --account=pi-spott
#SBATCH --partition=spott
#SBATCH --array=0-33
#SBATCH --output=/project/spott/cshan/fiber-seq/results/logs/fire_metagene_%A_%a.out

module load python/miniforge-25.3.0
source activate /project/spott/cshan/envs/pb-CpG-tools

FIRE_dir="/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/"
ref_genome="/project/spott/reference/human/GRCh38/hg38.fa"
gtf="/project/spott/cshan/annotations/gencode.v49.primary_assembly.annotation.gtf"
cpg_output_base="/project/spott/cshan/fiber-seq/results/fire_CpG"
plot_output_dir="/project/spott/cshan/fiber-seq/results/plots/metagene"
threads=16

# Create output directories
mkdir -p ${cpg_output_base}
mkdir -p ${plot_output_dir}
mkdir -p /project/spott/cshan/fiber-seq/results/logs

# Get list of all sample directories
mapfile -t sample_dirs < <(find ${FIRE_dir} -mindepth 1 -maxdepth 1 -type d | sort)

# Get the sample directory for this array task
sample_dir="${sample_dirs[$SLURM_ARRAY_TASK_ID]}"
sample=$(basename ${sample_dir})

echo "========================================="
echo "Array task ID: ${SLURM_ARRAY_TASK_ID}"
echo "Processing sample: ${sample}"
echo "Sample directory: ${sample_dir}"
echo "========================================="

# Find the CRAM file
cram_file=$(find ${sample_dir} -maxdepth 1 -name "*-fire-v0.1-filtered.cram" -o -name "*-fire-v0.1.1-filtered.cram" | head -n 1)

if [ -z "${cram_file}" ]; then
    echo "ERROR: No CRAM file found in ${sample_dir}"
    exit 1
fi

echo "Found CRAM: ${cram_file}"

# Set up CpG output directory for this sample
cpg_output_dir="${cpg_output_base}/${sample}"
mkdir -p ${cpg_output_dir}

# Run aligned_bam_to_cpg_scores if not already done
if [ ! -f "${cpg_output_dir}/${sample}_CPG.combined.bw" ]; then
    echo "Running aligned_bam_to_cpg_scores..."
    
    aligned_bam_to_cpg_scores \
        --bam ${cram_file} \
        --output-prefix ${cpg_output_dir}/${sample}_CPG \
        --ref ${ref_genome} \
        --threads ${threads}
    
    echo "CpG scores complete for ${sample}"
else
    echo "CpG scores already exist, skipping..."
fi

# Collect all BigWig files (FIRE trackHub + CpG outputs)
bw_dir="${sample_dir}/trackHub-v0.1/bw"

# Create list of all BigWig files to process
bw_files=()

# Add FIRE BigWigs
if [ -d "${bw_dir}" ]; then
    for bw in ${bw_dir}/*.bw; do
        [ -f "${bw}" ] && bw_files+=("${bw}")
    done
fi

# Add CpG BigWigs
for bw in ${cpg_output_dir}/*.bw; do
    [ -f "${bw}" ] && bw_files+=("${bw}")
done

# Check if we have any BigWig files
if [ ${#bw_files[@]} -eq 0 ]; then
    echo "ERROR: No BigWig files found for ${sample}"
    exit 1
fi

echo "Found ${#bw_files[@]} BigWig files to process"

# Switch to deeptools environment for plotting
source activate /project/spott/cshan/envs/deeptools

# Process each BigWig file
for bw in "${bw_files[@]}"; do
    label=$(basename ${bw} .bw)
    echo "  Processing ${label}..."
    
    # Compute matrix around genes (scale-regions mode)
    computeMatrix scale-regions \
        -S ${bw} \
        -R ${gtf} \
        --beforeRegionStartLength 3000 \
        --regionBodyLength 5000 \
        --afterRegionStartLength 3000 \
        --skipZeros \
        -p ${threads} \
        -o ${plot_output_dir}/${sample}_${label}.matrix.gz
    
    # Plot heatmap
    plotHeatmap -m ${plot_output_dir}/${sample}_${label}.matrix.gz \
        --plotFileFormat pdf \
        --samplesLabel ${label} \
        --heatmapWidth 8 \
        -out ${plot_output_dir}/${sample}_${label}_heatmap.pdf
    
    echo "  Created heatmap: ${plot_output_dir}/${sample}_${label}_heatmap.pdf"
    
    # Plot profile
    plotProfile -m ${plot_output_dir}/${sample}_${label}.matrix.gz \
        --plotFileFormat pdf \
        --samplesLabel ${label} \
        --perGroup \
        -out ${plot_output_dir}/${sample}_${label}_profile.pdf
    
    echo "  Created profile: ${plot_output_dir}/${sample}_${label}_profile.pdf"
done

echo "========================================="
echo "Completed ${sample}"
echo "CpG BigWigs saved to: ${cpg_output_dir}"
echo "Plots saved to: ${plot_output_dir}"
echo "========================================="
```

trying to plot TSS and TTS in the same metagene plot
/project/spott/cshan/fiber-seq/code/plot_metagene_Dnase_linker_nuc.sh

```{r setup, include=FALSE}
knitr::opts_chunk$set( echo = TRUE, warning = FALSE, message = FALSE )

library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(data.table)
library(Gviz)
```

look at nucleosome footprints at individual fire called by ft extract
```{r}
# first look at a single bed file from ft extract
nuc_bed_path <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL10_bc2178_19130/extracted_results/nuc_by_chr/AL10_bc2178_19130.ft_extracted_nuc.chr1.bed.gz"
tss_path     <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"

# ── 1. get nucleosome midpoints ────────────────────────────────────

# blockcount = # of nucleosomes in the region
# blockSizes = comma-separated sizes of each nucleosome
# blockStarts = comma-separated start positions of each nucleosome relative to chromStart

bed12 <- fread(
  nuc_bed_path,
  sep = "\t", header = FALSE,
  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                "thickStart","thickEnd","itemRgb",
                "blockCount","blockSizes","blockStarts")
)

# Expand each row into individual nucleosomes -- compute absolute start/end and midpoint for each nucleosome
nucs <- bed12[, {
  sizes  <- as.integer(strsplit(blockSizes,  ",")[[1]])
  starts <- as.integer(strsplit(blockStarts, ",")[[1]])
  # drop trailing empty strings (some rows end with ",")
  # remove the first and last block (The first and last block in each the bed12 record do not reflect real data, and exist only to mark the start and end positions of the read, so we want to drop them
  if (length(sizes) >= 2L) {
    idx <- seq_along(sizes)
    idx <- idx[-c(1L, length(idx))]
    sizes <- sizes[idx]
    starts <- starts[idx]
  }
  sizes  <- sizes[!is.na(sizes)]
  starts <- starts[!is.na(starts)]
  abs_start <- chromStart + starts          # 0-based
  abs_end   <- chromStart + starts + sizes  # 0-based exclusive
  midpoint  <- as.integer((abs_start + abs_end) / 2)
  .(nuc_mid = midpoint, nuc_size = sizes)
}, by = .(chrom, chromStart, chromEnd, name, strand)]

# GRanges of nucleosome midpoints (1-bp, 1-based)
nuc_gr <- GRanges(
  seqnames = nucs$chrom,
  ranges   = IRanges(nucs$nuc_mid + 1L, nucs$nuc_mid + 1L),  # 0->1 based
  strand   = "*",
  nuc_size = nucs$nuc_size
)

# ── 2. Load TSS annotations ───────────────────────────────────────────────────
tss <- fread(tss_path, header = FALSE,
             col.names = c("chrom","start","end","gene_id","score","strand"))

tss_gr <- GRanges(
  seqnames = tss$chrom,
  ranges   = IRanges(tss$start + 1L, tss$end),  # BED -> 1-based
  strand   = tss$strand
)

# ── 3. Find nucleosome midpoints within ±2 kb of each TSS ────────────────────
window_bp <- 2000L

tss_windows <- promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)

# pair overlaps between nucleosome midpoints and TSS windows
hits <- findOverlaps(nuc_gr, tss_windows)

# Compute strand-aware distance from TSS

# For "+" strand TSS: distance = nuc_pos - tss_pos
# For "-" strand TSS: distance = tss_pos - nuc_pos

# pull out nucl
nuc_pos  <- start(nuc_gr)[GenomicRanges::queryHits(hits)]
tss_pos  <- start(tss_gr)[GenomicRanges::subjectHits(hits)]   # TSS = 1-based position
tss_str  <- as.character(strand(tss_gr))[GenomicRanges::subjectHits(hits)]

rel_pos <- ifelse(tss_str == "+",
                  nuc_pos - tss_pos,
                  tss_pos - nuc_pos)

# ── 4. Meta-profile plot ──────────────────────────────────────────────────────
plot_dt <- data.table(rel_pos = rel_pos)

# bin into 10-bp bins
bin_size <- 10L
# assgin each nucleosome midpoint to a bin based on its distance from the TSS
plot_dt[, bin := round(rel_pos / bin_size) * bin_size]
# count number of nucleosome midpoints in each bin
meta <- plot_dt[, .N, by = bin]
# order bins for plotting
setorder(meta, bin)

AL10_chr1_nuc <- ggplot(meta, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.6) +
  geom_smooth(method = "loess", span = 0.05, se = FALSE, color = "steelblue", linewidth = 1) +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  labs(
    x    = "Distance to TSS",
    y    = "Nucleosome midpoint count (chr1)",
    title = "Nucleosome midpoint distribution around TSS",
    subtitle = sprintf("Sample: AL10_bc2178_19130 | n nucleosome calls = %s", format(length(nuc_gr), big.mark=","))
  ) +
  theme_bw(base_size = 12)

pdf("/project/spott/cshan/fiber-seq/results/plots/nuc/AL10_chr1_nuc_TSS.pdf", width = 6, height = 4)
print(AL10_chr1_nuc)
dev.off()

```

define functions
```{r define-functions}
# bed12 -> nucleosome midpoints
calculate_nucleosome_midpoints <- function(bed12) {
  bed12[, {
    parsed <- parse_bed12_blocks(blockSizes, blockStarts,
                                 drop_terminal_blocks = TRUE)
    sizes <- parsed$sizes
    starts <- parsed$starts
    if (!length(sizes)) return(NULL)
    abs_start <- chromStart + starts
    abs_end   <- chromStart + starts + sizes
    midpoint  <- as.integer((abs_start + abs_end) / 2)
    .(chrom = chrom, nuc_mid = midpoint, nuc_size = sizes)
  }, by = .(chrom, chromStart, chromEnd, name, strand)]
}


# load TSS BED -> GRanges
load_tss_annotations <- function(tss_path) {
  tss <- fread(tss_path, header = FALSE,
               col.names = c("chrom", "start", "end", "gene_id", "score", "strand"))
  GRanges(seqnames = tss$chrom,
          ranges   = IRanges(tss$start + 1L, tss$end),
          strand   = tss$strand,
          gene_id  = tss$gene_id)
}

# nuc midpoints near TSS windows

find_nucs_near_tss <- function(nuc_gr, tss_gr, window_bp = 2000L) {
  tss_windows <- promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)
  hits <- findOverlaps(nuc_gr, tss_windows)

  nuc_pos <- start(nuc_gr)[queryHits(hits)]
  tss_pos <- start(tss_gr)[subjectHits(hits)]
  tss_str <- as.character(strand(tss_gr))[subjectHits(hits)]

  data.table(
    seqnames = as.character(seqnames(tss_gr))[subjectHits(hits)],
    strand   = tss_str,
    nuc_pos  = nuc_pos,
    tss_pos  = tss_pos,
    rel_pos  = ifelse(tss_str == "+", nuc_pos - tss_pos, tss_pos - nuc_pos)
  )
}

# metaprofile plot of midpoints around TSS
create_meta_profile_plot <- function(plot_dt, window_bp = 2000L) {
  bin_size <- 10L
  plot_dt[, bin := round(rel_pos / bin_size) * bin_size]
  meta <- plot_dt[, .N, by = bin][order(bin)]
  ggplot(meta, aes(x = bin, y = N)) +
    geom_line(linewidth = 0.6) +
    geom_smooth(method = "loess", span = 0.05, se = FALSE,
                color = "steelblue", linewidth = 1) +
    scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                       labels = function(x) paste0(x/1000, " kb")) +
    geom_vline(xintercept = 0, linetype = "dashed",
               color = "red", linewidth = 0.5) +
    labs(x = "Distance to TSS", y = "Nucleosome midpoint count",
         title = "Nucleosome midpoint distribution around CAGE-defined TSS") +
    theme_bw(base_size = 12)
}

# per-TSS normalized metaprofile
make_meta_normalized_TSS <- function(nuc_gr, tss_gr, window_bp = 2000L,
                                     bin_size = 10L, label = NULL) {
  plot_dt <- find_nucs_near_tss(nuc_gr, tss_gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / bin_size) * bin_size]
  meta <- plot_dt[, .N, by = bin][order(bin)]
  meta[, N := N / length(tss_gr)]
  if (!is.null(label)) meta[, source := label]
  meta
}

# bed12 files -> nucleosome midpoint GRanges
load_nucleosome_annotations <- function(bed_files) {
  beds <- lapply(bed_files, fread, header = FALSE,
                 col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                               "thickStart","thickEnd","itemRgb","blockCount",
                               "blockSizes","blockStarts"))
  mids <- rbindlist(lapply(beds, calculate_nucleosome_midpoints))
  nuc_midpoints_to_granges(mids)
}


# raw counts per bin
make_meta <- function(tss_gr, label) {
  plot_dt <- find_nucs_near_tss(nuc_gr, tss_gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / 10L) * 10L]   # 10 bp bins
  meta <- plot_dt[, .N, by = bin][order(bin)]
  meta[, source := label]
  meta
}


# BED9 (chrom, start, end, name, score, strand, thickStart, thickEnd, itemRgb)
load_tss_annotations_bed9 <- function(tss_path) {
  cols <- c("chrom","start","end","name","score","strand",
            "thickStart","thickEnd","itemRgb")
  tss <- data.table::fread(tss_path, header = FALSE, col.names = cols)
  GenomicRanges::GRanges(
    seqnames = tss$chrom,
    ranges   = IRanges::IRanges(tss$start + 1L, tss$end),  # BED is 0‑based, half‑open
    strand   = tss$strand,
    tss_id   = tss$name,
    score    = tss$score
  )
}

# function used to load CAGE TSS annotations with gene_id in column 4 instead of name
load_tss_annotations_bed9_with_gene_id <- function(path) {
  cols <- c("chrom","start","end","gene_id","score","strand","thickStart","thickEnd","itemRgb")
  dt <- data.table::fread(path, header = FALSE, col.names = cols)
  GenomicRanges::GRanges(
    seqnames = dt$chrom,
    ranges   = IRanges::IRanges(dt$start + 1L, dt$end),
    strand   = dt$strand,
    gene_id  = dt$gene_id,
    score    = dt$score
  )
}

# ─── PER-READ AWARE FUNCTIONS  ────────────────────────────────

# Step 1: Build GRanges of reads (one row per sequencing molecule) with RID
build_reads_granges <- function(bed12) {
  GRanges(
    seqnames = bed12$chrom,
    ranges   = IRanges::IRanges(bed12$chromStart + 1L, bed12$chromEnd),
    strand   = "*",
    RID      = bed12$name
  )
}

# Step 2: Extract nucleosomes per read, preserving RID
extract_nucleosomes_per_read <- function(bed12) {
  nucs <- bed12[, {
    sizes  <- as.integer(strsplit(blockSizes, ",")[[1]])
    starts <- as.integer(strsplit(blockStarts, ",")[[1]])
    keep <- !is.na(sizes) & !is.na(starts)
    sizes <- sizes[keep]
    starts <- starts[keep]
    if (length(sizes) >= 2L) {
      idx <- seq_along(sizes)
      idx <- idx[-c(1L, length(idx))]  # Remove first and last block
      sizes <- sizes[idx]
      starts <- starts[idx]
    }
    if (!length(sizes)) return(NULL)
    abs_start <- chromStart + starts
    abs_end   <- chromStart + starts + sizes
    nuc_mid <- as.integer((abs_start + abs_end) / 2)
    .(RID = name, nuc_mid = nuc_mid, nuc_size = sizes)
  }, by = .(chrom, chromStart, chromEnd, strand)]
  nucs
}

# Main per-read function: Find nucleosomes on reads that overlap TSS windows
find_nucs_near_tss_per_read <- function(bed12, tss_gr, window_bp = 2000L, n_ranks = 5L) {
  # Step 1: Build reads GRanges
  reads_gr <- build_reads_granges(bed12)
  
  # Step 2: Extract nucleosomes per read
  nuc_read_cols <- extract_nucleosomes_per_read(bed12)
  
  # Step 3: Create TSS lookup table
  tss_dt <- data.table(
    tss_idx    = seq_along(tss_gr),
    tss_pos    = start(tss_gr),
    tss_strand = as.character(strand(tss_gr))
  )
  
  # Step 4: Find reads overlapping TSS windows
  tss_windows_gr <- promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)
  hits_tss <- findOverlaps(reads_gr, tss_windows_gr)
  
  read_tss_pairs <- data.table(
    RID     = reads_gr$RID[S4Vectors::queryHits(hits_tss)],
    tss_idx = S4Vectors::subjectHits(hits_tss)
  )
  
  # Step 5: Join reads, TSS, and nucleosomes
  rtp_nuc <- merge(read_tss_pairs, nuc_read_cols, by = "RID", allow.cartesian = TRUE)
  rtp_nuc <- merge(rtp_nuc, tss_dt, by = "tss_idx", allow.cartesian = TRUE)
  
  # Step 6: Compute strand-aware offsets
  rtp_nuc[, rel_pos  := fifelse(tss_strand == "+",
                                (nuc_mid + 1L) - tss_pos,
                                tss_pos - (nuc_mid + 1L))]
  rtp_nuc[, abs_dist := abs(rel_pos)]
  rtp_nuc <- rtp_nuc[abs_dist <= window_bp]
  rtp_nuc[, direction := fifelse(rel_pos < 0L, "upstream", "downstream")]

  # Step 7: Rank nucleosomes within each (TSS, read, direction)
  setorder(rtp_nuc, tss_idx, RID, direction, abs_dist)
  rtp_nuc[, rank := seq_len(.N), by = .(tss_idx, RID, direction)]
  rtp_nuc <- rtp_nuc[rank <= n_ranks]
  
  rtp_nuc[, signed_rank := fifelse(direction == "upstream", -rank, rank)]
  
  rtp_nuc
}

# Per-read version with gene information
find_nucs_near_tss_with_gene_per_read <- function(bed12, tss_gr, window_bp = 2000L, n_ranks = 5L) {
  # Build everything per-read first
  rtp_nuc <- find_nucs_near_tss_per_read(bed12, tss_gr, window_bp, n_ranks)
  
  # Add gene_id from TSS GRanges
  gene_dt <- data.table(
    tss_idx = seq_along(tss_gr),
    gene_id = as.character(mcols(tss_gr)$gene_id)
  )
  rtp_nuc <- merge(rtp_nuc, gene_dt, by = "tss_idx", all.x = TRUE)
  
  rtp_nuc
}

# ─── META-PROFILE FUNCTIONS FOR PER-READ DATA ──────────────────────────────

# Raw meta-profile: counts per bin (per-read aware)
# Returns: bin, N (count), n_reads (number of unique reads per bin)
make_meta_per_read_raw <- function(bed12, tss_gr, window_bp = 2000L, bin_size = 10L, label = NULL) {
  plot_dt <- find_nucs_near_tss_per_read(bed12, tss_gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / bin_size) * bin_size]
  
  meta <- plot_dt[, .(
    N = .N,                          # Total nucleosome observations
    n_reads = uniqueN(RID),          # Number of unique reads per bin
    n_tss = uniqueN(tss_idx)         # Number of unique TSS per bin
  ), by = bin][order(bin)]
  
  if (!is.null(label)) meta[, source := label]
  meta
}

# Per-read normalized meta-profile: mean nucleosomes per read
# Accounts for varying numbers of reads per TSS
make_meta_per_read_normalized <- function(bed12, tss_gr, window_bp = 2000L, bin_size = 10L, label = NULL) {
  plot_dt <- find_nucs_near_tss_per_read(bed12, tss_gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / bin_size) * bin_size]
  
  # Count unique (read, TSS) pairs first to normalize
  read_tss_pairs <- plot_dt[, .(n_reads = uniqueN(RID)), by = tss_idx]
  total_read_tss_pairs <- sum(read_tss_pairs$n_reads)
  
  meta <- plot_dt[, .(
    N = .N / total_read_tss_pairs,      # Normalize by total read-TSS pairs
    n_reads = uniqueN(RID),
    n_tss = uniqueN(tss_idx)
  ), by = bin][order(bin)]
  
  if (!is.null(label)) meta[, source := label]
  meta
}

# Generic meta-profile function: flexible normalization
# normalize_by: "none" (raw counts), "tss" (per TSS), "read_tss" (per read-TSS pair)
make_position_meta_per_read <- function(bed12, tss_gr, window_bp = 2000L, bin_size = 10L,
                                        normalize_by = "none", label = NULL) {
  plot_dt <- find_nucs_near_tss_per_read(bed12, tss_gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / bin_size) * bin_size]
  
  meta <- plot_dt[, .(
    N_raw = .N,
    n_reads = uniqueN(RID),
    n_tss = uniqueN(tss_idx)
  ), by = bin]
  
  # Apply normalization
  if (normalize_by == "tss") {
    meta[, N := N_raw / length(tss_gr)]
  } else if (normalize_by == "read_tss") {
    # Count unique (read, TSS) pairs
    read_tss_pairs <- plot_dt[, .(n_reads = uniqueN(RID)), by = tss_idx]
    total_read_tss_pairs <- sum(read_tss_pairs$n_reads)
    meta[, N := N_raw / total_read_tss_pairs]
  } else {
    meta[, N := N_raw]
  }
  
  setorder(meta, bin)
  meta[, N_raw := NULL]  # Drop intermediate column
  
  if (!is.null(label)) meta[, source := label]
  meta
}

# Gene-level per-read summary
summarize_gene_tss_per_read <- function(rtp_nuc_with_gene) {
  if (!nrow(rtp_nuc_with_gene) || !"gene_id" %in% names(rtp_nuc_with_gene)) {
    return(data.table(
      gene_id = character(),
      mean_rel_pos_bp = numeric(),
      median_rel_pos_bp = numeric(),
      mean_abs_rel_pos_bp = numeric(),
      median_abs_rel_pos_bp = numeric(),
      sd_rel_pos_bp = numeric(),
      n_nuc = integer(),
      n_reads = integer(),
      n_tss = integer()
    ))
  }
  
  rtp_nuc_with_gene[, .(
    mean_rel_pos_bp = mean(rel_pos, na.rm = TRUE),
    median_rel_pos_bp = as.integer(median(rel_pos, na.rm = TRUE)),
    mean_abs_rel_pos_bp = mean(abs_dist, na.rm = TRUE),
    median_abs_rel_pos_bp = as.integer(median(abs_dist, na.rm = TRUE)),
    sd_rel_pos_bp = sd(rel_pos, na.rm = TRUE),
    n_nuc = .N,
    n_reads = uniqueN(RID),
    n_tss = uniqueN(tss_idx)
  ), by = gene_id][order(mean_abs_rel_pos_bp, -n_nuc)]
}


```

for now, focus on one sample (AL10_bc2178_19130), but look at all chromsomes and look at the size distritbution of all nucleosomes
```{r}
# define file path for the sample
nuc_by_chr_dir <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL10_bc2178_19130/extracted_results/nuc_by_chr/"

# list all bed files for the sample
bed_files <- list.files(nuc_by_chr_dir, pattern = "ft_extracted_nuc.chr.*\\.bed\\.gz$", full.names = TRUE)

# initialize an empty data.table to hold all nucleosome sizes
all_nucs <- data.table()

# function to select one chr at a time and extract nucleosome sizes
extract_nuc_sizes <- function(bed12) {
    # split bloksizes by comma and convert to numeric
    bed12[, `blockSizes` := strsplit(`blockSizes`, ",")]
    bed12 <- bed12[, .(chrom, `blockSizes`)]  # keep only chrom and blockSizes for now
    bed12 <- bed12[, .(blockSizes = unlist(`blockSizes`)), by = chrom]  # unnest blockSizes
    bed12[, blockSizes := as.numeric(blockSizes)]  # convert to numeric
    
    return(bed12[, .(chrom, nuc_size = blockSizes)])
    }

# bin nucleosome sizes into biologically meaningful size classes
# canonical nucleosome wraps ~146 bp of DNA
bin_nuc_sizes <- function(dt) {
    dt[, size_class := fcase(
        nuc_size <  100,                   "< 100 bp",
        nuc_size >= 100 & nuc_size < 130,  "100-130 bp",
        nuc_size >= 130 & nuc_size < 160,  "130-160 bp",
        nuc_size >= 160 & nuc_size < 190,  "160-190 bp",
        nuc_size >= 190,                   "> 190 bp"
    )]
    dt[, size_class := factor(size_class, levels = c(
        "< 100 bp",
        "100-130 bp",
        "130-160 bp",
        "160-190 bp",
        "> 190 bp"
    ))]
    dt
}

# for a given chr, read the bed file, extract nucleosome sizes, and return a data.table with chrom and nuc_size
process_chr_bed <- function(bed_file) {
    bed12 <- fread(
    bed_file,
    header = FALSE,
    col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                  "thickStart","thickEnd","itemRgb",
                  "blockCount","blockSizes","blockStarts")
    )   

    # extract nucleosome sizes
    nuc_sizes <- extract_nuc_sizes(bed12)

    return(nuc_sizes)
}


# chr1_file <- bed_files[grepl("\\.chr1\\.", bed_files)]

# bed12_chr1 <- fread(
#     chr1_file,
#     header = FALSE,
#     col.names = c("chrom","chromStart","chromEnd","name","score","strand",
#                   "thickStart","thickEnd","itemRgb",
#                   "blockCount","blockSizes","blockStarts")
# )


# chr1_nucs <- extract_nuc_sizes(bed12_chr1)
# summary(chr1_nucs$nuc_size)

# plot nucleosome size distribution for a given chromosome
plot_nuc_size_dist <- function(bed_files, chr = "chr1", sample_name = "AL10_bc2178_19130") {
    # match exactly: e.g. "chr1" should not match "chr10", "chr11", etc.
    pattern <- paste0("\\.", chr, "\\.")
    matched <- bed_files[grepl(pattern, bed_files)]
    if (length(matched) == 0) stop("No bed file found for chromosome: ", chr)

    nucs <- rbindlist(lapply(matched, process_chr_bed))
    nucs <- bin_nuc_sizes(nucs)

    size_class_colors <- c(
        "< 100 bp"    = "#d62728",
        "100-130 bp"  = "#ff7f0e",
        "130-160 bp"  = "#2ca02c",
        "160-190 bp"  = "#1f77b4",
        "> 190 bp"    = "#9467bd"
    )

    counts <- nucs[, .N, by = size_class]

    ggplot(counts, aes(x = size_class, y = N, fill = size_class)) +
        geom_col() +
        geom_text(aes(label = format(N, big.mark = ",")), vjust = -0.4, size = 3) +
        scale_fill_manual(values = size_class_colors, guide = "none") +
        scale_y_continuous(expand = expansion(mult = c(0, 0.1)),
                           labels = scales::comma) +
        labs(
            x = "Size class",
            y = "Number of nucleosomes",
            title = sprintf("Nucleosome count per size class on %s", chr),
            subtitle = sprintf("Sample: %s | total = %s",
                               sample_name, format(nrow(nucs), big.mark = ","))
        ) +
        theme_bw(base_size = 12)
}

nuc_size_dist_chr1 <- plot_nuc_size_dist(bed_files, chr = "chr1")
print(nuc_size_dist_chr1)

pdf("/project/spott/cshan/fiber-seq/results/plots/nuc/AL10_nuc_chr1_size_distribution.pdf", width = 6, height = 4)
print(nuc_size_dist_chr1)
dev.off()

```


Use cage-seq to define TSS and look at nucleosome distribution around TSS
    - maps TSS and mesures promoter acitivity by sequence tags that are generated by reverse transcription of 5' capped RNA

cage-seq data for GM12878 is available from ENCODE (https://www.encodeproject.org/experiments/ENCSR000CKA/), and I used ENCFF715BEB.gff

```{bash, eval = FALSE}
# https://pughlab.mbg.cornell.edu/scriptmanager-docs/docs/Tools/coordinate-manipulation/gff-to-bed/ was used to convert the GFF to BED
# /project/spott/cshan/annotations

java -jar /project/spott/cshan/tools/ScriptManager-v0.15.jar coordinate-manipulation gff-to-bed \
  ENCFF715BEB.gff \
  -o ENCFF715BEB_CAGE.bed


# format the CAGE-seq BED file to have the same 6 columns as the TSS annotation BED file, and then intersect with TSS annotation to assign gene names to CAGE peaks

awk 'BEGIN{OFS="\t"} $1=="track"{next} {print $1,$2,$3,".",".",$6}' /project/spott/cshan/annotations/ENCFF715BEB_CAGE.bed \
  | /project/spott/cshan/envs/bedtools/bin/bedtools intersect -a stdin -b /project/spott/cshan/annotations/TSS.gencode.v49.bed -wa -wb \
  | awk 'BEGIN{OFS="\t"} {print $1,$2,$3,$10,".",$12}' \
  > /project/spott/cshan/annotations/ENCFF715BEB_CAGE_withGenes.bed


```

```{r CAGE_TSS}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

tss_path       <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"
sample_root    <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample <- "AL10_bc2178_19130"
default_chrom  <- "chr1"
window_bp      <- 2000L



selected_sample <- file.path(sample_root, default_sample)
nuc_by_chr_dir  <- file.path(selected_sample, "extracted_results/nuc_by_chr/")
bed_files <- list.files(nuc_by_chr_dir, pattern = "ft_extracted_nuc.chr.*\\.bed\\.gz$", full.names = TRUE)
if (!length(bed_files)) stop("No BED files found in ", nuc_by_chr_dir)

selected_chrom <- default_chrom
pattern  <- paste0("\\.", selected_chrom, "\\.")
bed_file <- bed_files[grepl(pattern, bed_files)][1]
if (is.na(bed_file) || !file.exists(bed_file)) stop("Could not find bed file for ", selected_chrom)

bed12 <- fread(bed_file, header = FALSE,
               col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                             "thickStart","thickEnd","itemRgb",
                             "blockCount","blockSizes","blockStarts"))

nucs    <- calculate_nucleosome_midpoints(bed12)
nuc_gr  <- nuc_midpoints_to_granges(nucs)
tss_gr  <- load_tss_annotations_bed9(tss_path)
plot_dt <- find_nucs_near_tss(nuc_gr, tss_gr, window_bp = window_bp)

# raw meta plot
meta_plot <- create_meta_profile_plot(plot_dt, window_bp = window_bp) +
  labs(subtitle = sprintf("Sample: %s | Chr: %s | n nucleosomes = %s",
                          default_sample, selected_chrom,
                          format(length(nuc_gr), big.mark = ",")))
print(meta_plot)

# per-TSS normalized plot
meta_norm <- make_meta_normalized_TSS(nuc_gr, tss_gr, window_bp = window_bp,
                                      bin_size = 10L, label = "CAGE TSS")
meta_plot_norm <- ggplot(meta_norm, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.7, color = "#d62728") +
  geom_smooth(method = "loess", span = 0.06, se = FALSE,
              linewidth = 1, color = "#d62728") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Avg nucleosome midpoints per TSS (10 bp bins)",
       title = "CAGE TSS",
       subtitle = sprintf("Sample: %s | Chr: %s", default_sample, selected_chrom)) +
  theme_bw(base_size = 12)
print(meta_plot_norm)

# save PDFs
out_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc/CAGE_TSS"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(out_dir,
                 sprintf("%s_%s_CAGE_TSS_nucleosome_midpoints_raw.pdf",
                         default_sample, selected_chrom)),
       plot = meta_plot, width = 6, height = 4)

ggsave(file.path(out_dir,
                 sprintf("%s_%s_CAGE_TSS_nucleosome_midpoints_perTSS.pdf",
                         default_sample, selected_chrom)),
       plot = meta_plot_norm, width = 6, height = 4)

```

# Per-read aware CAGE TSS analysis (single chromosome)

```{r CAGE_TSS_per_read}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

tss_path       <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"
sample_root    <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample <- "AL10_bc2178_19130"
default_chrom  <- "chr1"
window_bp      <- 2000L
bin_size       <- 10L

# Load BED12 for selected chromosome
selected_sample <- file.path(sample_root, default_sample)
nuc_by_chr_dir  <- file.path(selected_sample, "extracted_results/nuc_by_chr/")
bed_files <- list.files(nuc_by_chr_dir, pattern = "ft_extracted_nuc.chr.*\\.bed\\.gz$", full.names = TRUE)
if (!length(bed_files)) stop("No BED files found in ", nuc_by_chr_dir)

selected_chrom <- default_chrom
pattern  <- paste0("\\.", selected_chrom, "\\.")
bed_file <- bed_files[grepl(pattern, bed_files)][1]
if (is.na(bed_file) || !file.exists(bed_file)) stop("Could not find bed file for ", selected_chrom)

bed12 <- fread(bed_file, header = FALSE,
               col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                             "thickStart","thickEnd","itemRgb",
                             "blockCount","blockSizes","blockStarts"))

tss_gr <- load_tss_annotations_bed9(tss_path)

# ── Per-read raw meta-profile (counts per bin) ──
meta_per_read_raw <- make_meta_per_read_raw(bed12, tss_gr, window_bp = window_bp, 
                                             bin_size = bin_size, label = "CAGE TSS")

meta_plot_per_read_raw <- ggplot(meta_per_read_raw, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.6, color = "#2ca02c") +
  geom_smooth(method = "loess", span = 0.05, se = FALSE, color = "#2ca02c", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Nucleosome count (per-read aware)",
       title = "CAGE TSS (per-read method - raw counts)",
       subtitle = sprintf("Sample: %s | Chr: %s | Each read-TSS pair counted", 
                          default_sample, selected_chrom)) +
  theme_bw(base_size = 12)

print(meta_plot_per_read_raw)

# ── Per-read normalized meta-profile ──
meta_per_read_norm <- make_meta_per_read_normalized(bed12, tss_gr, window_bp = window_bp,
                                                     bin_size = bin_size, label = "CAGE TSS")

meta_plot_per_read_norm <- ggplot(meta_per_read_norm, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.7, color = "#d62728") +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, color = "#d62728", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Avg nucleosomes per read-TSS pair",
       title = "CAGE TSS (per-read)",
       subtitle = sprintf("Sample: %s | Chr: %s | Normalized by read-TSS pairs", 
                          default_sample, selected_chrom)) +
  theme_bw(base_size = 12)

print(meta_plot_per_read_norm)

# ── Show reads contributing to each bin ──
reads_per_bin <- meta_per_read_raw[, .(bin, n_reads)]
print(paste("Bin coverage info (sample):"))
print(head(reads_per_bin, 15))

# Save plots
out_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc/per_read_CAGE_TSS"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(out_dir,
                 sprintf("%s_%s_CAGE_TSS_per_read_raw.pdf",
                         default_sample, selected_chrom)),
       plot = meta_plot_per_read_raw, width = 6, height = 4)

ggsave(file.path(out_dir,
                 sprintf("%s_%s_CAGE_TSS_per_read_normalized.pdf",
                         default_sample, selected_chrom)),
       plot = meta_plot_per_read_norm, width = 6, height = 4)

# Save meta data
fwrite(meta_per_read_raw,
       file.path(out_dir,
                 sprintf("%s_%s_CAGE_TSS_per_read_raw_meta.tsv",
                         default_sample, selected_chrom)),
       sep = "\t")

fwrite(meta_per_read_norm,
       file.path(out_dir,
                 sprintf("%s_%s_CAGE_TSS_per_read_normalized_meta.tsv",
                         default_sample, selected_chrom)),
       sep = "\t")
```

Aggregate CAGE TSS distances by gene across all chromosomes:
1. Use the gene-annotated CAGE BED file so each TSS window carries a `gene_id`.
2. Load every chromosome-specific nucleosome BED12 file for the sample.
3. Convert all nucleosome blocks to midpoint coordinates and combine them across chromosomes.
4. Find midpoint overlaps within the same +/- 2 kb window around each CAGE TSS.
5. Summarize signed and absolute distance-to-TSS per gene and save the result for downstream analysis.

```{r CAGE_TSS_gene_aggregation_all_chr}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")


cage_tss_gene_path <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz"
sample_root_all_chr <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample_all_chr <- "AL10_bc2178_19130"
window_bp_all_chr <- 2000L

read_nuc_bed12 <- function(path) {
  fread(path, header = FALSE,
        col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                      "thickStart","thickEnd","itemRgb",
                      "blockCount","blockSizes","blockStarts"))
}

load_all_chr_nucleosome_midpoints <- function(bed_files) {
  nucs_all <- rbindlist(lapply(bed_files, function(path) {
    calculate_nucleosome_midpoints(read_nuc_bed12(path))
  }), use.names = TRUE, fill = TRUE)
  nucs_all[order(chrom, nuc_mid)]
}

granges_to_dt <- function(gr) {
  data.table(
    chrom = as.character(seqnames(gr)),
    start = start(gr) - 1L,
    end = end(gr),
    strand = as.character(strand(gr)),
    gene_id = as.character(mcols(gr)$gene_id),
    score = as.character(mcols(gr)$score)
  )
}

find_nucs_near_tss_with_gene <- function(nuc_gr, tss_gr, window_bp = 2000L) {
  tss_windows <- promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)
  hits <- findOverlaps(nuc_gr, tss_windows)
  if (!length(hits)) {
    return(data.table(
      gene_id = character(),
      gene_id_base = character(),
      seqnames = character(),
      strand = character(),
      tss_index = integer(),
      nuc_pos = integer(),
      tss_pos = integer(),
      rel_pos = integer()
    ))
  }

  query_idx <- queryHits(hits)
  subject_idx <- subjectHits(hits)
  nuc_pos <- start(nuc_gr)[query_idx]
  tss_pos <- start(tss_gr)[subject_idx]
  tss_str <- as.character(strand(tss_gr))[subject_idx]
  gene_id <- as.character(mcols(tss_gr)$gene_id[subject_idx])

  data.table(
    gene_id = gene_id,
    gene_id_base = sub("\\..*$", "", gene_id),
    seqnames = as.character(seqnames(tss_gr))[subject_idx],
    strand = tss_str,
    tss_index = subject_idx,
    nuc_pos = nuc_pos,
    tss_pos = tss_pos,
    rel_pos = ifelse(tss_str == "+", nuc_pos - tss_pos, tss_pos - nuc_pos)
  )
}

summarize_gene_tss_distance <- function(hit_dt) {
  if (!nrow(hit_dt)) {
    return(data.table(
      gene_id = character(),
      gene_id_base = character(),
      mean_rel_pos_bp = numeric(),
      median_rel_pos_bp = numeric(),
      mean_abs_rel_pos_bp = numeric(),
      median_abs_rel_pos_bp = numeric(),
      sd_rel_pos_bp = numeric(),
      n_midpoints = integer(),
      n_tss = integer(),
      n_chromosomes = integer()
    ))
  }

  hit_dt[, .(
    mean_rel_pos_bp = as.numeric(mean(rel_pos)),
    median_rel_pos_bp = as.numeric(median(rel_pos)),
    mean_abs_rel_pos_bp = as.numeric(mean(abs(rel_pos))),
    median_abs_rel_pos_bp = as.numeric(median(abs(rel_pos))),
    sd_rel_pos_bp = as.numeric(stats::sd(rel_pos)),
    n_midpoints = .N,
    n_tss = uniqueN(tss_index),
    n_chromosomes = uniqueN(seqnames)
  ), by = .(gene_id, gene_id_base)][order(mean_abs_rel_pos_bp, -n_midpoints, gene_id)]
}

selected_sample_all_chr <- file.path(sample_root_all_chr, default_sample_all_chr)
nuc_by_chr_dir_all_chr <- file.path(selected_sample_all_chr, "extracted_results/nuc_by_chr")
bed_files_all_chr <- sort(list.files(
  nuc_by_chr_dir_all_chr,
  pattern = "ft_extracted_nuc\\.chr.*\\.bed\\.gz$",
  full.names = TRUE
))
if (!length(bed_files_all_chr)) {
  stop("No chromosome-level nucleosome BED files found in ", nuc_by_chr_dir_all_chr)
}

tss_gr_all_chr <- load_tss_annotations_bed9_with_gene_id(cage_tss_gene_path)
tss_gr_all_chr <- tss_gr_all_chr[
  !is.na(mcols(tss_gr_all_chr)$gene_id) &
    mcols(tss_gr_all_chr)$gene_id != "NA"
]

nucs_all_chr <- load_all_chr_nucleosome_midpoints(bed_files_all_chr)
nuc_gr_all_chr <- nuc_midpoints_to_granges(nucs_all_chr)
cage_gene_hits_all_chr <- find_nucs_near_tss_with_gene(
  nuc_gr_all_chr,
  tss_gr_all_chr,
  window_bp = window_bp_all_chr
)
if (!nrow(cage_gene_hits_all_chr)) {
  stop("No nucleosome midpoints overlapped the CAGE TSS windows across all chromosomes.")
}

meta_raw_all_chr <- make_position_meta(
  nuc_gr_all_chr,
  tss_gr_all_chr,
  window_bp = window_bp_all_chr,
  bin_size = 10L,
  normalize_by_tss = FALSE,
  label = "CAGE TSS"
)
meta_norm_all_chr <- make_position_meta(
  nuc_gr_all_chr,
  tss_gr_all_chr,
  window_bp = window_bp_all_chr,
  bin_size = 10L,
  normalize_by_tss = TRUE,
  label = "CAGE TSS"
)

cage_gene_distance_summary <- summarize_gene_tss_distance(cage_gene_hits_all_chr)
print(head(cage_gene_distance_summary, 20))

meta_plot_all_chr <- ggplot(meta_raw_all_chr, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.6) +
  geom_smooth(method = "loess", span = 0.05, se = FALSE,
              color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_all_chr, window_bp_all_chr, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x = "Distance to TSS",
    y = "Nucleosome midpoint count",
    title = "Nucleosome midpoint distribution around CAGE TSS",
    subtitle = sprintf(
      "Sample: %s | all chromosomes | n nucleosomes = %s | n TSS = %s",
      default_sample_all_chr,
      format(length(nuc_gr_all_chr), big.mark = ","),
      format(length(tss_gr_all_chr), big.mark = ",")
    )
  ) +
  theme_bw(base_size = 12)
print(meta_plot_all_chr)

meta_plot_norm_all_chr <- ggplot(meta_norm_all_chr, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.7, color = "#d62728") +
  geom_smooth(method = "loess", span = 0.06, se = FALSE,
              linewidth = 1, color = "#d62728") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_all_chr, window_bp_all_chr, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x = "Distance to TSS",
    y = "Avg nucleosome midpoints per TSS (10 bp bins)",
    title = "CAGE TSS",
    subtitle = sprintf("Sample: %s | all chromosomes", default_sample_all_chr)
  ) +
  theme_bw(base_size = 12)
print(meta_plot_norm_all_chr)

gene_distance_plot <- ggplot(cage_gene_distance_summary, aes(x = mean_abs_rel_pos_bp)) +
  geom_histogram(binwidth = 25, fill = "#be1c54ff", color = "white") +
  labs(
    x = "Mean absolute distance to TSS per gene (bp)",
    y = "Number of genes",
    title = "Per-gene average nucleosome distance to CAGE TSS",
    subtitle = sprintf(
      "Sample: %s | all chromosomes | genes = %s",
      default_sample_all_chr,
      format(nrow(cage_gene_distance_summary), big.mark = ",")
    )
  ) +
  theme_bw(base_size = 12)
print(gene_distance_plot)

out_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(
  data.table(
    sample = default_sample_all_chr,
    chrom_bed = basename(bed_files_all_chr),
    chrom_bed_path = bed_files_all_chr
  ),
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_input_bed_files_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

fwrite(
  nucs_all_chr,
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_nucleosome_midpoints_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

fwrite(
  granges_to_dt(tss_gr_all_chr),
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_annotations_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

fwrite(
  cage_gene_hits_all_chr,
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_nucleosome_tss_hits_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

fwrite(
  meta_raw_all_chr,
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_nucleosome_midpoints_raw_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

fwrite(
  meta_norm_all_chr,
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_nucleosome_midpoints_perTSS_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

fwrite(
  cage_gene_distance_summary,
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_gene_mean_distance_all_chr.tsv",
                    default_sample_all_chr)),
  sep = "\t"
)

ggsave(
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_nucleosome_midpoints_raw_all_chr.pdf",
                    default_sample_all_chr)),
  plot = meta_plot_all_chr,
  width = 6,
  height = 4
)

ggsave(
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_nucleosome_midpoints_perTSS_all_chr.pdf",
                    default_sample_all_chr)),
  plot = meta_plot_norm_all_chr,
  width = 6,
  height = 4
)

ggsave(
  file.path(out_dir,
            sprintf("%s_CAGE_TSS_gene_mean_abs_distance_all_chr.pdf",
                    default_sample_all_chr)),
  plot = gene_distance_plot,
  width = 7,
  height = 4.5
)

```

# Per-read aware CAGE TSS gene aggregation (all chromosomes)

```{r CAGE_TSS_gene_aggregation_per_read_all_chr}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

cage_tss_gene_path <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz"
sample_root_per_read <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample_per_read <- "AL10_bc2178_19130"
window_bp_per_read <- 2000L
bin_size_per_read <- 10L

# Load all chromosome BED12 files
selected_sample_per_read <- file.path(sample_root_per_read, default_sample_per_read)
nuc_by_chr_dir_per_read <- file.path(selected_sample_per_read, "extracted_results/nuc_by_chr")
bed_files_per_read <- sort(list.files(
  nuc_by_chr_dir_per_read,
  pattern = "ft_extracted_nuc\\.chr.*\\.bed\\.gz$",
  full.names = TRUE
))
if (!length(bed_files_per_read)) {
  stop("No chromosome-level nucleosome BED files found in ", nuc_by_chr_dir_per_read)
}

# Load TSS with gene information
tss_gr_per_read <- load_tss_annotations_bed9_with_gene_id(cage_tss_gene_path)
tss_gr_per_read <- tss_gr_per_read[
  !is.na(mcols(tss_gr_per_read)$gene_id) &
    mcols(tss_gr_per_read)$gene_id != "NA"
]

# ── Per-read raw meta-profile (all chromosomes) ──
all_bed12_per_read <- rbindlist(lapply(bed_files_per_read, function(path) {
  fread(path, header = FALSE,
        col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                      "thickStart","thickEnd","itemRgb","blockCount",
                      "blockSizes","blockStarts"))
}))

meta_per_read_raw_all_chr <- make_meta_per_read_raw(
  all_bed12_per_read, tss_gr_per_read,
  window_bp = window_bp_per_read, bin_size = bin_size_per_read,
  label = "CAGE TSS"
)

meta_plot_per_read_raw_all_chr <- ggplot(meta_per_read_raw_all_chr, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.6, color = "#2ca02c") +
  geom_smooth(method = "loess", span = 0.05, se = FALSE, color = "#2ca02c", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_per_read, window_bp_per_read, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x = "Distance to TSS",
    y = "Nucleosome count",
    title = "Nucleosome distribution around CAGE TSS (per-read method - raw)",
    subtitle = sprintf("Sample: %s | all chromosomes | Each read-TSS pair counted",
                       default_sample_per_read)
  ) +
  theme_bw(base_size = 12)

print(meta_plot_per_read_raw_all_chr)

# ── Per-read normalized meta-profile (all chromosomes) ──
meta_per_read_norm_all_chr <- make_meta_per_read_normalized(
  all_bed12_per_read, tss_gr_per_read,
  window_bp = window_bp_per_read, bin_size = bin_size_per_read,
  label = "CAGE TSS"
)

meta_plot_per_read_norm_all_chr <- ggplot(meta_per_read_norm_all_chr, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.7, color = "#d62728") +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, color = "#d62728", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_per_read, window_bp_per_read, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x = "Distance to TSS",
    y = "Avg nucleosomes per read-TSS pair",
    title = "Nucleosome distribution around CAGE TSS (per-read method - normalized)",
    subtitle = sprintf("Sample: %s | all chromosomes | Normalized by read-TSS pairs",
                       default_sample_per_read)
  ) +
  theme_bw(base_size = 12)

print(meta_plot_per_read_norm_all_chr)

# ── Per-read gene-level analysis ──
rtp_nuc_with_gene <- find_nucs_near_tss_with_gene_per_read(
  all_bed12_per_read, tss_gr_per_read, 
  window_bp = window_bp_per_read
)

gene_summary_per_read <- summarize_gene_tss_per_read(rtp_nuc_with_gene)

print(paste("Top 20 genes by mean absolute distance:"))
print(head(gene_summary_per_read, 20))

# Visualization: gene-level distances per read
gene_distance_plot_per_read <- ggplot(gene_summary_per_read, aes(x = mean_abs_rel_pos_bp)) +
  geom_histogram(binwidth = 25, fill = "#be1c54ff", color = "white") +
  labs(
    x = "Mean absolute distance to TSS per gene (bp)",
    y = "Number of genes",
    title = "Per-gene average nucleosome distance to CAGE TSS (per-read method)",
    subtitle = sprintf(
      "Sample: %s | all chromosomes | genes = %s | Multiple reads per gene tracked",
      default_sample_per_read,
      format(nrow(gene_summary_per_read), big.mark = ",")
    )
  ) +
  theme_bw(base_size = 12)

print(gene_distance_plot_per_read)

# Scatter: number of reads vs. distance for each gene
gene_reads_scatter <- ggplot(gene_summary_per_read, aes(x = n_reads, y = mean_abs_rel_pos_bp, 
                                                         size = n_nuc, color = n_tss)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(name = "n nucleosomes") +
  scale_color_continuous(name = "n TSS", type = "viridis") +
  labs(
    x = "Number of reads covering gene TSS",
    y = "Mean absolute distance to TSS (bp)",
    title = "Per-gene nucleosome positioning by read coverage",
    subtitle = "Size = total nucleosomes; Color = number of TSS"
  ) +
  theme_bw(base_size = 12)

print(gene_reads_scatter)

# Save outputs
out_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

fwrite(meta_per_read_raw_all_chr,
       file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_raw_all_chr.tsv",
                         default_sample_per_read)),
       sep = "\t")

fwrite(meta_per_read_norm_all_chr,
       file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_normalized_all_chr.tsv",
                         default_sample_per_read)),
       sep = "\t")

fwrite(rtp_nuc_with_gene,
       file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_nuc_gene_hits_all_chr.tsv",
                         default_sample_per_read)),
       sep = "\t")

fwrite(gene_summary_per_read,
       file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_gene_summary_all_chr.tsv",
                         default_sample_per_read)),
       sep = "\t")

ggsave(file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_raw_all_chr.pdf",
                         default_sample_per_read)),
       plot = meta_plot_per_read_raw_all_chr, width = 7, height = 5)

ggsave(file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_normalized_all_chr.pdf",
                         default_sample_per_read)),
       plot = meta_plot_per_read_norm_all_chr, width = 7, height = 5)

ggsave(file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_gene_distance_all_chr.pdf",
                         default_sample_per_read)),
       plot = gene_distance_plot_per_read, width = 7, height = 5)

ggsave(file.path(out_dir,
                 sprintf("%s_CAGE_TSS_per_read_reads_vs_distance_all_chr.pdf",
                         default_sample_per_read)),
       plot = gene_reads_scatter, width = 8, height = 6)

```

Filter for nucleosomes that are only defined in fire-seq to have size of 140-150 bp, and compare with the global distribution of nucleosome sizes

```{r nuc_size_dist_filtered}
sample_root   <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
size_range    <- c(140L, 150L)          # midpoint-derived size filter
target_sample <- "AL10_bc2178_19130"
target_chrom  <- "chr1"
window_bp     <- 2000L                  # ± around TSS

# locate chromosome-specific nucleosome files
nuc_by_chr_dir <- file.path(sample_root, target_sample, "extracted_results/nuc_by_chr")
chr_pat        <- sprintf("ft_extracted_nuc\\.%s\\.bed\\.gz$", target_chrom)
bed_files      <- list.files(nuc_by_chr_dir, pattern = chr_pat, full.names = TRUE)
if (!length(bed_files)) stop("No nucleosome BED files found for ", target_chrom)

# compute midpoint + size per nucleosome, then filter by size_range
nuc_list <- lapply(bed_files, function(bf) {
  bed12 <- fread(bf, header = FALSE,
                 col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                               "thickStart","thickEnd","itemRgb","blockCount",
                               "blockSizes","blockStarts"))
  calculate_nucleosome_midpoints(bed12)
})
nucs   <- rbindlist(nuc_list)
nucs   <- nucs[nuc_size >= size_range[1] & nuc_size <= size_range[2]]
nuc_gr <- nuc_midpoints_to_granges(nucs)

# TSS sets
tss_gencode <- load_tss_annotations("/project/spott/cshan/annotations/TSS.gencode.v49.bed")
tss_cage    <- load_tss_annotations_bed9("/project/spott/cshan/annotations/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz")

# raw counts per bin
make_meta <- function(tss_gr, label) {
  plot_dt <- find_nucs_near_tss(nuc_gr, tss_gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / 10L) * 10L]   # 10 bp bins
  meta <- plot_dt[, .N, by = bin][order(bin)]
  meta[, source := label]
  meta
}
meta_all <- rbindlist(list(
  make_meta(tss_gencode, "GENCODE TSS"),
  make_meta(tss_cage,    "CAGE TSS")
))

meta_plot <- ggplot(meta_all, aes(x = bin, y = N, color = source)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  scale_color_manual(values = c("GENCODE TSS" = "#1f77b4", "CAGE TSS" = "#d62728")) +
  labs(x = "Distance to TSS",
       y = sprintf("Nucleosome midpoints (%d–%d bp)", size_range[1], size_range[2]),
       subtitle = sprintf("Sample: %s | Chr: %s", target_sample, target_chrom),
       color = "TSS source") +
  theme_bw(base_size = 12)
print(meta_plot)

# per‑TSS normalized overlay
meta_all_norm <- rbindlist(list(
  make_meta_normalized_TSS(nuc_gr, tss_gencode, window_bp = window_bp, bin_size = 10L, label = "GENCODE TSS"),
  make_meta_normalized_TSS(nuc_gr, tss_cage,    window_bp = window_bp, bin_size = 10L, label = "CAGE TSS")
))

meta_plot_norm <- ggplot(meta_all_norm, aes(x = bin, y = N, color = source)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  scale_color_manual(values = c("GENCODE TSS" = "#1f77b4", "CAGE TSS" = "#d62728")) +
  labs(x = "Distance to TSS",
       y = sprintf("Avg nucleosome midpoints per TSS", size_range[1], size_range[2]),
       subtitle = sprintf("Sample: %s | Chr: %s", target_sample, target_chrom),
       color = "TSS source") +
  theme_bw(base_size = 12)
print(meta_plot_norm)

# save PDFs
out_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
ggsave(file.path(out_dir,
                 sprintf("%s_%s_TSS_midpoints_%d_%d_bp.pdf",
                         target_sample, target_chrom,
                         size_range[1], size_range[2])),
       plot = meta_plot, width = 6, height = 4)

ggsave(file.path(out_dir,
                 sprintf("%s_%s_TSS_midpoints_%d_%d_bp_perTSS.pdf",
                         target_sample, target_chrom,
                         size_range[1], size_range[2])),
       plot = meta_plot_norm, width = 6, height = 4)

```


Based on gene expression profile, select highly expressed genes and lowly expressed genes, and plot the nucleosome distribution around TSS 
  - also compare between CAGE-defined TSS and GENCODE-defined TSS for each expression group
  - include both raw counts and per-TSS normalized profiles and use consistent colors across all plots:
    -  CAGE-Seq = #be1c54ff, GENCODE = #1f77b4
    - highly expressed = #FF7F0E, lowly expressed = #2CA02C
  - expression data is /project/spott/cshan/annotations/ENCFF873VWU.gene_quan.tsv
    - highly expressed genes: top 25% by TPM, lowly expressed genes: bottom 25% by TPM


```{bash, eval = FALSE}

# load/activate tools if not already
module load python/miniforge-25.3.0
source activate /project/spott/cshan/envs/bedtools

CAGE=/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz
TSS=/project/spott/cshan/annotations/TSS.gencode.v49.bed
OUT=/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz



# strand-aware nearest GENCODE TSS, preserve BED9 cols, replace col4 with gene_id
/project/spott/cshan/envs/bedtools/bin/bedtools closest -s -t first -d \
  -a <(zcat "$CAGE" | sort -k1,1 -k2,2n) \
  -b <(sort -k1,1 -k2,2n "$TSS") \
| awk 'BEGIN{OFS="\t"}{gene=$13; if(gene==""||gene==".") gene="NA"; print $1,$2,$3,gene,$5,$6,$7,$8,$9}' \
| /project/spott/cshan/envs/bedtools/bin/bgzip > "$OUT"

# index
/project/spott/cshan/envs/bedtools/bin/tabix -p bed "$OUT"

echo "Annotated file written to: $OUT"


```



```{r expr_stratified_TSS_profiles}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

gencode_tss_path <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"
cage_tss_path    <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz"
sample_root      <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
expr_path        <- "/project/spott/cshan/annotations/ENCFF873VWU.gene_quan.tsv"
default_sample   <- "AL10_bc2178_19130"
default_chrom    <- "chr1"
window_bp        <- 2000L

read_bed12 <- function(path) fread(path, header = FALSE,
  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                "thickStart","thickEnd","itemRgb","blockCount",
                "blockSizes","blockStarts"))

sample_dir <- file.path(sample_root, default_sample)
trees <- list.files(file.path(sample_dir, "extracted_results/nuc_by_chr"),
                    pattern = sprintf("ft_extracted_nuc\\.%s\\.bed\\.gz$", default_chrom),
                    full.names = TRUE)
bed12 <- read_bed12(trees[1])
nucs <- calculate_nucleosome_midpoints(bed12)
nuc_gr <- nuc_midpoints_to_granges(nucs)

expr <- fread(expr_path)
expr <- expr[grepl("^ENSG", gene_id)]
qs <- quantile(expr$TPM, probs = c(0.25, 0.75), na.rm = TRUE)
expr[, expr_group := fcase(
  TPM >= qs[2], "high",
  TPM <= qs[1], "low",
  default = NA_character_
)]
expr <- expr[!is.na(expr_group)]

load_and_annotate <- function(path) {
  dt <- fread(path, header = FALSE,
              col.names = c("chrom","start","end","gene_id","score","strand"))
  gr <- load_tss_annotations(path)
  mcols(gr)$gene_id <- dt$gene_id
  gr
}

tss_sets <- list(
  GENCODE = load_and_annotate(gencode_tss_path),
  CAGE    = load_tss_annotations_bed9_with_gene_id(cage_tss_path)
)

align_tss <- function(gr, source) {
  gene_ids <- mcols(gr)$gene_id
  idx <- match(gene_ids, expr$gene_id)
  valid <- !is.na(idx)
  if (!any(valid)) return(NULL)
  gr2 <- gr[valid]
  mcols(gr2)$expr_group <- expr$expr_group[idx[valid]]
  list(source = source, gr = gr2)
}

tss_by_group <- Filter(Negate(is.null), lapply(names(tss_sets), function(source) align_tss(tss_sets[[source]], source)))

build_meta_counts <- function(gr, source, expr_group) {
  plot_dt <- find_nucs_near_tss(nuc_gr, gr, window_bp = window_bp)
  plot_dt[, bin := round(rel_pos / 10L) * 10L]
  meta <- plot_dt[, .N, by = bin][order(bin)]
  meta[, N := N / length(gr)]
  meta[, `:=`(source = source, expr_group = expr_group)]
  meta
}

meta_raw_list <- list()
meta_norm_list <- list()
for (entry in tss_by_group) {
  for (grp in unique(mcols(entry$gr)$expr_group)) {
    subset <- entry$gr[mcols(entry$gr)$expr_group == grp]
    meta_raw_list[[length(meta_raw_list) + 1]] <- build_meta_counts(subset, entry$source, grp)
    meta_norm <- make_meta_normalized_TSS(nuc_gr, subset, window_bp = window_bp, bin_size = 10L)
    meta_norm[, `:=`(source = entry$source, expr_group = grp)]
    meta_norm_list[[length(meta_norm_list) + 1]] <- meta_norm
  }
}

meta_raw <- rbindlist(meta_raw_list, use.names = TRUE, fill = TRUE)
meta_norm <- rbindlist(meta_norm_list, use.names = TRUE, fill = TRUE)

raw_plot <- ggplot(meta_raw, aes(x = bin, y = N, color = source)) +
  geom_line(linewidth = 0.75) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 1) +
  facet_wrap(~expr_group, ncol = 1) +
  scale_color_manual(values = c("GENCODE" = "#1f77b4", "CAGE" = "#be1c54ff")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Avg midpoints per TSS (10 bp bins)",
       color = "TSS source",
       title = "Per-TSS normalized nucleosome midpoints by expression group")
print(raw_plot)

normalized_plot <- ggplot(meta_norm, aes(x = bin, y = N, color = source)) +
  geom_line(linewidth = 0.75) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 1) +
  facet_wrap(~expr_group, ncol = 1) +
  scale_color_manual(values = c("GENCODE" = "#1f77b4", "CAGE" = "#be1c54ff")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Avg midpoints per TSS (10 bp bins)",
       color = "TSS source",
       title = "Per-TSS normalized nucleosome midpoints")
print(normalized_plot)

expr_plot_raw <- ggplot(meta_raw, aes(x = bin, y = N, color = expr_group)) +
  geom_line(linewidth = 0.75) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 1) +
  facet_wrap(~source, ncol = 1) +
  scale_color_manual(values = c("high" = "#FF7F0E", "low" = "#2CA02C")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Avg midpoints per TSS (10 bp bins)",
       color = "Expression bin",
       title = "Expression-stratified per-TSS normalized counts")
print(expr_plot_raw)

expr_plot_norm <- ggplot(meta_norm, aes(x = bin, y = N, color = expr_group)) +
  geom_line(linewidth = 0.75) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 1) +
  facet_wrap(~source, ncol = 1) +
  scale_color_manual(values = c("high" = "#FF7F0E", "low" = "#2CA02C")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(breaks = seq(-window_bp, window_bp, 500),
                     labels = function(x) paste0(x/1000, " kb")) +
  labs(x = "Distance to TSS",
       y = "Avg midpoints per TSS (10 bp bins)",
       color = "Expression bin",
       title = "Expression-stratified per-TSS normalized counts")
print(expr_plot_norm)

out_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(out_dir, "raw_gencode_cage_expression_overlay.pdf"), plot = raw_plot, width = 8, height = 6)
ggsave(file.path(out_dir, "normalized_gencode_cage_expression_overlay.pdf"), plot = normalized_plot, width = 8, height = 6)
ggsave(file.path(out_dir, "expression_raw_by_source.pdf"), plot = expr_plot_raw, width = 8, height = 6)
ggsave(file.path(out_dir, "expression_norm_by_source.pdf"), plot = expr_plot_norm, width = 8, height = 6)
```


Plot metagene profiles for FIRE-seq coverage based on gene expression data
    - highly expressed genes should have higher coverage at the TSS and gene body compared to lowly expressed genes
    - /project/spott/cshan/fiber-seq/nucleosome/expression

```{r fire_seq_coverage_by_expression}
# load all required libraries
library(GenomicRanges)
library(rtracklayer)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(data.table)
library(Gviz)

# the goal is to define TSS sets stratified by expression, then plot FIRE-seq coverage around those TSS sets


expr_path <- "/project/spott/cshan/annotations/ENCFF873VWU.gene_quan.tsv"
expr <- fread(expr_path)
# grepl to filter for gene_id that starts with ENSG 
expr <- expr[grepl("^ENSG", gene_id)] 
# filter out genes with TPM = 0
expr <- expr[TPM > 0]

# rank genes into expression bins (e.g. quintiles) based on TPM values
expr[, TPM_bin := dplyr::ntile(TPM, 5)]
head(expr)

# merge expression groups with TSS annotations (using GENCODE TSS for this analysis)
gencode_tss_path <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"
tss <- fread(gencode_tss_path, header = FALSE,
col.names = c("chrom","start","end","gene_id","score","strand"))
tss_expr <- merge(tss, expr[, .(gene_id, TPM_bin)], by = "gene_id", all.x = TRUE)  
# remove NA values to keep only TSS with expression data
tss_expr <- tss_expr[!is.na(tss_expr$TPM_bin)]
str(tss_expr)

# convert each expression bin into a bed file for metagene plotting
for (bin in 1:5) {
  bin_tss <- tss_expr[TPM_bin == bin]
  bin_gr <- GRanges(seqnames = bin_tss$chrom,
  ranges = IRanges(start = bin_tss$start + 1L, end = bin_tss$end),
  strand = bin_tss$strand,
  gene_id = bin_tss$gene_id)
  # define export path for each bin and export as BED file
  export_path <- sprintf("/project/spott/cshan/fiber-seq/nucleosome/expression/TSS_bin_%d.bed", bin)
  export(bin_gr, con = export_path, format = "BED")
}
```

Plot the metagene plot for all the peaks
```{bash, eval = FALSE}
#!/bin/bash
#SBATCH --job-name=fire_seq_expr
#SBATCH --time=12:00:00
#SBATCH --mem=100G
#SBATCH --cpus-per-task=16
#SBATCH --account=pi-spott
#SBATCH --partition=caslake
#SBATCH --array=0-33
#SBATCH --output=/project/spott/cshan/fiber-seq/results/logs/fire_seq_expr_%A_%a.out

module load python/miniforge-25.3.0
source activate /project/spott/cshan/envs/deeptools

# roots and output dir
sample_root="/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/"
outdir="/project/spott/cshan/fiber-seq/nucleosome/expression"
mkdir -p "${outdir}"

# pick sample dir from array index
mapfile -t sample_dirs < <(find "${sample_root}" -mindepth 1 -maxdepth 1 -type d | sort)
if (( SLURM_ARRAY_TASK_ID >= ${#sample_dirs[@]} )); then
    echo "ERROR: SLURM_ARRAY_TASK_ID ${SLURM_ARRAY_TASK_ID} exceeds available samples (${#sample_dirs[@]})"
    exit 1
fi
sample_dir="${sample_dirs[$SLURM_ARRAY_TASK_ID]}"

# BigWig lives directly under sample_dir/trackHub-v0.1/bw
bw_dir="${sample_dir}/trackHub-v0.1/bw"
bw="${bw_dir}/all.fire.coverage.bw"

# outputs per sample
sample="$(basename "${sample_dir}")"
matrix="${outdir}/${sample}_fire_expr_matrix.gz"
plot="${outdir}/${sample}_fire_expr_profile.pdf"

echo "========================================="
echo "Processing sample: ${sample}"
echo "Sample directory: ${sample_dir}"
echo "BigWig directory: ${bw_dir}"
echo "BigWig file: ${bw}"
echo "Matrix: ${matrix}"
echo "Plot: ${plot}"
echo "========================================="

# verify BigWig
if [ ! -f "${bw}" ]; then
    echo "ERROR: BigWig file does not exist: ${bw}"
    exit 1
fi

computeMatrix scale-regions \
  -S "$bw" \
  -R \
    /project/spott/cshan/fiber-seq/nucleosome/expression/TSS_bin_1.bed \
    /project/spott/cshan/fiber-seq/nucleosome/expression/TSS_bin_2.bed \
    /project/spott/cshan/fiber-seq/nucleosome/expression/TSS_bin_3.bed \
    /project/spott/cshan/fiber-seq/nucleosome/expression/TSS_bin_4.bed \
    /project/spott/cshan/fiber-seq/nucleosome/expression/TSS_bin_5.bed \
  --beforeRegionStartLength 3000 \
  --regionBodyLength 5000 \
  --afterRegionStartLength 3000 \
  --skipZeros \
  --binSize 50 \
  --sortRegions keep \
  --samplesLabel "0-20%" "20-40%" "40-60%" "60-80%" "80-100%" \
  --outFileName "$matrix"

plotProfile \
  -m "$matrix" \
  -o "$plot" \
  --perGroup \
  --refPointLabel "TSS" \
  --colors "#999999,#7FC97F,#FDB462,#A6CEE3,#E31A1C"


echo "========================================="
echo "Finished processing sample: ${sample}"
echo "Matrix file: ${matrix}"
echo "Plot file: ${plot}"
echo "========================================="


```


Replicate the above analysis for nucleosome occpancy around TSS, stratified by expression level



Boxplots showing the distribution of nucleosome offsets for the first five nucleosomes upstream and downstream TSSs with a FIRE peak. 
  - Peak-adjacent nucleosomes are colored purple, while all others are shown in gray. 
  - The p values were computed across all distributions comparing the most proximal nucleosomes to all others using Wilcoxon rank sum (p < 0.0001)

- FIRE is scored 0–255 on MSPs; per FIRE docs, scores >230 are called FIRE elements (high-confidence regulatory hotspots)
- Overlapping those FIRE MSPs with TSS (here CAGE TSS) restricts the analysis to promoters that actually contain a FIRE element, matching the biological question: nucleosome positioning around FIRE‑positive promoters, not all promoters.
- `reduce()` merges overlapping MSP segments so each FIRE locus is treated once; `subsetByOverlaps()` then selects only TSS that intersect these high-confidence FIRE regions.


```{r fire_peak_nuc_offsets}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

gencode_tss_path <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"
cage_tss_path    <- "/project/spott/cshan/annotations/ENCFF715BEB_CAGE_withGenes.bed"
sample_root      <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample   <- "AL10_bc2178_19130"
default_chrom    <- "chr1"
window_bp        <- 2000L

# Directories
fire_peak_dir <- file.path(sample_root, default_sample, "extracted_results/all_by_chr")
sample_dir    <- file.path(sample_root, default_sample)
trees <- list.files(file.path(sample_dir, "extracted_results/nuc_by_chr"),
                    pattern = sprintf("ft_extracted_nuc\\.%s\\.bed\\.gz$", default_chrom),
                    full.names = TRUE)

# bed12 -> nucleosome midpoint GRanges
load_nucleosome_annotations <- function(bed_files) {
  beds <- lapply(bed_files, fread, header = FALSE,
  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
  "thickStart","thickEnd","itemRgb","blockCount",
  "blockSizes","blockStarts"))
  mids <- rbindlist(lapply(beds, calculate_nucleosome_midpoints))
  nuc_midpoints_to_granges(mids)
}


find_nucs_near_tss <- function(nuc_gr, tss_gr, window_bp = 2000L) {
  tss_windows <- promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)
  hits <- findOverlaps(nuc_gr, tss_windows)
  nuc_pos  <- start(nuc_gr)[queryHits(hits)]
  tss_pos  <- start(tss_gr)[subjectHits(hits)]
  tss_str  <- as.character(strand(tss_gr))[subjectHits(hits)]
  rel_pos <- ifelse(tss_str == "+", nuc_pos - tss_pos, tss_pos - nuc_pos)
  data.table(nuc_pos = nuc_pos, tss_pos = tss_pos, rel_pos = rel_pos)
}

# function to load TSS annotations from BED file and convert to GRanges
load_tss_annotations <- function(tss_path) {
    tss <- fread(tss_path, header = FALSE,
                 col.names = c("chrom","start","end","gene_id","score","strand"))
    GRanges(seqnames = tss$chrom,
            ranges   = IRanges(tss$start + 1L, tss$end),
            strand   = tss$strand,
            gene_id  = tss$gene_id)
}

# Build FIRE-peak TSS set (CAGE TSS overlapped by MSPs with FIRE score > 230)

# 1) load TSS (CAGE or GENCODE)
tss_gr <- load_tss_annotations(cage_tss_path)

# 2) load FIRE MSPs for the default chromosome
## the selected columns are: chrom, start, end, strand, fire score, ref_msp_starts, ref_msp_lengths

all_file <- file.path(fire_peak_dir,
  sprintf("%s.ft_extracted_all.%s.bed.gz",
  default_sample, default_chrom))


fire_cols <- fread(all_file, header = FALSE, select = c(1,2,3,6,25,26,27),
                     col.names = c("chrom","start","end","strand",
                                   "fire","ref_msp_starts","ref_msp_lengths"))

# expand MSPs with FIRE > 230 into GRanges
## for each row(individual fiber), split the comma-separated columns into vectors, filter for FIRE > 230, and create GRanges for those MSPs
make_fire_gr <- function(row) {
    # split on commons into a vector of starts
    starts  <- as.integer(strsplit(row[["ref_msp_starts"]], ",")[[1]])
    # lengths of the MSPs
    lens    <- as.integer(strsplit(row[["ref_msp_lengths"]], ",")[[1]])
    # scores of the FIRE signal for each MSP
    scores  <- as.integer(strsplit(row[["fire"]], ",")[[1]])
    # only keep MSPs with score > 230
    keep    <- !is.na(starts) & starts >= 0 & scores > 230
    if (!any(keep)) return(NULL)
    # create GRanges for the kept MSPs
    GRanges(seqnames = row[["chrom"]],
            ranges   = IRanges(starts[keep] + 1L, starts[keep] + lens[keep]),
            strand   = row[["strand"]],
            fire     = scores[keep])
  }

# split each row of fire_cols into its own group split(fire_cols, seq_len(nrow(fire_cols))), each element is a single row data table
# then apply make_fire_gr to each row to get a GRanges (or NULL), keep only the GRanges and combine them
fire_gr_list <- lapply(split(fire_cols, seq_len(nrow(fire_cols))), make_fire_gr)
fire_gr_list <- Filter(Negate(is.null), fire_gr_list)
if (!length(fire_gr_list)) stop("No FIRE peaks survived the filtering step.")
fire_peaks <- do.call(c, fire_gr_list)

# save the fire_peaks GRanges (takes long time to compute, so save it for reuse)
save_dir <- "/project/spott/cshan/fiber-seq/results/FIRE_peaks_GRanges"
dir.create(save_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(fire_peaks, file.path(save_dir, sprintf("%s.%s_FIRE_peaks.rds", default_sample, default_chrom)))

fire_peaks <- reduce(fire_peaks)  # merge overlaps
# 3) TSS that overlap FIRE peaks
fire_tss <- subsetByOverlaps(tss_gr, fire_peaks)


# Load nucleosomes and compute offsets
nuc_gr  <- load_nucleosome_annotations(trees)
fire_dt <- find_nucs_near_tss(nuc_gr, fire_tss, window_bp = window_bp)

fire_dt[, direction := fifelse(rel_pos < 0, "upstream", "downstream")]
fire_dt[, tss_id := paste0(seqnames, ":", tss_pos, ":", strand)]
setorder(fire_dt, tss_id, direction, abs(rel_pos))
fire_dt[, rank := seq_len(.N), by = .(tss_id, direction)]
fire_dt <- fire_dt[rank <= 5]
fire_dt[, peak_adjacent := rank == 1]


offset_plot <- ggplot(fire_dt, aes(x = factor(rank), y = rel_pos, fill = peak_adjacent)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~direction, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c("TRUE" = "#6a51a3", "FALSE" = "gray70"),
                    labels = c("Peak-adjacent", "Other")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(x = "Nucleosome rank (closest = 1)",
       y = "Offset from TSS (bp)",
       fill = NULL,
       title = "Offsets for the first 5 nucleosomes around FIRE TSS",
       subtitle = sprintf("Purple = closest nucleosome per TSS; gray = next four\nWilcoxon rank-sum: %s", p_lab)) +
  theme_bw(base_size = 12)

print(offset_plot)
ggsave(file.path("/project/spott/cshan/fiber-seq/results/plots/nuc",
                 "fire_TSS_nuc_offsets_boxplot.pdf"),
       plot = offset_plot, width = 7, height = 5)
```



# Same analysis as above, but skips FIRE peak filtering and just looks at the first five nucleosomes upstream and downstream of all CAGE TSS

How are nucleosomes positioned relative to CAGE defined TSS, specifically looking at the first five nucleosomes upstream and downstream of each TSS, and comparing the offsets of the closest nucleosome to the next four nucleosomes 

This is testing the hypothesis that the nucleosome closest to the TSS (rank 1) is more tightly positioned (i.e. has smaller offsets) compared to the next four nucleosomes (ranks 2–5), which may be more variably positioned. 
  - Offsets = the distance between a nucleosome and a TSS, telling you how chromatin is organized around gene starts

THIS CODE IS NOT WORKING BECASUE IT IS <LOOKINING FOR EACH TSS, WHAT ARE THE NUCLEOSOMES WITHIN 2KB, THEN RANKING THOSE NUCLEOSOMES BY ABSOLUTE OFFSET, THEN PLOTTING THE OFFSETS OF RANK 1 VS RANKS 2-5. >
BUT WE WANT *FOR EACH READ THAT OVERLAPS A TSS, WHERE ARE THE NUCLEOSOMES ON THE SAME READ*
```{r tss_nuc_offsets, eval=FALSE}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

sample_root    <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample <- "AL10_bc2178_19130"
default_chrom  <- "chr1"
window_bp      <- 2000L # remove restriction to FIRE peaks, so use a wider window to capture more nucleosomes around TSS
cage_tss_path  <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"

# Load single chromosome BED
nuc_by_chr_dir <- file.path(sample_root, default_sample, "extracted_results/nuc_by_chr")
pattern  <- sprintf("ft_extracted_nuc\\.%s\\.bed\\.gz$", default_chrom)
nuc_file <- list.files(nuc_by_chr_dir, pattern = pattern, full.names = TRUE)[1]
stopifnot(!is.na(nuc_file), file.exists(nuc_file))

bed12 <- fread(
  nuc_file,
  sep = "\t",
  header = FALSE,
  select = 1:12,
  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                "thickStart","thickEnd","itemRgb","blockCount","blockSizes","blockStarts")
)

nucs   <- calculate_nucleosome_midpoints(bed12)
nuc_gr <- nuc_midpoints_to_granges(nucs)
tss_gr <- load_tss_annotations_bed9(cage_tss_path)

# Offsets and ranks
tss_dt <- find_nucs_near_tss(nuc_gr, tss_gr, window_bp = window_bp)



tss_dt[, direction := fifelse(rel_pos < 0, "upstream", "downstream")]
tss_dt[, tss_id := paste0(seqnames, ":", tss_pos, ":", strand)]
setorder(tss_dt, tss_id, direction, abs(rel_pos))
tss_dt[, rank := seq_len(.N), by = .(tss_id, direction)]
tss_dt <- tss_dt[rank <= 5]
tss_dt[, peak_adjacent := rank == 1]
tss_dt[, signed_rank := ifelse(direction == "upstream", -rank, rank)]
tss_dt[, abs_offset := abs(rel_pos)]

# # Optional Wilcoxon p-value (commented out)
# tss_dt[, group := ifelse(rank == 1, "closest", "others")]
# tss_dt[, abs_offset := abs(rel_pos)]
# wilcox_res <- wilcox.test(abs_offset ~ group, data = tss_dt, exact = FALSE)
# p_lab <- if (wilcox_res$p.value < 1e-4) "p < 1e-4" else sprintf("p = %.3g", wilcox_res$p.value)

# Plot boxplot for offsets
offset_plot <- ggplot(tss_dt, aes(x = factor(signed_rank), y = abs_offset, fill = peak_adjacent)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = c("TRUE" = "#6a51a3", "FALSE" = "gray70"),
                    labels = c("Peak-adjacent (rank 1)", "Other (ranks 2–5)")) +
  labs(x = "Nucleosome rank (upstream = negative, downstream = positive)",
       y = "Nucleosome offset (bp)",
       fill = NULL,
       title = sprintf("Offsets for first 5 nucleosomes around TSS (%s)", default_chrom)) +
  theme_bw(base_size = 12)

print(offset_plot)
ggsave(file.path("/project/spott/cshan/fiber-seq/results/plots/nuc",
                 sprintf("TSS_nuc_offsets_boxplot_%s_%s_abs.pdf", default_sample, default_chrom)),
       plot = offset_plot, width = 7, height = 5)

# Plot histograms for nucleosome size distributions by direction
upstream_hist <- ggplot(tss_dt %>% filter(direction == "upstream", !is.na(nuc_size), nuc_size < 1000), 
                        aes(x = nuc_size)) +
  geom_histogram(bins = 20, fill = "lightblue", color = "black", alpha = 0.9) +
  labs(title = "Upstream nucleosome size distribution",
       x = "Nucleosome size (bp)",
       y = "Frequency") +
  theme_minimal()

```

# FANTOM5 CAGE-seq expression

TSS split into bins by CAGE score, then plot nucleosome midpoints around those TSS bins

```{bash, eval = FALSE}
mkdir -p /project/spott/cshan/annotations/fantom5 && cd /project/spott/cshan/annotations && \
wget -O fantom5.hg38.CAGE_peaks.bed.gz 'https://fantom.gsc.riken.jp/5/datafiles/reprocessed/hg38_latest/extra/CAGE_peaks/hg38_fair+new_CAGE_peaks_phase1and2.bed.gz' && \
wget -O fantom5.hg38.LCL.rep1.CTSS.bed.gz 'https://fantom.gsc.riken.jp/5/datafiles/reprocessed/hg38_latest/basic/human.cell_line.hCAGE/B%2520lymphoblastoid%2520cell%2520line%253a%2520GM12878%2520ENCODE%252c%2520biol_rep1.CNhs12331.10821-111C2.hg38.nobarcode.ctss.bed.gz' && \
wget -O fantom5.hg38.LCL.rep2.CTSS.bed.gz 'https://fantom.gsc.riken.jp/5/datafiles/reprocessed/hg38_latest/basic/human.cell_line.hCAGE/B%2520lymphoblastoid%2520cell%2520line%253a%2520GM12878%2520ENCODE%252c%2520biol_rep2.CNhs12332.10822-111C3.hg38.nobarcode.ctss.bed.gz' && \
wget -O fantom5.hg38.LCL.rep3.CTSS.bed.gz 'https://fantom.gsc.riken.jp/5/datafiles/reprocessed/hg38_latest/basic/human.cell_line.hCAGE/B%2520lymphoblastoid%2520cell%2520line%253a%2520GM12878%2520ENCODE%252c%2520biol_rep3.CNhs12333.10823-111C4.hg38.nobarcode.ctss.bed.gz'
```

```{r fantom5_cage_bins}
library(data.table)
library(GenomicRanges)
library(ggplot2)
library(Rsamtools)

# -----------------------------
# Paths and parameters
# -----------------------------
annotation_dir <- "/project/spott/cshan/annotations"
fantom_dir <- file.path(annotation_dir, "fantom5")
dir.create(fantom_dir, recursive = TRUE, showWarnings = FALSE)

# Raw files can be either in annotation_dir or annotation_dir/fantom5
pick_existing_path <- function(fname) {
  candidates <- c(file.path(fantom_dir, fname), file.path(annotation_dir, fname))
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) stop(sprintf("Missing required file: %s", fname))
  hit
}

cageBedGzPath <- pick_existing_path("fantom5.hg38.CAGE_peaks.bed.gz")
lclRep1CtssBedGzPath <- pick_existing_path("fantom5.hg38.LCL.rep1.CTSS.bed.gz")
lclRep2CtssBedGzPath <- pick_existing_path("fantom5.hg38.LCL.rep2.CTSS.bed.gz")
lclRep3CtssBedGzPath <- pick_existing_path("fantom5.hg38.LCL.rep3.CTSS.bed.gz")

# Consensus output always goes here
lclCageBedGzPath <- file.path(fantom_dir, "fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz")

minReplicateSupport <- 2L
minTotalScore <- 30L
scoreToWrite <- "sum"  # "sum" or "max"
useStrandForExactMatch <- TRUE

sample_root <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample <- "AL10_bc2178_19130"
default_chrom <- "chr1"
window_bp <- 2000L
bin_size <- 20L
n_score_bins <- 4L

# Directories/inputs consistent with prior chunks
fire_peak_dir <- file.path(sample_root, default_sample, "extracted_results/all_by_chr")
sample_dir <- file.path(sample_root, default_sample)
trees <- list.files(
  file.path(sample_dir, "extracted_results/nuc_by_chr"),
  pattern = sprintf("ft_extracted_nuc\\.%s\\.bed\\.gz$", default_chrom),
  full.names = TRUE
)
if (length(trees) == 0L) {
  stop(sprintf("No nucleosome BED files found for sample=%s chrom=%s", default_sample, default_chrom))
}

# -----------------------------
# Helper functions
# -----------------------------
readCtssBed6 <- function(ctssBedGzPath, replicateLabel) {
  ctssTable <- fread(cmd = paste("zcat", shQuote(ctssBedGzPath)), header = FALSE)
  setnames(ctssTable, c("chrom", "start", "end", "name", "score", "strand"))
  ctssTable[, `:=`(
    start = as.integer(start),
    end = as.integer(end),
    score = as.integer(score),
    replicate = replicateLabel
  )]
  ctssTable[]
}

buildConsensusCtss <- function(rep1Table, rep2Table, rep3Table,
                               minReplicateSupport, minTotalScore, scoreToWrite) {
  combinedCtss <- rbindlist(list(rep1Table, rep2Table, rep3Table), use.names = TRUE)
  combinedCtss[, cagePeakId := paste0(chrom, ":", start, "..", end, ",", strand)]

  consensusTable <- combinedCtss[
    , .(
      scoreSum = sum(score),
      scoreMax = max(score),
      nRep = sum(score > 0L)
    ),
    by = .(chrom, start, end, strand, cagePeakId)
  ][nRep >= minReplicateSupport & scoreSum >= minTotalScore]

  consensusTable[, scoreToUse := if (scoreToWrite == "max") scoreMax else scoreSum]

  bed9Table <- consensusTable[
    , .(
      chrom = chrom,
      peakStart = start,
      peakEnd = end,
      cagePeakId = cagePeakId,
      score = as.integer(scoreToUse),
      strand = strand,
      tssStart = start,
      tssEnd = end,
      itemRgb = "0,0,0"
    )
  ]
  setorder(bed9Table, chrom, peakStart, peakEnd, strand)
  bed9Table[]
}

writeBgzipAndTabix <- function(bed9Table, outputBedGzPath) {
  tempBedPath <- sub("\\.gz$", "", outputBedGzPath)
  fwrite(bed9Table, file = tempBedPath, sep = "\t", col.names = FALSE)
  Rsamtools::bgzip(tempBedPath, dest = outputBedGzPath, overwrite = TRUE)
  Rsamtools::indexTabix(outputBedGzPath, format = "bed")
  unlink(tempBedPath)
  invisible(outputBedGzPath)
}

# -----------------------------
# 1) Read all FANTOM5 CAGE peaks
# -----------------------------
allCageTable <- fread(cmd = paste("zcat", shQuote(cageBedGzPath)), header = FALSE)
setnames(allCageTable, c("chrom","peakStart","peakEnd","cagePeakId","score","strand","tssStart","tssEnd","itemRgb"))

# -----------------------------
# 2) Generate consensus file first, then load it from disk
# -----------------------------
rep1Ctss <- readCtssBed6(lclRep1CtssBedGzPath, "rep1")
rep2Ctss <- readCtssBed6(lclRep2CtssBedGzPath, "rep2")
rep3Ctss <- readCtssBed6(lclRep3CtssBedGzPath, "rep3")

lclConsensusBed9_to_write <- buildConsensusCtss(
  rep1Table = rep1Ctss,
  rep2Table = rep2Ctss,
  rep3Table = rep3Ctss,
  minReplicateSupport = minReplicateSupport,
  minTotalScore = minTotalScore,
  scoreToWrite = scoreToWrite
)
writeBgzipAndTabix(lclConsensusBed9_to_write, lclCageBedGzPath)

# Load consensus from written file (explicitly)
lclConsensusBed9 <- fread(cmd = paste("zcat", shQuote(lclCageBedGzPath)), header = FALSE)
setnames(lclConsensusBed9, c("chrom","peakStart","peakEnd","cagePeakId","score","strand","tssStart","tssEnd","itemRgb"))

# -----------------------------
# 3) Exact coordinate overlap summary
# -----------------------------
lclTssTable <- copy(lclConsensusBed9)[, .(
  chrom, tssStart, tssEnd, strand,
  lclScore = as.numeric(score),
  lclId = cagePeakId
)]
allTssTable <- copy(allCageTable)[, .(
  chrom, tssStart, tssEnd, strand,
  allScore = as.numeric(score),
  allId = cagePeakId
)]

if (!useStrandForExactMatch) {
  lclTssTable[, strand := "*"]
  allTssTable[, strand := "*"]
}

setkey(lclTssTable, chrom, tssStart, tssEnd, strand)
setkey(allTssTable, chrom, tssStart, tssEnd, strand)

exactMatchTable <- allTssTable[lclTssTable, nomatch = 0L]
lclOnlyExactTable <- lclTssTable[!allTssTable, on = .(chrom, tssStart, tssEnd, strand)]
exactOverlapSummary <- list(
  nLclCtss = nrow(lclTssTable),
  nAllTss = nrow(allTssTable),
  nExactMatches = nrow(exactMatchTable),
  fracLclWithExactMatch = nrow(unique(exactMatchTable[, .(chrom, tssStart, tssEnd, strand)])) / nrow(lclTssTable),
  nLclOnlyExact = nrow(lclOnlyExactTable)
)

# -----------------------------
# 4) TSS bins by score (selected chrom) + nucleosome profile
# uses helper functions from your {r define-functions} chunk:
#   - load_nucleosome_annotations()
#   - make_meta_normalized_TSS()
# -----------------------------
tss_bins <- copy(lclConsensusBed9)[chrom == default_chrom]
tss_bins[, score := as.numeric(score)]
tss_bins <- tss_bins[is.finite(score) & !is.na(score)]
if (nrow(tss_bins) == 0L) {
  stop(sprintf("No consensus TSS found on selected chromosome: %s", default_chrom))
}

nuc_gr <- load_nucleosome_annotations(trees)

tss_bins[, score_rank := frank(score, ties.method = "average")]
tss_bins[, scoreBinInt := pmax(1L, pmin(
  n_score_bins,
  as.integer(ceiling((score_rank / max(score_rank)) * n_score_bins))
))]
tss_bins[, scoreBin := factor(paste0("Q", scoreBinInt), levels = paste0("Q", seq_len(n_score_bins)))]

meta_counts <- rbindlist(
  lapply(paste0("Q", seq_len(n_score_bins)), function(bin_label) {
    tss_sub <- tss_bins[scoreBin == bin_label]
    if (nrow(tss_sub) == 0L) return(NULL)

    tss_gr_sub <- GRanges(
      seqnames = tss_sub$chrom,
      ranges = IRanges(start = tss_sub$tssStart + 1L, end = tss_sub$tssEnd),
      strand = tss_sub$strand
    )

    prof <- make_meta_normalized_TSS(
      nuc_gr = nuc_gr,
      tss_gr = tss_gr_sub,
      window_bp = window_bp,
      bin_size = bin_size,
      label = bin_label
    )
    prof[, nTss := length(tss_gr_sub)]
    prof
  }),
  fill = TRUE
)

if (nrow(meta_counts) == 0L) stop("No meta-profile points generated.")
setnames(meta_counts, c("bin", "N", "source"), c("rel_bin", "midpointsPerTss", "scoreBin"))
meta_counts[, scoreBin := factor(scoreBin, levels = paste0("Q", seq_len(n_score_bins)))]
setorder(meta_counts, scoreBin, rel_bin)

plot_title <- sprintf(
  "Nucleosome midpoint profiles around LCL consensus CAGE TSS bins (%s, %s)",
  default_sample, default_chrom
)

fantom5_nuc_plot <- ggplot(meta_counts, aes(x = rel_bin, y = midpointsPerTss, color = scoreBin)) +
  geom_line(linewidth = 0.6, alpha = 0.95) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.4) +
  scale_x_continuous(
    breaks = seq(-window_bp, window_bp, by = 500L),
    labels = function(x) paste0(x / 1000, " kb")
  ) +
  labs(
    x = "Distance to TSS",
    y = "Nucleosome midpoints per TSS",
    color = "CAGE score bin",
    title = plot_title,
    subtitle = sprintf("Bins: Q1 (lowest) to Q%d (highest); window = +/- %d bp", n_score_bins, window_bp)
  ) +
  theme_bw(base_size = 11)

print(fantom5_nuc_plot)

plot_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc"
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
ggsave(
  filename = file.path(
    plot_dir,
    sprintf("fantom5_LCL_cage_score_bins_nucleosome_midpoints_%s_%s.pdf", default_sample, default_chrom)
  ),
  plot = fantom5_nuc_plot,
  width = 8.5, height = 5.5
)

list(
  consensusBedGz = lclCageBedGzPath,
  consensusTabix = paste0(lclCageBedGzPath, ".tbi"),
  overlapSummary = exactOverlapSummary,
  firePeakDir = fire_peak_dir,
  treesUsed = trees,
  nNucMidpoints = length(nuc_gr),
  nTssUsed = nrow(tss_bins)
)


```


# c from fiber-seq paper to the above analysis, see if those features correlate with expression and/or nucleosome positioning around TSS

/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL10_bc2178_19130/extracted_results/cpg_by_chr
/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL10_bc2178_19130/extracted_results/m6a_by_chr
/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results/AL10_bc2178_19130/extracted_resulta/msp_by_chr

sample = AL10_bc2178_19130
chr = 1

/project/spott/cshan/fiber-seq/code/expr_stratified_TSS_profiles_with_fiber_features.r
/project/spott/cshan/fiber-seq/code/run_expr_stratified_TSS_profiles_with_fiber_features.sh
```{r expr_stratified_TSS_profiles_with_fiber_features}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

gencode_tss_path <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"
cage_tss_path    <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz"
sample_root      <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
expr_path        <- "/project/spott/cshan/annotations/ENCFF873VWU.gene_quan.tsv"
default_sample   <- "AL10_bc2178_19130"
default_chrom    <- "chr1"
window_bp        <- 2000L
bin_size         <- 10L

source_colors <- c("GENCODE" = "#1f77b4", "CAGE" = "#be1c54ff")
expr_colors   <- c("high" = "#FF7F0E", "low" = "#2CA02C")
feature_colors <- c(
  "Nucleosome" = "#4D4D4D",
  "m6A"        = "#7B3294",
  "CpG"        = "#8C510A",
  "MSP"        = "#1B9E77"
)
feature_levels <- c("Nucleosome", "m6A", "CpG", "MSP")

sample_extract_dir <- file.path(sample_root, default_sample, "extracted_results")
feature_files <- list(
  Nucleosome = file.path(sample_extract_dir, "nuc_by_chr",
                         sprintf("%s.ft_extracted_nuc.%s.bed.gz", default_sample, default_chrom)),
  m6A        = file.path(sample_extract_dir, "m6a_by_chr",
                         sprintf("%s.ft_extracted_m6a.%s.bed.gz", default_sample, default_chrom)),
  CpG        = file.path(sample_extract_dir, "cpg_by_chr",
                         sprintf("%s.ft_extracted_cpg.%s.bed.gz", default_sample, default_chrom)),
  MSP        = file.path(sample_extract_dir, "msp_by_chr",
                         sprintf("%s.ft_extracted_msp.%s.bed.gz", default_sample, default_chrom))
)
if (!all(file.exists(unlist(feature_files)))) {
  stop("Missing one or more feature BED12 files for ", default_sample, " ", default_chrom)
}

feature_grs <- list(
  Nucleosome = load_fiber_feature_positions(feature_files[["Nucleosome"]], feature_name = "Nucleosome"),
  m6A        = load_fiber_feature_positions(feature_files[["m6A"]],        feature_name = "m6A"),
  CpG        = load_fiber_feature_positions(feature_files[["CpG"]],        feature_name = "CpG"),
  MSP        = load_fiber_feature_positions(feature_files[["MSP"]],        feature_name = "MSP")
)

annotate_gene_id_base <- function(gr) {
  mcols(gr)$gene_id_base <- sub("\\..*$", "", mcols(gr)$gene_id)
  gr
}

expr <- fread(expr_path)
expr <- expr[grepl("^ENSG", gene_id)]
expr[, gene_id_base := sub("\\..*$", "", gene_id)]
qs <- quantile(expr$TPM, probs = c(0.25, 0.75), na.rm = TRUE)
expr[, expr_group := fcase(
  TPM >= qs[2], "high",
  TPM <= qs[1], "low",
  default = NA_character_
)]
expr <- expr[!is.na(expr_group), .(gene_id, gene_id_base, TPM, expr_group)]

tss_sets <- list(
  GENCODE = annotate_gene_id_base(load_tss_annotations(gencode_tss_path)),
  CAGE    = annotate_gene_id_base(load_tss_annotations_bed9_with_gene_id(cage_tss_path))
)

align_tss <- function(gr, source) {
  idx_full <- match(mcols(gr)$gene_id, expr$gene_id)
  idx_base <- match(mcols(gr)$gene_id_base, expr$gene_id_base)
  idx <- ifelse(!is.na(idx_full), idx_full, idx_base)
  valid <- !is.na(idx)
  if (!any(valid)) return(NULL)
  gr2 <- gr[valid]
  mcols(gr2)$expr_group <- expr$expr_group[idx[valid]]
  mcols(gr2)$TPM <- expr$TPM[idx[valid]]
  list(source = source, gr = gr2)
}

tss_by_group <- Filter(
  Negate(is.null),
  lapply(names(tss_sets), function(source) align_tss(tss_sets[[source]], source))
)

matched_sources <- vapply(tss_by_group, function(x) x$source, character(1))
if (!all(c("GENCODE", "CAGE") %in% matched_sources)) {
  stop("One or more TSS sources did not match the expression table; check CAGE gene_id annotation.")
}

meta_raw_list <- list()
meta_norm_list <- list()

for (entry in tss_by_group) {
  for (grp in c("high", "low")) {
    subset_gr <- entry$gr[mcols(entry$gr)$expr_group == grp]
    if (!length(subset_gr)) next

    for (feature_nm in names(feature_grs)) {
      raw_dt <- make_position_meta(
        feature_grs[[feature_nm]], subset_gr,
        window_bp = window_bp, bin_size = bin_size,
        normalize_by_tss = FALSE
      )
      raw_dt[, `:=`(feature = feature_nm, source = entry$source, expr_group = grp)]
      meta_raw_list[[length(meta_raw_list) + 1L]] <- raw_dt

      norm_dt <- make_position_meta(
        feature_grs[[feature_nm]], subset_gr,
        window_bp = window_bp, bin_size = bin_size,
        normalize_by_tss = TRUE
      )
      norm_dt[, `:=`(feature = feature_nm, source = entry$source, expr_group = grp)]
      meta_norm_list[[length(meta_norm_list) + 1L]] <- norm_dt
    }
  }
}

meta_raw <- rbindlist(meta_raw_list, use.names = TRUE, fill = TRUE)
meta_norm <- rbindlist(meta_norm_list, use.names = TRUE, fill = TRUE)

for (dt in list(meta_raw, meta_norm)) {
  dt[, source := factor(source, levels = c("GENCODE", "CAGE"))]
  dt[, expr_group := factor(expr_group, levels = c("high", "low"))]
  dt[, feature := factor(feature, levels = feature_levels)]
}

x_scale <- scale_x_continuous(
  breaks = seq(-window_bp, window_bp, 500),
  labels = function(x) paste0(x / 1000, " kb")
)

raw_source_plot <- ggplot(meta_raw, aes(x = bin, y = N, color = source)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 0.9) +
  facet_grid(feature ~ expr_group, scales = "free_y") +
  scale_color_manual(values = source_colors, drop = FALSE) +
  x_scale +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Distance to TSS",
    y = "Feature midpoints per 10 bp bin",
    color = "TSS source",
    title = "Fiber-seq features around TSS: raw counts by expression group",
    subtitle = sprintf("Sample: %s | Chromosome: %s", default_sample, default_chrom)
  ) +
  theme_bw(base_size = 12)

norm_source_plot <- ggplot(meta_norm, aes(x = bin, y = N, color = source)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 0.9) +
  facet_grid(feature ~ expr_group, scales = "free_y") +
  scale_color_manual(values = source_colors, drop = FALSE) +
  x_scale +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Distance to TSS",
    y = "Average feature midpoints per TSS (10 bp bins)",
    color = "TSS source",
    title = "Fiber-seq features around TSS: per-TSS normalized profiles",
    subtitle = sprintf("Sample: %s | Chromosome: %s", default_sample, default_chrom)
  ) +
  theme_bw(base_size = 12)

raw_expr_plot <- ggplot(meta_raw, aes(x = bin, y = N, color = expr_group)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 0.9) +
  facet_grid(feature ~ source, scales = "free_y") +
  scale_color_manual(values = expr_colors, drop = FALSE) +
  x_scale +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Distance to TSS",
    y = "Feature midpoints per 10 bp bin",
    color = "Expression bin",
    title = "Fiber-seq features around TSS: raw counts by TSS source",
    subtitle = sprintf("Sample: %s | Chromosome: %s", default_sample, default_chrom)
  ) +
  theme_bw(base_size = 12)

norm_expr_plot <- ggplot(meta_norm, aes(x = bin, y = N, color = expr_group)) +
  geom_line(linewidth = 0.7) +
  geom_smooth(method = "loess", span = 0.06, se = FALSE, linewidth = 0.9) +
  facet_grid(feature ~ source, scales = "free_y") +
  scale_color_manual(values = expr_colors, drop = FALSE) +
  x_scale +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Distance to TSS",
    y = "Average feature midpoints per TSS (10 bp bins)",
    color = "Expression bin",
    title = "Fiber-seq features around TSS: normalized by TSS source",
    subtitle = sprintf("Sample: %s | Chromosome: %s", default_sample, default_chrom)
  ) +
  theme_bw(base_size = 12)

meta_z <- copy(meta_norm)
meta_z[, z := if (sd(N) == 0) 0 else as.numeric(scale(N)),
       by = .(feature, source, expr_group)]

shape_overlay_plot <- ggplot(meta_z, aes(x = bin, y = z, color = feature)) +
  geom_line(linewidth = 0.8) +
  facet_grid(expr_group ~ source) +
  scale_color_manual(values = feature_colors, drop = FALSE) +
  x_scale +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Distance to TSS",
    y = "Z-scored normalized signal",
    color = "Feature",
    title = "Relative positioning of nucleosome, m6A, CpG, and MSP signals",
    subtitle = "Z-scores are computed within each feature, TSS source, and expression group"
  ) +
  theme_bw(base_size = 12)

safe_cor <- function(x, y, method = "pearson") {
  if (length(unique(x)) < 2L || length(unique(y)) < 2L) return(NA_real_)
  suppressWarnings(cor(x, y, method = method))
}

profile_corr <- rbindlist(lapply(setdiff(feature_levels, "Nucleosome"), function(feature_nm) {
  merged <- merge(
    meta_norm[feature == "Nucleosome", .(bin, source, expr_group, nuc_signal = N)],
    meta_norm[feature == feature_nm,   .(bin, source, expr_group, feature_signal = N)],
    by = c("bin", "source", "expr_group"),
    all = FALSE
  )
  merged[, .(
    pearson = safe_cor(nuc_signal, feature_signal, method = "pearson"),
    spearman = safe_cor(nuc_signal, feature_signal, method = "spearman")
  ), by = .(source, expr_group)][, feature := feature_nm]
}), use.names = TRUE, fill = TRUE)

profile_corr[, feature := factor(feature, levels = c("m6A", "CpG", "MSP"))]
corr_plot <- ggplot(profile_corr, aes(x = source, y = feature, fill = spearman)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", spearman)), size = 3.5) +
  facet_wrap(~expr_group, nrow = 1) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1), na.value = "grey90") +
  labs(
    x = "TSS source",
    y = "Feature",
    fill = "Spearman\nrho",
    title = "Correlation between nucleosome and feature profiles across TSS bins",
    subtitle = "Positive values indicate co-enrichment in the same TSS-relative bins"
  ) +
  theme_bw(base_size = 12)

feature_window_signal <- meta_norm[, .(signal_per_tss = sum(N)),
                                   by = .(feature, source, expr_group)]
expr_enrichment <- dcast(
  feature_window_signal,
  feature + source ~ expr_group,
  value.var = "signal_per_tss",
  fill = 0
)
expr_enrichment[, log2_high_vs_low := log2((high + 1e-6) / (low + 1e-6))]
expr_enrichment[, feature := factor(feature, levels = feature_levels)]

expr_enrichment_plot <- ggplot(expr_enrichment, aes(x = source, y = feature, fill = log2_high_vs_low)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", log2_high_vs_low)), size = 3.5) +
  scale_fill_gradient2(low = "#2CA02C", mid = "white", high = "#FF7F0E",
                       midpoint = 0) +
  labs(
    x = "TSS source",
    y = "Feature",
    fill = "log2(high/low)",
    title = "Expression association of fiber-seq features within +/-2 kb of TSS",
    subtitle = "Computed from per-TSS normalized signal summed across bins"
  ) +
  theme_bw(base_size = 12)

print(raw_source_plot)
print(norm_source_plot)
print(raw_expr_plot)
print(norm_expr_plot)
print(shape_overlay_plot)
print(corr_plot)
print(expr_enrichment_plot)

plot_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc"
table_dir <- "/project/spott/cshan/fiber-seq/results/tables/nuc"
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_raw_by_source_expr.pdf",
                         default_sample, default_chrom)),
       plot = raw_source_plot, width = 10, height = 10)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_norm_by_source_expr.pdf",
                         default_sample, default_chrom)),
       plot = norm_source_plot, width = 10, height = 10)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_raw_by_expr_source.pdf",
                         default_sample, default_chrom)),
       plot = raw_expr_plot, width = 10, height = 10)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_norm_by_expr_source.pdf",
                         default_sample, default_chrom)),
       plot = norm_expr_plot, width = 10, height = 10)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_zscore_overlay.pdf",
                         default_sample, default_chrom)),
       plot = shape_overlay_plot, width = 8, height = 5)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_vs_nucleosome_correlation_heatmap.pdf",
                         default_sample, default_chrom)),
       plot = corr_plot, width = 7, height = 4)

ggsave(file.path(plot_dir,
                 sprintf("%s_%s_fiber_features_expression_heatmap.pdf",
                         default_sample, default_chrom)),
       plot = expr_enrichment_plot, width = 7, height = 4.5)

fwrite(meta_raw,
       file.path(table_dir,
                 sprintf("%s_%s_fiber_features_raw_profiles.tsv",
                         default_sample, default_chrom)),
       sep = "\t")

fwrite(meta_norm,
       file.path(table_dir,
                 sprintf("%s_%s_fiber_features_norm_profiles.tsv",
                         default_sample, default_chrom)),
       sep = "\t")

fwrite(profile_corr,
       file.path(table_dir,
                 sprintf("%s_%s_fiber_features_vs_nucleosome_correlations.tsv",
                         default_sample, default_chrom)),
       sep = "\t")

fwrite(expr_enrichment,
       file.path(table_dir,
                 sprintf("%s_%s_fiber_features_high_vs_low_expression.tsv",
                         default_sample, default_chrom)),
       sep = "\t")

```

# Offset from TSS to closest nucleosome midpoint 

```{r find_closest_nucleosomes_to_TSS}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")


# ---- Parameters ----
sample_root    <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
default_sample <- "AL10_bc2178_19130"
default_chrom  <- "chr1"
window_bp      <- 2000L
cage_tss_path  <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"
n_ranks        <- 5L

# ---- Load nucleosome BED12 ----
# Each row = one sequencing read (molecule); blocks within the read = nucleosomes
nuc_by_chr_dir <- file.path(sample_root, default_sample, "extracted_results/nuc_by_chr")
pattern  <- sprintf("ft_extracted_nuc\\.%s\\.bed\\.gz$", default_chrom)
nuc_file <- list.files(nuc_by_chr_dir, pattern = pattern, full.names = TRUE)[1]
stopifnot(!is.na(nuc_file), file.exists(nuc_file))

bed12 <- fread(
  nuc_file,
  sep = "\t", header = FALSE, select = 1:12,
  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                "thickStart","thickEnd","itemRgb","blockCount","blockSizes","blockStarts")
)

# ---- Step 1: Build a GRanges of reads (one row per molecule) ----
# Analog of fire.gr in Kevin's get_nearest_nuc_to_snp:
#   Kevin uses fire.gr (reads) to findOverlaps with SNPs;
#   here we use reads_gr to findOverlaps with TSS windows.
# 'name' is the read ID (RID) — same as Kevin's RID field.
reads_gr <- GRanges(
  seqnames = bed12$chrom,
  ranges   = IRanges(bed12$chromStart + 1L, bed12$chromEnd),
  strand   = "*",
  RID      = bed12$name
)

# ---- Step 2: Build per-nucleosome table keyed by read ID ----
# calculate_nucleosome_midpoints parses BED12 blocks -> one row per nucleosome.
# 'name' (= RID) is carried through in the grouping key.
# nuc_mid is 0-based (raw BED arithmetic); we add 1L later to match GRanges 1-based coords
# midpoint is calculated as the average of the nucleosome start and end coordinates
nucs <- calculate_nucleosome_midpoints(bed12)
# rename name to RID for joining with reads_gr and TSS overlaps later
setnames(nucs, "name", "RID") 
# keep only read ID, nucleosome midpoint and size
nuc_read_cols <- nucs[, .(RID, nuc_mid, nuc_size)]

# ---- Step 3: Load TSS GRanges and restrict to working chromosome ----
tss_gr <- load_tss_annotations_bed9(cage_tss_path)
tss_gr <- tss_gr[as.character(seqnames(tss_gr)) == default_chrom]

# TSS lookup table for joining (tss_idx = integer index into tss_gr)
tss_dt <- data.table(
  # assign each TSS a unique index
  tss_idx    = seq_along(tss_gr),
  tss_pos    = start(tss_gr),            # 1-based (GRanges convention)
  tss_strand = as.character(strand(tss_gr))
)

# ---- Step 4: Find reads that overlap each TSS ± window ----
# Analog of: hits_snp <- findOverlaps(fire.gr, snps.gr)
# For each TSS, expand a window of ±window_bp; find all reads whose span overlaps it.
tss_windows_gr <- promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)
hits_tss <- findOverlaps(reads_gr, tss_windows_gr)

# One row per (read, TSS) overlap
read_tss_pairs <- data.table(
  RID     = reads_gr$RID[queryHits(hits_tss)], # read ID for each overlap
  tss_idx = subjectHits(hits_tss)               # index of the TSS in tss_gr that overlaps the read
)

# ---- Step 5: For each (read, TSS) pair, expand to all nucleosomes on that read ----
# Analog of Kevin's inner_join(snp_df, nuc_df, by = "RID", relationship = "many-to-many"):
#   each read may have many nucleosomes, and each read may overlap many TSS.
rtp_nuc <- merge(read_tss_pairs, nuc_read_cols, by = "RID", allow.cartesian = TRUE) # for each read-TSS pair, get all nucleosomes on that read
rtp_nuc <- merge(rtp_nuc, tss_dt, by = "tss_idx", allow.cartesian = TRUE)

# ---- Step 6: Compute strand-aware offset from TSS to nucleosome midpoint ----
# nuc_mid is 0-based; add 1L to convert to 1-based before comparing with tss_pos (1-based).
# On + strand: positive rel_pos = nucleosome is downstream of TSS.
# On - strand: flip sign so rel_pos is always in TSS-centric orientation
#              (negative = upstream of TSS regardless of genomic strand).
# Analog of Kevin's dist_nuc_mid, but signed rather than absolute.
rtp_nuc[, rel_pos  := fifelse(tss_strand == "+",
                               (nuc_mid + 1L) - tss_pos,
                               tss_pos - (nuc_mid + 1L))]
rtp_nuc[, abs_dist := abs(rel_pos)]
rtp_nuc[, direction := fifelse(rel_pos < 0L, "upstream", "downstream")]

# ---- Step 7: Rank nucleosomes by proximity within each (TSS, read, direction) ----
# Analog of Kevin's slice_min(dist_nuc_mid, n=1) for the single nearest nuc,
# extended here to keep the top n_ranks per direction.
# rank 1 = nucleosome closest to the TSS on that read in that direction.
setorder(rtp_nuc, tss_idx, RID, direction, abs_dist)
rtp_nuc[, rank := seq_len(.N), by = .(tss_idx, RID, direction)]
rtp_nuc <- rtp_nuc[rank <= n_ranks]

rtp_nuc[, signed_rank   := fifelse(direction == "upstream", -rank, rank)]
rtp_nuc[, peak_adjacent := rank == 1L]
rtp_nuc[, peak_adjacent_label := ifelse(peak_adjacent, "Closest to TSS (rank 1)", "Ranks 2–5")]
rtp_nuc[, peak_adjacent_label := factor(peak_adjacent_label, 
                                        levels = c("Closest to TSS (rank 1)", "Ranks 2–5"))]

# ---- Plot: offset distribution by rank ----
# Each observation is one (read × TSS) data point, not an aggregate.
# This tests whether rank-1 nucleosomes (closest to TSS) have smaller,
# tighter offsets compared to ranks 2–5 on the same reads.
# Use coord_cartesian to zoom into the bulk of the data while preserving outlier-checking
offset_plot <- ggplot(rtp_nuc, aes(x = factor(signed_rank), y = abs_dist,
                                    fill = peak_adjacent_label)) +
  geom_boxplot(outlier.size = 1.5, outlier.color = "grey50", outlier.alpha = 0.35) +
  scale_fill_manual(
    values = c("Closest to TSS (rank 1)" = "#6a51a3", "Ranks 2–5" = "gray70")
  ) +
  coord_cartesian(ylim = c(0, quantile(rtp_nuc$abs_dist, 0.95))) +
  labs(
    x     = "Nucleosome rank (upstream = negative, downstream = positive)",
    y     = "Distance to TSS midpoint (bp)",
    fill  = NULL,
    title = sprintf("Per-read nearest nucleosomes to CAGE TSS (%s, %s)",
                    default_sample, default_chrom),
    subtitle = sprintf("Each point = one read × TSS pair; window = ±%d bp (y-axis limited to 95th percentile)", window_bp)
  ) +
  theme_bw(base_size = 12)

print(offset_plot)

# ---- Plot: nucleosome size distributions by direction ----
upstream_size_hist <- ggplot(rtp_nuc %>% filter(direction == "upstream", !is.na(nuc_size), nuc_size < 1000),
                             aes(x = nuc_size)) +
  geom_histogram(bins = 20, fill = "lightblue", color = "black", alpha = 0.9) +
  labs(title = "Upstream nucleosome size distribution (per-read)",
       x = "Nucleosome size (bp)",
       y = "Frequency") +
  theme_minimal()

downstream_size_hist <- ggplot(rtp_nuc %>% filter(direction == "downstream", !is.na(nuc_size), nuc_size < 1000),
                               aes(x = nuc_size)) +
  geom_histogram(bins = 20, fill = "lightcoral", color = "black", alpha = 0.9) +
  labs(title = "Downstream nucleosome size distribution (per-read)",
       x = "Nucleosome size (bp)",
       y = "Frequency") +
  theme_minimal()

print(upstream_size_hist)
print(downstream_size_hist)

plot_dir <- "/project/spott/cshan/fiber-seq/results/plots/nuc/offset_per_read_TSS"
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(
  file.path(plot_dir,
            sprintf("TSS_nearest_nuc_per_read_%s_%s.pdf", default_sample, default_chrom)),
  plot = offset_plot, width = 10, height = 7
)

ggsave(
  file.path(plot_dir,
            sprintf("TSS_nearest_nuc_upstream_size_hist_per_read_%s_%s.pdf", default_sample, default_chrom)),
  plot = upstream_size_hist, width = 6, height = 4
)

ggsave(
  file.path(plot_dir,
            sprintf("TSS_nearest_nuc_downstream_size_hist_per_read_%s_%s.pdf", default_sample, default_chrom)),
  plot = downstream_size_hist, width = 6, height = 4
)


# perfomr wilcoxon test to see if rank 1 nucleosomes are significantly closer to TSS than ranks 2-5
wilcox_test <- wilcox.test(abs_dist ~ peak_adjacent_label, data = rtp_nuc)
wilcox_test_result <- data.table(
  statistic = wilcox_test$statistic,
  p_value = wilcox_test$p.value,
  method = wilcox_test$method,
  alternative = wilcox_test$alternative
)

# Build comparison pairs: upstream ranks vs -1, downstream ranks vs 1
# (rank_1 covers both -1 and 1, so split by direction)
other_srs        <- rtp_nuc[peak_adjacent == FALSE, sort(unique(signed_rank))]
upstream_comps   <- lapply(other_srs[other_srs < 0], function(r) c("-1", as.character(r)))
downstream_comps <- lapply(other_srs[other_srs > 0], function(r) c("1",  as.character(r)))
all_comps        <- c(upstream_comps, downstream_comps)

# Pre-filter to only significant pairs (p < 0.05) so ns brackets are hidden
sig_comps <- Filter(function(pair) {
  g1 <- rtp_nuc[as.character(signed_rank) == pair[1], abs_dist]
  g2 <- rtp_nuc[as.character(signed_rank) == pair[2], abs_dist]
  wilcox.test(g1, g2)$p.value < 0.05
}, all_comps)

# Expand y ceiling to accommodate stacked brackets without clipping
y_ceil <- quantile(rtp_nuc$abs_dist, 0.95, na.rm = TRUE) * (1 + 0.08 * length(sig_comps))

# Rebuild plot with expanded ylim (avoids replacing coord_cartesian mid-chain)
offset_plot_annotated <- ggplot(rtp_nuc, aes(x = factor(signed_rank), y = abs_dist,
                                              fill = peak_adjacent_label)) +
  geom_boxplot(outlier.size = 1.5, outlier.color = "grey50", outlier.alpha = 0.35) +
  scale_fill_manual(
    values = c("Closest to TSS (rank 1)" = "#6a51a3", "Ranks 2–5" = "gray70")
  ) +
  coord_cartesian(ylim = c(0, y_ceil)) +
  labs(
    x     = "Nucleosome rank (upstream = negative, downstream = positive)",
    y     = "Distance to TSS midpoint (bp)",
    fill  = NULL,
    title = sprintf("Per-read nearest nucleosomes to CAGE TSS (%s, %s)",
                    default_sample, default_chrom),
    subtitle = sprintf("Each point = one read × TSS pair; window = ±%d bp (y-axis limited to 95th percentile)", window_bp)
  ) +
  theme_bw(base_size = 12) +
  ggsignif::geom_signif(
    comparisons      = sig_comps,
    test             = "wilcox.test",
    map_signif_level = TRUE,
    step_increase    = 0.08,
    tip_length       = 0.01,
    color            = "red",
    textsize         = 3.5
  )

print(offset_plot_annotated)
```

# Plot FIRE, m6A, CpG on a genome wide scale

Genome-wide track plots showing:

- **FIRE score** (`all.percent.accessible.bw`) — fraction of reads with a FIRE element
- **m6A / MSP fraction** — derived from `fire_all.bed.gz` (MSP + FIRE element bases / total covered bases per bin)
- **CpG methylation** (`*_CPG.combined.bw`) — average CpG methylation fraction per bin
- **Nucleosome coverage** (`all.nucleosome.coverage.bw`)
- **FIRE peaks** annotated as tick marks below the FIRE track

```{r genome_wide_setup, message=FALSE, warning=FALSE}
## ── Packages ─────────────────────────────────────────────────────────────────
library(GenomicRanges)
library(IRanges)
library(rtracklayer)
library(ggplot2)
library(dplyr)
library(data.table)
library(patchwork)   # multi-panel layout
library(scales)      # axis formatting

## ── Paths ────────────────────────────────────────────────────────────────────
sample        <- "AL10_bc2178_19130"
fire_root     <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
cpg_root      <- "/project/spott/cshan/fiber-seq/results/fire_CpG"
chrom_sizes_f <- "/project/spott/reference/human/GRCh38/hg38.chrom.sizes"

sample_fire_dir <- file.path(fire_root, sample)
bw_dir          <- file.path(sample_fire_dir, "trackHub-v0.1", "bw")
extract_dir     <- file.path(sample_fire_dir, "extracted_results")

bw_fire    <- file.path(bw_dir, "all.percent.accessible.bw")
bw_nuc     <- file.path(bw_dir, "all.nucleosome.coverage.bw")
bw_cpg     <- file.path(cpg_root, sample, paste0(sample, "_CPG.combined.bw"))
peaks_file <- file.path(sample_fire_dir,
                        paste0(sample, "-fire-v0.1-peaks.bed.gz"))
fire_all_f <- file.path(extract_dir,
                        paste0(sample, ".fire_all.bed.gz"))
```


Tiles the genome into 100 kb bins, imports FIRE, nucleosome, and CpG bigwigs
```{r genome_wide_bins, message=FALSE, warning=FALSE}
## ── Select chromosome ─────────────────────────────────────────────────────────
## Change plot_chr to any canonical chromosome to switch the view
plot_chr <- "chr1"

BIN_SIZE <- 100e3

chrom_sizes <- fread(chrom_sizes_f, col.names = c("chrom", "size"))
chr_size    <- chrom_sizes[chrom == plot_chr, size]
stopifnot(length(chr_size) == 1)

## Build bins for the selected chromosome only
chr_gr  <- GRanges(plot_chr, IRanges(1, chr_size))
seqlengths(chr_gr) <- setNames(chr_size, plot_chr)

bins_gr <- tileGenome(seqlengths(chr_gr),
                      tilewidth  = BIN_SIZE,
                      cut.last.tile.in.chrom = TRUE)

## Helper: summarise a BigWig into bins for the selected chromosome.
## binnedAverage requires names(RleList) == seqlevels(bins) exactly,
## so we extract only the one chromosome and rebuild a matching RleList.
binnedBW <- function(bw_path, bins) {
  chr <- seqlevels(bins)                         # single chromosome name
  bw  <- import(bw_path, which = bins, as = "RleList")
  rle <- if (!is.null(bw[[chr]])) bw[[chr]] else Rle(0, seqlengths(bins)[chr])
  binnedAverage(bins, RleList(setNames(list(rle), chr)), "score")
}

message("Importing FIRE percent-accessible bigwig for ", plot_chr, " …")
bins_fire <- binnedBW(bw_fire, bins_gr)

message("Importing nucleosome coverage bigwig for ", plot_chr, " …")
bins_nuc    <- binnedBW(bw_nuc,  bins_gr)

message("Importing linker/MSP coverage bigwig for ", plot_chr, " …")
bw_linker   <- file.path(bw_dir, "all.linker.coverage.bw")
bins_linker <- binnedBW(bw_linker, bins_gr)

message("Importing FIRE element coverage bigwig for ", plot_chr, " …")
bw_fire_cov   <- file.path(bw_dir, "all.fire.coverage.bw")
bins_fire_cov <- binnedBW(bw_fire_cov, bins_gr)

message("Importing CpG methylation bigwig for ", plot_chr, " …")
bins_cpg  <- binnedBW(bw_cpg,  bins_gr)
```

```{r genome_wide_m6a, message=FALSE, warning=FALSE}
## ── Derive per-bin m6A and nucleosome fractions from coverage bigwigs ─────────
## Three mutually exclusive element classes per read:
##   FIRE element       (all.fire.coverage.bw)     → open, m6A-accessible
##   Linker / MSP       (all.linker.coverage.bw)   → open, m6A-accessible
##   Nucleosome         (all.nucleosome.coverage.bw) → closed
##
## Each bigwig score is a per-bin average coverage fraction after
## binnedAverage() summarises the signal across the 100 kb window.
## Let:
##   fc = all.fire.coverage.bw score in the bin
##   lc = all.linker.coverage.bw score in the bin
##   nc = all.nucleosome.coverage.bw score in the bin
##   tot = fc + lc + nc
##
## Then:
##   m6A fraction = (fc + lc) / tot
##   nucleosome fraction = nc / tot
##
## These are fractions, so the expected range is 0 to 1.

fc   <- bins_fire_cov$score
lc   <- bins_linker$score
nc   <- bins_nuc$score
tot  <- fc + lc + nc

bins_gr$m6a <- ifelse(tot > 0, (fc + lc) / tot, NA_real_)
bins_gr$nuc_frac <- ifelse(tot > 0, nc / tot, NA_real_)
```

```{r genome_wide_peaks, message=FALSE, warning=FALSE}
## ── Load FIRE peaks ──────────────────────────────────────────────────────────
peaks_dt <- fread(
  cmd       = paste("zcat", peaks_file),
  sep       = "\t",
  col.names = c("chr","peak_start","peak_end","start","end",
                "coverage","fire_coverage","score","nuc_coverage",
                "msp_coverage","coverage_H1","fire_coverage_H1","score_H1",
                "nuc_coverage_H1","msp_coverage_H1","coverage_H2",
                "fire_coverage_H2","score_H2","nuc_coverage_H2",
                "msp_coverage_H2","FDR","log_FDR","FIRE_size_mean",
                "FIRE_size_ssd","FIRE_start_ssd","FIRE_end_ssd",
                "local_max_count","peak_length","pass_coverage")
)[chr == plot_chr]

peaks_gr <- GRanges(peaks_dt$chr,
                    IRanges(peaks_dt$peak_start + 1, peaks_dt$peak_end),
                    score    = peaks_dt$score,
                    log_FDR  = peaks_dt$log_FDR,
                    pass_cov = peaks_dt$pass_coverage)
```

```{r genome_wide_tidy, message=FALSE, warning=FALSE}
## ── Tidy data frame for ggplot2 ──────────────────────────────────────────────
bins_df <- as.data.frame(bins_gr) |>
  transmute(
    chrom  = as.character(seqnames),
    start  = start,
    end    = end,
    mid    = (start + end) / 2,
    fire   = bins_fire$score,   # FIRE percent accessible (from all.percent.accessible.bw)
    m6a    = m6a,               # (fire_cov + linker_cov) / total_cov
    cpg    = bins_cpg$score / 100,    # CpG methylation fraction derived from percent
    nuc    = nuc_frac           # nuc_cov / total_cov
  )

## Peaks for rug marks
peaks_df <- as.data.frame(peaks_gr) |>
  transmute(
    mid        = (start + end) / 2,
    fire_score = .data$score
  )
```

```{r genome_wide_plot, fig.width=14, fig.height=7, message=FALSE, warning=FALSE}
## ── What this plot shows ──────────────────────────────────────────────────────
## Three stacked genome-browser-style tracks along the selected chromosome,
## each summarised in 10 kb non-overlapping bins:
##
##  TRACK 1 — m6A fraction (purple)
##    The fraction of PacBio CCS reads at each genomic bin whose underlying DNA
##    is marked by m6A methylation (i.e. nucleosome-free, accessible chromatin).
##    Derived from the FIRE pipeline coverage bigwigs:
##      (all.fire.coverage.bw score + all.linker.coverage.bw score)
##      / (all.fire.coverage.bw score + all.linker.coverage.bw score
##         + all.nucleosome.coverage.bw score)
##    High m6A = open, accessible chromatin; low m6A = packed nucleosomes.
##
##  TRACK 2 — CpG methylation / 5mC (green)
##    `*_CPG.combined.bw` stores CpG methylation as a percentage-like score.
##    The code converts it to a fraction with `bins_cpg$score / 100`, then
##    summarizes it into 10 kb bins via `binnedAverage()`.
##    High 5mC = silenced / constitutive heterochromatin (e.g. gene deserts,
##    repeat-rich regions); low 5mC = actively regulated or gene-rich regions.
##
##  TRACK 3 — Nucleosome fraction (grey)
##    The complement of m6A fraction: proportion of reads carrying a positioned
##    nucleosome at each bin (nucleosome coverage / total coverage).
##    Inversely correlated with m6A; peaks here correspond to troughs in track 1.
##
##  RUG MARKS (orange, bottom of each panel)
##    Positions of called FIRE peaks — loci where a statistically significant
##    fraction of single molecules simultaneously show open chromatin
##    (recurrent accessibility across the cell population).

bins_long <- bins_df |>
  tidyr::pivot_longer(cols = c(m6a, cpg, nuc),   # FIRE % accessible removed (redundant with m6A)
                      names_to  = "track",
                      values_to = "value") |>
  arrange(track, mid) |>
  group_by(track) |>
  mutate(
    value_smooth = data.table::frollmean(
      value,
      n = 11L,
      align = "center",
      fill = NA_real_,
      na.rm = TRUE
    )
  ) |>
  ungroup() |>
  mutate(track = factor(track,
                        levels = c("m6a", "cpg", "nuc"),
                        labels = c("m6A fraction\n(open chromatin)",
                                   "CpG methylation fraction\n(5mC)",
                                   "Nucleosome\nfraction")))

track_colors <- c(
  "m6A fraction\n(open chromatin)" = "#756BB1",
  "CpG methylation fraction\n(5mC)" = "#31A354",
  "Nucleosome\nfraction"           = "#636363"
)

plot_title <- paste0(
  "Single-molecule chromatin state landscape along ", plot_chr,
  "  |  ", sample,
  "  (", BIN_SIZE / 1e3, " kb bins)"
)

gw_plot <- ggplot(bins_long, aes(x = mid / 1e6, y = value, colour = track)) +
  geom_line(linewidth = 0.15, alpha = 0.2, na.rm = TRUE) +
  geom_line(aes(y = value_smooth), linewidth = 0.7, na.rm = TRUE) +
  geom_rug(data = peaks_df,
           aes(x = mid / 1e6, alpha = fire_score),
           inherit.aes = FALSE,
           colour = "#E6550D", linewidth = 0.25,
           sides  = "b", length = unit(0.06, "npc"),
           show.legend = FALSE) +
  scale_colour_manual(values = track_colors, guide = "none") +
  scale_x_continuous(
    name   = paste0(plot_chr, " position (Mb)"),
    labels = label_number(suffix = " Mb", scale = 1),
    expand = c(0, 0)
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(3)) +
  facet_wrap(~ track, ncol = 1, scales = "free_y", strip.position = "left") +
  labs(
    title    = plot_title,
    subtitle = "Orange rug marks = FIRE peaks (population-level recurrent accessibility)",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text.y.left = element_text(size = 9, angle = 0, hjust = 1),
    strip.placement   = "outside",
    panel.spacing.y   = unit(1, "mm"),
    axis.text.x       = element_text(size = 9),
    axis.text.y       = element_text(size = 8),
    plot.title        = element_text(face = "bold", size = 11)
  )

print(gw_plot)

dir.create("/project/spott/cshan/fiber-seq/results/plots/FIRE_genome_tracks", recursive = TRUE, showWarnings = FALSE)
ggsave(
  file.path("/project/spott/cshan/fiber-seq/results/plots/FIRE_genome_tracks",
            paste0(sample, "_", plot_chr, "_tracks.pdf")),
  gw_plot,
  width = 14, height = 8, units = "in"
)
```

# inspect pb-CpG tools output

```{r bw-plot-setup}
library(Gviz)
library(rtracklayer)
library(BSgenome.Hsapiens.UCSC.hg38)

# --- variables ---
sample <- "AL10_bc2178_19130"
chr    <- "chr1"
total_len <- seqlengths(Hsapiens)[chr]

fire_bw_dir <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
  sample, "trackHub-v0.1/bw"
)
cpg_bw_dir <- file.path(
  "/project/spott/cshan/fiber-seq/results/fire_CpG", sample
)

# bigwig paths and per-track colours (no hap tracks)
bw_files <- list(
  fire_coverage       = file.path(fire_bw_dir, "all.fire.coverage.bw"),
  linker_coverage     = file.path(fire_bw_dir, "all.linker.coverage.bw"),
  nucleosome_coverage = file.path(fire_bw_dir, "all.nucleosome.coverage.bw"),
  pct_accessible      = file.path(fire_bw_dir, "all.percent.accessible.bw"),
  cpg_combined        = file.path(cpg_bw_dir,  paste0(sample, "_CPG.combined.bw"))
)

track_colors <- c(
  fire_coverage       = "#E41A1C",  # red
  linker_coverage     = "#FF7F00",  # orange
  nucleosome_coverage = "#4DAF4A",  # green
  pct_accessible      = "#984EA3",  # purple
  cpg_combined        = "#377EB8"   # blue
)

# simple genome axis + one DataTrack per bigwig
axis_track <- GenomeAxisTrack()

data_tracks <- mapply(function(name, path, color) {
  DataTrack(
    range         = path,
    genome        = "hg38",
    chromosome    = chr,
    name          = name,
    type          = "histogram",
    col.histogram = color,
    fill.histogram = color,
    ylim          = if (name == "cpg_combined") c(0, 500) else NULL
  )
}, names(bw_files), bw_files, track_colors, SIMPLIFY = FALSE)



# plot first 5 Mb of chr1 as a quick overview
pdf("/project/spott/cshan/fiber-seq/results/plots/FIRE_genome_tracks/bw_track_plot_AL10_bc2178_19130_chr1_100Mb_150Mb.pdf",
    width = 10, height = 6)

plotTracks(
  c(list(axis_track), data_tracks),
  chromosome = chr,
  from = 100e6,
  to = 150e6,
  main = paste(sample, chr)
)

dev.off()
```

```{r cpg-structure}
library(data.table)
library(ggplot2)
library(patchwork)

file <- "/project/spott/cshan/fiber-seq/results/fire_CpG/AL10_bc2178_19130/AL10_bc2178_19130_CPG.combined.bed.gz"

dt <- fread(
  cmd = paste0("zcat '", file, "' | grep -v '^##'"),
  sep = "\t",
  header = TRUE
)
# The header line is "#chrom ..." — strip the leading '#' from the first column
setnames(dt, "#chrom", "chrom")

str(dt)
head(dt)
```

```{r cpg-qc-plots, fig.width=12, fig.height=10}
# 1. Distribution of continuous mod_score
library(ggplot2)
library(scales)
library(patchwork) 

# 1. mod_score distribution
p1 <- ggplot(dt, aes(x = mod_score)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = NA) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "mod_score distribution (BW signal)",
    x = "mod_score (0-100)", y = "# CpG sites"
  ) +
  theme_bw()

# 2. discretized_mod_score
p2 <- ggplot(dt, aes(x = discretized_mod_score)) +
  geom_histogram(binwidth = 1, fill = "coral", color = NA) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "discretized_mod_score distribution",
    x = "discretized_mod_score", y = "# CpG sites"
  ) +
  theme_bw()

# 3. Coverage distribution
p3 <- ggplot(dt[cov <= 100], aes(x = cov)) +
  geom_histogram(binwidth = 1, fill = "forestgreen", color = NA) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(
    title = "Coverage per CpG site",
    x = "Coverage", y = "# CpG sites"
  ) +
  theme_bw()

# Combine plots (auto-scaled y-axis for each)
p_qc <- (p1 / p2 / p3)  # vertical stack

# Save
ggsave(
  "/project/spott/cshan/fiber-seq/results/plots/pg_cpg/cpg_qc_AL10_bc2178_19130.png",
  plot   = p_qc,
  width  = 10,
  height = 15
)


```

```{r cpg-high-methylation-summary, fig.width=8, fig.height=5}
summary_tbl <- data.table(
  Metric = c(
    "discretized_mod_score == 100",
    "mod_score > 90",
    "mod_score > 80",
    "Coverage == 4 (min-cov)",
    "cov == 4 AND disc_score == 100"
  ),
  Pct = c(
    mean(dt$discretized_mod_score == 100),
    mean(dt$mod_score > 90),
    mean(dt$mod_score > 80),
    mean(dt$cov == 4),
    mean(dt$cov == 4 & dt$discretized_mod_score == 100)
  )
)

p_bar <- ggplot(summary_tbl, aes(x = reorder(Metric, Pct), y = Pct * 100)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(Pct * 100, 1), "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 105), labels = function(x) paste0(x, "%")) +
  labs(
    title = "CpG site summary: fraction meeting each criterion",
    x = NULL, y = "% of all CpG sites"
  ) +
  theme_bw()

ggsave(
  "/project/spott/cshan/fiber-seq/results/plots/pg_cpg/cpg_summary_barplot.pdf",
  plot = p_bar, width = 8, height = 5
)

p_bar
knitr::kable(summary_tbl[, .(Metric, `% of Sites` = paste0(round(Pct * 100, 1), "%"))],
             align = c("l", "r"))
```

# Coverage QC: why raw counts drop at the edges in the per-read plot
############################################################################################

The per-read normalized meta-profile drops at ±2 kb because reads must physically
span from the TSS to that distance to contribute nucleosomes there. Fewer long reads
reach the edges, so the denominator (total read-TSS pairs) is constant but the
numerator shrinks. The raw midpoint plot (CAGE_TSS_nucleosome_midpoints_raw.pdf)
does not suffer from this because each nucleosome midpoint is projected independently
to any TSS window it falls in — no read-length constraint.

## Step 1 — Reproduce {r CAGE_TSS} data

```{r coverage_qc_setup_raw}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

tss_path_qc    <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"
sample_root_qc <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
sample_qc      <- "AL10_bc2178_19130"
chrom_qc       <- "chr1"
window_bp_qc   <- 2000L
bin_size_qc    <- 10L

# locate chr1 BED12
nuc_dir_qc <- file.path(sample_root_qc, sample_qc, "extracted_results/nuc_by_chr")
bed_files_qc <- list.files(nuc_dir_qc,
                            pattern = "ft_extracted_nuc\\.chr.*\\.bed\\.gz$",
                            full.names = TRUE)
bed_file_qc <- bed_files_qc[grepl(paste0("\\.", chrom_qc, "\\."), bed_files_qc)][1]
stopifnot(!is.na(bed_file_qc), file.exists(bed_file_qc))

bed12_qc <- fread(bed_file_qc, header = FALSE,
                  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                                "thickStart","thickEnd","itemRgb",
                                "blockCount","blockSizes","blockStarts"))

nucs_qc   <- calculate_nucleosome_midpoints(bed12_qc)
nuc_gr_qc <- nuc_midpoints_to_granges(nucs_qc)
tss_gr_qc <- load_tss_annotations_bed9(tss_path_qc)

# raw midpoint meta (same as CAGE_TSS chunk)
plot_dt_qc <- find_nucs_near_tss(nuc_gr_qc, tss_gr_qc, window_bp = window_bp_qc)
plot_dt_qc[, bin := round(rel_pos / bin_size_qc) * bin_size_qc]

meta_raw_qc <- plot_dt_qc[, .(
  N_raw        = .N,
  n_tss_raw    = uniqueN(tss_pos)   # how many distinct TSS contributed to this bin
), by = bin][order(bin)]

cat(sprintf("Raw method: %d bins, total nuc-TSS hits = %d\n",
            nrow(meta_raw_qc), sum(meta_raw_qc$N_raw)))
cat(sprintf("Number of TSS in window: %d\n", length(tss_gr_qc)))
```
Number of TSS in window: 111640
Raw method: 401 bins, total nuc-TSS hits = 3994365


## Step 2 — Reproduce {r CAGE_TSS_per_read} data

```{r coverage_qc_setup_per_read}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

# (same paths as above — already in memory, just restate for clarity)
tss_path_qc    <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"
sample_root_qc <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
sample_qc      <- "AL10_bc2178_19130"
chrom_qc       <- "chr1"
window_bp_qc   <- 2000L
bin_size_qc    <- 10L

nuc_dir_qc <- file.path(sample_root_qc, sample_qc, "extracted_results/nuc_by_chr")
bed_files_qc <- list.files(nuc_dir_qc,
                            pattern = "ft_extracted_nuc\\.chr.*\\.bed\\.gz$",
                            full.names = TRUE)
bed_file_qc <- bed_files_qc[grepl(paste0("\\.", chrom_qc, "\\."), bed_files_qc)][1]

bed12_qc <- fread(bed_file_qc, header = FALSE,
                  col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                                "thickStart","thickEnd","itemRgb",
                                "blockCount","blockSizes","blockStarts"))
tss_gr_qc <- load_tss_annotations_bed9(tss_path_qc)

# per-read raw meta — already returns n_reads and n_tss per bin
meta_per_read_qc <- make_meta_per_read_raw(
  bed12_qc, tss_gr_qc,
  window_bp = window_bp_qc,
  bin_size  = bin_size_qc,
  label     = "CAGE TSS"
)

cat(sprintf("Per-read method: %d bins, total nuc-read-TSS hits = %d\n",
            nrow(meta_per_read_qc), sum(meta_per_read_qc$N)))

# also compute read spans for coverage model
c <- bed12_qc[, .(
  RID        = name,
  read_start = chromStart,
  read_end   = chromEnd,
  read_len   = chromEnd - chromStart
)]
cat(sprintf("Read length: median = %d bp, mean = %d bp, max = %d bp\n",
            as.integer(median(c$read_len)),
            as.integer(mean(c$read_len)),
            max(c$read_len)))
```

Per-read method: 401 bins, total nuc-read-TSS hits = 2377156
Read length: median = 10552 bp, mean = 11271 bp, max = 50859 bp



## Step 3 — QC plots: coverage by distance

```{r coverage_qc_plots, fig.width=10, fig.height=12}
out_dir_qc <- "/project/spott/cshan/fiber-seq/results/plots/nuc/coverage_qc"
dir.create(out_dir_qc, recursive = TRUE, showWarnings = FALSE)

# ── QC 1: n_reads per bin in the per-read method ──────────────────────────────
p_nreads <- ggplot(meta_per_read_qc, aes(x = bin, y = n_reads)) +
  geom_line(linewidth = 0.7, color = "#2ca02c") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_qc, window_bp_qc, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x     = "Distance to TSS",
    y     = "Number of unique reads contributing",
    title = "Per-read method: reads per bin",
    subtitle = sprintf("Sample: %s | Chr: %s | Shows coverage drop at edges",
                       sample_qc, chrom_qc)
  ) +
  theme_bw(base_size = 12)

print(p_nreads)

# ── QC 2: n_tss per bin in the raw method ─────────────────────────────────────
p_ntss_raw <- ggplot(meta_raw_qc, aes(x = bin, y = n_tss_raw)) +
  geom_line(linewidth = 0.7, color = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_qc, window_bp_qc, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x     = "Distance to TSS",
    y     = "Number of distinct TSS contributing",
    title = "Raw method: distinct TSS per bin",
    subtitle = sprintf("Sample: %s | Chr: %s | Flat = no TSS-coverage bias",
                       sample_qc, chrom_qc)
  ) +
  theme_bw(base_size = 12)

print(p_ntss_raw)

# ── QC 3: n_tss per bin in per-read method ────────────────────────────────────
p_ntss_perread <- ggplot(meta_per_read_qc, aes(x = bin, y = n_tss)) +
  geom_line(linewidth = 0.7, color = "#d62728") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_qc, window_bp_qc, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x     = "Distance to TSS",
    y     = "Number of distinct TSS contributing",
    title = "Per-read method: distinct TSS per bin",
    subtitle = sprintf("Sample: %s | Chr: %s | Also drops at edges", sample_qc, chrom_qc)
  ) +
  theme_bw(base_size = 12)

print(p_ntss_perread)

# ── QC 4: overlay raw N for both methods (scaled to compare shape) ────────────
# scale each to max=1 to compare the shape of coverage drop
meta_raw_qc[,      N_scaled := N_raw / max(N_raw, na.rm = TRUE)]
meta_per_read_qc[, N_scaled := N    / max(N,    na.rm = TRUE)]

meta_compare <- rbindlist(list(
  meta_raw_qc[,      .(bin, N_scaled, method = "Raw midpoints")],
  meta_per_read_qc[, .(bin, N_scaled, method = "Per-read raw")]
))

p_compare <- ggplot(meta_compare, aes(x = bin, y = N_scaled, color = method)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_color_manual(values = c("Raw midpoints" = "steelblue", "Per-read raw" = "#2ca02c")) +
  scale_x_continuous(breaks = seq(-window_bp_qc, window_bp_qc, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x     = "Distance to TSS",
    y     = "Scaled N (max = 1)",
    color = "Method",
    title = "Shape comparison: raw midpoints vs per-read",
    subtitle = "Both scaled to max=1 to isolate edge-drop effect"
  ) +
  theme_bw(base_size = 12)

print(p_compare)

# ── QC 5: reads per bin vs n_tss per bin in per-read method (should track) ───
p_reads_vs_tss <- ggplot(meta_per_read_qc, aes(x = bin)) +
  geom_line(aes(y = n_reads, color = "n_reads"), linewidth = 0.7) +
  geom_line(aes(y = n_tss,   color = "n_tss"),   linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_color_manual(values = c("n_reads" = "#2ca02c", "n_tss" = "#d62728")) +
  scale_x_continuous(breaks = seq(-window_bp_qc, window_bp_qc, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x     = "Distance to TSS",
    y     = "Count",
    color = NULL,
    title = "Per-read method: reads vs TSS contributing per bin",
    subtitle = "If n_reads drops more steeply than n_tss → long reads are scarce at the edges"
  ) +
  theme_bw(base_size = 12)

print(p_reads_vs_tss)

# ── QC 5: distinct TSS per bin in per-read and raw methods ──────────────────────────────
#      bin     N n_reads n_tss   source
#    <num> <int>   <int> <int>   <char>
# 1: -2000   696     289   657 CAGE TSS
# 2: -1990   987     345   910 CAGE TSS
# 3: -1980  1137     392  1066 CAGE TSS
# 4: -1970   931     346   899 CAGE TSS
# 5: -1960  1289     410  1190 CAGE TSS
# 6: -1950  1036     363   979 CAGE TSS



distinct_tss_compare <- rbindlist(list(
  meta_raw_qc[,      .(bin, N = n_tss_raw, method = "Raw midpoints")],
  meta_per_read_qc[, .(bin, N = n_tss,     method = "Per-read raw")]
))

# head(distinct_tss_compare)
#      bin     N        method
#    <num> <int>        <char>
# 1: -2000  4687 Raw midpoints
# 2: -1990  6212 Raw midpoints
# 3: -1980  6952 Raw midpoints
# 4: -1970  6248 Raw midpoints
# 5: -1960  6960 Raw midpoints
# 6: -1950  6219 Raw midpoints

# histogram of n_tss per bin for both methods
p_hist_ntss <- ggplot(distinct_tss_compare, aes(x = N, fill = method)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_fill_manual(values = c("Raw midpoints" = "red", "Per-read raw" = "blue")) +
  labs(
    x = "Number of distinct TSS contributing to bin",
    y = "# bins",
    fill = "Method",
    title = "Distribution of distinct TSS per bin for both methods"
  ) +
  theme_bw(base_size = 12)

print(p_hist_ntss)

# save all QC plots
ggsave(file.path(out_dir_qc, sprintf("%s_%s_qc_nreads_per_bin.pdf",       sample_qc, chrom_qc)), plot = p_nreads,       width = 6, height = 4)
ggsave(file.path(out_dir_qc, sprintf("%s_%s_qc_ntss_raw_per_bin.pdf",     sample_qc, chrom_qc)), plot = p_ntss_raw,     width = 6, height = 4)
ggsave(file.path(out_dir_qc, sprintf("%s_%s_qc_ntss_perread_per_bin.pdf", sample_qc, chrom_qc)), plot = p_ntss_perread, width = 6, height = 4)
ggsave(file.path(out_dir_qc, sprintf("%s_%s_qc_method_shape_compare.pdf", sample_qc, chrom_qc)), plot = p_compare,      width = 7, height = 4)
ggsave(file.path(out_dir_qc, sprintf("%s_%s_qc_reads_vs_tss.pdf",         sample_qc, chrom_qc)), plot = p_reads_vs_tss, width = 6, height = 4)
ggsave(file.path(out_dir_qc, sprintf("%s_%s_qc_hist_ntss_compare.pdf",     sample_qc, chrom_qc)), plot = p_hist_ntss,    width = 6, height = 4)

cat("QC plots saved to:", out_dir_qc, "\n")
```

# Per-read coverage-normalised CAGE TSS meta-profile
#
# Problem: make_meta_per_read_raw() divides the total nucleosome count per bin
# by a single global value (total read-TSS pairs), so bins near ±2 kb receive
# the same denominator as bins near 0, even though far fewer reads physically
# span that distance. The result is an artificial drop in apparent nucleosome
# density at the window edges.
#
# Fix: divide each bin's raw count (N) by the number of UNIQUE READS that
# actually contributed nucleosomes to that bin (n_reads). This converts the
# y-axis from "total nuc-read-TSS triplets" to "mean nucleosomes per read at
# this position", with a denominator that reflects actual coverage at each bin.

```{r per_read_coverage_normalised}
source("/project/spott/cshan/fiber-seq/plot_nuc_distribution_functions.R")

tss_path_norm    <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.bed.gz"
sample_root_norm <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results"
sample_norm      <- "AL10_bc2178_19130"
chrom_norm       <- "chr1"
window_bp_norm   <- 2000L
bin_size_norm    <- 10L

# ── Load data (same as {r CAGE_TSS} and {r CAGE_TSS_per_read}) ──────────────
nuc_dir_norm <- file.path(sample_root_norm, sample_norm, "extracted_results/nuc_by_chr")
bed_files_norm <- list.files(nuc_dir_norm,
                              pattern = "ft_extracted_nuc\\.chr.*\\.bed\\.gz$",
                              full.names = TRUE)
bed_file_norm <- bed_files_norm[grepl(paste0("\\.", chrom_norm, "\\."), bed_files_norm)][1]
stopifnot(!is.na(bed_file_norm), file.exists(bed_file_norm))

bed12_norm <- fread(bed_file_norm, header = FALSE,
                    col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                                  "thickStart","thickEnd","itemRgb",
                                  "blockCount","blockSizes","blockStarts"))

tss_gr_norm <- load_tss_annotations_bed9(tss_path_norm)

# ── Raw per-read counts (already carries n_reads per bin) ───────────────────
# make_meta_per_read_raw() returns:
#   bin     : 10 bp bin centre (relative to TSS)
#   N       : total (nucleosome × read × TSS) triplets in this bin
#   n_reads : number of distinct reads contributing to this bin
#   n_tss   : number of distinct TSS contributing to this bin
meta_raw_norm <- make_meta_per_read_raw(
  bed12_norm, tss_gr_norm,
  window_bp = window_bp_norm,
  bin_size  = bin_size_norm,
  label     = "CAGE TSS"
)

# ── Coverage normalisation: N_per_read = N / n_reads ─────────────────────────
# For each 10 bp bin, divide the raw count by however many unique reads
# contributed a nucleosome to that bin.  Bins near ±2 kb have low n_reads
# (only long reads reach there), so this corrects for the coverage gradient
# instead of spreading that gradient into the signal.
meta_raw_norm[, N_per_read := N / n_reads]

# ── Plot 1: raw per-read counts (uncorrected) ────────────────────────────────
p_uncorrected <- ggplot(meta_raw_norm, aes(x = bin, y = N)) +
  geom_line(linewidth = 0.6, color = "#2ca02c") +
  geom_smooth(method = "loess", span = 0.05, se = FALSE,
              color = "#2ca02c", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_norm, window_bp_norm, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x        = "Distance to TSS",
    y        = "Nucleosome count (nuc × read × TSS triplets)",
    title    = "Per-read raw counts — UNCORRECTED",
    subtitle = sprintf("Sample: %s | Chr: %s | drops at edges = read-length bias",
                       sample_norm, chrom_norm)
  ) +
  theme_bw(base_size = 12)

print(p_uncorrected)

# ── Plot 2: coverage-normalised (N / n_reads per bin) ────────────────────────
p_corrected <- ggplot(meta_raw_norm, aes(x = bin, y = N_per_read)) +
  geom_line(linewidth = 0.6, color = "#d62728") +
  geom_smooth(method = "loess", span = 0.05, se = FALSE,
              color = "#d62728", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_norm, window_bp_norm, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x        = "Distance to TSS",
    y        = "Mean nucleosomes per contributing read (per 10 bp bin)",
    title    = "Per-read coverage-normalised — CORRECTED",
    subtitle = sprintf("Sample: %s | Chr: %s | N / n_reads per bin",
                       sample_norm, chrom_norm)
  ) +
  theme_bw(base_size = 12)

print(p_corrected)

# ── Plot 3: side-by-side comparison (both scaled to max = 1) ─────────────────
meta_compare_norm <- rbindlist(list(
  meta_raw_norm[, .(bin, value = N        / max(N,         na.rm = TRUE), method = "Uncorrected (N)")],
  meta_raw_norm[, .(bin, value = N_per_read / max(N_per_read, na.rm = TRUE), method = "Corrected (N / n_reads)")]
))

p_compare_norm <- ggplot(meta_compare_norm, aes(x = bin, y = value, color = method)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_color_manual(values = c(
    "Uncorrected (N)"        = "#2ca02c",
    "Corrected (N / n_reads)" = "#d62728"
  )) +
  scale_x_continuous(breaks = seq(-window_bp_norm, window_bp_norm, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x        = "Distance to TSS",
    y        = "Scaled signal (max = 1)",
    color    = NULL,
    title    = "Effect of coverage normalisation on per-read meta-profile",
    subtitle = "Both scaled to max=1 to compare shape only"
  ) +
  theme_bw(base_size = 12)

print(p_compare_norm)

# ── Plot 4: n_reads per bin (the coverage curve being corrected for) ──────────
p_coverage_curve <- ggplot(meta_raw_norm, aes(x = bin, y = n_reads)) +
  geom_line(linewidth = 0.7, color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(breaks = seq(-window_bp_norm, window_bp_norm, 500),
                     labels = function(x) paste0(x / 1000, " kb")) +
  labs(
    x        = "Distance to TSS",
    y        = "n_reads (unique reads contributing to bin)",
    title    = "Read coverage by distance — the correction factor",
    subtitle = "Dividing N by this curve removes the read-length gradient"
  ) +
  theme_bw(base_size = 12)

print(p_coverage_curve)

# ── Save outputs ─────────────────────────────────────────────────────────────
out_dir_norm <- "/project/spott/cshan/fiber-seq/results/plots/nuc/coverage_normalised"
dir.create(out_dir_norm, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(out_dir_norm,
                 sprintf("%s_%s_per_read_uncorrected.pdf",  sample_norm, chrom_norm)),
       plot = p_uncorrected,   width = 6, height = 4)
ggsave(file.path(out_dir_norm,
                 sprintf("%s_%s_per_read_corrected.pdf",    sample_norm, chrom_norm)),
       plot = p_corrected,     width = 6, height = 4)
ggsave(file.path(out_dir_norm,
                 sprintf("%s_%s_per_read_shape_compare.pdf", sample_norm, chrom_norm)),
       plot = p_compare_norm,  width = 7, height = 4)
ggsave(file.path(out_dir_norm,
                 sprintf("%s_%s_per_read_coverage_curve.pdf", sample_norm, chrom_norm)),
       plot = p_coverage_curve, width = 6, height = 4)

# Save the data table with both raw and normalised columns
fwrite(meta_raw_norm,
       file.path(out_dir_norm,
                 sprintf("%s_%s_per_read_coverage_normalised_meta.tsv",
                         sample_norm, chrom_norm)),
       sep = "\t")

cat("Coverage-normalised outputs saved to:", out_dir_norm, "\n")
```



# Plot all chromsomes aggregated FIRE peaks

this is output from /project/spott/cshan/fiber-seq/code/all_chr_aggregated_FIRE_features.r

/project/spott/cshan/fiber-seq/results/plots/nuc/all_chr_aggregated_FIRE_features

```{r}
library(data.table)
library(ggplot2)

sample <- "AL10_bc2178_19130"
table_dir <- "/project/spott/cshan/fiber-seq/results/nuc/all_chr_aggregated_FIRE_features/AL10_bc2178_19130/all_chr"
plot_dir  <- "/project/spott/cshan/fiber-seq/results/plots/nuc/all_chr_aggregated_FIRE_features"

meta_norm <- fread(file.path(
  table_dir,
  sprintf("%s_all_chr_fiber_features_norm_profiles.tsv.gz", sample)
))

meta_norm <- meta_norm[feature %in% c("Nucleosome", "m6A", "CpG")]

meta_z <- copy(meta_norm)
meta_z[, z := if (sd(N) == 0) 0 else as.numeric(scale(N)), by = feature]

feature_colors <- c(
  "Nucleosome" = "#4D4D4D",
  "m6A" = "#7B3294",
  "CpG" = "#8C510A"
)

p_no_msp <- ggplot(meta_z, aes(x = bin, y = z, color = feature)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~source, nrow = 1) +
  scale_color_manual(values = feature_colors, drop = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  scale_x_continuous(
    breaks = seq(-2000, 2000, 500),
    labels = function(x) paste0(x / 1000, " kb")
  ) +
  labs(
    x = "Distance to TSS",
    y = "Z-scored normalized signal",
    color = "Feature",
    title = "Relative positioning of nucleosome, m6A, and CpG signals",
    subtitle = "MSP filtered out before z-scoring"
  ) +
  theme_bw(base_size = 12)

ggsave(
  file.path(plot_dir, sprintf("%s_all_chr_FIRE_features_overlay_no_MSP.pdf", sample)),
  plot = p_no_msp,
  width = 8,
  height = 5
)

print(p_no_msp)
```