---
title: "Leiden + Manhattan clustering of single molecules at custom regions"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(data.table)
library(Rsamtools)   
library(igraph)
library(ComplexHeatmap)
library(cowplot)
```

# Functions

`extract_ft_region_reads()`, `extract_ft_read_info()`, `get_sparse_met_mat()`
and `filter_met_mat()` come from the topic model

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

# promoter_table <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical/promoter_regions_gencode_v49.tsv"

output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/Leiden_Manhattan"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Window ("feature") size used for the custom-region run. The previous window
# sweep is kept below as a comment for reference.
# window_sizes <- c(0, 10, 20, 30, 50)
window_size <- 0L

# Leiden / graph parameters

# k_neighbors: 
k_neighbors       <- 10

# leiden_resolution: Raise it for finer clusters, lower it for coarser ones
leiden_resolution <- 1
leiden_seed       <- 1

# Exponential-kernel width for the affinities: NULL = mean of the retained KNN
# distances (scale-free across window sizes). Set to the number of features to
# reproduce scikit-learn's laplacian_kernel default (gamma = 1 / n_features).
kernel_sigma <- NULL

# Paper step 2, off for now: with full-span reads there is nothing to impute.
impute_missing <- FALSE
impute_k       <- 5

# Sweep grids for the "KNN graph colored by Leiden cluster" figures: one pdf
# per bin size x k x resolution combination under <region_id>/bin<size>/.
# Ranges come from a pilot sweep (k 5-50, resolution 0.1-5, both regions, all
# bin sizes, ~200 full-span reads each): k = 5 fragments the graph and k = 50
# merges everything below resolution 0.5 then shatters above 1; resolution 0.1
# gives one cluster almost everywhere and > 2 over-splits (> 15 clusters).
sweep_window_sizes <- c(0L, 10L, 20L, 30L, 50L)
sweep_k_neighbors  <- c(10, 15, 20, 30)
sweep_resolutions  <- c(0.25, 0.5, 1, 1.5, 2)

# Rolling-mean width for the m6A fraction line plot only (1 = raw profiles).
# At window size 0 the per-site cluster means are spiky; e.g. 25 makes the
# nucleosome-scale structure readable without changing any clustering input.
line_smooth_k <- 1

set.seed(123)
```

# Regions

Custom regions for the Leiden + Manhattan run.

```{r regions}
# promoters <- read.delim(promoter_table)
# promoters
regions <- data.frame(
  chr = c("chr4", "chr5"),
  start = c(112284963, 150068577),
  end = c(112285236, 150068792)
)
regions$region_id <- paste(regions$chr, regions$start, regions$end, sep = "_")
regions
```


## Testing functions

### 1. assemble_region_m6
```{r assemble_region_m6}
# Pick one region to test
i <- 1

region_id    <- regions$region_id[i]
region_chr   <- regions$chr[i]
region_start <- regions$start[i]
region_end   <- regions$end[i]

cat(
  "Testing:", region_id, "\n",
  "Region:", region_chr, region_start, region_end, "\n"
)

dat <- assemble_region_m6a(
  sample_names = sample_names,
  region_chr   = region_chr,
  region_start = region_start,
  region_end   = region_end,
  ft_result_dir = ft_result_dir,
  verbose = TRUE
)

head(dat$reads)
dim(dat$reads)
colnames(dat$reads)
```

### 2. bin_read_matrix

```{r bin_read_matrix}
binned50 <- bin_read_matrix(
  met_mat      = dat$met_mat,
  region_start = region_start,
  region_end   = region_end,
  window_size  = window_size
)

names(binned50)

head(binned50)
```


### 3. manhattan_knn_graph

```{r manhattan_knn_graph}
feat <- binned50$mat

knn <- manhattan_knn_graph(
  mat         = feat,
  k_neighbors = k_neighbors,   # 10
  sigma       = kernel_sigma,  
  verbose     = TRUE
)

knn$dist[1:5, 1:5]

edge_df <- igraph::as_data_frame(knn$graph, what = "edges")
head(edge_df)
```

### 4. leiden_partition
```{r leiden_partition}
part <- leiden_partition(
  knn$graph,
  resolution = leiden_resolution,
  seed = leiden_seed
)

# Leiden cluster for each read
membership <- part$membership

table(membership)

set.seed(1)

plot(
  knn$graph,
  vertex.color = membership,
  vertex.size = 4,
  vertex.label = NA,
  edge.width = 0.3,
  layout = layout_with_fr(knn$graph),
  main = "KNN graph colored by Leiden cluster"
)
```



# Parameter sweep: KNN graph colored by Leiden cluster

For each region and bin size, the KNN graph is built once per k (with a single
Fruchterman-Reingold layout per graph, so the plots for different resolutions
differ only in the coloring) and Leiden is re-run across the resolutions. One
pdf per k x resolution combination is saved in the region's `bin<size>` folder.

```{r knn-graph-sweep}
sweep_summary <- list()

for (i in seq_len(nrow(regions))) {
  region_id <- regions$region_id[i]
  dat <- assemble_region_m6a(
    sample_names = sample_names,
    region_chr   = regions$chr[i],
    region_start = regions$start[i],
    region_end   = regions$end[i],
    ft_result_dir = ft_result_dir,
    verbose = FALSE
  )
  met_mat_sweep <- filter_met_mat(dat$met_mat, verbose = FALSE)

  for (ws in sweep_window_sizes) {
    binned <- bin_read_matrix(met_mat_sweep, regions$start[i], regions$end[i], ws)
    bin_dir <- file.path(output_dir, region_id, paste0("bin", ws))
    if (!dir.exists(bin_dir)) dir.create(bin_dir, recursive = TRUE)

    for (k in sweep_k_neighbors) {
      knn <- manhattan_knn_graph(binned$mat, k_neighbors = k,
                                 sigma = kernel_sigma, verbose = FALSE)
      set.seed(1)
      lay <- layout_with_fr(knn$graph)

      for (r in sweep_resolutions) {
        part <- leiden_partition(knn$graph, resolution = r, seed = leiden_seed)
        membership <- part$membership
        n_cl <- length(unique(membership))

        pdf(file.path(bin_dir, sprintf("knn_graph_leiden_k%d_res%g_%s.pdf",
                                       k, r, region_id)),
            width = 7, height = 7)
        plot(knn$graph,
             vertex.color = membership,
             vertex.size = 4,
             vertex.label = NA,
             edge.width = 0.3,
             layout = lay,
             main = sprintf("%s bin%d k=%d res=%g: %d clusters",
                            region_id, ws, k, r, n_cl))
        dev.off()

        sweep_summary[[length(sweep_summary) + 1]] <- data.frame(
          region_id = region_id, bin = ws, k_neighbors = k, resolution = r,
          n_reads = nrow(binned$mat), n_clusters = n_cl)
      }
    }
  }
}

sweep_summary <- dplyr::bind_rows(sweep_summary)
write.table(sweep_summary, file.path(output_dir, "knn_graph_sweep_summary.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
sweep_summary %>%
  tidyr::pivot_wider(names_from = resolution, values_from = n_clusters,
                     names_prefix = "res") %>%
  as.data.frame()
```

# Clustering at each region

Run the same Leiden + Manhattan workflow on the explicit regions above, keeping
the same single `window_size` throughout. Outputs per region are written under
`<region_id>/bin<size>/k<k>/`.

```{r cluster-regions, fig.width=9, fig.height=9}
res <- leiden_manhattan_cluster_regions(
  sample_names = sample_names,
  regions = regions,
  ft_result_dir = ft_result_dir,
  window_size = window_size
)

for (region_id in names(res)) {
  region_res <- res[[region_id]]
  lab <- paste0("bin", window_size)
  k_dir <- file.path(output_dir, region_id, lab, paste0("k", region_res$n_clusters))
  if (!dir.exists(k_dir)) dir.create(k_dir, recursive = TRUE)

  region_res$assignments$timepoint <- sample_table$timepoint[
    match(region_res$assignments$sample_name, sample_table$sample_name)
  ]
  res[[region_id]] <- region_res

  saveRDS(region_res, file.path(k_dir, paste0("leiden_manhattan_", region_id, ".rds")))
  write.table(region_res$assignments,
              file.path(k_dir, paste0("read_cluster_assignments_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(cluster = rownames(region_res$profiles), region_res$profiles,
                         check.names = FALSE),
              file.path(k_dir, paste0("cluster_profiles_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(region_res$window_anno,
              file.path(k_dir, paste0("window_annotation_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(RID = rownames(region_res$feat_mat),
                         sample_name = region_res$assignments$sample_name,
                         cluster = region_res$assignments$cluster,
                         region_res$feat_mat, check.names = FALSE),
              file.path(k_dir, paste0("feature_matrix_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(region_res$assignments %>% dplyr::count(cluster, sample_name, timepoint),
              file.path(k_dir, paste0("cluster_timepoint_counts_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(parameter = names(region_res$params),
                         value = vapply(region_res$params, as.character, character(1))),
              file.path(k_dir, paste0("run_parameters_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)

  gg <- plot_cluster_met_profiles(region_res, region_res$site_met_mat, main = region_id)
  ggsave(file.path(k_dir, paste0("cluster_met_profiles_", region_id, ".pdf")), gg,
         width = 6, height = 1.4 * region_res$n_clusters + 1.4, limitsize = FALSE)

  gg <- plot_cluster_composition(region_res, main = region_id)
  ggsave(file.path(k_dir, paste0("cluster_composition_", region_id, ".pdf")), gg,
         width = 7, height = 8)
  gg <- plot_cluster_structure(region_res, main = region_id)
  ggsave(file.path(k_dir, paste0("cluster_structure_", region_id, ".pdf")), gg,
         width = 9, height = 3)

  pdf(file.path(k_dir, paste0("heatmap_", region_id, ".pdf")), width = 8, height = 9)
  ComplexHeatmap::draw(plot_cluster_heatmap(region_res, main = region_id))
  dev.off()

  gg <- plot_met_fraction_lines(region_res, main = region_id, smooth_k = line_smooth_k)
  ggsave(file.path(k_dir, paste0("met_fraction_lines_", region_id, ".pdf")), gg,
         width = 8, height = 1.4 * region_res$n_clusters + 1, limitsize = FALSE)
}

names(res)
```

# Cluster sizes and timepoint composition across regions

`n_clusters` is the number of Leiden clusters for that region at the current
`window_size`; `n` counts reads.

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(res), function(region_id) {
  region_res <- res[[region_id]]
  region_res$assignments %>%
    dplyr::count(cluster) %>%
    dplyr::mutate(region_id = region_id,
                  bins = paste0("bin", window_size),
                  n_clusters = region_res$n_clusters,
                  .before = 1)
}))
write.table(cluster_summary, file.path(output_dir, "cluster_summary_all_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = cluster, values_from = n)

timepoint_summary <- dplyr::bind_rows(lapply(names(res), function(region_id) {
  region_res <- res[[region_id]]
  region_res$assignments %>%
    dplyr::count(cluster, sample_name, timepoint) %>%
    dplyr::mutate(region_id = region_id,
                  bins = paste0("bin", window_size),
                  .before = 1)
}))
write.table(timepoint_summary,
            file.path(output_dir, "cluster_timepoint_summary_all_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

run_summary <- dplyr::bind_rows(lapply(names(res), function(region_id) {
  p <- res[[region_id]]$params
  data.frame(region_id = region_id, bins = paste0("bin", window_size),
             chr = p$region_chr, start = p$region_start, end = p$region_end,
             n_reads = p$n_reads, n_features = p$n_features, k_eff = p$k_eff,
             resolution = p$resolution, sigma = p$sigma,
             mean_knn_dist = p$mean_knn_dist, quality = p$quality,
             n_clusters = res[[region_id]]$n_clusters)
}))
write.table(run_summary, file.path(output_dir, "run_summary_all_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
run_summary
```

# Session info

```{r session-info}
sessionInfo()
```
