---
title: "Leiden + Manhattan clustering of single molecules at early-response promoters (TSS +/- 150 bp)"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(data.table)
library(Rsamtools)   
library(igraph)
library(ComplexHeatmap)
library(cowplot)
```

# Functions

`extract_ft_region_reads()`, `extract_ft_read_info()`, `get_sparse_met_mat()`
and `filter_met_mat()` come from the topic model

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

promoter_table <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical/promoter_regions_gencode_v49.tsv"

output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/Leiden_Manhattan/TSS_300bp"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Promoter half-width: the 2-kb run used the table's TSS +/- 1000 bp columns;
# here the window is rebuilt as TSS +/- 150 bp (a 300-bp region).
flank <- 150L

# Window ("feature") size sweep: one full clustering run per promoter per
# entry. Positive = window size in bp (the paper: 50); 0 = no windowing, i.e.
# each m6A site is its own feature (the default, and the first entry here).
# Output folders: <gene>/bin<size>/k<k>/, k = the number of Leiden clusters.
window_sizes <- c(0, 10, 20, 30, 50)

# Leiden / graph parameters (the paper's steps 5-7).
# k_neighbors: the paper's k = 50 would connect ~25% of the read set at these
# promoters (~150-250 full-span reads each) instead of the tiny fraction it
# connects in the paper, giving a densely connected graph and coarse clusters,
# so k = 10 is used here.
# leiden_resolution: the paper used 2.1-3, but on read sets orders of magnitude
# larger, where those values shatter a graph this size; 1 is standard
# modularity. Raise it for finer clusters, lower it for coarser ones.
k_neighbors       <- 10
leiden_resolution <- 1
leiden_seed       <- 1

# Exponential-kernel width for the affinities: NULL = mean of the retained KNN
# distances (scale-free across window sizes). Set to the number of features to
# reproduce scikit-learn's laplacian_kernel default (gamma = 1 / n_features).
kernel_sigma <- NULL

# Paper step 2, off for now: with full-span reads there is nothing to impute.
impute_missing <- FALSE
impute_k       <- 5

# Rolling-mean width for the m6A fraction line plot only (1 = raw profiles).
# At window size 0 the per-site cluster means are spiky; e.g. 25 makes the
# nucleosome-scale structure readable without changing any clustering input.
line_smooth_k <- 1

set.seed(123)
```

# Promoter regions

Same table and genes as the 2-kb run (GENCODE v49 TSS, `make_promoter_table.R`),
with the window narrowed to TSS +/- `r flank` bp.

```{r regions}
promoters <- read.delim(promoter_table)
promoters$start <- promoters$tss - flank
promoters$end   <- promoters$tss + flank
promoters
```


## Testing functions

### 1. assemble_region_m6
```{r assemble_region_m6}
# Pick one promoter to test
i <- 1

gene         <- promoters$gene[i]
region_chr   <- promoters$chr[i]
region_start <- promoters$start[i]
region_end   <- promoters$end[i]

cat(
  "Testing:", gene, "\n",
  "Region:", region_chr, region_start, region_end, "\n"
)

dat <- assemble_region_m6a(
  sample_names = sample_names,
  region_chr   = region_chr,
  region_start = region_start,
  region_end   = region_end,
  ft_result_dir = ft_result_dir,
  verbose = TRUE
)

head(dat$reads)
dim(dat$reads)
colnames(dat$reads)
```

### 2. bin_read_matrix

```{r bin_read_matrix}
binned50 <- bin_read_matrix(
  met_mat      = dat$met_mat,
  region_start = region_start,
  region_end   = region_end,
  window_size  = 10
)

names(binned50)

head(binned50)
```


### 3. manhattan_knn_graph

```{r manhattan_knn_graph}
feat <- binned50$mat

knn <- manhattan_knn_graph(
  mat         = feat,
  k_neighbors = k_neighbors,   # 10
  sigma       = kernel_sigma,  
  verbose     = TRUE
)

knn$dist[1:5, 1:5]

edge_df <- igraph::as_data_frame(knn$graph, what = "edges")
head(edge_df)
```

### 4. leiden_partition
```{r leiden_partition}
part <- leiden_partition(
  knn$graph,
  resolution = leiden_resolution,
  seed = leiden_seed
)

# Leiden cluster for each read
membership <- part$membership

table(membership)

set.seed(1)

plot(
  knn$graph,
  vertex.color = membership,
  vertex.size = 4,
  vertex.label = NA,
  edge.width = 0.3,
  layout = layout_with_fr(knn$graph),
  main = "KNN graph colored by Leiden cluster"
)
```



# Clustering at each promoter

Per gene and per `window_sizes` entry: assemble the pooled m6A matrix once per
gene, apply the full-span filter, build the read x feature matrix (window size
0 uses the m6A site matrix directly), then run the paper's Manhattan / KNN /
exponential-kernel / Leiden pipeline. Outputs per gene x window size, under
`<gene>/bin<size>/k<k>/`: the result object (.rds), per-read cluster
assignments **with the read's timepoint**, cluster profiles, the feature
matrix, the window annotation, cluster x timepoint counts, the run parameters
(.tsv), and the five plots.

```{r cluster-promoters, fig.width=9, fig.height=9}
all_results <- list()

# loop through every promoter and extract m6a
for (i in seq_len(nrow(promoters))) {
  gene <- promoters$gene[i]
  cat("\n=====", gene, "=====\n")

  dat <- assemble_region_m6a(sample_names, promoters$chr[i],
                             promoters$start[i], promoters$end[i],
                             ft_result_dir, verbose = FALSE)
  cat(gene, ": ", nrow(dat$rids_df), " reads x ", ncol(dat$met_mat),
      " m6A sites before filtering\n", sep = "")
  
  # keep fully spanning reads
  met_mat <- filter_met_mat(dat$met_mat, verbose = TRUE)
  rids_df <- dat$rids_df[match(rownames(met_mat), as.character(dat$rids_df$RID)), ]
  cat(gene, ": ", nrow(met_mat), " full-span reads x ", ncol(met_mat),
      " m6A sites\n", sep = "")

  # bin m6a into different window sizes
  for (ws in window_sizes) {
    lab      <- paste0("bin", ws)
    win_desc <- if (ws == 0) "per-site, no windows" else paste0(ws, "-bp windows")
    gene_lab <- paste0(gene, " (", win_desc, ")")
    cat("--- ", gene, ", ", win_desc, " ---\n", sep = "")

    # run Leiden clustering
    res <- leiden_manhattan_cluster(met_mat, rids_df,
                                    region_start = promoters$start[i],
                                    region_end   = promoters$end[i],
                                    window_size  = ws,
                                    k_neighbors  = k_neighbors,
                                    resolution   = leiden_resolution,
                                    impute_missing = impute_missing,
                                    impute_k     = impute_k,
                                    sigma        = kernel_sigma,
                                    seed         = leiden_seed)

    # every read keeps its timepoint alongside its cluster
    res$assignments$sample_name <- factor(res$assignments$sample_name,
                                          levels = sample_names)
    res$assignments$timepoint <- sample_table$timepoint[
      match(res$assignments$sample_name, sample_table$sample_name)]
    all_results[[gene]][[lab]] <- res

    # k subfolder = the number of Leiden clusters this run produced
    k_dir <- file.path(output_dir, gene, lab, paste0("k", res$n_clusters))
    if (!dir.exists(k_dir)) dir.create(k_dir, recursive = TRUE)

    ## tables + rds
    saveRDS(res, file.path(k_dir, paste0("leiden_manhattan_", gene, ".rds")))
    write.table(res$assignments,
                file.path(k_dir, paste0("read_cluster_assignments_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(cluster = rownames(res$profiles), res$profiles,
                           check.names = FALSE),
                file.path(k_dir, paste0("cluster_profiles_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$window_anno,
                file.path(k_dir, paste0("window_annotation_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(RID = rownames(res$feat_mat),
                           sample_name = res$assignments$sample_name,
                           cluster = res$assignments$cluster,
                           res$feat_mat, check.names = FALSE),
                file.path(k_dir, paste0("feature_matrix_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$assignments %>% dplyr::count(cluster, sample_name, timepoint),
                file.path(k_dir, paste0("cluster_timepoint_counts_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(parameter = names(res$params),
                           value = vapply(res$params, as.character, character(1))),
                file.path(k_dir, paste0("run_parameters_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)

    ## 1. per-cluster m6A methylation profiles (topic-model style)
    gg <- plot_cluster_met_profiles(res, met_mat, tss = promoters$tss[i],
                                    main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_met_profiles_", gene, ".pdf")), gg,
           width = 6, height = 1.4 * res$n_clusters + 1.4, limitsize = FALSE)

    ## 2. cluster composition by timepoint, and cluster structure
    gg <- plot_cluster_composition(res, main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_composition_", gene, ".pdf")), gg,
           width = 7, height = 8)
    gg <- plot_cluster_structure(res, main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_structure_", gene, ".pdf")), gg,
           width = 9, height = 3)

    ## 3. read x feature methylation heatmap
    pdf(file.path(k_dir, paste0("heatmap_", gene, ".pdf")), width = 8, height = 9)
    ComplexHeatmap::draw(plot_cluster_heatmap(res, main = gene_lab))
    dev.off()

    ## 4. m6A fraction line plot (one panel per cluster)
    gg <- plot_met_fraction_lines(res, tss = promoters$tss[i], main = gene_lab,
                                  smooth_k = line_smooth_k)
    ggsave(file.path(k_dir, paste0("met_fraction_lines_", gene, ".pdf")), gg,
           width = 8, height = 1.4 * res$n_clusters + 1, limitsize = FALSE)
  }
}
```

# Cluster sizes and timepoint composition across genes

`n_clusters` is the number of Leiden clusters for that gene x window size at
`leiden_resolution`; `n` counts reads.

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    all_results[[g]][[lab]]$assignments %>%
      dplyr::count(cluster) %>%
      dplyr::mutate(gene = g, bins = lab,
                    n_clusters = all_results[[g]][[lab]]$n_clusters,
                    .before = 1)
  }))
}))
write.table(cluster_summary, file.path(output_dir, "cluster_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = cluster, values_from = n)

timepoint_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    all_results[[g]][[lab]]$assignments %>%
      dplyr::count(cluster, sample_name, timepoint) %>%
      dplyr::mutate(gene = g, bins = lab, .before = 1)
  }))
}))
write.table(timepoint_summary,
            file.path(output_dir, "cluster_timepoint_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

## run parameters per gene x window size (resolution, kernel width, graph size)
run_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    p <- all_results[[g]][[lab]]$params
    data.frame(gene = g, bins = lab, n_reads = p$n_reads,
               n_features = p$n_features, k_eff = p$k_eff,
               resolution = p$resolution, sigma = p$sigma,
               mean_knn_dist = p$mean_knn_dist, quality = p$quality,
               n_clusters = all_results[[g]][[lab]]$n_clusters)
  }))
}))
write.table(run_summary, file.path(output_dir, "run_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
run_summary
```

# Session info

```{r session-info}
sessionInfo()
```
