---
title: "Leiden + Manhattan clustering of single molecules at early-response promoters"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

Replicates the single-molecule read clustering of Raviram, Jiang, Schippke,
Cova & Skok (2026), "Cohesin collisions maintain ordered nucleosome
architecture at boundaries and promoters" (bioRxiv 2026.05.22.727261, Methods /
"Clustering reads"), on THP-1 Fiber-seq m6A at the early-response promoters
instead of nano-NOMe GpC (GCH) accessibility at CTCF sites and TSSs. The
procedure is the paper's:

1. per-read accessibility binned into windows across the 2-kb region, each
   window summarised by the mean methylation call;
2. remaining missing bins filled by KNN imputation (k = 5);
3. conditions balanced by sampling an equal number of reads per condition,
   then pooled;
4. read-read similarity = Manhattan distance over the bins,
   `d(i,j) = sum_b |x_ib - x_jb|`;
5. a KNN graph over reads (k = 50 nearest neighbours);
6. Manhattan distances turned into edge weights ("affinities") by an
   exponential kernel;
7. Leiden community detection (`RBConfigurationVertexPartition`) at a set
   resolution, each cluster a recurring single-molecule accessibility pattern;
8. cluster profiles = the mean accessibility profile of each cluster's reads.

Deviations from the paper (all switches in the Settings chunk):

* Data type: Fiber-seq m6A (1 = m6A / accessible, 0 = covered but unmodified)
  instead of nano-NOMe GpC methylation.
* Molecules: the full-span read set (`filter_met_mat()`, the topic model's row
  filter, also used by the hamming and spearman clusterings), so all four
  clusterings share one read universe.
* No condition balancing (step 3, `balance_conditions` is not implemented -
  the paper subsamples each condition to equal depth). All four LPS timepoints
  are pooled and clustered jointly, as in the topic model, at their observed
  depths. Every read keeps its timepoint (`sample_name` / `timepoint`) next to
  its cluster in `read_cluster_assignments_<gene>.tsv`, so for any read both
  its cluster and its timepoint are recoverable, and the paper's
  condition-composition plots become timepoint-composition plots.
* Window-size sweep (`window_sizes`) instead of the paper's fixed 50 bp: each
  promoter is clustered once per entry - a window size in bp, or 0 for no
  windowing at all, i.e. each m6A site is its own feature. 0 is the default and
  the first entry of the sweep. Outputs go to `<gene>/bin<size>/k<k>/`, where
  k is the number of Leiden clusters that run produced.
* KNN imputation (step 2, `impute_missing`) is optional and **off** for now.
  With full-span reads it is a no-op: every read covers every m6A site, so the
  only missing bins are windows containing no m6A site at all, and those are
  dropped as features (they inform no read) rather than imputed.
* No meta-clustering: the paper groups Leiden clusters into meta-clusters by
  FFT + hierarchical clustering for display; here the Leiden clusters are used
  directly. Clusters are renumbered largest-first so the labels are
  deterministic.
* Graph and resolution (`k_neighbors`, `leiden_resolution`): the paper's
  k = 50 neighbours and resolution 2.1-3 are set on read sets orders of
  magnitude larger than these promoters (~150-250 full-span reads each), where
  k = 50 connects a quarter of all reads and those resolutions shatter the
  graph into dozens of tiny clusters. Here k = 10 and resolution 1 (standard
  modularity); both are the knobs to turn if the clusters look too coarse or
  too fine.

See `leiden_manhattan_functions.r` for the step-by-step implementation
(including the two details the Methods leave open - the kernel width and the
symmetrisation of the KNN graph) and `leiden_manhattan_plots.r` for the plots.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
# igraph needs libglpk at load time; if library(igraph) fails with
# "libglpk.so.40: cannot open shared object file", add glpk's lib directory
# (/software/glpk-5.0-el8-x86_64/lib on this cluster) to LD_LIBRARY_PATH before
# starting R.
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(data.table)
library(Rsamtools)   # extract_ft_region_reads() reads the tabix-indexed beds
library(igraph)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
```

# Functions

`extract_ft_region_reads()`, `extract_ft_read_info()`, `get_sparse_met_mat()`
and `filter_met_mat()` come from the topic model, as in the hamming and
spearman analyses.

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

promoter_table <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical/promoter_regions_gencode_v49.tsv"

output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/Leiden_Manhattan"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Window ("feature") size sweep: one full clustering run per promoter per
# entry. Positive = window size in bp (the paper: 50); 0 = no windowing, i.e.
# each m6A site is its own feature (the default, and the first entry here).
# Output folders: <gene>/bin<size>/k<k>/, k = the number of Leiden clusters.
window_sizes <- c(0, 10, 20, 30, 50)

# Leiden / graph parameters (the paper's steps 5-7).
# k_neighbors: the paper's k = 50 would connect ~25% of the read set at these
# promoters (~150-250 full-span reads each) instead of the tiny fraction it
# connects in the paper, giving a densely connected graph and coarse clusters,
# so k = 10 is used here.
# leiden_resolution: the paper used 2.1-3, but on read sets orders of magnitude
# larger, where those values shatter a graph this size; 1 is standard
# modularity. Raise it for finer clusters, lower it for coarser ones.
k_neighbors       <- 10
leiden_resolution <- 1
leiden_seed       <- 1

# Exponential-kernel width for the affinities: NULL = mean of the retained KNN
# distances (scale-free across window sizes). Set to the number of features to
# reproduce scikit-learn's laplacian_kernel default (gamma = 1 / n_features).
kernel_sigma <- NULL

# Paper step 2, off for now: with full-span reads there is nothing to impute.
impute_missing <- FALSE
impute_k       <- 5

# Rolling-mean width for the m6A fraction line plot only (1 = raw profiles).
# At window size 0 the per-site cluster means are spiky; e.g. 25 makes the
# nucleosome-scale structure readable without changing any clustering input.
line_smooth_k <- 1

set.seed(123)
```

# Promoter regions

Same table and genes as the hamming and spearman analyses: TSS +/- 1000 bp
from GENCODE v49 (`make_promoter_table.R`).

```{r regions}
promoters <- read.delim(promoter_table)
promoters
```

# Clustering at each promoter

Per gene and per `window_sizes` entry: assemble the pooled m6A matrix once per
gene, apply the full-span filter, build the read x feature matrix (window size
0 uses the m6A site matrix directly), then run the paper's Manhattan / KNN /
exponential-kernel / Leiden pipeline. Outputs per gene x window size, under
`<gene>/bin<size>/k<k>/`: the result object (.rds), per-read cluster
assignments **with the read's timepoint**, cluster profiles, the feature
matrix, the window annotation, cluster x timepoint counts, the run parameters
(.tsv), and the five plots.

```{r cluster-promoters, fig.width=9, fig.height=9}
all_results <- list()
for (i in seq_len(nrow(promoters))) {
  gene <- promoters$gene[i]
  cat("\n=====", gene, "=====\n")

  dat <- assemble_region_m6a(sample_names, promoters$chr[i],
                             promoters$start[i], promoters$end[i],
                             ft_result_dir, verbose = FALSE)
  cat(gene, ": ", nrow(dat$rids_df), " reads x ", ncol(dat$met_mat),
      " m6A sites before filtering\n", sep = "")

  met_mat <- filter_met_mat(dat$met_mat, verbose = TRUE)
  rids_df <- dat$rids_df[match(rownames(met_mat), as.character(dat$rids_df$RID)), ]
  cat(gene, ": ", nrow(met_mat), " full-span reads x ", ncol(met_mat),
      " m6A sites\n", sep = "")

  for (ws in window_sizes) {
    lab      <- paste0("bin", ws)
    win_desc <- if (ws == 0) "per-site, no windows" else paste0(ws, "-bp windows")
    gene_lab <- paste0(gene, " (", win_desc, ")")
    cat("--- ", gene, ", ", win_desc, " ---\n", sep = "")

    res <- leiden_manhattan_cluster(met_mat, rids_df,
                                    region_start = promoters$start[i],
                                    region_end   = promoters$end[i],
                                    window_size  = ws,
                                    k_neighbors  = k_neighbors,
                                    resolution   = leiden_resolution,
                                    impute_missing = impute_missing,
                                    impute_k     = impute_k,
                                    sigma        = kernel_sigma,
                                    seed         = leiden_seed)

    # every read keeps its timepoint alongside its cluster
    res$assignments$sample_name <- factor(res$assignments$sample_name,
                                          levels = sample_names)
    res$assignments$timepoint <- sample_table$timepoint[
      match(res$assignments$sample_name, sample_table$sample_name)]
    all_results[[gene]][[lab]] <- res

    # k subfolder = the number of Leiden clusters this run produced
    k_dir <- file.path(output_dir, gene, lab, paste0("k", res$n_clusters))
    if (!dir.exists(k_dir)) dir.create(k_dir, recursive = TRUE)

    ## tables + rds
    saveRDS(res, file.path(k_dir, paste0("leiden_manhattan_", gene, ".rds")))
    write.table(res$assignments,
                file.path(k_dir, paste0("read_cluster_assignments_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(cluster = rownames(res$profiles), res$profiles,
                           check.names = FALSE),
                file.path(k_dir, paste0("cluster_profiles_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$window_anno,
                file.path(k_dir, paste0("window_annotation_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(RID = rownames(res$feat_mat),
                           sample_name = res$assignments$sample_name,
                           cluster = res$assignments$cluster,
                           res$feat_mat, check.names = FALSE),
                file.path(k_dir, paste0("feature_matrix_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$assignments %>% dplyr::count(cluster, sample_name, timepoint),
                file.path(k_dir, paste0("cluster_timepoint_counts_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(parameter = names(res$params),
                           value = vapply(res$params, as.character, character(1))),
                file.path(k_dir, paste0("run_parameters_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)

    ## 1. per-cluster m6A methylation profiles (topic-model style)
    gg <- plot_cluster_met_profiles(res, met_mat, tss = promoters$tss[i],
                                    main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_met_profiles_", gene, ".pdf")), gg,
           width = 6, height = 1.4 * res$n_clusters + 1.4, limitsize = FALSE)

    ## 2. cluster composition by timepoint, and cluster structure
    gg <- plot_cluster_composition(res, main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_composition_", gene, ".pdf")), gg,
           width = 7, height = 8)
    gg <- plot_cluster_structure(res, main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_structure_", gene, ".pdf")), gg,
           width = 9, height = 3)

    ## 3. read x feature methylation heatmap
    pdf(file.path(k_dir, paste0("heatmap_", gene, ".pdf")), width = 8, height = 9)
    ComplexHeatmap::draw(plot_cluster_heatmap(res, main = gene_lab))
    dev.off()

    ## 4. m6A fraction line plot
    gg <- plot_met_fraction_lines(res, tss = promoters$tss[i], main = gene_lab,
                                  smooth_k = line_smooth_k)
    ggsave(file.path(k_dir, paste0("met_fraction_lines_", gene, ".pdf")), gg,
           width = 8, height = 4)
  }
}
```

# Cluster sizes and timepoint composition across genes

`n_clusters` is the number of Leiden clusters for that gene x window size at
`leiden_resolution`; `n` counts reads.

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    all_results[[g]][[lab]]$assignments %>%
      dplyr::count(cluster) %>%
      dplyr::mutate(gene = g, bins = lab,
                    n_clusters = all_results[[g]][[lab]]$n_clusters,
                    .before = 1)
  }))
}))
write.table(cluster_summary, file.path(output_dir, "cluster_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = cluster, values_from = n)

timepoint_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    all_results[[g]][[lab]]$assignments %>%
      dplyr::count(cluster, sample_name, timepoint) %>%
      dplyr::mutate(gene = g, bins = lab, .before = 1)
  }))
}))
write.table(timepoint_summary,
            file.path(output_dir, "cluster_timepoint_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

## run parameters per gene x window size (resolution, kernel width, graph size)
run_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    p <- all_results[[g]][[lab]]$params
    data.frame(gene = g, bins = lab, n_reads = p$n_reads,
               n_features = p$n_features, k_eff = p$k_eff,
               resolution = p$resolution, sigma = p$sigma,
               mean_knn_dist = p$mean_knn_dist, quality = p$quality,
               n_clusters = all_results[[g]][[lab]]$n_clusters)
  }))
}))
write.table(run_summary, file.path(output_dir, "run_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
run_summary
```

# Session info

```{r session-info}
sessionInfo()
```
