---
title: "vis_fire_freq"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggpubr)
library(ggrepel)
library(tidyverse)
library(DT)
```

# FIRE frequency

chrom start   end        region_id timepoint n_reads n_fire     freq
1  chr1 11261 11486 chr1:11261-11486     LPS_0      12      2 0.166667
2  chr1 21178 21359 chr1:21178-21359     LPS_0      72      9 0.125000
3  chr1 28997 29166 chr1:28997-29166     LPS_0      26     12 0.461538
4  chr1 29283 29478 chr1:29283-29478     LPS_0      25     23 0.920000
5  chr1 29964 30121 chr1:29964-30121     LPS_0      24      6 0.250000
6  chr1 38601 38804 chr1:38601-38804     LPS_0      19      5 0.263158

region represents universe FIRE regions across all timepoints
```{r load_data}
input_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency"
plot_dir  <- file.path(input_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE)

fire_freq <- read.csv(file.path(input_dir, "fire_frequency_long.tsv.gz"), sep = "\t")
head(fire_freq)

# group by region_id and timepoint, and calculate the mean frequency
fire_freq_summary <- fire_freq %>%
  group_by(region_id, timepoint) %>%
  summarise(mean_freq = mean(freq, na.rm = TRUE)) %>%
  ungroup()


# rank region by mean frequency for each timepoint
fire_freq_summary <- fire_freq_summary %>%
  group_by(timepoint) %>%
  mutate(rank = rank(-mean_freq, ties.method = "first")) %>%
  ungroup()

head(fire_freq_summary)

```


# cumulative distribution of FIRE frequency for each timepoint
for a given frequency x, what fraction of FIRE-universe regions have frequency ≤ x?

Left-shifted curve = more regions accumulate at lower x values → overall lower FIRE frequencies
```{r CDF_fire_freq}

fire_freq_cdf <- fire_freq_summary %>%
  filter(!is.na(mean_freq)) %>%
  group_by(timepoint) %>%
  arrange(mean_freq, .by_group = TRUE) %>%
  mutate(
    cdf = row_number() / n()
  ) %>%
  ungroup()

head(fire_freq_cdf)



p_freq_cdf <- ggplot(fire_freq_cdf, aes(x = mean_freq, color = timepoint)) +
  stat_ecdf(na.rm = TRUE) +
  labs(
    x = "FIRE frequency",
    y = "Cumulative fraction of regions",
    color = "Timepoint"
  ) +
  theme_classic()
p_freq_cdf
ggsave(file.path(plot_dir, "fire_freq_cdf.pdf"), p_freq_cdf, width = 7, height = 4)
```


# Pairwise differential FIRE frequency: top 100 regions

Top 100 regions per comparison, ranked by Fisher exact p-value among
coverage-passing regions (6 comparisons x 100 rows).

```{r load_pairwise}
pairwise <- read_tsv(file.path(input_dir, "pairwise_fisher.tsv.gz"),
                     show_col_types = FALSE)

comparison_levels <- c("LPS_0_vs_LPS_5", "LPS_0_vs_LPS_10", "LPS_0_vs_LPS_15",
                       "LPS_5_vs_LPS_10", "LPS_5_vs_LPS_15", "LPS_10_vs_LPS_15")
pairwise <- pairwise %>%
  mutate(comparison = factor(comparison, levels = comparison_levels))

# only coverage-passing regions were ranked in 02_pairwise_differential.R
passing <- pairwise %>% filter(pass_coverage)

# top 100 per comparison by raw p
top100 <- passing %>%
  group_by(comparison) %>%
  arrange(p_value, .by_group = TRUE) %>%
  slice_head(n = 100) %>%
  ungroup() %>%
  mutate(direction = ifelse(delta_freq > 0, "gain", "loss"))

# gain = higher FIRE frequency at the later timepoint
table(top100$comparison, top100$direction)
```


# Regions tested vs significant

Tested = coverage-passing regions in that comparison; significant = unadjusted
Fisher p < 0.05. "any comparison" is the union across the 6 comparisons, i.e.
the definition used in the cCRE section below.

```{r tested_vs_signif_bar, fig.width=8, fig.height=4.5}
p_sig <- 0.05

counts_by_comparison <- passing %>%
  group_by(comparison) %>%
  summarise(tested      = n(),
            significant = sum(p_value < p_sig),
            .groups = "drop") %>%
  mutate(group = as.character(comparison))

counts_union <- tibble(
  group       = "any comparison",
  tested      = n_distinct(passing$region_id),
  significant = n_distinct(passing$region_id[passing$p_value < p_sig]))

region_counts <- bind_rows(counts_by_comparison %>% select(group, tested, significant),
                           counts_union) %>%
  mutate(group = factor(group, levels = c(comparison_levels, "any comparison")))

region_counts_long <- region_counts %>%
  pivot_longer(c(tested, significant), names_to = "category", values_to = "n") %>%
  mutate(category = factor(category, levels = c("tested", "significant")))

p_region_counts <- ggplot(region_counts_long, aes(x = group, y = n, fill = category)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7,
           color = "grey30", linewidth = 0.2) +
  geom_text(aes(label = scales::comma(n)), position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c(tested = "grey70", significant = "firebrick")) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "FIRE regions", fill = NULL,
       subtitle = paste0("FIRE universe: ", n_distinct(pairwise$region_id),
                         " regions; significant = p < ", p_sig)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_region_counts
ggsave(file.path(plot_dir, "regions_tested_vs_signif.pdf"), p_region_counts,
       width = 8, height = 4.5)

region_counts %>% mutate(frac_signif = round(significant / tested, 3))
```


# P-value distribution per comparison

Histogram of the unadjusted Fisher p-values 

```{r pvalue_hist, fig.width=12, fig.height=6}
expected_bin <- passing %>%
  count(comparison) %>%
  mutate(expected = n * 0.025)

p_pval_hist <- ggplot(passing, aes(x = p_value)) +
  geom_histogram(binwidth = 0.025, boundary = 0,
                 fill = "grey70", color = "grey30", linewidth = 0.2) +
  geom_hline(data = expected_bin, aes(yintercept = expected),
             linetype = "dashed", color = "red") +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = "Fisher exact p (unadjusted)", y = "Regions") +
  theme_classic()
p_pval_hist
ggsave(file.path(plot_dir, "pvalue_histogram.pdf"), p_pval_hist, width = 12, height = 6)
```


# Volcano: effect size vs significance

Effect size = delta_freq (later - earlier FIRE frequency); significance =
-log10 of the unadjusted Fisher p. Coverage-passing regions only. Points are
coloured by direction among the regions with p < 0.05.

```{r volcano_all, fig.width=12, fig.height=6}
volcano <- passing %>%
  mutate(neglog10_p = -log10(p_value),
         status = case_when(p_value >= p_sig  ~ "n.s.",
                            delta_freq > 0    ~ "gain (up in later)",
                            TRUE              ~ "loss (down in later)"),
         status = factor(status, levels = c("n.s.", "gain (up in later)",
                                            "loss (down in later)")))

status_cols <- c("n.s." = "grey75",
                 "gain (up in later)" = "firebrick",
                 "loss (down in later)" = "steelblue")

p_volcano <- ggplot(volcano, aes(x = delta_freq, y = neglog10_p, color = status)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_hline(yintercept = -log10(p_sig), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~ comparison, nrow = 2) +
  scale_color_manual(values = status_cols) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = expression(Delta ~ "FIRE frequency (later - earlier)"),
       y = expression(-log[10] ~ "Fisher p (unadjusted)"), color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
p_volcano
ggsave(file.path(plot_dir, "volcano_all_comparisons.pdf"), p_volcano,
       width = 12, height = 6)
```

## LPS_0 vs LPS_15: regions gained and lost at 15 min

Same volcano restricted to the 0 vs 15 min comparison. Labelled points are the
top 10 in each direction: p < 0.05 and the largest |delta_freq|, i.e. the
regions with the highest FIRE frequency at 15 min relative to 0 (gains) and the
lowest at 15 min relative to 0 (losses).

```{r volcano_0_vs_15, fig.width=8.5, fig.height=6}
n_label <- 10

v15 <- volcano %>% filter(comparison == "LPS_0_vs_LPS_15")

top_gain_15 <- v15 %>%
  filter(p_value < p_sig, delta_freq > 0) %>%
  arrange(desc(delta_freq), p_value) %>%
  mutate(rank = row_number(), direction = "gain")

top_loss_15 <- v15 %>%
  filter(p_value < p_sig, delta_freq < 0) %>%
  arrange(delta_freq, p_value) %>%
  mutate(rank = row_number(), direction = "loss")

labelled_15 <- bind_rows(head(top_gain_15, n_label), head(top_loss_15, n_label))

p_volcano_15 <- ggplot(v15, aes(x = delta_freq, y = neglog10_p, color = status)) +
  geom_point(size = 0.5, alpha = 0.4) +
  geom_hline(yintercept = -log10(p_sig), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(data = labelled_15, color = "black", size = 1.4) +
  geom_text_repel(data = labelled_15, aes(label = region_id),
                  color = "black", size = 2.6, min.segment.length = 0,
                  max.overlaps = Inf, box.padding = 0.4) +
  scale_color_manual(values = status_cols) +
  # extra room on both sides so the repelled labels clear the point cloud
  scale_x_continuous(expand = expansion(mult = c(0.35, 0.35))) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = expression(Delta ~ "FIRE frequency (15 min - 0 min)"),
       y = expression(-log[10] ~ "Fisher p (unadjusted)"), color = NULL,
       subtitle = paste0(nrow(top_gain_15), " gains / ", nrow(top_loss_15),
                         " losses at p < ", p_sig,
                         "; top ", n_label, " each labelled")) +
  theme_classic() +
  theme(legend.position = "top")
p_volcano_15
ggsave(file.path(plot_dir, "volcano_LPS_0_vs_LPS_15.pdf"), p_volcano_15,
       width = 8.5, height = 6)
```

```{r top_15_tables}
top_15 <- bind_rows(top_gain_15, top_loss_15)
write_tsv(top_15, file.path(input_dir, "top_changed_LPS_0_vs_LPS_15.tsv"))

# shared column layout for the two direction tables
format_top15 <- function(d) {
  d %>%
    transmute(rank, region_id,
              n_reads_0 = n_reads_A, freq_0 = round(freq_A, 3),
              n_reads_15 = n_reads_B, freq_15 = round(freq_B, 3),
              delta_freq = round(delta_freq, 3), log2_or = round(log2_or, 2),
              p_value = signif(p_value, 3))
}
```

Gains: higher FIRE frequency at 15 min than at 0, ranked by delta_freq.

```{r top_gain_15_table}
datatable(format_top15(top_gain_15), filter = "top", rownames = FALSE,
          caption = paste0("LPS_0 vs LPS_15 gains (p < ", p_sig, "): ",
                           nrow(top_gain_15), " regions"),
          options = list(pageLength = 10))
```

Losses: lower FIRE frequency at 15 min than at 0, ranked by the most negative
delta_freq.

```{r top_loss_15_table}
datatable(format_top15(top_loss_15), filter = "top", rownames = FALSE,
          caption = paste0("LPS_0 vs LPS_15 losses (p < ", p_sig, "): ",
                           nrow(top_loss_15), " regions"),
          options = list(pageLength = 10))
```


# QQ plot: observed vs null p-values

Observed -log10 p against the value expected under the null of no change,
one point per coverage-passing region. The diagonal is the null; points above
it in the upper right are the excess of small p-values that no-effect cannot
account for.

Caveat: the uniform null is only approximate here. Fisher's exact test on small
read counts is discrete and conservative, so even with no real effect the bulk
of the curve is expected to sit slightly *below* the diagonal. Departure in the
tail is still interpretable; mild deflation in the body is not evidence of
anything.

`pi0` is a Storey-style estimate of the fraction of regions with no true change
(from the density of p > 0.5); `excess` is the number of p < 0.05 regions
beyond what that null fraction predicts.

```{r pvalue_qq, fig.width=12, fig.height=6}
qq <- passing %>%
  group_by(comparison) %>%
  arrange(p_value, .by_group = TRUE) %>%
  mutate(observed = -log10(p_value),
         expected = -log10((row_number() - 0.5) / n())) %>%
  ungroup()

pi0_tab <- passing %>%
  group_by(comparison) %>%
  summarise(tested = n(),
            # capped at 1: discreteness piles null p-values up against 1
            pi0    = min(1, mean(p_value > 0.5) / 0.5),
            signif = sum(p_value < p_sig),
            .groups = "drop") %>%
  mutate(expected_null_signif = round(pi0 * tested * p_sig),
         excess               = signif - expected_null_signif)

# all of the tail, a 1-in-20 subsample of the bulk, so the pdf stays small
qq_thin <- qq %>%
  group_by(comparison) %>%
  filter(row_number() <= 5000 | row_number() %% 20 == 0) %>%
  ungroup()

pi0_lab <- pi0_tab %>%
  mutate(label = paste0("pi0 = ", round(pi0, 2),
                        "\nexcess p<", p_sig, ": ", scales::comma(excess)))

p_qq <- ggplot(qq_thin, aes(x = expected, y = observed)) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  geom_point(size = 0.4, alpha = 0.4, color = "grey30") +
  geom_text(data = pi0_lab, aes(x = 0, y = max(qq$observed), label = label),
            hjust = 0, vjust = 1, size = 3, inherit.aes = FALSE) +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = expression("Expected" ~ -log[10] ~ "p (uniform null)"),
       y = expression("Observed" ~ -log[10] ~ "Fisher p")) +
  theme_classic()
p_qq
ggsave(file.path(plot_dir, "pvalue_qq.pdf"), p_qq, width = 12, height = 6)

pi0_tab %>% mutate(pi0 = round(pi0, 3))
```


# Significant regions vs ENCODE cCREs

Significant = unadjusted Fisher p < 0.05 in at least one of the 6 pairwise
comparisons

`ccre_class` / `ccre_id` report that best cCRE.

```{r ccre_overlap}
suppressPackageStartupMessages(library(GenomicRanges))

ccre_file <- "/project/spott/cshan/annotations/GRCh38-cCREs.bed"
p_sig <- 0.05

signif_rows <- passing %>% filter(p_value < p_sig)

# region level: keep the samllest p comparison for ranking
signif_regions <- signif_rows %>%
  group_by(region_id, chrom, start, end) %>%
  summarise(n_signif_comparisons = n(),
            min_p           = min(p_value),
            best_comparison = comparison[which.min(p_value)],
            delta_freq_best = delta_freq[which.min(p_value)],
            .groups = "drop")

ccre <- data.table::fread(ccre_file, header = FALSE,
                          col.names = c("chr", "start", "end",
                                        "ccre_dcc", "ccre_id", "ccre_class"))

reg.gr  <- GRanges(signif_regions$chrom,
                   IRanges(signif_regions$start + 1L, signif_regions$end))
ccre.gr <- GRanges(ccre$chr, IRanges(ccre$start + 1L, ccre$end))
hits <- findOverlaps(reg.gr, ccre.gr)
ov_frac <- width(pintersect(reg.gr[queryHits(hits)], ccre.gr[subjectHits(hits)])) /
  width(reg.gr)[queryHits(hits)]

# best cCRE per region (largest covered fraction)
best <- data.frame(ri = queryHits(hits), frac = ov_frac,
                   ccre_id = ccre$ccre_id[subjectHits(hits)],
                   ccre_class = ccre$ccre_class[subjectHits(hits)]) %>%
  group_by(ri) %>%
  slice_max(frac, n = 1, with_ties = FALSE) %>%
  ungroup()

signif_regions <- signif_regions %>%
  mutate(ccre_frac = 0, ccre_id = NA_character_, ccre_class = NA_character_)
signif_regions$ccre_frac[best$ri]  <- best$frac
signif_regions$ccre_id[best$ri]    <- best$ccre_id
signif_regions$ccre_class[best$ri] <- best$ccre_class

signif_ccre <- signif_regions %>% filter(ccre_frac >= 0.5)

tibble(n_signif_regions = nrow(signif_regions),
       any_ccre_overlap = sum(signif_regions$ccre_frac > 0),
       ccre_50          = nrow(signif_ccre),
       frac_ccre_50     = round(nrow(signif_ccre) / nrow(signif_regions), 3))
```

Nearest gencode v46 gene for the cCRE-supported regions (distance 0 =
overlapping the gene body).

```{r ccre_nearest_gene}
gtf_file <- "/project/spott/cshan/annotations/gencode.v46.annotation.gtf.gz"
gtf <- data.table::fread(cmd = paste("zcat", gtf_file, "| awk -F'\\t' '$3==\"gene\"'"),
                         header = FALSE, sep = "\t")
genes.gr <- GRanges(gtf$V1, IRanges(gtf$V4, gtf$V5),
                    gene_name = sub('.*gene_name "([^"]+)".*', "\\1", gtf$V9),
                    gene_type = sub('.*gene_type "([^"]+)".*', "\\1", gtf$V9))

sc.gr <- GRanges(signif_ccre$chrom, IRanges(signif_ccre$start + 1L, signif_ccre$end))
d <- distanceToNearest(sc.gr, genes.gr)
signif_ccre$nearest_gene  <- NA_character_
signif_ccre$nearest_gene_type <- NA_character_
signif_ccre$gene_distance <- NA_integer_
signif_ccre$nearest_gene[queryHits(d)]      <- genes.gr$gene_name[subjectHits(d)]
signif_ccre$nearest_gene_type[queryHits(d)] <- genes.gr$gene_type[subjectHits(d)]
signif_ccre$gene_distance[queryHits(d)] <- mcols(d)$distance

write_tsv(signif_ccre %>% arrange(min_p),
          file.path(input_dir, "signif_ccre_regions.tsv"))
```

```{r ccre_class_dist, fig.width=6, fig.height=4}
p_ccre_class <- signif_ccre %>%
  count(ccre_class) %>%
  ggplot(aes(x = reorder(ccre_class, -n), y = n)) +
  geom_col(fill = "grey70", color = "grey30") +
  labs(x = "ENCODE cCRE class",
       y = "Significant regions (>= 50% cCRE overlap)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_ccre_class
ggsave(file.path(plot_dir, "signif_ccre_class_dist.pdf"), p_ccre_class,
       width = 6, height = 4)
```

Top 100 cCRE-supported significant regions by best p (full table in
`signif_ccre_regions.tsv`).

```{r ccre_table}
signif_ccre %>%
  arrange(min_p) %>%
  head(100) %>%
  transmute(region_id, best_comparison, n_signif_comparisons,
            min_p = signif(min_p, 3),
            delta_freq_best = round(delta_freq_best, 3),
            ccre_id, ccre_class, ccre_frac = round(ccre_frac, 2),
            nearest_gene, gene_distance) %>%
  datatable(filter = "top", rownames = FALSE, options = list(pageLength = 20))
```


# Read-level plots: top 5 cCRE-supported differential regions

The 5 cCRE-supported significant regions with the smallest p-value and are assoicated with protein coding genes

```{r region_functions}
code_dir <- "/project/spott/cshan/fiber-seq/code/topic_model/read_region_extraction"
source(file.path(code_dir, "region_data_utils.R"))
source(file.path(code_dir, "region_plot_functions.R"))
```

Extraction (tabix/BAM slices per region x sample) runs
`extract_region_result_macrophage.sh` and is skipped when the parsed output
already exists, so re-knits are cheap.

```{r top5_extract}
options(scipen = 999)

extract_script <- file.path(code_dir, "extract_region_result_macrophage.sh")
extract_root <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency/region_plots"
sample_names <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")
pad <- 1000

# only regions whose nearest gene is protein-coding; all five turn out to be
# intronic (gene_distance 0), so no distance is drawn on the plots
top5 <- signif_ccre %>%
  filter(nearest_gene_type == "protein_coding") %>%
  arrange(min_p) %>%
  head(5) %>%
  mutate(outname = paste(chrom, start, end, sep = "_"),
         window_start = as.integer(start - pad),
         window_end   = as.integer(end + pad))
top5 %>% select(region_id, best_comparison, min_p, delta_freq_best,
                ccre_class, ccre_frac, nearest_gene, gene_distance)

for (i in seq_len(nrow(top5))) {
  for (s in sample_names) {
    parsed <- file.path(extract_root, top5$outname[i], s, "parsed")
    if (file.exists(file.path(parsed, "fire.bed"))) next
    region_str <- paste0(top5$chrom[i], ":", top5$window_start[i], "-",
                         top5$window_end[i])
    cat("extracting", top5$outname[i], s, "\n")
    rc <- system2("bash", c(extract_script, s, region_str, top5$outname[i],
                            extract_root))
    if (rc != 0) stop("extraction failed for ", top5$outname[i], " ", s)
  }
}
```

```{r top5_region_plots, fig.width=10, fig.height=16}
for (i in seq_len(nrow(top5))) {
  row <- top5[i, ]
  cat("=== ", row$region_id, " (", row$ccre_class, ", nearest gene: ",
      row$nearest_gene, ") ===\n", sep = "")

  res <- load_region_results(
    region       = data.frame(chr = row$chrom, start = row$window_start,
                              end = row$window_end),
    sample_names = sample_names,
    result_dirs  = file.path(extract_root, row$outname, sample_names, "parsed"),
    fiberHMM_feature      = "tf",
    min_fiberHMM_fp_score = 50,
    fiberHMM_size_breaks  = c(10, 30, 60, 80))

  # facet the read-level panels and the pileup by timepoint
  res <- add_read_groups(res,
                         labels = setNames(as.character(res$rids_df$sample_name),
                                           as.character(res$rids_df$RID)),
                         group_col = "timepoint",
                         group_levels = sample_names)

  # cCREs in the window, one row per cCRE labelled with its class; the tested region
  # itself is the yellow band behind every panel
  peaks_df <- ccre %>%
    filter(chr == row$chrom, end > row$window_start, start < row$window_end) %>%
    arrange(ccre_class, start) %>%
    transmute(chr, start, end, group = ccre_class)

  p <- plot_region_panels(res,
                          plot_name = paste0(row$region_id, "  (", row$ccre_class,
                                             ", nearest gene: ", row$nearest_gene, ")"),
                          peaks_df = peaks_df,
                          highlight = c(row$start, row$end),
                          fiberHMM_cols = c("purple", "magenta", "#41c4e1"),
                          heights = c(pileup = 10, peaks = 2, fire_peaks = 1,
                                      reads = 8, fire_fiberHMM = 8))
  print(p$combined)
  ggsave(file.path(plot_dir, paste0("region_", row$outname, "_by_timepoint.pdf")),
         p$combined, width = 10, height = 16, limitsize = FALSE)
}
```