---
title: "vis_m6a"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(data.table)
library(tidyverse)
library(DT)
```

# Average m6A level at the union FIRE regions

Region-level table from 04_m6a_binned.py: for each union FIRE region and
timepoint, n_cov = read x A/T-site events, n_mod = called m6A events,
m6a = n_mod / n_cov, n_reads = distinct reads overlapping the region.

```{r load_data}
input_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency/methylation"
fire_dir  <- "/project/spott/cshan/fiber-seq/macrophage_project/fire_frequency"
plot_dir  <- file.path(input_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE)

m6a_region <- fread(file.path(input_dir, "m6a_region_long.tsv.gz"))
head(m6a_region)

# one timepoint -> color mapping shared by every plot
timepoint_levels <- c("LPS_0", "LPS_5", "LPS_10", "LPS_15")
timepoint_colors <- setNames(scales::hue_pal()(4), timepoint_levels)
m6a_region$timepoint <- factor(m6a_region$timepoint, levels = timepoint_levels)
```


# Cumulative distribution of region m6A level for each timepoint

for a given m6A level x, what fraction of FIRE-universe regions have level <= x?
Left-shifted curve = overall lower m6A (less accessible) at that timepoint.

```{r CDF_m6a}
p_m6a_cdf <- ggplot(m6a_region, aes(x = m6a, color = timepoint)) +
  stat_ecdf(na.rm = TRUE) +
  scale_color_manual(values = timepoint_colors) +
  labs(
    x = "Average m6A level (n_mod / n_cov)",
    y = "Cumulative fraction of regions",
    color = "Timepoint"
  ) +
  theme_classic()
p_m6a_cdf
ggsave(file.path(plot_dir, "m6a_level_cdf.pdf"), p_m6a_cdf, width = 7, height = 4)
```


# Pairwise differential m6A level

Wilcoxon rank-sum test on per-read m6A fractions (05): reads are the
independent unit, so the p-values are calibrated at read level (unlike a
Fisher test on pooled read x site events, which within-read correlation
inflates).

```{r load_pairwise}
pairwise_m6a <- read_tsv(file.path(input_dir, "pairwise_wilcoxon_m6a.tsv.gz"),
                         show_col_types = FALSE)

comparison_levels <- c("LPS_0_vs_LPS_5", "LPS_0_vs_LPS_10", "LPS_0_vs_LPS_15",
                       "LPS_5_vs_LPS_10", "LPS_5_vs_LPS_15", "LPS_10_vs_LPS_15")
pairwise_m6a <- pairwise_m6a %>%
  mutate(comparison = factor(comparison, levels = comparison_levels))

passing <- pairwise_m6a %>% filter(pass_coverage, !is.na(p_value))

# top 100 per comparison by raw p; head(100) instead of rank_p <= 100 because
# rank_p uses ties.method = "min" and can return != 100 rows at tie boundaries
top100 <- passing %>%
  group_by(comparison) %>%
  arrange(p_value, .by_group = TRUE) %>%
  slice_head(n = 100) %>%
  ungroup() %>%
  mutate(direction = ifelse(delta_m6a > 0, "gain", "loss"))

# gain = higher m6A level at the later timepoint
table(top100$comparison, top100$direction)
write_tsv(top100, file.path(input_dir, "pairwise_m6a_top100.tsv"))
```


# P-value distribution per comparison

Histogram of the unadjusted Wilcoxon p-values over coverage-passing regions.
The dashed red line is the count expected per bin if all p-values were uniform
(no signal). An excess in the leftmost bins = real differential signal (the
remaining caveat is biological: merged pools, no replication).

```{r pvalue_hist, fig.width=12, fig.height=6}
expected_bin <- passing %>%
  count(comparison) %>%
  mutate(expected = n * 0.025)

p_pval_hist <- ggplot(passing, aes(x = p_value)) +
  geom_histogram(binwidth = 0.025, boundary = 0,
                 fill = "grey70", color = "grey30", linewidth = 0.2) +
  geom_hline(data = expected_bin, aes(yintercept = expected),
             linetype = "dashed", color = "red") +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = "Wilcoxon rank-sum p (unadjusted)", y = "Regions") +
  theme_classic()
p_pval_hist
ggsave(file.path(plot_dir, "pvalue_histogram_m6a.pdf"), p_pval_hist, width = 12, height = 6)
```


# Distribution of m6A level changes

```{r delta_violin, fig.width=10, fig.height=5}
p_delta_violin <- ggplot(passing, aes(x = comparison, y = delta_m6a)) +
  geom_violin(fill = "grey80") +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = NULL, y = "delta m6A (later - earlier)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_delta_violin
ggsave(file.path(plot_dir, "delta_m6a_violin.pdf"), p_delta_violin, width = 10, height = 5)
```

```{r volcano, fig.width=12, fig.height=6}
p_volcano <- ggplot(passing, aes(x = delta_m6a, y = -log10(p_value))) +
  geom_point(size = 0.3, alpha = 0.3) +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = "delta m6A (later - earlier)", y = "-log10 Wilcoxon p") +
  theme_classic()
p_volcano
ggsave(file.path(plot_dir, "volcano_pairwise_m6a.pdf"), p_volcano, width = 12, height = 6)
```


# Does m6A change where FIRE frequency changes?

Join with the FIRE-frequency Fisher results by region x comparison. FIRE
elements are called from m6A density, so a positive delta_freq / delta_m6a
correlation is expected by construction; the informative regions are the
discordant ones.

```{r fire_overlay, fig.width=12, fig.height=6}
pairwise_fire <- read_tsv(file.path(fire_dir, "pairwise_fisher.tsv.gz"),
                          show_col_types = FALSE) %>%
  select(region_id, comparison, delta_freq, p_fire = p_value,
         pass_coverage_fire = pass_coverage)

both <- passing %>%
  inner_join(pairwise_fire, by = c("region_id", "comparison")) %>%
  filter(pass_coverage_fire) %>%
  mutate(fire_signif = p_fire < 0.05)

# delta_freq vs delta_m6a (binned 2D counts; ~90k regions x 6 facets)
p_dd <- ggplot(both, aes(x = delta_freq, y = delta_m6a)) +
  geom_bin2d(bins = 80) +
  scale_fill_viridis_c(trans = "log10") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.3) +
  facet_wrap(~ comparison, nrow = 2) +
  labs(x = "delta FIRE frequency", y = "delta m6A", fill = "Regions") +
  theme_classic()
p_dd
ggsave(file.path(plot_dir, "delta_freq_vs_delta_m6a.pdf"), p_dd, width = 12, height = 6)

both %>%
  group_by(comparison) %>%
  summarise(spearman = cor(delta_freq, delta_m6a, method = "spearman",
                           use = "complete.obs"))
```

```{r fire_signif_violin, fig.width=12, fig.height=5}
# |delta_m6a| at FIRE-significant regions (raw Fisher p < 0.05 in that same
# comparison) vs the rest of the coverage-passing universe
p_sig <- ggplot(both, aes(x = fire_signif, y = delta_m6a, fill = fire_signif)) +
  geom_violin() +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ comparison, nrow = 1) +
  labs(x = "FIRE frequency change significant (raw p < 0.05)", y = "delta m6A") +
  theme_classic() +
  theme(legend.position = "none")
p_sig
ggsave(file.path(plot_dir, "delta_m6a_by_fire_sig.pdf"), p_sig, width = 12, height = 5)

# regions significant in both tests (raw p < 0.05, per comparison)
both %>%
  group_by(comparison) %>%
  summarise(n_pass = n(),
            fire_sig = sum(fire_signif),
            m6a_sig = sum(p_value < 0.05),
            both_sig = sum(fire_signif & p_value < 0.05))
```


# 10 bp binned profiles of the top differential regions

Per-bin m6A level across the region, one line per timepoint, for the 5 regions
with the smallest Fisher p over all comparisons.

```{r top5_profiles, fig.width=10, fig.height=8}
m6a_bins <- fread(file.path(input_dir, "m6a_bins_long.tsv.gz"))

top5_ids <- passing %>%
  arrange(p_value) %>%
  distinct(region_id) %>%
  slice_head(n = 5) %>%
  pull(region_id)
top5_ids

p_top5 <- m6a_bins %>%
  filter(region_id %in% top5_ids) %>%
  mutate(bin_mid = (bin_start + bin_end) / 2,
         timepoint = factor(timepoint, levels = timepoint_levels)) %>%
  ggplot(aes(x = bin_mid, y = m6a, color = timepoint)) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 0.6, na.rm = TRUE) +
  scale_color_manual(values = timepoint_colors) +
  facet_wrap(~ region_id, ncol = 1, scales = "free_x") +
  labs(x = "Position (10 bp bin midpoint)", y = "m6A level", color = "Timepoint") +
  theme_classic()
p_top5
ggsave(file.path(plot_dir, "top5_m6a_profiles.pdf"), p_top5, width = 10, height = 8)
```


# Meta-profile across scaled region positions

Bins mapped to their relative position within the region (0 = region start,
1 = region end), n_cov-weighted mean m6A per position bin per timepoint.
Flat separation between timepoints = a uniform shift; edge-vs-center
differences = shape changes at the accessible region.

```{r metaprofile, fig.width=7, fig.height=4}
region_coords <- m6a_region %>%
  distinct(region_id, start, end)

meta <- m6a_bins %>%
  inner_join(region_coords, by = "region_id") %>%
  mutate(rel_pos = ((bin_start + bin_end) / 2 - start) / (end - start),
         pos_bin = cut(rel_pos, breaks = seq(0, 1, 0.05), include.lowest = TRUE)) %>%
  group_by(timepoint, pos_bin) %>%
  summarise(m6a = sum(n_mod) / sum(n_cov), .groups = "drop") %>%
  mutate(pos_mid = (as.integer(pos_bin) - 0.5) * 0.05,
         timepoint = factor(timepoint, levels = timepoint_levels))

p_meta <- ggplot(meta, aes(x = pos_mid, y = m6a, color = timepoint)) +
  geom_line() +
  scale_color_manual(values = timepoint_colors) +
  labs(x = "Relative position in region", y = "Mean m6A level (pooled)",
       color = "Timepoint") +
  theme_classic()
p_meta
ggsave(file.path(plot_dir, "m6a_metaprofile.pdf"), p_meta, width = 7, height = 4)
```
