---
title: "Hamming + Leiden clustering of single molecules at early-response promoters"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

A hybrid of the two existing single-molecule clusterings, on THP-1 Fiber-seq
m6A at the early-response promoters:

* **Distance**: the masked Hamming distance of the hamming_hierarchical
  analysis (Bellini, Do & Skok 2025; `masked_hamming()` in
  `phasing_functions.r`), computed on the same binary-bin feature matrices
  that analysis clusters (a read's bin value is 1 if the read has any m6A
  call in the bin - fractional bins compress all masked Hamming distances,
  see that notebook's deviations).
* **Clustering and number of clusters**: the Leiden_Manhattan analysis's
  graph pipeline (Raviram et al. 2026, Methods steps 5-7): a KNN graph over
  reads (k = 10 nearest neighbours), Manhattan/Hamming distances turned into
  edge weights by an exponential kernel `exp(-d / sigma)`, and Leiden
  community detection (`RBConfigurationVertexPartition`) at resolution 1.
  As there, the number of clusters is whatever Leiden returns at the set
  resolution - not the hamming analysis's within-cluster dispersion elbow -
  and clusters are renumbered largest-first.

Shared with both parents: the full-span read set (`filter_met_mat()`, the
topic model's row filter), all four LPS timepoints pooled at their observed
depths with each read keeping its timepoint tag, the TSS +/- 1000 bp
promoter table, and the bin-size sweep (`bin_sizes`: 0 = no binning, i.e.
the 0/1 m6A site matrix itself, where binarization is a no-op).

No KNN imputation step: masking is the masked Hamming distance's own way of
handling missing features (each pair is compared over co-observed features
only), so nothing needs imputing even off the full-span read set.

Note: on full-span reads the feature matrices have no missing values, so the
masked Hamming distance is exactly the Manhattan distance divided by the
number of features, and the default kernel width (mean retained KNN distance)
is scale-free - with `binarize_bins = FALSE` this pipeline would reproduce
Leiden_Manhattan exactly. The binary bins are what make it different at
bin_sizes > 0; the bin-0 run differs from Leiden_Manhattan's bin-0 run only
through the (scale-free) distance normalisation, i.e. not at all.

Outputs go to `<gene>/bin<size>/k<k>/` (k = the number of Leiden clusters
that run produced) under the output directory, with the same tables and the
same five plots as the Leiden_Manhattan notebook
(`leiden_manhattan_plots.r`).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
# igraph needs libglpk at load time; if library(igraph) fails with
# "libglpk.so.40: cannot open shared object file", add glpk's lib directory
# (/software/glpk-5.0-el8-x86_64/lib on this cluster) to LD_LIBRARY_PATH before
# starting R.
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(data.table)
library(Rsamtools)   # extract_ft_region_reads() reads the tabix-indexed beds
library(igraph)
library(ComplexHeatmap)
library(cowplot)
```

# Functions

`masked_hamming()` comes from the hamming analysis; `bin_read_matrix()`,
`leiden_partition()`, `assemble_region_m6a()`, `cluster_site_profiles()` and
the plots come from the Leiden_Manhattan analysis. Source order matters:
`phasing_functions.r` also defines `plot_cluster_met_profiles()` /
`plot_cluster_composition()`, so it is sourced before
`leiden_manhattan_plots.r` and the Leiden versions (which take this
notebook's result objects) win.

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_hierarchical/phasing_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
source("/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/hamming_leiden/hamming_leiden_functions.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

promoter_table <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_hierarchical/promoter_regions_gencode_v49.tsv"

output_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/hamming_leiden"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Bin-size sweep: one full clustering run per promoter per entry. Positive =
# bin size in bp; 0 = no binning, i.e. each m6A site is its own feature (the
# default, and the first entry here). Output folders: <gene>/bin<size>/k<k>/,
# k = the number of Leiden clusters.
bin_sizes <- c(0, 10, 20, 30, 50)

# Binary bins (the hamming analysis's binarize_bins, on): a read's bin value
# is 1 if the read has any m6A call in the bin, instead of the mean site call.
# No effect on the bin-0 run, whose site calls are already 0/1.
binarize_bins <- TRUE

# Leiden / graph parameters - identical to the Leiden_Manhattan notebook, so
# any clustering difference comes from the distance/features, not the graph:
# k = 10 neighbours (the paper's k = 50 would connect ~25% of these ~150-250
# full-span read sets) and resolution 1 (standard modularity; the paper's
# 2.1-3 shatters graphs this size). Raise the resolution for finer clusters.
k_neighbors       <- 10
leiden_resolution <- 1
leiden_seed       <- 1

# Exponential-kernel width for the affinities: NULL = mean of the retained KNN
# distances (scale-free across bin sizes). Set to the number of features to
# reproduce scikit-learn's laplacian_kernel default (gamma = 1 / n_features).
kernel_sigma <- NULL

# Rolling-mean width for the m6A fraction line plot only (1 = raw profiles).
line_smooth_k <- 1

set.seed(123)
```

# Promoter regions

Same table and genes as the hamming, spearman and Leiden_Manhattan analyses:
TSS +/- 1000 bp from GENCODE v49 (`make_promoter_table.R`).

```{r regions}
promoters <- read.delim(promoter_table)
promoters
```

# Clustering at each promoter

Per gene and per `bin_sizes` entry: assemble the pooled m6A matrix once per
gene, apply the full-span filter, build the read x feature matrix (binary
bins; bin size 0 uses the m6A site matrix directly), then masked Hamming
distances, the KNN graph with exponential-kernel affinities, and Leiden.
Outputs per gene x bin size, under `<gene>/bin<size>/k<k>/`: the result
object (.rds), per-read cluster assignments **with the read's timepoint**,
cluster profiles, the feature matrix, the window annotation, cluster x
timepoint counts, the run parameters (.tsv), and the five plots.

```{r cluster-promoters, fig.width=9, fig.height=9}
all_results <- list()
for (i in seq_len(nrow(promoters))) {
  gene <- promoters$gene[i]
  cat("\n=====", gene, "=====\n")

  dat <- assemble_region_m6a(sample_names, promoters$chr[i],
                             promoters$start[i], promoters$end[i],
                             ft_result_dir, verbose = FALSE)
  cat(gene, ": ", nrow(dat$rids_df), " reads x ", ncol(dat$met_mat),
      " m6A sites before filtering\n", sep = "")

  met_mat <- filter_met_mat(dat$met_mat, verbose = TRUE)
  rids_df <- dat$rids_df[match(rownames(met_mat), as.character(dat$rids_df$RID)), ]
  cat(gene, ": ", nrow(met_mat), " full-span reads x ", ncol(met_mat),
      " m6A sites\n", sep = "")

  for (bs in bin_sizes) {
    lab      <- paste0("bin", bs)
    bin_desc <- if (bs == 0) "per-site, no bins" else paste0(bs, "-bp binary bins")
    gene_lab <- paste0(gene, " (", bin_desc, ")")
    cat("--- ", gene, ", ", bin_desc, " ---\n", sep = "")

    res <- hamming_leiden_cluster(met_mat, rids_df,
                                  region_start  = promoters$start[i],
                                  region_end    = promoters$end[i],
                                  bin_size      = bs,
                                  binarize_bins = binarize_bins,
                                  k_neighbors   = k_neighbors,
                                  resolution    = leiden_resolution,
                                  sigma         = kernel_sigma,
                                  seed          = leiden_seed)

    # every read keeps its timepoint alongside its cluster
    res$assignments$sample_name <- factor(res$assignments$sample_name,
                                          levels = sample_names)
    res$assignments$timepoint <- sample_table$timepoint[
      match(res$assignments$sample_name, sample_table$sample_name)]
    all_results[[gene]][[lab]] <- res

    # k subfolder = the number of Leiden clusters this run produced
    k_dir <- file.path(output_dir, gene, lab, paste0("k", res$n_clusters))
    if (!dir.exists(k_dir)) dir.create(k_dir, recursive = TRUE)

    ## tables + rds
    saveRDS(res, file.path(k_dir, paste0("hamming_leiden_", gene, ".rds")))
    write.table(res$assignments,
                file.path(k_dir, paste0("read_cluster_assignments_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(cluster = rownames(res$profiles), res$profiles,
                           check.names = FALSE),
                file.path(k_dir, paste0("cluster_profiles_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$window_anno,
                file.path(k_dir, paste0("window_annotation_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(RID = rownames(res$feat_mat),
                           sample_name = res$assignments$sample_name,
                           cluster = res$assignments$cluster,
                           res$feat_mat, check.names = FALSE),
                file.path(k_dir, paste0("feature_matrix_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(res$assignments %>% dplyr::count(cluster, sample_name, timepoint),
                file.path(k_dir, paste0("cluster_timepoint_counts_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
    write.table(data.frame(parameter = names(res$params),
                           value = vapply(res$params, as.character, character(1))),
                file.path(k_dir, paste0("run_parameters_", gene, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)

    ## 1. per-cluster m6A methylation profiles (topic-model style)
    gg <- plot_cluster_met_profiles(res, met_mat, tss = promoters$tss[i],
                                    main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_met_profiles_", gene, ".pdf")), gg,
           width = 6, height = 1.4 * res$n_clusters + 1.4, limitsize = FALSE)

    ## 2. cluster composition by timepoint, and cluster structure
    gg <- plot_cluster_composition(res, main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_composition_", gene, ".pdf")), gg,
           width = 7, height = 8)
    gg <- plot_cluster_structure(res, main = gene_lab)
    ggsave(file.path(k_dir, paste0("cluster_structure_", gene, ".pdf")), gg,
           width = 9, height = 3)

    ## 3. read x feature methylation heatmap
    pdf(file.path(k_dir, paste0("heatmap_", gene, ".pdf")), width = 8, height = 9)
    ComplexHeatmap::draw(plot_cluster_heatmap(res, main = gene_lab))
    dev.off()

    ## 4. m6A fraction line plot (one panel per cluster)
    gg <- plot_met_fraction_lines(res, tss = promoters$tss[i], main = gene_lab,
                                  smooth_k = line_smooth_k)
    if (binarize_bins && bs > 0)   # binary bins: the mean is a read fraction
      gg <- gg + ylab(sprintf("fraction of reads with m6A per %d-bp bin", bs))
    ggsave(file.path(k_dir, paste0("met_fraction_lines_", gene, ".pdf")), gg,
           width = 8, height = 1.4 * res$n_clusters + 1, limitsize = FALSE)
  }
}
```

# Cluster sizes and timepoint composition across genes

`n_clusters` is the number of Leiden clusters for that gene x bin size at
`leiden_resolution`; `n` counts reads.

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    all_results[[g]][[lab]]$assignments %>%
      dplyr::count(cluster) %>%
      dplyr::mutate(gene = g, bins = lab,
                    n_clusters = all_results[[g]][[lab]]$n_clusters,
                    .before = 1)
  }))
}))
write.table(cluster_summary, file.path(output_dir, "cluster_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = cluster, values_from = n)

timepoint_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    all_results[[g]][[lab]]$assignments %>%
      dplyr::count(cluster, sample_name, timepoint) %>%
      dplyr::mutate(gene = g, bins = lab, .before = 1)
  }))
}))
write.table(timepoint_summary,
            file.path(output_dir, "cluster_timepoint_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

## run parameters per gene x bin size (resolution, kernel width, graph size)
run_summary <- dplyr::bind_rows(lapply(names(all_results), function(g) {
  dplyr::bind_rows(lapply(names(all_results[[g]]), function(lab) {
    p <- all_results[[g]][[lab]]$params
    data.frame(gene = g, bins = lab, n_reads = p$n_reads,
               n_features = p$n_features, k_eff = p$k_eff,
               resolution = p$resolution, sigma = p$sigma,
               mean_knn_dist = p$mean_knn_dist, quality = p$quality,
               n_clusters = all_results[[g]][[lab]]$n_clusters)
  }))
}))
write.table(run_summary, file.path(output_dir, "run_summary_all_genes.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
run_summary
```

# Session info

```{r session-info}
sessionInfo()
```
