---
title: "LCL Fiber-seq: top 10 heterozygous AS-FIRE regions"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    self_contained: true
params:
  regenerate_plots: true
---

```{r lcl-knit, include=FALSE}
knit_report <- function(inputFile, encoding = "UTF-8") {
  lines <- readLines(inputFile, warn = FALSE)
  start <- which(trimws(lines) == "```{r lcl-output-loader}")
  stopifnot(length(start) == 1L)
  end <- which(seq_along(lines) > start & trimws(lines) == "```")[1L]
  stopifnot(!is.na(end), end > start + 1L)
  eval(parse(text = lines[seq.int(start + 1L, end - 1L)]))
  source_lcl_output_config(inputFile)
  source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_functions.r", local = TRUE)
  source("/project/spott/cshan/fiber-seq/code/haplotype_phasing/LCL_phasing.r", local = TRUE)
  source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r", local = TRUE)
  # RStudio's Knit button uses a separate report for the top-ten analysis.
  render_report <- function(input, ...) {
    root <- lcl_output_path(LCL_TOP10_ROOT)
    dir.create(root, recursive = TRUE, showWarnings = FALSE)
    intermediates <- lcl_output_path(file.path(root, "report_intermediates"))
    dir.create(intermediates, recursive = TRUE, showWarnings = FALSE)
    rmarkdown::render(input, output_file = "leiden_LCL_top10_asfire_het_2kb.html",
      output_dir = "/project/spott/cshan/fiber-seq/code/docs",
      intermediates_dir = intermediates, knit_root_dir = root, ...)
  }
  render_report(inputFile, encoding = encoding)
}
```

```{r manual-render, eval=FALSE, include=FALSE}
knit_report("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_LCL.Rmd")
```

```{r lcl-output-loader}
#| include: false
source_lcl_output_config <- function(notebook, envir = parent.frame()) {
  lines <- readLines(notebook, warn = FALSE)
  start <- which(trimws(lines) == "```{r lcl-output-config}")
  if (length(start) != 1L) stop("Expected one LCL output configuration chunk")
  end <- which(seq_along(lines) > start & trimws(lines) == "```")[1L]
  if (is.na(end) || end <= start + 1L) stop("Missing LCL output configuration code")
  eval(parse(text = lines[seq.int(start + 1L, end - 1L)]), envir = envir)
  invisible(NULL)
}
```

```{r lcl-output-config}
#| include: false
LCL_OUTPUT_ROOT <- "/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan"
LCL_TOP10_ROOT <- Sys.getenv("LCL_REPORT_ROOT", unset = LCL_OUTPUT_ROOT)
lcl_output_path <- function(path) {
  if (any(strsplit(path, "/", fixed = TRUE)[[1]] %in% c(".", "..")))
    stop("Output path must not contain dot components")
  root <- normalizePath(LCL_OUTPUT_ROOT, mustWork = TRUE)
  ancestor <- path
  suffix <- character()
  while (!file.exists(ancestor)) {
    suffix <- c(basename(ancestor), suffix)
    parent <- dirname(ancestor)
    if (identical(parent, ancestor)) stop("Cannot resolve output path")
    ancestor <- parent
  }
  resolved <- do.call(file.path, as.list(c(normalizePath(ancestor), suffix)))
  if (!(identical(resolved, root) || startsWith(resolved, paste0(root, "/"))))
    stop("Output must be inside ", root, ": ", path)
  resolved
}
```

The AS-FIRE analyses use only fully spanning, read-level phased heterozygotes:
HP1/HP2, a focal-SNP genotype of `0|1` or `1|0`, and a resolved SNP allele.
`lcl_filter_focal_heterozygotes()` applies this requirement before Manhattan
clustering for every selected region. Fourier and autocorrelation now apply
the same filter before their transforms, so their summaries and plots also
exclude unphased, unresolved and nonheterozygous reads. This policy applies to
future AS-FIRE regions as well; each must supply focal-SNP and sample metadata.
The promoter analysis below instead pools all fully spanning reads without phasing.

```{r setup, include=FALSE}
code_dir <- "/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan"
suppressPackageStartupMessages(library(dplyr))
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r", local = TRUE)
source(file.path(code_dir, "leiden_manhattan_functions.r"), local = TRUE)
source("/project/spott/cshan/fiber-seq/code/haplotype_phasing/LCL_phasing.r", local = TRUE)
source(file.path(code_dir, "leiden_manhattan_plots.r"), local = TRUE)
output_dir <- lcl_output_path(LCL_TOP10_ROOT)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
  fig.path = file.path(output_dir, "report_figures/"), dev = "pdf", cache = FALSE)
options(scipen = 999)
data.table::setDTthreads(2L)
```



knn plot colored by alleles
```{r plot-knn-by-allele, include=TRUE}
plot_knn_allele_composition <- function(summary, main = NULL) {
  composition <- summary$composition
  counts <- tapply(composition$n_reads, composition$group, sum)
  ggplot2::ggplot(composition, ggplot2::aes(group, fraction, fill = allele)) +
    ggplot2::geom_col(width = 0.75, position = ggplot2::position_stack(reverse = TRUE)) +
    ggplot2::geom_text(ggplot2::aes(label = ifelse(fraction >= 0.05, scales::percent(fraction, accuracy = 1), "")),
      position = ggplot2::position_stack(vjust = 0.5, reverse = TRUE), color = "white", size = 3) +
    ggplot2::scale_fill_manual(values = fiberseq_category_palette(summary$allele_levels), drop = FALSE) +
    ggplot2::scale_x_discrete(labels = function(labels) paste0(labels, "\n(n=", counts[labels], ")")) +
    ggplot2::scale_y_continuous(labels = scales::percent, limits = c(0, 1),
      expand = ggplot2::expansion(mult = c(0, 0.02))) +
    ggplot2::labs(title = main, x = "Leiden cluster", y = "Proportion of retained reads", fill = "Focal SNP allele",
      subtitle = "All reads: region-wide baseline; each bar sums to 100%") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
}

plot_knn_allele_connectivity <- function(summary, main = NULL) {
  connectivity <- summary$connectivity
  connected <- connectivity[connectivity$n_reads_with_neighbors > 0L, , drop = FALSE]
  if (!nrow(connected)) stop("No reads have neighbors in the saved KNN graph")
  bars <- dplyr::bind_rows(
    data.frame(allele = connected$allele, relationship = "Same allele", fraction = connected$same_fraction),
    data.frame(allele = connected$allele, relationship = "Opposite allele", fraction = connected$opposite_fraction))
  bars$relationship <- factor(bars$relationship, levels = c("Same allele", "Opposite allele"))
  counts <- setNames(connectivity$n_reads_with_neighbors, as.character(connectivity$allele))
  ggplot2::ggplot(bars, ggplot2::aes(allele, fraction, fill = relationship)) +
    ggplot2::geom_col(width = 0.65, position = ggplot2::position_stack(reverse = TRUE)) +
    ggplot2::geom_text(ggplot2::aes(label = ifelse(fraction >= 0.05, scales::percent(fraction, accuracy = 1), "")),
      position = ggplot2::position_stack(vjust = 0.5, reverse = TRUE), color = "white", size = 3) +
    ggplot2::geom_point(data = connected, ggplot2::aes(x = allele, y = expected_same_fraction),
      inherit.aes = FALSE, shape = 23, size = 3, fill = "white", color = "black") +
    ggplot2::scale_fill_manual(values = c("Same allele" = "#0072B2", "Opposite allele" = "#D55E00"), drop = FALSE) +
    ggplot2::scale_x_discrete(labels = function(labels) paste0(labels, "\n(n=", counts[labels], ")")) +
    ggplot2::scale_y_continuous(labels = scales::percent, limits = c(0, 1),
      expand = ggplot2::expansion(mult = c(0, 0.02))) +
    ggplot2::labs(title = main, x = "Source read's focal SNP allele", y = "Mean fraction of neighbors", fill = "Neighbor relationship",
      subtitle = "Unweighted neighbor fractions, averaged equally across reads within each allele",
      caption = paste0("Uses all undirected graph neighbors, including connections between Leiden clusters.\n",
        "Diamonds: region-wide random-mixing baseline excluding the source read, not a significance test.\n",
        sum(connectivity$n_isolated_reads), " isolated reads excluded from connectivity means; retained in composition bars.")) +
    ggplot2::theme_bw(base_size = 11)
}

```

```{r lcl-figure-helpers, include=FALSE}
save_plots <- function(result, footprints, sample_colors, output_dir, include_haplotype = FALSE) {
  output_dir <- lcl_output_path(output_dir)
  plot_dir <- file.path(output_dir, "plots")
  dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
  ggplot2::ggsave(file.path(plot_dir, "knn_graph.pdf"), plot_knn_graph(result),
    width = 12, height = 9, bg = "white")
  signals <- signal_profiles(result, footprints)
  profiles <- signals$profiles
  records <- signals$records
  matched <- match(records$RID, result$assignments$RID)
  for (column in intersect(c("cluster", "sample_name", "haplotype", "phase_set", "group_id"),
                           names(result$assignments))) records[[column]] <- result$assignments[[column]][matched]
  summary <- methylation_by_cluster(result)
  ggplot2::ggsave(file.path(plot_dir, "methylation_proportion_by_cluster.pdf"),
    plot_methylation_by_cluster(summary, result$region), width = 8, height = 4.5)
  heatmap_file <- if (include_haplotype) "heatmap_m6a_footprints.pdf" else "heatmap_m6a.pdf"
  heatmap_height <- if (include_haplotype) 11 + 0.55 * result$n_clusters else 11
  heatmap <- plot_genomic_cluster_heatmap(result, result$region, sample_colors,
    include_haplotype = include_haplotype, variants = result$variants,
    show_cluster_profiles = include_haplotype)
  grDevices::pdf(file.path(plot_dir, heatmap_file), width = 14, height = heatmap_height)
  tryCatch(ComplexHeatmap::draw(heatmap, newpage = FALSE), finally = grDevices::dev.off())
  ggplot2::ggsave(file.path(plot_dir, "fig1_read_footprints.pdf"),
    plot_smf_reads(result, records, sample_colors, include_haplotype = include_haplotype),
    width = 12, height = max(7, 0.03 * nrow(result$assignments) + 3.5 + 0.3 * result$n_clusters), limitsize = FALSE)
  ggplot2::ggsave(file.path(plot_dir, "fig2_occupancy_by_cluster.pdf"),
    plot_signal_profile(profiles, result$region, levels(result$assignments$cluster)),
    width = 10, height = 1.4 * result$n_clusters + 2.5, limitsize = FALSE)
  ggplot2::ggsave(file.path(plot_dir, "cluster_sample_composition.pdf"),
    plot_sample_composition(result, sample_colors, result$region$annotation), width = 12, height = 5)
  invisible(signals)
}


save_all_plots <- function(result, footprints, output_dir, sample_colors,
                               embedding = NULL) {
  plotted <- save_fiberseq_plots(result, footprints, output_dir, sample_colors,
    embedding = embedding)
  view <- fiberseq_display_result(result)
  short_colors <- setNames(unname(sample_colors), sub("_.*$", "", names(sample_colors)))
  save_plots(view, footprints, short_colors, output_dir, include_haplotype = TRUE)
  invisible(plotted)
}

```

```{r lcl-report-helpers, include=FALSE}
report_url <- function(path) {
  path <- lcl_output_path(path)
  if (!file.exists(path)) stop("Missing report artifact: ", path)
  paste0("https://gate.rcc.uchicago.edu/spott/file_show?path=",
    utils::URLencode(path, reserved = TRUE))
}
```

<style>
.main-container { max-width: 1700px; padding: 0 24px; }
table { display: block; overflow-x: auto; }
</style>

# Select the top 10 AS-FIRE regions

*Definition of AS-FIRE and how it is selected:*

1. AS-FIRE: whehther a FIRE peak is accessible at different frequencies on reads carrying  a SNP's reference vs alternate allele
2. retains genotypes 0|1 or 1|0, retains reads tagged H1/H2, and pairs peaks with nearby variants using a 2,000-bp distance setting
  - Each SNP is associated with a set of contributing samples based on its phase and the phase of the reads
  - Different SNPs can therefore have different contributing samples
3. Counts are combined across contributing samples for each SNP–peak pair 
  - Fisher’s exact test compares reference/alternate alleles against FIRE/non-FIRE observations
4. Significance filters
  - combined coverage >= 20 reads
  - reference and alternate alleles each have >= 0 reads
  - q value < 0.1
  - FIRE frequency >= 0.1
  
  
AS-FIRE from kevin:
  - /project/spott/kevinluo/Fiber_seq/results/QTL/fireQTL/AS_fire_freq_results/combined_results/combined_AS_fire_freq_merged_peaks_window2000_firefreq0.1_het_only.rds
  
| peak_name | rsID | snp_pos | ref | alt | coverage_ref | coverage_alt | fisher_pvalue |
|---|---|---:|---|---|---:|---:|---:|
| chr1_11161_11494 | rs180734498 | 13302 | C | T | 2 | 3 | 0.5238095 |
| chr1_28837_29482 | rs140337953 | 30923 | G | T | 19 | 25 | 0.6412728 |
| chr1_88661_88783 | rs146836579 | 87647 | T | C | 15 | 5 | 0.6286665 |
| chr1_88661_88783 | rs1524602 | 87190 | G | A | 15 | 5 | 0.6286665 |
| chr1_88661_88783 | rs59529791 | 88136 | G | A | 15 | 5 | 0.6286665 |


*Selection of top 10 regions:*
Only biallelic A/C/G/T SNPs with a valid rsID and a complete 2 kb reference window
are eligible. Each region is defined by the lead SNP's position and the FIRE peak, 
in this case, it is the SNP with the smallest Fisher p-value within the peak

  - allele label: each phased read is assigned to H1 or H2 based on the focal SNP's
  phased genotype in the sample
  - lowest p-value: the SNP with the lowest Fisher p-value within the peak is selected as the focal SNP for that region
  - before clustering, homozygous and unphased reads are removed

example block table:
  - `/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples/AL10_bc2178_19130/blocks.tsv`


| source_block_index | sample_name | phase_block_id | chrom | start | end | num_variants |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| 1 | NA19130 | 52238 | chr1 | 52238 | 103905 | 16 |
| 1 | NA19130 | 126108 | chr1 | 126108 | 135032 | 2 |
| 4 | NA19130 | 591460 | chr1 | 591460 | 599165 | 5 |
| 5 | NA19130 | 638771 | chr1 | 638771 | 638771 | 1 |
| 5 | NA19130 | 665098 | chr1 | 665098 | 683083 | 3 |
| 6 | NA19130 | 779825 | chr1 | 779825 | 1190918 | 437 |
| 8 | NA19130 | 1285767 | chr1 | 1285767 | 1290261 | 2 |
| 9 | NA19130 | 1339884 | chr1 | 1339884 | 1339911 | 2 |
| 12 | NA19130 | 1427454 | chr1 | 1427454 | 1427474 | 3 |
| 14 | NA19130 | 1457071 | chr1 | 1457071 | 1457071 | 1 |
| 15 | NA19130 | 1483249 | chr1 | 1483249 | 1492007 | 3 |
| 16 | NA19130 | 1514366 | chr1 | 1514366 | 1516323 | 2 |
| 17 | NA19130 | 1543702 | chr1 | 1543702 | 1922424 | 259 |

example read level haplotype file:
  - `/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples/AL10_bc2178_19130/read-level-phasing.tsv`
  - haplotag 1 = first allele in GT
  
| source_block_index | sample_name | chrom | phase_block_id | read_name | haplotag |
|---:|---|---|---:|---|---:|
| 4 | NA19130 | chr1 | 591460 | m84241_250311_201728_s2/81201882/ccs | 1 |
| 4 | NA19130 | chr1 | 591460 | m84241_250311_201728_s2/219218451/ccs | 1 |
| 4 | NA19130 | chr1 | 591460 | m84241_250311_201728_s2/65275399/ccs | 2 |
| 4 | NA19130 | chr1 | 591460 | m84241_250311_201728_s2/7869529/ccs | 1 |
| 4 | NA19130 | chr1 | 591460 | m84241_250311_201728_s2/56759899/ccs | 1 |

example vcf file:
  - `/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples/AL10_bc2178_19130/AL10_bc2178_19130.5mC.6mA.aligned.phased.vcf.gz`
    - GT = 0|1 means HP1 carries reference allele and HP2 carries alternate allele
    - PS =  a cluster of phased genotypes that belong to the same local chromosome block
    - GT:PS = 0|1:100 with A and T, and GT:PS = 0|1:100 with C and G means haplotype A reads A-C and haplotype B reads T-G
  
  
The block file and vcf file are used to determine the phased genotype of each read at the focal SNP
  - they are matched based on the sample name, chr, and phase_block_id - groups of variants whose alleles
  have been assgned to the same haplotype (H1 or H2) based on the phased genotype of the sample


example RDS file produced by cache_lcl_haplotags
  - finds the block containing the focal SNP, then matches its ID to the VCF and read haplotags to determine each read’s allele
  - `/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan/phase_summary/AL10_bc2178_19130_haplotags.rds`: a list contain the blocks file and the tags file with haplotype label for each read



```{r settings}
asfire_file <- "/project/spott/kevinluo/Fiber_seq/results/QTL/fireQTL/AS_fire_freq_results/combined_results/combined_AS_fire_freq_merged_peaks_window2000_firefreq0.1_het_only.rds"
sample_csv <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv"
sample_table <- read.csv(sample_csv, stringsAsFactors = FALSE)
stopifnot(!anyDuplicated(sample_table$sample_name),
          !anyDuplicated(sub("_.*$", "", sample_table$sample_name)))
phasing_root <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples"
signature_file <- file.path(output_dir, "summary tables", "selection_signature.rds")
if (file.exists(signature_file)) {
  signature <- readRDS(signature_file)
  regions <- signature$regions
  sample_table <- signature$sample_table
} else {
  asfire_results <- readRDS(asfire_file)
  regions <- select_lcl_top_asfire_regions(asfire_results, top_n = 10L, width_bp = 2000L)
  rm(asfire_results)
}
knitr::kable(regions[, c("rank", "region_id", "peak_name", "focal_snp", "focal_pos",
  "fisher_pvalue", "fisher_qvalue", "coverage", "coverage_ref", "coverage_alt",
  "n_listed_samples", "analysis_start", "analysis_end", "overlapping_selected_windows",
  "overlapping_selected_peaks", "nearest_selected_snp_distance_bp", "same_focal_snp_as_another_peak")])
```

plot methylation by cluster
```{r methylation-figures, include=TRUE}

plot_methylation_by_cluster <- function(summary, region) {
  summary$cluster <- factor(summary$cluster, levels = summary$cluster)
  ggplot2::ggplot(summary, ggplot2::aes(cluster, methylation_proportion, fill = cluster)) +
    ggplot2::geom_col(width = 0.7) +
    ggplot2::scale_fill_manual(values = cluster_id_palette(levels(summary$cluster)), guide = "none") +
    ggplot2::scale_y_continuous(limits = c(0, 1), expand = ggplot2::expansion(mult = c(0, 0.03))) +
    ggplot2::labs(x = "m6A-defined cluster", y = "Mean m6A call proportion", title = region$annotation,
      subtitle = "Mean across observed m6A sites; all cluster reads included") +
    cowplot::theme_cowplot(font_size = 10) + cowplot::panel_border()
}

methylation_by_cluster <- function(result) {
  dplyr::bind_rows(lapply(levels(result$assignments$cluster), function(cluster) {
    ids <- result$assignments$RID[result$assignments$cluster == cluster]
    m <- result$site_met_mat[ids, , drop = FALSE]
    n_calls <- sum(m > 0)
    denominator <- nrow(m) * ncol(m)
    data.frame(cluster = cluster, n_reads = nrow(m), n_sites = ncol(m),
      n_m6a_calls = n_calls, denominator = denominator,
      methylation_proportion = if (denominator) n_calls / denominator else NA_real_)
  }))
}

```

# Main results and haplotype examples

```{r generate-results, include=FALSE}
if (params$regenerate_plots) {
  if (file.exists(file.path(output_dir, "summary tables", "fiberseq_result_paths.rds"))) {
    # Redraw the saved clusters without rebuilding matrices or clustering.
    result_paths <- local({
      tracks_dir <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks"
      output_dir <- lcl_output_path(output_dir)
      summary_dir <- file.path(output_dir, "summary tables")
      paths <- readRDS(file.path(summary_dir, "fiberseq_result_paths.rds"))
      signature <- readRDS(file.path(summary_dir, "selection_signature.rds"))
      stopifnot(all(signature$regions$region_id %in% names(paths)))
      paths <- paths[signature$regions$region_id]
      caches <- cache_footprint_tracks(signature$regions, tracks_dir,
        file.path(output_dir, "footprint_summary"), workers = 3L, reuse = TRUE)
      colors <- sample_palette(signature$sample_table$sample_name)
      audit <- list()
      for (rid in names(paths)) {
        message("Reproducing saved-cluster plots: ", rid)
        original_path <- paths[[rid]]
        result <- readRDS(original_path)
        original <- result$assignments
        original_hash <- unname(tools::md5sum(original_path))
        out <- lcl_output_path(file.path(output_dir, rid, "bin0", "k10", "resolution1"))
        table_dir <- file.path(out, "tables")
        dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
        # Consolidate the existing annotations and main-analysis audits into one region folder.
        for (source_dir in unique(c(dirname(original_path),
          file.path(output_dir, "results", rid, "bin0", "k10", "resolution1", "tables")))) {
          if (dir.exists(source_dir) && normalizePath(source_dir) != normalizePath(table_dir)) {
            entries <- list.files(source_dir, full.names = TRUE)
            entries <- entries[!file.exists(file.path(table_dir, basename(entries)))]
            if (length(entries) && !all(file.copy(entries, table_dir, recursive = TRUE))) stop("Table consolidation failed")
          }
        }
        result <- fiberseq_display_result(result)
        sample_colors <- colors[names(colors) %in% result$assignments$sample_label]
        nuc <- extract_nucleosomes(lcl_region_samples(result$region, signature$sample_table),
          result$region, result$assignments)
        tf <- region_footprints(caches, result$region, result$assignments)$records
        footprints <- dplyr::bind_rows(nuc, tf)
        embedding <- fiberseq_umap(result, seed = signature$view_params$seed)
        embedding <- embedding[match(result$assignments$RID, embedding$RID), , drop = FALSE]
        full_colors <- setNames(unname(colors[sub("_.*$", "", signature$sample_table$sample_name)]),
          signature$sample_table$sample_name)
        full_colors <- full_colors[names(full_colors) %in% result$assignments$sample_name]
        save_all_plots(result, footprints, out, full_colors,
          embedding = embedding)
        paths[[rid]] <- file.path(table_dir, "fiberseq_annotated_clustering.rds")
        stopifnot(identical(result$assignments$RID, original$RID),
          identical(result$assignments$cluster, original$cluster),
          identical(unname(tools::md5sum(original_path)), original_hash),
          identical(readRDS(paths[[rid]])$assignments, original))
        audit[[rid]] <- data.frame(region_id = rid, n_reads = nrow(original),
          n_clusters = result$n_clusters, assignments_unchanged = TRUE,
          original_result_md5 = original_hash, nucleosome_footprints = nrow(nuc), tf_footprints = nrow(tf))
        data.table::fwrite(dplyr::bind_rows(audit), file.path(summary_dir, "plot_reproduction_validation.tsv"), sep = "\t")
      }
      saveRDS(paths, file.path(summary_dir, "fiberseq_result_paths.rds"))
      main_ids <- intersect(signature$regions$region_id, names(paths))
      main_paths <- setNames(file.path(dirname(paths[main_ids]), "clustering.rds"), main_ids)
      stopifnot(all(file.exists(main_paths)))
      saveRDS(main_paths, file.path(summary_dir, "result_paths.rds"))
      old_summary <- file.path(output_dir, "results", "summary tables", "run_summary.tsv")
      if (file.exists(old_summary)) file.copy(old_summary, file.path(summary_dir, "run_summary.tsv"), overwrite = TRUE)
      paths
    })
  } else {
    # Build the top-ten results when no saved result paths exist.
    result_paths <- local({
      seed <- 1L
      reuse_cache <- TRUE
      output_dir <- lcl_output_path(output_dir)
      stopifnot(nrow(regions) > 0L, all(regions$width == 2000L),
        all(regions$fisher_qvalue < .1), !anyDuplicated(regions$peak_name))
      summary_dir <- file.path(output_dir, "summary tables")
      results_dir <- output_dir
      haplotype_dir <- output_dir
      matrix_dir <- lcl_output_path(file.path(output_dir, "m6a_summary"))
      dir.create(summary_dir, recursive = TRUE, showWarnings = FALSE)
      # Both the main results and haplotype examples derive from this one selection.
      data.table::fwrite(regions, file.path(summary_dir, "selected_top10_regions.tsv"), sep = "\t")
      datasets <- lapply(seq_len(nrow(regions)), function(i) {
        assemble_region_m6a(sample_table = lcl_region_samples(regions[i, ], sample_table),
          region = regions[i, ], matrix_dir = matrix_dir, reuse = reuse_cache, full_span = TRUE)
      })
      sample_ids <- unique(unlist(strsplit(regions$contributing_samples, ",", fixed = TRUE)))
      used_samples <- sample_table[match(sample_ids, sample_table$sample_name), , drop = FALSE]
      phases <- cache_lcl_haplotags(used_samples, datasets, phasing_root,
        lcl_output_path(file.path(output_dir, "phase_summary")), reuse = reuse_cache)
      rm(datasets)
      main_paths <- run_lcl_clustering(regions, sample_table, matrix_dir, results_dir,
        window_size = 0L, k_neighbors = 10L, leiden_resolution = 1, leiden_seed = seed,
        reuse_cache = reuse_cache, phase_cache = phases)
      short_colors <- sample_palette(sample_table$sample_name)
      sample_colors <- setNames(unname(short_colors[sub("_.*$", "", sample_table$sample_name)]), sample_table$sample_name)
      footprint_summaries <- cache_footprint_tracks(regions,
        "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks",
        file.path(output_dir, "footprint_summary"), workers = 3L, reuse = reuse_cache)
      paths <- main_paths
      audit <- list()
      for (rid in names(main_paths)) {
        message("Top AS-FIRE Fiber-seq views: ", rid)
        result <- fiberseq_display_result(readRDS(main_paths[[rid]]))
        assignments <- result$assignments
        region_colors <- sample_colors[names(sample_colors) %in% assignments$sample_name]
        source_hash <- tools::md5sum(main_paths[[rid]])
        selected_samples <- lcl_region_samples(result$region, sample_table)
        out <- lcl_output_path(file.path(haplotype_dir, rid, "bin0", "k10", "resolution1"))
        dir.create(file.path(out, "tables"), recursive = TRUE, showWarnings = FALSE)
        focal <- result$focal
        result$markers <- focal_markers(result)
        footprints <- dplyr::bind_rows(
          extract_nucleosomes(selected_samples, result$region, assignments),
          region_footprints(footprint_summaries, result$region, assignments)$records)
        embedding <- fiberseq_umap(result, seed = seed)
        # Main and haplotype outputs have identical retained reads and main labels.
        save_all_plots(result, footprints, out, region_colors, embedding = embedding)
        paths[[rid]] <- file.path(out, "tables", "fiberseq_annotated_clustering.rds")
        saveRDS(result, paths[[rid]])
        stopifnot(identical(result$assignments, assignments),
                  identical(tools::md5sum(main_paths[[rid]]), source_hash),
                  all(assignments$allele_status == "phased_focal_genotype"),
                  all(assignments$focal_genotype %in% c("0|1", "1|0")))
        audit[[rid]] <- data.frame(region_id = rid, rank = result$region$rank,
          focal_snp = result$region$focal_snp, n_reads = nrow(assignments),
          n_samples = length(unique(assignments$sample_name)), n_clusters = result$n_clusters,
          main_haplotype_assignments_identical = TRUE, n_unresolved_plotted = 0L,
          n_ref_reads = sum(assignments$allele_label == paste0(focal$variant_id, ": ", focal$ref)),
          n_alt_reads = sum(assignments$allele_label == paste0(focal$variant_id, ": ", focal$alt)))
        data.table::fwrite(dplyr::bind_rows(audit), file.path(summary_dir, "fiberseq_validation.tsv"), sep = "\t")
      }
      saveRDS(paths, file.path(summary_dir, "fiberseq_result_paths.rds"))
      # A selection signature prevents plots-only rendering from loading the previous examples.
      saveRDS(list(regions = regions, sample_table = sample_table,
        view_params = list(seed = seed),
        asfire_file = asfire_file),
        file.path(summary_dir, "selection_signature.rds"))
      paths
    })
  }
} else {
  result_paths <- readRDS(file.path(output_dir, "summary tables", "fiberseq_result_paths.rds"))
}
stopifnot(all(regions$region_id %in% names(result_paths)))
result_paths <- result_paths[regions$region_id]
# Reject cached results that do not satisfy the current LCL cohort policy.
for (path in result_paths) {
  retained <- readRDS(path)$assignments
  stopifnot(nrow(retained) > 0L,
    all(c("haplotype", "focal_genotype", "allele_status") %in% names(retained)),
    all(retained$haplotype %in% c("HP1", "HP2")),
    all(retained$focal_genotype %in% c("0|1", "1|0")),
    all(retained$allele_status == "phased_focal_genotype"))
}
main_paths <- readRDS(file.path(output_dir, "summary tables", "result_paths.rds"))
```

```{r results-summary}
knitr::kable(data.table::fread(file.path(output_dir, "summary tables", "run_summary.tsv")))
validation_file <- file.path(output_dir, "summary tables", "plot_reproduction_validation.tsv")
if (!file.exists(validation_file)) validation_file <- file.path(output_dir, "summary tables", "fiberseq_validation.tsv")
knitr::kable(data.table::fread(validation_file))
```

# Visualization of results
- Accessibility is the fraction of retained cluster reads with an m6A call at an observed site over all the reads in the cluster
- Nucleosome occupancy is the fraction covered by a 130–160 bp footprint at each base, counting overlapping footprints on a read once

```{r gallery-helper, include = FALSE}
pdftopng <- "/software/xpdf-4.05-el8-x86_64/bin/pdftopng"
xpdfrc <- tempfile("xpdfrc-")
sans <- "/usr/share/fonts/liberation-sans/LiberationSans-"
mono <- "/usr/share/fonts/dejavu/DejaVuSansMono"
writeLines(c(
  paste0("fontFile Helvetica ", sans, "Regular.ttf"),
  paste0("fontFile Helvetica-Bold ", sans, "Bold.ttf"),
  paste0("fontFile Helvetica-Oblique ", sans, "Italic.ttf"),
  paste0("fontFile Helvetica-BoldOblique ", sans, "BoldItalic.ttf"),
  paste0("fontFile Times-Roman ", sans, "Regular.ttf"),
  paste0("fontFile Times-Bold ", sans, "Bold.ttf"),
  paste0("fontFile Times-Italic ", sans, "Italic.ttf"),
  paste0("fontFile Times-BoldItalic ", sans, "BoldItalic.ttf"),
  paste0("fontFile Courier ", mono, ".ttf"),
  paste0("fontFile Courier-Bold ", mono, "-Bold.ttf"),
  paste0("fontFile Courier-Oblique ", mono, "-Oblique.ttf"),
  paste0("fontFile Courier-BoldOblique ", mono, "-BoldOblique.ttf"),
  "fontFile Symbol /usr/share/fonts/dejavu/DejaVuSans.ttf",
  "fontFile ZapfDingbats /usr/share/fonts/dejavu/DejaVuSans.ttf"), xpdfrc)

inline_png <- function(path, label, dpi = 120) {
  stopifnot(file.exists(path), file.info(path)$size > 0)
  prefix <- tempfile("fig-")
  status <- system2(pdftopng, c("-cfg", shQuote(xpdfrc), "-q", "-r", dpi, "-f", 1, "-l", 1,
    shQuote(path), shQuote(prefix)))
  png <- paste0(prefix, "-000001.png")
  stopifnot(status == 0, file.exists(png))
  on.exit(unlink(png))
  cat(as.character(htmltools::tags$figure(
    htmltools::tags$img(
      src = base64enc::dataURI(file = png, mime = "image/png"),
      alt = label, loading = "lazy",
      style = "display:block;max-width:100%;height:auto;"),
    htmltools::tags$figcaption(label))), "\n\n")
  invisible(NULL)
}

figure_gallery <- function(name, title) {
  cat("\n## ", title, "\n\n", sep = "")
  for (rid in names(result_paths)) {
    cat("\n### ", rid, "\n\n", sep = "")
    plot_dir <- file.path(dirname(dirname(result_paths[[rid]])), "plots")
    pdf <- lcl_output_path(file.path(plot_dir, paste0(name, ".pdf")))
    stopifnot(file.exists(pdf))
    inline_png(pdf, paste(rid, title, sep = " — "))
    cat(as.character(htmltools::tags$p(htmltools::tags$a(
      href = report_url(pdf), paste("Download original PDF:", name)))), "\n\n")
  }
}
```

```{r gallery-haplotype-example, results='asis'}
figure_gallery("fiberseq_haplotype_example", "Haplotype overview")
```

```{r gallery-read-footprints, results='asis'}
figure_gallery("fig1_read_footprints", "Read-level footprints")
```

```{r gallery-occupancy-profiles, results='asis'}
figure_gallery("fig2_occupancy_by_cluster", "Occupancy profiles by cluster")
```


```{r gallery-m6a-footprint-heatmap, results='asis'}
figure_gallery("heatmap_m6a_footprints", "m6A and footprint heatmaps")
```

# KNN graph

- how similar individual reads are to each other based on their m6A footprint patterns
  - each dot is one read, an edge connects two reads if either selects the other as one of its 10 nearest neighbors. 
  - a read can have more than 10 edges if it is selected as a neighbor by multiple other reads

```{r gallery-knn-graph, results='asis'}
for (rid in names(result_paths)) {
  plot_dir <- lcl_output_path(file.path(dirname(dirname(result_paths[[rid]])), "plots"))
  graph_pdf <- file.path(plot_dir, "knn_graph.pdf")
  if (!file.exists(graph_pdf)) {
    result <- readRDS(result_paths[[rid]])
    dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
    ggplot2::ggsave(graph_pdf, plot_knn_graph(result), width = 12, height = 9, bg = "white")
  }
}
figure_gallery("knn_graph", "Manhattan KNN graphs colored by Leiden cluster")
```


# KNN graph colored by haplotype

Within each group of connected reads in the KNN graph, what fraction carries each lead SNP allele?

```{r knn-haplotype-summaries, include=FALSE}
knn_haplotype_views <- setNames(vector("list", length(result_paths)), names(result_paths))
for (rid in names(result_paths)) {
  result <- readRDS(result_paths[[rid]])
  result <- result[c("graph", "assignments", "params", "region", "focal")]
  summary <- knn_allele_summary(result)
  knn_haplotype_views[[rid]] <- list(result = result, summary = summary)
  table_dir <- lcl_output_path(dirname(result_paths[[rid]]))
  data.table::fwrite(summary$composition, file.path(table_dir, "knn_allele_composition.tsv"), sep = "\t")
  data.table::fwrite(summary$connectivity, file.path(table_dir, "knn_allele_connectivity.tsv"), sep = "\t")
  data.table::fwrite(summary$reads, file.path(table_dir, "knn_read_allele_connectivity.tsv"), sep = "\t")
}
```

```{r knn-graph-by-haplotype, results='asis'}
for (rid in names(knn_haplotype_views)) {
  result <- knn_haplotype_views[[rid]]$result
  plot_dir <- lcl_output_path(file.path(dirname(dirname(result_paths[[rid]])), "plots"))
  dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
  ggplot2::ggsave(file.path(plot_dir, "knn_graph_by_allele.pdf"),
    plot_knn_graph(result, color_by = "allele"), width = 12, height = 9, bg = "white")
}
figure_gallery("knn_graph_by_allele", "KNN graph colored by focal SNP allele")
```

```{r knn-haplotype-cluster-composition, results='asis'}
for (rid in names(knn_haplotype_views)) {
  summary <- knn_haplotype_views[[rid]]$summary
  plot_dir <- lcl_output_path(file.path(dirname(dirname(result_paths[[rid]])), "plots"))
  ggplot2::ggsave(file.path(plot_dir, "knn_allele_composition.pdf"),
    plot_knn_allele_composition(summary, paste(rid, "Focal SNP allele proportions", sep = "\n")),
    width = max(10, 0.7 * nlevels(summary$composition$group) + 3), height = 6, bg = "white")
}
figure_gallery("knn_allele_composition", "Focal SNP allele proportions within each Leiden cluster")
```

```{r knn-haplotype-connectivity, results='asis'}
for (rid in names(knn_haplotype_views)) {
  summary <- knn_haplotype_views[[rid]]$summary
  plot_dir <- lcl_output_path(file.path(dirname(dirname(result_paths[[rid]])), "plots"))
  ggplot2::ggsave(file.path(plot_dir, "knn_allele_connectivity.pdf"),
    plot_knn_allele_connectivity(summary, paste(rid, "Allele mixing in the KNN graph", sep = "\n")),
    width = 11, height = 7, bg = "white")
}
figure_gallery("knn_allele_connectivity", "Same- versus opposite-allele graph connectivity")
```



# PRDX1 and ALAS1 promoters: all reads, without haplotype information

Pool all 31 LCL samples, including samples without genotypes, and retain every
read spanning the complete promoter window. No haplotype, SNP, genotype or
allele filtering is applied, and reads are not downsampled. Clustering uses
the existing m6A Manhattan/Leiden helpers with `window_size = 0`, `k = 10`,
resolution 1 and seed 1, as in the selected AS-FIRE figures.

The canonical TSS BED encodes 20 bp intervals `[TSS - 10, TSS + 10)` in
zero-based coordinates. Use their centers (`start + 11` in one-based coordinates)
and the exactly 2,000 bp half-open window `[TSS0 - 1000, TSS0 + 1000)`.
Read-footprint and occupancy plots orient the coordinates by gene strand.
Dashed lines mark the TSS (0 bp on these relative axes), including each
cluster slice and m6A profile of the genomic heatmap. UMAPs and the pie chart
have no genomic coordinate axis.


```{r lcl-promoters-all-reads, results='hide'}
local({
  genes <- c("PRDX1", "ALAS1")
  tss_bed <- data.table::fread(
    "/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed",
    header = FALSE, col.names = c("chr", "start", "end", "annotation", "score", "strand"),
    data.table = FALSE)
  tss_bed$gene <- vapply(strsplit(tss_bed$annotation, ";", fixed = TRUE), `[`, character(1), 3L)
  selected <- tss_bed[tss_bed$gene %in% genes, , drop = FALSE]
  stopifnot(nrow(selected) == length(genes), !anyDuplicated(selected$gene),
    all(selected$end - selected$start == 20L))
  selected <- selected[match(genes, selected$gene), , drop = FALSE]
  tss0 <- selected$start + 10L
  promoters <- data.frame(region_id = genes, gene = genes, chr = selected$chr,
    start = tss0 - 1000L, end = tss0 + 1000L,
    analysis_start = tss0 - 999L, analysis_end = tss0 + 1000L,
    width = 2000L, tss = tss0 + 1L, strand = selected$strand,
    region_type = "promoter", annotation = paste(genes, "canonical promoter"))
  stopifnot(all(promoters$analysis_end - promoters$analysis_start + 1L == 2000L))

  # Read the complete sample table afresh, independently of AS-FIRE contributors.
  samples <- read.csv(
    "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv",
    stringsAsFactors = FALSE)[, c("sample_name", "fire_dir")]
  stopifnot(!anyDuplicated(samples$sample_name),
    !anyDuplicated(sub("_.*$", "", samples$sample_name)))
  sample_colors <- sample_palette(samples$sample_name)
  full_colors <- setNames(unname(sample_colors[sub("_.*$", "", samples$sample_name)]), samples$sample_name)
  root <- lcl_output_path(file.path(LCL_OUTPUT_ROOT, "test_promoters"))
  dir.create(root, recursive = TRUE, showWarnings = FALSE)
  # The existing footprint extractor requires a cache; remove it after plotting
  # so the only persistent outputs are the requested PDFs.
  cache_dir <- tempfile(".footprint-cache-", tmpdir = root)
  # Plot layout helpers may open a device before export; avoid an extra Rplots.pdf.
  grDevices::pdf(NULL)
  layout_device <- grDevices::dev.cur()
  tryCatch({
    caches <- cache_footprint_tracks(promoters,
      "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks",
      cache_dir, workers = 2L, reuse = FALSE)
    for (i in seq_len(nrow(promoters))) {
      region <- promoters[i, , drop = FALSE]
      dat <- assemble_region_m6a(sample_table = samples, region = region,
        full_span = TRUE, matrix_dir = NULL, reuse = FALSE)
      result <- leiden_manhattan_cluster(dat$met_mat, dat$rids_df,
        region$analysis_start, region$analysis_end, window_size = 0L,
        k_neighbors = 10L, resolution = 1, seed = 1L, impute_missing = FALSE)
      result$region <- region
      result$site_met_mat <- dat$met_mat
      result$assignments$original_RID <- dat$rids_df$original_RID[
        match(result$assignments$RID, dat$rids_df$RID)]
      stopifnot(identical(result$assignments$RID, dat$rids_df$RID),
        nrow(result$assignments) == sum(dat$qc$full_span_reads))
      message(region$gene, ": retained all ", nrow(result$assignments),
        " full-span reads from ", length(unique(result$assignments$sample_name)), " samples")
      footprints <- dplyr::bind_rows(
        extract_nucleosomes(samples, region, result$assignments),
        region_footprints(caches, region, result$assignments)$records)
      signals <- signal_profiles(result, footprints)
      plot_dir <- lcl_output_path(file.path(root, region$gene))
      dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

      embedding <- fiberseq_umap(result, seed = 1L)
      cluster_colors <- cluster_id_palette(levels(result$assignments$cluster))
      counts <- as.data.frame(table(result$assignments$cluster),
        responseName = "n_reads", stringsAsFactors = FALSE)
      names(counts)[1L] <- "cluster"
      counts$proportion <- counts$n_reads / sum(counts$n_reads)
      counts$label <- sprintf("%s (n=%d; %.1f%%)", counts$cluster,
        counts$n_reads, 100 * counts$proportion)
      # There is no standalone pie helper; compose it with the sourced UMAPs.
      pie <- ggplot2::ggplot(counts, ggplot2::aes(x = "", y = proportion, fill = cluster)) +
        ggplot2::geom_col(width = 1, color = "white",
          position = ggplot2::position_stack(reverse = TRUE)) +
        ggplot2::coord_polar(theta = "y") +
        ggplot2::scale_fill_manual(values = cluster_colors, breaks = counts$cluster, labels = counts$label) +
        ggplot2::labs(title = paste0("Cluster proportions: ", nrow(embedding), " reads"), fill = NULL) +
        ggplot2::theme_void() + ggplot2::theme(legend.text = ggplot2::element_text(size = 8))
      overview <- cowplot::plot_grid(
        plot_umap(embedding, "sample_name", full_colors, "UMAP: LCL sample",
          labels = sub("_.*$", "", names(full_colors)), show_legend = TRUE),
        plot_umap(embedding, "cluster", cluster_colors, "UMAP: cluster", show_legend = TRUE),
        pie, ncol = 3, rel_widths = c(1.3, 1, 1))
      ggplot2::ggsave(file.path(plot_dir, "fiberseq_haplotype_example.pdf"),
        overview, width = 20, height = 7, bg = "white")
      tss_caption <- paste0("Dashed line: canonical TSS ", region$chr, ":", region$tss, " (0 bp)")
      ggplot2::ggsave(file.path(plot_dir, "fig1_read_footprints.pdf"),
        plot_smf_reads(result, signals$records, sample_colors, include_haplotype = FALSE) +
          ggplot2::labs(caption = tss_caption),
        width = 12, height = max(7, 0.03 * nrow(result$assignments) + 3.5 + 0.3 * result$n_clusters),
        limitsize = FALSE)
      ggplot2::ggsave(file.path(plot_dir, "fig2_occupancy_by_cluster.pdf"),
        plot_signal_profile(signals$profiles, region, levels(result$assignments$cluster)) +
          ggplot2::labs(subtitle = "m6A calls and footprint occupancy; all full-span reads",
            caption = tss_caption),
        width = 10, height = 1.4 * result$n_clusters + 2.5, limitsize = FALSE)
      heatmap <- plot_genomic_cluster_heatmap(result, region, sample_colors,
        include_haplotype = FALSE, variants = NULL, show_cluster_profiles = TRUE)
      grDevices::pdf(file.path(plot_dir, "heatmap_m6a_footprints.pdf"),
        width = 14, height = 11 + 0.55 * result$n_clusters)
      tryCatch({
        ComplexHeatmap::draw(heatmap, newpage = FALSE)
        # Column centers occupy (index - 0.5) / width of each heatmap slice.
        tss_x <- (region$tss - region$analysis_start + 0.5) / region$width
        for (slice in seq_len(result$n_clusters)) {
          ComplexHeatmap::decorate_heatmap_body("m6A", {
            grid::grid.lines(x = c(tss_x, tss_x), y = c(0, 1),
              gp = grid::gpar(col = "#D55E00", lty = "dashed", lwd = 1.2))
          }, slice = slice)
        }
        for (cluster in levels(result$assignments$cluster)) {
          ComplexHeatmap::decorate_annotation(paste0(cluster, " m6A"), {
            grid::grid.lines(x = c(tss_x, tss_x), y = c(0, 1),
              gp = grid::gpar(col = "#D55E00", lty = "dashed", lwd = 1.2))
          })
        }
      }, finally = grDevices::dev.off())
    }
  }, finally = {
    grDevices::dev.off(layout_device)
    unlink(cache_dir, recursive = TRUE)
  })
})
```

#### Outputs

The final output location is
`/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan`.
Each of the ten region folders contains `bin0/k10/resolution1/plots` and `tables`

`tables` has saved clustering objects (`clustering.rds` and
`fiberseq_annotated_clustering.rds`)

Selection follows [Kevin's heterozygous AS-FIRE examples](https://kevinlkx.github.io/fiber-seq-analysis/example_figures_with_merged_samples.html#AS-fire_results).

```{r session}
sessionInfo()
```
