---
title: "Leiden + Manhattan clustering at GENCODE v46 canonical TSSs and custom regions"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(Matrix)
library(GenomicRanges)
library(data.table)
library(Rsamtools)   
library(igraph)
library(ComplexHeatmap)
library(cowplot)
```

# Functions

`extract_ft_region_reads()`, `extract_ft_read_info()`, `get_sparse_met_mat()`
and `filter_met_mat()` come from the topic model

```{r functions}
source("/project/spott/cshan/fiber-seq/code/topic_model/topic_modelling_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_functions.r")
source("/project/spott/cshan/fiber-seq/code/clustering_methods/Leiden_Manhattan/leiden_manhattan_plots.r")
```

# Settings

```{r settings}
options(scipen = 999)

sample_table <- tibble::tribble(
  ~sample_name, ~condition, ~timepoint,
  "LPS_0",      "LPS",       0L,
  "LPS_5",      "LPS",       5L,
  "LPS_10",     "LPS",       10L,
  "LPS_15",     "LPS",       15L
)
sample_names <- sample_table$sample_name[sample_table$condition == "LPS"]

ft_result_dir <- "/project/spott/cshan/fiber-seq/macrophage_project/FiberHMM/extract/ft_result_dir"

# The eight early-response genes used in this notebook (IL8 = CXCL8).
early_response_genes <- c(IL1B = "IL1B", TNF = "TNF", IL1A = "IL1A", PTGS2 = "PTGS2",
                          IL8 = "CXCL8", CXCL2 = "CXCL2", IL6 = "IL6", RELB = "RELB")

# GENCODE v46 Ensembl-canonical TSS: promoter windows are canonical TSS +/- tss_flank
gencode_tss_bed <- "/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed"
tss_flank <- 1000L

output_dir <- file.path(
  "/project/spott/cshan/fiber-seq/macrophage_project/clustering_TSS_methods/Leiden_Manhattan",
  paste0("TSS_", tss_flank)
)
summary_dir <- file.path(output_dir, "summary")
if (!dir.exists(summary_dir)) {
  dir.create(summary_dir, recursive = TRUE)
}

# Window ("feature") size used for the custom-region run. The previous window
# sweep is kept below as a comment for reference.
# window_sizes <- c(0, 10, 20, 30, 50)
window_size <- 0L

# Leiden / graph parameters

# k_neighbors: 
k_neighbors       <- 10

# leiden_resolution: Raise it for finer clusters, lower it for coarser ones
leiden_resolution <- 1
leiden_seed       <- 1

# Exponential-kernel width for the affinities: NULL = mean of the retained KNN
# distances (scale-free across window sizes). Set to the number of features to
# reproduce scikit-learn's laplacian_kernel default (gamma = 1 / n_features).
kernel_sigma <- NULL

# Paper step 2, off for now: with full-span reads there is nothing to impute.
impute_missing <- FALSE
impute_k       <- 5

# Rolling-mean width for the m6A fraction line plot only (1 = raw profiles).
# At window size 0 the per-site cluster means are spiky; e.g. 25 makes the
# nucleosome-scale structure readable without changing any clustering input.
line_smooth_k <- 1

set.seed(123)
```

# Regions

All eight early-response promoters listed in `early_response_genes`, using
GENCODE v46 canonical TSS +/- `tss_flank` bp, plus the two custom coordinate
regions. Coordinates are 1-based inclusive and are exported to
`summary/regions_bin0.tsv`. Results are written under `TSS_1000/` for this run.

```{r regions}
# Read gene names and ENSG directly from the canonical TSS annotation; the old
# GENCODE v49 promoter-table path is no longer present. The BED holds a 20-bp
# interval per gene,
# [TSS - 10, TSS + 10) in 0-based coordinates, so the 1-based TSS is start + 11.
gencode_tss <- read.delim(gencode_tss_bed, header = FALSE,
                          col.names = c("chr", "start", "end", "annotation", "score", "strand"))
gencode_tss$ensg <- sub("\\..*$", "", sub(";.*$", "", gencode_tss$annotation))
gencode_tss$gene <- vapply(strsplit(gencode_tss$annotation, ";", fixed = TRUE),
                          function(x) x[[3]], character(1))
gencode_tss$tss  <- gencode_tss$start + 11L
gencode_tss <- gencode_tss[gencode_tss$chr %in% paste0("chr", c(1:22, "X", "Y")) &
                            gencode_tss$gene %in% early_response_genes, ]
stopifnot(!anyDuplicated(gencode_tss$gene))
idx <- match(unname(early_response_genes), gencode_tss$gene)
stopifnot(!anyNA(idx))
promoters <- data.frame(
  gene = names(early_response_genes), gencode_name = unname(early_response_genes),
  ensg = gencode_tss$ensg[idx], chr = gencode_tss$chr[idx],
  gene_strand = gencode_tss$strand[idx], canonical_tss = gencode_tss$tss[idx]
)
promoters$start <- promoters$canonical_tss - tss_flank
promoters$end   <- promoters$canonical_tss + tss_flank
promoters[, c("gene", "gencode_name", "ensg", "chr", "gene_strand", "canonical_tss",
              "start", "end")]

custom <- data.frame(
  chr = c("chr4", "chr5"),
  start = c(112284963, 150068577),
  end = c(112285236, 150068792)
)
custom$region_id <- paste(custom$chr, custom$start, custom$end, sep = "_")

regions <- rbind(
  data.frame(chr = promoters$chr, start = promoters$start, end = promoters$end,
             region_id = promoters$gene),
  custom
)
write.table(promoters, file.path(summary_dir, "promoter_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
write.table(regions, file.path(summary_dir, paste0("regions_bin", window_size, ".tsv")),
            sep = "\t", quote = FALSE, row.names = FALSE)
regions
```


## Testing functions

### 1. assemble_region_m6
```{r assemble_region_m6}
# Pick one region to test
i <- 1

region_id    <- regions$region_id[i]
region_chr   <- regions$chr[i]
region_start <- regions$start[i]
region_end   <- regions$end[i]

cat(
  "Testing:", region_id, "\n",
  "Region:", region_chr, region_start, region_end, "\n"
)

dat <- assemble_region_m6a(
  sample_names = sample_names,
  region_chr   = region_chr,
  region_start = region_start,
  region_end   = region_end,
  ft_result_dir = ft_result_dir,
  verbose = TRUE
)

head(dat$reads)
dim(dat$reads)
colnames(dat$reads)
```

### 2. bin_read_matrix

```{r bin_read_matrix}
binned50 <- bin_read_matrix(
  met_mat      = filter_met_mat(dat$met_mat, verbose = TRUE),
  region_start = region_start,
  region_end   = region_end,
  window_size  = window_size
)

names(binned50)

dim(binned50$mat)
binned50$mat[1:5, 1:5]
head(binned50$window_anno)
```


### 3. manhattan_knn_graph

```{r manhattan_knn_graph}
feat <- binned50$mat

knn <- manhattan_knn_graph(
  mat         = feat,
  k_neighbors = k_neighbors,   # 10
  sigma       = kernel_sigma,  
  verbose     = TRUE
)

knn$dist[1:5, 1:5]

edge_df <- igraph::as_data_frame(knn$graph, what = "edges")
head(edge_df)
```

### 4. leiden_partition
```{r leiden_partition}
part <- leiden_partition(
  knn$graph,
  resolution = leiden_resolution,
  seed = leiden_seed
)

# Leiden cluster for each read
membership <- part$membership

table(membership)

set.seed(1)

plot(
  knn$graph,
  vertex.color = membership,
  vertex.size = 4,
  vertex.label = NA,
  edge.width = 0.3,
  layout = layout_with_fr(knn$graph),
  main = "KNN graph colored by Leiden cluster"
)
```

# Clustering at each region

Run the same Leiden + Manhattan workflow on the explicit regions above, keeping
the same single `window_size` throughout. Outputs per region are written under
`<region_id>/bin<size>/k<n_clusters>/`, with TSV files in `tables/`, PDFs in
`plots/`, and the RDS result directly in the `k<n_clusters>/` directory.
Run-level TSVs are saved in `summary/`. Each `knn_edges_<region_id>.tsv` contains
one row per undirected edge in the union kNN graph, with endpoint read IDs
(`from`, `to`), the original `manhattan_distance`, and the affinity `weight`
used by Leiden (`exp(-manhattan_distance / sigma)`). Reciprocal neighbor pairs
are saved once; zero-distance edges are retained. At bin 0 the feature columns
are genomic site positions and values are the original 0/1 m6A calls.

```{r cluster-regions, fig.width=9, fig.height=9}
res <- leiden_manhattan_cluster_regions(
  sample_names = sample_names,
  regions = regions,
  ft_result_dir = ft_result_dir,
  window_size = window_size,
  k_neighbors = k_neighbors,
  resolution = leiden_resolution,
  impute_missing = impute_missing,
  impute_k = impute_k,
  sigma = kernel_sigma,
  seed = leiden_seed,
  sample_table = sample_table
)

for (region_id in names(res)) {
  region_res <- res[[region_id]]
  lab <- paste0("bin", window_size)
  k_dir <- file.path(output_dir, region_id, lab, paste0("k", region_res$n_clusters))
  tables_dir <- file.path(k_dir, "tables")
  plots_dir <- file.path(k_dir, "plots")
  if (!dir.exists(tables_dir)) dir.create(tables_dir, recursive = TRUE)
  if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

  region_res$assignments$timepoint <- sample_table$timepoint[
    match(region_res$assignments$sample_name, sample_table$sample_name)
  ]
  res[[region_id]] <- region_res

  edges <- igraph::as_data_frame(region_res$graph, what = "edges")
  stopifnot("manhattan_distance" %in% names(edges),
            nrow(edges) == igraph::ecount(region_res$graph),
            all(is.finite(edges$manhattan_distance)),
            all(edges$manhattan_distance >= 0),
            all(edges$from %in% region_res$assignments$RID),
            all(edges$to %in% region_res$assignments$RID),
            identical(rownames(region_res$feat_mat), region_res$assignments$RID))
  write.table(edges,
              file.path(tables_dir, paste0("knn_edges_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)

  saveRDS(region_res, file.path(k_dir, paste0("leiden_manhattan_", region_id, ".rds")))
  write.table(region_res$assignments,
              file.path(tables_dir, paste0("read_cluster_assignments_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(cluster = rownames(region_res$profiles), region_res$profiles,
                         check.names = FALSE),
              file.path(tables_dir, paste0("cluster_profiles_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(region_res$window_anno,
              file.path(tables_dir, paste0("window_annotation_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(RID = rownames(region_res$feat_mat),
                         sample_name = region_res$assignments$sample_name,
                         timepoint = region_res$assignments$timepoint,
                         cluster = region_res$assignments$cluster,
                         region_res$feat_mat, check.names = FALSE),
              file.path(tables_dir, paste0("feature_matrix_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(region_res$assignments %>% dplyr::count(cluster, sample_name, timepoint),
              file.path(tables_dir, paste0("cluster_timepoint_counts_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)
  write.table(data.frame(parameter = names(region_res$params),
                         value = vapply(region_res$params, as.character, character(1))),
              file.path(tables_dir, paste0("run_parameters_", region_id, ".tsv")),
              sep = "\t", quote = FALSE, row.names = FALSE)

  gg <- plot_cluster_met_profiles(region_res, region_res$site_met_mat, main = region_id)
  ggsave(file.path(plots_dir, paste0("cluster_met_profiles_", region_id, ".pdf")), gg,
         width = 6, height = 1.4 * region_res$n_clusters + 1.4, limitsize = FALSE)

  gg <- plot_cluster_composition(region_res, main = region_id)
  ggsave(file.path(plots_dir, paste0("cluster_composition_", region_id, ".pdf")), gg,
         width = 7, height = 8)
  gg <- plot_cluster_structure(region_res, main = region_id)
  ggsave(file.path(plots_dir, paste0("cluster_structure_", region_id, ".pdf")), gg,
         width = 9, height = 3)

  pdf(file.path(plots_dir, paste0("heatmap_", region_id, ".pdf")), width = 8, height = 9)
  ComplexHeatmap::draw(plot_cluster_heatmap(region_res, main = region_id))
  dev.off()

  gg <- plot_met_fraction_lines(region_res, main = region_id, smooth_k = line_smooth_k)
  ggsave(file.path(plots_dir, paste0("met_fraction_lines_", region_id, ".pdf")), gg,
         width = 8, height = 1.4 * region_res$n_clusters + 1, limitsize = FALSE)
}

names(res)
```

# Cluster sizes and timepoint composition across regions

`n_clusters` is the number of Leiden clusters for that region at the current
`window_size`; `n` counts reads.

```{r cluster-summary}
cluster_summary <- dplyr::bind_rows(lapply(names(res), function(region_id) {
  region_res <- res[[region_id]]
  region_res$assignments %>%
    dplyr::count(cluster) %>%
    dplyr::mutate(region_id = region_id,
                  bins = paste0("bin", window_size),
                  n_clusters = region_res$n_clusters,
                  .before = 1)
}))
write.table(cluster_summary, file.path(summary_dir, "cluster_summary_all_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
cluster_summary %>% tidyr::pivot_wider(names_from = cluster, values_from = n)

timepoint_summary <- dplyr::bind_rows(lapply(names(res), function(region_id) {
  region_res <- res[[region_id]]
  region_res$assignments %>%
    dplyr::count(cluster, sample_name, timepoint) %>%
    dplyr::mutate(region_id = region_id,
                  bins = paste0("bin", window_size),
                  .before = 1)
}))
write.table(timepoint_summary,
            file.path(summary_dir, "cluster_timepoint_summary_all_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

run_summary <- dplyr::bind_rows(lapply(names(res), function(region_id) {
  p <- res[[region_id]]$params
  data.frame(region_id = region_id, bins = paste0("bin", window_size),
             chr = p$region_chr, start = p$region_start, end = p$region_end,
             n_reads = p$n_reads, n_features = p$n_features, k_eff = p$k_eff,
             n_edges = igraph::ecount(res[[region_id]]$graph),
             resolution = p$resolution, sigma = p$sigma,
             mean_knn_dist = p$mean_knn_dist, quality = p$quality,
             n_clusters = res[[region_id]]$n_clusters)
}))
write.table(run_summary, file.path(summary_dir, "run_summary_all_regions.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
run_summary
```

# Session info

```{r session-info}
sessionInfo()
```
