---
title: "LCL m6A clustering and footprint profiles"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    self_contained: true
params:
  plots_only: false
  regenerate_plots: true
---

<style>
.main-container { max-width: 100%; padding: 0 24px; }
.lcl-gallery { overflow-x: auto; margin-bottom: 2em; }
.lcl-gallery table { width: 100%; table-layout: fixed; min-width: 1250px; }
.lcl-gallery th, .lcl-gallery td { padding: 8px; border: 1px solid #ddd; vertical-align: top; overflow-wrap: anywhere; }
.lcl-gallery th:first-child { width: 160px; }
.lcl-preview { padding: 0; border: 0; background: white; width: 100%; cursor: zoom-in; }
.lcl-preview img { width: 100%; height: auto; }
#lcl-zoom { max-width: 96vw; max-height: 96vh; overflow: auto; }
#lcl-zoom img { display: block; width: auto; max-width: 90vw; height: auto; }
#lcl-zoom::backdrop { background: rgba(0,0,0,.65); }
</style>
<dialog id="lcl-zoom"><form method="dialog"><button>Close preview</button></form><img alt="Enlarged plot preview"></dialog>

## Setup and code organization

1. `leiden_LCL_functions.R`: construct genomic windows, m6A matrices
2. `process_LCL_footprints.R` computes occupancy and aggregate profiles
3. `plot_clustering_LCL.R`: plot heatmap and aggregate profiles for footprints and nucleosomes
4. `leiden_manhattan_functions.r`:  calculate manhattan distance and run leiden clustering


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
options(scipen = 999)
suppressPackageStartupMessages({
  library(dplyr)
  library(Matrix)
  library(GenomicRanges)
  library(Rsamtools)
  library(igraph)
  library(ComplexHeatmap)
  library(ggplot2)
  library(cowplot)
})
code_dir <- "/project/spott/cshan/fiber-seq/code/clustering_TSS_methods/Leiden_Manhattan/LCL"
source(file.path(code_dir, "leiden_LCL_functions.R"))
source("/project/spott/cshan/fiber-seq/code/nucleosome_positioning/LCL/process_LCL_footprints.R")
source(file.path(code_dir, "plot_clustering_LCL.R"))
data.table::setDTthreads(2L)
```

## Regions and samples and parameters

Heatmaps are in genomic coordinates, footprint and profile plots use TSS relative coordinates

```{r settings}
sample_csv <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/Data/LCL_sample_metatable_merged_samples_31samples.csv"
tss_bed <- "/project/spott/cshan/annotations/gencodev46_Ensembl_canonical_TSS.bed"
tracks_dir <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_tracks"
matrix_dir <- "/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan/m6a_matrix"
output_dir <- "/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan/results"
include_haplotype <- TRUE
haplotype_output_dir <- "/project/spott/cshan/fiber-seq/LCL_project/Leiden_manhattan/haplotype"
phasing_root <- "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples"
tss_flank <- 1000L
window_size <- 0L
k_neighbors <- 10L
leiden_resolution <- 1
leiden_seed <- 1L
kernel_sigma <- NULL
footprint_workers <- 2L
reuse_cache <- TRUE
promoter_genes <- c("EIF4A2", "RPS17", "NPM1")
coordinate_regions <- tibble::tribble(
  ~region_id, ~chr, ~start, ~end, ~annotation,
  "rs4953223", "chr2", 45289157L, 45291257L, "dsQTL: rs4953223; NFKB example",
  "rs1530562", "chr2", 119291054L, 119291154L, "dsQTL: rs1530562",
  "rs6513945", "chr20", 44683645L, 44683745L, "dsQTL: rs6513945; CTCF example 1",
  "rs749579", "chr2", 86220700L, 86221700L, "dsQTL: rs749579",
  "rs7603346", "chr2", 212990848L, 212991848L, "dsQTL: rs7603346",
  "chr3_102233378_102238378", "chr3", 102233378L, 102238378L, "chr3:102233378-102238378 region"
)
regions <- construct_lcl_windows(coordinate_regions, promoter_genes, tss_bed, tss_flank)
sample_table <- read.csv(sample_csv, stringsAsFactors = FALSE)
stopifnot(nrow(sample_table) == 31L, !anyDuplicated(sample_table$sample_name),
          all(grepl("^AL[0-9]+(AL[0-9]+)*_", sample_table$sample_name)))
sample_table$sample_label <- sub("_.*$", "", sample_table$sample_name)
sample_colors <- lcl_sample_palette(sample_table$sample_name)
dir.create(file.path(output_dir, "summary tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(matrix_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(haplotype_output_dir, "summary tables"), recursive = TRUE, showWarnings = FALSE)
data.table::fwrite(regions, file.path(output_dir, "summary tables", "regions.tsv"), sep = "\t")
data.table::fwrite(sample_table, file.path(output_dir, "summary tables", "samples.tsv"), sep = "\t")
data.table::fwrite(data.frame(sample = names(sample_colors), color = unname(sample_colors)),
                   file.path(output_dir, "summary tables", "sample_colors.tsv"), sep = "\t")
input_paths <- unlist(lapply(unique(regions$chr), function(chromosome) {
  unlist(lapply(c("m6a", "nuc"), function(feature) {
    paths <- lcl_extracted_path(sample_table, chromosome, feature)
    c(paths, paste0(paths, ".tbi"))
  }))
}))
data.table::fwrite(lcl_source_signature(c(sample_csv, tss_bed, input_paths)),
                   file.path(output_dir, "summary tables", "input_manifest.tsv"), sep = "\t")
regions
```

## Pooled m6A clustering

All 31 samples are pooled with fully spanning reads

  * bin = 0, k = 10, resolution = 1

```{r clustering}
if (params$plots_only) {
  result_paths <- readRDS(file.path(output_dir, "summary tables", "result_paths.rds"))
} else {
  result_paths <- run_lcl_clustering(regions, sample_table, matrix_dir, output_dir,
    window_size, k_neighbors, leiden_resolution, leiden_seed, kernel_sigma, reuse_cache)
}
data.table::fread(file.path(output_dir, "summary tables", "run_summary.tsv"))
```

## Footprint processing

130-160 bp nucleosomes and TF footprints of 10-30, 40-60 and 60-80 bp

```{r footprints}
if (!params$plots_only) process_lcl_footprints(result_paths, sample_table, regions,
  tracks_dir, output_dir, workers = footprint_workers, reuse = reuse_cache)
```

## Pooled plots

- `heatmap_m6a.pdf`: m6A only
- `fig1_read_footprints.pdf`: nucleosomes and the selected TF sizes, with the same m6A-defined clusters 
- `aggregate_profile.pdf`: all features combined barplot
- `methylation_proportion_by_cluster.pdf`: m6A calls divided by the number of reads times the number of observed sites,

```{r pooled-plots, include=FALSE}
if (params$regenerate_plots) for (path in result_paths) {
  result <- readRDS(path)
  message("Plotting pooled region: ", result$region$region_id)
  footprints <- data.table::fread(file.path(dirname(path), "footprint_positions.tsv.gz"), data.table = FALSE)
  save_lcl_plots(result, footprints, sample_colors, dirname(dirname(path)))
}
```


```{r pooled-gallery, results='asis', echo=FALSE}
lcl_report_gallery(result_paths)
```

<!--
## Haplotype annotation and plots

Haplotype computation and display are temporarily disabled in this report.

HP1, HP2 and unphased reads retain the pooled m6A assignments and ordering.
Haplotype identifiers are local to each sample/phase block. HP1 is light blue,
HP2 light yellow. VCF SNP positions, REF/ALT bases and per-cluster m6A profiles
annotate `heatmap_m6a_footprints.pdf`; despite its filename, its matrix contains
only m6A. Fig 1 contains only protected footprints. Fig 2 uses lines for all
five tracks; `aggregate_profile.pdf` uses footprint bars with an m6A line.
The methylation summary barplot is also written for each annotated region.

```{r haplotype-plots, eval=FALSE, echo=FALSE}
if (include_haplotype) {
  if (params$plots_only) {
    haplotype_paths <- setNames(vapply(result_paths, function(path) {
      result <- readRDS(path)
      file.path(haplotype_output_dir, result$region$region_id, "pooled",
        paste0("bin", result$params$window_size), paste0("k", result$params$k_neighbors),
        paste0("resolution", result$params$resolution), "tables", "clustering.rds")
    }, character(1)), names(result_paths))
  } else {
    haplotype_paths <- annotate_lcl_results(result_paths, sample_table, phasing_root,
      haplotype_output_dir, reuse_cache)
  }
  if (params$regenerate_plots) for (rid in names(haplotype_paths)) {
    result <- readRDS(haplotype_paths[[rid]])
    message("Plotting haplotype annotations: ", rid)
    footprints <- data.table::fread(file.path(dirname(result_paths[[rid]]), "footprint_positions.tsv.gz"), data.table = FALSE)
    save_lcl_plots(result, footprints, sample_colors, dirname(dirname(haplotype_paths[[rid]])), include_haplotype = TRUE)
  }
}
```

```{r haplotype-gallery, eval=FALSE, echo=FALSE}
if (include_haplotype) lcl_report_gallery(haplotype_paths, include_haplotype = TRUE)
```

-->

## Output index and session

Each region/run contains `plots/` for PDFs and `tables/` for tabular outputs
and saved clustering objects. Shared tables and file-location indexes are in
`summary tables/` under both `results/` and `haplotype/`.

```{r output-index, include=FALSE}
write_lcl_result_index(output_dir, include_haplotype = FALSE)
```

```{r session}
sessionInfo()
```
